---
title: "Add cell states and remove contamination"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "Add_cell_states_remove_contamination"

# Load libraries
library(here)
library(dplyr)
library(Seurat)
library(qs2)

#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```

## Load dataset

Load the complete DCM object generated in QC_integration_annotation.

```{r Load data, eval=TRUE, echo = TRUE, results= "hide"}
DCM_RV <- qs_read(here::here("output", "QC_integration_annotation", "DCM_RV.qs2"), nthreads = 32)
```

```{r cell numbers before removal, eval=TRUE}
dim(DCM_RV)
table(DCM_RV$cell_type, useNA = "ifany")
```

## Load cell state annotations and contamination lists

```{r Load cell states, eval=TRUE}
CM_cell_states <- read.csv(here::here("output", "Cardiomyocytes_subclustering", "CM_cell_states.csv"), stringsAsFactors = FALSE)
FB_cell_states <- read.csv(here::here("output", "Fibroblasts_subclustering", "FB_cell_states.csv"), stringsAsFactors = FALSE)
EC_cell_states <- read.csv(here::here("output", "Endothelial_cells_subclustering", "EC_cell_states.csv"), stringsAsFactors = FALSE)
MC_cell_states <- read.csv(here::here("output", "Mural_cells_subclustering", "MC_cell_states.csv"), stringsAsFactors = FALSE)
MP_cell_states <- read.csv(here::here("output", "Macrophages_subclustering", "MP_cell_states.csv"), stringsAsFactors = FALSE)
LC_cell_states <- read.csv(here::here("output", "Lymphocytes_subclustering", "LC_cell_states.csv"), stringsAsFactors = FALSE)

cell_states <- bind_rows(CM_cell_states, FB_cell_states, EC_cell_states, MC_cell_states, MP_cell_states, LC_cell_states)
```

```{r Load contamination, eval=TRUE}
FB_contamination <- read.csv(here::here("output", "Fibroblasts_subclustering", "FB_cell_states_contamination.csv"), stringsAsFactors = FALSE)
EC_contamination <- read.csv(here::here("output", "Endothelial_cells_subclustering", "EC_cell_states_contamination.csv"), stringsAsFactors = FALSE)
MP_contamination <- read.csv(here::here("output", "Macrophages_subclustering", "MP_cell_states_contamination.csv"), stringsAsFactors = FALSE)
LC_contamination <- read.csv(here::here("output", "Lymphocytes_subclustering", "LC_cell_states_contamination.csv"), stringsAsFactors = FALSE)

cell_states_contamination <- bind_rows(FB_contamination, EC_contamination, MP_contamination, LC_contamination)

table(cell_states_contamination$cell_state)
```

## Check cell annotations

```{r Check cell annotations, eval=TRUE}
if (!all(c("cell", "cell_state") %in% colnames(cell_states))) stop("The cell state files do not contain the expected cell and cell_state columns")
if (!all(c("cell", "cell_state") %in% colnames(cell_states_contamination))) stop("The contamination files do not contain the expected cell and cell_state columns")

if (anyDuplicated(cell_states$cell) > 0) stop("Duplicated cell barcodes were found in the cell state files")
if (anyDuplicated(cell_states_contamination$cell) > 0) stop("Duplicated cell barcodes were found in the contamination files")
if (any(cell_states$cell %in% cell_states_contamination$cell)) stop("Some cells are present in both the final cell state files and the contamination files")

missing_cells <- setdiff(cell_states$cell, colnames(DCM_RV))
missing_contamination <- setdiff(cell_states_contamination$cell, colnames(DCM_RV))

if (length(missing_cells) > 0) stop(paste(length(missing_cells), "cells from the cell state files are not present in DCM_RV"))
if (length(missing_contamination) > 0) stop(paste(length(missing_contamination), "cells from the contamination files are not present in DCM_RV"))

length(cell_states$cell)
length(cell_states_contamination$cell)
```

## Remove contamination

```{r Remove contamination, eval=TRUE}
DCM_RV <- subset(DCM_RV, cells = setdiff(colnames(DCM_RV), cell_states_contamination$cell))

dim(DCM_RV)
table(DCM_RV$cell_type, useNA = "ifany")
```

Check that all remaining cells from the subclustered cell types have a cell state annotation.

```{r Check final cell state coverage, eval=TRUE}
subclustered_cells <- rownames(DCM_RV@meta.data)[DCM_RV$cell_type %in% c("Cardiomyocytes", "Fibroblasts", "Endothelial cells", "Pericytes", "Smooth muscle cells", "Macrophages", "T lymphocytes", "B lymphocytes")]

cells_without_cell_state <- setdiff(subclustered_cells, cell_states$cell)

if (length(cells_without_cell_state) > 0) stop(paste(length(cells_without_cell_state), "cells from subclustered cell types do not have a cell state annotation"))
```

## Add cell state annotations

Cells from cell types that were not subclustered are left as NA.

```{r Add cell states, eval=TRUE}
DCM_RV$cell_state <- NA_character_
DCM_RV@meta.data[cell_states$cell, "cell_state"] <- cell_states$cell_state
```

```{r Final cell numbers, eval=TRUE}
dim(DCM_RV)

table(DCM_RV$cell_type, useNA = "ifany")

table(DCM_RV$cell_state, useNA = "ifany")

table(DCM_RV$cell_type, DCM_RV$cell_state, useNA = "ifany")
```

## Save final objects

```{r save DCM_RV cell states, eval = TRUE}
qs_save(DCM_RV,
        file = here::here(output_dir_data, "DCM_RV_cell_states.qs2"),
        nthreads = 32
        )
```

```{r save DCM_RV cell states RNA, eval = TRUE}
DCM_RV_RNA <- DCM_RV
DCM_RV_RNA@assays[["SCT"]] <- NULL
DefaultAssay(DCM_RV_RNA) <- "RNA"
qs_save(DCM_RV_RNA,
        file = here::here(output_dir_data, "DCM_RV_cell_states_RNA.qs2"),
        nthreads = 32
        )
rm(DCM_RV_RNA)
```
