---
title: "Analysis of matrifibrocyte markers in fibroblasts from Amrute et al., Nature CVR 2024"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
current_file <- "Amrute_2024_matrifibrocyte_markers"

library(here)
library(Matrix)
library(Seurat)
library(scCustomize)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(stringr)
library(ggplot2)
library(ggpubr)
library(patchwork)
library(edgeR)
library(limma)
library(reactable)
library(scales)
library(RColorBrewer)
library(harmony)
library(clustree)
library(qs2)
library(broom)

output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"), recursive = TRUE)
output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs, recursive = TRUE)

data_dir <- here::here("data", "Heart_datasets", "Amrute_2024")
metadata_file <- file.path(data_dir, "GSE217494_cell.metadata.csv")

matrifibrocyte_genes <- c("COMP", "CILP", "CHAD", "THBS4")
min_pb_libsize <- 10000
min_fibroblast_cells <- 100

etiology_levels <- c("Donor", "ICM", "NICM", "AMI")
etiology_cols <- c(
  "Donor" = "grey85",
  "ICM" = "#004C8C",
  "NICM" = "#A63A2A",
  "AMI" = "#33A02C"
)

disease_cols <- c("HC" = "grey85", "Disease" = "#A63A2A")
dataset_cols <- c("Amrute" = "#FA95C0")

cell_type_cols <- c(
  "#d05050", "#33A02C", "#1F78B4", "#CAB2D6", "#FF7F00", "#B2DF8A",
  "#B15928", "#FFD000", "#FA95C0", "#FDBF6F", "#6A3D9A", "#7aCDF9", "#E6E600",
  "#8DD3C7", "#FB8072", "#80B1D3", "#BC80BD", "#CCEBC5", "#FFED6F",
  "#A6CEE3", "#F781BF"
)
```

## Load and prepare metadata

```{r load metadata}
cell_meta <- readr::read_csv(metadata_file, show_col_types = FALSE)

required_cols <- c("bc", "sample", "celltype", "HF.etiology")
missing_cols <- setdiff(required_cols, colnames(cell_meta))
if (length(missing_cols) > 0) {
  stop("Missing required metadata columns: ", paste(missing_cols, collapse = ", "))
}

cell_meta <- cell_meta %>%
  dplyr::mutate(
    bc = as.character(bc),
    sample = as.character(sample),
    celltype = as.character(celltype),
    HF.etiology = as.character(HF.etiology),
    HF.etiology = dplyr::case_when(
      HF.etiology %in% c("Donor", "donor", "DONOR") ~ "Donor",
      HF.etiology %in% c("ICM", "ischemic", "Ischemic") ~ "ICM",
      HF.etiology %in% c("NICM", "non-ischemic", "Non-ischemic") ~ "NICM",
      HF.etiology %in% c("AMI", "MI", "acute myocardial infarction") ~ "AMI",
      TRUE ~ HF.etiology
    ),
    HF.etiology = factor(HF.etiology, levels = etiology_levels),
    Group = HF.etiology,
    Disease_status = dplyr::if_else(as.character(HF.etiology) == "Donor", "HC", "Disease"),
    Disease_status = factor(Disease_status, levels = c("HC", "Disease")),
    Dataset = factor("Amrute", levels = "Amrute"),
    Sample = sample
  )

cell_meta %>%
  dplyr::count(HF.etiology, celltype, sort = TRUE) %>%
  reactable::reactable()
```

## Helper functions to read 10x-style MTX files

The dataset is CITE-seq, but this script keeps only RNA / gene-expression rows. If the `features.tsv.gz` file contains a feature-type column, the script keeps rows annotated as `Gene Expression`; otherwise it keeps all rows.

```{r helper functions}
read_one_gse217494_sample <- function(sample_id, data_dir, cell_meta) {
  mtx_file <- file.path(data_dir, paste0("GSE217494_", sample_id, ".matrix.mtx.gz"))
  feature_file <- file.path(data_dir, paste0("GSE217494_", sample_id, ".features.tsv.gz"))
  barcode_file <- file.path(data_dir, paste0("GSE217494_", sample_id, ".barcodes.tsv.gz"))

  if (!file.exists(mtx_file) || !file.exists(feature_file) || !file.exists(barcode_file)) {
    stop("Missing one or more files for ", sample_id)
  }

  message("Reading ", sample_id)

  mat <- Matrix::readMM(mtx_file)
  mat <- methods::as(mat, "dgCMatrix")

  features <- readr::read_tsv(feature_file, col_names = FALSE, show_col_types = FALSE)
  barcodes <- readr::read_tsv(barcode_file, col_names = FALSE, show_col_types = FALSE)[[1]]

  if (nrow(features) != nrow(mat)) {
    stop(sample_id, ": number of features does not match matrix rows.")
  }

  if (length(barcodes) != ncol(mat)) {
    stop(sample_id, ": number of barcodes does not match matrix columns.")
  }

  if (ncol(features) >= 3 && any(features[[3]] == "Gene Expression")) {
    rna_rows <- which(features[[3]] == "Gene Expression")
  } else {
    rna_rows <- seq_len(nrow(features))
  }

  gene_symbols <- if (ncol(features) >= 2) features[[2]] else features[[1]]
  gene_symbols <- as.character(gene_symbols)
  gene_symbols[is.na(gene_symbols) | gene_symbols == ""] <- as.character(features[[1]][is.na(gene_symbols) | gene_symbols == ""])

  mat <- mat[rna_rows, , drop = FALSE]
  rownames(mat) <- make.unique(gene_symbols[rna_rows])

  barcodes <- as.character(barcodes)
  bc_for_metadata <- ifelse(
    grepl("-1$", barcodes),
    sub("-1$", paste0("-", sample_id), barcodes),
    paste0(barcodes, "-", sample_id)
  )

  colnames(mat) <- bc_for_metadata

  keep_cells <- intersect(colnames(mat), cell_meta$bc)
  if (length(keep_cells) == 0) {
    warning(sample_id, ": no barcodes matched metadata.")
    return(NULL)
  }

  mat[, keep_cells, drop = FALSE]
}

get_p_label <- function(p) {
  if (is.na(p)) return("p = NA")
  if (p < 0.001) return("p < 0.001")
  paste0("p = ", formatC(p, format = "f", digits = 3))
}

get_fdr_label <- function(p) {
  if (is.na(p)) return(NA_character_)
  if (p < 0.001) return("FDR < 0.001")
  paste0("FDR = ", formatC(p, format = "f", digits = 3))
}
```

## Read RNA counts and create Seurat object

```{r read counts create seurat}
sample_ids_from_files <- list.files(
  data_dir,
  pattern = "^GSE217494_sample[0-9]+\\.matrix\\.mtx\\.gz$",
  full.names = FALSE
) %>%
  stringr::str_match("^GSE217494_(sample[0-9]+)\\.matrix\\.mtx\\.gz$") %>%
  .[, 2] %>%
  unique() %>%
  sort()

sample_ids_from_metadata <- unique(cell_meta$sample)
sample_ids <- intersect(sample_ids_from_files, sample_ids_from_metadata)

length(sample_ids)
sample_ids

counts_list <- lapply(sample_ids, function(sid) {
  read_one_gse217494_sample(
    sample_id = sid,
    data_dir = data_dir,
    cell_meta = cell_meta
  )
})
names(counts_list) <- sample_ids
counts_list <- counts_list[!vapply(counts_list, is.null, logical(1))]

common_genes <- Reduce(intersect, lapply(counts_list, rownames))
length(common_genes)

counts_all <- do.call(
  cbind,
  lapply(counts_list, function(x) x[common_genes, , drop = FALSE])
)

metadata_all_matched <- cell_meta %>%
  dplyr::filter(bc %in% colnames(counts_all)) %>%
  dplyr::arrange(match(bc, colnames(counts_all))) %>%
  as.data.frame()

counts_all <- counts_all[, metadata_all_matched$bc, drop = FALSE]
rownames(metadata_all_matched) <- metadata_all_matched$bc

stopifnot(identical(colnames(counts_all), rownames(metadata_all_matched)))

amrute <- CreateSeuratObject(
  counts = counts_all,
  meta.data = metadata_all_matched,
  project = "Amrute_2024"
)

amrute$percent.mt <- PercentageFeatureSet(amrute, pattern = "^MT-")
amrute$Dataset <- factor(amrute$Dataset, levels = "Amrute")
amrute$Group <- factor(amrute$Group, levels = etiology_levels)
amrute$HF.etiology <- factor(amrute$HF.etiology, levels = etiology_levels)
amrute$Disease_status <- factor(amrute$Disease_status, levels = c("HC", "Disease"))

amrute@meta.data %>%
  dplyr::count(HF.etiology, celltype, sort = TRUE) %>%
  reactable::reactable()
```

```{r save full seurat object, eval=FALSE}
qs_save(
  amrute,
  file = here::here(output_dir_data, "amrute_full.qs2"),
  nthreads = 32
)
```

## Subset fibroblasts

```{r subset fibroblasts}
amrute_fb <- subset(amrute, subset = celltype == "Fibroblast")

rm(amrute, counts_all, counts_list)
gc()

amrute_fb <- RenameCells(amrute_fb, add.cell.id = "Amrute")
amrute_fb$Dataset <- factor(amrute_fb$Dataset, levels = "Amrute")
amrute_fb$Group <- factor(amrute_fb$Group, levels = etiology_levels)
amrute_fb$HF.etiology <- factor(amrute_fb$HF.etiology, levels = etiology_levels)
amrute_fb$Disease_status <- factor(amrute_fb$Disease_status, levels = c("HC", "Disease"))
amrute_fb$Sample <- amrute_fb$sample

amrute_fb@meta.data %>%
  dplyr::count(HF.etiology, Sample, name = "n_fibroblast_cells") %>%
  dplyr::arrange(HF.etiology, Sample) %>%
  reactable::reactable()
```

## Basic RNA normalization and QC

```{r normalize fibroblasts}
DefaultAssay(amrute_fb) <- "RNA"

amrute_fb <- NormalizeData(amrute_fb, verbose = FALSE)
```

```{r qc violin fibroblasts, fig.width=9, fig.height=4}
VlnPlot(
  amrute_fb,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  group.by = "HF.etiology",
  cols = etiology_cols,
  pt.size = 0,
  ncol = 3
)
```

## SCTransform by sample

```{r sctransform fibroblasts by sample}
DefaultAssay(amrute_fb) <- "RNA"
FB_list <- SplitObject(amrute_fb, split.by = "Sample")

for (i in seq_along(FB_list)) {
  FB_list[[i]] <- SCTransform(
    FB_list[[i]],
    vst.flavor = "v2",
    vars.to.regress = "percent.mt",
    return.only.var.genes = TRUE,
    variable.features.n = 3000,
    min_cells = 0,
    verbose = FALSE
  )
}

features_fb <- SelectIntegrationFeatures(object.list = FB_list, nfeatures = 3000)

if (length(FB_list) == 1) {
  FB_combined <- FB_list[[1]]
} else {
  FB_combined <- merge(
    x = FB_list[[1]],
    y = FB_list[2:length(FB_list)],
    merge.data = TRUE
  )
}

VariableFeatures(FB_combined) <- features_fb
rm(FB_list, amrute_fb)
gc()
```

## PCA

```{r pca fibroblasts combined, fig.width=6, fig.height=4}
FB_combined <- RunPCA(FB_combined, npcs = 50, verbose = FALSE)
ElbowPlot(FB_combined, ndims = 50)
```

## Harmony integration by sample before contamination removal

```{r harmony fibroblasts combined raw}
DefaultAssay(FB_combined) <- "SCT"

FB_combined <- harmony::RunHarmony(
  object = FB_combined,
  assay.use = "SCT",
  group.by.vars = "Sample",
  dims.use = 1:50,
  plot_convergence = TRUE,
  .options = harmony::harmony_options(max.iter.cluster = 100)
)
```

```{r remove old sct resolutions raw}
resolutions_to_keep <- paste0("SCT_snn_res.", c("0.1", "0.2", "0.3", "0.4", "0.5", "0.6", "0.7", "0.8"))

sct_res_cols <- grep("^SCT_snn[\\._]res\\.", colnames(FB_combined@meta.data), value = TRUE)
cols_to_remove <- setdiff(sct_res_cols, resolutions_to_keep)

if (length(cols_to_remove) > 0) {
  FB_combined@meta.data <- FB_combined@meta.data[, !colnames(FB_combined@meta.data) %in% cols_to_remove, drop = FALSE]
}

grep("^SCT_snn[\\._]res\\.", colnames(FB_combined@meta.data), value = TRUE)
```

## Clustering before contamination removal

```{r clustering fibroblasts combined raw}
DefaultAssay(FB_combined) <- "SCT"

FB_combined <- RunUMAP(
  FB_combined,
  reduction = "harmony",
  dims = 1:30,
  reduction.name = "umap",
  reduction.key = "UMAP_"
)

FB_combined <- FindNeighbors(
  FB_combined,
  reduction = "harmony",
  dims = 1:30
)

FB_combined <- FindClusters(
  FB_combined,
  resolution = seq(0.1, 0.8, by = 0.1)
)

Idents(FB_combined) <- "SCT_snn_res.0.3"
```

```{r clustree raw, fig.width=8, fig.height=10}
clustree::clustree(
  FB_combined@meta.data[, grep("SCT_snn_res", colnames(FB_combined@meta.data))],
  prefix = "SCT_snn_res."
)
```

## UMAPs before contamination removal

```{r umaps fibroblasts combined raw clusters, fig.width=11, fig.height=10}
p1 <- DimPlot(
  FB_combined,
  reduction = "umap",
  group.by = "SCT_snn_res.0.3",
  label = TRUE,
  repel = TRUE,
  shuffle = TRUE,
  raster = FALSE,
  pt.size = 0.1
) +
  NoLegend() +
  ggtitle("Fibroblast subclusters, res 0.3") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

p2 <- DimPlot(
  FB_combined,
  reduction = "umap",
  group.by = "Sample",
  shuffle = TRUE,
  raster = FALSE,
  pt.size = 0.1
) +
  ggtitle("Sample") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

p3 <- DimPlot(
  FB_combined,
  reduction = "umap",
  group.by = "HF.etiology",
  cols = etiology_cols,
  shuffle = TRUE,
  raster = FALSE,
  pt.size = 0.1
) +
  ggtitle("HF etiology") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

p4 <- DimPlot(
  FB_combined,
  reduction = "umap",
  split.by = "HF.etiology",
  group.by = "SCT_snn_res.0.3",
  label = FALSE,
  shuffle = TRUE,
  pt.size = 0.1,
  raster = FALSE,
  ncol = 2
) +
  NoLegend() +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 12, face = "bold")
  )

(p1 + p2) / (p3 + p4)
```

## Cluster markers before contamination removal

```{r markers fibroblast raw clusters}
DefaultAssay(FB_combined) <- "RNA"
FB_combined <- NormalizeData(FB_combined, verbose = FALSE)
Idents(FB_combined) <- "SCT_snn_res.0.3"

FB_markers_raw_clusters <- FindAllMarkers(
  FB_combined,
  assay = "RNA",
  only.pos = TRUE,
  min.pct = 0.1,
  logfc.threshold = 0.25,
  return.thresh = 0.05
)

FB_markers_raw_clusters <- FB_markers_raw_clusters %>%
  dplyr::arrange(cluster, desc(avg_log2FC))

FB_markers_top10_raw_clusters <- FB_markers_raw_clusters %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 10, with_ties = FALSE) %>%
  dplyr::ungroup()

FB_markers_top3_raw_clusters <- FB_markers_raw_clusters %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 3, with_ties = FALSE) %>%
  dplyr::ungroup()

FB_markers_raw_clusters %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_head(n = 20) %>%
  reactable::reactable()
```

```{r write raw cluster markers, eval=TRUE}
write.csv(
  FB_markers_raw_clusters,
  here::here(output_dir_data, "FB_markers_amrute_raw_res_0_3_all.csv"),
  row.names = FALSE
)
write.csv(
  FB_markers_top10_raw_clusters,
  here::here(output_dir_data, "FB_markers_amrute_raw_res_0_3_top10.csv"),
  row.names = FALSE
)
write.csv(
  FB_markers_top3_raw_clusters,
  here::here(output_dir_data, "FB_markers_amrute_raw_res_0_3_top3.csv"),
  row.names = FALSE
)
```

```{r heatmap top markers raw, fig.width=9, fig.height=7}
DefaultAssay(FB_combined) <- "SCT"
Idents(FB_combined) <- "SCT_snn_res.0.3"

heatmap_cols <- colorRampPalette(RColorBrewer::brewer.pal(9, "RdBu"))(256)
heatmap_cols <- rev(heatmap_cols[1:256])

set.seed(123)
cells_use_heatmap_raw <- colnames(FB_combined)
if (length(cells_use_heatmap_raw) > 20000) {
  cells_use_heatmap_raw <- sample(cells_use_heatmap_raw, 20000)
}

DoHeatmap(
  subset(FB_combined, cells = cells_use_heatmap_raw),
  label = TRUE,
  draw.lines = FALSE,
  features = unique(FB_markers_top10_raw_clusters$gene),
  assay = "SCT",
  size = 4
) +
  scale_fill_gradientn(colours = heatmap_cols) +
  theme(
    text = element_text(size = 12),
    axis.text.y = element_text(size = 8),
    plot.margin = unit(c(0.2, 0, 0, 0), "inches")
  ) +
  guides(fill = guide_colorbar(barwidth = 1, barheight = 4)) +
  guides(color = "none")
```

## QC genes before contamination removal

```{r qc selected features raw, fig.height=5, fig.width=12}
features_qc <- c("nCount_RNA", "nFeature_RNA", "percent.mt", "TTN", "PECAM1", "VWF")
DefaultAssay(FB_combined) <- "RNA"
Idents(FB_combined) <- "SCT_snn_res.0.3"

VlnPlot_scCustom(
  FB_combined,
  features = features_qc,
  num_columns = 3,
  pt.size = 0,
  colors_use = cell_type_cols
) + theme(axis.title.x = element_blank())
```

## Flag contamination clusters at resolution 0.3

Clusters 5 and 7 are flagged as contamination and removed before reclustering the cleaned fibroblast object.

```{r flag contamination clusters res 0.3}
FB_combined$cell_type_contamination <- dplyr::case_when(
  FB_combined$SCT_snn_res.0.3 %in% c("5", "7") ~ "yes",
  TRUE ~ "no"
)

FB_combined$cell_type_contamination <- factor(
  FB_combined$cell_type_contamination,
  levels = c("no", "yes")
)

table(FB_combined$SCT_snn_res.0.3, FB_combined$cell_type_contamination)
```

```{r UMAP contamination flag, fig.width=5.5, fig.height=5, dpi=300}
DimPlot(
  FB_combined,
  group.by = "cell_type_contamination",
  reduction = "umap",
  cols = c("no" = "grey75", "yes" = "#A63A2A"),
  label = FALSE,
  shuffle = TRUE,
  raster = FALSE,
  pt.size = 0.1
) +
  ggtitle("Contamination flag before removal") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )
```

## Remove contamination clusters

```{r remove contamination fibroblasts before reclustering}
FB_combined_before_contamination_filter <- FB_combined

FB_combined <- subset(
  FB_combined,
  cell_type_contamination == "no"
)

FB_combined$cell_type_contamination <- droplevels(FB_combined$cell_type_contamination)
DefaultAssay(FB_combined) <- "SCT"

# Remove pre-filter clustering columns so that the next clustree only shows the clean reclustering.
sct_res_cols <- grep("^SCT_snn[\\._]res\\.", colnames(FB_combined@meta.data), value = TRUE)
if (length(sct_res_cols) > 0) {
  FB_combined@meta.data <- FB_combined@meta.data[, !colnames(FB_combined@meta.data) %in% sct_res_cols, drop = FALSE]
}

table(FB_combined$cell_type_contamination, FB_combined$HF.etiology)
```

## Harmony integration by sample after contamination removal

```{r harmony fibroblasts combined cleaned}
DefaultAssay(FB_combined) <- "SCT"

FB_combined <- harmony::RunHarmony(
  object = FB_combined,
  assay.use = "SCT",
  group.by.vars = "Sample",
  dims.use = 1:50,
  plot_convergence = TRUE,
  reduction.save = "harmony",
  .options = harmony::harmony_options(max.iter.cluster = 100)
)
```

## Reclustering after contamination removal

```{r clustering fibroblasts combined cleaned}
DefaultAssay(FB_combined) <- "SCT"

FB_combined <- RunUMAP(
  FB_combined,
  reduction = "harmony",
  dims = 1:30,
  reduction.name = "umap",
)

FB_combined <- FindNeighbors(
  FB_combined,
  reduction = "harmony",
  dims = 1:30
)

FB_combined <- FindClusters(
  FB_combined,
  resolution = seq(0.1, 0.8, by = 0.1)
)

Idents(FB_combined) <- "SCT_snn_res.0.3"
```

```{r clustree cleaned, fig.width=8, fig.height=10}
clustree::clustree(
  FB_combined@meta.data[, grep("SCT_snn_res", colnames(FB_combined@meta.data))],
  prefix = "SCT_snn_res."
)
```

## UMAPs after reclustering cleaned fibroblasts

```{r umaps fibroblasts cleaned clusters, fig.width=11, fig.height=10}
p1 <- DimPlot(
  FB_combined,
  reduction = "umap",
  group.by = "SCT_snn_res.0.3",
  label = TRUE,
  repel = TRUE,
  shuffle = TRUE,
  raster = FALSE,
  pt.size = 0.1
) +
  NoLegend() +
  ggtitle("Fibroblast subclusters, res 0.3") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

p2 <- DimPlot(
  FB_combined,
  reduction = "umap",
  group.by = "Sample",
  shuffle = TRUE,
  raster = FALSE,
  pt.size = 0.1
) +
  ggtitle("Sample") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

p3 <- DimPlot(
  FB_combined,
  reduction = "umap",
  group.by = "HF.etiology",
  cols = etiology_cols,
  shuffle = TRUE,
  raster = FALSE,
  pt.size = 0.1
) +
  ggtitle("HF etiology") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

p4 <- DimPlot(
  FB_combined,
  reduction = "umap",
  split.by = "HF.etiology",
  group.by = "SCT_snn_res.0.3",
  label = FALSE,
  shuffle = TRUE,
  pt.size = 0.1,
  raster = FALSE,
  ncol = 2
) +
  NoLegend() +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 12, face = "bold")
  )

(p1 + p2) / (p3 + p4)
```

## Annotate final fibroblast cell types after reclustering

```{r annotate fibroblast cell types cleaned reclustered res 0.3}
DefaultAssay(FB_combined) <- "SCT"
Idents(FB_combined) <- "SCT_snn_res.0.3"

clean_clusters <- levels(factor(FB_combined$SCT_snn_res.0.3))
clean_clusters <- clean_clusters[order(as.numeric(clean_clusters))]
clean_cluster_map <- setNames(paste0("FB", seq_along(clean_clusters)), clean_clusters)

FB_combined$cell_type <- unname(clean_cluster_map[as.character(FB_combined$SCT_snn_res.0.3)])
FB_combined$cell_type <- factor(
  FB_combined$cell_type,
  levels = paste0("FB", seq_along(clean_clusters))
)

Idents(FB_combined) <- "cell_type"

table(FB_combined$SCT_snn_res.0.3, FB_combined$cell_type)
table(FB_combined$cell_type, FB_combined$HF.etiology)
```

```{r UMAP cell types cleaned reclustered, fig.width=4.2, fig.height=4, dpi=300}
p <- DimPlot(
  FB_combined,
  group.by = "cell_type",
  reduction = "umap",
  cols = cell_type_cols,
  label = FALSE,
  shuffle = TRUE,
  raster = FALSE,
  pt.size = 0.1
) +
  NoLegend() +
  ggtitle(expression(italic("Amrute et al.") ~ " Nature CVR 2024")) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

LabelClusters(
  p,
  id = "cell_type",
  fontface = "bold",
  size = 5,
  repel = TRUE
)
```

```{r umap cleaned fibroblasts metadata reclustered, fig.width=11, fig.height=5}
p1 <- DimPlot(
  FB_combined,
  reduction = "umap",
  group.by = "Sample",
  shuffle = TRUE,
  raster = FALSE,
  pt.size = 0.1
) +
  ggtitle("Sample") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

p2 <- DimPlot(
  FB_combined,
  reduction = "umap",
  group.by = "HF.etiology",
  cols = etiology_cols,
  shuffle = TRUE,
  raster = FALSE,
  pt.size = 0.1
) +
  ggtitle("HF etiology") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

p1 + p2
```

## Cluster markers after contamination removal and reclustering

```{r markers fibroblast cell types cleaned reclustered}
DefaultAssay(FB_combined) <- "RNA"
FB_combined <- NormalizeData(FB_combined, verbose = FALSE)
Idents(FB_combined) <- "cell_type"

FB_markers_combined <- FindAllMarkers(
  FB_combined,
  assay = "RNA",
  only.pos = TRUE,
  min.pct = 0.1,
  logfc.threshold = 0.25,
  return.thresh = 0.05
)

FB_markers_combined <- FB_markers_combined %>%
  dplyr::arrange(cluster, desc(avg_log2FC))

FB_markers_top10_combined <- FB_markers_combined %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 10, with_ties = FALSE) %>%
  dplyr::ungroup()

FB_markers_top3_combined <- FB_markers_combined %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 3, with_ties = FALSE) %>%
  dplyr::ungroup()

FB_markers_combined %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_head(n = 20) %>%
  reactable::reactable()
```

```{r write cluster markers cleaned reclustered, eval=TRUE}
write.csv(
  FB_markers_combined,
  here::here(output_dir_data, "FB_markers_amrute_cleaned_reclustered_cell_type_all.csv"),
  row.names = FALSE
)
write.csv(
  FB_markers_top10_combined,
  here::here(output_dir_data, "FB_markers_amrute_cleaned_reclustered_cell_type_top10.csv"),
  row.names = FALSE
)
write.csv(
  FB_markers_top3_combined,
  here::here(output_dir_data, "FB_markers_amrute_cleaned_reclustered_cell_type_top3.csv"),
  row.names = FALSE
)
```

```{r heatmap top markers cleaned reclustered, fig.width=9, fig.height=7}
DefaultAssay(FB_combined) <- "SCT"
Idents(FB_combined) <- "cell_type"

heatmap_cols <- colorRampPalette(RColorBrewer::brewer.pal(9, "RdBu"))(256)
heatmap_cols <- rev(heatmap_cols[1:256])

set.seed(123)
cells_use_heatmap <- colnames(FB_combined)
if (length(cells_use_heatmap) > 20000) {
  cells_use_heatmap <- sample(cells_use_heatmap, 20000)
}

DoHeatmap(
  subset(FB_combined, cells = cells_use_heatmap),
  label = TRUE,
  draw.lines = FALSE,
  features = unique(FB_markers_top10_combined$gene),
  assay = "SCT",
  size = 4
) +
  scale_fill_gradientn(colours = heatmap_cols) +
  theme(
    text = element_text(size = 12),
    axis.text.y = element_text(size = 8),
    plot.margin = unit(c(0.2, 0, 0, 0), "inches")
  ) +
  guides(fill = guide_colorbar(barwidth = 1, barheight = 4)) +
  guides(color = "none")
```

## QC genes after contamination removal and reclustering

```{r qc selected features cleaned reclustered, fig.height=5, fig.width=12}
features_qc <- c("nCount_RNA", "nFeature_RNA", "percent.mt", "TTN", "PECAM1", "VWF")
DefaultAssay(FB_combined) <- "RNA"
Idents(FB_combined) <- "cell_type"

VlnPlot_scCustom(
  FB_combined,
  features = features_qc,
  num_columns = 3,
  pt.size = 0,
  colors_use = cell_type_cols
) + theme(axis.title.x = element_blank())
```

```{r dotplot top markers cleaned reclustered, fig.width=7, fig.height=4}
id_backup <- Idents(FB_combined)
Idents(FB_combined) <- factor(Idents(FB_combined), levels = rev(levels(Idents(FB_combined))))

p_dot <- DotPlot_scCustom(
  FB_combined,
  features = unique(FB_markers_top3_combined$gene),
  x_lab_rotate = TRUE
)

p_dot + theme(
  axis.text.x = element_text(size = 11, angle = 60, vjust = 1, hjust = 1),
  axis.text.y = element_text(size = 12)
)

Idents(FB_combined) <- id_backup
```

```{r featureplot top marker genes cleaned reclustered, fig.width=12, fig.height=16}
DefaultAssay(FB_combined) <- "RNA"

FeaturePlot_scCustom(
  FB_combined,
  features = unique(FB_markers_top3_combined$gene),
  num_columns = 3,
  order = TRUE
)
```


## Marker features on cleaned integrated fibroblasts

```{r selected marker featureplots cleaned, fig.width=8.4, fig.height=8}
DefaultAssay(FB_combined) <- "SCT"

matrifibrocyte_genes_present <- matrifibrocyte_genes[
  matrifibrocyte_genes %in% rownames(FB_combined)
]

FeaturePlot(
  FB_combined,
  features = matrifibrocyte_genes_present,
  reduction = "umap",
  order = TRUE,
  cols = c("grey90", "#A63A2A"),
  pt.size = 0.5,
  ncol = 2,
  raster = FALSE
)+
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

```{r featureplot ACTA2 cleaned, fig.width=4.2, fig.height=4, dpi=300}
FeaturePlot(
  FB_combined,
  features = "ACTA2",
  reduction = "umap",
  order = TRUE,
  cols = c("grey90", "#A63A2A"),
  pt.size = 0.5,
  raster = FALSE
) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

```{r featureplot MXRA5 cleaned, fig.width=4.2, fig.height=4, dpi=300}
FeaturePlot(
  FB_combined,
  features = "MXRA5",
  reduction = "umap",
  order = TRUE,
  cols = c("grey90", "#A63A2A"),
  pt.size = 0.5,
  raster = FALSE
) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

## Matrifibrocyte score on cleaned integrated fibroblasts

```{r add matrifibrocyte score cleaned}
DefaultAssay(FB_combined) <- "SCT"

if (length(matrifibrocyte_genes_present) == 0) {
  stop("None of the matrifibrocyte genes were found in the integrated object.")
}

if ("Matrifibrocyte_score" %in% colnames(FB_combined@meta.data)) {
  FB_combined$Matrifibrocyte_score <- NULL
}
if ("Matrifibrocyte_score1" %in% colnames(FB_combined@meta.data)) {
  FB_combined$Matrifibrocyte_score1 <- NULL
}

FB_combined <- AddModuleScore(
  object = FB_combined,
  features = list(matrifibrocyte_genes_present),
  name = "Matrifibrocyte_score",
  assay = "SCT",
  slot = "data",
  ctrl = 25,
  nbin = 24,
  seed = 1
)

FB_combined$Matrifibrocyte_score <- FB_combined$Matrifibrocyte_score1
FB_combined$Matrifibrocyte_score1 <- NULL
```

```{r feature plot matrifibrocyte score cleaned, fig.width=4.2, fig.height=4, dpi=300}
FeaturePlot(
  FB_combined,
  features = "Matrifibrocyte_score",
  reduction = "umap",
  order = TRUE,
  cols = c("grey90", "#A63A2A"),
  pt.size = 0.5,
  raster = FALSE
) +
  ggtitle("Matrifibrocyte score") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

```{r violin matrifibrocyte genes by cell type cleaned alpha dots, fig.width=12, fig.height=2.8, dpi=300}
Idents(FB_combined) <- "cell_type"

p_list <- VlnPlot(
  FB_combined,
  features = matrifibrocyte_genes_present,
  group.by = "cell_type",
  assay = "RNA",
  cols = cell_type_cols,
  pt.size = 0.001,
  add.noise = FALSE,
  ncol = length(matrifibrocyte_genes_present),
  combine = FALSE
)

p_list <- lapply(p_list, function(p) {
  for (j in seq_along(p$layers)) {
    geom_name <- class(p$layers[[j]]$geom)[1]

    if (geom_name %in% c("GeomPoint", "GeomJitter")) {
      p$layers[[j]]$aes_params$alpha <- 0.1
      p$layers[[j]]$aes_params$colour <- scales::alpha("black", 0.01)
    }
  }

  p +
    theme_classic() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", size = 12),
      axis.text.y = element_text(face = "bold", size = 10),
      axis.title.y = element_text(face = "bold"),
      axis.title.x = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.position = "none"
    )
})

wrap_plots(p_list, ncol = length(p_list))
```

```{r violin ACTA2 by cell type cleaned alpha dots, fig.width=3, fig.height=2.8, dpi=300}
Idents(FB_combined) <- "cell_type"

p_acta2 <- VlnPlot(
  FB_combined,
  features = "ACTA2",
  assay = "RNA",
  group.by = "cell_type",
  cols = cell_type_cols,
  pt.size = 0.001,
  add.noise = FALSE
)

for (j in seq_along(p_acta2$layers)) {
  geom_name <- class(p_acta2$layers[[j]]$geom)[1]

  if (geom_name %in% c("GeomPoint", "GeomJitter")) {
    p_acta2$layers[[j]]$aes_params$alpha <- 0.1
    p_acta2$layers[[j]]$aes_params$colour <- scales::alpha("black", 0.01)
  }
}

p_acta2 +
  ggtitle("ACTA2") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 10),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "none"
  )

```

```{r violin MXRA5 by cell type cleaned alpha dots, fig.width=3, fig.height=2.8, dpi=300}
Idents(FB_combined) <- "cell_type"

p_mxra5 <- VlnPlot(
  FB_combined,
  features = "MXRA5",
  assay = "RNA",
  group.by = "cell_type",
  cols = cell_type_cols,
  pt.size = 0.001,
  add.noise = FALSE
)

for (j in seq_along(p_mxra5$layers)) {
  geom_name <- class(p_mxra5$layers[[j]]$geom)[1]

  if (geom_name %in% c("GeomPoint", "GeomJitter")) {
    p_mxra5$layers[[j]]$aes_params$alpha <- 0.1
    p_mxra5$layers[[j]]$aes_params$colour <- scales::alpha("black", 0.01)
  }
}

p_mxra5 +
  ggtitle("MXRA5") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 10),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "none"
  )
```

```{r violin matrifibrocyte score by cell type cleaned alpha dots, fig.width=3, fig.height=2.8, dpi=300}
Idents(FB_combined) <- "cell_type"

p_score <- VlnPlot(
  FB_combined,
  features = "Matrifibrocyte_score",
  group.by = "cell_type",
  cols = cell_type_cols,
  pt.size = 0.00,
  add.noise = FALSE
)

for (j in seq_along(p_score$layers)) {
  geom_name <- class(p_score$layers[[j]]$geom)[1]

  if (geom_name %in% c("GeomPoint", "GeomJitter")) {
    p_score$layers[[j]]$aes_params$alpha <- 0.1
    p_score$layers[[j]]$aes_params$colour <- scales::alpha("black", 0.01)
  }
}

p_score +
  ggtitle("Matrifibrocyte score") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 10),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "none"
  )
```

```{r dotplot selected genes by cell type cleaned flipped, fig.width=8, fig.height=4, dpi=300}
DefaultAssay(FB_combined) <- "SCT"
Idents(FB_combined) <- "cell_type"

DotPlot(
  FB_combined,
  dot.scale = 13,
  features = matrifibrocyte_genes,
  assay = "SCT",
  cols = c("grey90", "#A63A2A")
) +
  coord_flip() +
  theme_classic() +
  theme(
    axis.text.x = element_text(face = "bold", size = 19),
    axis.text.y = element_text(face = "bold", size = 19),
    axis.title = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

## Cell-type proportions by HF etiology

```{r cell type proportions metadata cleaned}
cluster_var <- "cell_type"

prop_df <- FB_combined@meta.data %>%
  dplyr::mutate(cluster = .data[[cluster_var]]) %>%
  dplyr::count(HF.etiology, Sample, cluster, name = "n") %>%
  dplyr::group_by(HF.etiology, Sample) %>%
  dplyr::mutate(
    total_n = sum(n),
    freq = n / total_n
  ) %>%
  dplyr::ungroup()

prop_df$cluster <- factor(
  prop_df$cluster,
  levels = levels(FB_combined$cell_type)
)

prop_df$HF.etiology <- factor(as.character(prop_df$HF.etiology), levels = etiology_levels)
```

```{r cell type proportions by etiology cleaned, fig.width=8, fig.height=3, dpi=300}
pd  <- position_dodge(width = 0.75)
pjd <- position_jitterdodge(jitter.width = 0.12, dodge.width = 0.75, seed = 123)

ggplot(prop_df, aes(x = cluster, y = freq, fill = HF.etiology)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, width = 0.7, position = pd) +
  geom_point(position = pjd, size = 0.8, alpha = 0.7, shape = 16) +
  scale_fill_manual(values = etiology_cols, drop = FALSE) +
  labs(x = "Cell type", y = "Fraction of fibroblast cells") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.title = element_blank()
  )
```

## Pseudobulk fibroblasts by sample after contamination removal

```{r pseudobulk fibroblasts cleaned}
DefaultAssay(FB_combined) <- "RNA"
counts_fb <- GetAssayData(FB_combined, assay = "RNA", slot = "counts")
metadata_fb_matched <- FB_combined@meta.data %>%
  tibble::rownames_to_column("cell")

sample_vec <- metadata_fb_matched$Sample
sample_levels <- unique(sample_vec)

counts_pb <- do.call(
  cbind,
  lapply(sample_levels, function(sid) {
    Matrix::rowSums(counts_fb[, sample_vec == sid, drop = FALSE])
  })
)

counts_pb <- methods::as(counts_pb, "dgCMatrix")
rownames(counts_pb) <- rownames(counts_fb)
colnames(counts_pb) <- sample_levels

sample_meta <- metadata_fb_matched %>%
  dplyr::select(Sample, sample, HF.etiology, Group, Disease_status, Dataset, dplyr::any_of(c("Race", "Sex", "Age"))) %>%
  dplyr::distinct() %>%
  dplyr::arrange(match(Sample, colnames(counts_pb)))

sample_meta$n_fibroblast_cells <- as.numeric(
  table(factor(sample_vec, levels = colnames(counts_pb)))[sample_meta$Sample]
)
sample_meta$libsize <- Matrix::colSums(counts_pb)[sample_meta$Sample]

sample_meta <- sample_meta %>%
  dplyr::filter(
    libsize >= min_pb_libsize,
    n_fibroblast_cells >= min_fibroblast_cells
  )

counts_pb <- counts_pb[, sample_meta$Sample, drop = FALSE]

sample_meta$HF.etiology <- factor(as.character(sample_meta$HF.etiology), levels = etiology_levels)
sample_meta$Group <- factor(as.character(sample_meta$Group), levels = etiology_levels)
sample_meta$Disease_status <- factor(as.character(sample_meta$Disease_status), levels = c("HC", "Disease"))
rownames(sample_meta) <- sample_meta$Sample

write.csv(
  sample_meta,
  file.path(
    output_dir_data,
    paste0(
      "cleaned_fibroblast_pseudobulk_sample_metadata_min_libsize_",
      min_pb_libsize,
      "_min_fibroblast_cells_",
      min_fibroblast_cells,
      ".csv"
    )
  ),
  row.names = FALSE
)

sample_meta %>%
  dplyr::count(HF.etiology, name = "n_samples")

sample_meta %>% reactable::reactable()
```

## logCPM normalization and pseudobulk scores

```{r logcpm normalization cleaned}
dge <- edgeR::DGEList(counts = counts_pb)
dge <- edgeR::calcNormFactors(dge)
logcpm <- edgeR::cpm(dge, log = TRUE, prior.count = 1)

genes_present <- intersect(matrifibrocyte_genes, rownames(logcpm))
missing_genes <- setdiff(matrifibrocyte_genes, rownames(logcpm))

extra_pseudobulk_genes <- c("ACTA2", "MXRA5")
extra_pseudobulk_genes_present <- intersect(extra_pseudobulk_genes, rownames(logcpm))
mxra5_present <- "MXRA5" %in% rownames(logcpm)

genes_present
missing_genes
extra_pseudobulk_genes_present

if (length(genes_present) == 0) {
  stop("None of the matrifibrocyte genes were found in the RNA matrix.")
}

matrifibrocyte_z <- t(scale(t(logcpm[genes_present, , drop = FALSE])))
matrifibrocyte_pb_score <- colMeans(matrifibrocyte_z, na.rm = TRUE)
```

```{r gene and score expression table cleaned}
genes_for_plot <- unique(c(genes_present, extra_pseudobulk_genes_present))

gene_plot_df <- logcpm[genes_for_plot, , drop = FALSE] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene") %>%
  tidyr::pivot_longer(cols = -gene, names_to = "Sample", values_to = "logCPM") %>%
  dplyr::left_join(sample_meta, by = "Sample") %>%
  dplyr::mutate(
    gene = factor(gene, levels = unique(c(matrifibrocyte_genes, extra_pseudobulk_genes))),
    HF.etiology = factor(as.character(HF.etiology), levels = etiology_levels),
    Disease_status = factor(as.character(Disease_status), levels = c("HC", "Disease"))
  )

score_plot_df <- tibble::tibble(
  Sample = names(matrifibrocyte_pb_score),
  Matrifibrocyte_score = as.numeric(matrifibrocyte_pb_score)
) %>%
  dplyr::left_join(sample_meta, by = "Sample") %>%
  dplyr::mutate(
    HF.etiology = factor(as.character(HF.etiology), levels = etiology_levels),
    Disease_status = factor(as.character(Disease_status), levels = c("HC", "Disease"))
  )

write.csv(
  gene_plot_df,
  file.path(
    output_dir_data,
    paste0(
      "cleaned_fibroblast_pseudobulk_logCPM_selected_genes_min_libsize_",
      min_pb_libsize,
      "_min_fibroblast_cells_",
      min_fibroblast_cells,
      ".csv"
    )
  ),
  row.names = FALSE
)

write.csv(
  score_plot_df,
  file.path(
    output_dir_data,
    paste0(
      "cleaned_fibroblast_pseudobulk_matrifibrocyte_score_min_libsize_",
      min_pb_libsize,
      "_min_fibroblast_cells_",
      min_fibroblast_cells,
      ".csv"
    )
  ),
  row.names = FALSE
)
```

## limma comparisons across HF etiology groups

This model is intentionally simple: `gene ~ 0 + HF.etiology`. The contrasts include each disease group versus Donor and all pairwise disease-group comparisons. No pooled HC versus disease plot is generated.

```{r limma gene comparisons cleaned}
model_meta <- sample_meta %>%
  dplyr::filter(!is.na(HF.etiology))

model_meta$HF.etiology <- droplevels(
  factor(as.character(model_meta$HF.etiology), levels = etiology_levels)
)

feature_mat <- logcpm[genes_for_plot, model_meta$Sample, drop = FALSE]

if (length(unique(model_meta$HF.etiology)) < 2) {
  stop("Fewer than two HF.etiology groups remain after filtering.")
}

design <- model.matrix(~ 0 + HF.etiology, data = model_meta)
colnames(design) <- gsub("^HF.etiology", "", colnames(design))

contrast_strings <- c(
  "ICM_vs_Donor" = "ICM - Donor",
  "NICM_vs_Donor" = "NICM - Donor",
  "AMI_vs_Donor" = "AMI - Donor",
  "ICM_vs_NICM" = "ICM - NICM",
  "AMI_vs_ICM" = "AMI - ICM",
  "AMI_vs_NICM" = "AMI - NICM"
)

contrast_strings <- contrast_strings[vapply(contrast_strings, function(x) {
  groups_needed <- trimws(unlist(strsplit(x, "-")))
  all(groups_needed %in% colnames(design))
}, logical(1))]

contrast_matrix <- limma::makeContrasts(
  contrasts = contrast_strings,
  levels = design
)

fit <- limma::lmFit(feature_mat, design)
fit2 <- limma::contrasts.fit(fit, contrast_matrix)
fit2 <- limma::eBayes(fit2)

limma_results <- dplyr::bind_rows(lapply(colnames(contrast_matrix), function(contrast_name) {
  limma::topTable(
    fit2,
    coef = contrast_name,
    number = Inf,
    sort.by = "none"
  ) %>%
    tibble::rownames_to_column("gene") %>%
    dplyr::rename(
      coefficient_logFC = logFC,
      raw_p_value = P.Value,
      FDR_BH_within_contrast = adj.P.Val,
      B_statistic = B
    ) %>%
    dplyr::mutate(
      comparison = contrast_name,
      model_formula = "gene_logCPM ~ 0 + HF.etiology",
      min_pb_libsize = min_pb_libsize,
      min_fibroblast_cells = min_fibroblast_cells,
      n_samples_used = nrow(model_meta),
      samples_used = paste(model_meta$Sample, collapse = ";")
    )
})) %>%
  dplyr::select(
    comparison,
    gene,
    coefficient_logFC,
    AveExpr,
    t,
    raw_p_value,
    FDR_BH_within_contrast,
    B_statistic,
    model_formula,
    min_pb_libsize,
    min_fibroblast_cells,
    n_samples_used,
    samples_used,
    dplyr::everything()
  )

write.csv(
  limma_results,
  file.path(
    output_dir_data,
    paste0(
      "cleaned_fibroblast_selected_genes_limma_HF_etiology_contrasts_min_libsize_",
      min_pb_libsize,
      "_min_fibroblast_cells_",
      min_fibroblast_cells,
      ".csv"
    )
  ),
  row.names = FALSE
)

limma_results %>% reactable::reactable()
```

## Simple score comparisons

```{r score comparisons cleaned}
score_mat <- matrix(
  score_plot_df$Matrifibrocyte_score,
  nrow = 1,
  dimnames = list("Matrifibrocyte_score", score_plot_df$Sample)
)

score_model_meta <- score_plot_df %>%
  dplyr::filter(Sample %in% colnames(score_mat))

score_mat <- score_mat[, score_model_meta$Sample, drop = FALSE]
score_design <- model.matrix(~ 0 + HF.etiology, data = score_model_meta)
colnames(score_design) <- gsub("^HF.etiology", "", colnames(score_design))

score_contrast_strings <- contrast_strings[vapply(contrast_strings, function(x) {
  groups_needed <- trimws(unlist(strsplit(x, "-")))
  all(groups_needed %in% colnames(score_design))
}, logical(1))]

score_contrast_matrix <- limma::makeContrasts(
  contrasts = score_contrast_strings,
  levels = score_design
)

score_fit <- limma::lmFit(score_mat, score_design)
score_fit2 <- limma::contrasts.fit(score_fit, score_contrast_matrix)
score_fit2 <- limma::eBayes(score_fit2)

score_limma_results <- dplyr::bind_rows(lapply(colnames(score_contrast_matrix), function(contrast_name) {
  limma::topTable(
    score_fit2,
    coef = contrast_name,
    number = Inf,
    sort.by = "none"
  ) %>%
    tibble::rownames_to_column("feature") %>%
    dplyr::rename(
      coefficient_logFC = logFC,
      raw_p_value = P.Value,
      FDR_BH_within_contrast = adj.P.Val,
      B_statistic = B
    ) %>%
    dplyr::mutate(
      comparison = contrast_name,
      model_formula = "pseudobulk_score ~ 0 + HF.etiology",
      n_samples_used = nrow(score_model_meta)
    )
}))

write.csv(
  score_limma_results,
  file.path(
    output_dir_data,
    paste0(
      "cleaned_fibroblast_matrifibrocyte_score_limma_HF_etiology_contrasts_min_libsize_",
      min_pb_libsize,
      "_min_fibroblast_cells_",
      min_fibroblast_cells,
      ".csv"
    )
  ),
  row.names = FALSE
)

score_limma_results %>% reactable::reactable()
```

## Boxplots of pseudobulk matrifibrocyte genes with all comparison lines and FDR labels

```{r boxplots matrifibrocyte genes cleaned, fig.width=9, fig.height=3.5}
comparison_map <- tibble::tibble(
  comparison = c(
    "ICM_vs_Donor",
    "NICM_vs_Donor",
    "AMI_vs_Donor",
    "ICM_vs_NICM",
    "AMI_vs_ICM",
    "AMI_vs_NICM"
  ),
  x = c(1, 1, 1, 2, 2, 3),
  xend = c(2, 3, 4, 3, 4, 4)
)

gene_plot_df_mf <- gene_plot_df %>%
  dplyr::filter(gene %in% matrifibrocyte_genes)

gene_ypos <- gene_plot_df_mf %>%
  dplyr::group_by(gene) %>%
  dplyr::summarise(
    y_min = min(logCPM, na.rm = TRUE),
    y_max = max(logCPM, na.rm = TRUE),
    y_range = y_max - y_min,
    .groups = "drop"
  ) %>%
  dplyr::mutate(y_range = dplyr::if_else(y_range == 0, 1, y_range))

stat_annot_df <- limma_results %>%
  dplyr::filter(gene %in% matrifibrocyte_genes) %>%
  dplyr::left_join(comparison_map, by = "comparison") %>%
  dplyr::filter(!is.na(x), !is.na(xend)) %>%
  dplyr::left_join(gene_ypos, by = "gene") %>%
  dplyr::mutate(
    label = vapply(FDR_BH_within_contrast, get_fdr_label, character(1)),
    is_sig = !is.na(FDR_BH_within_contrast) & FDR_BH_within_contrast < 0.1,
    span = xend - x
  ) %>%
  dplyr::filter(is_sig) %>%
  dplyr::group_by(gene) %>%
  dplyr::arrange(span, x, xend, .by_group = TRUE) %>%
  dplyr::mutate(
    stack_id = dplyr::row_number(),
    line_y = y_max + (0.12 + 0.15 * (stack_id - 1)) * y_range,
    tick_height = 0.035 * y_range,
    label_y = line_y + 0.055 * y_range
  ) %>%
  dplyr::ungroup()

ggplot(gene_plot_df_mf, aes(x = HF.etiology, y = logCPM, fill = HF.etiology)) +
  geom_boxplot(width = 0.65, outlier.shape = NA, color = "black", linewidth = 0.45) +
  geom_jitter(width = 0.12, size = 1.9, alpha = 0.9, color = "black") +
  geom_segment(
    data = stat_annot_df,
    aes(x = x, xend = xend, y = line_y, yend = line_y),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_segment(
    data = stat_annot_df,
    aes(x = x, xend = x, y = line_y - tick_height, yend = line_y),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_segment(
    data = stat_annot_df,
    aes(x = xend, xend = xend, y = line_y - tick_height, yend = line_y),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_text(
    data = stat_annot_df,
    aes(x = (x + xend) / 2, y = label_y, label = label),
    inherit.aes = FALSE,
    size = 3.0,
    fontface = "bold"
  ) +
  facet_wrap(~gene, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = etiology_cols, drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.08))) +
  labs(
    title = "Matrifibrocyte markers",
    x = NULL,
    y = "logCPM"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title.y = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )
```

## MXRA5 pseudobulk differences between HF etiology groups

```{r boxplot MXRA5 pseudobulk cleaned, fig.width=3, fig.height=3.5}
if (!mxra5_present) {
  stop("MXRA5 was not found in the pseudobulk logCPM matrix.")
}

mxra5_plot_df <- gene_plot_df %>%
  dplyr::filter(gene == "MXRA5")

mxra5_ypos <- mxra5_plot_df %>%
  dplyr::summarise(
    y_min = min(logCPM, na.rm = TRUE),
    y_max = max(logCPM, na.rm = TRUE),
    y_range = y_max - y_min
  ) %>%
  dplyr::mutate(
    y_range = dplyr::if_else(y_range == 0, 1, y_range)
  )

mxra5_annot_df <- limma_results %>%
  dplyr::filter(gene == "MXRA5") %>%
  dplyr::left_join(comparison_map, by = "comparison") %>%
  dplyr::filter(!is.na(x), !is.na(xend)) %>%
  dplyr::mutate(
    label = vapply(FDR_BH_within_contrast, get_fdr_label, character(1)),
    is_sig = !is.na(FDR_BH_within_contrast) & FDR_BH_within_contrast < 0.1,
    span = xend - x
  ) %>%
  dplyr::filter(is_sig) %>%
  dplyr::arrange(span, x, xend) %>%
  dplyr::mutate(
    stack_id = dplyr::row_number(),
    line_y = mxra5_ypos$y_max + (0.12 + 0.15 * (stack_id - 1)) * mxra5_ypos$y_range,
    tick_height = 0.035 * mxra5_ypos$y_range,
    label_y = line_y + 0.055 * mxra5_ypos$y_range
  )

ggplot(mxra5_plot_df, aes(x = HF.etiology, y = logCPM, fill = HF.etiology)) +
  geom_boxplot(width = 0.65, outlier.shape = NA, color = "black", linewidth = 0.45) +
  geom_jitter(width = 0.12, size = 2.0, alpha = 0.9, color = "black") +
  geom_segment(
    data = mxra5_annot_df,
    aes(x = x, xend = xend, y = line_y, yend = line_y),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_segment(
    data = mxra5_annot_df,
    aes(x = x, xend = x, y = line_y - tick_height, yend = line_y),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_segment(
    data = mxra5_annot_df,
    aes(x = xend, xend = xend, y = line_y - tick_height, yend = line_y),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_text(
    data = mxra5_annot_df,
    aes(x = (x + xend) / 2, y = label_y, label = label),
    inherit.aes = FALSE,
    size = 3.0,
    fontface = "bold"
  ) +
  scale_fill_manual(values = etiology_cols, drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.08))) +
  labs(
    title = "MXRA5",
    x = NULL,
    y = "logCPM"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "none"
  )
```

## Boxplot of pseudobulk matrifibrocyte score

```{r boxplot matrifibrocyte score cleaned, fig.width=3, fig.height=3.5}
score_ypos <- score_plot_df %>%
  dplyr::summarise(
    y_min = min(Matrifibrocyte_score, na.rm = TRUE),
    y_max = max(Matrifibrocyte_score, na.rm = TRUE),
    y_range = y_max - y_min
  ) %>%
  dplyr::mutate(
    y_range = dplyr::if_else(y_range == 0, 1, y_range)
  )

score_annot_df <- score_limma_results %>%
  dplyr::left_join(comparison_map, by = "comparison") %>%
  dplyr::filter(!is.na(x), !is.na(xend)) %>%
  dplyr::mutate(
    label = vapply(FDR_BH_within_contrast, get_fdr_label, character(1)),
    is_sig = !is.na(FDR_BH_within_contrast) & FDR_BH_within_contrast < 0.1,
    span = xend - x
  ) %>%
  dplyr::filter(is_sig) %>%
  dplyr::arrange(span, x, xend) %>%
  dplyr::mutate(
    stack_id = dplyr::row_number(),
    line_y = score_ypos$y_max + (0.12 + 0.15 * (stack_id - 1)) * score_ypos$y_range,
    tick_height = 0.035 * score_ypos$y_range,
    label_y = line_y + 0.055 * score_ypos$y_range
  )

ggplot(score_plot_df, aes(x = HF.etiology, y = Matrifibrocyte_score, fill = HF.etiology)) +
  geom_boxplot(width = 0.65, outlier.shape = NA, color = "black", linewidth = 0.45) +
  geom_jitter(width = 0.12, size = 2.0, alpha = 0.9, color = "black") +
  geom_segment(
    data = score_annot_df,
    aes(x = x, xend = xend, y = line_y, yend = line_y),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_segment(
    data = score_annot_df,
    aes(x = x, xend = x, y = line_y - tick_height, yend = line_y),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_segment(
    data = score_annot_df,
    aes(x = xend, xend = xend, y = line_y - tick_height, yend = line_y),
    inherit.aes = FALSE,
    linewidth = 0.45
  ) +
  geom_text(
    data = score_annot_df,
    aes(x = (x + xend) / 2, y = label_y, label = label),
    inherit.aes = FALSE,
    size = 3.0,
    fontface = "bold"
  ) +
  scale_fill_manual(values = etiology_cols, drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.08))) +
  labs(
    title = "Matrifibrocyte score",
    x = NULL,
    y = "Mean z-score of COMP/CILP/CHAD/THBS4"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title.y = element_text(face = "bold", size = 11.7),
    legend.position = "none"
  )
```

```{r results tables cleaned}
reactable::reactable(
  dplyr::select(limma_results, -samples_used),
  filterable = TRUE,
  searchable = FALSE
)

reactable::reactable(
  score_limma_results,
  filterable = TRUE,
  searchable = FALSE
)
```

## Save cleaned integrated fibroblast object

```{r save FB_combined final cleaned, eval=FALSE}
DefaultAssay(FB_combined) <- "RNA"
qs_save(
  FB_combined,
  file = here::here(output_dir_data, "amrute_fb_integrated_cleaned.qs2"),
  nthreads = 32
)
```
