---
title: "Differential abundance and differential expression analysis in cardiomyocytes"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "Cardiomyocytes_DA_DE"

# Load libraries
library(here)
library(readr)
library(readxl)
library(xlsx)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(ArchR)
library(clustree)
library(harmony)
library(gprofiler2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationHub)
library(ReactomePA)
library(statmod)
library(edgeR)
library(speckle)
library(EnhancedVolcano)
library(decoupleR)
library(OmnipathR)
library(dorothea)
library(enrichplot)
library(png)
library(reactable)
library(UpSetR)
library(ComplexHeatmap)
library(msigdbr)
library(simplifyEnrichment)
library(miloR)
library(igraph)
library(gridExtra) 
library(ggpubr)
library(DOSE)

#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```

Load cardiomyocytes dataset

```{r read CM}
CM <- readRDS(here::here("output", "Cardiomyocytes_subclustering_26", "CM.rds"))
```

```{r NormalizeData}
DefaultAssay(CM) <- "RNA"
CM <- NormalizeData(CM)
```

## Differential abundance analysis

```{r metadata}
metadata <- CM@meta.data %>%
  dplyr::select(Sample, Batch_processing, Batch_sequencing, Age_years, Sex, Group, mPAP_mmHg, sPAP_mmHg, PCW_mmHg, PVR_WU, TAPSE_mm, RVID_mm, LVEF_percent, LVEDD_mm) %>%
  unique() %>%
  dplyr::arrange(match(Sample, names(table(CM$Sample))))

rownames(metadata) <- metadata$Sample

all.equal(metadata$Sample, names(table(CM$Sample)))

reactable(metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

metadata_DCM <- metadata %>%
  dplyr::filter(Group == "DCM")
```

```{r Speckle_all}
differential_abundance <- data.frame()
props <- getTransformedProps(CM$cell_state, CM$Sample, transform="logit")

for (i in 7:(length(metadata_DCM))) {
  metadata_subset <- metadata_DCM[complete.cases(metadata_DCM[[i]]),]
  props_subset <- props
  props_subset$Counts <- props$Counts[, metadata_subset$Sample, drop = FALSE]
  props_subset$TransformedProps <- props$TransformedProps[, metadata_subset$Sample, drop = FALSE]
  props_subset$Proportions <- props$Proportions[, metadata_subset$Sample, drop = FALSE]
  metadata_subset <- metadata_subset %>% 
    arrange(Sample, colnames(props_subset$Proportions))
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  fit <- lmFit(props_subset$TransformedProps, design)
  fit <- eBayes(fit, robust=TRUE)
  differential_abundance_temp <- topTable(fit, n = Inf, coef = 2)
  differential_abundance_temp$metadata <- colnames(metadata_subset[i])
  differential_abundance_temp$cell_state <- rownames(differential_abundance_temp)
  differential_abundance <- rbind(differential_abundance, differential_abundance_temp)
}

rm(metadata_subset, props_subset, design, fit, differential_abundance_temp)

rownames(differential_abundance) <- NULL
differential_abundance <- dplyr::arrange(differential_abundance, adj.P.Val)

write.csv(differential_abundance, file = here::here(output_dir_data, "CM_Differential_Abundance.csv"), row.names = F)

reactable(differential_abundance, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Speckle_LVEDD_mm, fig.height=4, fig.width=20}
metadata_subset <- metadata[complete.cases(metadata[["LVEDD_mm"]]),]

props_subset <- props
props_subset$Counts <- props$Counts[, metadata_subset$Sample, drop = FALSE]
props_subset$TransformedProps <- props$TransformedProps[, metadata_subset$Sample, drop = FALSE]
props_subset$Proportions <- props$Proportions[, metadata_subset$Sample, drop = FALSE]

metadata_subset <- metadata_subset %>% 
  arrange(Sample, colnames(props_subset$Proportions))

design <- model.matrix(~ metadata_subset$LVEDD_mm + metadata_subset$Age_years)
fit.prop <- lmFit(props_subset$Proportions,design)

par(mfrow=c(1,5), mar = c(5, 5, 3, 1))
for(i in seq(1,5,1)){
  plot(metadata_subset$LVEDD_mm, props_subset$Proportions[i,], main = rownames(props_subset$Proportions)[i],
       pch=16, cex=2, ylab="Proportions", xlab="LVEDD_mm", cex.lab=2, cex.axis=1.5,
       cex.main=2.5)
  abline(a=fit.prop$coefficients[i,1], b=fit.prop$coefficients[i,2], col=4,
         lwd=1)
}
rm(metadata_subset, props_subset, design, fit.prop)
```

```{r Speckle_LVEF_percent, fig.height=4, fig.width=20}
metadata_subset <- metadata[complete.cases(metadata[["LVEF_percent"]]),]

props_subset <- props
props_subset$Counts <- props$Counts[, metadata_subset$Sample, drop = FALSE]
props_subset$TransformedProps <- props$TransformedProps[, metadata_subset$Sample, drop = FALSE]
props_subset$Proportions <- props$Proportions[, metadata_subset$Sample, drop = FALSE]

metadata_subset <- metadata_subset %>% 
  arrange(Sample, colnames(props_subset$Proportions))

design <- model.matrix(~ metadata_subset$LVEF_percent + metadata_subset$Age_years)
fit.prop <- lmFit(props_subset$Proportions,design)

par(mfrow=c(1,5), mar = c(5, 5, 3, 1))
for(i in seq(1,5,1)){
  plot(metadata_subset$LVEF_percent, props_subset$Proportions[i,], main = rownames(props_subset$Proportions)[i],
       pch=16, cex=2, ylab="Proportions", xlab="LVEF_percent", cex.lab=2, cex.axis=1.5,
       cex.main=2.5)
  abline(a=fit.prop$coefficients[i,1], b=fit.prop$coefficients[i,2], col=4,
         lwd=1)
}
rm(metadata_subset, props_subset, design, fit.prop)
```

```{r Speckle_mPAP_mmHg, fig.height=4, fig.width=20}
metadata_subset <- metadata[complete.cases(metadata[["mPAP_mmHg"]]),]

props_subset <- props
props_subset$Counts <- props$Counts[, metadata_subset$Sample, drop = FALSE]
props_subset$TransformedProps <- props$TransformedProps[, metadata_subset$Sample, drop = FALSE]
props_subset$Proportions <- props$Proportions[, metadata_subset$Sample, drop = FALSE]

metadata_subset <- metadata_subset %>% 
  arrange(Sample, colnames(props_subset$Proportions))

design <- model.matrix(~ metadata_subset$mPAP_mmHg + metadata_subset$Age_years)
fit.prop <- lmFit(props_subset$Proportions,design)

par(mfrow=c(1,5), mar = c(5, 5, 3, 1))
for(i in seq(1,5,1)){
  plot(metadata_subset$mPAP_mmHg, props_subset$Proportions[i,], main = rownames(props_subset$Proportions)[i],
       pch=16, cex=2, ylab="Proportions", xlab="mPAP_mmHg", cex.lab=2, cex.axis=1.5,
       cex.main=2.5)
  abline(a=fit.prop$coefficients[i,1], b=fit.prop$coefficients[i,2], col=4,
         lwd=1)
}
rm(metadata_subset, props_subset, design, fit.prop)
```

```{r Speckle_Group, fig.height=4, fig.width=20}
design <- model.matrix(~ 0 + metadata$Group)
colnames(design) <- c("DCM", "HC")
mycontr <- makeContrasts(DCM-HC, levels=design)
differential_abundance_HC <- propeller.ttest(props, design, contrasts = mycontr, robust=TRUE, trend=FALSE, sort=TRUE)

write.csv(differential_abundance_HC, file = here::here(output_dir_data, "CM_Differential_Abundance_HC.csv"), row.names = F)

reactable(differential_abundance_HC, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

par(mfrow = c(1, 5), mar = c(3, 5, 3, 1))
for (i in seq(1, 5, 1)) {
  stripchart(props$Proportions[i,] ~ metadata$Group,
             vertical = TRUE, pch = 16, method = "jitter", 
             col = c("orange", "purple"),
             cex = 2, ylab = "Proportions", cex.lab = 2, cex.axis = 2)
  title(rownames(props$Proportions)[i], cex.main = 2.5)
}
rm(design, mycontr)
```

```{r CM_proportions}
CM_proportions <- as.data.frame(props$Proportions) %>%   
  pivot_wider(
    names_from  = clusters,  
    values_from = Freq       
  )

write.csv(CM_proportions, file = here::here(output_dir_data, "CM_Proportions.csv"), row.names = F)
```

```{r rm differential_abundance, eval = FALSE}
rm(differential_abundance, differential_abundance_HC, props, CM_proportions)
```


## Differential expression analysis

```{r percent_stats}
expr_matrix <- GetAssayData(object = CM, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- CM[which(rownames(CM) %in% keep_genes),]
rm(keep_genes, non_zero_cells, expr_matrix, total_cells)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```


Create and prepare pseudobulk object for differential expression analysis

```{r pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r save CM_pseudocounts, eval=FALSE}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "CM_pseudocounts.rds"))
```

```{r filter pseudocounts}
keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r DE_all}
tested_param <- c("mPAP (mmHg)",
                  "sPAP (mmHg)",
                  "PCWP (mmHg)",
                  "PVR (WU)",
                  "TAPSE (mm)",
                  "RVID (mm)",
                  "LVEF (%)",
                  "LVEDD (mm)"
                  )
pseudocounts_DCM <- pseudocounts[, metadata_DCM$Sample]
all.equal(metadata_DCM$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM))) {
  metadata_subset <- metadata_DCM[complete.cases(metadata_DCM[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("CM_DE_Results_", colnames(metadata[i]), ".csv")), row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("CM_DE_Significant_", colnames(metadata[i]), ".csv")), row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"

  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  ylab = bquote(~-Log[10] ~ italic(FDR)),
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 0,
                  axisLabSize = 12,
                  captionLabSize = 0,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Cardiomyocytes", "\n", tested_param[i-6], "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp, keyvals)
```

```{r n_de_genes}
write.csv(n_de_genes, file = here::here(output_dir_data, "n_de_genes.csv"), row.names = F)
```

```{r Volcano_all_metadata, fig.height=9, fig.width=12}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]]) / 
  (volcano[[5]] | volcano[[6]] |  volcano[[7]] | volcano[[8]]))
```

```{r Volcano_PH, fig.height=4.5, fig.width=12}
print(volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]])
```

```{r Volcano_RV, fig.height=4.5, fig.width=6}
print(volcano[[5]] | volcano[[6]])
```

```{r Volcano_LV, fig.height=4.5, fig.width=6}
print(volcano[[7]] | volcano[[8]])
```

```{r UpSet_plot_DE_up, fig.height=4, fig.width=6}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- dplyr::filter(signif[[i]], logFC > 0)$gene
  names(gene_sets)[i] <- tested_param[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 1]

UpSet(m, column_title = "Cardiomyocytes - Upregulated genes", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r UpSet_plot_DE_down, fig.height=4, fig.width=6}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- dplyr::filter(signif[[i]], logFC < 0)$gene
  names(gene_sets)[i] <- tested_param[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 1]

UpSet(m, column_title = "Cardiomyocytes - Downregulated genes", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r rm upset, eval = FALSE}
rm(m, gene_sets)
```

## Over-representation analysis

```{r universe}
universe_genes <- results[["mPAP_mmHg"]]$gene
universe_entrez <- bitr(universe_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)$ENTREZID
```

```{r ENTREZID}
up_genes <- filter(signif[["mPAP_mmHg"]], logFC > 0.5)$gene
up_genes <- bitr(up_genes, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")$ENTREZID
down_genes <- filter(signif[["mPAP_mmHg"]], logFC < -0.5)$gene
down_genes <- bitr(down_genes, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")$ENTREZID
```

```{r reorder_GO_by_pvalue}
reorder_GO_by_pvalue <- function(enrichGO_result) {
  go_results <- enrichGO_result@result
  go_results_sorted <- go_results[order(go_results$p.adjust), ]
  enrichGO_result@result <- go_results_sorted
  return(enrichGO_result)
}
```

```{r save ORA_GO_mPAP, eval=FALSE}
mPAP_GO_up <- enrichGO(up_genes, OrgDb = "org.Hs.eg.db", universe = universe_entrez, ont = "ALL", readable = TRUE)
mPAP_GO_up <- gsfilter(mPAP_GO_up, by = 'Count', min = 3)
mPAP_GO_up <- reorder_GO_by_pvalue(mPAP_GO_up)
saveRDS(mPAP_GO_up, here::here(output_dir_data, "CM_ORA_GO_mPAP_up.rds"))

mPAP_GO_down <- enrichGO(down_genes, OrgDb = "org.Hs.eg.db", universe = universe_entrez, ont = "ALL", readable = TRUE)
mPAP_GO_down <- gsfilter(mPAP_GO_down, by = 'Count', min = 3)
mPAP_GO_down <- reorder_GO_by_pvalue(mPAP_GO_down)
saveRDS(mPAP_GO_down, here::here(output_dir_data, "CM_ORA_GO_mPAP_down.rds"))
```

```{r ORA_GO_mPAP, fig.height=8, fig.width=12}
mPAP_GO_up <- readRDS(here::here(output_dir_data, "CM_ORA_GO_mPAP_up.rds"))
mPAP_GO_down <- readRDS(here::here(output_dir_data, "CM_ORA_GO_mPAP_down.rds"))
nrow(mPAP_GO_up)
nrow(mPAP_GO_down)
p1 <- dotplot(mPAP_GO_up, showCategory = 10, title = paste0("GO - Over-representation analysis", "\nmPAP - Upnregulated genes"), label_format = 27, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"))
p2 <- dotplot(mPAP_GO_down, showCategory = 10, title = paste0("GO - Over-representation analysis", "\nmPAP - Downregulated genes"), label_format = 27, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"))
print(p1|p2)
```

```{r ORA_GO_mPAP_simplified, fig.height=8, fig.width=12}
mPAP_GO_simplified_up <- clusterProfiler::simplify(
  mPAP_GO_up,
  cutoff = 0.75,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)
mPAP_GO_simplified_down <- clusterProfiler::simplify(
  mPAP_GO_down,
  cutoff = 0.75,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)
p1 <- dotplot(mPAP_GO_simplified_up, showCategory = 10, title = paste0("GO - Over-representation analysis", "\nmPAP - Upnregulated genes"), label_format = 27, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"))
p2 <- dotplot(mPAP_GO_simplified_down, showCategory = 10, title = paste0("GO - Over-representation analysis", "\nmPAP - Downregulated genes"), label_format = 27, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"))
print(p1 | p2)
```

```{r save ORA_KEGG_mPAP, eval=FALSE}
mPAP_KEGG_up <- enrichKEGG(up_genes, organism = 'hsa', universe = universe_entrez)
mPAP_KEGG_up <- gsfilter(mPAP_KEGG_up, by = 'Count', min = 3)
saveRDS(mPAP_KEGG_up, here::here(output_dir_data, "CM_ORA_KEGG_mPAP_up.rds"))
mPAP_KEGG_down <- enrichKEGG(down_genes,  organism = 'hsa', universe = universe_entrez)
mPAP_KEGG_down <- gsfilter(mPAP_KEGG_down, by = 'Count', min = 3)
saveRDS(mPAP_KEGG_down, here::here(output_dir_data, "CM_ORA_KEGG_mPAP_down.rds"))
```

```{r ORA_KEGG_mPAP, fig.height=8, fig.width=6}
mPAP_KEGG_up <- readRDS(here::here(output_dir_data, "CM_ORA_KEGG_mPAP_up.rds"))
mPAP_KEGG_down <- readRDS(here::here(output_dir_data, "CM_ORA_KEGG_mPAP_down.rds"))
nrow(mPAP_KEGG_up)
nrow(mPAP_KEGG_down)
p1 <- dotplot(mPAP_KEGG_down, showCategory = 10, title = paste0("KEGG Pathways", "\nmPAP - Downregulated genes"), label_format = 27, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"))
print(p1)
```

```{r save ORA_REACTOME_mPAP, eval=FALSE}
mPAP_REACTOME_up <- enrichPathway(up_genes, organism = 'human', universe = universe_entrez, readable = TRUE)
mPAP_REACTOME_up <- gsfilter(mPAP_REACTOME_up, by = 'Count', min = 3)
saveRDS(mPAP_REACTOME_up, here::here(output_dir_data, "CM_ORA_REACTOME_mPAP_up.rds"))
mPAP_REACTOME_down <- enrichPathway(down_genes, organism = 'human', universe = universe_entrez, readable = TRUE)
mPAP_REACTOME_down <- gsfilter(mPAP_REACTOME_down, by = 'Count', min = 3)
saveRDS(mPAP_REACTOME_up, here::here(output_dir_data, "CM_ORA_REACTOME_mPAP_down.rds"))
```

```{r ORA_REACTOME_mPAP, fig.height=8, fig.width=6}
mPAP_REACTOME_up <- readRDS(here::here(output_dir_data, "CM_ORA_REACTOME_mPAP_up.rds"))
mPAP_REACTOME_down <- readRDS(here::here(output_dir_data, "CM_ORA_REACTOME_mPAP_down.rds"))
nrow(mPAP_REACTOME_up)
nrow(mPAP_REACTOME_down)
```

```{r rm_GO, eval = FALSE}
rm(mPAP_GO_up, mPAP_GO_down, mPAP_GO_simplified_up, mPAP_GO_simplified_down, mPAP_KEGG_up, mPAP_KEGG_down, mPAP_REACTOME_up, mPAP_REACTOME_down)
```


## Gene set enrichment analysis

```{r ranked_genes}
results_mPAP <- results[["mPAP_mmHg"]]
results_mPAP$metric <- results_mPAP$logFC*-log10(results_mPAP$PValue)
entrezid <- bitr(results_mPAP$gene, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db", drop = TRUE)
results_mPAP <- merge(results_mPAP, entrezid, by.x = "gene", by.y = "SYMBOL", all.x = TRUE)
results_mPAP[duplicated(results_mPAP$metric) | duplicated(results_mPAP$metric, fromLast = TRUE),]
results_mPAP <- results_mPAP[!duplicated(results_mPAP$metric), ]
write.csv(results_mPAP, file = here::here(output_dir_data, "CM_results_mPAP.csv"), row.names = F)
reactable(results_mPAP, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
ranked_genes <- results_mPAP[!is.na(results_mPAP$ENTREZID), ]
genelist <- ranked_genes$metric
names(genelist) <-ranked_genes$ENTREZID
genelist <- genelist[order(genelist, decreasing = T)]
```

```{r save GSEA_GO, eval=FALSE}
GSEA_GO <- gseGO(geneList = genelist,
                OrgDb = org.Hs.eg.db,
                ont = "ALL",
                keyType = "ENTREZID",
                nPermSimple = 100000,
                minGSSize = 10)
GSEA_GO <- reorder_GO_by_pvalue(GSEA_GO)

saveRDS(GSEA_GO, here::here(output_dir_data, "CM_GSEA_GO_mPAP.rds"))
```

```{r GSEA_GO, fig.height=8, fig.width=8}
GSEA_GO <- readRDS(here::here(output_dir_data, "CM_GSEA_GO_mPAP.rds"))

dotplot(GSEA_GO, showCategory = 10, split=".sign", title = paste0("GO Enrichment Analysis", "\n mPAP-associated genes"), label_format = 30, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
```

```{r classify_terms}
classify_terms <- function(df) {
  df$sign <- ifelse(df$NES > 0, "activated", "suppressed")
  return(df)
}
```

```{r GSEA_GO_emapplot, fig.height=10, fig.width=20}
GSEA_GO@result <- classify_terms(GSEA_GO@result)

activated_terms <- subset(GSEA_GO@result, sign == "activated")
suppressed_terms <- subset(GSEA_GO@result, sign == "suppressed")

GSEA_GO_activated <- GSEA_GO
GSEA_GO_activated@result <- activated_terms
GSEA_GO_activated <- pairwise_termsim(GSEA_GO_activated)

GSEA_GO_suppressed <- GSEA_GO
GSEA_GO_suppressed@result <- suppressed_terms
GSEA_GO_suppressed <- pairwise_termsim(GSEA_GO_suppressed)

p1 <- emapplot(GSEA_GO_activated, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

p2 <- emapplot(GSEA_GO_suppressed, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Suppressed Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

combined_plot <- p1 | p2
print(combined_plot)
```

```{r GSEA_GO_simplified, fig.height=8, fig.width=8}
GSEA_GO_simplified <- clusterProfiler::simplify(
  GSEA_GO,
  cutoff = 0.75,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)
p1 <- dotplot(GSEA_GO_simplified, showCategory = 15, split=".sign", title = paste0("GO Enrichment Analysis", "\n mPAP-associated genes"), label_format = 35, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
print(p1)
```


```{r GSEA_GO_simplified_emapplot, fig.height=8, fig.width=16}
GSEA_GO_simplified@result <- classify_terms(GSEA_GO_simplified@result)

activated_terms <- subset(GSEA_GO_simplified@result, sign == "activated")
suppressed_terms <- subset(GSEA_GO_simplified@result, sign == "suppressed")

GSEA_GO_activated <- GSEA_GO_simplified
GSEA_GO_activated@result <- activated_terms
GSEA_GO_activated <- pairwise_termsim(GSEA_GO_activated)

GSEA_GO_suppressed <- GSEA_GO_simplified
GSEA_GO_suppressed@result <- suppressed_terms
GSEA_GO_suppressed <- pairwise_termsim(GSEA_GO_suppressed)

p1 <- emapplot(GSEA_GO_activated, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

p2 <- emapplot(GSEA_GO_suppressed, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Suppressed Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

combined_plot <- p1 | p2
print(combined_plot)
```

```{r GSEA_GO_simplified_emapplot_unlabeled, fig.height=8, fig.width=14}
emapplot(GSEA_GO_activated, showCategory = 100, cex_label_category = 0, min_edge = 0.15) +
  ggtitle("Gene Ontology Enrichment Analysis\nHigh mPAP-associated terms") + 
  theme(plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
    legend.title = element_text(size = 16),
    legend.text  = element_text(size = 14),
    legend.key.size = unit(0.9, "cm")        
  )
```

```{r GSEA_GO_simplified_ridgeplot, fig.height=12, fig.width=12}
p1 <- ridgeplot(GSEA_GO_simplified, showCategory = 30) + labs(x = "enrichment distribution")
print(p1)
```

```{r reorder_GO_by_NES}
reorder_GO_by_NES <- function(enrichGO_result) {
  go_results <- enrichGO_result@result
  go_results_sorted <- go_results[order(desc(go_results$NES)), ]
  enrichGO_result@result <- go_results_sorted
  return(enrichGO_result)
}
```

```{r default_labeller}
default_labeller <- function(n) {
    function(str){
        str <- gsub("_", " ", str)
        ep_str_wrap(str, n)
    }
}
```

```{r barplot.enrichResult}
barplot.enrichResult <- function (height, x = "Count", color = "p.adjust", showCategory = 8, 
  font.size = 12, title = "", label_format = 30, ...) 
{
  object <- height
  colorBy <- match.arg(color, c("pvalue", "p.adjust", "qvalue"))
  if (x == "geneRatio" || x == "GeneRatio") {
    x <- "GeneRatio"
  }
  else if (x == "count" || x == "Count") {
    x <- "Count"
  }
  df <- fortify(object, showCategory = showCategory, by = x, 
    ...)
  if (colorBy %in% colnames(df)) {
    p <- ggplot(df, aes_string(x = x, y = "Description", 
      fill = colorBy)) + theme_dose(font.size) + set_enrichplot_color(type = "fill", 
      name = color)
  }
  else {
    p <- ggplot(df, aes_string(x = x, y = "Description", 
      fill = "Description")) + theme_dose(font.size) + 
      theme(legend.position = "none")
  }
  label_func <- default_labeller(label_format)
  if (is.function(label_format)) {
    label_func <- label_format
  }
  p + geom_col() + scale_y_discrete(labels = label_func) + 
    ggtitle(title) + ylab(NULL)
}
```

```{r ep_str_wrap}
ep_str_wrap <- function(string, width) {
    x <- gregexpr(' ', string)
    vapply(seq_along(x),
           FUN = function(i) {
               y <- x[[i]]
               n <- nchar(string[i])
               len <- (c(y,n) - c(0, y)) ## length + 1
               idx <- len > width
               j <- which(!idx)
               if (length(j) && max(j) == length(len)) {
                   j <- j[-length(j)]
               }
               if (length(j)) {
                   idx[j] <- len[j] + len[j+1] > width
               }
               idx <- idx[-length(idx)] ## length - 1
               start <- c(1, y[idx] + 1)
               end <- c(y[idx] - 1, n)
               words <- substring(string[i], start, end)
               paste0(words, collapse="\n")
           },
           FUN.VALUE = character(1)
    )
}
```

```{r GSEA_GO_simplified_barplot, fig.height=12, fig.width=8}
barplot.enrichResult(reorder_GO_by_NES(GSEA_GO_simplified), x="NES", showCategory = 100, label_format = 45) + 
  ggtitle("Gene Ontology Enrichment Analysis\nHigh mPAP-associated terms") + 
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
```

```{r GSEA_GO_simplified_barplot_top20, fig.width=8, fig.height=5}
top_terms <- GSEA_GO_simplified@result %>%
  dplyr::arrange(desc(abs(NES))) %>%
  dplyr::slice(1:20)

GSEA_GO_top20 <- GSEA_GO_simplified
GSEA_GO_top20@result <- top_terms
barplot.enrichResult(reorder_GO_by_NES(GSEA_GO_top20), x='NES', showCategory = 100, label_format = 45) +  
  ggtitle("Gene Ontology Enrichment Analysis\nHigh mPAP-associated terms") + 
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
        axis.title.x  = element_text(size = 14)) + 
  labs(x = "Normalized Enrichment Score")
```

```{r GSEA_GO_simplified_treeplot, fig.height=12, fig.width=15}
GSEA_GO_simplified <- pairwise_termsim(GSEA_GO_simplified)
treeplot(GSEA_GO_simplified, hclust_method = "ward.D2", showCategory = 100, nCluster = 9) +  ggtitle("Gene Ontology Enrichment Analysis\nHigh mPAP-associated terms") + 
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
```

```{r save GSEA_KEGG, eval=FALSE}
GSEA_KEGG <- GSEA_KEGG(geneList = genelist,
                organism = "hsa",
                keyType = "kegg",
                nPermSimple = 100000)
saveRDS(GSEA_KEGG, here::here(output_dir_data, "CM_GSEA_KEGG_mPAP.rds"))
```

```{r GSEA_KEGG, fig.height=8, fig.width=8}
GSEA_KEGG <- readRDS(here::here(output_dir_data, "CM_GSEA_KEGG_mPAP.rds"))
p1 <- dotplot(GSEA_KEGG, showCategory = 10, split=".sign", title = paste0("GSEA\nKEGG Pathways", "\n mPAP-associated genes"), label_format = 35, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
print(p1)
```

```{r GSEA_KEGG_emapplot, fig.height=8, fig.width=16}
GSEA_KEGG@result <- classify_terms(GSEA_KEGG@result)

activated_terms <- subset(GSEA_KEGG@result, sign == "activated")
suppressed_terms <- subset(GSEA_KEGG@result, sign == "suppressed")

GSEA_KEGG_activated <- GSEA_KEGG
GSEA_KEGG_activated@result <- activated_terms
GSEA_KEGG_activated <- pairwise_termsim(GSEA_KEGG_activated)

GSEA_KEGG_suppressed <- GSEA_KEGG
GSEA_KEGG_suppressed@result <- suppressed_terms
GSEA_KEGG_suppressed <- pairwise_termsim(GSEA_KEGG_suppressed)

p1 <- emapplot(GSEA_KEGG_activated, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

p2 <- emapplot(GSEA_KEGG_suppressed, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Suppressed Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

combined_plot <- p1 | p2
print(combined_plot)
```

```{r GSEA_KEGG_emapplot_unlabeled, fig.height=8, fig.width=14}
emapplot(GSEA_KEGG_activated, showCategory = 100, cex_label_category = 0, min_edge = 0.15) +
  ggtitle("KEGG pathways Enrichment Analysis\nHigh mPAP-associated terms") + 
  theme(plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
    legend.title = element_text(size = 16),
    legend.text  = element_text(size = 14),
    legend.key.size = unit(0.9, "cm")        
  )
```

```{r GSEA_KEGG_ridgeplot, fig.height=12, fig.width=12}
p1 <- ridgeplot(GSEA_KEGG, showCategory = 30) + labs(x = "enrichment distribution")
print(p1)
```

```{r GSEA_KEGG_barplot, fig.height=5, fig.width=8}
barplot.enrichResult(reorder_GO_by_NES(GSEA_KEGG), x="NES", showCategory = 100, label_format = 45) +  ggtitle("KEGG pathways Enrichment Analysis\nHigh mPAP-associated terms") + 
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
        axis.title.x  = element_text(size = 14)) + 
  labs(x = "Normalized Enrichment Score")
```

```{r save GSEA_Reactome, eval=FALSE}
GSEA_Reactome <- gsePathway(geneList = genelist,
                organism = "human",
                nPermSimple = 100000,
                minGSSize = 10)
saveRDS(GSEA_Reactome, here::here(output_dir_data, "CM_GSEA_Reactome_mPAP.rds"))
```

```{r GSEA_Reactome, fig.height=8, fig.width=8}
GSEA_Reactome <- readRDS(here::here(output_dir_data, "CM_GSEA_Reactome_mPAP.rds"))
p1 <- dotplot(GSEA_Reactome, showCategory = 10, split=".sign", title = paste0("GSEA\nReactome Pathways", "\n mPAP-associated genes"), label_format = 40, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
print(p1)
```

```{r GSEA_Reactome_emapplot, fig.height=8, fig.width=16}
GSEA_Reactome@result <- classify_terms(GSEA_Reactome@result)

activated_terms <- subset(GSEA_Reactome@result, sign == "activated")
suppressed_terms <- subset(GSEA_Reactome@result, sign == "suppressed")

GSEA_GO_activated <- GSEA_Reactome
GSEA_GO_activated@result <- activated_terms
GSEA_GO_activated <- pairwise_termsim(GSEA_GO_activated)

GSEA_GO_suppressed <- GSEA_Reactome
GSEA_GO_suppressed@result <- suppressed_terms
GSEA_GO_suppressed <- pairwise_termsim(GSEA_GO_suppressed)

p1 <- emapplot(GSEA_GO_activated, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

p2 <- emapplot(GSEA_GO_suppressed, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Suppressed Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

combined_plot <- p1 | p2
print(combined_plot)
```
```{r GSEA_Reactome_ridgeplot, fig.height=12, fig.width=12}
p1 <- ridgeplot(GSEA_Reactome, showCategory = 30) + labs(x = "enrichment distribution")
print(p1)
```

```{r GSEA_Reactome_barplot_top20, fig.width=8, fig.height=5}
barplot.enrichResult(reorder_GO_by_NES(GSEA_Reactome), x='NES', showCategory = 100, label_format = 45) +  
  ggtitle("Reactome Pathways Enrichment Analysis\nHigh mPAP-associated terms") + 
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
        axis.title.x  = element_text(size = 14)) + 
  labs(x = "Normalized Enrichment Score")
```


```{r GO_similarity, fig.height=5, fig.width=7}
gseaGO_up_terms <- dplyr::filter(GSEA_GO@result, NES > 0)
table(gseaGO_up_terms$ONTOLOGY)
gseaGO_up_terms <- gseaGO_up_terms$ID
mat1 = GO_similarity(gseaGO_up_terms, ont = "BP")
BP <- simplifyGO(mat1, column_title = "57 GO Biological Process terms")
mat2 = GO_similarity(gseaGO_up_terms, ont = "MF")
MF <- simplifyGO(mat2, column_title = "17 GO Molecular Function terms")
mat3 = GO_similarity(gseaGO_up_terms, ont = "CC")
CC <- simplifyGO(mat3, column_title = "18 GO Cellular Component terms")
```

```{r GSEA_Combined}
GSEA_GO <- setReadable(GSEA_GO, OrgDb = org.Hs.eg.db)
GSEA_KEGG <- setReadable(GSEA_KEGG, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
GSEA_Reactome <- setReadable(GSEA_Reactome, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

GSEA_GO_df <- GSEA_GO@result
GSEA_KEGG_df <- GSEA_KEGG@result
GSEA_Reactome_df <- GSEA_Reactome@result

GSEA_KEGG_df$ONTOLOGY <- "KEGG"
GSEA_Reactome_df$ONTOLOGY <- "Reactome"


GSEA_Combined <- bind_rows(GSEA_GO_df, GSEA_KEGG_df, GSEA_Reactome_df)

write.csv(GSEA_Combined, here::here(output_dir_data, "CM_GSEA_ALL.csv"))
```

## Interesting genes

```{r pseudocounts_mPAP}
metadata_mPAP <- metadata[complete.cases(metadata[["mPAP_mmHg"]]),]

pseudocounts_mPAP <- pseudocounts[, metadata_mPAP$Sample]
pseudocounts_mPAP <- normLibSizes(pseudocounts_mPAP)

counts_mPAP <- cpm(pseudocounts_mPAP, log = TRUE)
```

```{r heatmap_cols, fig.width=10, fig.height=10}
heatmap_cols <- colorRampPalette(c("#004C8C", "white", "#A63A2A"))(101)

scales::show_col(heatmap_cols)
```

```{r metadata_heatmap}
metadata_heatmap <- metadata_mPAP %>% dplyr::select(Sample, mPAP_mmHg) %>% dplyr::arrange(mPAP_mmHg)
```

```{r ann_col}
ann_col <- metadata_mPAP %>%                 
  arrange(mPAP_mmHg) %>%                   
  dplyr::select(mPAP_mmHg)                   

ann_colors <- list(
  mPAP_mmHg = colorRampPalette(c("grey80", "grey20"))(100)
)
```

```{r autophagy_genes}
keywords <- "autophag|ampk"
filtered_terms <- GSEA_Combined %>%
  filter(str_detect(Description, regex(keywords, ignore_case = TRUE)))

autophagy_genes <- unique(unlist(strsplit(filtered_terms$core_enrichment, split = "/", fixed = TRUE)))
autophagy_selected <- c("ULK1", "WIPI1", "MAP1LC3A", "ATG4D", "ATG2A", "ATG7", "ATG4B")
```

```{r Heatmap_autophagy, fig.width=8, fig.height=3}
counts_heatmap <- counts_mPAP[autophagy_selected, ]
counts_heatmap <- counts_heatmap[, metadata_heatmap$Sample, drop = FALSE]

title <- textGrob("Autophagy", gp = gpar(fontsize = 18, fontface = "bold"))

heatmap <- pheatmap::pheatmap(
  name = " ",
  angle_col = "45",
  counts_heatmap,
  color = heatmap_cols,   
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,          
  annotation_legend = TRUE,          
  border_color = NA,
  fontsize = 10,
  fontsize_row = 13,
  fontsize_col = 10,
  cellheight = 15,
  gaps_col = c(0,0,0),
  silent = TRUE
  )

grid.arrange(grobs = list(title, heatmap[[4]]), heights = c(1, 5, 1))
```

```{r transport_genes}
keywords <- "endo|transp"
filtered_terms <- GSEA_Combined %>%
  filter(str_detect(Description, regex(keywords, ignore_case = TRUE)))

transport_genes <- unique(unlist(strsplit(filtered_terms$core_enrichment, split = "/", fixed = TRUE)))
transport_selected <- c("RAB11FIP3", "RAB35", "RAB7A", "RAB11FIP4", "RAB11FIP1", "RAB11B", "VPS37C", "VPS37B", "VPS45", "CHMP1A", "CHMP4B", "MVB12A", "DNM2", "CAV1", "EPS15L1", "EPN1", "EPN2")
```

```{r Heatmap_transport, fig.width=8, fig.height=5}
counts_heatmap <- counts_mPAP[transport_selected, ]
counts_heatmap <- counts_heatmap[, metadata_heatmap$Sample, drop = FALSE]

title <- textGrob("Vesicle trafficking", gp = gpar(fontsize = 18, fontface = "bold"))

heatmap <- pheatmap::pheatmap(
  name = " ",
  angle_col = "45",
  counts_heatmap,
  color = heatmap_cols,   
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,          
  annotation_legend = TRUE,          
  border_color = NA,
  fontsize = 10,
  fontsize_row = 13,
  fontsize_col = 10,
  cellheight = 15,
  gaps_col = c(0,0,0),
  silent = TRUE
  )

grid.arrange(grobs = list(title, heatmap[[4]]), heights = c(0.5, 5, 0.5))
```

```{r sarcomere_genes}
keywords <- "muscle|cardi|actin|heart|fiber|fibril|I band|Z disc|sarcomere|cytoskelet|myosin"
filtered_terms <- GSEA_Combined %>%
  filter(str_detect(Description, regex(keywords, ignore_case = TRUE)))

sarcomere_genes <- unique(unlist(strsplit(filtered_terms$core_enrichment, split = "/", fixed = TRUE)))
sarcomere_selected <- c("ACTN1", "ACTN4", "FLNC", "MYL9", "MYH9", "MYH14", "MYO18B")
```

```{r Heatmap_sarcomere, fig.width=8, fig.height=3}
counts_heatmap <- counts_mPAP[sarcomere_selected, ]
counts_heatmap <- counts_heatmap[, metadata_heatmap$Sample, drop = FALSE]

title <- textGrob("Sarcomere and cytoskeleton remodeling", gp = gpar(fontsize = 18, fontface = "bold"))

heatmap <- pheatmap::pheatmap(
  name = " ",
  angle_col = "45",
  counts_heatmap,
  color = heatmap_cols,   
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,          
  annotation_legend = TRUE,          
  border_color = NA,
  fontsize = 10,
  fontsize_row = 13,
  fontsize_col = 10,
  cellheight = 15,
  gaps_col = c(0,0,0),
  silent = TRUE
  )

grid.arrange(grobs = list(title, heatmap[[4]]), heights = c(1, 5, 1))
```

```{r glucose_genes}
keywords <- "glucose"
filtered_terms <- GSEA_Combined %>%
  filter(str_detect(Description, regex(keywords, ignore_case = TRUE)))

glucose_genes <- unique(unlist(strsplit(filtered_terms$core_enrichment, split = "/", fixed = TRUE)))
glucose_selected <- c("HK2", "GPT", "GPT2", "PDHA1", "H6PD", "ACACB")
```

```{r Heatmap_glucose, fig.width=8, fig.height=3}
counts_heatmap <- counts_mPAP[glucose_selected, ]
counts_heatmap <- counts_heatmap[, metadata_heatmap$Sample, drop = FALSE]

title <- textGrob("Glucose metabolism", gp = gpar(fontsize = 18, fontface = "bold"))

heatmap <- pheatmap::pheatmap(
  name = " ",
  angle_col = "45",
  counts_heatmap,
  color = heatmap_cols,   
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,          
  annotation_legend = TRUE,          
  border_color = NA,
  fontsize = 10,
  fontsize_row = 13,
  fontsize_col = 10,
  cellheight = 15,
  gaps_col = c(0,0,0),
  silent = TRUE
  )

grid.arrange(grobs = list(title, heatmap[[4]]), heights = c(1.5, 5, 1.5))
```

```{r adrenergic_genes}
keywords <- "adrenergic"
filtered_terms <- GSEA_Combined %>%
  filter(str_detect(Description, regex(keywords, ignore_case = TRUE)))

adrenergic_genes <- unique(unlist(strsplit(filtered_terms$core_enrichment, split = "/", fixed = TRUE)))
adrenergic_selected <- c("CACNA1C", "CAMK2A", "CAMK2B", "ATP2A2", "ATP1B3", "ATP1A3", "KCNQ1", "KCNE1")
```

```{r Heatmap_adrenergic, fig.width=8, fig.height=3}
counts_heatmap <- counts_mPAP[adrenergic_selected, ]
counts_heatmap <- counts_heatmap[, metadata_heatmap$Sample, drop = FALSE]

title <- textGrob("Adrenergic signaling", gp = gpar(fontsize = 18, fontface = "bold"))

heatmap <- pheatmap::pheatmap(
  name = " ",
  angle_col = "45",
  counts_heatmap,
  color = heatmap_cols,   
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,          
  annotation_legend = TRUE,          
  border_color = NA,
  fontsize = 10,
  fontsize_row = 13,
  fontsize_col = 10,
  cellheight = 15,
  gaps_col = c(0,0,0),
  silent = TRUE
  )

grid.arrange(grobs = list(title, heatmap[[4]]), heights = c(1, 5, 0.5))
```

```{r all_genes}
all_selected <- c(sarcomere_selected, autophagy_selected, adrenergic_selected, glucose_selected, transport_selected)
length(sarcomere_selected)
length(autophagy_selected)
length(adrenergic_selected)
length(glucose_selected)
length(transport_selected)
```

```{r Heatmap_all, fig.width=8, fig.height=12}
counts_heatmap <- counts_mPAP[all_selected, ]
counts_heatmap <- counts_heatmap[, metadata_heatmap$Sample, drop = FALSE]

pheatmap::pheatmap(
  name = " ",
  angle_col = "45",
  counts_heatmap,
  color = heatmap_cols,   
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,          
  annotation_legend = TRUE,          
  border_color = NA,
  fontsize = 16,
  fontsize_row = 16,
  fontsize_col = 12,
  cellheight = 15,
  cellwidth = 15,
  gaps_col = c(0,0,0,0),
  gaps_row = c(7,7,
               14,14,
               22,22,
               28,28)
  )
```

```{r lm_scores_mPAP, fig.width=9, fig.height=6}
pathways <- list(
  "Cytoskeleton and sarcomere remodeling" = sarcomere_genes,
  "Autophagy" = autophagy_genes,
  "Adrenergic signaling" = adrenergic_genes,
  "Glucose metabolism" = glucose_genes,
  "Vesicle transport" = transport_genes)

plots <- list()

for (pathway_name in names(pathways)) {
  genes <- pathways[[pathway_name]]
  gene_counts <- counts_mPAP[rownames(counts_mPAP) %in% genes, ]
  z_scores <- t(scale(t(gene_counts)))
  z_scores[is.nan(z_scores)] <- 0
  wrapped_name <- str_wrap(pathway_name, width = 20)
  gene_score <- colMeans(z_scores)
  gene_score_df <- data.frame(Sample = colnames(z_scores), Score = gene_score, mPAP = metadata_mPAP$mPAP_mmHg)
  lm_model <- lm(Score ~ mPAP, data = gene_score_df)
  
  lm_summary <- summary(lm_model)
  r_squared <- lm_summary$r.squared
  p_value <- lm_summary$coefficients[2, 4]

  plot <- ggplot(gene_score_df, aes(x = mPAP, y = Score)) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE, color = "firebrick", size = 0.4) + 
    theme_minimal() +
    labs(
      title = paste(wrapped_name, "\nR² =", round(r_squared, 3), "p =", round(p_value, 5)),
      x = "mPAP (mmHg)", 
      y = "Gene Set Score"
    )
  plots[[pathway_name]] <- plot
}

grid.arrange(grobs = plots, ncol = 3)
```

```{r lm_scores_mPAP_2, fig.width=15, fig.height=3}
grid.arrange(grobs = plots, ncol = 5)
```

## Heatmaps

```{r Heatmap_DEGs_mPAP, fig.width=8, fig.height=10}
counts_heatmap <- counts_mPAP[arrange(signif[["mPAP_mmHg"]], logFC)$gene, ]
counts_heatmap <- counts_heatmap[, metadata_heatmap$Sample, drop = FALSE]

pheatmap::pheatmap(
  name = " ",
  angle_col = "45",
  counts_heatmap,
  color = heatmap_cols,   
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE, 
  show_rownames = FALSE,
  annotation_legend = TRUE,          
  border_color = NA,
  fontsize = 10,
  fontsize_row = 13,
  fontsize_col = 10,
  cellheight = 0.2,
  gaps_col = c(0,0,0)
  )
```

```{r Heatmap_top_DEGs_mPAP, fig.width=10, fig.height=5}
DEGs_mPAP_top_up <- ranked_genes %>% 
  dplyr::arrange(desc(metric)) %>%
  slice_head(n=10)
DEGs_mPAP_top_down <- ranked_genes %>% 
  dplyr::arrange(metric) %>%
  slice_head(n=10)

DEGs_mPAP_top <- dplyr::bind_rows(DEGs_mPAP_top_up, DEGs_mPAP_top_down)
rm(DEGs_mPAP_top_up, DEGs_mPAP_top_down)

counts_heatmap <- counts_mPAP[DEGs_mPAP_top$gene, ]
counts_heatmap <- counts_heatmap[, metadata_heatmap$Sample, drop = FALSE]

pheatmap::pheatmap(
  name = " ",
  angle_col = "45",
  counts_heatmap,
  color = heatmap_cols,   
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE, 
  show_rownames = TRUE,
  annotation_legend = TRUE,          
  border_color = NA,
  fontsize = 10,
  fontsize_row = 13,
  fontsize_col = 10,
  cellheight = 15,
  gaps_col = c(0,0,0)
  )
```

## Transcription factor inference analysis

```{r net}
net <- get_collectri(organism='human', split_complexes=FALSE)
```

```{r ranked_genes_TF}
ranked_genes_TF <- ranked_genes %>%
  dplyr::select(gene, logFC, metric, PValue) %>% 
  remove_rownames() %>%
  column_to_rownames(var = "gene") %>%
  as.matrix()
```

```{r contrast_acts}
contrast_acts <- run_ulm(mat=ranked_genes_TF[, 'metric', drop=FALSE], net=net, .source='source', .target='target', .mor='mor', minsize = 10)
```

```{r mPAP_TF_barplot, fig.height=6, fig.width=12}
n_tfs <- 40

f_contrast_acts <- contrast_acts %>%
  mutate(rnk = NA)
msk <- f_contrast_acts$score > 0
f_contrast_acts[msk, 'rnk'] <- rank(-f_contrast_acts[msk, 'score'])
f_contrast_acts[!msk, 'rnk'] <- rank(-abs(f_contrast_acts[!msk, 'score']))
tfs <- f_contrast_acts %>%
  arrange(rnk) %>%
  head(n_tfs) %>%
  pull(source)
f_contrast_acts <- f_contrast_acts %>%
  filter(source %in% tfs)

ggplot(f_contrast_acts, aes(x = reorder(source, score), y = score)) + 
    geom_bar(aes(fill = score), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("TFs")
```

```{r FeaturePlot_TF, fig.width=24, fig.height=15}
FeaturePlot(CM, features = f_contrast_acts$source, ncol = 8)
```

```{r contrast_acts_filtered}
contrast_acts_filtered <- contrast_acts %>%
  filter(source %in% filter(percent_stats, percentCells > 5)$gene)
```

```{r TF_cols}
TF_cols <- rev(colorRampPalette(brewer.pal(5, "BrBG"))(101))
```

```{r TF_barplot_filtered, fig.height=5, fig.width=8}
n_tfs <- 40
f_contrast_acts <- contrast_acts_filtered %>%
  mutate(rnk = NA) %>%
  filter(p_value < 0.05)
msk <- f_contrast_acts$score > 0
f_contrast_acts[msk, 'rnk'] <- rank(-f_contrast_acts[msk, 'score'])
f_contrast_acts[!msk, 'rnk'] <- rank(-abs(f_contrast_acts[!msk, 'score']))
tfs <- f_contrast_acts %>%
  arrange(rnk) %>%
  head(n_tfs) %>%
  pull(source)
f_contrast_acts <- f_contrast_acts %>%
  filter(source %in% tfs) 

ggplot(f_contrast_acts, aes(x = reorder(source, score), y = score)) + 
    geom_bar(aes(fill = score), stat = "identity") +
    scale_fill_gradientn(colors = TF_cols) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 18),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =15, face= "plain", colour = "black"),
        axis.text.y = element_text(size =15, face= "plain", colour = "black"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Transcription Factor")
```

```{r TF_MEF2A, fig.height=3, fig.width=5}
tf <- 'MEF2A'

df <- net %>%
  filter(source == tf) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(ranked_genes_TF),rownames(df)))
df <- df[inter, ]
df[,c('logfc', 't_value', 'p_value')] <- ranked_genes_TF[inter, ]
df <- df %>%
  mutate(color = if_else(mor > 0 & t_value > 0, '1', color)) %>%
  mutate(color = if_else(mor > 0 & t_value < 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value > 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value < 0, '1', color))

ggplot(df, aes(logfc, -log10(p_value), colour = color)) +
  geom_point(size = 1) +                
  scale_colour_manual(values = c("#A63A2A", "#004C8C", "white")) +
  geom_label_repel(aes(label = ID), size = 2.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ggtitle(tf)
```

```{r TF_SRF, fig.height=3, fig.width=5}
tf <- 'SRF'

df <- net %>%
  filter(source == tf) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(ranked_genes_TF),rownames(df)))
df <- df[inter, ]
df[,c('logfc', 't_value', 'p_value')] <- ranked_genes_TF[inter, ]
df <- df %>%
  mutate(color = if_else(mor > 0 & t_value > 0, '1', color)) %>%
  mutate(color = if_else(mor > 0 & t_value < 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value > 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value < 0, '1', color))

ggplot(df, aes(logfc, -log10(p_value), colour = color)) +
  geom_point(size = 1) +                 
  scale_colour_manual(values = c("#A63A2A", "#004C8C", "white")) +
  geom_label_repel(aes(label = ID), size = 2.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ggtitle(tf)
```

https://www.nature.com/articles/s41419-023-05665-8

```{r TF_HIF1A, fig.height=3, fig.width=5}
tf <- 'HIF1A'

df <- net %>%
  filter(source == tf) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(ranked_genes_TF),rownames(df)))
df <- df[inter, ]
df[,c('logfc', 't_value', 'p_value')] <- ranked_genes_TF[inter, ]
df <- df %>%
  mutate(color = if_else(mor > 0 & t_value > 0, '1', color)) %>%
  mutate(color = if_else(mor > 0 & t_value < 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value > 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value < 0, '1', color))

ggplot(df, aes(logfc, -log10(p_value), colour = color)) +
  geom_point(size = 1) +              
  scale_colour_manual(values = c("#A63A2A", "#004C8C", "white")) +
  geom_label_repel(aes(label = ID), size = 2.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ggtitle(tf)
```

```{r TF_SP1, fig.height=3, fig.width=5}
tf <- 'SP1'

df <- net %>%
  filter(source == tf) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(ranked_genes_TF),rownames(df)))
df <- df[inter, ]
df[,c('logfc', 't_value', 'p_value')] <- ranked_genes_TF[inter, ]
df <- df %>%
  mutate(color = if_else(mor > 0 & t_value > 0, '1', color)) %>%
  mutate(color = if_else(mor > 0 & t_value < 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value > 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value < 0, '1', color))

ggplot(df, aes(logfc, -log10(p_value), colour = color)) +
  geom_point(size = 1) +              
  scale_colour_manual(values = c("#A63A2A", "#004C8C", "white")) +
  geom_label_repel(aes(label = ID), size = 2.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ggtitle(tf)
```

```{r TF_PPARG, fig.height=3, fig.width=5}
tf <- 'PPARG'

df <- net %>%
  filter(source == tf) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(ranked_genes_TF),rownames(df)))
df <- df[inter, ]
df[,c('logfc', 't_value', 'p_value')] <- ranked_genes_TF[inter, ]
df <- df %>%
  mutate(color = if_else(mor > 0 & t_value > 0, '1', color)) %>%
  mutate(color = if_else(mor > 0 & t_value < 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value > 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value < 0, '1', color))

ggplot(df, aes(logfc, -log10(p_value), colour = color)) +
  geom_point(size = 1) +               
  scale_colour_manual(values = c("#A63A2A", "#004C8C", "white")) +
  geom_label_repel(aes(label = ID), size = 2.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ggtitle(tf)
```

````{r TF_heatmap_Sample, fig.width=8, fig.height=5}
sample_acts <- run_ulm(mat=counts_mPAP, net=net, .source='source', .target='target',
                  .mor='mor', minsize = 10)

sample_acts_mat <- sample_acts %>%
  pivot_wider(id_cols = 'condition', names_from = 'source',
              values_from = 'score') %>%
  column_to_rownames('condition') %>%
  as.matrix()

tfs <- arrange(f_contrast_acts, desc(score))$source
sample_acts_mat <- sample_acts_mat[,tfs]

sample_acts_mat <- scale(sample_acts_mat)
sample_acts_mat <- t(sample_acts_mat)
sample_acts_mat <- sample_acts_mat[, metadata_heatmap$Sample, drop = FALSE]

my_breaks <- c(seq(-3, 0, length.out=ceiling(101/2) + 1),
               seq(0.05, 3, length.out=floor(101/2)))

pheatmap::pheatmap(
  name = " ",
  angle_col = "45",
  sample_acts_mat,
  breaks = my_breaks,
  color = TF_cols,   
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,          
  annotation_legend = TRUE,          
  border_color = NA,
  fontsize = 16,
  fontsize_row = 16,
  fontsize_col = 12,
  cellheight = 15,
  cellwidth = 15,
  gaps_col = c(0,0,0,0),
  gaps_row = c(11,11)
  )
```

```{r TF_tfsulm}
mat <- as.matrix(CM@assays$RNA@data)

acts <- decoupleR::run_ulm(mat = mat, 
                           net = net, 
                           .source = 'source', 
                           .target = 'target',
                           .mor='mor', 
                           minsize = 10)

CM[['tfsulm']] <- acts %>%
                    tidyr::pivot_wider(id_cols = 'source', 
                                       names_from = 'condition',
                                       values_from = 'score') %>%
                    tibble::column_to_rownames('source') %>%
                    Seurat::CreateAssayObject(.)

DefaultAssay(object = CM) <- "tfsulm"

CM <- Seurat::ScaleData(CM)
CM@assays$tfsulm@data <- CM@assays$tfsulm@scale.data
```

```{r TF_Feat_MEF2A, fig.width=12, fig.height=4}
p1 <- Seurat::DimPlot(CM, 
                      reduction = "umap", 
                      label = TRUE, 
                      pt.size = 0.1) + 
      Seurat::NoLegend() + 
      ggplot2::ggtitle('Cell states')

p2 <- Seurat::FeaturePlot(CM, features = c("MEF2A"), pt.size = 0.1) + 
      ggplot2::scale_colour_gradient2(low = TF_cols[1], high = TF_cols[101], mid = "white") +
      ggplot2::ggtitle('MEF2A activity')

DefaultAssay(object = CM) <- "RNA"
p3 <- Seurat::FeaturePlot(CM, 
                          features = c("MEF2A"), pt.size = 0.1) + 
      ggplot2::ggtitle('MEF2A expression')

Seurat::DefaultAssay(CM) <- "tfsulm"

p <- p1 | p2 | p3
p
```

```{r TF_heatmap_cluster_10, fig.width=4, fig.height=2.5}
n_tfs <- 10

contrast_acts_filtered <- contrast_acts %>%
  filter(source %in% filter(percent_stats, percentCells > 5)$gene)

df <- t(as.matrix(CM@assays$tfsulm@data)) %>%
      as.data.frame() %>%
      dplyr::mutate(cluster = Seurat::Idents(CM)) %>%
      tidyr::pivot_longer(cols = -cluster, 
                          names_to = "source", 
                          values_to = "score") %>%
      dplyr::group_by(cluster, source) %>%
      dplyr::summarise(mean = mean(score))
df <- df %>%
  filter(source %in% filter(percent_stats, percentCells > 5)$gene)

tfs <- df %>%
       dplyr::group_by(source) %>%
       dplyr::summarise(std = stats::sd(mean)) %>%
       dplyr::arrange(-abs(std)) %>%
       head(n_tfs) %>%
       dplyr::pull(source)

top_acts_mat <- df %>%
                dplyr::filter(source %in% tfs) %>%
                tidyr::pivot_wider(id_cols = 'cluster', 
                                   names_from = 'source',
                                   values_from = 'mean') %>%
                tibble::column_to_rownames('cluster') %>%
                as.matrix()

my_breaks <- c(seq(-2, 0, length.out = ceiling(101 / 2) + 1),
               seq(0.05, 2, length.out = floor(101 / 2)))

# Plot
pheatmap::pheatmap(mat = top_acts_mat,
                   color = TF_cols,
                   border_color = "white",
                   breaks = my_breaks,
                   cellwidth = 15,
                   cellheight = 15,
                   treeheight_row = 20,
                   treeheight_col = 20) 
```

```{r TF_heatmap_cluster_20, fig.width=6, fig.height=2.5}
n_tfs <- 20

contrast_acts_filtered <- contrast_acts %>%
  filter(source %in% filter(percent_stats, percentCells > 5)$gene)

df <- t(as.matrix(CM@assays$tfsulm@data)) %>%
      as.data.frame() %>%
      dplyr::mutate(cluster = Seurat::Idents(CM)) %>%
      tidyr::pivot_longer(cols = -cluster, 
                          names_to = "source", 
                          values_to = "score") %>%
      dplyr::group_by(cluster, source) %>%
      dplyr::summarise(mean = mean(score))
df <- df %>%
  filter(source %in% filter(percent_stats, percentCells > 5)$gene)

tfs <- df %>%
       dplyr::group_by(source) %>%
       dplyr::summarise(std = stats::sd(mean)) %>%
       dplyr::arrange(-abs(std)) %>%
       head(n_tfs) %>%
       dplyr::pull(source)

top_acts_mat <- df %>%
                dplyr::filter(source %in% tfs) %>%
                tidyr::pivot_wider(id_cols = 'cluster', 
                                   names_from = 'source',
                                   values_from = 'mean') %>%
                tibble::column_to_rownames('cluster') %>%
                as.matrix()

my_breaks <- c(seq(-2, 0, length.out = ceiling(100 / 2) + 1),
               seq(0.05, 2, length.out = floor(100 / 2)))

# Plot
pheatmap::pheatmap(mat = top_acts_mat,
                   color = TF_cols,
                   border_color = "white",
                   breaks = my_breaks,
                   cellwidth = 15,
                   cellheight = 15,
                   treeheight_row = 20,
                   treeheight_col = 20) 
```

## Pathway activity inference

```{r net_pathway}
net_path <- get_progeny(organism = 'human', top = 100000)
```

```{r contrast_acts_pathway}
contrast_acts <- run_ulm(mat=ranked_genes_TF[, 'metric', drop=FALSE], net=net_path, .source='source', .target='target', .mor='weight', minsize = 10)
```

```{r Pathway_cols}
Pathway_cols <- colorRampPalette(c("#28A2EB", "white", "#FFB233"))(101)
```

```{r Pathway_barplot, fig.height=5, fig.width=8}
ggplot(contrast_acts, aes(x = reorder(source, score), y = score)) + 
    geom_bar(aes(fill = score), stat = "identity") +
    scale_fill_gradientn(colors = Pathway_cols) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 18),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =15, face= "plain", colour = "black"),
        axis.text.y = element_text(size =15, face= "plain", colour = "black"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Pathway")
```

```{r Pathway_TGFb, fig.height=3, fig.width=5}
pathway <- 'TGFb'

df <- net_path %>%
  filter(source == pathway) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(ranked_genes_TF),rownames(df)))
df <- df[inter, ]
df['metric'] <- ranked_genes_TF[, 'metric', drop=FALSE][inter, ]
df <- df %>%
  mutate(color = if_else(weight > 0 & metric > 0, '1', color)) %>%
  mutate(color = if_else(weight > 0 & metric < 0, '2', color)) %>%
  mutate(color = if_else(weight < 0 & metric > 0, '2', color)) %>%
  mutate(color = if_else(weight < 0 & metric < 0, '1', color))

ggplot(df, aes(x = weight, y = metric, color = color)) +
  geom_point(size = 1) +                
  scale_colour_manual(values = c("#A63A2A", "#004C8C", "white")) +
  geom_label_repel(aes(label = ID), size = 2.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ggtitle(pathway)
```

```{r Pathway_Hypoxia, fig.height=3, fig.width=5}
pathway <- 'Hypoxia'

df <- net_path %>%
  filter(source == pathway) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(ranked_genes_TF),rownames(df)))
df <- df[inter, ]
df['metric'] <- ranked_genes_TF[, 'metric', drop=FALSE][inter, ]
df <- df %>%
  mutate(color = if_else(weight > 0 & metric > 0, '1', color)) %>%
  mutate(color = if_else(weight > 0 & metric < 0, '2', color)) %>%
  mutate(color = if_else(weight < 0 & metric > 0, '2', color)) %>%
  mutate(color = if_else(weight < 0 & metric < 0, '1', color))

ggplot(df, aes(x = weight, y = metric, color = color)) +
  geom_point(size = 1) +                
  scale_colour_manual(values = c("#A63A2A", "#004C8C", "white")) +
  geom_label_repel(aes(label = ID), size = 2.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ggtitle(pathway)
```

```{r Pathway_Trail, fig.height=3, fig.width=5}
pathway <- 'Trail'

df <- net_path %>%
  filter(source == pathway) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(ranked_genes_TF),rownames(df)))
df <- df[inter, ]
df['metric'] <- ranked_genes_TF[, 'metric', drop=FALSE][inter, ]
df <- df %>%
  mutate(color = if_else(weight > 0 & metric > 0, '1', color)) %>%
  mutate(color = if_else(weight > 0 & metric < 0, '2', color)) %>%
  mutate(color = if_else(weight < 0 & metric > 0, '2', color)) %>%
  mutate(color = if_else(weight < 0 & metric < 0, '1', color))

ggplot(df, aes(x = weight, y = metric, color = color)) +
  geom_point(size = 1) +                
  scale_colour_manual(values = c("#A63A2A", "#004C8C", "white")) +
  geom_label_repel(aes(label = ID), size = 2.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ggtitle(pathway)
```

```{r Pathway_JAK_STAT, fig.height=3, fig.width=5}
pathway <- 'JAK-STAT'

df <- net_path %>%
  filter(source == pathway) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(ranked_genes_TF),rownames(df)))
df <- df[inter, ]
df['metric'] <- ranked_genes_TF[, 'metric', drop=FALSE][inter, ]
df <- df %>%
  mutate(color = if_else(weight > 0 & metric > 0, '1', color)) %>%
  mutate(color = if_else(weight > 0 & metric < 0, '2', color)) %>%
  mutate(color = if_else(weight < 0 & metric > 0, '2', color)) %>%
  mutate(color = if_else(weight < 0 & metric < 0, '1', color))

ggplot(df, aes(x = weight, y = metric, color = color)) +
  geom_point(size = 1) +                
  scale_colour_manual(values = c("#A63A2A", "#004C8C", "white")) +
  geom_label_repel(aes(label = ID), size = 2.5) +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ggtitle(pathway)
```

````{r Pathway_heatmap_Sample, fig.width=8, fig.height=5}
sample_acts <- run_ulm(mat=counts_mPAP, net=net_path, .source='source', .target='target',
                  .mor='weight', minsize = 5)

sample_acts_mat <- sample_acts %>%
  pivot_wider(id_cols = 'condition', names_from = 'source',
              values_from = 'score') %>%
  column_to_rownames('condition') %>%
  as.matrix()

sample_acts_mat <- scale(sample_acts_mat)
sample_acts_mat <- t(sample_acts_mat)
sample_acts_mat <- sample_acts_mat[, metadata_heatmap$Sample, drop = FALSE]

my_breaks <- c(seq(-3, 0, length.out=ceiling(101/2) + 1),
               seq(0.05, 3, length.out=floor(101/2)))

ordered_sources <- contrast_acts %>%
  arrange(desc(score)) %>%
  pull(source)

sample_acts_mat <- sample_acts_mat[ordered_sources, , drop = FALSE]

pheatmap::pheatmap(
  name = " ",
  angle_col = "45",
  sample_acts_mat,
  breaks = my_breaks,
  color = Pathway_cols,   
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,          
  annotation_legend = TRUE,          
  border_color = NA,
  fontsize = 16,
  fontsize_row = 16,
  fontsize_col = 12,
  cellheight = 15,
  cellwidth = 15,
  gaps_col = c(0,0,0,0),
  gaps_row = c(3,3)
  )
```

```{r Pathway_pathwaysmlm}
acts_pathway <- decoupleR::run_mlm(mat = mat, 
                           net = net_path, 
                           .source = 'source', 
                           .target = 'target',
                           .mor = 'weight', 
                           minsize = 10)

CM[['pathwaysmlm']] <- acts_pathway %>%
                         tidyr::pivot_wider(id_cols = 'source', 
                                            names_from = 'condition',
                                            values_from = 'score') %>%
                         tibble::column_to_rownames(var = 'source') %>%
                         Seurat::CreateAssayObject(.)

Seurat::DefaultAssay(object = CM) <- "pathwaysmlm"

CM <- Seurat::ScaleData(CM)
CM@assays$pathwaysmlm@data <- CM@assays$pathwaysmlm@scale.data
```

```{r Pathway_Feat_TGFb, fig.width=8, fig.height=4}
p1 <- Seurat::DimPlot(CM, 
                      reduction = "umap", 
                      label = TRUE, 
                      pt.size = 0.11) + 
      Seurat::NoLegend() + 
      ggplot2::ggtitle('Cell states')

p2 <- Seurat::FeaturePlot(CM, features = c("TGFb"), 
                      pt.size = 0.1) + 
      ggplot2::scale_colour_gradientn(colors = Pathway_cols
                                      ) +
      ggplot2::ggtitle('TGFb activity')

p <- p1 | p2
p
```

```{r Pathway_heatmap_cluster, fig.width=6, fig.height=2.5}
df_pathway <- t(as.matrix(CM@assays$pathwaysmlm@data)) %>%
      as.data.frame() %>%
      dplyr::mutate(cluster = Seurat::Idents(CM)) %>%
      tidyr::pivot_longer(cols = -cluster, 
                          names_to = "source", 
                          values_to = "score") %>%
      dplyr::group_by(cluster, source) %>%
      dplyr::summarise(mean = mean(score))

top_acts_mat <- df_pathway %>%
                tidyr::pivot_wider(id_cols = 'cluster', 
                                   names_from = 'source',
                                   values_from = 'mean') %>%
                tibble::column_to_rownames(var = 'cluster') %>%
                as.matrix()

my_breaks <- c(seq(-1.25, 0, length.out = ceiling(101 / 2) + 1),
               seq(0.05, 1.25, length.out = floor(101 / 2)))

pheatmap::pheatmap(mat = top_acts_mat,
                   color = Pathway_cols,
                   border_color = "white",
                   breaks = my_breaks,
                   cellwidth = 20,
                   cellheight = 20,
                   treeheight_row = 20,
                   treeheight_col = 20) 
```

