---
title: "Cardiomyocytes subclustering"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 600, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "Cardiomyocytes_subclustering"

# Load libraries
library(here)
library(readr)
library(readxl)
library(xlsx)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(ArchR)
library(clustree)
library(harmony)
library(gprofiler2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationHub)
library(ReactomePA)
library(speckle)

#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```


Load dataset and subset cardiomyocytes

```{r Load data, echo = TRUE, results= "hide"}
DCM_RV_integrated <- readRDS(here::here("output", "QC_integration_annotation", "DCM_RV_integrated_RNA.rds"))
DefaultAssay(DCM_RV_integrated) <- "RNA"
CM <- subset(DCM_RV_integrated, cell_type == "Cardiomyocytes")
```

## Integration and clustering

Re-normalize and re-scale the subsetted object using SCTransform

```{r SCTransform, eval=FALSE, message=FALSE}
CM <- SplitObject(CM, split.by = "Sample")
for (i in 1:length(CM)) {
CM[[i]] <- SCTransform(CM[[i]], vst.flavor = "v2", vars.to.regress = c("percent.mt", "percent.rp"), return.only.var.genes = FALSE)
}
```

PCA

```{r PCA, eval=FALSE}
features <- SelectIntegrationFeatures(object.list = CM, nfeatures = 3000)
CM <- merge(x = CM[[1]], y = CM[2:length(CM)], merge.data=TRUE)
VariableFeatures(CM) <- features
CM <- RunPCA(CM)
```

```{r save CM, eval=FALSE}
saveRDS(CM, 
        here::here(output_dir_data, "CM.rds"))
```

```{r read CM}
CM <- readRDS(here::here(output_dir_data, "CM.rds"))
```


```{r Elbow_CM, fig.width=4, fig.height=4}
ElbowPlot(CM, ndims = 50)
```

Reintegrate the samples using Harmony

```{r RunHarmony, eval=FALSE}
CM <- RunHarmony(CM, assay.use="SCT", group.by.vars = "Sample", dims.use = 1:50)
```

Clustering after integration

```{r clustering integrated, eval=FALSE}
DefaultAssay(CM) <- "SCT"
CM <- RunUMAP(CM, dims = 1:30, reduction = "harmony")
CM <- FindNeighbors(CM, dims = 1:30, reduction = "harmony")
CM <- FindClusters(CM, resolution = seq(0.1, 0.8, by=0.1))
```

```{r Clustree_CM, fig.width=6, fig.height=9}
clustree::clustree(CM@meta.data[,grep("SCT_snn_res", colnames(CM@meta.data))], prefix = "SCT_snn_res.")
```

```{r UMAPs_CM, fig.width=12, fig.height=10}
DimPlot(CM, reduction = "umap", shuffle = T,
        group.by = c("SCT_snn_res.0.2", "Sample", "Batch_processing", "Batch_sequencing"), ncol = 2)
```

```{r save CM 2, eval=FALSE}
saveRDS(CM, 
        here::here(output_dir_data, "CM.rds"))
```

Find the markers that characterize each cell population

```{r FindAllMarkers}
DefaultAssay(CM) <- "SCT"
CM <- PrepSCTFindMarkers(CM)
Idents(CM) <- "SCT_snn_res.0.2"
Markers <- FindAllMarkers(CM, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 0.25)
write.csv(Markers, here::here(output_dir_data, "CM_Markers_all.csv"))
Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
write.csv(Markers_top10, here::here(output_dir_data, "CM_Markers_top10.csv"))
Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
write.csv(Markers_top3, here::here(output_dir_data, "CM_Markers_top3.csv"))
```

```{r Heatmap_CM, fig.width=15, fig.height=10}
mapal <- colorRampPalette(RColorBrewer::brewer.pal(9,"RdBu"))(256)
mapal <- rev(mapal[1:256])
Heatmap <- DoHeatmap(subset(CM, cells = sample(colnames(CM), 20000)), label = FALSE, draw.line = F, features = Markers_top10$gene) +
  scale_fill_gradientn(colours = mapal) +
  theme(axis.text.y = element_text(size = 5)) +
  theme(plot.margin = unit(c(0.1, 0, 0, 0), 
                           "inches"))
Heatmap
```

Check QC parameters in each cluster

```{r QC_VlnPlots_CM, fig.width=8, fig.height=6}
VlnPlot(CM, features = c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.rp"), ncol = 2, pt.size = 0) + theme(axis.title.x = element_blank())
```

Check cell cycle genes

```{r Cell_cycle_UMAP, fig.width=5, fig.height=4}
DimPlot(CM, reduction = "umap", shuffle = T,
        group.by = "Phase")
```

## Cell type annotation

Annotate the clusters

```{r, eval=FALSE}
CM@meta.data[CM@meta.data$SCT_snn_res.0.2 %in% c("0"), "cell_state"] = "CM1"
CM@meta.data[CM@meta.data$SCT_snn_res.0.2 %in% c("1"), "cell_state"] = "CM2"
CM@meta.data[CM@meta.data$SCT_snn_res.0.2 %in% c("2"), "cell_state"] = "CM3"
CM@meta.data[CM@meta.data$SCT_snn_res.0.2 %in% c("3"), "cell_state"] = "CM4"
CM@meta.data[CM@meta.data$SCT_snn_res.0.2 %in% c("4"), "cell_state"] = "CM5"

CM$cell_state <- factor(CM$cell_state, levels = c("CM1", "CM2", "CM3","CM4","CM5"))
Idents(CM) <- CM$cell_state
```

```{r FindAllMarkers cell_state}
DefaultAssay(CM) <- "SCT"
Idents(CM) <- "cell_state"
Markers <- FindAllMarkers(CM, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 0.25)
write.csv(Markers, here::here(output_dir_data, "CM_Markers_all_cell_state.csv"))
Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
write.csv(Markers_top10, here::here(output_dir_data, "CM_Markers_top10_cell_state.csv"))
Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
write.csv(Markers_top3, here::here(output_dir_data, "CM_Markers_top3_cell_state.csv"))
```

```{r Heatmap_cell_state, fig.width=12, fig.height=8}
Heatmap <- DoHeatmap(subset(CM, cells = sample(colnames(CM), 20000)), label = TRUE, draw.line = F, features = Markers_top10$gene) + 
  scale_fill_gradientn(colours = mapal) +
  theme(text = element_text(size = 12), 
        axis.text.y = element_text(size = 8)) +
  theme(plot.margin = unit(c(0.2, 0, 0, 0), "inches")) +
  guides(fill = guide_colorbar(barwidth = 1, barheight = 4))
Heatmap
```

```{r UMAP_cell_state, fig.width=6.5, fig.height=6}
p <- DimPlot(CM, group.by = "cell_state", reduction = "umap", label = F) + 
  NoLegend() + 
  theme(axis.text=element_text(size=10, face = "bold"), axis.title = element_text(size = 14, face = "bold")) + 
  theme(plot.title = element_blank())
LabelClusters(p, id = "cell_state", fontface = "bold", size = 7, repel = T)
```

```{r DotPlot_markers, fig.width=6, fig.height=8}
DotPlot(CM, assay = "SCT", features = rev(Markers_top3$gene), dot.scale = 5, cluster.idents = FALSE) +
  RotatedAxis() +
  coord_flip() +
  theme(axis.title = element_blank(), axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 13), legend.text = element_text(size = 9), legend.title = element_text(size = 11), plot.margin = unit(c(0, 0, 0, 0.1), 
                           "inches"))
```

```{r VlnPlot_markers_CM, fig.width=12, fig.height=4}
VlnPlot(CM, features = c("LINC02388", "TNNT2", "ARHGAP24", "NRAP"), ncol = 4)
```

```{r FeatPlot_markers_CM, fig.width=12, fig.height=3}
FeaturePlot(CM, features = c("LINC02388", "TNNT2", "ARHGAP24", "NRAP"), ncol = 4)
```

```{r Barplot_CM_Sample, fig.height=6, fig.width=15, message=FALSE}
color_palette_all <- c("#F8766D", "#00C1AB", "#24B700", "#E18A00", "#00ACFC", "#D575FE", "#BE9C00", "#00BE70", 
                   "#8CAB00", "#00BBDA", "#FF65AC", "#8B93FF", "#EA8331")
color_palette <- c("#F8766D", "#24B700", "#E18A00", "#00ACFC", "#D575FE")

props <- getTransformedProps(CM$cell_state, CM$Sample, transform="logit")

par(mar = c(7, 5, 1, 8), xpd = TRUE)
barplot(props$Proportions, legend = FALSE, ylab = "Proportions", col = color_palette, 
        cex.names = 2, las = 2, font.lab = 2, font.axis = 2, cex.axis = 1.2, cex.lab = 2)
legend("topright", inset = c(-0.1, 0.1), legend = rownames(props$Proportions), fill = color_palette, bty = "n", cex = 2)
```

```{r Barplot_FB_Cell_cycle, fig.height=6, fig.width=8, message=FALSE}
color_palette_all <- c("#F8766D", "#00C1AB", "#24B700", "#E18A00", "#00ACFC", "#D575FE", "#BE9C00", "#00BE70", 
                   "#8CAB00", "#00BBDA", "#FF65AC", "#8B93FF", "#EA8331")
color_palette <- c("#F8766D", "#24B700", "#00ACFC")

props <- getTransformedProps(CM$Phase, CM$cell_state, transform="logit")

par(mar = c(5, 5, 3, 11), xpd = TRUE)
barplot(props$Proportions, legend = FALSE, ylab = "Proportions", col = color_palette, 
        cex.names = 2, las = 2, font.lab = 2, font.axis = 2, cex.axis = 1.2, cex.lab = 2)
legend("topright", inset = c(-0.35, 0.3), legend = rownames(props$Proportions), fill = color_palette, bty = "n", cex = 2)
```

```{r}
df <- Markers[,7:6]
dfsample <- split(df$gene,df$cluster)
length(dfsample)

dfsample$CM1 = bitr(dfsample$CM1, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
dfsample$CM2 = bitr(dfsample$CM2, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
dfsample$CM3 = bitr(dfsample$CM3, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
dfsample$CM4 = bitr(dfsample$CM4, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
dfsample$CM5 = bitr(dfsample$CM5, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")

#do the same here, a line like below for each cluster
genelist <- list("CM1" = dfsample$CM1$ENTREZID, 
                 "CM2" = dfsample$CM2$ENTREZID,
                 "CM3" = dfsample$CM3$ENTREZID,
                 "CM4" = dfsample$CM4$ENTREZID,
                 "CM5" = dfsample$CM5$ENTREZID
)

# Calculate gene counts across cells
gene_counts <- rowSums(CM@assays$RNA@counts > 0)

# Filter genes to include only those expressed in at least 3 cells
universe_genes <- names(gene_counts[gene_counts >= 3])
write.csv(as.data.frame(universe_genes), here::here(output_dir_data, "CM_universe_genes.csv"))

# Convert universe gene symbols to Entrez IDs
universe_entrez <- bitr(universe_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)$ENTREZID
write.csv(as.data.frame(universe_entrez), here::here(output_dir_data, "CM_universe_entrez.csv"))

```

```{r Pathways_GO_MF, fig.height=8, fig.width=6}
GO_MF <- compareCluster(geneCluster = genelist, fun = "enrichGO", ont = "MF",  OrgDb = "org.Hs.eg.db", universe = universe_entrez)
GO_MF <- gsfilter(GO_MF, by = 'Count', min = 3)
dotplot(GO_MF, label_format = 35, title = "GO - Molecular Function")
```

```{r Pathways_GO_MF_simplified, fig.height=8, fig.width=6}
GO_MF_simplified <- simplify(
  GO_MF,
  cutoff = 0.5,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)
dotplot(GO_MF_simplified, label_format = 35, title = "GO - Molecular Function")
```

```{r Pathways_GO_BP, fig.height=8, fig.width=6}
GO_BP <- compareCluster(geneCluster = genelist, fun = "enrichGO", ont = "BP", OrgDb = "org.Hs.eg.db", universe = universe_entrez)
GO_BP <- gsfilter(GO_BP, by = 'Count', min = 3)
dotplot(GO_BP, label_format = 35, title = "GO - Biological process")
```

```{r Pathways_GO_BP_simplified, fig.height=8, fig.width=6}
GO_BP_simplified <- simplify(
  GO_BP,
  cutoff = 0.5,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)
dotplot(GO_BP_simplified, label_format = 35, title = "GO - Biological process")
```

```{r Pathways_KEGG, fig.height=8, fig.width=6}
KEGG <- compareCluster(geneCluster = genelist, fun = "enrichKEGG", organism = 'hsa', universe = universe_entrez)
KEGG <- gsfilter(KEGG, by = 'Count', min = 3)
dotplot(KEGG, label_format = 35, title = "KEGG Pathways")
```

```{r Pathways_REACTOME, fig.height=8, fig.width=6}
REACTOME <- compareCluster(geneCluster = genelist, fun = "enrichPathway", organism = 'human', universe = universe_entrez)
REACTOME <- gsfilter(REACTOME, by = 'Count', min = 3)
dotplot(REACTOME, label_format = 40, title = "REACTOME", showCategory = 4)
```

```{r save CM final, eval = FALSE}
saveRDS(CM, 
        here::here(output_dir_data, "CM.rds"))
```
