---
title: "Differential abundance and differential expression analysis in cardiomyocytes"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 600, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "Cardiomyocytes_DA_DE"

# Load libraries
library(here)
library(readr)
library(readxl)
library(xlsx)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(ArchR)
library(clustree)
library(harmony)
library(gprofiler2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationHub)
library(ReactomePA)
library(statmod)
library(edgeR)
library(speckle)
library(EnhancedVolcano)
library(decoupleR)
library(OmnipathR)
library(dorothea)
library(enrichplot)
library(png)


#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```


Load cardiomyocytes dataset

```{r read CM}
CM <- readRDS(here::here("output", "Cardiomyocytes_subclustering", "CM.rds"))
```


## Differential abundance analysis

```{r Speckle_all}
differential_abundance <- data.frame()
metadata <- unique(subset(CM@meta.data, select = c(Sample:CO_l.min)))
samples_index <- match(metadata$Sample, names(table(CM$Sample)))
metadata <- metadata[samples_index, ]
rownames(metadata) <- metadata$Sample
metadata <- metadata[, -which(names(metadata) %in% c("LVID_mm", "RVIT_mm"))]
rm(samples_index)

for (i in 4:(length(metadata))) {
  metadata_subset <- metadata[complete.cases(metadata[[i]]),]

  data <- subset(CM, Sample %in% metadata_subset$Sample)
  props <- getTransformedProps(data$cell_state, data$Sample, transform="logit")
  
  
  design <- model.matrix(~ metadata_subset[[i]])
  fit <- lmFit(props$TransformedProps, design)
  fit <- eBayes(fit, robust=TRUE)
  differential_abundance_temp <- topTable(fit, n = Inf, coef = 2)
  differential_abundance_temp$metadata <- colnames(metadata_subset[i])
  differential_abundance_temp$cell_state <- rownames(differential_abundance_temp)
  differential_abundance <- rbind(differential_abundance, differential_abundance_temp)
  
  fit.prop <- lmFit(props$Proportions,design)
  fit.prop <- eBayes(fit.prop, robust=TRUE)
}
rownames(differential_abundance) <- NULL
differential_abundance <- dplyr::arrange(differential_abundance, adj.P.Val)
head(differential_abundance, 10)
```

```{r Speckle_mPAP, fig.height=4, fig.width=20}
metadata_subset <- metadata[complete.cases(metadata[["mPAP_mmHg"]]),]

data <- subset(CM, Sample %in% metadata_subset$Sample)
props <- getTransformedProps(data$cell_state, data$Sample, transform="logit")


design <- model.matrix(~ metadata_subset$mPAP_mmHg)
fit <- lmFit(props$TransformedProps, design)
fit <- eBayes(fit, robust=TRUE)

fit.prop <- lmFit(props$Proportions,design)
fit.prop <- eBayes(fit.prop, robust=TRUE)

par(mfrow=c(1,5))
for(i in seq(1,5,1)){
  plot(metadata_subset$mPAP_mmHg, props$Proportions[i,], main = rownames(props$Proportions)[i],
       pch=16, cex=2, ylab="Proportions", xlab="mPAP_mmHg", cex.lab=1.5, cex.axis=1.5,
       cex.main=2)
  abline(a=fit.prop$coefficients[i,1], b=fit.prop$coefficients[i,2], col=4,
         lwd=1)
}
```

```{r Speckle_PVR, fig.height=4, fig.width=20}
metadata_subset <- metadata[complete.cases(metadata[["PVR_WU"]]),]

data <- subset(CM, Sample %in% metadata_subset$Sample)
props <- getTransformedProps(data$cell_state, data$Sample, transform="logit")


design <- model.matrix(~ metadata_subset$PVR_WU)
fit <- lmFit(props$TransformedProps, design)
fit <- eBayes(fit, robust=TRUE)

fit.prop <- lmFit(props$Proportions,design)
fit.prop <- eBayes(fit.prop, robust=TRUE)

par(mfrow=c(1,5))
for(i in seq(1,5,1)){
  plot(metadata_subset$PVR_WU, props$Proportions[i,], main = rownames(props$Proportions)[i],
       pch=16, cex=2, ylab="Proportions", xlab="PVR_WU", cex.lab=1.5, cex.axis=1.5,
       cex.main=2)
  abline(a=fit.prop$coefficients[i,1], b=fit.prop$coefficients[i,2], col=4,
         lwd=1)
}
```

```{r Speckle_RVS, fig.height=4, fig.width=20}
metadata_subset <- metadata[complete.cases(metadata[["RVS_cm.s"]]),]

data <- subset(CM, Sample %in% metadata_subset$Sample)
props <- getTransformedProps(data$cell_state, data$Sample, transform="logit")


design <- model.matrix(~ metadata_subset$RVS_cm.s)
fit <- lmFit(props$TransformedProps, design)
fit <- eBayes(fit, robust=TRUE)

fit.prop <- lmFit(props$Proportions,design)
fit.prop <- eBayes(fit.prop, robust=TRUE)

par(mfrow=c(1,5))
for(i in seq(1,5,1)){
  plot(metadata_subset$RVS_cm.s, props$Proportions[i,], main = rownames(props$Proportions)[i],
       pch=16, cex=2, ylab="Proportions", xlab="RVS_cm.s", cex.lab=1.5, cex.axis=1.5,
       cex.main=2)
  abline(a=fit.prop$coefficients[i,1], b=fit.prop$coefficients[i,2], col=4,
         lwd=1)
}
```


## Differential expression analysis

Extract metadata table

```{r}
metadata <- unique(subset(CM@meta.data, select = c(Sample:CO_l.min)))
samples_index <- match(metadata$Sample, names(table(CM$Sample)))
metadata <- metadata[samples_index, ]
rownames(metadata) <- metadata$Sample
metadata <- metadata[, -which(names(metadata) %in% c("LVID_mm", "RVIT_mm"))]
rm(samples_index)
```

Remove genes expressed in <1% of cells and MT and RP genes

```{r}
percent_stats <- Percent_Expressing(seurat_object = CM, features = rownames(CM), assay = "RNA", entire_object = TRUE)
percent_stats$gene <- rownames(percent_stats)

keep_genes <- rownames(dplyr::filter(percent_stats, All_Cells > 1))

data <- CM[which(rownames(CM) %in% keep_genes),]

rm(keep_genes)
```

Create and prepare pseudobulk object for differential expression analysis

```{r}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
keep.samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep.samples]
keep.genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep.genes, ]
```

Create empty objects for the DE output

```{r}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r}
for (i in 5:(length(metadata))) {
  metadata_subset <- metadata[complete.cases(metadata[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i]] <- topTags(fit, n = Inf)$table
  results[[i]] <- merge(results[[i]], percent_stats[, c("gene", "All_Cells")], by = "gene", all.x = FALSE)
  signif[[i]] <- results[[i]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i] <- colnames(metadata_subset[i])
  names(signif)[i] <- colnames(metadata_subset[i])
  
  write.csv(results[[i]], file = here::here(output_dir_data, paste0("CM_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i]], file = here::here(output_dir_data, paste0("CM_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i]]$logFC < -0.5),
                                n_tot = nrow(signif[[i]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  volcano[[i]] <- EnhancedVolcano(results[[i]],
                  lab = results[[i]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 3.5,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  col = c("black", "black", "black", "red"),
                  colAlpha = 1,
                  drawConnectors = TRUE,
                  widthConnectors = 0.1,
                  boxedLabels = TRUE,
                  subtitle = NULL,
                  maxoverlapsConnectors = 100,
                  max.overlaps = 100,
                  title = paste0("Cardiomyocytes", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i] <- colnames(metadata_subset[i])
}
```

```{r Volcano_all_metadata, fig.height=12, fig.width=28, dpi=300}
combined_plots <- (volcano[[5]] | volcano[[6]] | volcano[[7]] | volcano[[8]] | volcano[[9]] | volcano[[10]] | volcano[[11]]) / 
  (volcano[[12]] | volcano[[13]] |  volcano[[14]] | volcano[[15]] | volcano[[16]] | volcano[[17]])
print(combined_plots)
```


## Over-representation analysis

```{r}
# Calculate gene counts across cells
gene_counts <- rowSums(CM@assays$RNA@counts > 0)

# Filter genes to include only those expressed in at least 3 cells
universe_genes <- names(gene_counts[gene_counts >= 3])
write.csv(as.data.frame(universe_genes), here::here(output_dir_data, "CM_universe_genes.csv"))

# Convert universe gene symbols to Entrez IDs
universe_entrez <- bitr(universe_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)$ENTREZID
write.csv(as.data.frame(universe_entrez), here::here(output_dir_data, "CM_universe_entrez.csv"))
```

```{r}
up_genes <- filter(signif[["mPAP_mmHg"]], logFC > 0.5)$gene
up_genes <- bitr(up_genes, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")$ENTREZID
down_genes <- filter(signif[["mPAP_mmHg"]], logFC < -0.5)$gene
down_genes <- bitr(down_genes, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")$ENTREZID
```

```{r mPAP_GO_MF, fig.height=8, fig.width=6}
mPAP_GO_MF_up <- enrichGO(up_genes, OrgDb = "org.Hs.eg.db", universe = universe_entrez, ont = "MF", readable = TRUE)
mPAP_GO_MF_up <- gsfilter(mPAP_GO_MF_up, by = 'Count', min = 3)
mPAP_GO_MF_down <- enrichGO(down_genes, OrgDb = "org.Hs.eg.db", universe = universe_entrez, ont = "MF", readable = TRUE)
mPAP_GO_MF_down <- gsfilter(mPAP_GO_MF_down, by = 'Count', min = 3)
p1 <- dotplot(mPAP_GO_MF_up, showCategory = 10, title = paste0("GO - Molecular Function", "\nHigh mPAP - associated genes"), label_format = 27, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"))
print(p1)
```

```{r mPAP_GO_MF_simplified, fig.height=8, fig.width=6}
mPAP_GO_MF_simplified_up <- simplify(
  mPAP_GO_MF_up,
  cutoff = 0.5,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)
p1 <- dotplot(mPAP_GO_MF_simplified_up, showCategory = 10, title = paste0("GO - Molecular Function", "\nHigh mPAP - associated genes"), label_format = 27, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"))
print(p1)
```

```{r mPAP_GO_BP, fig.height=8, fig.width=6}
mPAP_GO_BP_up <- enrichGO(up_genes, OrgDb = "org.Hs.eg.db", universe = universe_entrez, ont = "BP", readable = TRUE)
mPAP_GO_BP_up <- gsfilter(mPAP_GO_BP_up, by = 'Count', min = 3)
mPAP_GO_BP_down <- enrichGO(down_genes, OrgDb = "org.Hs.eg.db", universe = universe_entrez, ont = "BP", readable = TRUE)
mPAP_GO_BP_down <- gsfilter(mPAP_GO_BP_down, by = 'Count', min = 3)
p1 <- dotplot(mPAP_GO_BP_up, showCategory = 10, title = paste0("GO - Biological Process", "\nHigh mPAP - associated genes"), label_format = 27, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"))
print(p1)
```

```{r mPAP_GO_BP_simplified, fig.height=8, fig.width=6}
mPAP_GO_BP_simplified_up <- simplify(
  mPAP_GO_BP_up,
  cutoff = 0.5,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)
p1 <- dotplot(mPAP_GO_BP_simplified_up, showCategory = 10, title = paste0("GO - Biological Process", "\nHigh mPAP - associated genes"), label_format = 27, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"))
print(p1)
```

```{r mPAP_KEGG, fig.height=8, fig.width=6}
mPAP_KEGG_up <- enrichKEGG(up_genes, organism = 'hsa', universe = universe_entrez)
mPAP_KEGG_up <- gsfilter(mPAP_KEGG_up, by = 'Count', min = 3)
mPAP_KEGG_down <- enrichKEGG(down_genes,  organism = 'hsa', universe = universe_entrez)
mPAP_KEGG_down <- gsfilter(mPAP_KEGG_down, by = 'Count', min = 3)
p1 <- dotplot(mPAP_KEGG_up, showCategory = 10, title = paste0("KEGG Pathways", "\nHigh mPAP - associated genes"), label_format = 27, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"))
print(p1)
```

```{r mPAP_REACTOME, fig.height=8, fig.width=6}
mPAP_REACTOME_up <- enrichPathway(up_genes, organism = 'human', universe = universe_entrez, readable = TRUE, pvalueCutoff = 0.1)
mPAP_REACTOME_up <- gsfilter(mPAP_REACTOME_up, by = 'Count', min = 3)
mPAP_REACTOME_down <- enrichPathway(down_genes, organism = 'human', universe = universe_entrez, readable = TRUE, pvalueCutoff = 0.1)
mPAP_REACTOME_down <- gsfilter(mPAP_REACTOME_down, by = 'Count', min = 3)
p1 <- dotplot(mPAP_REACTOME_up, showCategory = 10, title = paste0("REACTOME", "\nHigh mPAP - associated genes"), label_format = 27, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"))
print(p1)
```


## Gene set enrichment analysis


```{r}
ranked_genes <- results[["mPAP_mmHg"]]
ranked_genes$metric <- ranked_genes$logFC*-log10(ranked_genes$PValue)
entrezid <- bitr(ranked_genes$gene, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db", drop = TRUE)
ranked_genes <- merge(ranked_genes, entrezid, by.x = "gene", by.y = "SYMBOL")
ranked_genes <- ranked_genes[!duplicated(ranked_genes$metric), ]
genelist <- ranked_genes$metric
names(genelist) <- ranked_genes$ENTREZID
genelist <- genelist[order(genelist, decreasing = T)]
```


```{r}
gseaGO_MF <- gseGO(geneList = genelist,
                OrgDb = org.Hs.eg.db,
                ont = "MF",
                keyType = "ENTREZID",
                nPermSimple = 1000000)
```

```{r mPAP_gseaGO_MF_simplified, fig.height=8, fig.width=8}
gseaGO_MF_simplified <- simplify(
  gseaGO_MF,
  cutoff = 0.7,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)
p1 <- dotplot(gseaGO_MF_simplified, showCategory = 10, split=".sign", title = paste0("GSEA\nGO - Molecular Function", "\n mPAP-associated genes"), label_format = 30, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
print(p1)
```

```{r mPAP_gseaGO_MF_emapplot, fig.height=6, fig.width=15}
classify_terms <- function(df) {
  df$sign <- ifelse(df$NES > 0, "activated", "suppressed")
  return(df)
}
gseaGO_MF@result <- classify_terms(gseaGO_MF@result)
# Split the GSEA results
activated_terms <- subset(gseaGO_MF@result, sign == "activated")
suppressed_terms <- subset(gseaGO_MF@result, sign == "suppressed")

# Create two separate enrichResult objects
gseaGO_MF_activated <- gseaGO_MF
gseaGO_MF_activated@result <- activated_terms

gseaGO_MF_suppressed <- gseaGO_MF
gseaGO_MF_suppressed@result <- suppressed_terms

# Plot activated terms
p1 <- emapplot(pairwise_termsim(gseaGO_MF_activated), showCategory = 20, cex_label_category = 0.7) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

# Plot suppressed terms
p2 <- emapplot(pairwise_termsim(gseaGO_MF_suppressed), showCategory = 20, cex_label_category = 0.7) +
  ggtitle("Suppressed Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

# Combine the plots
combined_plot <- p1 | p2
print(combined_plot)
```

```{r mPAP_gseaGO_MF_ridgeplot, fig.height=8, fig.width=12}
p1 <- ridgeplot(gseaGO_MF_simplified, showCategory = 20) + labs(x = "enrichment distribution")
print(p1)
```

```{r}
gseaGO_BP <- gseGO(geneList = genelist,
                OrgDb = org.Hs.eg.db,
                ont = "BP",
                keyType = "ENTREZID",
                nPermSimple = 1000000)
```

```{r mPAP_gseaGO_BP_simplified, fig.height=8, fig.width=8}
gseaGO_BP_simplified <- simplify(
  gseaGO_BP,
  cutoff = 0.7,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)
p1 <- dotplot(gseaGO_BP_simplified, showCategory = 10, split=".sign", title = paste0("GSEA\nGO - Biological Process", "\n mPAP-associated genes"), label_format = 30, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
print(p1)
```

```{r mPAP_gseaGO_BP_emapplot, fig.height=6, fig.width=15}
gseaGO_BP@result <- classify_terms(gseaGO_BP@result)
# Split the GSEA results
activated_terms <- subset(gseaGO_BP@result, sign == "activated")
suppressed_terms <- subset(gseaGO_BP@result, sign == "suppressed")

# Create two separate enrichResult objects
gseaGO_MF_activated <- gseaGO_BP
gseaGO_MF_activated@result <- activated_terms

gseaGO_MF_suppressed <- gseaGO_BP
gseaGO_MF_suppressed@result <- suppressed_terms

# Plot activated terms
p1 <- emapplot(pairwise_termsim(gseaGO_MF_activated), showCategory = 20, cex_label_category = 0.7) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

# Plot suppressed terms
p2 <- emapplot(pairwise_termsim(gseaGO_MF_suppressed), showCategory = 20, cex_label_category = 0.7) +
  ggtitle("Suppressed Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

# Combine the plots
combined_plot <- p1 | p2
print(combined_plot)
```

```{r mPAP_gseaGO_BP_ridgeplot, fig.height=8, fig.width=12}
p1 <- ridgeplot(gseaGO_BP_simplified, showCategory = 20) + labs(x = "enrichment distribution")
print(p1)
```

```{r}
gseaKEGG <- gseKEGG(geneList = genelist,
                organism = "hsa",
                keyType = "kegg",
                nPermSimple = 100000)
```

```{r gseaKEGG, fig.height=8, fig.width=8}
p1 <- dotplot(gseaKEGG, showCategory = 10, split=".sign", title = paste0("GSEA\nKEGG Pathways", "\n mPAP-associated genes"), label_format = 30, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
print(p1)
```

```{r mPAP_gseaKEGG_emapplot, fig.height=6, fig.width=15}
gseaKEGG@result <- classify_terms(gseaKEGG@result)
# Split the GSEA results
activated_terms <- subset(gseaKEGG@result, sign == "activated")
suppressed_terms <- subset(gseaKEGG@result, sign == "suppressed")

# Create two separate enrichResult objects
gseaKEGG_activated <- gseaKEGG
gseaKEGG_activated@result <- activated_terms

gseaKEGG_suppressed <- gseaKEGG
gseaKEGG_suppressed@result <- suppressed_terms

# Plot activated terms
p1 <- emapplot(pairwise_termsim(gseaKEGG_activated), showCategory = 20, cex_label_category = 0.7) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

# Plot suppressed terms
p2 <- emapplot(pairwise_termsim(gseaKEGG_suppressed), showCategory = 20, cex_label_category = 0.7) +
  ggtitle("Suppressed Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

# Combine the plots
combined_plot <- p1 | p2
print(combined_plot)
```

```{r mPAP_gseaKEGG_ridgeplot, fig.height=8, fig.width=12}
p1 <- ridgeplot(gseaKEGG, showCategory = 20) + labs(x = "enrichment distribution")
print(p1)
```


## Transcription factor inference analysis

```{r}
net <- get_collectri(organism='human', split_complexes=FALSE)
net
```

```{r}
metadata_subset <- metadata[complete.cases(metadata[["mPAP_mmHg"]]),]
metadata_subset[,i] <- rescale(metadata_subset[,"mPAP_mmHg"])
design <- model.matrix(~ metadata_subset[["mPAP_mmHg"]])
colnames(design)[2] <- colnames(metadata_subset["mPAP_mmHg"])
  
pseudocounts_subset <- pseudocounts[, metadata_subset$Sample]
pseudocounts_subset <- normLibSizes(pseudocounts_subset)
pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)

counts <- cpm(pseudocounts_subset, log = TRUE)
head(counts)
```

```{r}
deg <- ranked_genes %>%
  dplyr::select(gene, logFC, metric, PValue) %>% 
  remove_rownames() %>%
  column_to_rownames(var = "gene") %>%
  as.matrix()
head(deg)
```


```{r}
# Run ulm
contrast_acts <- run_ulm(mat=deg[, 'metric', drop=FALSE], net=net, .source='source', .target='target',
                  .mor='mor', minsize = 5)
head(contrast_acts)
```

```{r mPAP_TF barplot, fig.height=6, fig.width=12}
n_tfs <- 40

# Filter top TFs in both signs
f_contrast_acts <- contrast_acts %>%
  mutate(rnk = NA)
msk <- f_contrast_acts$score > 0
f_contrast_acts[msk, 'rnk'] <- rank(-f_contrast_acts[msk, 'score'])
f_contrast_acts[!msk, 'rnk'] <- rank(-abs(f_contrast_acts[!msk, 'score']))
tfs <- f_contrast_acts %>%
  arrange(rnk) %>%
  head(n_tfs) %>%
  pull(source)
f_contrast_acts <- f_contrast_acts %>%
  filter(source %in% tfs)

# Plot
ggplot(f_contrast_acts, aes(x = reorder(source, score), y = score)) + 
    geom_bar(aes(fill = score), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("TFs")
```

```{r TF_TCF21, fig.height=6, fig.width=9}
tf <- 'TCF21'

df <- net %>%
  filter(source == tf) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(deg),rownames(df)))
df <- df[inter, ]
df[,c('logfc', 't_value', 'p_value')] <- deg[inter, ]
df <- df %>%
  mutate(color = if_else(mor > 0 & t_value > 0, '1', color)) %>%
  mutate(color = if_else(mor > 0 & t_value < 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value > 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value < 0, '1', color))

ggplot(df, aes(x = logfc, y = -log10(p_value), color = color, size=abs(mor))) +
  geom_point() +
  scale_colour_manual(values = c("red","royalblue3","grey")) +
  geom_label_repel(aes(label = ID, size=1)) + 
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = 'dotted') +
  geom_hline(yintercept = 0, linetype = 'dotted') +
  ggtitle(tf)
```

```{r TF_MYOCD, fig.height=6, fig.width=9}
tf <- 'MYOCD'

df <- net %>%
  filter(source == tf) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(deg),rownames(df)))
df <- df[inter, ]
df[,c('logfc', 't_value', 'p_value')] <- deg[inter, ]
df <- df %>%
  mutate(color = if_else(mor > 0 & t_value > 0, '1', color)) %>%
  mutate(color = if_else(mor > 0 & t_value < 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value > 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value < 0, '1', color))

ggplot(df, aes(x = logfc, y = -log10(p_value), color = color, size=abs(mor))) +
  geom_point() +
  scale_colour_manual(values = c("red","royalblue3","grey")) +
  geom_label_repel(aes(label = ID, size=1)) + 
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = 'dotted') +
  geom_hline(yintercept = 0, linetype = 'dotted') +
  ggtitle(tf)
```

```{r TF_HIF1A, fig.height=6, fig.width=9}
tf <- 'HIF1A'

df <- net %>%
  filter(source == tf) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(deg),rownames(df)))
df <- df[inter, ]
df[,c('logfc', 't_value', 'p_value')] <- deg[inter, ]
df <- df %>%
  mutate(color = if_else(mor > 0 & t_value > 0, '1', color)) %>%
  mutate(color = if_else(mor > 0 & t_value < 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value > 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value < 0, '1', color))

ggplot(df, aes(x = logfc, y = -log10(p_value), color = color, size=abs(mor))) +
  geom_point() +
  scale_colour_manual(values = c("red","royalblue3","grey")) +
  geom_label_repel(aes(label = ID, size=1)) + 
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = 'dotted') +
  geom_hline(yintercept = 0, linetype = 'dotted') +
  ggtitle(tf)
```

```{r}
tf <- 'MEF2A'

df <- net %>%
  filter(source == tf) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')

inter <- sort(intersect(rownames(deg),rownames(df)))
df <- df[inter, ]
df[,c('logfc', 't_value', 'p_value')] <- deg[inter, ]
df <- df %>%
  mutate(color = if_else(mor > 0 & t_value > 0, '1', color)) %>%
  mutate(color = if_else(mor > 0 & t_value < 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value > 0, '2', color)) %>%
  mutate(color = if_else(mor < 0 & t_value < 0, '1', color))

ggplot(df, aes(x = logfc, y = -log10(p_value), color = color, size=abs(mor))) +
  geom_point() +
  scale_colour_manual(values = c("red","royalblue3","grey")) +
  geom_label_repel(aes(label = ID, size=1)) + 
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = 'dotted') +
  geom_hline(yintercept = 0, linetype = 'dotted') +
  ggtitle(tf)
```

https://www.nature.com/articles/s41419-023-05665-8

## Pathway inference analysis

```{r save CM final, eval = FALSE}
saveRDS(CM, 
        here::here(output_dir_data, "CM.rds"))
```




