---
title: "Cardiomyocytes subclustering"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "Cardiomyocytes_subclustering"

# Load libraries
library(here)
library(dplyr)
library(ggplot2)
library(tibble)
library(readr)
library(readxl)
library(knitr)
library(patchwork)
library(scales)

library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)


library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(Matrix.utils)
library(scater)
library(statmod)
library(clustree)
library(harmony)
library(gprofiler2)
library(AnnotationHub)
library(ReactomePA)
library(speckle)
library(limma)
library(reactable)
library(decoupleR)
library(OmnipathR)
library(dorothea)
library(edgeR)
library(qs2)


#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```


Load dataset and subset cardiomyocytes

```{r Load data, echo = TRUE, results= "hide"}
DCM_RV <- qs_read(here::here("output", "QC_integration_annotation", "DCM_RV_RNA.qs2"), nthreads = 32)
DefaultAssay(DCM_RV) <- "RNA"
CM <- subset(DCM_RV, cell_type == "Cardiomyocytes")
```

## Integration and clustering

Re-normalize and re-scale the subsetted object using SCTransform

```{r SCTransform, message=FALSE, warning=FALSE}
CM <- SplitObject(CM, split.by = "Sample")
for (i in 1:length(CM)) {
CM[[i]] <- SCTransform(CM[[i]], vst.flavor = "v2", 
                       vars.to.regress = "percent.mt", 
                       return.only.var.genes = TRUE, 
                       variable.features.n = 2000)
}
```

PCA

```{r PCA}
features <- SelectIntegrationFeatures(object.list = CM, nfeatures = 2000)
CM <- merge(x = CM[[1]], y = CM[2:length(CM)], merge.data=TRUE)
VariableFeatures(CM) <- features
CM <- RunPCA(CM, npcs = 30)
```

```{r Elbow_CM, fig.width=4, fig.height=4}
ElbowPlot(CM, ndims = 30)
```

Reintegrate the samples using Harmony

```{r RunHarmony}
CM <- RunHarmony(CM, 
                 group.by.vars = "Sample", 
                 dims.use = 1:30, 
                 plot_convergence=TRUE, 
                 .options = harmony_options(
                   max.iter.cluster = 100,
                   )
                 )
```

```{r read CM}
CM <- qs_read(here::here(output_dir_data, "CM.qs2"), nthreads = 32)
```

Clustering after integration

```{r clustering integrated, eval=FALSE}
DefaultAssay(CM) <- "SCT"
CM <- RunUMAP(CM, dims = 1:30, reduction = "harmony")
CM <- FindNeighbors(CM, dims = 1:30, reduction = "harmony")
CM <- FindClusters(CM, resolution = seq(0.1, 0.8, by=0.1))
```

```{r Clustree_CM, fig.width=6, fig.height=9}
clustree::clustree(CM@meta.data[,grep("SCT_snn_res", colnames(CM@meta.data))], prefix = "SCT_snn_res.")
```

```{r UMAPs_CM, fig.width=12, fig.height=10}
DimPlot(CM, reduction = "umap", shuffle = T,
        group.by = c("SCT_snn_res.0.2", "Sample", "Batch_processing", "Batch_sequencing"), ncol = 2)
```

```{r FindAllMarkers}
DefaultAssay(CM) <- "RNA"
CM <- NormalizeData(CM)
Idents(CM) <- "SCT_snn_res.0.2"
Markers <- FindAllMarkers(CM, only.pos = TRUE, min.pct = 0.1, logfc.threshold = 0.25)
Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
```

```{r Heatmap_CM, fig.width=15, fig.height=10}
heatmap_cols <- colorRampPalette(RColorBrewer::brewer.pal(9,"RdBu"))(256)
heatmap_cols <- rev(heatmap_cols[1:256])
Heatmap <- DoHeatmap(subset(CM, cells = sample(colnames(CM), 20000)), label = FALSE, draw.line = F, features = Markers_top10$gene, assay = "SCT") +
  scale_fill_gradientn(colours = heatmap_cols) +
  theme(axis.text.y = element_text(size = 5)) +
  theme(plot.margin = unit(c(0.1, 0, 0, 0), 
                           "inches"))
Heatmap
```

```{r cell_type_cols}
cell_type_cols <- c("#d05050", "#33A02C", "#1F78B4", "#CAB2D6", "#FF7F00", "#B2DF8A", "#B15928", "#FFD000", "#FA95C0", "#FDBF6F", "#6A3D9A", "#7aCDF9")  
```

Check QC parameters in each cluster

```{r QC_VlnPlots_1, fig.width=8, fig.height=6}
VlnPlot_scCustom(CM, features = c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.rp"), num_columns = 2, pt.size = 0, colors_use = cell_type_cols) + theme(axis.title.x = element_blank())
```

```{r QC_VlnPlots_2, fig.width=8, fig.height=6}
VlnPlot_scCustom(CM, features = c("COL6A3", "VWF", "RGS5", "scDblFinder.score"), assay="RNA", num_columns = 2, pt.size = 0, colors_use = cell_type_cols) + theme(axis.title.x = element_blank())
```

## Cell type annotation

Annotate the clusters

```{r cell_state}
CM@meta.data[CM@meta.data$SCT_snn_res.0.2 %in% c("0"), "cell_state"] = "CM1"
CM@meta.data[CM@meta.data$SCT_snn_res.0.2 %in% c("1"), "cell_state"] = "CM2"
CM@meta.data[CM@meta.data$SCT_snn_res.0.2 %in% c("2"), "cell_state"] = "CM3"
CM@meta.data[CM@meta.data$SCT_snn_res.0.2 %in% c("3"), "cell_state"] = "CM4"
CM@meta.data[CM@meta.data$SCT_snn_res.0.2 %in% c("4"), "cell_state"] = "CM5"

CM$cell_state <- factor(CM$cell_state, levels = c("CM1", "CM2", "CM3","CM4","CM5"))
Idents(CM) <- CM$cell_state
```

```{r FindAllMarkers cell_state}
DefaultAssay(CM) <- "RNA"
Markers <- FindAllMarkers(CM, only.pos = TRUE, min.pct = 0.1, logfc.threshold = 0.25)
write.csv(Markers, here::here(output_dir_data, "CM_Markers_all_cell_state.csv"))
Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
write.csv(Markers_top10, here::here(output_dir_data, "CM_Markers_top10_cell_state.csv"))
Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
write.csv(Markers_top3, here::here(output_dir_data, "CM_Markers_top3_cell_state.csv"))
```

```{r Heatmap_cell_state, fig.width=8, fig.height=6}
Heatmap <- DoHeatmap(subset(CM, cells = sample(colnames(CM), 20000)), label = TRUE, draw.line = F, features = Markers_top10$gene, assay = "SCT") + 
  scale_fill_gradientn(colours = heatmap_cols) +
  theme(text = element_text(size = 12), 
        axis.text.y = element_text(size = 8)) +
  theme(plot.margin = unit(c(0.2, 0, 0, 0), "inches")) +
  guides(fill = guide_colorbar(barwidth = 1, barheight = 4)) + guides(color = "none") 
Heatmap
```

```{r UMAP_cell_type_2, fig.width=6, fig.height=6}
p <- DimPlot(
        CM,
        group.by  = "cell_state",
        reduction = "umap",
        cols      = cell_type_cols,       
        label     = FALSE,
        shuffle=T,
        pt.size = 0.1
        ) +
     NoLegend() +
     theme(axis.text  = element_text(size = 8, face = "bold"),
           axis.title = element_text(size = 12, face = "bold"),
           plot.title = element_blank())

LabelClusters(p,
              id       = "cell_state",
              fontface = "bold",
              size     = 8,
              repel    = TRUE)
```

```{r DotPlot_markers, fig.width=8, fig.height=4}
Idents(CM) <- factor(
  Idents(CM),
  levels = rev(levels(Idents(CM)))
)

p <- DotPlot_scCustom(
        CM,
        features       = Markers_top3$gene,
        x_lab_rotate   = TRUE
)

p + theme(
      axis.text.x  = element_text(size = 12,        
                                  angle = 60,          
                                  vjust = 1,
                                  hjust = 1),
      axis.text.y  = element_text(size = 15))  
Idents(CM) <- CM$cell_state
```


```{r VlnPlot_markers_CM, fig.width=12, fig.height=4}
VlnPlot_scCustom(CM, features = c("LINC02388", "ARHGAP24", "THSD4", "NRAP"), num_columns = 4, colors_use = cell_type_cols, pt.size = 0)
```

```{r FeatPlot_markers_CM, fig.width=13, fig.height=3}
FeaturePlot_scCustom(CM, features = c("LINC02388", "ARHGAP24", "THSD4", "NRAP"), num_columns = 4)
```

```{r Barplot_CM_Sample, fig.height=6, fig.width=15, message=FALSE}
props <- getTransformedProps(CM$cell_state, CM$Sample, transform="logit")

par(mar = c(7, 5, 1, 8), xpd = TRUE)
barplot(props$Proportions, legend = FALSE, ylab = "Proportions", col = cell_type_cols, 
        cex.names = 2, las = 2, font.lab = 2, font.axis = 2, cex.axis = 1.2, cex.lab = 2)
legend("topright", inset = c(-0.1, 0.1), legend = rownames(props$Proportions), fill = cell_type_cols, bty = "n", cex = 2)
```

```{r bitr}
df <- Markers[,7:6]
dfsample <- split(df$gene,df$cluster)
length(dfsample)

dfsample$CM1 = bitr(dfsample$CM1, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
dfsample$CM2 = bitr(dfsample$CM2, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
dfsample$CM3 = bitr(dfsample$CM3, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
dfsample$CM4 = bitr(dfsample$CM4, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
dfsample$CM5 = bitr(dfsample$CM5, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")

genelist <- list("CM1" = dfsample$CM1$ENTREZID, 
                 "CM2" = dfsample$CM2$ENTREZID,
                 "CM3" = dfsample$CM3$ENTREZID,
                 "CM4" = dfsample$CM4$ENTREZID,
                 "CM5" = dfsample$CM5$ENTREZID
)

gene_counts <- rowSums(CM@assays$RNA@counts > 0)

universe_genes <- names(gene_counts[gene_counts >= 3])

universe_entrez <- bitr(universe_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)$ENTREZID
```

```{r Pathways_GO, fig.height=8, fig.width=6}
GO <- compareCluster(geneCluster = genelist, fun = "enrichGO", ont = "ALL",  OrgDb = "org.Hs.eg.db", universe = universe_entrez)
GO <- gsfilter(GO, by = 'Count', min = 3)
dotplot(GO, label_format = 35, title = "GO - Over-representation analysis")
```

```{r Pathways_GO_simplified, fig.height=8, fig.width=6}
GO_simplified <- clusterProfiler::simplify(
  GO,
  cutoff = 0.5,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)
dotplot(GO_simplified, label_format = 35, title = "GO - Over-representation analysis")
```

```{r Pathways_KEGG, fig.height=8, fig.width=6}
KEGG <- compareCluster(geneCluster = genelist, fun = "enrichKEGG", organism = 'hsa', universe = universe_entrez)
KEGG <- gsfilter(KEGG, by = 'Count', min = 3)
dotplot(KEGG, label_format = 35, title = "KEGG Pathways")
```

```{r Pathways_REACTOME, fig.height=8, fig.width=6}
REACTOME <- compareCluster(geneCluster = genelist, fun = "enrichPathway", organism = 'human', universe = universe_entrez)
REACTOME <- gsfilter(REACTOME, by = 'Count', min = 3)
dotplot(REACTOME, label_format = 40, title = "REACTOME", showCategory = 4)
```

## Differential abundance analysis

```{r metadata}
metadata <- CM@meta.data %>%
  dplyr::select(Sample, Batch_processing, Batch_sequencing, Age_years, Sex, Group, mPAP_mmHg, sPAP_mmHg, PCW_mmHg, PVR_WU, TAPSE_mm, RVID_mm, LVEF_percent, LVEDD_mm) %>%
  unique() %>%
  dplyr::arrange(match(Sample, names(table(CM$Sample))))

rownames(metadata) <- metadata$Sample

all.equal(metadata$Sample, names(table(CM$Sample)))

reactable(metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

metadata_DCM <- metadata %>%
  dplyr::filter(Group == "DCM")
```

```{r Speckle_all}
differential_abundance <- data.frame()
props <- getTransformedProps(CM$cell_state, CM$Sample, transform="logit")

for (i in 7:(length(metadata_DCM))) {
  metadata_subset <- metadata_DCM[complete.cases(metadata_DCM[[i]]),]
  props_subset <- props
  props_subset$Counts <- props$Counts[, metadata_subset$Sample, drop = FALSE]
  props_subset$TransformedProps <- props$TransformedProps[, metadata_subset$Sample, drop = FALSE]
  props_subset$Proportions <- props$Proportions[, metadata_subset$Sample, drop = FALSE]
  metadata_subset <- metadata_subset %>% 
    arrange(Sample, colnames(props_subset$Proportions))
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset$Age_years)
  fit <- lmFit(props_subset$TransformedProps, design)
  fit <- eBayes(fit, robust=TRUE)
  differential_abundance_temp <- topTable(fit, n = Inf, coef = 2)
  differential_abundance_temp$metadata <- colnames(metadata_subset[i])
  differential_abundance_temp$cell_state <- rownames(differential_abundance_temp)
  differential_abundance <- rbind(differential_abundance, differential_abundance_temp)
}

rm(metadata_subset, props_subset, design, fit, differential_abundance_temp)

rownames(differential_abundance) <- NULL
differential_abundance <- dplyr::arrange(differential_abundance, adj.P.Val)

write.csv(differential_abundance, file = here::here(output_dir_data, "CM_Differential_Abundance.csv"), row.names = F)

reactable(differential_abundance, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Speckle_LVEDD_mm, fig.height=4, fig.width=20}
metadata_subset <- metadata[complete.cases(metadata[["LVEDD_mm"]]),]

props_subset <- props
props_subset$Counts <- props$Counts[, metadata_subset$Sample, drop = FALSE]
props_subset$TransformedProps <- props$TransformedProps[, metadata_subset$Sample, drop = FALSE]
props_subset$Proportions <- props$Proportions[, metadata_subset$Sample, drop = FALSE]

metadata_subset <- metadata_subset %>% 
  arrange(Sample, colnames(props_subset$Proportions))

design <- model.matrix(~ metadata_subset$LVEDD_mm + metadata_subset$Age_years)
fit.prop <- lmFit(props_subset$Proportions,design)

par(mfrow=c(1,5), mar = c(5, 5, 3, 1))
for(i in seq(1,5,1)){
  plot(metadata_subset$LVEDD_mm, props_subset$Proportions[i,], main = rownames(props_subset$Proportions)[i],
       pch=16, cex=2, ylab="Proportions", xlab="LVEDD_mm", cex.lab=2, cex.axis=1.5,
       cex.main=2.5)
  abline(a=fit.prop$coefficients[i,1], b=fit.prop$coefficients[i,2], col=4,
         lwd=1)
}
rm(metadata_subset, props_subset, design, fit.prop)
```

```{r Speckle_LVEF_percent, fig.height=4, fig.width=20}
metadata_subset <- metadata[complete.cases(metadata[["LVEF_percent"]]),]

props_subset <- props
props_subset$Counts <- props$Counts[, metadata_subset$Sample, drop = FALSE]
props_subset$TransformedProps <- props$TransformedProps[, metadata_subset$Sample, drop = FALSE]
props_subset$Proportions <- props$Proportions[, metadata_subset$Sample, drop = FALSE]

metadata_subset <- metadata_subset %>% 
  arrange(Sample, colnames(props_subset$Proportions))

design <- model.matrix(~ metadata_subset$LVEF_percent + metadata_subset$Age_years)
fit.prop <- lmFit(props_subset$Proportions,design)

par(mfrow=c(1,5), mar = c(5, 5, 3, 1))
for(i in seq(1,5,1)){
  plot(metadata_subset$LVEF_percent, props_subset$Proportions[i,], main = rownames(props_subset$Proportions)[i],
       pch=16, cex=2, ylab="Proportions", xlab="LVEF_percent", cex.lab=2, cex.axis=1.5,
       cex.main=2.5)
  abline(a=fit.prop$coefficients[i,1], b=fit.prop$coefficients[i,2], col=4,
         lwd=1)
}
rm(metadata_subset, props_subset, design, fit.prop)
```

```{r Speckle_mPAP_mmHg, fig.height=4, fig.width=20}
metadata_subset <- metadata[complete.cases(metadata[["mPAP_mmHg"]]),]

props_subset <- props
props_subset$Counts <- props$Counts[, metadata_subset$Sample, drop = FALSE]
props_subset$TransformedProps <- props$TransformedProps[, metadata_subset$Sample, drop = FALSE]
props_subset$Proportions <- props$Proportions[, metadata_subset$Sample, drop = FALSE]

metadata_subset <- metadata_subset %>% 
  arrange(Sample, colnames(props_subset$Proportions))

design <- model.matrix(~ metadata_subset$mPAP_mmHg + metadata_subset$Age_years)
fit.prop <- lmFit(props_subset$Proportions,design)

par(mfrow=c(1,5), mar = c(5, 5, 3, 1))
for(i in seq(1,5,1)){
  plot(metadata_subset$mPAP_mmHg, props_subset$Proportions[i,], main = rownames(props_subset$Proportions)[i],
       pch=16, cex=2, ylab="Proportions", xlab="mPAP_mmHg", cex.lab=2, cex.axis=1.5,
       cex.main=2.5)
  abline(a=fit.prop$coefficients[i,1], b=fit.prop$coefficients[i,2], col=4,
         lwd=1)
}
rm(metadata_subset, props_subset, design, fit.prop)
```

```{r CM_proportions}
CM_proportions <- as.data.frame(props$Proportions) %>%   
  pivot_wider(
    names_from  = clusters,  
    values_from = Freq       
  )

write.csv(CM_proportions, file = here::here(output_dir_data, "CM_Proportions.csv"), row.names = F)
```

```{r rm differential_abundance, eval = FALSE}
rm(differential_abundance, differential_abundance_HC, props, CM_proportions)
```


## Transcription factor inference analysis

```{r net}
net <- get_collectri(organism='human', split_complexes=FALSE)
```

```{r percent_stats}
expr_matrix <- GetAssayData(object = CM, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
```

```{r filter percent_stats}
net <- net %>% 
  dplyr::filter(source %in% filter(percent_stats, percentCells > 1)$gene)
```

```{r TF_cols}
TF_cols <- rev(colorRampPalette(RColorBrewer::brewer.pal(5, "BrBG"))(101))
```

```{r TF_tfsulm, eval=FALSE}
mat <- as.matrix(CM@assays$RNA@data)

acts <- decoupleR::run_ulm(mat = mat, 
                           net = net, 
                           .source = 'source', 
                           .target = 'target',
                           .mor='mor', 
                           minsize = 10)

CM[['tfsulm']] <- acts %>%
                    tidyr::pivot_wider(id_cols = 'source', 
                                       names_from = 'condition',
                                       values_from = 'score') %>%
                    tibble::column_to_rownames('source') %>%
                    Seurat::CreateAssayObject(.)

DefaultAssay(object = CM) <- "tfsulm"

CM <- Seurat::ScaleData(CM)
CM@assays$tfsulm@data <- CM@assays$tfsulm@scale.data
```

```{r TF_heatmap_cluster_10, fig.width=4, fig.height=2.5}
n_tfs <- 10

df <- t(as.matrix(CM@assays$tfsulm@data)) %>%
      as.data.frame() %>%
      dplyr::mutate(cluster = Seurat::Idents(CM)) %>%
      tidyr::pivot_longer(cols = -cluster, 
                          names_to = "source", 
                          values_to = "score") %>%
      dplyr::group_by(cluster, source) %>%
      dplyr::summarise(mean = mean(score))

tfs <- df %>%
       dplyr::group_by(source) %>%
       dplyr::summarise(std = stats::sd(mean)) %>%
       dplyr::arrange(-abs(std)) %>%
       head(n_tfs) %>%
       dplyr::pull(source)

top_acts_mat <- df %>%
                dplyr::filter(source %in% tfs) %>%
                tidyr::pivot_wider(id_cols = 'cluster', 
                                   names_from = 'source',
                                   values_from = 'mean') %>%
                tibble::column_to_rownames('cluster') %>%
                as.matrix()

my_breaks <- c(seq(-2, 0, length.out = ceiling(101 / 2) + 1),
               seq(0.05, 2, length.out = floor(101 / 2)))

# Plot
pheatmap::pheatmap(mat = top_acts_mat,
                   color = TF_cols,
                   border_color = "white",
                   breaks = my_breaks,
                   cellwidth = 15,
                   cellheight = 15,
                   treeheight_row = 20,
                   treeheight_col = 20) 
```

```{r TF_heatmap_cluster_20, fig.width=6, fig.height=2.5}
n_tfs <- 20

df <- t(as.matrix(CM@assays$tfsulm@data)) %>%
      as.data.frame() %>%
      dplyr::mutate(cluster = Seurat::Idents(CM)) %>%
      tidyr::pivot_longer(cols = -cluster, 
                          names_to = "source", 
                          values_to = "score") %>%
      dplyr::group_by(cluster, source) %>%
      dplyr::summarise(mean = mean(score))

tfs <- df %>%
       dplyr::group_by(source) %>%
       dplyr::summarise(std = stats::sd(mean)) %>%
       dplyr::arrange(-abs(std)) %>%
       head(n_tfs) %>%
       dplyr::pull(source)

top_acts_mat <- df %>%
                dplyr::filter(source %in% tfs) %>%
                tidyr::pivot_wider(id_cols = 'cluster', 
                                   names_from = 'source',
                                   values_from = 'mean') %>%
                tibble::column_to_rownames('cluster') %>%
                as.matrix()

my_breaks <- c(seq(-2, 0, length.out = ceiling(100 / 2) + 1),
               seq(0.05, 2, length.out = floor(100 / 2)))

# Plot
pheatmap::pheatmap(mat = top_acts_mat,
                   color = TF_cols,
                   border_color = "white",
                   breaks = my_breaks,
                   cellwidth = 15,
                   cellheight = 15,
                   treeheight_row = 20,
                   treeheight_col = 20) 
```

```{r TF_Feat_MEF2A, fig.width=12, fig.height=4}
Seurat::DefaultAssay(CM) <- "tfsulm"
p1 <- Seurat::DimPlot(CM, 
                      reduction = "umap", 
                      label = TRUE, 
                      pt.size = 0.1) + 
      Seurat::NoLegend() + 
      ggplot2::ggtitle('Cell states')

p2 <- Seurat::FeaturePlot(CM, features = c("MEF2A"), pt.size = 0.1, order=F) + 
      ggplot2::scale_colour_gradient2(low = TF_cols[1], high = TF_cols[101], mid = "white") +
      ggplot2::ggtitle('MEF2A activity')

DefaultAssay(object = CM) <- "RNA"
p3 <- FeaturePlot_scCustom(CM, 
                          features = c("MEF2A"), pt.size = 0.1) + 
      ggplot2::ggtitle('MEF2A expression')

Seurat::DefaultAssay(CM) <- "tfsulm"

p <- p1 | p2 | p3
p
```

```{r TF_Feat_SRF, fig.width=12, fig.height=4}
Seurat::DefaultAssay(CM) <- "tfsulm"
p1 <- Seurat::DimPlot(CM, 
                      reduction = "umap", 
                      label = TRUE, 
                      pt.size = 0.1) + 
      Seurat::NoLegend() + 
      ggplot2::ggtitle('Cell states')

p2 <- Seurat::FeaturePlot(CM, features = c("SRF"), pt.size = 0.1, order=F) + 
      ggplot2::scale_colour_gradient2(low = TF_cols[1], high = TF_cols[101], mid = "white") +
      ggplot2::ggtitle('SRF activity')

DefaultAssay(object = CM) <- "RNA"
p3 <- FeaturePlot_scCustom(CM, 
                          features = c("SRF"), pt.size = 0.1, order = F) + 
      ggplot2::ggtitle('SRF expression')

Seurat::DefaultAssay(CM) <- "tfsulm"

p <- p1 | p2 | p3
p
```

## Pathway activity inference

```{r net_pathway}
net_path <- get_progeny(organism = 'human', top = 100000)
```

```{r Pathway_cols}
Pathway_cols <- colorRampPalette(c("#28A2EB", "white", "#FFB233"))(101)
```

```{r Pathway_pathwaysmlm, eval=FALSE}
acts_pathway <- decoupleR::run_mlm(mat = mat, 
                           net = net_path, 
                           .source = 'source', 
                           .target = 'target',
                           .mor = 'weight', 
                           minsize = 10)

CM[['pathwaysmlm']] <- acts_pathway %>%
                         tidyr::pivot_wider(id_cols = 'source', 
                                            names_from = 'condition',
                                            values_from = 'score') %>%
                         tibble::column_to_rownames(var = 'source') %>%
                         Seurat::CreateAssayObject(.)

Seurat::DefaultAssay(object = CM) <- "pathwaysmlm"

CM <- Seurat::ScaleData(CM)
CM@assays$pathwaysmlm@data <- CM@assays$pathwaysmlm@scale.data
```

```{r Pathway_heatmap_cluster, fig.width=6, fig.height=2.5}
df_pathway <- t(as.matrix(CM@assays$pathwaysmlm@data)) %>%
      as.data.frame() %>%
      dplyr::mutate(cluster = Seurat::Idents(CM)) %>%
      tidyr::pivot_longer(cols = -cluster, 
                          names_to = "source", 
                          values_to = "score") %>%
      dplyr::group_by(cluster, source) %>%
      dplyr::summarise(mean = mean(score))

top_acts_mat <- df_pathway %>%
                tidyr::pivot_wider(id_cols = 'cluster', 
                                   names_from = 'source',
                                   values_from = 'mean') %>%
                tibble::column_to_rownames(var = 'cluster') %>%
                as.matrix()

my_breaks <- c(seq(-1.25, 0, length.out = ceiling(101 / 2) + 1),
               seq(0.05, 1.25, length.out = floor(101 / 2)))

pheatmap::pheatmap(mat = top_acts_mat,
                   color = Pathway_cols,
                   border_color = "white",
                   breaks = my_breaks,
                   cellwidth = 20,
                   cellheight = 20,
                   treeheight_row = 20,
                   treeheight_col = 20) 
```

```{r Pathway_Feat_Trail, fig.width=8, fig.height=4}
Seurat::DefaultAssay(object = CM) <- "pathwaysmlm"

p1 <- Seurat::DimPlot(CM, 
                      reduction = "umap", 
                      label = TRUE, 
                      pt.size = 0.11) + 
      Seurat::NoLegend() + 
      ggplot2::ggtitle('Cell states')

p2 <- Seurat::FeaturePlot(CM, features = c("Trail"), 
                      pt.size = 0.1, order=F) + 
      ggplot2::scale_colour_gradientn(colors = Pathway_cols
                                      ) +
      ggplot2::ggtitle('Trail activity')

p <- p1 | p2
p
```

```{r CM_cell_states.csv, eval=FALSE}
cell_states <- data.frame(
  cell = colnames(CM),
  cell_state = CM$cell_state
)

write.csv(
  cell_states,
  here::here(output_dir_data, "CM_cell_states.csv"),
  row.names = FALSE
)
```

```{r save CM final, eval=FALSE}
DefaultAssay(CM) <- "RNA"
qs_save(CM,
        file = here::here(output_dir_data, "CM.qs2"),
        nthreads = 32
        )
```

