---
title: "Cell cell communication analysis using MultiNicheNet"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "Cell_cell_communication"

# Load libraries
library(here)
library(readr)
library(readxl)
library(xlsx)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(ArchR)
library(clustree)
library(harmony)
library(gprofiler2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationHub)
library(ReactomePA)
library(statmod)
library(edgeR)
library(speckle)
library(EnhancedVolcano)
library(decoupleR)
library(OmnipathR)
library(dorothea)
library(enrichplot)
library(png)
library(reactable)
library(UpSetR)
library(ComplexHeatmap)
library(rtracklayer)
library(cowplot)
library(nichenetr)
library(multinichenetr)
library(muscat)


#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```


Load seurat object

```{r read DCM_RV_integrated}
DCM_RV_integrated <- readRDS(here::here("output", "QC_integration_annotation", "DCM_RV_integrated.rds"))
```

```{r}
DCM_RV_integrated@meta.data <- DCM_RV_integrated@meta.data %>%
  mutate(Group_CCC = case_when(
    Group == "HC" ~ "HC",
    Group == "DCM" & !is.na(mPAP_mmHg) & mPAP_mmHg > 35 ~ "mPAP35_hi",
    Group == "DCM" & is.na(mPAP_mmHg) ~ "no_mPAP_data",
    Group == "DCM" & !is.na(mPAP_mmHg) & mPAP_mmHg <= 35 ~ "mPAP35_lo",
    TRUE ~ NA_character_
  ))

DCM_RV_integrated@meta.data$cell_type_2 <- make.names(DCM_RV_integrated@meta.data$cell_type)

DCM_RV_integrated@meta.data %>%
  dplyr::select(Group_CCC, Sample) %>%
  distinct() %>%                
  dplyr::count(Group_CCC)  
```

```{r}
organism = "human"

if(organism == "human"){
  
  lr_network_all = 
    readRDS(url(
      "https://zenodo.org/record/10229222/files/lr_network_human_allInfo_30112033.rds"
      )) %>% 
    mutate(
      ligand = convert_alias_to_symbols(ligand, organism = organism), 
      receptor = convert_alias_to_symbols(receptor, organism = organism))
  
  lr_network_all = lr_network_all  %>% 
    mutate(ligand = make.names(ligand), receptor = make.names(receptor)) 
  
  lr_network = lr_network_all %>% 
    distinct(ligand, receptor)
  
  ligand_target_matrix = readRDS(url(
    "https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds"
    ))
  
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  
  lr_network = lr_network %>% filter(ligand %in% colnames(ligand_target_matrix))
  ligand_target_matrix = ligand_target_matrix[, lr_network$ligand %>% unique()]
  
} else if(organism == "mouse"){
  
  lr_network_all = readRDS(url(
    "https://zenodo.org/record/10229222/files/lr_network_mouse_allInfo_30112033.rds"
    )) %>% 
    mutate(
      ligand = convert_alias_to_symbols(ligand, organism = organism), 
      receptor = convert_alias_to_symbols(receptor, organism = organism))
  
  lr_network_all = lr_network_all  %>% 
    mutate(ligand = make.names(ligand), receptor = make.names(receptor)) 
  lr_network = lr_network_all %>% 
    distinct(ligand, receptor)
  
  ligand_target_matrix = readRDS(url(
    "https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final_mouse.rds"
    ))
  
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  
  lr_network = lr_network %>% filter(ligand %in% colnames(ligand_target_matrix))
  ligand_target_matrix = ligand_target_matrix[, lr_network$ligand %>% unique()]
  
}
```

```{r}
sce <- Seurat::as.SingleCellExperiment(DCM_RV_integrated, assay = "RNA")

sce = alias_to_symbol_SCE(sce, "human") %>% makenames_SCE()

sce = scuttle::logNormCounts(sce)
```

```{r}
group_id = "Group_CCC"
sample_id = "Sample"
celltype_id = "cell_type_2"
covariates = "Age_years"
batches = NA
```

```{r}
contrasts_oi = c("'mPAP35_hi-mPAP35_lo','mPAP35_lo-mPAP35_hi'")
contrast_tbl = tibble(contrast = 
                        c("mPAP35_hi-mPAP35_lo","mPAP35_lo-mPAP35_hi"), 
                      group = c("mPAP35_hi","mPAP35_lo"))
```

```{r}
senders_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
receivers_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% 
            c(senders_oi, receivers_oi)
          ]
```

```{r}
conditions_keep = c("mPAP35_lo", "mPAP35_hi")
sce = sce[, SummarizedExperiment::colData(sce)[,group_id] %in% 
            conditions_keep
          ]
```

```{r}
min_cells = 50
```

```{r}
abundance_info = get_abundance_info(
  sce = sce, 
  sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, 
  min_cells = min_cells, 
  senders_oi = senders_oi, receivers_oi = receivers_oi, 
  batches = batches
  )
```

```{r, fig.width=8, fig.height=10}
abundance_info$abund_plot_sample
```

```{r}
abundance_df_summarized = abundance_info$abundance_data %>% 
  dplyr::mutate(keep = as.logical(keep)) %>% 
  dplyr::group_by(group_id, celltype_id) %>% 
  dplyr::summarise(samples_present = sum((keep)))

celltypes_absent_one_condition = abundance_df_summarized %>% 
  dplyr::filter(samples_present == 0) %>% pull(celltype_id) %>% unique() 
# find truly condition-specific cell types by searching for cell types 
# truely absent in at least one condition

celltypes_present_one_condition = abundance_df_summarized %>% 
  dplyr::filter(samples_present >= 2) %>% pull(celltype_id) %>% unique() 
# require presence in at least 2 samples of one group so 
# it is really present in at least one condition

condition_specific_celltypes = intersect(
  celltypes_absent_one_condition, 
  celltypes_present_one_condition)

total_nr_conditions = SummarizedExperiment::colData(sce)[,group_id] %>% 
  unique() %>% length() 

absent_celltypes = abundance_df_summarized %>% 
  dplyr::filter(samples_present < 2) %>% 
  dplyr::group_by(celltype_id) %>% 
  count() %>% 
  dplyr::filter(freq == total_nr_conditions) %>% 
  pull(celltype_id)
  
print("condition-specific celltypes:")
## [1] "condition-specific celltypes:"
print(condition_specific_celltypes)
## character(0)
  
print("absent celltypes:")
## [1] "absent celltypes:"
print(absent_celltypes)
## character(0)
```

```{r}
analyse_condition_specific_celltypes = FALSE

if(analyse_condition_specific_celltypes == TRUE){
  senders_oi = senders_oi %>% setdiff(absent_celltypes)
  receivers_oi = receivers_oi %>% setdiff(absent_celltypes)
} else {
  senders_oi = senders_oi %>% 
    setdiff(union(absent_celltypes, condition_specific_celltypes))
  receivers_oi = receivers_oi %>% 
    setdiff(union(absent_celltypes, condition_specific_celltypes))
}

sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% 
            c(senders_oi, receivers_oi)
          ]
```

```{r}
min_sample_prop = 0.50
fraction_cutoff = 0.05
```

```{r}
frq_list = get_frac_exprs(
  sce = sce, 
  sample_id = sample_id, celltype_id =  celltype_id, group_id = group_id, 
  batches = batches, 
  min_cells = min_cells, 
  fraction_cutoff = fraction_cutoff, min_sample_prop = min_sample_prop)

```

```{r}
genes_oi = frq_list$expressed_df %>% 
  filter(expressed == TRUE) %>% pull(gene) %>% unique() 
sce = sce[genes_oi, ]
```


```{r, eval=FALSE}
abundance_expression_info = process_abundance_expression_info(
  sce = sce, 
  sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, 
  min_cells = min_cells, 
  senders_oi = senders_oi, receivers_oi = receivers_oi, 
  lr_network = lr_network, 
  batches = batches, 
  frq_list = frq_list, 
  abundance_info = abundance_info)
```

```{r, eval=FALSE}
abundance_expression_info$celltype_info$pb_df %>% head()
```

```{r, eval=FALSE}
abundance_expression_info$celltype_info$pb_df_group %>% head()
```

```{r, eval=FALSE}
abundance_expression_info$sender_receiver_info$pb_df %>% head()

abundance_expression_info$sender_receiver_info$pb_df_group %>% head()
```


```{r, eval=FALSE}
DE_info = get_DE_info(
  sce = sce, 
  sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, 
  batches = batches, covariates = covariates, 
  contrasts_oi = contrasts_oi, 
  min_cells = min_cells, 
  expressed_df = frq_list$expressed_df)
```

```{r, eval=FALSE}
DE_info$celltype_de$de_output_tidy %>% head()
```

```{r, fig.width=15, fig.height=6, eval=FALSE}
DE_info$hist_pvals
```

```{r, eval=FALSE}
empirical_pval = FALSE
if(empirical_pval == TRUE){
  DE_info_emp = get_empirical_pvals(DE_info$celltype_de$de_output_tidy)
  celltype_de = DE_info_emp$de_output_tidy_emp %>% select(-p_val, -p_adj) %>% 
    rename(p_val = p_emp, p_adj = p_adj_emp)
} else {
  celltype_de = DE_info$celltype_de$de_output_tidy
} 
```

```{r, eval=FALSE}
sender_receiver_de = combine_sender_receiver_de(
  sender_de = celltype_de,
  receiver_de = celltype_de,
  senders_oi = senders_oi,
  receivers_oi = receivers_oi,
  lr_network = lr_network
)
sender_receiver_de %>% head(20)
```

```{r}
logFC_threshold = 0.5
p_val_threshold = 0.05
p_val_adj = FALSE 
```

```{r, eval=FALSE}
geneset_assessment = 	
contrast_tbl$contrast %>% 
  lapply(
    process_geneset_data, 
    celltype_de, logFC_threshold, p_val_adj, p_val_threshold
  ) %>% 
  bind_rows() 
geneset_assessment
```

```{r, eval=FALSE}
geneset_assessment_adjustedPval = contrast_tbl$contrast %>% 
  lapply(
    process_geneset_data, 
    celltype_de, logFC_threshold, p_val_adj = TRUE, p_val_threshold
    ) %>% 
  bind_rows() 
geneset_assessment_adjustedPval
```


```{r}
top_n_target = 250
```

```{r, eval=FALSE}
verbose = TRUE
cores_system = 32
n.cores = min(cores_system, celltype_de$cluster_id %>% unique() %>% length()) 
```

```{r, eval=FALSE}
ligand_activities_targets_DEgenes = suppressMessages(suppressWarnings(
  get_ligand_activities_targets_DEgenes(
    receiver_de = celltype_de,
    receivers_oi = intersect(receivers_oi, celltype_de$cluster_id %>% unique()),
    ligand_target_matrix = ligand_target_matrix,
    logFC_threshold = logFC_threshold,
    p_val_threshold = p_val_threshold,
    p_val_adj = p_val_adj,
    top_n_target = top_n_target,
    verbose = verbose, 
    n.cores = n.cores
  )
))
```

```{r, eval=FALSE}
ligand_activities_targets_DEgenes$ligand_activities %>% head(20)
```

```{r, eval=FALSE}
ligand_activity_down = FALSE
```

```{r, eval=FALSE}
sender_receiver_tbl = sender_receiver_de %>% distinct(sender, receiver)

metadata_combined = SummarizedExperiment::colData(sce) %>% tibble::as_tibble()

if(!is.na(batches)){
  grouping_tbl = metadata_combined[,c(sample_id, group_id, batches)] %>% 
    tibble::as_tibble() %>% distinct()
  colnames(grouping_tbl) = c("sample","group",batches)
} else {
  grouping_tbl = metadata_combined[,c(sample_id, group_id)] %>% 
    tibble::as_tibble() %>% distinct()
  colnames(grouping_tbl) = c("sample","group")
}

prioritization_tables = suppressMessages(generate_prioritization_tables(
    sender_receiver_info = abundance_expression_info$sender_receiver_info,
    sender_receiver_de = sender_receiver_de,
    ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes,
    contrast_tbl = contrast_tbl,
    sender_receiver_tbl = sender_receiver_tbl,
    grouping_tbl = grouping_tbl,
    scenario = "regular", # all prioritization criteria will be weighted equally
    fraction_cutoff = fraction_cutoff, 
    abundance_data_receiver = abundance_expression_info$abundance_data_receiver,
    abundance_data_sender = abundance_expression_info$abundance_data_sender,
    ligand_activity_down = ligand_activity_down
  ))
```

```{r, eval=FALSE}
prioritization_tables$group_prioritization_tbl %>% head(20)
```

```{r, eval=FALSE}
lr_target_prior_cor = lr_target_prior_cor_inference(
  receivers_oi = prioritization_tables$group_prioritization_tbl$receiver %>% unique(), 
  abundance_expression_info = abundance_expression_info, 
  celltype_de = celltype_de, 
  grouping_tbl = grouping_tbl, 
  prioritization_tables = prioritization_tables, 
  ligand_target_matrix = ligand_target_matrix, 
  logFC_threshold = logFC_threshold, 
  p_val_threshold = p_val_threshold, 
  p_val_adj = p_val_adj
  )
```

```{r, eval=FALSE}
path = output_dir_data

multinichenet_output = list(
    celltype_info = abundance_expression_info$celltype_info,
    celltype_de = celltype_de,
    sender_receiver_info = abundance_expression_info$sender_receiver_info,
    sender_receiver_de =  sender_receiver_de,
    ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes,
    prioritization_tables = prioritization_tables,
    grouping_tbl = grouping_tbl,
    lr_target_prior_cor = lr_target_prior_cor
  ) 
multinichenet_output = make_lite_output(multinichenet_output)

save = TRUE
if(save == TRUE){
  saveRDS(multinichenet_output, paste0(path, "/multinichenet_output.rds"))

}
```

```{r}
multinichenet_output <- readRDS(here::here(output_dir_data, "multinichenet_output.rds"))
```

```{r}
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 50, 
  rank_per_group = FALSE
  )
```

```{r, fig.height=6, fig.width=6}
prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{r}
group_oi = "mPAP35_hi"
```

```{r}
prioritized_tbl_oi_hi_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=21, fig.height=12}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_hi_50 %>% inner_join(lr_network_all)
  )
plot_oi
```

```{r}
group_oi = "mPAP35_lo"
```

```{r}
prioritized_tbl_oi_lo_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=21, fig.height=5}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_lo_50 %>% inner_join(lr_network_all)
  )
plot_oi
```

## Cardiomyocytes as receivers

```{r, fig.width=20, fig.height=15}
group_oi = "mPAP35_hi"
prioritized_tbl_oi_hi_50 = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  25, 
  groups_oi = group_oi, 
  receivers_oi  = "Cardiomyocytes"
  ) 
```

```{r, fig.width=21, fig.height=12}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_hi_50 %>% inner_join(lr_network_all)
  )
plot_oi
```

```{r, fig.height=6, fig.width=6}
prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_hi_50$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_hi_50)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{r, fig.width=12, fig.height=8}
VlnPlot(DCM_RV_integrated, c("TGFBR1", "TGFBR2", "TGFBR3", "TGFB1", "TGFB2", "TGFB3"), group.by="cell_type", pt.size=0.001, ncol = 3)
```
```{r, fig.width=12, fig.height=8}
DotPlot(DCM_RV_integrated, c("TGFBR1", "TGFBR2", "TGFBR3", "TGFB1", "TGFB2", "TGFB3"), group.by="cell_type", assay="SCT")
```

## Fibroblasts as receivers

```{r}
group_oi = "mPAP35_hi"
prioritized_tbl_oi_hi_50 = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  25, 
  groups_oi = group_oi, 
  receivers_oi  = "Fibroblasts"
  ) 
```

```{r, fig.width=21, fig.height=12}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_hi_50 %>% inner_join(lr_network_all)
  )
plot_oi
```

```{r, fig.height=6, fig.width=6}
prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_hi_50$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_hi_50)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

## Endothelial cells as receivers

```{r, fig.width=20, fig.height=15}
group_oi = "mPAP35_hi"
prioritized_tbl_oi_hi_50 = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  25, 
  groups_oi = group_oi, 
  receivers_oi  = "Endothelial.cells"
  ) 
```

```{r, fig.width=21, fig.height=12}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_hi_50 %>% inner_join(lr_network_all)
  )
plot_oi
```

```{r, fig.height=6, fig.width=6}
prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_hi_50$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_hi_50)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{r}
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 50, 
  rank_per_group = FALSE
  )

lr_target_prior = prioritized_tbl_oi_all %>% inner_join(
        multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>%
          distinct(ligand, target, direction_regulation, contrast) %>% inner_join(contrast_tbl) %>% ungroup() 
        ) 
lr_target_df = lr_target_prior %>% distinct(group, sender, receiver, ligand, receptor, id, target, direction_regulation) 
```

```{r}
lr_target_df %>% filter(target %in% union(lr_network$ligand, lr_network$receptor))
```

```{r}
network = infer_intercellular_regulatory_network(lr_target_df, prioritized_tbl_oi_all)
network$links %>% head()
network$nodes %>% head()
```

```{r, fig.width=15, fig.height=8}
colors_sender["Lymphocytes"] = "pink" # the  original yellow background with white font is not very readable
network_graph = visualize_network(network, colors_sender)
network_graph$plot
```

```{r}
lr_target_prior_cor_filtered = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl$group %>% unique() %>% 
  lapply(function(group_oi){
    lr_target_prior_cor_filtered = multinichenet_output$lr_target_prior_cor %>%
      inner_join(
        multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>%
          distinct(ligand, target, direction_regulation, contrast)
        ) %>% 
      inner_join(contrast_tbl) %>% filter(group == group_oi)
    
    lr_target_prior_cor_filtered_up = lr_target_prior_cor_filtered %>% 
      filter(direction_regulation == "up") %>% 
      filter( (rank_of_target < top_n_target) & (pearson > 0.33))
    
    lr_target_prior_cor_filtered_down = lr_target_prior_cor_filtered %>% 
      filter(direction_regulation == "down") %>% 
      filter( (rank_of_target < top_n_target) & (pearson < -0.33))
    lr_target_prior_cor_filtered = bind_rows(
      lr_target_prior_cor_filtered_up, 
      lr_target_prior_cor_filtered_down
      )
}) %>% bind_rows()

lr_target_df = lr_target_prior_cor_filtered %>% 
  distinct(group, sender, receiver, ligand, receptor, id, target, direction_regulation) 
```

```{r}
network = infer_intercellular_regulatory_network(lr_target_df, prioritized_tbl_oi_all)
network$links %>% head()
network$nodes %>% head()
```

```{r, fig.width=10, fig.height=8}
colors_sender["Lymphocytes"] = "pink" # the  original yellow background with white font is not very readable
colors_sender["Macrophages"] = "blue"
network_graph = visualize_network(network, colors_sender)
network_graph$plot
```

```{r}
network$nodes %>% filter(gene == "TGFB1")
```

```{r}
network$prioritized_lr_interactions
```

```{r}
prioritized_tbl_oi_network = prioritized_tbl_oi_all %>% inner_join(
  network$prioritized_lr_interactions)
prioritized_tbl_oi_network
```

```{r, fig.width=15}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_network %>% inner_join(lr_network_all)
  )
plot_oi

```

To summarize: this interpretable bubble plot is an important and helpful plot because: 1) these LR interactions are all in the overall top50 of condition-specific interactions 2) they are a likely interaction inducing one or more other prioritized LR interaction and/or they are regulated by one or more other prioritized LR interactions. Because of this, interactions in this plot may be interesting candidates for follow-up experimental validation.

```{r, fig.width=15, fig.height=10}
ligands_oi = multinichenet_output$prioritization_tables$ligand_activities_target_de_tbl %>% 
  inner_join(contrast_tbl) %>% 
  group_by(group, receiver) %>% filter(direction_regulation == "up") %>% 
  distinct(ligand, receiver, group, activity) %>% 
  top_n(5, activity) %>% 
  pull(ligand) %>% unique()

plot_oi = make_ligand_activity_plots(
  multinichenet_output$prioritization_tables, 
  ligands_oi, 
  contrast_tbl,
  widths = NULL)
plot_oi
```

```{r}
group_oi = "mPAP35_hi"
receiver_oi = "Cardiomyocytes"
prioritized_tbl_oi_M_10 = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  20, 
  groups_oi = group_oi, 
  receivers_oi = receiver_oi)
```

```{r, fig.width=20, fig.height=10}
combined_plot = make_ligand_activity_target_plot(
  group_oi, 
  receiver_oi, 
  prioritized_tbl_oi_M_10,
  multinichenet_output$prioritization_tables, 
  multinichenet_output$ligand_activities_targets_DEgenes, contrast_tbl, 
  multinichenet_output$grouping_tbl, 
  multinichenet_output$celltype_info, 
  ligand_target_matrix, 
  plot_legend = FALSE)
combined_plot
## $combined_plot
```





