---
title: "Cell cell communication analysis using MultiNicheNet"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "Cell_cell_communication"

# Load libraries
library(here)
library(Seurat)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(SingleCellExperiment)
library(RColorBrewer)
library(nichenetr)
library(multinichenetr)
library(muscat)
library(writexl)


#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```

## Load Seurat object and define mPAP groups

```{r read_DCM_RV, eval=TRUE}
DCM_RV <- qs2::qs_read(here::here("output", "Add_cell_states_remove_contamination", "DCM_RV_cell_states_RNA.qs2"), nthreads = 32)
```

```{r define_groups, eval=TRUE}
DCM_RV@meta.data <- DCM_RV@meta.data %>%
  mutate(Group_CCC = case_when(
    !is.na(mPAP_mmHg) & mPAP_mmHg >= 35 ~ "mPAP35_hi",
    !is.na(mPAP_mmHg) & mPAP_mmHg < 35 ~ "mPAP35_lo",
    is.na(mPAP_mmHg) ~ "no_mPAP_data",
    TRUE ~ NA_character_
  ))

DCM_RV@meta.data$cell_type_2 <- make.names(DCM_RV@meta.data$cell_type)

DCM_RV@meta.data %>%
  dplyr::select(Group_CCC, Sample) %>%
  distinct() %>%
  dplyr::count(Group_CCC)
```

## MultiNicheNet networks

```{r multinichenet_networks, eval=TRUE}
organism = "human"

lr_network_all = 
  readRDS(url(
    "https://zenodo.org/record/10229222/files/lr_network_human_allInfo_30112033.rds"
    )) %>% 
  mutate(
    ligand = convert_alias_to_symbols(ligand, organism = organism), 
    receptor = convert_alias_to_symbols(receptor, organism = organism))

lr_network_all = lr_network_all  %>% 
  mutate(ligand = make.names(ligand), receptor = make.names(receptor)) 

lr_network = lr_network_all %>% 
  distinct(ligand, receptor)

ligand_target_matrix = readRDS(url(
  "https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds"
  ))

colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% 
  convert_alias_to_symbols(organism = organism) %>% make.names()
rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% 
  convert_alias_to_symbols(organism = organism) %>% make.names()

lr_network = lr_network %>% filter(ligand %in% colnames(ligand_target_matrix))
ligand_target_matrix = ligand_target_matrix[, lr_network$ligand %>% unique()]
```

## Prepare SingleCellExperiment

```{r prepare_sce, eval=TRUE}
sce <- Seurat::as.SingleCellExperiment(DCM_RV, assay = "RNA")

sce = alias_to_symbol_SCE(sce, "human") %>% makenames_SCE()

sce = scuttle::logNormCounts(sce)
```

```{r multinichenet_parameters, eval=TRUE}
group_id = "Group_CCC"
sample_id = "Sample"
celltype_id = "cell_type_2"
covariates = "Age_years"
batches = NA

contrasts_oi = c("'mPAP35_hi-mPAP35_lo','mPAP35_lo-mPAP35_hi'")
contrast_tbl = tibble(contrast = 
                        c("mPAP35_hi-mPAP35_lo","mPAP35_lo-mPAP35_hi"), 
                      group = c("mPAP35_hi","mPAP35_lo"))

conditions_keep = c("mPAP35_lo", "mPAP35_hi")

min_cells = 50
min_sample_prop = 0.50
fraction_cutoff = 0.05

logFC_threshold = 0.5
p_val_threshold = 0.05
p_val_adj = FALSE

top_n_target = 250
```

## MultiNicheNet analysis

Run this chunk interactively when the analysis needs to be recalculated. The resulting object is saved and loaded below for rendering.

```{r multinichenet_analysis, eval=TRUE}
senders_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
receivers_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()

sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% 
            c(senders_oi, receivers_oi)
          ]

sce = sce[, SummarizedExperiment::colData(sce)[,group_id] %in% 
            conditions_keep
          ]

abundance_info = get_abundance_info(
  sce = sce, 
  sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, 
  min_cells = min_cells, 
  senders_oi = senders_oi, receivers_oi = receivers_oi, 
  batches = batches
  )

abundance_df_summarized = abundance_info$abundance_data %>% 
  dplyr::mutate(keep = as.logical(keep)) %>% 
  dplyr::group_by(group_id, celltype_id) %>% 
  dplyr::summarise(samples_present = sum((keep)))

celltypes_absent_one_condition = abundance_df_summarized %>% 
  dplyr::filter(samples_present == 0) %>% pull(celltype_id) %>% unique()

celltypes_present_one_condition = abundance_df_summarized %>% 
  dplyr::filter(samples_present >= 2) %>% pull(celltype_id) %>% unique()

condition_specific_celltypes = intersect(
  celltypes_absent_one_condition, 
  celltypes_present_one_condition)

total_nr_conditions = SummarizedExperiment::colData(sce)[,group_id] %>% 
  unique() %>% length()

absent_celltypes = abundance_df_summarized %>% 
  dplyr::filter(samples_present < 2) %>% 
  dplyr::group_by(celltype_id) %>% 
  dplyr::count() %>% 
  dplyr::filter(n == total_nr_conditions) %>% 
  pull(celltype_id)

analyse_condition_specific_celltypes = FALSE

if(analyse_condition_specific_celltypes == TRUE){
  senders_oi = senders_oi %>% setdiff(absent_celltypes)
  receivers_oi = receivers_oi %>% setdiff(absent_celltypes)
} else {
  senders_oi = senders_oi %>% 
    setdiff(union(absent_celltypes, condition_specific_celltypes))
  receivers_oi = receivers_oi %>% 
    setdiff(union(absent_celltypes, condition_specific_celltypes))
}

sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% 
            c(senders_oi, receivers_oi)
          ]

frq_list = get_frac_exprs(
  sce = sce, 
  sample_id = sample_id, celltype_id =  celltype_id, group_id = group_id, 
  batches = batches, 
  min_cells = min_cells, 
  fraction_cutoff = fraction_cutoff, min_sample_prop = min_sample_prop)

genes_oi = frq_list$expressed_df %>% 
  filter(expressed == TRUE) %>% pull(gene) %>% unique()
sce = sce[genes_oi, ]

abundance_expression_info = process_abundance_expression_info(
  sce = sce, 
  sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, 
  min_cells = min_cells, 
  senders_oi = senders_oi, receivers_oi = receivers_oi, 
  lr_network = lr_network, 
  batches = batches, 
  frq_list = frq_list, 
  abundance_info = abundance_info)

DE_info = get_DE_info(
  sce = sce, 
  sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, 
  batches = batches, covariates = covariates, 
  contrasts_oi = contrasts_oi, 
  min_cells = min_cells, 
  expressed_df = frq_list$expressed_df)

empirical_pval = FALSE
if(empirical_pval == TRUE){
  DE_info_emp = get_empirical_pvals(DE_info$celltype_de$de_output_tidy)
  celltype_de = DE_info_emp$de_output_tidy_emp %>% select(-p_val, -p_adj) %>% 
    rename(p_val = p_emp, p_adj = p_adj_emp)
} else {
  celltype_de = DE_info$celltype_de$de_output_tidy
}

sender_receiver_de = combine_sender_receiver_de(
  sender_de = celltype_de,
  receiver_de = celltype_de,
  senders_oi = senders_oi,
  receivers_oi = receivers_oi,
  lr_network = lr_network
)

verbose = TRUE
cores_system = 32
n.cores = min(cores_system, celltype_de$cluster_id %>% unique() %>% length())

ligand_activities_targets_DEgenes = suppressMessages(suppressWarnings(
  get_ligand_activities_targets_DEgenes(
    receiver_de = celltype_de,
    receivers_oi = intersect(receivers_oi, celltype_de$cluster_id %>% unique()),
    ligand_target_matrix = ligand_target_matrix,
    logFC_threshold = logFC_threshold,
    p_val_threshold = p_val_threshold,
    p_val_adj = p_val_adj,
    top_n_target = top_n_target,
    verbose = verbose, 
    n.cores = n.cores
  )
))

ligand_activity_down = FALSE

sender_receiver_tbl = sender_receiver_de %>% distinct(sender, receiver)

metadata_combined = SummarizedExperiment::colData(sce) %>% tibble::as_tibble()

grouping_tbl = metadata_combined[,c(sample_id, group_id)] %>% 
  tibble::as_tibble() %>% distinct()
colnames(grouping_tbl) = c("sample","group")

prioritization_tables = suppressMessages(generate_prioritization_tables(
    sender_receiver_info = abundance_expression_info$sender_receiver_info,
    sender_receiver_de = sender_receiver_de,
    ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes,
    contrast_tbl = contrast_tbl,
    sender_receiver_tbl = sender_receiver_tbl,
    grouping_tbl = grouping_tbl,
    scenario = "regular",
    fraction_cutoff = fraction_cutoff, 
    abundance_data_receiver = abundance_expression_info$abundance_data_receiver,
    abundance_data_sender = abundance_expression_info$abundance_data_sender,
    ligand_activity_down = ligand_activity_down
  ))

lr_target_prior_cor = lr_target_prior_cor_inference(
  receivers_oi = prioritization_tables$group_prioritization_tbl$receiver %>% unique(), 
  abundance_expression_info = abundance_expression_info, 
  celltype_de = celltype_de, 
  grouping_tbl = grouping_tbl, 
  prioritization_tables = prioritization_tables, 
  ligand_target_matrix = ligand_target_matrix, 
  logFC_threshold = logFC_threshold, 
  p_val_threshold = p_val_threshold, 
  p_val_adj = p_val_adj
  )

multinichenet_output = list(
    celltype_info = abundance_expression_info$celltype_info,
    celltype_de = celltype_de,
    sender_receiver_info = abundance_expression_info$sender_receiver_info,
    sender_receiver_de = sender_receiver_de,
    ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes,
    prioritization_tables = prioritization_tables,
    grouping_tbl = grouping_tbl,
    lr_target_prior_cor = lr_target_prior_cor
  )

multinichenet_output = make_lite_output(multinichenet_output)

qs2::qs_save(
  multinichenet_output,
  here::here(output_dir_data, "multinichenet_output_clean.qs2"),
  nthreads = 32
)
```

```{r read_multinichenet_output, eval=TRUE}
multinichenet_output <- qs2::qs_read(
  here::here(output_dir_data, "multinichenet_output_clean.qs2"),
  nthreads = 32
)
```

## Top prioritized ligand-receptor interactions

```{r top_LR_interactions, eval=TRUE}
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 50, 
  rank_per_group = FALSE
  )

prioritized_tbl_oi_hi_50 = prioritized_tbl_oi_all %>% 
  filter(group == "mPAP35_hi")

prioritized_tbl_oi_lo_50 = prioritized_tbl_oi_all %>% 
  filter(group == "mPAP35_lo")

celltype_cols <- c(
  "Cardiomyocytes"      = "#D8576B",
  "Endothelial.cells"   = "#EE8424",
  "Fibroblasts"         = "#D2AE2B",
  "Macrophages"         = "#7AA641",
  "Neuronal.cells"      = "#5ABF9A",
  "Pericytes"           = "#3FA7B5",
  "Smooth.muscle.cells" = "#4C78A8",
  "T.lymphocytes"       = "#9C6ADE"
)
```

## Chord diagrams

```{r custom_circos_function, eval=TRUE}
make_circos_group_comparison_custom <- function(prioritized_tbl_oi, colors_sender, colors_receiver) {
  
  requireNamespace("dplyr")
  requireNamespace("ggplot2")
  requireNamespace("circlize")
  
  prioritized_tbl_oi = prioritized_tbl_oi %>% dplyr::ungroup()
  
  grid_col_tbl_ligand = tibble::tibble(
    sender = colors_sender %>% names(),
    color_ligand_type = colors_sender
  )
  
  grid_col_tbl_receptor = tibble::tibble(
    receiver = colors_receiver %>% names(),
    color_receptor_type = colors_receiver
  )
  
  groups_oi = prioritized_tbl_oi$group %>% unique()
  
  all_plots = groups_oi %>% lapply(function(group_oi) {
    
    title = group_oi
    
    circos_links_oi = prioritized_tbl_oi %>%
      dplyr::filter(group == group_oi)
    
    circos_links = circos_links_oi %>%
      dplyr::rename(weight = prioritization_score)
    
    df = circos_links
    
    ligand.uni = unique(df$ligand)
    
    for (i in 1:length(ligand.uni)) {
      
      df.i = df[df$ligand == ligand.uni[i], ]
      sender.uni = unique(df.i$sender)
      
      for (j in 1:length(sender.uni)) {
        
        df.i.j = df.i[df.i$sender == sender.uni[j], ]
        
        df.i.j$ligand = paste0(
          df.i.j$ligand,
          paste(rep(" ", j - 1), collapse = "")
        )
        
        df$ligand[df$id %in% df.i.j$id] = df.i.j$ligand
      }
    }
    
    receptor.uni = unique(df$receptor)
    
    for (i in 1:length(receptor.uni)) {
      
      df.i = df[df$receptor == receptor.uni[i], ]
      receiver.uni = unique(df.i$receiver)
      
      for (j in 1:length(receiver.uni)) {
        
        df.i.j = df.i[df.i$receiver == receiver.uni[j], ]
        
        df.i.j$receptor = paste0(
          df.i.j$receptor,
          paste(rep(" ", j - 1), collapse = "")
        )
        
        df$receptor[df$id %in% df.i.j$id] = df.i.j$receptor
      }
    }
    
    intersecting_ligands_receptors = generics::intersect(
      unique(df$ligand),
      unique(df$receptor)
    )
    
    while (length(intersecting_ligands_receptors) > 0) {
      
      df_unique = df %>%
        dplyr::filter(!receptor %in% intersecting_ligands_receptors)
      
      df_duplicated = df %>%
        dplyr::filter(receptor %in% intersecting_ligands_receptors)
      
      df_duplicated = df_duplicated %>%
        dplyr::mutate(receptor = paste(" ", receptor, sep = ""))
      
      df = dplyr::bind_rows(df_unique, df_duplicated)
      
      intersecting_ligands_receptors = generics::intersect(
        unique(df$ligand),
        unique(df$receptor)
      )
    }
    
    circos_links = df
    
    circos_links = circos_links %>%
      dplyr::inner_join(grid_col_tbl_ligand) %>%
      dplyr::inner_join(grid_col_tbl_receptor)
    
    links_circle = circos_links %>%
      dplyr::distinct(ligand, receptor, weight)
    
    ligand_color = circos_links %>%
      dplyr::distinct(ligand, color_ligand_type)
    
    grid_ligand_color = ligand_color$color_ligand_type %>%
      magrittr::set_names(ligand_color$ligand)
    
    receptor_color = circos_links %>%
      dplyr::distinct(receptor, color_receptor_type)
    
    grid_receptor_color = receptor_color$color_receptor_type %>%
      magrittr::set_names(receptor_color$receptor)
    
    grid_col = c(grid_ligand_color, grid_receptor_color)
    
    transparency = circos_links %>%
      dplyr::mutate(
        weight = (weight - min(weight)) / (max(weight) - min(weight))
      ) %>%
      dplyr::mutate(transparency = 1 - weight) %>%
      .$transparency
    
    ligand_order = prioritized_tbl_oi$sender %>%
      unique() %>%
      sort() %>%
      lapply(function(sender_oi) {
        
        ligands = circos_links %>%
          dplyr::filter(sender == sender_oi) %>%
          dplyr::arrange(ligand) %>%
          dplyr::distinct(ligand)
      }) %>%
      unlist()
    
    receptor_order = prioritized_tbl_oi$receiver %>%
      unique() %>%
      sort() %>%
      lapply(function(receiver_oi) {
        
        receptors = circos_links %>%
          dplyr::filter(receiver == receiver_oi) %>%
          dplyr::arrange(receptor) %>%
          dplyr::distinct(receptor)
      }) %>%
      unlist()
    
    order = c(ligand_order, receptor_order)
    
    width_same_cell_same_ligand_type = 0.275
    width_different_cell = 3
    width_ligand_receptor = 9
    width_same_cell_same_receptor_type = 0.275
    
    sender_gaps = prioritized_tbl_oi$sender %>%
      unique() %>%
      sort() %>%
      lapply(function(sender_oi) {
        
        sector = rep(
          width_same_cell_same_ligand_type,
          times = (
            circos_links %>%
              dplyr::filter(sender == sender_oi) %>%
              dplyr::distinct(ligand) %>%
              nrow()
          ) - 1
        )
        
        gap = width_different_cell
        
        return(c(sector, gap))
      }) %>%
      unlist()
    
    sender_gaps = sender_gaps[-length(sender_gaps)]
    
    receiver_gaps = prioritized_tbl_oi$receiver %>%
      unique() %>%
      sort() %>%
      lapply(function(receiver_oi) {
        
        sector = rep(
          width_same_cell_same_receptor_type,
          times = (
            circos_links %>%
              dplyr::filter(receiver == receiver_oi) %>%
              dplyr::distinct(receptor) %>%
              nrow()
          ) - 1
        )
        
        gap = width_different_cell
        
        return(c(sector, gap))
      }) %>%
      unlist()
    
    receiver_gaps = receiver_gaps[-length(receiver_gaps)]
    
    gaps = c(
      sender_gaps,
      width_ligand_receptor,
      receiver_gaps,
      width_ligand_receptor
    )
    
    if (
      length(gaps) !=
      length(
        union(circos_links$ligand, circos_links$receptor) %>%
          unique()
      )
    ) {
      warning(
        "Specified gaps have different length than combined total of ligands and receptors - This is probably due to duplicates in ligand-receptor names"
      )
    }
    
   links_circle$weight[links_circle$weight == 0] = 0.01

graphics::par(mar = c(1, 1, 4, 1))

circlize::circos.clear()

circlize::circos.par(
  gap.degree = gaps,
  canvas.xlim = c(-1.30, 1.30),
  canvas.ylim = c(-1.30, 1.42)
)

circlize::chordDiagram(
  links_circle,
  directional = 1,
  order = order,
  link.sort = TRUE,
  link.decreasing = TRUE,
  grid.col = grid_col,
  transparency = transparency,
  diffHeight = 0.0075,
  direction.type = c("diffHeight", "arrows"),
  link.visible = links_circle$weight > 0.01,
  annotationTrack = "grid",
  preAllocateTracks = list(track.height = 0.175),
  grid.border = "gray35",
  link.arr.length = 0.05,
  link.arr.type = "big.arrow",
  link.lwd = 1.25,
  link.lty = 1,
  link.border = "gray35",
  reduce = 0,
  scale = TRUE
)

circlize::circos.track(
  track.index = 1,
  panel.fun = function(x, y) {
    circlize::circos.text(
      circlize::CELL_META$xcenter,
      circlize::CELL_META$ylim[1],
      circlize::CELL_META$sector.index,
      facing = "clockwise",
      niceFacing = TRUE,
      adj = c(0, 0.5),
      cex = 0.8
    )
  },
  bg.border = NA
)

graphics::text(
  x = 0,
  y = 1.32,
  labels = title,
  cex = 1.3,
  font = 2
)

p_circos = grDevices::recordPlot()

return(p_circos)
})
  
  names(all_plots) = groups_oi
  
  graphics::plot(
    NULL,
    xaxt = "n",
    yaxt = "n",
    bty = "n",
    ylab = "",
    xlab = "",
    xlim = 0:1,
    ylim = 0:1
  )
  
  legend = ComplexHeatmap::Legend(
    at = prioritized_tbl_oi$receiver %>% unique() %>% sort(),
    type = "grid",
    legend_gp = grid::gpar(
      fill = colors_receiver[
        prioritized_tbl_oi$receiver %>% unique() %>% sort()
      ]
    ),
    labels_gp = grid::gpar(fontsize = 11),
    title_gp = grid::gpar(fontsize = 13, fontface = "bold"),
    grid_height = grid::unit(5, "mm"),
    grid_width = grid::unit(5, "mm"),
    gap = grid::unit(2, "mm"),
    title_position = "topleft",
    title = "Receiver"
  )
  
  ComplexHeatmap::draw(
    legend,
    just = c("left", "bottom")
  )
  
  legend = ComplexHeatmap::Legend(
    at = prioritized_tbl_oi$sender %>% unique() %>% sort(),
    type = "grid",
    legend_gp = grid::gpar(
      fill = colors_sender[
        prioritized_tbl_oi$sender %>% unique() %>% sort()
      ]
    ),
    labels_gp = grid::gpar(fontsize = 11),
    title_gp = grid::gpar(fontsize = 13, fontface = "bold"),
    grid_height = grid::unit(5, "mm"),
    grid_width = grid::unit(5, "mm"),
    gap = grid::unit(2, "mm"),
    title_position = "topleft",
    title = "Sender"
  )
  
  ComplexHeatmap::draw(
    legend,
    just = c("left", "top")
  )
  
  p_legend = grDevices::recordPlot()
  
  all_plots$legend = p_legend
  
  return(all_plots)
}
```

```{r chord_diagrams, fig.height=5, fig.width=4.5, eval=TRUE}
prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_all)

prioritized_tbl_oi$prioritization_score[
  is.na(prioritized_tbl_oi$prioritization_score)
] = 0

prioritized_tbl_oi$group <- dplyr::recode(
  prioritized_tbl_oi$group,
  "mPAP35_hi" = "mPAP ≥ 35",
  "mPAP35_lo" = "mPAP < 35"
)

senders_receivers <- union(
  unique(prioritized_tbl_oi$sender),
  unique(prioritized_tbl_oi$receiver)
) %>%
  sort()

colors_sender <- celltype_cols[senders_receivers]
colors_receiver <- celltype_cols[senders_receivers]

circos_list <- make_circos_group_comparison_custom(
  prioritized_tbl_oi,
  colors_sender,
  colors_receiver
)

circos_list[["mPAP ≥ 35"]]

circos_list[["mPAP < 35"]]

circos_list$legend
```

## Ligand-receptor bubble heatmaps

```{r bubble_plot_settings, eval=TRUE}
bubble_widths_plot <- c(6, 3.4, 3.4, 2.2)

bubble_row_height <- 0.28
bubble_base_height <- 4.2

n_interactions_hi <- prioritized_tbl_oi_hi_50 %>%
  distinct(sender, receiver, ligand, receptor) %>%
  nrow()

n_interactions_lo <- prioritized_tbl_oi_lo_50 %>%
  distinct(sender, receiver, ligand, receptor) %>%
  nrow()

bubble_height_hi <- bubble_base_height + n_interactions_hi * bubble_row_height
bubble_height_lo <- bubble_base_height + n_interactions_lo * bubble_row_height

n_interactions_hi
n_interactions_lo
bubble_height_hi
bubble_height_lo
```

### mPAP ≥ 35

```{r mPAP_hi_bubble, fig.width=20.5, fig.height=17.08, eval=TRUE}
plot_oi_hi <- make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables,
  prioritized_tbl_oi_hi_50 %>% inner_join(lr_network_all),
  widths = bubble_widths_plot
)

p1_hi <- plot_oi_hi[[1]]
p2_hi <- plot_oi_hi[[2]]
p3_hi <- plot_oi_hi[[3]]
p4_hi <- plot_oi_hi[[4]]

p1_hi$data$group <- factor(
  dplyr::recode(
    as.character(p1_hi$data$group),
    "mPAP35_hi" = "mPAP ≥ 35",
    "mPAP35_lo" = "mPAP < 35"
  ),
  levels = c("mPAP ≥ 35", "mPAP < 35")
)

p2_hi$data$group <- factor(
  dplyr::recode(
    as.character(p2_hi$data$group),
    "mPAP35_hi" = "mPAP ≥ 35",
    "mPAP35_lo" = "mPAP < 35"
  ),
  levels = c("mPAP ≥ 35", "mPAP < 35")
)

p3_hi$data$group <- factor(
  dplyr::recode(
    as.character(p3_hi$data$group),
    "mPAP35_hi" = "mPAP ≥ 35",
    "mPAP35_lo" = "mPAP < 35"
  ),
  levels = c("mPAP ≥ 35", "mPAP < 35")
)

p1_hi <- p1_hi +
  theme(
    axis.text.y = element_text(face = "bold.italic", size = 15),
    axis.text.x.top = element_text(
      size = 12,
      angle = 45,
      hjust = 0,
      vjust = 0,
      margin = margin(b = 2)
    ),
    strip.text.x.top = element_text(size = 16, face = "bold"),
    strip.text.y.left = element_text(size = 16, face = "bold"),
    panel.spacing.y = grid::unit(0.18, "lines")
  ) +
  guides(
    color = guide_colorbar(
      direction = "vertical",
      title.position = "top",
      barheight = grid::unit(25, "mm"),
      barwidth = grid::unit(4, "mm")
    ),
    size = guide_legend(
      direction = "vertical",
      title.position = "top",
      ncol = 1
    )
  )

p2_hi <- p2_hi +
  theme(
    axis.text.y = element_blank(),
    axis.text.x.top = element_text(
      size = 12,
      angle = 45,
      hjust = 0,
      vjust = 0,
      margin = margin(b = 2)
    ),
    strip.text.x = element_text(size = 16, face = "bold"),
    panel.spacing.y = grid::unit(0.18, "lines")
  ) +
  guides(
    fill = guide_colorbar(
      direction = "vertical",
      title.position = "top",
      barheight = grid::unit(25, "mm"),
      barwidth = grid::unit(4, "mm")
    )
  )

p3_hi <- p3_hi +
  theme(
    axis.text.x.top = element_text(
      size = 12,
      angle = 45,
      hjust = 0,
      vjust = 0,
      margin = margin(b = 2)
    ),
    strip.text.x = element_text(size = 16, face = "bold"),
    panel.spacing.y = grid::unit(0.18, "lines")
  ) +
  guides(
    color = guide_colorbar(
      direction = "vertical",
      title.position = "top",
      barheight = grid::unit(25, "mm"),
      barwidth = grid::unit(4, "mm")
    ),
    size = guide_legend(
      direction = "vertical",
      title.position = "top",
      ncol = 1
    )
  )

p4_hi <- p4_hi +
  theme(
    axis.text.x.top = element_text(
      size = 12,
      angle = 45,
      hjust = 0,
      vjust = 0,
      margin = margin(b = 2)
    ),
    strip.text.x = element_text(size = 16, face = "bold"),
    panel.spacing.y = grid::unit(0.18, "lines")
  ) +
  guides(
    size = guide_legend(
      direction = "vertical",
      title.position = "top",
      ncol = 1
    )
  )

plot_hi <- patchwork::wrap_plots(
  p1_hi, p2_hi, p3_hi, p4_hi,
  nrow = 1,
  widths = bubble_widths_plot,
  guides = "collect"
) &
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.box.just = "left",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    legend.spacing.x = grid::unit(0.6, "cm"),
    legend.spacing.y = grid::unit(0.15, "cm"),
    legend.box.spacing = grid::unit(0.3, "cm"),
    plot.margin = margin(t = 5, r = 11.5, b = 5, l = 1)
  )

plot_hi
```

### mPAP < 35

```{r mPAP_lo_bubble, fig.width=20.5, fig.height=4.5, eval=TRUE}
plot_oi_lo <- make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables,
  prioritized_tbl_oi_lo_50 %>% inner_join(lr_network_all),
  widths = bubble_widths_plot
)

p1_lo <- plot_oi_lo[[1]]
p2_lo <- plot_oi_lo[[2]]
p3_lo <- plot_oi_lo[[3]]
p4_lo <- plot_oi_lo[[4]]

p1_lo$data$group <- factor(
  dplyr::recode(
    as.character(p1_lo$data$group),
    "mPAP35_hi" = "mPAP ≥ 35",
    "mPAP35_lo" = "mPAP < 35"
  ),
  levels = c("mPAP ≥ 35", "mPAP < 35")
)

p2_lo$data$group <- factor(
  dplyr::recode(
    as.character(p2_lo$data$group),
    "mPAP35_hi" = "mPAP ≥ 35",
    "mPAP35_lo" = "mPAP < 35"
  ),
  levels = c("mPAP ≥ 35", "mPAP < 35")
)

p3_lo$data$group <- factor(
  dplyr::recode(
    as.character(p3_lo$data$group),
    "mPAP35_hi" = "mPAP ≥ 35",
    "mPAP35_lo" = "mPAP < 35"
  ),
  levels = c("mPAP ≥ 35", "mPAP < 35")
)

p1_lo <- p1_lo +
  theme(
    axis.text.y = element_text(face = "bold.italic", size = 15),
    axis.text.x.top = element_text(
      size = 12,
      angle = 45,
      hjust = 0,
      vjust = 0,
      margin = margin(b = 2)
    ),
    strip.text.x.top = element_text(size = 16, face = "bold"),
    strip.text.y.left = element_text(size = 16, face = "bold"),
    panel.spacing.y = grid::unit(0.18, "lines")
  ) +
  guides(
    color = guide_colorbar(
      direction = "vertical",
      title.position = "top",
      barheight = grid::unit(25, "mm"),
      barwidth = grid::unit(4, "mm")
    ),
    size = guide_legend(
      direction = "vertical",
      title.position = "top",
      ncol = 1
    )
  )

p2_lo <- p2_lo +
  theme(
    axis.text.y = element_blank(),
    axis.text.x.top = element_text(
      size = 12,
      angle = 45,
      hjust = 0,
      vjust = 0,
      margin = margin(b = 2)
    ),
    strip.text.x = element_text(size = 16, face = "bold"),
    panel.spacing.y = grid::unit(0.18, "lines")
  ) +
  guides(
    fill = guide_colorbar(
      direction = "vertical",
      title.position = "top",
      barheight = grid::unit(25, "mm"),
      barwidth = grid::unit(4, "mm")
    )
  )

p3_lo <- p3_lo +
  theme(
    axis.text.x.top = element_text(
      size = 12,
      angle = 45,
      hjust = 0,
      vjust = 0,
      margin = margin(b = 2)
    ),
    strip.text.x = element_text(size = 16, face = "bold"),
    panel.spacing.y = grid::unit(0.18, "lines")
  ) +
  guides(
    color = guide_colorbar(
      direction = "vertical",
      title.position = "top",
      barheight = grid::unit(25, "mm"),
      barwidth = grid::unit(4, "mm")
    ),
    size = guide_legend(
      direction = "vertical",
      title.position = "top",
      ncol = 1
    )
  )

p4_lo <- p4_lo +
  theme(
    axis.text.x.top = element_text(
      size = 12,
      angle = 45,
      hjust = 0,
      vjust = 0,
      margin = margin(b = 2)
    ),
    strip.text.x = element_text(size = 16, face = "bold"),
    panel.spacing.y = grid::unit(0.18, "lines")
  ) +
  guides(
    size = guide_legend(
      direction = "vertical",
      title.position = "top",
      ncol = 1
    )
  )

plot_lo <- patchwork::wrap_plots(
  p1_lo, p2_lo, p3_lo, p4_lo,
  nrow = 1,
  widths = bubble_widths_plot,
  guides = "collect"
) &
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.box.just = "left",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    legend.spacing.x = grid::unit(0.6, "cm"),
    legend.spacing.y = grid::unit(0.15, "cm"),
    legend.box.spacing = grid::unit(0.3, "cm"),
    plot.margin = margin(t = 5, r = 11.5, b = 5, l = 1)
  )

plot_lo
```

## mPAP-high intercellular regulatory network

```{r mPAP_hi_network, eval=TRUE}
group_oi = "mPAP35_hi"

lr_target_prior_cor_filtered = multinichenet_output$lr_target_prior_cor %>%
  inner_join(
    multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>%
      distinct(ligand, target, direction_regulation, contrast)
    ) %>%
  inner_join(contrast_tbl) %>%
  filter(group == group_oi)

lr_target_prior_cor_filtered_up = lr_target_prior_cor_filtered %>%
  filter(direction_regulation == "up") %>%
  filter((rank_of_target < top_n_target) & (pearson > 0.33))

lr_target_prior_cor_filtered_down = lr_target_prior_cor_filtered %>%
  filter(direction_regulation == "down") %>%
  filter((rank_of_target < top_n_target) & (pearson < -0.33))

lr_target_prior_cor_filtered = bind_rows(
  lr_target_prior_cor_filtered_up,
  lr_target_prior_cor_filtered_down
  )

lr_target_df = lr_target_prior_cor_filtered %>%
  distinct(group, sender, receiver, ligand, receptor, id, target, direction_regulation)

network = infer_intercellular_regulatory_network(lr_target_df, prioritized_tbl_oi_hi_50)
```

```{r mPAP_hi_network_plot, fig.width=8, fig.height=5, eval=TRUE}
network_plot_data <- network

network_plot_data$links$group <- dplyr::recode(
  as.character(network_plot_data$links$group),
  "mPAP35_hi" = "mPAP ≥ 35"
)

network_graph <- visualize_network(network_plot_data, celltype_cols)

network_graph$plot$layers[[1]]$aes_params$edge_width <- 1.2
network_graph$plot$layers[[1]]$geom_params$arrow <- grid::arrow(
  angle = 30,
  length = grid::unit(4, "mm"),
  ends = "last",
  type = "open"
)

network_graph$plot$layers[[2]]$aes_params$size <- 4.1
network_graph$plot$layers[[2]]$geom_params$label.padding <- grid::unit(0.18, "lines")
network_graph$plot$layers[[2]]$geom_params$label.r <- grid::unit(0.10, "lines")

network_plot <- network_graph$plot +
  scale_x_continuous(expand = expansion(mult = c(0.06, 0.06))) +
  scale_y_continuous(expand = expansion(mult = c(0.06, 0.06))) +
  theme(
    strip.text = element_text(
      size = 17,
      face = "bold",
      colour = "white",
      margin = margin(t = 4, b = 4)
    ),
    legend.title = element_text(
      size = 13,
      face = "bold"
    ),
    legend.text = element_text(
      size = 11
    ),
    legend.key.height = grid::unit(0.55, "cm"),
    legend.key.width = grid::unit(0.45, "cm"),
    legend.spacing.y = grid::unit(0.15, "cm"),
    plot.margin = margin(t = 5, r = 2, b = 5, l = 2)
  )

network_plot
```


## Supplementary Data export

```{r Supplementary_data_export, eval=TRUE}
ligand_target_activity_supp <- multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>%
  inner_join(contrast_tbl)

writexl::write_xlsx(
  list(
    Prioritized_LR_all = multinichenet_output$prioritization_tables$group_prioritization_tbl,
    Top50_LR_mPAP_hi = prioritized_tbl_oi_hi_50,
    Top50_LR_mPAP_lo = prioritized_tbl_oi_lo_50,
    Ligand_target_activity = ligand_target_activity_supp,
    LR_target_correlations = lr_target_prior_cor_filtered,
    Network_links = network$links,
    Network_nodes = network$nodes,
    Network_LR_interactions = network$prioritized_lr_interactions
  ),
  here::here(output_dir_data, "Cell_cell_communication_SuppData.xlsx")
)
```
