---
title: "Cell cell communication analysis using CellChat"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "CellChat"

# Load libraries
library(here)
library(readr)
library(readxl)
library(xlsx)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(gprofiler2)
library(org.Hs.eg.db)
library(AnnotationHub)
library(statmod)
library(edgeR)
library(speckle)
library(OmnipathR)
library(dorothea)
library(enrichplot)
library(png)
library(reactable)
library(UpSetR)
library(ComplexHeatmap)
library(rtracklayer)
library(cowplot)
library(nichenetr)
library(CellChat)


#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```

## Reichart

Load seurat object

```{r}
Reichart <- readRDS("/scratch/gbonaz/Heart_datasets/Reichart_2022/DCM_ACM_heart_cell_atlas.rds")
Reichart_gene_names <- data.frame(
  ensembl_id = rownames(Reichart),
  gene_symbol = Reichart@assays[["RNA"]]@meta.features$feature_name
)
rownames(Reichart@assays[["RNA"]]@data) <- Reichart@assays[["RNA"]]@meta.features$feature_name
rownames(Reichart@assays[["RNA"]]@counts) <- Reichart@assays[["RNA"]]@meta.features$feature_name

Idents(Reichart) <- Reichart$cell_type
```

```{r}
Clinical_data <- read_excel("/scratch/gbonaz/Heart_datasets/Reichart_2022/Reichart_et_al_Science_2022_Selected_clinical_info.xlsx")
add_metadata <- left_join(Reichart[["donor_id"]], Clinical_data)
row.names(add_metadata) <- row.names(Reichart[[]])
Reichart <- AddMetaData(Reichart, metadata = add_metadata)
rm(add_metadata, Clinical_data)
```

```{r}
Reichart <- subset(
  Reichart, 
  disease %in% c("dilated cardiomyopathy", "normal") &
    tissue == "heart left ventricle" &
    cell_type != "unknown"
)

table(Reichart$cell_type, Reichart$disease)
```

```{r}
Reichart@meta.data <- Reichart@meta.data %>%
  mutate(Group_CCC = case_when(
    disease == "normal" ~ "HC",
    disease == "dilated cardiomyopathy" ~ "DCM"
  ))

Reichart@meta.data$cell_type_2 <- make.names(Reichart@meta.data$cell_type)

Reichart@meta.data %>%
  dplyr::select(Group_CCC, donor_id) %>%
  distinct() %>%                
  dplyr::count(Group_CCC)  
```


## Koenig

```{r}
load("/scratch/gbonaz/Heart_datasets/Koenig_2022/GSE183852_DCM_Nuclei.Robj")
Koenig <- nuclei
rm(nuclei)
```

```{r}
Clinical_data <- read_excel("/scratch/gbonaz/Heart_datasets/Koenig_2022/Koenig_et_al_2022_Selected_clinical_info.xlsx")
add_metadata <- left_join(Koenig[["orig.ident"]], Clinical_data)
row.names(add_metadata) <- row.names(Koenig[[]])
Koenig <- AddMetaData(Koenig, metadata = add_metadata)
rm(add_metadata, Clinical_data)
```

```{r}
table(Koenig$Names, Koenig$Condition)
```

```{r}
Koenig@meta.data <- Koenig@meta.data %>%
  mutate(Group_CCC = case_when(
    Condition == "Donor" ~ "HC",
    Condition == "DCM" ~ "DCM"
  ))

Koenig@meta.data$cell_type_2 <- make.names(Koenig@meta.data$Names)

Koenig@meta.data %>%
  dplyr::select(Group_CCC, orig.ident) %>%
  distinct() %>%                
  dplyr::count(Group_CCC)  
```

```{r}
Koenig$Sample <- make.names(Koenig$orig.ident)
table(Koenig$Sample)
```

```{r}
DefaultAssay(Koenig) <- "RNA"
Koenig <- NormalizeData(Koenig)
```

## CellChat Reichart

```{r}
options(stringsAsFactors = FALSE)
```

```{r}
data.input <- Reichart[["RNA"]]@data # normalized data matrix
# For Seurat version >= “5.0.0”, get the normalized data via `seurat_object[["RNA"]]$data`
Reichart$cell_type <- droplevels(Reichart$cell_type)
labels <- Reichart$cell_type
meta <- data.frame(labels = labels, row.names = names(labels)) # create a dataframe of the cell labels
```

```{r}
cellchat <- createCellChat(object = Reichart, group.by = "cell_type", assay = "RNA")
cellchat <- addMeta(cellchat, meta = meta)
cellchat <- setIdent(cellchat, ident.use = "labels") # set "labels" as default cell identity
levels(cellchat@idents) # show factor levels of the cell labels
groupSize <- as.numeric(table(cellchat@idents)) # number of cells in each cell group
groupSize
```

```{r}
CellChatDB <- CellChatDB.human 
cellchat@DB <- CellChatDB
```

### Whole CellChat database

```{r}
# subset the expression data of signaling genes for saving computation cost
cellchat <- subsetData(cellchat) # This step is necessary even if using the whole database
#future::plan("multisession", workers = 4) # do parallel
cellchat <- identifyOverExpressedGenes(cellchat)
cellchat <- identifyOverExpressedInteractions(cellchat)
```

```{r}
cellchat <- computeCommunProb(cellchat, type = "triMean")
cellchat <- filterCommunication(cellchat, min.cells = 10)
```

```{r}
df.net <- subsetCommunication(cellchat, slot.name = "net")
```


```{r}
cellchat <- computeCommunProbPathway(cellchat)
```

```{r}
cellchat <- aggregateNet(cellchat)
```

```{r}
df.pathways <- subsetCommunication(cellchat, slot.name = "netP", thresh = 0.05)
```

### Aggregated cell-cell communication network

```{r, fig.width=9, fig.height=3}
groupSize <- as.numeric(table(cellchat@idents))
par(mfrow = c(1,2), mar = c(0, 0, 2, 0), xpd=TRUE)
netVisual_circle(cellchat@net$count, vertex.weight = groupSize, weight.scale = T, label.edge= F, title.name = "Number of interactions")
netVisual_circle(cellchat@net$weight, vertex.weight = groupSize, weight.scale = T, label.edge= F, title.name = "Interaction weights/strength")
```

### Cell-cell communication network split by sender

```{r, fig.width=9, fig.height=5}
mat <- cellchat@net$weight
par(mfrow = c(2,3), mar = c(0, 4, 0, 4), xpd=TRUE)
for (i in 1:nrow(mat)) {
  mat2 <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  mat2[i, ] <- mat[i, ]
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = T, edge.weight.max = max(mat), title.name = rownames(mat)[i])
}
```


### Interaction strength and number of interactions for each cell type pair

```{r}
netVisual_heatmap(cellchat, color.heatmap = "Reds", slot.name = "net", measure = "weight")
```

```{r}
netVisual_heatmap(cellchat, color.heatmap = "Reds", slot.name = "net", measure = "count")
```


### Dominant senders (sources) and receivers (targets)

```{r}
# Run the analysis without parallel processing
cellchat <- netAnalysis_computeCentrality(cellchat, slot.name = "netP")

gg1 <- netAnalysis_signalingRole_scatter(cellchat)
gg1
```


### Signals contributing the most to outgoing or incoming signaling

```{r, fig.height=13, fig.width=11}
# Signaling role analysis on the aggregated cell-cell communication network from all signaling pathways
ht1 <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "outgoing", width = 10,
  height = 25)
ht2 <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "incoming", width = 10,
  height = 25)
ht1 + ht2
```

```{r, fig.height=6, fig.width=8}
df <- rankNet(cellchat, mode = "single", stacked = T, return.data = T)$signaling.contribution %>%
  arrange(desc(contribution))
ht1 <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "outgoing", width = 7,
  height = 8, signaling = df[1:25,]$name)
ht2 <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "incoming", width = 7,
  height = 8, signaling = df[1:25,]$name)
ht1 + ht2
```


### Top interactions between all cell types

```{r, fig.height=6, fig.width=6}
net <- df.net %>%
  arrange(desc(prob))
netVisual_chord_gene(cellchat, lab.cex = 1, legend.pos.y = 130, legend.pos.x = 5, slot.name = "net", net = net[1:25,])
```

```{r, fig.height=6, fig.width=6}
net <- df.pathways %>%
  arrange(desc(prob))
netVisual_chord_gene(cellchat, lab.cex = 0.65, legend.pos.y = 110, legend.pos.x = 5, slot.name = "netP", net = net[1:25,], small.gap = 2)
```


### Top interactions with endothelial cells as senders

```{r, fig.height=6, fig.width=6}
net <- df.net %>%
  filter(target == "endothelial cell") %>%
  arrange(desc(prob))
netVisual_chord_gene(cellchat, lab.cex = 1, legend.pos.y = 130, legend.pos.x = 5, slot.name = "net", net = net[1:25,], sources.use = "endothelial cell")
```

```{r, fig.height=8, fig.width=10}
net <- df.pathways %>%
  arrange(desc(prob))
netVisual_chord_gene(cellchat, sources.use = "endothelial cell", lab.cex = 1,legend.pos.y = 150, slot.name = "netP", net = net[1:100,])
```

```{r}
net <- df.net %>%
  filter(source == "endothelial cell") %>%
  arrange(desc(prob))
df <- as.data.frame(net[1:25,]$interaction_name)
colnames(df) <- "interaction_name"
netVisual_bubble(cellchat, remove.isolate = FALSE, pairLR.use = df, sources.use = "endothelial cell", angle.x = 45)
```


## CellChat Koenig

```{r}
options(stringsAsFactors = FALSE)
```

```{r}
data.input <- Koenig[["RNA"]]@data # normalized data matrix
# For Seurat version >= “5.0.0”, get the normalized data via `seurat_object[["RNA"]]$data`
labels <- Koenig$Names
meta <- data.frame(labels = labels, row.names = names(labels)) # create a dataframe of the cell labels
```

```{r}
cellchat <- createCellChat(object = Koenig, group.by = "Names", assay = "RNA")
cellchat <- addMeta(cellchat, meta = meta)
cellchat <- setIdent(cellchat, ident.use = "labels") # set "labels" as default cell identity
levels(cellchat@idents) # show factor levels of the cell labels
groupSize <- as.numeric(table(cellchat@idents)) # number of cells in each cell group
groupSize
```

```{r}
CellChatDB <- CellChatDB.human 
cellchat@DB <- CellChatDB
```

### Whole CellChat database

```{r}
# subset the expression data of signaling genes for saving computation cost
cellchat <- subsetData(cellchat) # This step is necessary even if using the whole database
#future::plan("multisession", workers = 4) # do parallel
cellchat <- identifyOverExpressedGenes(cellchat)
cellchat <- identifyOverExpressedInteractions(cellchat)
```

```{r}
cellchat <- computeCommunProb(cellchat, type = "triMean")
cellchat <- filterCommunication(cellchat, min.cells = 10)
```

```{r}
df.net <- subsetCommunication(cellchat, slot.name = "net")
```


```{r}
cellchat <- computeCommunProbPathway(cellchat)
```

```{r}
cellchat <- aggregateNet(cellchat)
```

```{r}
df.pathways <- subsetCommunication(cellchat, slot.name = "netP", thresh = 0.05)
```

### Aggregated cell-cell communication network

```{r, fig.width=9, fig.height=3}
groupSize <- as.numeric(table(cellchat@idents))
par(mfrow = c(1,2), mar = c(0, 0, 2, 0), xpd=TRUE)
netVisual_circle(cellchat@net$count, vertex.weight = groupSize, weight.scale = T, label.edge= F, title.name = "Number of interactions")
netVisual_circle(cellchat@net$weight, vertex.weight = groupSize, weight.scale = T, label.edge= F, title.name = "Interaction weights/strength")
```

### Cell-cell communication network split by sender

```{r, fig.width=9, fig.height=5}
mat <- cellchat@net$weight
par(mfrow = c(2,3), mar = c(0, 4, 0, 4), xpd=TRUE)
for (i in 1:nrow(mat)) {
  mat2 <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  mat2[i, ] <- mat[i, ]
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = T, edge.weight.max = max(mat), title.name = rownames(mat)[i])
}
```


### Interaction strength and number of interactions for each cell type pair

```{r}
netVisual_heatmap(cellchat, color.heatmap = "Reds", slot.name = "net", measure = "weight")
```

```{r}
netVisual_heatmap(cellchat, color.heatmap = "Reds", slot.name = "net", measure = "count")
```


### Dominant senders (sources) and receivers (targets)

```{r}
# Run the analysis without parallel processing
cellchat <- netAnalysis_computeCentrality(cellchat, slot.name = "netP")

gg1 <- netAnalysis_signalingRole_scatter(cellchat)
gg1
```


### Signals contributing the most to outgoing or incoming signaling

```{r, fig.height=13, fig.width=11}
# Signaling role analysis on the aggregated cell-cell communication network from all signaling pathways
ht1 <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "outgoing", width = 10,
  height = 25)
ht2 <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "incoming", width = 10,
  height = 25)
ht1 + ht2
```

```{r, fig.height=6, fig.width=8}
df <- rankNet(cellchat, mode = "single", stacked = T, return.data = T)$signaling.contribution %>%
  arrange(desc(contribution))
ht1 <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "outgoing", width = 7,
  height = 8, signaling = df[1:25,]$name)
ht2 <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "incoming", width = 7,
  height = 8, signaling = df[1:25,]$name)
ht1 + ht2
```


### Top interactions between all cell types

```{r, fig.height=6, fig.width=6}
net <- df.net %>%
  arrange(desc(prob))
netVisual_chord_gene(cellchat, lab.cex = 1, legend.pos.y = 130, legend.pos.x = 5, slot.name = "net", net = net[1:25,])
```

```{r, fig.height=6, fig.width=6}
net <- df.pathways %>%
  arrange(desc(prob))
netVisual_chord_gene(cellchat, lab.cex = 0.65, legend.pos.y = 110, legend.pos.x = 5, slot.name = "netP", net = net[1:25,], small.gap = 2)
```


### Top interactions with endothelial cells as senders

```{r, fig.height=6, fig.width=6}
net <- df.net %>%
  filter(target == "Endothelium") %>%
  arrange(desc(prob))
netVisual_chord_gene(cellchat, lab.cex = 1, legend.pos.y = 130, legend.pos.x = 5, slot.name = "net", net = net[1:25,], sources.use = "Endothelium")
```

```{r, fig.height=8, fig.width=10}
net <- df.pathways %>%
  arrange(desc(prob))
netVisual_chord_gene(cellchat, sources.use = "Endothelium", lab.cex = 1,legend.pos.y = 150, slot.name = "netP", net = net[1:100,])
```

```{r}
net <- df.net %>%
  filter(source == "Endothelium") %>%
  arrange(desc(prob))
df <- as.data.frame(net[1:25,]$interaction_name)
colnames(df) <- "interaction_name"
netVisual_bubble(cellchat, remove.isolate = FALSE, pairLR.use = df, sources.use = "Endothelium", angle.x = 45)
```
