---
title: "Cell cell communication analysis using MultiNicheNet"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "Cell_cell_communication_LV2"

# Load libraries
library(here)
library(readr)
library(readxl)
library(xlsx)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(ArchR)
library(clustree)
library(harmony)
library(gprofiler2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationHub)
library(ReactomePA)
library(statmod)
library(edgeR)
library(speckle)
library(EnhancedVolcano)
library(decoupleR)
library(OmnipathR)
library(dorothea)
library(enrichplot)
library(png)
library(reactable)
library(UpSetR)
library(ComplexHeatmap)
library(rtracklayer)
library(cowplot)
library(nichenetr)
library(multinichenetr)
library(muscat)


#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```

## Reichart

Load seurat object

```{r Import Reichart}
Reichart <- readRDS("/scratch/gbonaz/Heart_datasets/Reichart_2022/DCM_ACM_heart_cell_atlas.rds")
Reichart_gene_names <- data.frame(
  ensembl_id = rownames(Reichart),
  gene_symbol = Reichart@assays[["RNA"]]@meta.features$feature_name
)
rownames(Reichart@assays[["RNA"]]@data) <- Reichart@assays[["RNA"]]@meta.features$feature_name
rownames(Reichart@assays[["RNA"]]@counts) <- Reichart@assays[["RNA"]]@meta.features$feature_name

Idents(Reichart) <- Reichart$cell_type
```

```{r add metadata Reichart}
Clinical_data <- read_excel("/scratch/gbonaz/Heart_datasets/Reichart_2022/Reichart_et_al_Science_2022_Selected_clinical_info.xlsx")
add_metadata <- left_join(Reichart[["donor_id"]], Clinical_data)
row.names(add_metadata) <- row.names(Reichart[[]])
Reichart <- AddMetaData(Reichart, metadata = add_metadata)
rm(add_metadata, Clinical_data)
```

```{r subset Reichart}
Reichart <- subset(
  Reichart, 
  disease %in% c("dilated cardiomyopathy", "normal") &
    tissue == "heart left ventricle" &
    cell_type != "unknown"
)

table(Reichart$cell_type, Reichart$disease)
```

```{r}
Reichart@meta.data <- Reichart@meta.data %>%
  mutate(Group_CCC = case_when(
    disease == "normal" ~ "HC",
    disease == "dilated cardiomyopathy" ~ "DCM"
  ))

Reichart@meta.data$cell_type_2 <- make.names(Reichart@meta.data$cell_type)

Reichart@meta.data %>%
  dplyr::select(Group_CCC, donor_id) %>%
  distinct() %>%                
  dplyr::count(Group_CCC)  
```

```{r}
organism = "human"

if(organism == "human"){
  
  lr_network_all = 
    readRDS(url(
      "https://zenodo.org/record/10229222/files/lr_network_human_allInfo_30112033.rds"
      )) %>% 
    mutate(
      ligand = convert_alias_to_symbols(ligand, organism = organism), 
      receptor = convert_alias_to_symbols(receptor, organism = organism))
  
  lr_network_all = lr_network_all  %>% 
    mutate(ligand = make.names(ligand), receptor = make.names(receptor)) 
  
  lr_network = lr_network_all %>% 
    distinct(ligand, receptor)
  
  ligand_target_matrix = readRDS(url(
    "https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds"
    ))
  
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  
  lr_network = lr_network %>% filter(ligand %in% colnames(ligand_target_matrix))
  ligand_target_matrix = ligand_target_matrix[, lr_network$ligand %>% unique()]
  
} else if(organism == "mouse"){
  
  lr_network_all = readRDS(url(
    "https://zenodo.org/record/10229222/files/lr_network_mouse_allInfo_30112033.rds"
    )) %>% 
    mutate(
      ligand = convert_alias_to_symbols(ligand, organism = organism), 
      receptor = convert_alias_to_symbols(receptor, organism = organism))
  
  lr_network_all = lr_network_all  %>% 
    mutate(ligand = make.names(ligand), receptor = make.names(receptor)) 
  lr_network = lr_network_all %>% 
    distinct(ligand, receptor)
  
  ligand_target_matrix = readRDS(url(
    "https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final_mouse.rds"
    ))
  
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  
  lr_network = lr_network %>% filter(ligand %in% colnames(ligand_target_matrix))
  ligand_target_matrix = ligand_target_matrix[, lr_network$ligand %>% unique()]
  
}
```

```{r}
sce <- Seurat::as.SingleCellExperiment(Reichart, assay = "RNA")

sce = alias_to_symbol_SCE(sce, "human") %>% makenames_SCE()

sce = scuttle::logNormCounts(sce)
```

```{r}
group_id = "Group_CCC"
sample_id = "donor_id"
celltype_id = "cell_type_2"
covariates = NA
batches = NA
```

```{r}
contrasts_oi = c("'DCM-HC','HC-DCM'")
contrast_tbl = tibble(contrast = 
                        c("DCM-HC","HC-DCM"), 
                      group = c("DCM","HC"))
```

```{r}
senders_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
receivers_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% 
            c(senders_oi, receivers_oi)
          ]
```

```{r}
conditions_keep = c("HC", "DCM")
sce = sce[, SummarizedExperiment::colData(sce)[,group_id] %in% 
            conditions_keep
          ]
```

```{r}
multinichenet_output <- readRDS(here::here("output", "Cell_cell_communication_LV", "multinichenet_output_Reichart.rds"))
```

### Endothelial cells as receivers

```{r}
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 200, 
  receivers_oi  = "endothelial.cell",
  rank_per_group = FALSE
  )
```

```{r, fig.height=5, fig.width=5}
prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

par(mai = c(0.5, 0.4, 0.6, 0.4), xpd = NA)
circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{r}
group_oi = "DCM"
```

```{r}
prioritized_tbl_oi_hi_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=18, fig.height=40}
plot_oi <- make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables,
  prioritized_tbl_oi_hi_50 %>% inner_join(lr_network_all),
  widths = c(18, 3, 3,3)   # sample panel, activity tiles, CS panel, Omnipath
)
plot_oi
```

```{r}
group_oi = "HC"
```

```{r}
prioritized_tbl_oi_lo_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=18, fig.height=40}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_lo_50 %>% inner_join(lr_network_all),
  widths = c(18, 3, 3,3)
  )
plot_oi
```

### Endothelial cells as senders

```{r}
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 200, 
  senders_oi  = "endothelial.cell",
  rank_per_group = FALSE
  )
```

```{r, fig.height=8, fig.width=8}
prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

par(mai = c(0.5, 0.4, 0.6, 0.4), xpd = NA)
circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{r}
group_oi = "DCM"
```

```{r}
prioritized_tbl_oi_hi_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=18, fig.height=40}
plot_oi <- make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables,
  prioritized_tbl_oi_hi_50 %>% inner_join(lr_network_all),
  widths = c(18, 3, 3,3)   # sample panel, activity tiles, CS panel, Omnipath
)
plot_oi
```

```{r}
group_oi = "HC"
```

```{r}
prioritized_tbl_oi_lo_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=18, fig.height=40}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_lo_50 %>% inner_join(lr_network_all),
  widths = c(18, 3, 3,3)
  )
plot_oi
```


### Cardiomyocytes & Fibroblasts

```{r}
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 50, 
  senders_oi  = c("cardiac.muscle.cell", "fibroblast.of.cardiac.tissue"),
  receivers_oi  = c("cardiac.muscle.cell", "fibroblast.of.cardiac.tissue"),
  rank_per_group = FALSE
  )
```

```{r, fig.height=5, fig.width=5}
prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

par(mai = c(0.5, 0.4, 0.6, 0.4), xpd = NA)
circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{r}
group_oi = "DCM"
```

```{r}
prioritized_tbl_oi_hi_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=18, fig.height=10}
plot_oi <- make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables,
  prioritized_tbl_oi_hi_50 %>% inner_join(lr_network_all),
  widths = c(18, 3, 3,3)   # sample panel, activity tiles, CS panel, Omnipath
)
plot_oi
```

```{r}
group_oi = "HC"
```

```{r}
prioritized_tbl_oi_lo_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=18, fig.height=10}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_lo_50 %>% inner_join(lr_network_all),
  widths = c(18, 3, 3,3)
  )
plot_oi
```

## Koenig

```{r Import Koenig}
load("/scratch/gbonaz/Heart_datasets/Koenig_2022/GSE183852_DCM_Nuclei.Robj")
Koenig <- nuclei
rm(nuclei)
```

```{r add_metadata Koenig}
Clinical_data <- read_excel("/scratch/gbonaz/Heart_datasets/Koenig_2022/Koenig_et_al_2022_Selected_clinical_info.xlsx")
add_metadata <- left_join(Koenig[["orig.ident"]], Clinical_data)
row.names(add_metadata) <- row.names(Koenig[[]])
Koenig <- AddMetaData(Koenig, metadata = add_metadata)
rm(add_metadata, Clinical_data)
```

```{r}
table(Koenig$Names, Koenig$Condition)
```

```{r}
Koenig@meta.data <- Koenig@meta.data %>%
  mutate(Group_CCC = case_when(
    Condition == "Donor" ~ "HC",
    Condition == "DCM" ~ "DCM"
  ))

Koenig@meta.data$cell_type_2 <- make.names(Koenig@meta.data$Names)

Koenig@meta.data %>%
  dplyr::select(Group_CCC, orig.ident) %>%
  distinct() %>%                
  dplyr::count(Group_CCC)  
```

```{r}
Koenig$Sample <- make.names(Koenig$orig.ident)
table(Koenig$Sample)
```

```{r}
sce <- Seurat::as.SingleCellExperiment(Koenig, assay = "RNA")

sce = alias_to_symbol_SCE(sce, "human") %>% makenames_SCE()

sce = scuttle::logNormCounts(sce)
```

```{r}
group_id = "Group_CCC"
sample_id = "Sample"
celltype_id = "cell_type_2"
covariates = NA
batches = NA
```

```{r}
contrasts_oi = c("'DCM-HC','HC-DCM'")
contrast_tbl = tibble(contrast = 
                        c("DCM-HC","HC-DCM"), 
                      group = c("DCM","HC"))
```

```{r}
senders_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
receivers_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% 
            c(senders_oi, receivers_oi)
          ]
```

```{r}
conditions_keep = c("HC", "DCM")
sce = sce[, SummarizedExperiment::colData(sce)[,group_id] %in% 
            conditions_keep
          ]
```

```{r}
multinichenet_output_K <- readRDS(here::here("output", "Cell_cell_communication_LV", "multinichenet_output_Koenig.rds"))
```

### Endothelial cells as receivers

```{r}
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output_K$prioritization_tables, 
  top_n = 200, 
  receivers_oi  = "Endothelium",
  rank_per_group = FALSE
  )
```

```{r, fig.height=8, fig.width=8}
prioritized_tbl_oi = 
  multinichenet_output_K$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender <- colorRampPalette(RColorBrewer::brewer.pal(11, "Spectral"))(length(senders_receivers)) %>%
  magrittr::set_names(senders_receivers)
colors_receiver <- colorRampPalette(RColorBrewer::brewer.pal(11, "Spectral"))(length(senders_receivers)) %>%
  magrittr::set_names(senders_receivers)

par(mai = c(0.5, 0.4, 0.6, 0.4), xpd = NA)
circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{r}
group_oi = "DCM"
```

```{r}
prioritized_tbl_oi_hi_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=18, fig.height=40}
plot_oi <- make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output_K$prioritization_tables,
  prioritized_tbl_oi_hi_50 %>% inner_join(lr_network_all),
  widths = c(18, 3, 3,3)   # sample panel, activity tiles, CS panel, Omnipath
)
plot_oi
```

```{r}
group_oi = "HC"
```

```{r}
prioritized_tbl_oi_lo_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=18, fig.height=40}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output_K$prioritization_tables, 
  prioritized_tbl_oi_lo_50 %>% inner_join(lr_network_all),
  widths = c(18, 3, 3,3)
  )
plot_oi
```

### Endothelial cells as senders

```{r}
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output_K$prioritization_tables, 
  top_n = 200, 
  senders_oi  = "Endothelium",
  rank_per_group = FALSE
  )
```

```{r, fig.height=8, fig.width=8}
prioritized_tbl_oi = 
  multinichenet_output_K$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender <- colorRampPalette(RColorBrewer::brewer.pal(11, "Spectral"))(length(senders_receivers)) %>%
  magrittr::set_names(senders_receivers)
colors_receiver <- colorRampPalette(RColorBrewer::brewer.pal(11, "Spectral"))(length(senders_receivers)) %>%
  magrittr::set_names(senders_receivers)

par(mai = c(0.5, 0.4, 0.6, 0.4), xpd = NA)
circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{r}
group_oi = "DCM"
```

```{r}
prioritized_tbl_oi_hi_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=18, fig.height=40}
plot_oi <- make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output_K$prioritization_tables,
  prioritized_tbl_oi_hi_50 %>% inner_join(lr_network_all),
  widths = c(18, 3, 3,3)   # sample panel, activity tiles, CS panel, Omnipath
)
plot_oi
```

```{r}
group_oi = "HC"
```

```{r}
prioritized_tbl_oi_lo_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=18, fig.height=40}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output_K$prioritization_tables, 
  prioritized_tbl_oi_lo_50 %>% inner_join(lr_network_all),
  widths = c(18, 3, 3,3)
  )
plot_oi
```

### Cardiomyocytes & Fibroblasts

```{r}
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output_K$prioritization_tables, 
  top_n = 50, 
  senders_oi  = c("Cardiomyocytes", "Fibroblasts"),
  receivers_oi  = c("Cardiomyocytes", "Fibroblasts"),
  rank_per_group = FALSE
  )
```

```{r, fig.height=5, fig.width=5}
prioritized_tbl_oi = 
  multinichenet_output_K$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

par(mai = c(0.5, 0.4, 0.6, 0.4), xpd = NA)
circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{r}
group_oi = "DCM"
```

```{r}
prioritized_tbl_oi_hi_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=18, fig.height=10}
plot_oi <- make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output_K$prioritization_tables,
  prioritized_tbl_oi_hi_50 %>% inner_join(lr_network_all),
  widths = c(18, 3, 3,3)   # sample panel, activity tiles, CS panel, Omnipath
)
plot_oi
```

```{r}
group_oi = "HC"
```

```{r}
prioritized_tbl_oi_lo_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=18, fig.height=10}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output_K$prioritization_tables, 
  prioritized_tbl_oi_lo_50 %>% inner_join(lr_network_all),
  widths = c(18, 3, 3,3)
  )
plot_oi
```



## DYSF DDAH2

```{r}
DefaultAssay(Koenig) <- "RNA"
Koenig <- NormalizeData(Koenig)
```


```{r, fig.height=4, fig.width=8}
ggplot2::update_geom_defaults("point", list(size = 0.001, alpha = 0.02))

p1 <- VlnPlot(subset(Reichart, cell_type_2 %in% c("cardiac.muscle.cell")), features = c("DYSF","DDAH2"),
                      group.by = "Group_CCC", assay = "RNA",
                      pt.size = 0.001)

p2 <- VlnPlot(subset(Koenig, Names %in% c("Cardiomyocytes")), features = c("DYSF","DDAH2"),
                      group.by = "Group_CCC", assay = "RNA",
                      pt.size = 0.001)

# simple title grobs (one per column)
t1 <- ggplot() + labs(title = "Reichart et al. 2022\nCardiomyocytes") +
  theme_void() + theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15))

t2 <- ggplot() + labs(title = "Koenig et al. 2022\nCardiomyocytes") +
  theme_void() + theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15))

# arrange: titles on top row, plots below
((t1 | t2) / (p1 | p2)) + plot_layout(heights = c(0.08, 1))

```

```{r, fig.height=4, fig.width=8}
ggplot2::update_geom_defaults("point", list(size = 0.001, alpha = 0.02))

p1 <- VlnPlot(subset(Reichart, cell_type_2 %in% c("fibroblast.of.cardiac.tissue")), features = c("DYSF","DDAH2"),
                      group.by = "Group_CCC", assay = "RNA",
                      pt.size = 0.001)

p2 <- VlnPlot(subset(Koenig, Names %in% c("Fibroblasts")), features = c("DYSF","DDAH2"),
                      group.by = "Group_CCC", assay = "RNA",
                      pt.size = 0.001)

# simple title grobs (one per column)
t1 <- ggplot() + labs(title = "Reichart et al. 2022\nFibroblasts") +
  theme_void() + theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15))

t2 <- ggplot() + labs(title = "Koenig et al. 2022\nFibroblasts") +
  theme_void() + theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15))

# arrange: titles on top row, plots below
((t1 | t2) / (p1 | p2)) + plot_layout(heights = c(0.08, 1))

```

```{r, fig.height=4, fig.width=8}
ggplot2::update_geom_defaults("point", list(size = 0.001, alpha = 0.02))

p1 <- VlnPlot(subset(Reichart, cell_type_2 %in% c("endothelial.cell")), features = c("DYSF","DDAH2"),
                      group.by = "Group_CCC", assay = "RNA",
                      pt.size = 0.001)

p2 <- VlnPlot(subset(Koenig, Names %in% c("Endothelium")), features = c("DYSF","DDAH2"),
                      group.by = "Group_CCC", assay = "RNA",
                      pt.size = 0.001)

# simple title grobs (one per column)
t1 <- ggplot() + labs(title = "Reichart et al. 2022\nEndothelial cells") +
  theme_void() + theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15))

t2 <- ggplot() + labs(title = "Koenig et al. 2022\nEndothelial cells") +
  theme_void() + theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15))

# arrange: titles on top row, plots below
((t1 | t2) / (p1 | p2)) + plot_layout(heights = c(0.08, 1))
ggplot2::update_geom_defaults("point", list(size = 0.5, alpha = 1))

```

## Reichart Fibroblasts DE DCM vs HC

```{r, eval=FALSE}
library(zellkonverter)

sce <- readH5AD("/scratch/gbonaz/Heart_datasets/Reichart_2022/c6851627-0ee9-40ff-9ce4-0dbdec73aa0b.h5ad")

fb <- as.Seurat(sce, counts = "X", data = NULL)
DefaultAssay(fb) <- "originalexp"

Reichart_gene_names <- data.frame(
  ensembl_id = rownames(fb),
  gene_symbol = fb@assays[["originalexp"]]@meta.features$feature_name
)
rownames(fb@assays[["originalexp"]]@data) <- fb@assays[["originalexp"]]@meta.features$feature_name
rownames(fb@assays[["originalexp"]]@counts) <- fb@assays[["originalexp"]]@meta.features$feature_name
Clinical_data <- read_excel("/scratch/gbonaz/Heart_datasets/Reichart_2022/Reichart_et_al_Science_2022_Selected_clinical_info.xlsx")
add_metadata <- left_join(fb[["donor_id"]], Clinical_data)
row.names(add_metadata) <- row.names(fb[[]])
fb <- AddMetaData(fb, metadata = add_metadata)
rm(add_metadata, Clinical_data)
saveRDS(fb, here::here(output_dir_data, "Reichart_FB.rds"))
```

```{r}
fb <- readRDS(here::here(output_dir_data, "Reichart_FB.rds"))
DefaultAssay(fb) <- "originalexp"

Idents(fb) <- fb$cell_states
```

```{r}
fb <- subset(
  fb, 
  disease %in% c("dilated cardiomyopathy", "normal") &
    tissue == "heart left ventricle"
)

table(fb$cell_type, fb$disease)
```

```{r}
fb@meta.data <- fb@meta.data %>%
  mutate(Group_CCC = case_when(
    disease == "normal" ~ "HC",
    disease == "dilated cardiomyopathy" ~ "DCM"
  ))

fb@meta.data$cell_states_2 <- make.names(fb@meta.data$cell_states)

```

```{r, fig.width=9, fig.height=4}
fb$Group <- fb$Group_CCC 
DimPlot(fb, group.by = c("Group", "cell_states"), reduction = "X_umap", label = F, shuffle = TRUE, pt.size = 0.1) 
```

```{r, fig.width=8, fig.height=4}
DimPlot(fb, group.by = c("cell_states"), reduction = "X_umap", label = F, shuffle = TRUE, split.by = "Group", pt.size = 0.001) + 
  theme(plot.title = element_blank())
```

Prepare matadata for differential expression testing

```{r}
fb_metadata <- fb@meta.data %>%
  dplyr::select(donor_id, Primary.Genetic.Diagnosis, disease, sex, Group_CCC) %>%
  unique()

rownames(fb_metadata) <- fb_metadata$donor_id
table(fb$donor_id)[table(fb$donor_id) > 0]

all.equal(rownames(fb_metadata), names(table(fb$donor_id)[table(fb$donor_id) > 0]))

reactable(fb_metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

Remove genes expressed in <1% of cells

```{r}
expr_matrix <- GetAssayData(object = fb, assay = "originalexp", slot = "counts")

# Calculate the number of cells where each gene is expressed (non-zero counts)
non_zero_cells <- rowSums(expr_matrix > 0)

# Calculate the total number of cells
total_cells <- ncol(expr_matrix)

# Calculate the percentage of cells expressing each gene
percent_cells <- (non_zero_cells / total_cells) * 100

# Create a dataframe to store the results
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- fb[which(rownames(fb) %in% keep_genes),]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

Create and prepare pseudobulk object for differential expression analysis

```{r}
data <- RenameAssays(data, originalexp = "RNA")

Assays(data)   
DefaultAssay(data) <- "RNA"

pseudocounts <- Seurat2PB(data, sample="donor_id", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(fb_metadata$donor_id, pseudocounts$samples$sample)]
all(fb_metadata$donor_id == pseudocounts$samples$sample)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

DE analysis

```{r}
fb_metadata$disease <- factor(fb_metadata$disease, levels = c("normal", "dilated cardiomyopathy"))
design<- model.matrix(~ disease, data = fb_metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_fb <- topTags(fit, n = Inf)$table
results_fb <- merge(results_fb, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_fb <- results_fb %>%
  dplyr::select(-feature_biotype, -feature_is_filtered, -feature_name, -feature_reference, -feature_length)
results_fb$up_down <- ifelse(
  results_fb$logFC > 0.5 & results_fb$FDR < 0.05, "Increased",
  ifelse(results_fb$logFC < -0.5 & results_fb$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )

signif_fb <- results_fb %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

write.csv(results_fb, file = here::here(output_dir_data, "fb_DE_Results.csv"), quote=F, row.names = F)
write.csv(signif_fb, file = here::here(output_dir_data, "fb_DE_Significant.csv"), quote=F, row.names = F)
```

```{r, fig.height=6, fig.width=4}
keyvals <- rep("grey", nrow(results_fb))
names(keyvals) <- rep("Not significant", nrow(results_fb))

keyvals[results_fb$up_down == "Increased"] <- "#d62728"
names(keyvals)[keyvals == "#d62728"] <- "Increased"
keyvals[results_fb$up_down == "Decreased"] <- "#1f77b4"
names(keyvals)[keyvals == "#1f77b4"] <- "Decreased"

volcano <- EnhancedVolcano(results_fb,
                lab = results_fb$gene,
                x = "logFC",
                y = "FDR",
                labSize = 0,
                legendLabSize = 8.5,
                titleLabSize = 15,
                subtitleLabSize = 14,
                axisLabSize = 12,
                captionLabSize = 9,
                pointSize = 0.5,
                FCcutoff = 0.5,
                pCutoff  = 0.05,
                ylim = c(0, 16),
                colAlpha = 1,
                drawConnectors = FALSE,
                subtitle = NULL,
                title = "LV Fibroblasts\nDCM (n=49) vs HC (n=19)",
                colCustom = keyvals
) + theme(legend.position = "bottom")

volcano
```

```{r, fig.width=10, fig.height=5}
signif_fb$metric <- signif_fb$logFC*-log10(signif_fb$PValue)

DEGs_top_up <- signif_fb %>% 
  dplyr::arrange(desc(metric)) %>%
  slice_head(n=10)
DEGs_top_down <- signif_fb %>% 
  dplyr::arrange(metric) %>%
  slice_head(n=10)

DEGs_top <- dplyr::bind_rows(DEGs_top_up, DEGs_top_down)
rm(DEGs_top_up, DEGs_top_down)

counts <- cpm(pseudocounts, log = TRUE)

fb_metadata$Group <- fb_metadata$Group_CCC

fb_metadata <- fb_metadata %>%
  dplyr::arrange(Group)
counts_heatmap <- counts[DEGs_top$gene, ]
counts_heatmap <- counts_heatmap[, fb_metadata$donor_id, drop = FALSE]


# 1) Build the annotation data.frame aligned to your matrix columns
ann_col <- fb_metadata %>%
  dplyr::filter(donor_id %in% colnames(counts)) %>%                            # if you have a donor_id key
  dplyr::arrange(match(donor_id, colnames(counts))) %>%
  dplyr::select(Group) %>%
  as.data.frame()
rownames(ann_col) <- colnames(counts)

# (optional) lock desired order
ann_col$Group <- factor(ann_col$Group, levels = c("HC","DCM"))

# 2) Make a named color vector that matches the present levels
lvl <- levels(ann_col$Group)
# if you specifically want HC/DCM colors:
palette_group <- c(HC = "#1f77b4", DCM = "#d62728")
ann_colors <- list(Group = setNames(palette_group[lvl], lvl))
# If levels are unknown beforehand, use this instead:
# vals <- unique(as.character(ann_col$Group))
# ann_colors <- list(Group = setNames(c("#1f77b4","#d62728","#2ca02c","#9467bd")[seq_along(vals)], vals))

# 3) Heatmap colors
heatmap_cols <- colorRampPalette(c("#004C8C", "white", "#A63A2A"))(101)

# 4) Plot
pheatmap::pheatmap(
  counts_heatmap,
  name = " ",
  angle_col = 45,                 # numeric is fine
  color = heatmap_cols,
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,
  show_rownames = TRUE,
  annotation_legend = TRUE,
  border_color = NA,
  fontsize = 10,
  fontsize_row = 13,
  fontsize_col = 6,
  cellheight = 15,
  gaps_col = c(0,0,0,0)
  )

```

```{r, fig.height=4, fig.width=20}
fb_metadata$Group_CCC <- factor(fb_metadata$Group_CCC, levels = c("HC", "DCM"))

props <- getTransformedProps(fb$cell_states, fb$donor_id, transform="logit")


DA_metadata <- fb_metadata %>%
  dplyr::arrange(match(donor_id, colnames(props$Counts)))

identical(colnames(props$Counts), DA_metadata$donor_id)

design <- model.matrix(~ 0 + DA_metadata$Group_CCC)
colnames(design) <- c("HC", "DCM")
mycontr <- makeContrasts(DCM-HC, levels=design)
differential_abundance <- propeller.ttest(props, design, contrasts = mycontr, robust=TRUE, trend=FALSE, sort=TRUE)

write.csv(differential_abundance, file = here::here(output_dir_data, "Differential_Abundance_fb.csv"), row.names = F)

reactable(differential_abundance, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

```

```{r, fig.height=3, fig.width=16}
prop_mat <- as.matrix(props$Proportions)
stopifnot(is.matrix(prop_mat))

# 2) Long format
prop_long <- as.data.frame.matrix(prop_mat) %>%           # <-- important: *.matrix
  tibble::rownames_to_column("state") %>%
  pivot_longer(cols = -state, names_to = "donor_id", values_to = "prop") %>%
  mutate(donor_id = as.character(donor_id),
         state    = factor(state, levels = rownames(prop_mat))) %>%
  left_join(DA_metadata %>% dplyr::select(donor_id, Group_CCC) %>% distinct(),
            by = "donor_id") %>%
  mutate(Group_CCC = factor(Group_CCC, levels = c("HC","DCM")))

# ---- Build labels from propeller results ----
# rownames of differential_abundance should be the state names
lab_df <- differential_abundance %>%
  mutate(state = rownames(differential_abundance),
         label = ifelse(FDR < 0.05, paste0("FDR=", signif(FDR, 2)), "ns")) %>%
  dplyr::select(state, label)

# y-position for labels (a little above the highest point per facet)
y_pos <- prop_long %>%
  group_by(state) %>%
  summarise(y = max(prop, na.rm = TRUE) * 1.08, .groups = "drop")

lab_plot <- left_join(lab_df, y_pos, by = "state")

# ---- Colors (classic blue/red) ----
cols <- c(HC = "#1f77b4", DCM = "#d62728")

# ---- Plot ----
p <- ggplot(prop_long, aes(x = Group_CCC, y = prop, fill = Group_CCC, color = Group_CCC)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.35) +
  geom_jitter(width = 0.12, size = 1.8, alpha = 0.9) +
  facet_wrap(~ state, nrow = 1, scales = "free_y") +
  geom_text(data = lab_plot, aes(x = 1.5, y = y, label = label),
            inherit.aes = FALSE, size = 4.5) +
  scale_fill_manual(values = cols) +
  scale_color_manual(values = cols) +
  labs(x = NULL, y = "Proportions") +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 14),
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 13),
    panel.spacing = unit(1.5, "lines")
  )

p
```

```{r}
DefaultAssay(fb) <- "originalexp"
Idents(fb) <- "cell_states"
Markers <- FindAllMarkers(fb, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 0.5)
write.csv(Markers, here::here(output_dir_data, "Reichart_fb_Markers.csv"))
Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
write.csv(Markers_top10, here::here(output_dir_data, "Reichart_fb_top10.csv"))
Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
write.csv(Markers_top3, here::here(output_dir_data, "Reichart_fb_top10.csv"))
```

```{r}
Idents(fb) <- factor(
  Idents(fb),
  levels = rev(levels(Idents(fb)))
)
```

```{r, fig.width=10.5, fig.height=6}
p <- DotPlot_scCustom(
        fb,
        features       = Markers_top3$gene,
        x_lab_rotate   = TRUE
)

p + theme(
      axis.text.x  = element_text(size = 12,        
                                  angle = 60,          
                                  vjust = 1,
                                  hjust = 1),
      axis.text.y  = element_text(size = 15))         
```

```{r}
heatmap_cols <- colorRampPalette(RColorBrewer::brewer.pal(9,"RdBu"))(256)
heatmap_cols <- rev(heatmap_cols[1:256])
```

```{r}
fb <- ScaleData(fb)
```

```{r, fig.width=8, fig.height=6}
genes.use <- intersect(
  Markers_top10$gene,
  rownames(fb[["originalexp"]]@scale.data)
)
DefaultAssay(fb) <- "originalexp"
Heatmap <- DoHeatmap(subset(fb, cells = sample(colnames(fb), 10000)),
                     features = genes.use,
                     label = FALSE,
                     draw.line = TRUE,
                     group.by = "cell_states",
                     lines.width = 1) +
  scale_fill_gradientn(colours = heatmap_cols) + 
  theme(text = element_text(size = 15), 
        axis.text.y = element_text(size = 7))
Heatmap
```

## Koenig Fibroblasts DE DCM vs HC

```{r}
fb_K <- subset(
  Koenig, 
  Names == "Fibroblasts"
)

table(fb_K$Names, fb_K$Condition)
```

```{r}
fb_K@meta.data <- fb_K@meta.data %>%
  mutate(Group = case_when(
    Condition == "Donor" ~ "HC",
    Condition == "DCM" ~ "DCM"
  ))
```

```{r, fig.width=5, fig.height=4}
DimPlot(fb_K, group.by = c("Group"), reduction = "umap", label = F, shuffle = TRUE, pt.size = 0.1) 
```

Prepare matadata for differential expression testing

```{r}
fb_K_metadata <- fb_K@meta.data %>%
  dplyr::select(orig.ident, Group) %>%
  unique()

rownames(fb_K_metadata) <- fb_K_metadata$orig.ident

reactable(fb_K_metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

Remove genes expressed in <1% of cells

```{r}
expr_matrix <- GetAssayData(object = fb_K, assay = "RNA", slot = "counts")

# Calculate the number of cells where each gene is expressed (non-zero counts)
non_zero_cells <- rowSums(expr_matrix > 0)

# Calculate the total number of cells
total_cells <- ncol(expr_matrix)

# Calculate the percentage of cells expressing each gene
percent_cells <- (non_zero_cells / total_cells) * 100

# Create a dataframe to store the results
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- fb_K[which(rownames(fb_K) %in% keep_genes),]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

Create and prepare pseudobulk object for differential expression analysis

```{r}
DefaultAssay(data) <- "RNA"

pseudocounts <- Seurat2PB(data, sample="orig.ident", cluster = "Names")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(fb_K_metadata$orig.ident, pseudocounts$samples$sample)]
all(fb_K_metadata$orig.ident == pseudocounts$samples$sample)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

DE analysis

```{r}
fb_K_metadata$Group <- factor(fb_K_metadata$Group, levels = c("HC", "DCM"))
design<- model.matrix(~ Group, data = fb_K_metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_fb_K <- topTags(fit, n = Inf)$table
results_fb_K <- merge(results_fb_K, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_fb_K$up_down <- ifelse(
  results_fb_K$logFC > 0.5 & results_fb_K$FDR < 0.05, "Increased",
  ifelse(results_fb_K$logFC < -0.5 & results_fb_K$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )

signif_fb_K <- results_fb_K %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

write.csv(results_fb_K, file = here::here(output_dir_data, "fb_K_DE_Results.csv"), quote=F, row.names = F)
write.csv(signif_fb_K, file = here::here(output_dir_data, "fb_K_DE_Significant.csv"), quote=F, row.names = F)
```

```{r, fig.height=6, fig.width=4}
keyvals <- rep("grey", nrow(results_fb_K))
names(keyvals) <- rep("Not significant", nrow(results_fb_K))

keyvals[results_fb_K$up_down == "Increased"] <- "#d62728"
names(keyvals)[keyvals == "#d62728"] <- "Increased"
keyvals[results_fb_K$up_down == "Decreased"] <- "#1f77b4"
names(keyvals)[keyvals == "#1f77b4"] <- "Decreased"

volcano <- EnhancedVolcano(results_fb_K,
                lab = results_fb_K$gene,
                x = "logFC",
                y = "FDR",
                labSize = 0,
                legendLabSize = 8.5,
                titleLabSize = 15,
                subtitleLabSize = 14,
                axisLabSize = 12,
                captionLabSize = 9,
                pointSize = 0.5,
                FCcutoff = 0.5,
                pCutoff  = 0.05,
                ylim = c(0, 9),
                colAlpha = 1,
                drawConnectors = FALSE,
                subtitle = NULL,
                title = "LV Fibroblasts\nDCM (n=13) vs HC (n=25)",
                colCustom = keyvals
) + theme(legend.position = "bottom")

volcano
```

```{r, fig.width=10, fig.height=5}
signif_fb_K$metric <- signif_fb_K$logFC*-log10(signif_fb_K$PValue)

DEGs_top_up <- signif_fb_K %>% 
  dplyr::arrange(desc(metric)) %>%
  slice_head(n=10)
DEGs_top_down <- signif_fb_K %>% 
  dplyr::arrange(metric) %>%
  slice_head(n=10)

DEGs_top <- dplyr::bind_rows(DEGs_top_up, DEGs_top_down)
rm(DEGs_top_up, DEGs_top_down)

counts <- cpm(pseudocounts, log = TRUE)

fb_K_metadata <- fb_K_metadata %>%
  dplyr::arrange(Group)
counts_heatmap <- counts[DEGs_top$gene, ]
counts_heatmap <- counts_heatmap[, fb_K_metadata$orig.ident, drop = FALSE]

# 1) Build the annotation data.frame aligned to your matrix columns
ann_col <- fb_K_metadata %>%
  dplyr::filter(orig.ident %in% colnames(counts)) %>%                            # if you have a orig.ident key
  dplyr::arrange(match(orig.ident, colnames(counts))) %>%
  dplyr::select(Group) %>%
  as.data.frame()
rownames(ann_col) <- colnames(counts)

# (optional) lock desired order
ann_col$Group <- factor(ann_col$Group, levels = c("HC","DCM"))

# 2) Make a named color vector that matches the present levels
lvl <- levels(ann_col$Group)
# if you specifically want HC/DCM colors:
palette_group <- c(HC = "#1f77b4", DCM = "#d62728")
ann_colors <- list(Group = setNames(palette_group[lvl], lvl))
# If levels are unknown beforehand, use this instead:
# vals <- unique(as.character(ann_col$Group))
# ann_colors <- list(Group = setNames(c("#1f77b4","#d62728","#2ca02c","#9467bd")[seq_along(vals)], vals))

# 3) Heatmap colors
heatmap_cols <- colorRampPalette(c("#004C8C", "white", "#A63A2A"))(101)

# 4) Plot
pheatmap::pheatmap(
  counts_heatmap,
  name = " ",
  angle_col = 45,                 # numeric is fine
  color = heatmap_cols,
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,
  show_rownames = TRUE,
  annotation_legend = TRUE,
  border_color = NA,
  fontsize = 10,
  fontsize_row = 13,
  fontsize_col = 6,
  cellheight = 15,
  gaps_col = c(0,0,0,0)
)

```
