---
title: "lncRNAs in sn/scRNA-seq dataset of human cardiac tissue"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "CM_lncRNAs"

# Load libraries
library(here)
library(readr)
library(readxl)
library(xlsx)
library(Seurat)
library(Matrix)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(org.Hs.eg.db)
library(AnnotationHub)
library(ReactomePA)
library(statmod)
library(SeuratDisk)
library(speckle)
library(rtracklayer)
library(png)
library(reactable)
library(UpSetR)
library(msigdbr)
library(simplifyEnrichment)
library(biomaRt)
library(igraph)
library(gridExtra) 
library(ggpubr)
library(edgeR)

#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```

```{r reference_genes}
reference_genes <- readGFF(here::here("data", "refdata-gex-GRCh38-2020-A", "genes", "genes.gtf"))
```

```{r lncrna}
lncrna_ens <- c("ENSG00000254002", "ENSG00000228692", "ENSG00000226900", "ENSG00000215246", "ENSG00000253807")
lncrna_ref <- reference_genes %>%
  dplyr::filter(gene_id %in% lncrna_ens) %>%
  dplyr::select(gene_id, gene_name) %>%
  unique() %>%
  dplyr::mutate(gene_id = factor(gene_id, levels = lncrna_ens)) %>%
  dplyr::arrange(gene_id)
lncrna_symbol <- lncrna_ref$gene_name
lncrna_symbol
```


## Litvinukova et al. 2020

```{r import Litvinukova}
Convert("/scratch/gbonaz/Heart_datasets/Litvinukova_2020/global_raw_v1.h5ad", 
        dest = "h5seurat", overwrite = TRUE )
Litvinukova <- LoadH5Seurat("/scratch/gbonaz/Heart_datasets/Litvinukova_2020/global_raw_v1.h5seurat")
DefaultAssay(Litvinukova) <- "RNA"
Litvinukova <- NormalizeData(Litvinukova)
```

```{r Litvinukova_dimplot, fig.width=8, fig.height=5}
Litvinukova_dimplot <- DimPlot(Litvinukova, group.by = "cell_type") + ggtitle("Litvinukova et al. 2020")
Litvinukova_dimplot
```

```{r genes_Litvinukova}
genes_Litvinukova <- data.frame(gene = rownames(Litvinukova))
lncrna_symbol %in% genes_Litvinukova$gene
```

```{r Litvinukova_dotplot, fig.width=6, fig.height=5}
Litvinukova_dotplot <- DotPlot(Litvinukova, features=lncrna_symbol, group.by="cell_type") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +  
  scale_y_discrete(limits = rev) +
  ggtitle("Litvinukova et al. 2020")
Litvinukova_dotplot
```

```{r Litvinukova_vlnplot, fig.width=6, fig.height=5}
Idents(Litvinukova) <- "cell_type"
Litvinukova_vlnplot <- Stacked_VlnPlot(Litvinukova, features = lncrna_symbol, x_lab_rotate = TRUE, pt.size = 0.001, plot_spacing = 0.05)
Litvinukova_vlnplot[[1]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Litvinukova_vlnplot[[2]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Litvinukova_vlnplot[[3]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Litvinukova_vlnplot[[4]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Litvinukova_vlnplot[[5]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Litvinukova_vlnplot <- wrap_plots(Litvinukova_vlnplot) + 
  plot_annotation(title = "Litvinukova et al. 2020") & 
  theme(plot.title = element_text(hjust = 0.5,      
                                  face = "bold",     
                                  size = 15)
  )
print(Litvinukova_vlnplot)
```

```{r Litvinukova_featureplot, fig.width=9, fig.height=6}
Litvinukova_featureplot <- FeaturePlot_scCustom(Litvinukova, features = lncrna_symbol, num_columns = 3)
Litvinukova_featureplot <- wrap_plots(Litvinukova_featureplot) + 
  plot_annotation(title = "Litvinukova et al. 2020") & 
  theme(plot.title = element_text(hjust = 0.5,      
                                  face = "bold",     
                                  size = 20)
  )
print(Litvinukova_featureplot)
```

```{r subset Litvinukova FindMarkers}
Litvinukova_subset_V <- subset(Litvinukova,  cell_type == "Ventricular_Cardiomyocyte" & region %in% c("RV", "LV"))
Litvinukova_subset_A <- subset(Litvinukova, cell_type == "Atrial_Cardiomyocyte" & region %in% c("RA", "LA"))

Idents(Litvinukova_subset_V) <- Litvinukova_subset_V$region
Idents(Litvinukova_subset_A) <- Litvinukova_subset_A$region

Litvinukova_RV_vs_LV <- FindMarkers(Litvinukova_subset_V, ident.1 = "RV", ident.2 = "LV", features = lncrna_symbol, logfc.threshold = 0, min.pct = 0.001)

Litvinukova_RA_vs_LA <- FindMarkers(Litvinukova_subset_A, ident.1 = "RA", ident.2 = "LA", features = lncrna_symbol, logfc.threshold = 0, min.pct = 0.001)
```

```{r Litvinukova_RV_vs_LV }
reactable(Litvinukova_RV_vs_LV, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Litvinukova_RA_vs_LA}
reactable(Litvinukova_RA_vs_LA, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Litvinukova_vlnplot_RV_vs_LV, fig.width=10, fig.height=3}
Litvinukova_vlnplot_RV_vs_LV <- VlnPlot(Litvinukova_subset_V, features = lncrna_symbol, group.by = "region", ncol = 5, pt.size = 0.0001)
Litvinukova_vlnplot_RV_vs_LV[[1]]$layers[[2]]$aes_params$alpha <- 0.2
Litvinukova_vlnplot_RV_vs_LV[[2]]$layers[[2]]$aes_params$alpha <- 0.2
Litvinukova_vlnplot_RV_vs_LV[[3]]$layers[[2]]$aes_params$alpha <- 0.2
Litvinukova_vlnplot_RV_vs_LV[[4]]$layers[[2]]$aes_params$alpha <- 0.2
Litvinukova_vlnplot_RV_vs_LV[[5]]$layers[[2]]$aes_params$alpha <- 0.2
Litvinukova_vlnplot_RV_vs_LV
```

```{r Litvinukova_vlnplot_RA_vs_LA, fig.width=10, fig.height=3}
Litvinukova_vlnplot_RA_vs_LA <- VlnPlot(Litvinukova_subset_V, features = lncrna_symbol, group.by = "region", ncol = 5, pt.size = 0.0001)
Litvinukova_vlnplot_RA_vs_LA[[1]]$layers[[2]]$aes_params$alpha <- 0.2
Litvinukova_vlnplot_RA_vs_LA[[2]]$layers[[2]]$aes_params$alpha <- 0.2
Litvinukova_vlnplot_RA_vs_LA[[3]]$layers[[2]]$aes_params$alpha <- 0.2
Litvinukova_vlnplot_RA_vs_LA[[4]]$layers[[2]]$aes_params$alpha <- 0.2
Litvinukova_vlnplot_RA_vs_LA[[5]]$layers[[2]]$aes_params$alpha <- 0.2
Litvinukova_vlnplot_RA_vs_LA
```

```{r Litvinukova_dotplot_RV_vs_LV, fig.width=6, fig.height=4}
Litvinukova_dotplot_RV_vs_LV <- DotPlot(Litvinukova_subset_V, features=lncrna_symbol, group.by="region", scale=F) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) 
  ggtitle("Litvinukova et al. 2020")
Litvinukova_dotplot_RV_vs_LV
```

```{r Litvinukova_dotplot_RA_vs_LA, fig.width=6, fig.height=4}
Litvinukova_dotplot_RA_vs_LA  <- DotPlot(Litvinukova_subset_A, features=lncrna_symbol, group.by="region", scale=F) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) 
  ggtitle("Litvinukova et al. 2020")
Litvinukova_dotplot_RA_vs_LA 
```
```{r rm Litvinukova}
rm(Litvinukova)
```


## Koenig et al. 2022

```{r import Koenig}
load("/scratch/gbonaz/Heart_datasets/Koenig_2022/GSE183852_DCM_Integrated.Robj")
Koenig <- RefMerge
rm(RefMerge)
DefaultAssay(Koenig) <- "RNA"
Koenig <- NormalizeData(Koenig)
```

```{r Koenig_dimplot, fig.width=6, fig.height=5}
Koenig_dimplot <- DimPlot(Koenig, group.by = "Names") + ggtitle("Koenig et al. 2022")
Koenig_dimplot
```

```{r genes_Koenig}
genes_Koenig <- data.frame(gene = rownames(Koenig))
lncrna_symbol %in% genes_Koenig$gene
```

```{r Koenig_dotplot, fig.width=6, fig.height=5}
Koenig_dotplot <- DotPlot(Koenig, features=lncrna_symbol, group.by="Names") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +  
  scale_y_discrete(limits = rev) +
  ggtitle("Koenig et al. 2022")
Koenig_dotplot
```

```{r Koenig_vlnplot, fig.width=6, fig.height=5}
Koenig_vlnplot <- Stacked_VlnPlot(Koenig, features = lncrna_symbol, x_lab_rotate = TRUE, pt.size = 0.001, plot_spacing = 0.05)
Koenig_vlnplot[[1]][[1]]$layers[[2]]$aes_params$alpha <- 0.2
Koenig_vlnplot[[2]][[1]]$layers[[2]]$aes_params$alpha <- 0.2
Koenig_vlnplot[[3]][[1]]$layers[[2]]$aes_params$alpha <- 0.2
Koenig_vlnplot[[4]][[1]]$layers[[2]]$aes_params$alpha <- 0.2
Koenig_vlnplot[[5]][[1]]$layers[[2]]$aes_params$alpha <- 0.2
Koenig_vlnplot <- wrap_plots(Koenig_vlnplot) + 
  plot_annotation(title = "Koenig et al. 2022") & 
  theme(plot.title = element_text(hjust = 0.5,      
                                  face = "bold",     
                                  size = 15)
  )
print(Koenig_vlnplot)
```

```{r Koenig_featureplot, fig.width=9, fig.height=6}
Koenig_featureplot <- FeaturePlot_scCustom(Koenig, features = lncrna_symbol, num_columns = 3)
Koenig_featureplot <- wrap_plots(Koenig_featureplot) + 
  plot_annotation(title = "Koenig et al. 2022") & 
  theme(plot.title = element_text(hjust = 0.5,      
                                  face = "bold",     
                                  size = 20)
  )
print(Koenig_featureplot)
```

```{r Koenig metadata}
Koenig_CM <- subset(Koenig, Names %in% c("Cardiomyocytes") & CondTech %in% c("DCM_SN", "Donor_SN"))

metadata <- Koenig_CM@meta.data %>%
  dplyr::select(orig.ident, condition) %>%
  unique()
rownames(metadata) <- metadata$orig.ident
reactable(metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

Remove genes expressed in <1% of cells

```{r Koenig_pseudocounts, eval=TRUE}
expr_matrix <- GetAssayData(object = Koenig_CM, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- Koenig_CM[which(rownames(Koenig_CM) %in% keep_genes),]
rm(keep_genes, non_zero_cells, expr_matrix, total_cells, percent_cells)
pseudocounts <- Seurat2PB(data, sample="orig.ident", cluster = "Names")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(metadata$orig.ident, pseudocounts$samples$sample)]
all(metadata$donor_id == pseudocounts$samples$sample)
keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

```{r Koenig_DE}
metadata$condition <- factor(metadata$condition, levels = c("Donor", "DCM"))
design<- model.matrix(~ condition, data = metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_Koenig <- topTags(fit, n = Inf)$table
results_Koenig <- merge(results_Koenig, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_Koenig$up_down <- ifelse(
  results_Koenig$logFC > 0.5 & results_Koenig$FDR < 0.05, "Increased",
  ifelse(results_Koenig$logFC < -0.5 & results_Koenig$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )
results_Koenig_lncRNA <- results_Koenig %>%
  dplyr::filter(gene %in% lncrna_symbol)
reactable(results_Koenig_lncRNA, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Koenig_heatmap_DCM_vs_HC, fig.width=6, fig.height=1}
counts <- cpm(pseudocounts, log = TRUE)

counts <- counts[rownames(counts) %in% results_Koenig_lncRNA$gene, , drop = FALSE]

counts <- counts[, dplyr::arrange(metadata, condition)$orig.ident, drop = FALSE]

heat_colors <- rev(brewer.pal(11, "PuOr"))
pheatmap::pheatmap(counts, 
         color = heat_colors, 
         cluster_rows = F, 
         annotation_col = dplyr::select(metadata, condition), 
         border_color = NA, 
         fontsize = 10, 
         scale = "row",
         cluster_cols=F,
         fontsize_row = 15, 
         show_rownames = T,
         show_colnames = F,
         cellheight = 15,
         legend_breaks = c(-1.5, 0, 1.5)
         )
```

```{r Koenig_vlnplot_DE, fig.width=2.5, fig.height=4}
Koenig_dotplot_DE <- VlnPlot(Koenig_CM, features=results_Koenig_lncRNA$gene, group.by="condition") + 
  ggtitle("Koenig et al. 2022") + NoLegend()
Koenig_dotplot_DE
```

```{r rm Koenig}
rm(Koenig)
```

## Chaffin et al. 2022

```{r import Chaffin}
Chaffin <- LoadH5Seurat("/scratch/gbonaz/Heart_datasets/Chaffin_2022/human_dcm_hcm_scportal_03.17.2022.h5seurat")
Chaffin@assays$cellranger_raw <- NULL
saveRDS(Chaffin, "/scratch/gbonaz/Heart_datasets/Chaffin_2022/Chaffin_2022.rds")
Chaffin <- readRDS("/scratch/gbonaz/Heart_datasets/Chaffin_2022/Chaffin_2022.rds")
DefaultAssay(Chaffin) <- "RNA"
Chaffin <- NormalizeData(Chaffin)
```

```{r Chaffin_dimplot, fig.width=8, fig.height=5}
Chaffin_dimplot <- DimPlot(Chaffin, group.by = "cell_type_leiden0.6") + ggtitle("Chaffin et al. 2022")
Chaffin_dimplot
```

```{r genes_Chaffin}
genes_Chaffin <- data.frame(gene = rownames(Chaffin))
lncrna_symbol %in% genes_Chaffin$gene
```

```{r Chaffin_dotplot, fig.width=6, fig.height=5}
Chaffin_dotplot <- DotPlot(Chaffin, features=lncrna_symbol, group.by="cell_type_leiden0.6") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +  
  scale_y_discrete(limits = rev) +
  ggtitle("Chaffin et al. 2022")
Chaffin_dotplot
```

```{r Chaffin_vlnplot, fig.width=6, fig.height=5}
Idents(Chaffin) <- "cell_type_leiden0.6"
Chaffin_vlnplot <- Stacked_VlnPlot(Chaffin, features = lncrna_symbol, x_lab_rotate = TRUE, pt.size = 0.001, plot_spacing = 0.05)
Chaffin_vlnplot[[1]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Chaffin_vlnplot[[2]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Chaffin_vlnplot[[3]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Chaffin_vlnplot[[4]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Chaffin_vlnplot[[5]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Chaffin_vlnplot <- wrap_plots(Chaffin_vlnplot) + 
  plot_annotation(title = "Chaffin et al. 2022") & 
  theme(plot.title = element_text(hjust = 0.5,      
                                  face = "bold",     
                                  size = 15)
  )
print(Chaffin_vlnplot)
```

```{r Chaffin_featureplot, fig.width=9, fig.height=6}
Chaffin_featureplot <- FeaturePlot_scCustom(Chaffin, features = lncrna_symbol, num_columns = 3)
Chaffin_featureplot <- wrap_plots(Chaffin_featureplot) + 
  plot_annotation(title = "Chaffin et al. 2022") & 
  theme(plot.title = element_text(hjust = 0.5,      
                                  face = "bold",     
                                  size = 20)
  )
print(Chaffin_featureplot)
```

```{r Chaffin metadata}
Chaffin_CM <- subset(Chaffin, cell_type_leiden0.6 %in% c("Cardiomyocyte_I", "Cardiomyocyte_II", "Cardiomyocyte_III"))
Chaffin_CM$cell_type <- "Cardiomyocytes"
metadata <- Chaffin_CM@meta.data %>%
  dplyr::select(donor_id, disease) %>%
  unique()
rownames(metadata) <- metadata$donor_id
reactable(metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

Remove genes expressed in <1% of cells

```{r Chaffin_pseudocounts, eval=TRUE}
expr_matrix <- GetAssayData(object = Chaffin_CM, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- Chaffin_CM[which(rownames(Chaffin_CM) %in% keep_genes),]
rm(keep_genes, non_zero_cells, expr_matrix, total_cells, percent_cells)
pseudocounts <- Seurat2PB(data, sample="donor_id", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(metadata$donor_id, pseudocounts$samples$sample)]
all(metadata$donor_id == pseudocounts$samples$sample)
keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

```{r Chaffin_DE}
metadata$disease <- factor(metadata$disease, levels = c("NF", "DCM", "HCM"))
design<- model.matrix(~ disease, data = metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)

contrast_DCM_vs_NF <- makeContrasts(diseaseDCM, levels=design)
results_Chaffin_DCM_vs_NF <- glmQLFTest(fit, contrast=contrast_DCM_vs_NF)
results_Chaffin_DCM_vs_NF <- topTags(results_Chaffin_DCM_vs_NF, n = Inf)$table
results_Chaffin_DCM_vs_NF <- merge(results_Chaffin_DCM_vs_NF, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_Chaffin_DCM_vs_NF$up_down <- ifelse(
  results_Chaffin_DCM_vs_NF$logFC > 0.5 & results_Chaffin_DCM_vs_NF$FDR < 0.05, "Increased",
  ifelse(results_Chaffin_DCM_vs_NF$logFC < -0.5 & results_Chaffin_DCM_vs_NF$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )
results_Chaffin_DCM_vs_NF_lncRNA <- results_Chaffin_DCM_vs_NF %>%
  dplyr::filter(gene %in% lncrna_symbol)
results_Chaffin_DCM_vs_NF_lncRNA$Contrast <- "DCM_vs_NF"


contrast_HCM_vs_NF <- makeContrasts(diseaseHCM, levels=design)
results_Chaffin_HCM_vs_NF <- glmQLFTest(fit, contrast=contrast_HCM_vs_NF)
results_Chaffin_HCM_vs_NF <- topTags(results_Chaffin_HCM_vs_NF, n = Inf)$table
results_Chaffin_HCM_vs_NF <- merge(results_Chaffin_HCM_vs_NF, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_Chaffin_HCM_vs_NF$up_down <- ifelse(
  results_Chaffin_HCM_vs_NF$logFC > 0.5 & results_Chaffin_HCM_vs_NF$FDR < 0.05, "Increased",
  ifelse(results_Chaffin_HCM_vs_NF$logFC < -0.5 & results_Chaffin_HCM_vs_NF$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )
results_Chaffin_HCM_vs_NF_lncRNA <- results_Chaffin_HCM_vs_NF %>%
  dplyr::filter(gene %in% lncrna_symbol)
results_Chaffin_HCM_vs_NF_lncRNA$Contrast <- "HCM_vs_NF"


contrast_DCM_vs_HCM <- makeContrasts(diseaseDCM - diseaseHCM, levels=design)
results_Chaffin_DCM_vs_HCM <- glmQLFTest(fit, contrast=contrast_DCM_vs_HCM)
results_Chaffin_DCM_vs_HCM <- topTags(results_Chaffin_DCM_vs_HCM, n = Inf)$table
results_Chaffin_DCM_vs_HCM <- merge(results_Chaffin_DCM_vs_HCM, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_Chaffin_DCM_vs_HCM$up_down <- ifelse(
  results_Chaffin_DCM_vs_HCM$logFC > 0.5 & results_Chaffin_DCM_vs_HCM$FDR < 0.05, "Increased",
  ifelse(results_Chaffin_DCM_vs_HCM$logFC < -0.5 & results_Chaffin_DCM_vs_HCM$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )
results_Chaffin_DCM_vs_HCM_lncRNA <- results_Chaffin_DCM_vs_HCM %>%
  dplyr::filter(gene %in% lncrna_symbol)
results_Chaffin_DCM_vs_HCM_lncRNA$Contrast <- "DCM_vs_HCM"


reactable(results_Chaffin_DCM_vs_NF_lncRNA, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

reactable(results_Chaffin_HCM_vs_NF_lncRNA, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

reactable(results_Chaffin_DCM_vs_HCM_lncRNA, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Chaffin_heatmap_DCM_vs_HC, fig.width=6, fig.height=2}
counts <- cpm(pseudocounts, log = TRUE)

counts <- counts[rownames(counts) %in% results_Chaffin_DCM_vs_NF_lncRNA$gene, , drop = FALSE]

metadata_matched <- metadata[match(colnames(counts), metadata$donor_id), ]

# Step 2: Check they align
stopifnot(all(colnames(counts) == metadata_matched$donor_id))

# Step 3: Reorder both by disease
order_idx <- order(metadata_matched$disease)

counts <- counts[, order_idx]
metadata_matched <- metadata_matched[order_idx, ]

# Step 4: Heatmap with matched annotations
heat_colors <- rev(brewer.pal(11, "PuOr"))
pheatmap::pheatmap(
  counts,
  color = heat_colors,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = dplyr::select(metadata_matched, disease),
  border_color = NA,
  fontsize = 10,
  scale = "row",
  fontsize_row = 15,
  show_rownames = TRUE,
  show_colnames = FALSE,
  cellheight = 15,
  legend_breaks = c(-1.5, 0, 1.5)
)
```

```{r Chaffin_vlnplot_DE, fig.width=10, fig.height=4}
Chaffin_dotplot_DE <- VlnPlot(Chaffin_CM, features=results_Chaffin_DCM_vs_HCM_lncRNA$gene, group.by="disease", ncol = 4, pt.size = 0.001) + 
  ggtitle("Chaffin et al. 2022") + NoLegend()
Chaffin_dotplot_DE
```

```{r rm Chaffin}
rm(Chaffin)
```

## Reichart et al. 2022

```{r import Reichart}
Reichart <- readRDS("/scratch/gbonaz/Heart_datasets/Reichart_2022/DCM_ACM_heart_cell_atlas.rds")
DefaultAssay(Reichart) <- "RNA"
Reichart <- NormalizeData(Reichart)
```

```{r Reichart_dimplot, fig.width=6, fig.height=5}
Reichart_dimplot <- DimPlot(Reichart, group.by = "cell_type") + ggtitle("Reichart et al. 2022")
Reichart_dimplot
```

```{r genes_Reichart}
genes_Reichart <- data.frame(gene = rownames(Reichart))
lncrna_symbol %in% genes_Reichart$gene
lncrna_ens %in% genes_Reichart$gene
```

```{r Reichart_dotplot, fig.width=6, fig.height=5}
Reichart_dotplot <- DotPlot(Reichart, features=lncrna_ens, group.by="cell_type") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +  
  scale_y_discrete(limits = rev) +
  ggtitle("Reichart et al. 2022")
Reichart_dotplot
```

```{r Reichart_vlnplot, fig.width=6, fig.height=5}
Idents(Reichart) <- "cell_type"
Reichart_vlnplot <- Stacked_VlnPlot(Reichart, features = lncrna_ens, x_lab_rotate = TRUE, pt.size = 0.001, plot_spacing = 0.05)
Reichart_vlnplot[[1]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Reichart_vlnplot[[2]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Reichart_vlnplot[[3]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Reichart_vlnplot[[4]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Reichart_vlnplot[[5]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Reichart_vlnplot <- wrap_plots(Reichart_vlnplot) + 
  plot_annotation(title = "Reichart et al. 2022") & 
  theme(plot.title = element_text(hjust = 0.5,      
                                  face = "bold",     
                                  size = 15)
  )
print(Reichart_vlnplot)
```

```{r Reichart_featureplot, fig.width=9, fig.height=6}
Reichart_featureplot <- FeaturePlot_scCustom(Reichart, features = lncrna_ens, num_columns = 3)
Reichart_featureplot <- wrap_plots(Reichart_featureplot) + 
  plot_annotation(title = "Reichart et al. 2022") & 
  theme(plot.title = element_text(hjust = 0.5,      
                                  face = "bold",     
                                  size = 20)
  )
print(Reichart_featureplot)
```

```{r subset Reichart FindMarkers}
Reichart_subset_V <- subset(Reichart,  cell_type == "cardiac muscle cell" & Region_x %in% c("RV", "LV"))

Idents(Reichart_subset_V) <- Reichart_subset_V$Region_x

Reichart_RV_vs_LV <- FindMarkers(Reichart_subset_V, ident.1 = "RV", ident.2 = "LV", features = lncrna_ens, logfc.threshold = 0, min.pct = 0.001)
```

```{r Reichart_RV_vs_LV }
reactable(Reichart_RV_vs_LV, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Reichart_vlnplot_RV_vs_LV, fig.width=12, fig.height=3}
Reichart_vlnplot_RV_vs_LV <- VlnPlot(Reichart_subset_V, features = lncrna_ens, group.by = "Region_x", ncol = 5, pt.size = 0.0001)
Reichart_vlnplot_RV_vs_LV[[1]]$layers[[2]]$aes_params$alpha <- 0.2
Reichart_vlnplot_RV_vs_LV[[2]]$layers[[2]]$aes_params$alpha <- 0.2
Reichart_vlnplot_RV_vs_LV[[3]]$layers[[2]]$aes_params$alpha <- 0.2
Reichart_vlnplot_RV_vs_LV[[4]]$layers[[2]]$aes_params$alpha <- 0.2
Reichart_vlnplot_RV_vs_LV[[5]]$layers[[2]]$aes_params$alpha <- 0.2
Reichart_vlnplot_RV_vs_LV
```

```{r Reichart_dotplot_RV_vs_LV, fig.width=6.5, fig.height=4}
Reichart_dotplot_RV_vs_LV <- DotPlot(Reichart_subset_V, features=lncrna_ens, group.by="Region_x", scale=F) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0.5), "inch")) 
  ggtitle("Reichart et al. 2020")
Reichart_dotplot_RV_vs_LV
```

```{r Reichart metadata RV}
Reichart_CM_RV <- subset(Reichart, cell_type %in% c("cardiac muscle cell") & disease %in% c("normal", "dilated cardiomyopathy") & Region_x == "RV")
metadata_RV <- Reichart_CM_RV@meta.data %>%
  dplyr::select(donor_id, disease) %>%
  unique()
rownames(metadata_RV) <- metadata_RV$donor_id
reactable(metadata_RV, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Reichart metadata LV}
Reichart_CM_LV <- subset(Reichart, cell_type %in% c("cardiac muscle cell") & disease %in% c("normal", "dilated cardiomyopathy") & Region_x == "LV")
metadata_LV <- Reichart_CM_LV@meta.data %>%
  dplyr::select(donor_id, disease) %>%
  unique()
rownames(metadata_LV) <- metadata_LV$donor_id
reactable(metadata_LV, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Reichart_pseudocounts_RV}
expr_matrix <- GetAssayData(object = Reichart_CM_RV, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- Reichart_CM_RV[which(rownames(Reichart_CM_RV) %in% keep_genes),]
rm(keep_genes, non_zero_cells, expr_matrix, total_cells, percent_cells)
pseudocounts <- Seurat2PB(data, sample="donor_id", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(metadata_RV$donor_id, pseudocounts$samples$sample)]
all(metadata$donor_id == pseudocounts$samples$sample)
keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

```{r Reichart_DE_RV}
metadata_RV$disease <- factor(metadata_RV$disease, levels = c("normal", "dilated cardiomyopathy"))
design<- model.matrix(~ disease, data = metadata_RV)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_Reichart_RV <- topTags(fit, n = Inf)$table
results_Reichart_RV <- merge(results_Reichart_RV, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_Reichart_RV$up_down <- ifelse(
  results_Reichart_RV$logFC > 0.5 & results_Reichart_RV$FDR < 0.05, "Increased",
  ifelse(results_Reichart_RV$logFC < -0.5 & results_Reichart_RV$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )
results_Reichart_RV_lncRNA <- results_Reichart_RV %>%
  dplyr::filter(gene %in% lncrna_ens)
reactable(results_Reichart_RV_lncRNA, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Reichart_heatmap_DCM_vs_HC_RV, fig.width=8, fig.height=2}
counts <- cpm(pseudocounts, log = TRUE)

counts <- counts[rownames(counts) %in% results_Reichart_RV_lncRNA$gene, , drop = FALSE]

metadata_matched <- metadata_RV[match(colnames(counts), metadata_RV$donor_id), ]

stopifnot(all(colnames(counts) == metadata_matched$donor_id))

order_idx <- order(metadata_matched$disease)

counts <- counts[, order_idx]
metadata_matched <- metadata_matched[order_idx, ]

heat_colors <- rev(brewer.pal(11, "PuOr"))
pheatmap::pheatmap(counts, 
         color = heat_colors, 
         cluster_rows = F, 
         annotation_col = dplyr::select(metadata_matched, disease), 
         border_color = NA, 
         fontsize = 10, 
         scale = "row",
         cluster_cols=F,
         fontsize_row = 15, 
         show_rownames = T,
         show_colnames = F,
         cellheight = 15,
         legend_breaks = c(-1.5, 0, 1.5)
         )
```

```{r Reichart_pseudocounts_LV}
expr_matrix <- GetAssayData(object = Reichart_CM_LV, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- Reichart_CM_LV[which(rownames(Reichart_CM_LV) %in% keep_genes),]
rm(keep_genes, non_zero_cells, expr_matrix, total_cells, percent_cells)
pseudocounts <- Seurat2PB(data, sample="donor_id", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(metadata_LV$donor_id, pseudocounts$samples$sample)]
all(metadata$donor_id == pseudocounts$samples$sample)
keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

```{r Reichart_DE_LV}
metadata_LV$disease <- factor(metadata_LV$disease, levels = c("normal", "dilated cardiomyopathy"))
design<- model.matrix(~ disease, data = metadata_LV)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_Reichart_LV <- topTags(fit, n = Inf)$table
results_Reichart_LV <- merge(results_Reichart_LV, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_Reichart_LV$up_down <- ifelse(
  results_Reichart_LV$logFC > 0.5 & results_Reichart_LV$FDR < 0.05, "Increased",
  ifelse(results_Reichart_LV$logFC < -0.5 & results_Reichart_LV$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )
results_Reichart_LV_lncRNA <- results_Reichart_LV %>%
  dplyr::filter(gene %in% lncrna_ens)
reactable(results_Reichart_LV_lncRNA, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Reichart_heatmap_DCM_vs_HC_LV, fig.width=8, fig.height=2}
counts <- cpm(pseudocounts, log = TRUE)

counts <- counts[rownames(counts) %in% results_Reichart_LV_lncRNA$gene, , drop = FALSE]

metadata_matched <- metadata_LV[match(colnames(counts), metadata_LV$donor_id), ]

stopifnot(all(colnames(counts) == metadata_matched$donor_id))

order_idx <- order(metadata_matched$disease)

counts <- counts[, order_idx]
metadata_matched <- metadata_matched[order_idx, ]

heat_colors <- rev(brewer.pal(11, "PuOr"))
pheatmap::pheatmap(counts, 
         color = heat_colors, 
         cluster_rows = F, 
         annotation_col = dplyr::select(metadata_matched, disease), 
         border_color = NA, 
         fontsize = 10, 
         scale = "row",
         cluster_cols=F,
         fontsize_row = 15, 
         show_rownames = T,
         show_colnames = F,
         cellheight = 15,
         legend_breaks = c(-1.5, 0, 1.5)
         )
```

```{r Reichart_vlnplot_DE_RV, fig.width=10, fig.height=6}
Reichart_dotplot_DE <- VlnPlot(Reichart_CM_RV, features=results_Reichart_RV_lncRNA$gene, group.by="disease", ncol = 4) + 
  ggtitle("Reichart et al. 2022") + NoLegend()
Reichart_dotplot_DE
```

```{r Reichart_vlnplot_DE_LV, fig.width=10, fig.height=6}
Reichart_dotplot_DE <- VlnPlot(Reichart_CM_LV, features=results_Reichart_LV_lncRNA$gene, group.by="disease", ncol = 4) + 
  ggtitle("Reichart et al. 2022") + NoLegend()
Reichart_dotplot_DE
```

```{r rm Reichart}
rm(Reichart)
```

## DCM RV

```{r import DCM_RV}
DCM_RV <- readRDS(here::here("output", "QC_integration_annotation_26", "DCM_RV_integrated.rds"))
DCM_RV@assays$RAW <- NULL
DCM_RV@assays$SCT <- NULL
DefaultAssay(DCM_RV) <- "RNA"
DCM_RV <- NormalizeData(DCM_RV)
```

```{r DCM_RV_dimplot, fig.width=6, fig.height=5}
DCM_RV_dimplot <- DimPlot(DCM_RV, group.by = "cell_type") + ggtitle("DCM RV")
DCM_RV_dimplot
```

```{r genes_DCM_RV}
genes_DCM_RV <- data.frame(gene = rownames(DCM_RV))
lncrna_symbol %in% genes_DCM_RV$gene
```

```{r DCM_RV_dotplot, fig.width=6, fig.height=5}
DCM_RV_dotplot <- DotPlot(DCM_RV, features=lncrna_symbol, group.by="cell_type") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +  
  scale_y_discrete(limits = rev) +
  ggtitle("DCM RV")
DCM_RV_dotplot
```

```{r DCM_RV_vlnplot, fig.width=6, fig.height=5}
DCM_RV_vlnplot <- Stacked_VlnPlot(DCM_RV, features = lncrna_symbol, x_lab_rotate = TRUE, pt.size = 0.001, plot_spacing = 0.05)
DCM_RV_vlnplot[[1]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
DCM_RV_vlnplot[[2]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
DCM_RV_vlnplot[[3]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
DCM_RV_vlnplot[[4]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
DCM_RV_vlnplot[[5]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
DCM_RV_vlnplot <- wrap_plots(DCM_RV_vlnplot) + 
  plot_annotation(title = "DCM RV") & 
  theme(plot.title = element_text(hjust = 0.5,      
                                  face = "bold",     
                                  size = 15)
  )
print(DCM_RV_vlnplot)
```

```{r DCM_RV_featureplot, fig.width=9, fig.height=6}
DCM_RV_featureplot <- FeaturePlot_scCustom(DCM_RV, features = lncrna_symbol, num_columns = 3)
DCM_RV_featureplot <- wrap_plots(DCM_RV_featureplot) + 
  plot_annotation(title = "DCM RV") & 
  theme(plot.title = element_text(hjust = 0.5,      
                                  face = "bold",     
                                  size = 20)
  )
print(DCM_RV_featureplot)
```

```{r DCM_RV metadata}
DCM_RV_CM <- subset(DCM_RV, cell_type %in% c("Cardiomyocytes"))

metadata <- DCM_RV_CM@meta.data %>%
  dplyr::select(Sample, Group) %>%
  unique()
rownames(metadata) <- metadata$Sample
reactable(metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

Remove genes expressed in <1% of cells

```{r DCM_RV_pseudocounts, eval=TRUE}
expr_matrix <- GetAssayData(object = DCM_RV_CM, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- DCM_RV_CM[which(rownames(DCM_RV_CM) %in% keep_genes),]
rm(keep_genes, non_zero_cells, expr_matrix, total_cells, percent_cells)
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(metadata$Sample, pseudocounts$samples$sample)]
all(metadata$donor_id == pseudocounts$samples$sample)
keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

```{r DCM_RV_DE}
metadata$Group <- factor(metadata$Group, levels = c("HC", "DCM"))
design<- model.matrix(~ Group, data = metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_DCM_RV <- topTags(fit, n = Inf)$table
results_DCM_RV <- merge(results_DCM_RV, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_DCM_RV$up_down <- ifelse(
  results_DCM_RV$logFC > 0.5 & results_DCM_RV$FDR < 0.05, "Increased",
  ifelse(results_DCM_RV$logFC < -0.5 & results_DCM_RV$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )
results_DCM_RV_lncRNA <- results_DCM_RV %>%
  dplyr::filter(gene %in% lncrna_symbol)
reactable(results_DCM_RV_lncRNA, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r DCM_RV_heatmap_DCM_vs_HC, fig.width=6, fig.height=2}
counts <- cpm(pseudocounts, log = TRUE)

counts <- counts[rownames(counts) %in% results_DCM_RV_lncRNA$gene, , drop = FALSE]

metadata_matched <- metadata[match(colnames(counts), metadata$Sample), ]

stopifnot(all(colnames(counts) == metadata_matched$Sample))

order_idx <- order(metadata_matched$Group)

counts <- counts[, order_idx]
metadata_matched <- metadata_matched[order_idx, ]

heat_colors <- rev(brewer.pal(11, "PuOr"))
pheatmap::pheatmap(counts, 
         color = heat_colors, 
         cluster_rows = F, 
         annotation_col = dplyr::select(metadata, Group), 
         border_color = NA, 
         fontsize = 10, 
         scale = "row",
         cluster_cols=F,
         fontsize_row = 15, 
         show_rownames = T,
         show_colnames = F,
         cellheight = 15,
         legend_breaks = c(-1.5, 0, 1.5)
         )
```

```{r rm DCM_RV}
rm(DCM_RV)
```

## Cao et al. 2020

```{r import Cao}
Cao <- readRDS("/scratch/gbonaz/Heart_datasets/Cao_2020/Heart_gene_count.RDS")
Cao_genes <- readRDS("/scratch/gbonaz/Heart_datasets/Cao_2020/df_gene.RDS")
Cao_cell <- readRDS("/scratch/gbonaz/Heart_datasets/Cao_2020/df_cell.RDS")

Cao_cell <- Cao_cell[match(colnames(Cao), Cao_cell$sample), ]
Cao_genes <- Cao_genes[match(rownames(Cao), rownames(Cao_genes)), ]
cao_seurat <- CreateSeuratObject(
  counts = Cao,
  meta.data = Cao_cell,
  assay = "RNA",
  project = "Cao_2020"
)

cao_seurat[["RNA"]]@meta.features <- Cao_genes
Cao <- cao_seurat
DefaultAssay(Cao)  <- "RNA"
Cao <- NormalizeData(Cao)
```

```{r Cao_dimplot, fig.width=8, fig.height=5, eval=FALSE}
Cao_dimplot <- DimPlot(Cao, group.by = "Main_cluster_name") + ggtitle("Cao et al. 2020")
Cao_dimplot
```

```{r genes_Cao}
lncrna_Cao <- c("ENSG00000254002.1", "ENSG00000228692.2", "ENSG00000226900.1", "ENSG00000215246.4", "ENSG00000253807.1")
lncrna_Cao %in% Cao_genes$gene_id
```

```{r NA_Cao}
table(is.na(Cao$Main_cluster_name))
Cao$Main_cluster_name[is.na(Cao$Main_cluster_name)] <- "Unassigned"
```


```{r Cao_dotplot, fig.width=6, fig.height=5}
Cao_dotplot <- DotPlot(Cao, features=lncrna_Cao, group.by="Main_cluster_name", scale = F, assay = "RNA") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +  
  ggtitle("Cao et al. 2020")
Cao_dotplot
```

```{r Cao_vlnplot, fig.width=6, fig.height=5}
Idents(Cao) <- "Main_cluster_name"
Cao_vlnplot <- Stacked_VlnPlot(Cao, features = lncrna_Cao, x_lab_rotate = TRUE, pt.size = 0.001, plot_spacing = 0.05)
Cao_vlnplot[[1]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Cao_vlnplot[[2]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Cao_vlnplot[[3]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Cao_vlnplot[[4]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Cao_vlnplot[[5]][[1]]$layers[[2]]$aes_params$alpha <- 0.3
Cao_vlnplot <- wrap_plots(Cao_vlnplot) + 
  plot_annotation(title = "Cao et al. 2020") & 
  theme(plot.title = element_text(hjust = 0.5,      
                                  face = "bold",     
                                  size = 15)
  )
print(Cao_vlnplot)
```


```{r Cao metadata}
Cao_CM <- subset(Cao, Main_cluster_name %in% c("Cardiomyocytes"))

metadata <- Cao_CM@meta.data %>%
  dplyr::select(Fetus_id, Development_day) %>%
  unique()
rownames(metadata) <- metadata$Fetus_id
reactable(metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

Remove genes expressed in <1% of cells

```{r Cao_pseudocounts, eval=TRUE}
expr_matrix <- GetAssayData(object = Cao_CM, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- Cao_CM[which(rownames(Cao_CM) %in% keep_genes),]
rm(keep_genes, non_zero_cells, expr_matrix, total_cells, percent_cells)
pseudocounts <- Seurat2PB(data, sample="Fetus_id", cluster = "Main_cluster_name")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(metadata$Fetus_id, pseudocounts$samples$sample)]
all(metadata$donor_id == pseudocounts$samples$sample)
keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

```{r Cao_DE}
design<- model.matrix(~ metadata$Development_day)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_Cao <- topTags(fit, n = Inf)$table
results_Cao <- merge(results_Cao, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_Cao$up_down <- ifelse(
  results_Cao$logFC > 0.5 & results_Cao$FDR < 0.05, "Increased",
  ifelse(results_Cao$logFC < -0.5 & results_Cao$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )
results_Cao_lncRNA <- results_Cao %>%
  dplyr::filter(gene %in% lncrna_Cao)
reactable(results_Cao_lncRNA, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Cao_heatmap_DCM_vs_HC, fig.width=6, fig.height=2}
counts <- cpm(pseudocounts, log = TRUE)

counts <- counts[rownames(counts) %in% results_Cao_lncRNA$gene, , drop = FALSE]

metadata_matched <- metadata[match(colnames(counts), metadata$Fetus_id), ]

stopifnot(all(colnames(counts) == metadata_matched$Fetus_id))

order_idx <- order(metadata_matched$Development_day)

counts <- counts[, order_idx]
metadata_matched <- metadata_matched[order_idx, ]

heat_colors <- rev(brewer.pal(11, "PuOr"))
pheatmap::pheatmap(counts, 
         color = heat_colors, 
         cluster_rows = F, 
         annotation_col = dplyr::select(metadata, Development_day), 
         border_color = NA, 
         fontsize = 10, 
         scale = "row",
         cluster_cols=F,
         fontsize_row = 15, 
         show_rownames = T,
         show_colnames = F,
         cellheight = 15,
         legend_breaks = c(-1.5, 0, 1.5)
         )
```

```{r rm Cao}
rm(Cao)
```
