---
title: "Differential abundance and differential expression analysis in all cell types with all clinical parameters"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "DA_DE_all_cell_types_and_parameters"


library(here)
library(dplyr)
library(ggplot2)
library(tibble)
library(readr)
library(readxl)
library(knitr)
library(patchwork)
library(scales)

library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)


library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(Matrix.utils)
library(scater)
library(statmod)
library(clustree)
library(harmony)
library(gprofiler2)
library(AnnotationHub)
library(ReactomePA)
library(speckle)
library(limma)
library(reactable)
library(decoupleR)
library(OmnipathR)
library(dorothea)
library(edgeR)
library(EnhancedVolcano)
library(ComplexHeatmap)
library(UpSetR)
library(cowplot)
library(GGally)
library(rtracklayer)
library(biomaRt)
library(ggpubr)
library(qs2)


#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```


Load DCM_RV dataset

```{r read CM}
DCM_RV <- qs_read(here::here("output", "QC_integration_annotation", "DCM_RV_RNA.qs2"), nthreads = 32)
```

```{r NormalizeData}
DefaultAssay(DCM_RV) <- "RNA"
DCM_RV <- NormalizeData(DCM_RV)
```

## Differential abundance analysis

```{r metadata}
DCM_RV$TAPSE_sPAP <- DCM_RV$TAPSE_mm/DCM_RV$sPAP_mmHg
table(DCM_RV$TAPSE_sPAP)

metadata <- DCM_RV@meta.data %>%
  dplyr::select(Sample:CO_l.min, TAPSE_sPAP) %>%
  unique() %>%
  dplyr::select(-LVID_mm, -RVIT_mm) 
metadata <- metadata %>%
  dplyr::select(1:3, 5, 6, 4, 7:ncol(metadata)) %>%
  dplyr::arrange(match(Sample, names(table(DCM_RV$Sample))))

rownames(metadata) <- metadata$Sample

# Check if metadata$Sample matches names(table(DCM_RV$Sample))
all.equal(metadata$Sample, names(table(DCM_RV$Sample)))

reactable(metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

# Filter for Group == "DCM"
metadata_DCM <- metadata %>%
  dplyr::filter(Group == "DCM")
```

```{r Speckle_all}
# Initialize an empty data frame for storing differential abundance results
differential_abundance <- data.frame()
props <- getTransformedProps(DCM_RV$cell_type, DCM_RV$Sample, transform="logit")

# Loop through each metadata column from Age_years to CO_l.min
for (i in 7:(length(metadata_DCM))) {
  metadata_subset <- metadata_DCM[complete.cases(metadata_DCM[[i]]),]
  props_subset <- props
  props_subset$Counts <- props$Counts[, metadata_subset$Sample, drop = FALSE]
  props_subset$TransformedProps <- props$TransformedProps[, metadata_subset$Sample, drop = FALSE]
  props_subset$Proportions <- props$Proportions[, metadata_subset$Sample, drop = FALSE]
  metadata_subset <- metadata_subset %>% 
    arrange(Sample, colnames(props_subset$Proportions))
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  fit <- lmFit(props_subset$TransformedProps, design)
  fit <- eBayes(fit, robust=TRUE)
  differential_abundance_temp <- topTable(fit, n = Inf, coef = 2)
  differential_abundance_temp$metadata <- colnames(metadata_subset[i])
  differential_abundance_temp$cell_state <- rownames(differential_abundance_temp)
  differential_abundance <- rbind(differential_abundance, differential_abundance_temp)
}

# Clean up the environment
rm(metadata_subset, props_subset, design, fit, differential_abundance_temp)

# Reset row names and arrange the results by adjusted p-value
rownames(differential_abundance) <- NULL
differential_abundance <- dplyr::arrange(differential_abundance, adj.P.Val)

write.csv(differential_abundance, file = here::here(output_dir_data, "DCM_RV_Differential_Abundance.csv"), quote=F, row.names = F)

reactable(differential_abundance, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Speckle_CO_l.min, fig.height=4, fig.width=4}
metadata_subset <- metadata[complete.cases(metadata[["CO_l.min"]]),]

props_subset <- props
props_subset$Counts <- props$Counts[, metadata_subset$Sample, drop = FALSE]
props_subset$TransformedProps <- props$TransformedProps[, metadata_subset$Sample, drop = FALSE]
props_subset$Proportions <- props$Proportions[, metadata_subset$Sample, drop = FALSE]

metadata_subset <- metadata_subset %>% 
  arrange(Sample, colnames(props_subset$Proportions))

design <- model.matrix(~ metadata_subset$CO_l.min + metadata_subset[["Age_years"]])
fit.prop <- lmFit(props_subset$Proportions,design)

par(mfrow=c(1,1), mar = c(5, 5, 3, 1))
  plot(metadata_subset$CO_l.min, props_subset$Proportions[10,], main = rownames(props_subset$Proportions)[10],
       pch=16, cex=2, ylab="Proportions", xlab="CO_l.min", cex.lab=2, cex.axis=1.5,
       cex.main=2.5)
  abline(a=fit.prop$coefficients[10,1], b=fit.prop$coefficients[10,2], col=4,
         lwd=1)

rm(metadata_subset, props_subset, design, fit.prop)
```

```{r Cell_type_sample_table, fig.height=4, fig.width=10, message=FALSE, warning=FALSE}
# Extract metadata
metadata_tbl <- DCM_RV@meta.data

# Create a contingency table with cell types as rows and samples as columns
cell_type_sample_table <- table(metadata_tbl$cell_type, metadata_tbl$Sample)

# Convert the table to a data frame for easier manipulation
cell_type_sample_df <- as.data.frame.matrix(cell_type_sample_table)

# Add the totals per sample (column-wise sum) as the last row
cell_type_sample_df["Total", ] <- colSums(cell_type_sample_df)

# Add the totals per cell type (row-wise sum) as the last column
cell_type_sample_df$Total <- rowSums(cell_type_sample_df)

# Create table grob
table_plot <- gridExtra::tableGrob(cell_type_sample_df)

# Show directly in the knitted document
grid::grid.newpage()
grid::grid.draw(table_plot)

# Save separately if you still want the png file
ggplot2::ggsave(
  here::here(output_dir_figs, "cell_type_sample_table.png"),
  plot = table_plot,
  dpi = 600,
  width = 22,
  height = 6
)

rm(table_plot, metadata_tbl)
```


## DE Cardiomyocytes

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Cardiomyocytes}
data <- subset(DCM_RV, cell_type == "Cardiomyocytes")
```

```{r Cardiomyocytes save percent_stats, eval=FALSE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Cardiomyocytes_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Cardiomyocytes pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Cardiomyocytes filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Cardiomyocytes_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Cardiomyocytes empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Cardiomyocytes DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Cardiomyocytes_pseudocounts.rds"))
pseudocounts_DCM <- pseudocounts[, metadata_DCM$Sample]
all.equal(metadata_DCM$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM))) {
  metadata_subset <- metadata_DCM[complete.cases(metadata_DCM[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Cardiomyocytes_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Cardiomyocytes_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Cardiomyocytes", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Cardiomyocytes_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Cardiomyocytes_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Cardiomyocytes_UpSet_plot_DE, fig.height=5, fig.width=9}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- signif[[i]]$gene
  names(gene_sets)[i] <- names(signif)[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 10]
UpSet(m, column_title = "Cardiomyocytes", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r barplots_DEGs, fig.height=4.5, fig.width=10}
n_de_genes <- data.frame()
for (i in seq_along(signif)) {
  parameter <- c(rep(names(signif)[i], 2))
  effect <- c("Upregulated", "Downregulated")
  n <- c(nrow(dplyr::filter(signif[[i]], logFC > 0)), nrow(dplyr::filter(signif[[i]], logFC < 0)))
  de_tot <- nrow(signif[[i]])
  n_de_genes <- rbind(n_de_genes, data.frame(parameter,effect,n, de_tot))
}
n_de_genes <- n_de_genes %>%
  arrange(desc(de_tot))
n_de_genes$parameter <- factor(n_de_genes$parameter, levels = unique(n_de_genes$parameter))

# Plot with customizations
ggplot(n_de_genes, aes(fill=effect, y=n, x=parameter)) + 
  geom_bar(position="stack", stat="identity") +
  scale_fill_manual(values = c("Upregulated" = "#A63A2A", "Downregulated" = "#004C8C")) +
  theme(text = element_text(size = 15), axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Clinical parameter", y = "Number of Genes", fill = "Effect") +
  guides(fill = guide_legend(title = NULL))
```

```{r Cardiomyocytes rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m)
```


## DE Fibroblasts

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Fibroblasts}
data <- subset(DCM_RV, cell_type == "Fibroblasts")
```

```{r Fibroblasts save percent_stats, eval=FALSE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Fibroblasts_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Fibroblasts pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Fibroblasts filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Fibroblasts_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Fibroblasts empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Fibroblasts DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Fibroblasts_pseudocounts.rds"))
pseudocounts_DCM <- pseudocounts[, metadata_DCM$Sample]
all.equal(metadata_DCM$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM))) {
  metadata_subset <- metadata_DCM[complete.cases(metadata_DCM[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Fibroblasts_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Fibroblasts_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Fibroblasts", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Fibroblasts_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Fibroblasts_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Fibroblasts_UpSet_plot_DE, fig.height=5, fig.width=9}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- signif[[i]]$gene
  names(gene_sets)[i] <- names(signif)[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 10]
UpSet(m, column_title = "Fibroblasts", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r Fibroblasts rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m)
```




## DE Endothelial cells

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Endothelial_cells}
data <- subset(DCM_RV, cell_type == "Endothelial cells")
```

```{r Endothelial_cells save percent_stats, eval=FALSE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Endothelial_cells_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Endothelial_cells pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Endothelial_cells filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Endothelial_cells_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Endothelial_cells empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Endothelial_cells DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Endothelial_cells_pseudocounts.rds"))
pseudocounts_DCM <- pseudocounts[, metadata_DCM$Sample]
all.equal(metadata_DCM$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM))) {
  metadata_subset <- metadata_DCM[complete.cases(metadata_DCM[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Endothelial_cells_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Endothelial_cells_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Endothelial_cells", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Endothelial_cells_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Endothelial_cells_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Endothelial_cells_UpSet_plot_DE, fig.height=5, fig.width=9}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- signif[[i]]$gene
  names(gene_sets)[i] <- names(signif)[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 10]
UpSet(m, column_title = "Endothelial_cells", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r Endothelial_cells rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m)
```


## DE Pericytes

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Pericytes}
data <- subset(DCM_RV, cell_type == "Pericytes")
```

```{r Pericytes save percent_stats, eval=FALSE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Pericytes_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Pericytes pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Pericytes filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Pericytes_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Pericytes empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Pericytes DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Pericytes_pseudocounts.rds"))
metadata_DCM_subset <- dplyr::filter(metadata_DCM, Sample %in% pseudocounts$samples$sample)
pseudocounts_DCM <- pseudocounts[, metadata_DCM_subset$Sample]
all.equal(metadata_DCM_subset$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM_subset))) {
  metadata_subset <- metadata_DCM_subset[complete.cases(metadata_DCM_subset[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Pericytes_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Pericytes_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Pericytes", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Pericytes_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Pericytes_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Pericytes_UpSet_plot_DE, fig.height=5, fig.width=9}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- signif[[i]]$gene
  names(gene_sets)[i] <- names(signif)[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 10]
UpSet(m, column_title = "Pericytes", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r Pericytes rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m, metadata_DCM_subset)
```


## DE Macrophages

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Macrophages}
data <- subset(DCM_RV, cell_type == "Macrophages")
```

```{r Macrophages save percent_stats, eval=FALSE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Macrophages_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Macrophages pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Macrophages filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Macrophages_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Macrophages empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Macrophages DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Macrophages_pseudocounts.rds"))
metadata_DCM_subset <- dplyr::filter(metadata_DCM, Sample %in% pseudocounts$samples$sample)
pseudocounts_DCM <- pseudocounts[, metadata_DCM_subset$Sample]
all.equal(metadata_DCM_subset$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM_subset))) {
  metadata_subset <- metadata_DCM_subset[complete.cases(metadata_DCM_subset[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Macrophages_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Macrophages_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Macrophages", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Macrophages_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Macrophages_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Macrophages_UpSet_plot_DE, fig.height=5, fig.width=9}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- signif[[i]]$gene
  names(gene_sets)[i] <- names(signif)[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 10]
UpSet(m, column_title = "Macrophages", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r Macrophages rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m, metadata_DCM_subset)
```


## DE Lymphocytes

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Lymphocytes}
data <- subset(DCM_RV, cell_type == "T lymphocytes")
```

```{r Lymphocytes save percent_stats, eval=FALSE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "T_lymphocytes_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Lymphocytes pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Lymphocytes filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "T_lymphocytes_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Lymphocytes empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Lymphocytes DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "T_lymphocytes_pseudocounts.rds"))
metadata_DCM_subset <- dplyr::filter(metadata_DCM, Sample %in% pseudocounts$samples$sample)
pseudocounts_DCM <- pseudocounts[, metadata_DCM_subset$Sample]
all.equal(metadata_DCM_subset$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM_subset))) {
  metadata_subset <- metadata_DCM_subset[complete.cases(metadata_DCM_subset[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("T_lymphocytes_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("T_lymphocytes_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("T lymphocytes", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "T_lymphocytes_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Lymphocytes_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```


```{r Lymphocytes rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m, metadata_DCM_subset)
```


## DE Smooth muscle cells

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Smooth_muscle_cells}
data <- subset(DCM_RV, cell_type == "Smooth muscle cells")
```

```{r Smooth_muscle_cells save percent_stats, eval=FALSE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Smooth_muscle_cells_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Smooth_muscle_cells pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Smooth_muscle_cells filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Smooth_muscle_cells_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Smooth_muscle_cells empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Smooth_muscle_cells DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Smooth_muscle_cells_pseudocounts.rds"))
metadata_DCM_subset <- dplyr::filter(metadata_DCM, Sample %in% pseudocounts$samples$sample)
pseudocounts_DCM <- pseudocounts[, metadata_DCM_subset$Sample]
all.equal(metadata_DCM_subset$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM_subset))) {
  metadata_subset <- metadata_DCM_subset[complete.cases(metadata_DCM_subset[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Smooth_muscle_cells_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Smooth_muscle_cells_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Smooth_muscle_cells", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Smooth_muscle_cells_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Smooth_muscle_cells_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Smooth_muscle_cells_UpSet_plot_DE, fig.height=5, fig.width=9}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- signif[[i]]$gene
  names(gene_sets)[i] <- names(signif)[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 10]
UpSet(m, column_title = "Smooth_muscle_cells", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r Smooth_muscle_cells rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m, metadata_DCM_subset)
```


## DE Neuronal cells

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Neuronal_cells}
data <- subset(DCM_RV, cell_type == "Neuronal cells")
```

```{r Neuronal_cells save percent_stats, eval=FALSE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Neuronal_cells_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Neuronal_cells pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Neuronal_cells filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Neuronal_cells_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Neuronal_cells empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Neuronal_cells DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Neuronal_cells_pseudocounts.rds"))
metadata_DCM_subset <- dplyr::filter(metadata_DCM, Sample %in% pseudocounts$samples$sample)
pseudocounts_DCM <- pseudocounts[, metadata_DCM_subset$Sample]
all.equal(metadata_DCM_subset$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM_subset))) {
  metadata_subset <- metadata_DCM_subset[complete.cases(metadata_DCM_subset[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Neuronal_cells_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Neuronal_cells_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Neuronal_cells", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Neuronal_cells_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Neuronal_cells_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Neuronal_cells rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m, metadata_DCM_subset)
```


## DE Endocardial cells

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Endocardial_cells}
data <- subset(DCM_RV, cell_type == "Endocardial cells")
```

```{r Endocardial_cells save percent_stats, eval=FALSE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Endocardial_cells_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Endocardial_cells pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Endocardial_cells filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Endocardial_cells_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Endocardial_cells empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Endocardial_cells DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Endocardial_cells_pseudocounts.rds"))
metadata_DCM_subset <- dplyr::filter(metadata_DCM, Sample %in% pseudocounts$samples$sample)
pseudocounts_DCM <- pseudocounts[, metadata_DCM_subset$Sample]
all.equal(metadata_DCM_subset$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM_subset))) {
  metadata_subset <- metadata_DCM_subset[complete.cases(metadata_DCM_subset[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Endocardial_cells_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Endocardial_cells_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Endocardial_cells", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Endocardial_cells_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Endocardial_cells_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Endocardial_cells rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m, metadata_DCM_subset)
```


## Heatmap 

```{r, fig.height=6, fig.width=6}
# ---- Define cell types you analyzed ----
cell_types <- c("Cardiomyocytes", "Fibroblasts", "Endothelial_cells", "Pericytes", 
                "Macrophages", "T_lymphocytes", "Smooth_muscle_cells", "Neuronal_cells",
                "Endocardial_cells")  # update as needed

# ---- Read in all summary CSVs and build DEG matrix ----
deg_summary_list <- lapply(cell_types, function(celltype) {
  file <- file.path(output_dir_data, paste0(celltype, "_n_de_genes.csv"))
  df <- read.csv(file)
  df <- df %>% dplyr::select(metadata, n_tot) %>%
    dplyr::rename(parameter = metadata) %>%
    dplyr::mutate(cell_type = celltype)
  return(df)
})

# Combine into long format and pivot
deg_summary <- bind_rows(deg_summary_list)

deg_mat <- deg_summary %>%
  pivot_wider(names_from = cell_type, values_from = n_tot, values_fill = 0) %>%
  column_to_rownames("parameter") %>%
  as.matrix()

# ---- Plot heatmap ----
pheatmap(deg_mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("grey95", "red", "darkred"))(100),
         fontsize = 10,
         main = "DEG Counts by Parameter and Cell Type",
         border_color = NA,
         display_numbers = TRUE,           # <- this adds the numbers
         number_format = "%.0f",           # format as integer (no decimal)
         number_color = "black")
```

