---
title: "Differential abundance and differential expression analysis in all cell types with all clinical parameters"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "DA_DE_all_cell_types_and_parameters"


library(here)
library(dplyr)
library(ggplot2)
library(tibble)
library(readr)
library(readxl)
library(knitr)
library(patchwork)
library(scales)

library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)


library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(Matrix.utils)
library(scater)
library(statmod)
library(clustree)
library(harmony)
library(gprofiler2)
library(AnnotationHub)
library(ReactomePA)
library(speckle)
library(limma)
library(reactable)
library(decoupleR)
library(OmnipathR)
library(dorothea)
library(edgeR)
library(EnhancedVolcano)
library(ComplexHeatmap)
library(UpSetR)
library(cowplot)
library(GGally)
library(rtracklayer)
library(biomaRt)
library(ggpubr)
library(qs2)


#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```


Load DCM_RV dataset

```{r read CM}
DCM_RV <- qs_read(here::here("output", "Add_cell_states_remove_contamination", "DCM_RV_cell_states_RNA.qs2"), nthreads = 32)
```

```{r NormalizeData}
DefaultAssay(DCM_RV) <- "RNA"
DCM_RV <- NormalizeData(DCM_RV)
```

```{r}
Clinical_data <- read_excel(here::here("data", "DCM_Clinical_data_22.xlsx"))

# make cell-level metadata by matching Sample
add_metadata <- DCM_RV[[]] %>%
  dplyr::select(Sample) %>%
  left_join(Clinical_data, by = "Sample")

row.names(add_metadata) <- row.names(DCM_RV[[]])

# remove old columns that are present in the new clinical file
cols_to_overwrite <- setdiff(colnames(Clinical_data), "Sample")

DCM_RV@meta.data[, cols_to_overwrite] <- NULL

# add corrected metadata
DCM_RV <- AddMetaData(
  DCM_RV,
  metadata = add_metadata
)

rm(add_metadata, Clinical_data, cols_to_overwrite)
```

## Differential abundance analysis

```{r metadata}
DCM_RV$TAPSE_sPAP <- DCM_RV$TAPSE_mm/DCM_RV$sPAP_mmHg
table(DCM_RV$TAPSE_sPAP)

metadata <- DCM_RV@meta.data %>%
  dplyr::select(Sample, Batch_processing:CO_l.min, TAPSE_sPAP) %>%
  unique() %>%
  dplyr::select(-LVID_mm, -RVOT_mm) 
metadata <- metadata %>%
  dplyr::select(1:3, 5, 6, 4, 7:ncol(metadata)) %>%
  dplyr::arrange(match(Sample, names(table(DCM_RV$Sample))))

rownames(metadata) <- metadata$Sample

# Check if metadata$Sample matches names(table(DCM_RV$Sample))
all.equal(metadata$Sample, names(table(DCM_RV$Sample)))

reactable(metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

# Filter for Group == "DCM"
metadata_DCM <- metadata %>%
  dplyr::filter(Group == "DCM")
```

```{r Speckle_all}
# Initialize an empty data frame for storing differential abundance results
differential_abundance <- data.frame()
props <- getTransformedProps(DCM_RV$cell_type, DCM_RV$Sample, transform="logit")

# Loop through each metadata column from Age_years to CO_l.min
for (i in 7:(length(metadata_DCM))) {
  metadata_subset <- metadata_DCM[complete.cases(metadata_DCM[[i]]),]
  props_subset <- props
  props_subset$Counts <- props$Counts[, metadata_subset$Sample, drop = FALSE]
  props_subset$TransformedProps <- props$TransformedProps[, metadata_subset$Sample, drop = FALSE]
  props_subset$Proportions <- props$Proportions[, metadata_subset$Sample, drop = FALSE]
  metadata_subset <- metadata_subset %>% 
    arrange(Sample, colnames(props_subset$Proportions))
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  fit <- lmFit(props_subset$TransformedProps, design)
  fit <- eBayes(fit, robust=TRUE)
  differential_abundance_temp <- topTable(fit, n = Inf, coef = 2)
  differential_abundance_temp$metadata <- colnames(metadata_subset[i])
  differential_abundance_temp$cell_state <- rownames(differential_abundance_temp)
  differential_abundance <- rbind(differential_abundance, differential_abundance_temp)
}

# Clean up the environment
rm(metadata_subset, props_subset, design, fit, differential_abundance_temp)

# Reset row names and arrange the results by adjusted p-value
rownames(differential_abundance) <- NULL
differential_abundance <- dplyr::arrange(differential_abundance, adj.P.Val)

write.csv(differential_abundance, file = here::here(output_dir_data, "DCM_RV_Differential_Abundance.csv"), quote=F, row.names = F)

reactable(differential_abundance, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r Speckle_CO_l.min, fig.height=4, fig.width=4}
metadata_subset <- metadata[complete.cases(metadata[["CO_l.min"]]),]

props_subset <- props
props_subset$Counts <- props$Counts[, metadata_subset$Sample, drop = FALSE]
props_subset$TransformedProps <- props$TransformedProps[, metadata_subset$Sample, drop = FALSE]
props_subset$Proportions <- props$Proportions[, metadata_subset$Sample, drop = FALSE]

metadata_subset <- metadata_subset %>% 
  arrange(Sample, colnames(props_subset$Proportions))

design <- model.matrix(~ metadata_subset$CO_l.min + metadata_subset[["Age_years"]])
fit.prop <- lmFit(props_subset$Proportions,design)

par(mfrow=c(1,1), mar = c(5, 5, 3, 1))
  plot(metadata_subset$CO_l.min, props_subset$Proportions[10,], main = rownames(props_subset$Proportions)[10],
       pch=16, cex=2, ylab="Proportions", xlab="CO_l.min", cex.lab=2, cex.axis=1.5,
       cex.main=2.5)
  abline(a=fit.prop$coefficients[10,1], b=fit.prop$coefficients[10,2], col=4,
         lwd=1)

rm(metadata_subset, props_subset, design, fit.prop)
```

```{r Cell_type_sample_table, fig.height=4, fig.width=10, message=FALSE, warning=FALSE}
# Extract metadata
metadata_tbl <- DCM_RV@meta.data

# Create a contingency table with cell types as rows and samples as columns
cell_type_sample_table <- table(metadata_tbl$cell_type, metadata_tbl$Sample)

# Convert the table to a data frame for easier manipulation
cell_type_sample_df <- as.data.frame.matrix(cell_type_sample_table)

# Add the totals per sample (column-wise sum) as the last row
cell_type_sample_df["Total", ] <- colSums(cell_type_sample_df)

# Add the totals per cell type (row-wise sum) as the last column
cell_type_sample_df$Total <- rowSums(cell_type_sample_df)

# Create table grob
table_plot <- gridExtra::tableGrob(cell_type_sample_df)

# Show directly in the knitted document
grid::grid.newpage()
grid::grid.draw(table_plot)

# Save separately if you still want the png file
ggplot2::ggsave(
  here::here(output_dir_figs, "cell_type_sample_table.png"),
  plot = table_plot,
  dpi = 600,
  width = 22,
  height = 6
)

rm(table_plot, metadata_tbl)
```


## DE Cardiomyocytes

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Cardiomyocytes}
data <- subset(DCM_RV, cell_type == "Cardiomyocytes")
```

```{r Cardiomyocytes save percent_stats, eval=TRUE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Cardiomyocytes_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Cardiomyocytes pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Cardiomyocytes filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Cardiomyocytes_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Cardiomyocytes empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Cardiomyocytes DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Cardiomyocytes_pseudocounts.rds"))
pseudocounts_DCM <- pseudocounts[, metadata_DCM$Sample]
all.equal(metadata_DCM$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM))) {
  metadata_subset <- metadata_DCM[complete.cases(metadata_DCM[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Cardiomyocytes_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Cardiomyocytes_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Cardiomyocytes", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Cardiomyocytes_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Cardiomyocytes_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Cardiomyocytes_UpSet_plot_DE, fig.height=5, fig.width=9}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- signif[[i]]$gene
  names(gene_sets)[i] <- names(signif)[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 10]
UpSet(m, column_title = "Cardiomyocytes", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r barplots_DEGs, fig.height=4.5, fig.width=10}
n_de_genes <- data.frame()
for (i in seq_along(signif)) {
  parameter <- c(rep(names(signif)[i], 2))
  effect <- c("Upregulated", "Downregulated")
  n <- c(nrow(dplyr::filter(signif[[i]], logFC > 0)), nrow(dplyr::filter(signif[[i]], logFC < 0)))
  de_tot <- nrow(signif[[i]])
  n_de_genes <- rbind(n_de_genes, data.frame(parameter,effect,n, de_tot))
}
n_de_genes <- n_de_genes %>%
  arrange(desc(de_tot))
n_de_genes$parameter <- factor(n_de_genes$parameter, levels = unique(n_de_genes$parameter))

# Plot with customizations
ggplot(n_de_genes, aes(fill=effect, y=n, x=parameter)) + 
  geom_bar(position="stack", stat="identity") +
  scale_fill_manual(values = c("Upregulated" = "#A63A2A", "Downregulated" = "#004C8C")) +
  theme(text = element_text(size = 15), axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Clinical parameter", y = "Number of Genes", fill = "Effect") +
  guides(fill = guide_legend(title = NULL))
```

```{r Cardiomyocytes rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m)
```


## DE Fibroblasts

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Fibroblasts}
data <- subset(DCM_RV, cell_type == "Fibroblasts")
```

```{r Fibroblasts save percent_stats, eval=TRUE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Fibroblasts_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Fibroblasts pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Fibroblasts filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Fibroblasts_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Fibroblasts empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Fibroblasts DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Fibroblasts_pseudocounts.rds"))
pseudocounts_DCM <- pseudocounts[, metadata_DCM$Sample]
all.equal(metadata_DCM$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM))) {
  metadata_subset <- metadata_DCM[complete.cases(metadata_DCM[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Fibroblasts_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Fibroblasts_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Fibroblasts", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Fibroblasts_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Fibroblasts_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Fibroblasts_UpSet_plot_DE, fig.height=5, fig.width=9}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- signif[[i]]$gene
  names(gene_sets)[i] <- names(signif)[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 10]
UpSet(m, column_title = "Fibroblasts", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r Fibroblasts rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m)
```




## DE Endothelial cells

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Endothelial_cells}
data <- subset(DCM_RV, cell_type == "Endothelial cells")
```

```{r Endothelial_cells save percent_stats, eval=TRUE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Endothelial_cells_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Endothelial_cells pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Endothelial_cells filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Endothelial_cells_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Endothelial_cells empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Endothelial_cells DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Endothelial_cells_pseudocounts.rds"))
pseudocounts_DCM <- pseudocounts[, metadata_DCM$Sample]
all.equal(metadata_DCM$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM))) {
  metadata_subset <- metadata_DCM[complete.cases(metadata_DCM[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Endothelial_cells_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Endothelial_cells_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Endothelial_cells", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Endothelial_cells_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Endothelial_cells_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Endothelial_cells_UpSet_plot_DE, fig.height=5, fig.width=9}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- signif[[i]]$gene
  names(gene_sets)[i] <- names(signif)[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 10]
UpSet(m, column_title = "Endothelial_cells", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r Endothelial_cells rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m)
```


## DE Pericytes

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Pericytes}
data <- subset(DCM_RV, cell_type == "Pericytes")
```

```{r Pericytes save percent_stats, eval=TRUE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Pericytes_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Pericytes pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Pericytes filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Pericytes_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Pericytes empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Pericytes DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Pericytes_pseudocounts.rds"))
metadata_DCM_subset <- dplyr::filter(metadata_DCM, Sample %in% pseudocounts$samples$sample)
pseudocounts_DCM <- pseudocounts[, metadata_DCM_subset$Sample]
all.equal(metadata_DCM_subset$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM_subset))) {
  metadata_subset <- metadata_DCM_subset[complete.cases(metadata_DCM_subset[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Pericytes_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Pericytes_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Pericytes", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Pericytes_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Pericytes_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Pericytes_UpSet_plot_DE, fig.height=5, fig.width=9}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- signif[[i]]$gene
  names(gene_sets)[i] <- names(signif)[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 10]
UpSet(m, column_title = "Pericytes", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r Pericytes rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m, metadata_DCM_subset)
```


## DE Macrophages

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Macrophages}
data <- subset(DCM_RV, cell_type == "Macrophages")
```

```{r Macrophages save percent_stats, eval=TRUE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Macrophages_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Macrophages pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Macrophages filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Macrophages_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Macrophages empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Macrophages DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Macrophages_pseudocounts.rds"))
metadata_DCM_subset <- dplyr::filter(metadata_DCM, Sample %in% pseudocounts$samples$sample)
pseudocounts_DCM <- pseudocounts[, metadata_DCM_subset$Sample]
all.equal(metadata_DCM_subset$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM_subset))) {
  metadata_subset <- metadata_DCM_subset[complete.cases(metadata_DCM_subset[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Macrophages_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Macrophages_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Macrophages", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Macrophages_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Macrophages_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Macrophages_UpSet_plot_DE, fig.height=5, fig.width=9}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- signif[[i]]$gene
  names(gene_sets)[i] <- names(signif)[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 10]
UpSet(m, column_title = "Macrophages", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r Macrophages rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m, metadata_DCM_subset)
```


## DE Lymphocytes

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Lymphocytes}
data <- subset(DCM_RV, cell_type == "T lymphocytes")
```

```{r Lymphocytes save percent_stats, eval=TRUE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "T_lymphocytes_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Lymphocytes pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Lymphocytes filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "T_lymphocytes_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Lymphocytes empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Lymphocytes DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "T_lymphocytes_pseudocounts.rds"))
metadata_DCM_subset <- dplyr::filter(metadata_DCM, Sample %in% pseudocounts$samples$sample)
pseudocounts_DCM <- pseudocounts[, metadata_DCM_subset$Sample]
all.equal(metadata_DCM_subset$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM_subset))) {
  metadata_subset <- metadata_DCM_subset[complete.cases(metadata_DCM_subset[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("T_lymphocytes_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("T_lymphocytes_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("T lymphocytes", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "T_lymphocytes_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Lymphocytes_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```


```{r Lymphocytes rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m, metadata_DCM_subset)
```


## DE Smooth muscle cells

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Smooth_muscle_cells}
data <- subset(DCM_RV, cell_type == "Smooth muscle cells")
```

```{r Smooth_muscle_cells save percent_stats, eval=TRUE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Smooth_muscle_cells_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Smooth_muscle_cells pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Smooth_muscle_cells filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Smooth_muscle_cells_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Smooth_muscle_cells empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Smooth_muscle_cells DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Smooth_muscle_cells_pseudocounts.rds"))
metadata_DCM_subset <- dplyr::filter(metadata_DCM, Sample %in% pseudocounts$samples$sample)
pseudocounts_DCM <- pseudocounts[, metadata_DCM_subset$Sample]
all.equal(metadata_DCM_subset$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM_subset))) {
  metadata_subset <- metadata_DCM_subset[complete.cases(metadata_DCM_subset[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Smooth_muscle_cells_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Smooth_muscle_cells_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Smooth_muscle_cells", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Smooth_muscle_cells_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Smooth_muscle_cells_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Smooth_muscle_cells_UpSet_plot_DE, fig.height=5, fig.width=9}
gene_sets <- list()
for (i in 1:length(signif)) {
  gene_sets[[i]] <- signif[[i]]$gene
  names(gene_sets)[i] <- names(signif)[i]
  }
m <- make_comb_mat(gene_sets, mode = "intersect", min_set_size = 10)
m <- m[comb_degree(m) >= 2]
m <- m[comb_size(m) >= 10]
UpSet(m, column_title = "Smooth_muscle_cells", comb_order = order(comb_size(m)), 
                    top_annotation = upset_top_annotation(m, add_numbers = TRUE),
                    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
```

```{r Smooth_muscle_cells rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m, metadata_DCM_subset)
```


## DE Neuronal cells

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Neuronal_cells}
data <- subset(DCM_RV, cell_type == "Neuronal cells")
```

```{r Neuronal_cells save percent_stats, eval=TRUE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Neuronal_cells_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Neuronal_cells pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Neuronal_cells filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Neuronal_cells_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Neuronal_cells empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Neuronal_cells DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Neuronal_cells_pseudocounts.rds"))
metadata_DCM_subset <- dplyr::filter(metadata_DCM, Sample %in% pseudocounts$samples$sample)
pseudocounts_DCM <- pseudocounts[, metadata_DCM_subset$Sample]
all.equal(metadata_DCM_subset$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM_subset))) {
  metadata_subset <- metadata_DCM_subset[complete.cases(metadata_DCM_subset[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Neuronal_cells_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Neuronal_cells_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Neuronal_cells", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Neuronal_cells_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Neuronal_cells_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Neuronal_cells rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m, metadata_DCM_subset)
```


## DE Endocardial cells

Remove genes expressed in <1% of cells and MT and RP genes

```{r subset Endocardial_cells}
data <- subset(DCM_RV, cell_type == "Endocardial cells")
```

```{r Endocardial_cells save percent_stats, eval=TRUE}
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
write.csv(percent_stats, file = here::here(output_dir_data, "Endocardial_cells_percent_stats.csv"), quote=F, row.names = F)
```

Create and prepare pseudobulk object for differential expression analysis

```{r Endocardial_cells pseudocounts}
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
```

```{r Endocardial_cells filter pseudocounts}
percent_stats <- read.csv(here::here(output_dir_data, "Endocardial_cells_percent_stats.csv"))
rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```

Create empty objects for the DE output

```{r Endocardial_cells empty_objects_DE}
results <- list()
signif <- list()
volcano <- list()
n_de_genes <- data.frame()
```

```{r Endocardial_cells DE_all}
saveRDS(pseudocounts, 
        here::here(output_dir_data, "Endocardial_cells_pseudocounts.rds"))
metadata_DCM_subset <- dplyr::filter(metadata_DCM, Sample %in% pseudocounts$samples$sample)
pseudocounts_DCM <- pseudocounts[, metadata_DCM_subset$Sample]
all.equal(metadata_DCM_subset$Sample, colnames(pseudocounts_DCM$counts))

for (i in 7:(length(metadata_DCM_subset))) {
  metadata_subset <- metadata_DCM_subset[complete.cases(metadata_DCM_subset[[i]]),]
  metadata_subset[,i] <- rescale(metadata_subset[,i])
  design <- model.matrix(~ metadata_subset[[i]] + metadata_subset[["Age_years"]])
  colnames(design)[2] <- colnames(metadata_subset[i])
  
  pseudocounts_subset <- pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)
  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)
  
  fit <- glmQLFit(pseudocounts_subset, design, robust=TRUE)
  fit <- glmQLFTest(fit, coef = 2)
  
  results[[i-6]] <- topTags(fit, n = Inf)$table
  results[[i-6]] <- merge(results[[i-6]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
  signif[[i-6]] <- results[[i-6]] %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)

  names(results)[i-6] <- colnames(metadata_subset[i])
  names(signif)[i-6] <- colnames(metadata_subset[i])
  
  write.csv(results[[i-6]], file = here::here(output_dir_data, paste0("Endocardial_cells_DE_Results_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  write.csv(signif[[i-6]], file = here::here(output_dir_data, paste0("Endocardial_cells_DE_Significant_", colnames(metadata[i]), ".csv")), quote=F, row.names = F)
  
  n_de_genes_temp <- data.frame(metadata = colnames(metadata_subset[i]),
                                n_upregulated = sum(signif[[i-6]]$logFC > 0.5),
                                n_downregulated = sum(signif[[i-6]]$logFC < -0.5),
                                n_tot = nrow(signif[[i-6]])
                                )
  n_de_genes <- rbind(n_de_genes, n_de_genes_temp)
  
  keyvals <- rep("grey", nrow(results[[i-6]]))
  names(keyvals) <- rep("Not significant", nrow(results[[i-6]]))
  keyvals[results[[i-6]]$logFC > 0.5 & results[[i-6]]$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results[[i-6]]$logFC < -0.5 & results[[i-6]]$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"
  
  volcano[[i-6]] <- EnhancedVolcano(results[[i-6]],
                  lab = results[[i-6]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 0.5,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 3.5),
                  colCustom = keyvals,
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = paste0("Endocardial_cells", "\n", colnames(metadata_subset[i]), "\nn = ", nrow(metadata_subset))
  ) + theme(legend.position = "none")
  names(volcano)[i-6] <- colnames(metadata_subset[i])
}
write.csv(n_de_genes, file = here::here(output_dir_data, "Endocardial_cells_n_de_genes.csv"), quote=F, row.names = F)
rm(metadata_subset, design, pseudocounts_subset, fit, n_de_genes_temp)
```

```{r Endocardial_cells_Volcano_all_metadata, fig.height=9, fig.width=21, dpi=300}
print((volcano[[1]] | volcano[[2]] | volcano[[3]] | volcano[[4]] | volcano[[5]] | volcano[[6]] | volcano[[7]]) / 
  (volcano[[8]] | volcano[[9]] | volcano[[10]] |  volcano[[11]] | volcano[[12]] | volcano[[13]] | volcano[[14]]))
```

```{r Endocardial_cells rm}
rm(pseudocounts, pseudocounts_DCM, percent_stats, results, signif, volcano, n_de_genes, gene_sets, m, metadata_DCM_subset)
```


## Heatmap 

```{r, fig.height=6, fig.width=6}
# ---- Define cell types you analyzed ----
cell_types <- c("Cardiomyocytes", "Fibroblasts", "Endothelial_cells", "Pericytes", 
                "Macrophages", "T_lymphocytes", "Smooth_muscle_cells", "Neuronal_cells",
                "Endocardial_cells")  # update as needed

# ---- Read in all summary CSVs and build DEG matrix ----
deg_summary_list <- lapply(cell_types, function(celltype) {
  file <- file.path(output_dir_data, paste0(celltype, "_n_de_genes.csv"))
  df <- read.csv(file)
  df <- df %>% dplyr::select(metadata, n_tot) %>%
    dplyr::rename(parameter = metadata) %>%
    dplyr::mutate(cell_type = celltype)
  return(df)
})

# Combine into long format and pivot
deg_summary <- bind_rows(deg_summary_list)

deg_mat <- deg_summary %>%
  pivot_wider(names_from = cell_type, values_from = n_tot, values_fill = 0) %>%
  column_to_rownames("parameter") %>%
  as.matrix()

# ---- Plot heatmap ----
pheatmap(deg_mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("grey95", "red", "darkred"))(100),
         fontsize = 10,
         main = "DEG Counts by Parameter and Cell Type",
         border_color = NA,
         display_numbers = TRUE,           # <- this adds the numbers
         number_format = "%.0f",           # format as integer (no decimal)
         number_color = "black")
```

```{r deg_count_heatmap_mainfig, fig.height=4.6, fig.width=6.4, dpi = 300}
# ---- Cell types in desired order ----
cell_types <- c(
  "Cardiomyocytes",
  "Fibroblasts",
  "Endothelial_cells",
  "Pericytes",
  "Smooth_muscle_cells",
  "Macrophages",
  "Neuronal_cells",
  "T_lymphocytes",
  "Endocardial_cells"
)

# ---- Variables shown in the main DEG/volcano figure ----
vars_keep <- c(
  "mPAP_mmHg",
  "sPAP_mmHg",
  "PCW_mmHg",
  "PVR_WU",
  "TAPSE_mm",
  "RVIT_mm",
  "LVEF_percent",
  "LVEDD_mm"
)

# ---- Nice labels ----
pretty_vars <- c(
  "mPAP_mmHg"    = "mPAP (mmHg)",
  "sPAP_mmHg"    = "sPAP (mmHg)",
  "PCW_mmHg"     = "PCWP (mmHg)",
  "PVR_WU"       = "PVR (WU)",
  "TAPSE_mm"     = "TAPSE (mm)",
  "RVIT_mm"      = "RVIT (mm)",
  "LVEF_percent" = "LVEF (%)",
  "LVEDD_mm"     = "LVEDD (mm)"
)

pretty_celltypes <- c(
  "Cardiomyocytes"       = "Cardiomyocytes",
  "Fibroblasts"          = "Fibroblasts",
  "Endothelial_cells"    = "Endothelial cells",
  "Pericytes"            = "Pericytes",
  "Smooth_muscle_cells"  = "Smooth muscle cells",
  "Macrophages"          = "Macrophages",
  "Neuronal_cells"       = "Neuronal cells",
  "T_lymphocytes"        = "T lymphocytes",
  "Endocardial_cells"    = "Endocardial cells"
)

# ---- Read summary CSVs ----
deg_summary_list <- lapply(cell_types, function(celltype) {
  file <- file.path(output_dir_data, paste0(celltype, "_n_de_genes.csv"))

  read.csv(file) %>%
    dplyr::select(metadata, n_tot) %>%
    dplyr::rename(parameter = metadata) %>%
    dplyr::mutate(cell_type = celltype)
})

deg_summary <- dplyr::bind_rows(deg_summary_list)

# ---- Build matrix ----
deg_mat <- deg_summary %>%
  dplyr::filter(parameter %in% vars_keep) %>%
  dplyr::mutate(
    parameter = factor(parameter, levels = vars_keep),
    cell_type = factor(cell_type, levels = cell_types)
  ) %>%
  tidyr::pivot_wider(
    names_from = cell_type,
    values_from = n_tot,
    values_fill = 0
  ) %>%
  dplyr::arrange(parameter) %>%
  tibble::column_to_rownames("parameter") %>%
  as.matrix()

deg_mat <- deg_mat[vars_keep, cell_types, drop = FALSE]

# ---- Optional: hide zeros in the displayed numbers ----
deg_numbers <- ifelse(deg_mat == 0, "", deg_mat)

# ---- Proteomics-style red heatmap palette ----
deg_cols <- colorRampPalette(c("white", "#A63A2A"))(100)

# ---- Plot ----
pheatmap::pheatmap(
  deg_mat,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = deg_cols,
  breaks = seq(0, max(deg_mat, na.rm = TRUE), length.out = 101),
  fontsize = 10,
  fontsize_row = 11,
  fontsize_col = 10,
  border_color = "grey90",
  display_numbers = deg_numbers,
  number_color = "black",
  labels_row = pretty_vars[rownames(deg_mat)],
  labels_col = pretty_celltypes[colnames(deg_mat)],
  angle_col = 45,
  main = NA,
  legend = TRUE
)
```

```{r deg_count_heatmap_simple_mainfig_narrow, fig.width=5, fig.height=5.4, dpi=300}
cell_types <- c(
  "Cardiomyocytes",
  "Fibroblasts",
  "Endothelial_cells",
  "Pericytes",
  "Smooth_muscle_cells",
  "Macrophages",
  "Neuronal_cells",
  "T_lymphocytes",
  "Endocardial_cells"
)

# ---- Variables in desired order ----
# PVR and PCWP inverted
vars_keep <- c(
  "mPAP_mmHg",
  "sPAP_mmHg",
  "PVR_WU",
  "PCW_mmHg",
  "TAPSE_mm",
  "RVIT_mm",
  "LVEF_percent",
  "LVEDD_mm"
)

# ---- Nice labels ----
pretty_vars <- c(
  "mPAP_mmHg"    = "mPAP (mmHg)",
  "sPAP_mmHg"    = "sPAP (mmHg)",
  "PVR_WU"       = "PVR (WU)",
  "PCW_mmHg"     = "PCWP (mmHg)",
  "TAPSE_mm"     = "TAPSE (mm)",
  "RVIT_mm"      = "RVIT (mm)",
  "LVEF_percent" = "LVEF (%)",
  "LVEDD_mm"     = "LVEDD (mm)"
)

pretty_celltypes <- c(
  "Cardiomyocytes"       = "Cardiomyocytes",
  "Fibroblasts"          = "Fibroblasts",
  "Endothelial_cells"    = "Endothelial cells",
  "Pericytes"            = "Pericytes",
  "Smooth_muscle_cells"  = "Smooth muscle cells",
  "Macrophages"          = "Macrophages",
  "Neuronal_cells"       = "Neuronal cells",
  "T_lymphocytes"        = "T lymphocytes",
  "Endocardial_cells"    = "Endocardial cells"
)

# ---- Read all DEG count summaries ----
deg_summary_list <- lapply(cell_types, function(celltype) {
  file <- file.path(output_dir_data, paste0(celltype, "_n_de_genes.csv"))

  read.csv(file) %>%
    dplyr::select(metadata, n_tot) %>%
    dplyr::rename(parameter = metadata) %>%
    dplyr::mutate(cell_type = celltype)
})

deg_summary <- dplyr::bind_rows(deg_summary_list) %>%
  dplyr::filter(parameter %in% vars_keep) %>%
  dplyr::mutate(
    parameter = factor(parameter, levels = vars_keep),
    parameter_label = factor(
      pretty_vars[as.character(parameter)],
      levels = rev(pretty_vars[vars_keep])
    ),
    cell_type = factor(cell_type, levels = cell_types),
    cell_type_label = factor(
      pretty_celltypes[as.character(cell_type)],
      levels = pretty_celltypes[cell_types]
    )
  )

# ---- Make sure missing combinations are shown as 0 ----
deg_summary_plot <- deg_summary %>%
  tidyr::complete(
    parameter_label,
    cell_type_label,
    fill = list(n_tot = 0)
  )

ggplot(deg_summary_plot, aes(x = cell_type_label, y = parameter_label, fill = n_tot)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(
    aes(label = ifelse(n_tot == 0, "", n_tot)),
    size = 3.5,
    color = "black"
  ) +
  scale_fill_gradient(
    low = "grey95",
    high = "#A63A2A",
    limits = c(0, max(deg_summary_plot$n_tot, na.rm = TRUE))
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(
    axis.line = element_blank(),
    panel.border = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1,
      face = "bold",
      size = 11,
      color = "black"
    ),
    axis.text.y = element_text(
      face = "bold",
      size = 12,
      color = "black"
    ),
    legend.position = "bottom",
    legend.direction = "horizontal",
    plot.margin = margin(t = 5, r = 6, b = 5, l = 5)
  ) +
  guides(
    fill = guide_colorbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth = unit(4.0, "cm"),
      barheight = unit(0.35, "cm")
    )
  )
```

```{r}
deg_scale_max <- max(unlist(lapply(cell_types, function(celltype) {
  file <- file.path(output_dir_data, paste0(celltype, "_n_de_genes.csv"))

  x <- read.csv(file)

  c(
    x$n_upregulated[x$metadata %in% vars_keep],
    x$n_downregulated[x$metadata %in% vars_keep]
  )
})), na.rm = TRUE)
```

```{r deg_count_heatmap_simple_mainfig_narrow_upregulated, fig.width=5, fig.height=5.4, dpi=300}
cell_types <- c(
  "Cardiomyocytes",
  "Fibroblasts",
  "Endothelial_cells",
  "Pericytes",
  "Smooth_muscle_cells",
  "Macrophages",
  "Neuronal_cells",
  "T_lymphocytes",
  "Endocardial_cells"
)

vars_keep <- c(
  "mPAP_mmHg",
  "sPAP_mmHg",
  "PVR_WU",
  "PCW_mmHg",
  "TAPSE_mm",
  "RVIT_mm",
  "LVEF_percent",
  "LVEDD_mm"
)

pretty_vars <- c(
  "mPAP_mmHg"    = "mPAP (mmHg)",
  "sPAP_mmHg"    = "sPAP (mmHg)",
  "PVR_WU"       = "PVR (WU)",
  "PCW_mmHg"     = "PCWP (mmHg)",
  "TAPSE_mm"     = "TAPSE (mm)",
  "RVIT_mm"      = "RVIT (mm)",
  "LVEF_percent" = "LVEF (%)",
  "LVEDD_mm"     = "LVEDD (mm)"
)

pretty_celltypes <- c(
  "Cardiomyocytes"       = "Cardiomyocytes",
  "Fibroblasts"          = "Fibroblasts",
  "Endothelial_cells"    = "Endothelial cells",
  "Pericytes"            = "Pericytes",
  "Smooth_muscle_cells"  = "Smooth muscle cells",
  "Macrophages"          = "Macrophages",
  "Neuronal_cells"       = "Neuronal cells",
  "T_lymphocytes"        = "T lymphocytes",
  "Endocardial_cells"    = "Endocardial cells"
)

deg_summary_list <- lapply(cell_types, function(celltype) {
  file <- file.path(output_dir_data, paste0(celltype, "_n_de_genes.csv"))

  read.csv(file) %>%
    dplyr::select(metadata, n_upregulated) %>%
    dplyr::rename(parameter = metadata, n_tot = n_upregulated) %>%
    dplyr::mutate(cell_type = celltype)
})

deg_summary <- dplyr::bind_rows(deg_summary_list) %>%
  dplyr::filter(parameter %in% vars_keep) %>%
  dplyr::mutate(
    parameter = factor(parameter, levels = vars_keep),
    parameter_label = factor(
      pretty_vars[as.character(parameter)],
      levels = rev(pretty_vars[vars_keep])
    ),
    cell_type = factor(cell_type, levels = cell_types),
    cell_type_label = factor(
      pretty_celltypes[as.character(cell_type)],
      levels = pretty_celltypes[cell_types]
    )
  )

deg_summary_plot <- deg_summary %>%
  tidyr::complete(
    parameter_label,
    cell_type_label,
    fill = list(n_tot = 0)
  )

ggplot(deg_summary_plot, aes(x = cell_type_label, y = parameter_label, fill = n_tot)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(
    aes(label = ifelse(n_tot == 0, "", n_tot)),
    size = 3.5,
    color = "black"
  ) +
  scale_fill_gradient(
    low = "grey95",
    high = "#A63A2A",
    limits = c(0, deg_scale_max)
  ) +
  labs(
    title = "Upregulated genes",
    x = NULL,
    y = NULL,
    fill = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.line = element_blank(),
    panel.border = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1,
      face = "bold",
      size = 11,
      color = "black"
    ),
    axis.text.y = element_text(
      face = "bold",
      size = 12,
      color = "black"
    ),
    legend.position = "bottom",
    legend.direction = "horizontal",
    plot.margin = margin(t = 5, r = 6, b = 5, l = 5)
  ) +
  guides(
    fill = guide_colorbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth = unit(4.0, "cm"),
      barheight = unit(0.35, "cm")
    )
  )
```

```{r deg_count_heatmap_simple_mainfig_narrow_downregulated, fig.width=5, fig.height=5.4, dpi=300}
deg_summary_list <- lapply(cell_types, function(celltype) {
  file <- file.path(output_dir_data, paste0(celltype, "_n_de_genes.csv"))

  read.csv(file) %>%
    dplyr::select(metadata, n_downregulated) %>%
    dplyr::rename(parameter = metadata, n_tot = n_downregulated) %>%
    dplyr::mutate(cell_type = celltype)
})

deg_summary <- dplyr::bind_rows(deg_summary_list) %>%
  dplyr::filter(parameter %in% vars_keep) %>%
  dplyr::mutate(
    parameter = factor(parameter, levels = vars_keep),
    parameter_label = factor(
      pretty_vars[as.character(parameter)],
      levels = rev(pretty_vars[vars_keep])
    ),
    cell_type = factor(cell_type, levels = cell_types),
    cell_type_label = factor(
      pretty_celltypes[as.character(cell_type)],
      levels = pretty_celltypes[cell_types]
    )
  )

deg_summary_plot <- deg_summary %>%
  tidyr::complete(
    parameter_label,
    cell_type_label,
    fill = list(n_tot = 0)
  )

ggplot(deg_summary_plot, aes(x = cell_type_label, y = parameter_label, fill = n_tot)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(
    aes(label = ifelse(n_tot == 0, "", n_tot)),
    size = 3.5,
    color = "black"
  ) +
  scale_fill_gradient(
    low = "grey95",
    high = "#A63A2A",
    limits = c(0, deg_scale_max)
  ) +
  labs(
    title = "Downregulated genes",
    x = NULL,
    y = NULL,
    fill = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.line = element_blank(),
    panel.border = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1,
      face = "bold",
      size = 11,
      color = "black"
    ),
    axis.text.y = element_text(
      face = "bold",
      size = 12,
      color = "black"
    ),
    legend.position = "bottom",
    legend.direction = "horizontal",
    plot.margin = margin(t = 5, r = 6, b = 5, l = 5)
  ) +
  guides(
    fill = guide_colorbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth = unit(4.0, "cm"),
      barheight = unit(0.35, "cm")
    )
  )
```

## Cardiomyocyte threshold-based differential expression


```{r Cardiomyocytes_threshold_DE_setup}
if (!"TAPSE_sPAP" %in% colnames(metadata_DCM)) {
  metadata_DCM$TAPSE_sPAP <- metadata_DCM$TAPSE_mm / metadata_DCM$sPAP_mmHg
}

cm_threshold_contrasts <- list(
  mPAP_ge35 = list(
    variable = "mPAP_mmHg",
    test_fun = function(x) x >= 35,
    test_label = "mPAP >=35",
    reference_label = "<35"
  ),
  TAPSE_lt16 = list(
    variable = "TAPSE_mm",
    test_fun = function(x) x < 16,
    test_label = "TAPSE <16",
    reference_label = ">=16"
  ),
  RVIT_ge42 = list(
    variable = "RVIT_mm",
    test_fun = function(x) x >= 42,
    test_label = "RVIT >=42",
    reference_label = "<42"
  ),
  TAPSE_sPAP_le032 = list(
    variable = "TAPSE_sPAP",
    test_fun = function(x) x <= 0.32,
    test_label = "TAPSE/sPAP <=0.32",
    reference_label = ">0.32"
  )
)

# ---- Read cardiomyocyte pseudobulk object generated above ----
cm_pseudocounts <- readRDS(here::here(output_dir_data, "Cardiomyocytes_pseudocounts.rds"))

cm_percent_stats <- read.csv(here::here(output_dir_data, "Cardiomyocytes_percent_stats.csv"))
rownames(cm_percent_stats) <- cm_percent_stats$gene

cm_samples <- intersect(metadata_DCM$Sample, colnames(cm_pseudocounts$counts))
metadata_DCM_cm <- metadata_DCM %>%
  dplyr::filter(Sample %in% cm_samples) %>%
  dplyr::arrange(match(Sample, cm_samples))

cm_pseudocounts_DCM <- cm_pseudocounts[, metadata_DCM_cm$Sample]
all.equal(metadata_DCM_cm$Sample, colnames(cm_pseudocounts_DCM$counts))
```

```{r Cardiomyocytes_threshold_DE}
cm_threshold_results <- list()
cm_threshold_signif <- list()
cm_threshold_volcano <- list()
cm_threshold_n_de_genes <- data.frame()

for (contrast_name in names(cm_threshold_contrasts)) {

  contrast_info <- cm_threshold_contrasts[[contrast_name]]
  variable <- contrast_info$variable

  metadata_subset <- metadata_DCM_cm %>%
    dplyr::filter(!is.na(.data[[variable]]), !is.na(Age_years)) %>%
    dplyr::mutate(
      DE_group = ifelse(contrast_info$test_fun(.data[[variable]]), "test", "reference"),
      DE_group = factor(DE_group, levels = c("reference", "test"))
    )

  n_test <- sum(metadata_subset$DE_group == "test")
  n_reference <- sum(metadata_subset$DE_group == "reference")

  contrast_label <- paste0(
    contrast_info$test_label, " (n=", n_test, ") vs ",
    contrast_info$reference_label, " (n=", n_reference, ")"
  )

  pseudocounts_subset <- cm_pseudocounts_DCM[, metadata_subset$Sample]
  pseudocounts_subset <- normLibSizes(pseudocounts_subset)

  design <- model.matrix(~ DE_group + Age_years, data = metadata_subset)

  pseudocounts_subset <- estimateDisp(pseudocounts_subset, design)

  fit <- glmQLFit(pseudocounts_subset, design, robust = TRUE)
  fit <- glmQLFTest(fit, coef = "DE_grouptest")

  res <- topTags(fit, n = Inf)$table

  if (!"gene" %in% colnames(res)) {
    res <- res %>%
      tibble::rownames_to_column("gene")
  }

  res <- res %>%
    dplyr::left_join(cm_percent_stats[, c("gene", "percentCells")], by = "gene") %>%
    dplyr::arrange(FDR)

  sig <- res %>%
    dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>%
    dplyr::arrange(FDR)

  cm_threshold_results[[contrast_name]] <- res
  cm_threshold_signif[[contrast_name]] <- sig

  write.csv(
    res,
    file = here::here(output_dir_data, paste0("Cardiomyocytes_DE_Results_threshold_", contrast_name, ".csv")),
    quote = FALSE,
    row.names = FALSE
  )

  write.csv(
    sig,
    file = here::here(output_dir_data, paste0("Cardiomyocytes_DE_Significant_threshold_", contrast_name, ".csv")),
    quote = FALSE,
    row.names = FALSE
  )

  cm_threshold_n_de_genes_temp <- data.frame(
    contrast = contrast_name,
    variable = variable,
    comparison = contrast_label,
    n_test = n_test,
    n_reference = n_reference,
    n_upregulated = sum(sig$logFC > 0.5),
    n_downregulated = sum(sig$logFC < -0.5),
    n_tot = nrow(sig)
  )

  cm_threshold_n_de_genes <- rbind(cm_threshold_n_de_genes, cm_threshold_n_de_genes_temp)

  keyvals <- rep("grey", nrow(res))
  names(keyvals) <- rep("Not significant", nrow(res))

  keyvals[res$logFC > 0.5 & res$FDR < 0.05] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"

  keyvals[res$logFC < -0.5 & res$FDR < 0.05] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"

  cm_threshold_volcano[[contrast_name]] <- EnhancedVolcano(
    res,
    lab = res$gene,
    x = "logFC",
    y = "FDR",
    labSize = 0,
    titleLabSize = 13,
    subtitleLabSize = 8,
    axisLabSize = 12,
    captionLabSize = 9,
    pointSize = 0.5,
    FCcutoff = 0.5,
    pCutoff = 0.05,
    ylim = c(0, 3.5),
    colCustom = keyvals,
    colAlpha = 1,
    drawConnectors = FALSE,
    title = "Cardiomyocytes",
    subtitle = contrast_label
  ) +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, size = 8)
    )

  ggsave(
    filename = here::here(output_dir_figs, paste0("Cardiomyocytes_volcano_threshold_", contrast_name, ".png")),
    plot = cm_threshold_volcano[[contrast_name]],
    width = 3,
    height = 4.5,
    dpi = 600
  )
}

write.csv(
  cm_threshold_n_de_genes,
  file = here::here(output_dir_data, "Cardiomyocytes_threshold_n_de_genes.csv"),
  quote = FALSE,
  row.names = FALSE
)

reactable(
  cm_threshold_n_de_genes,
  filterable = TRUE,
  searchable = TRUE,
  showPageSizeOptions = TRUE
)

rm(
  contrast_name, contrast_info, variable, metadata_subset, n_test, n_reference,
  contrast_label, pseudocounts_subset, design, fit, res, sig, keyvals,
  cm_threshold_n_de_genes_temp
)
```

```{r Cardiomyocytes_threshold_volcano_mPAP_ge35, fig.width=3, fig.height=4.5, dpi=300}
cm_threshold_volcano[["mPAP_ge35"]]
```

```{r Cardiomyocytes_threshold_volcano_TAPSE_lt16, fig.width=3, fig.height=4.5, dpi=300}
cm_threshold_volcano[["TAPSE_lt16"]]
```

```{r Cardiomyocytes_threshold_volcano_RVIT_ge42, fig.width=3, fig.height=4.5, dpi=300}
cm_threshold_volcano[["RVIT_ge42"]]
```

```{r Cardiomyocytes_threshold_volcano_TAPSE_sPAP_le032, fig.width=3, fig.height=4.5, dpi=300}
cm_threshold_volcano[["TAPSE_sPAP_le032"]]
```

## mPAP threshold differential expression in all cell types

Repeat the mPAP >=35 versus <35 comparison in all cell types included in the continuous differential expression analysis. The cardiomyocyte analysis above is kept unchanged and is used for the figures.

```{r mPAP_threshold_all_cell_types, eval=TRUE}
threshold_cell_types <- c(
  "Fibroblasts",
  "Endothelial_cells",
  "Pericytes",
  "Macrophages",
  "T_lymphocytes",
  "Smooth_muscle_cells",
  "Neuronal_cells",
  "Endocardial_cells"
)

mPAP_threshold_all_summary <- cm_threshold_n_de_genes %>%
  dplyr::filter(contrast == "mPAP_ge35") %>%
  dplyr::mutate(cell_type = "Cardiomyocytes", .before = 1)

for (celltype in threshold_cell_types) {
  threshold_pseudocounts <- readRDS(
    here::here(output_dir_data, paste0(celltype, "_pseudocounts.rds"))
  )
  threshold_percent_stats <- read.csv(
    here::here(output_dir_data, paste0(celltype, "_percent_stats.csv"))
  )

  threshold_samples <- intersect(
    metadata_DCM$Sample,
    colnames(threshold_pseudocounts$counts)
  )

  threshold_metadata <- metadata_DCM %>%
    dplyr::filter(
      Sample %in% threshold_samples,
      !is.na(mPAP_mmHg),
      !is.na(Age_years)
    ) %>%
    dplyr::arrange(match(Sample, threshold_samples)) %>%
    dplyr::mutate(
      DE_group = ifelse(mPAP_mmHg >= 35, "test", "reference"),
      DE_group = factor(DE_group, levels = c("reference", "test"))
    )

  n_test <- sum(threshold_metadata$DE_group == "test")
  n_reference <- sum(threshold_metadata$DE_group == "reference")

  if (n_test < 2L | n_reference < 2L) {
    stop(
      "Fewer than two samples are available in one of the mPAP threshold groups for ",
      celltype,
      "."
    )
  }

  threshold_pseudocounts_subset <- threshold_pseudocounts[, threshold_metadata$Sample]
  threshold_pseudocounts_subset <- normLibSizes(threshold_pseudocounts_subset)

  threshold_design <- model.matrix(~ DE_group + Age_years, data = threshold_metadata)
  threshold_pseudocounts_subset <- estimateDisp(threshold_pseudocounts_subset, threshold_design)

  threshold_fit <- glmQLFit(threshold_pseudocounts_subset, threshold_design, robust = TRUE)
  threshold_fit <- glmQLFTest(threshold_fit, coef = "DE_grouptest")

  threshold_results <- topTags(threshold_fit, n = Inf)$table

  if (!"gene" %in% colnames(threshold_results)) {
    threshold_results <- threshold_results %>%
      tibble::rownames_to_column("gene")
  }

  threshold_results <- threshold_results %>%
    dplyr::left_join(
      threshold_percent_stats[, c("gene", "percentCells")],
      by = "gene"
    ) %>%
    dplyr::arrange(FDR)

  threshold_significant <- threshold_results %>%
    dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>%
    dplyr::arrange(FDR)

  write.csv(
    threshold_results,
    file = here::here(
      output_dir_data,
      paste0(celltype, "_DE_Results_threshold_mPAP_ge35.csv")
    ),
    quote = FALSE,
    row.names = FALSE
  )

  write.csv(
    threshold_significant,
    file = here::here(
      output_dir_data,
      paste0(celltype, "_DE_Significant_threshold_mPAP_ge35.csv")
    ),
    quote = FALSE,
    row.names = FALSE
  )

  threshold_summary <- data.frame(
    cell_type = celltype,
    contrast = "mPAP_ge35",
    variable = "mPAP_mmHg",
    comparison = paste0(
      "mPAP >=35 (n=", n_test, ") vs <35 (n=", n_reference, ")"
    ),
    n_test = n_test,
    n_reference = n_reference,
    n_upregulated = sum(threshold_significant$logFC > 0.5),
    n_downregulated = sum(threshold_significant$logFC < -0.5),
    n_tot = nrow(threshold_significant)
  )

  mPAP_threshold_all_summary <- rbind(
    mPAP_threshold_all_summary,
    threshold_summary
  )
}

write.csv(
  mPAP_threshold_all_summary,
  file = here::here(output_dir_data, "mPAP_ge35_all_cell_types_summary.csv"),
  quote = FALSE,
  row.names = FALSE
)

reactable(
  mPAP_threshold_all_summary,
  filterable = TRUE,
  searchable = TRUE,
  showPageSizeOptions = TRUE
)

rm(
  celltype, threshold_pseudocounts, threshold_percent_stats,
  threshold_samples, threshold_metadata, n_test, n_reference,
  threshold_pseudocounts_subset, threshold_design, threshold_fit,
  threshold_results, threshold_significant, threshold_summary
)
```

## mPAP linear versus mPAP threshold comparison

```{r Cardiomyocytes_mPAP_linear_threshold_prepare}
cm_mpap_linear <- read.csv(
  here::here(output_dir_data, "Cardiomyocytes_DE_Results_mPAP_mmHg.csv")
)

cm_mpap_threshold <- cm_threshold_results[["mPAP_ge35"]]

cm_mpap_linear_sig <- cm_mpap_linear %>%
  dplyr::filter(FDR < 0.05, abs(logFC) > 0.5)

cm_mpap_threshold_sig <- cm_mpap_threshold %>%
  dplyr::filter(FDR < 0.05, abs(logFC) > 0.5)

cm_mpap_gene_sets <- list(
  "Linear mPAP up" = cm_mpap_linear_sig %>%
    dplyr::filter(logFC > 0.5) %>%
    dplyr::pull(gene),
  "mPAP >=35 up" = cm_mpap_threshold_sig %>%
    dplyr::filter(logFC > 0.5) %>%
    dplyr::pull(gene),
  "Linear mPAP down" = cm_mpap_linear_sig %>%
    dplyr::filter(logFC < -0.5) %>%
    dplyr::pull(gene),
  "mPAP >=35 down" = cm_mpap_threshold_sig %>%
    dplyr::filter(logFC < -0.5) %>%
    dplyr::pull(gene)
)

cm_mpap_overlap_table <- tibble(
  gene = unique(unlist(cm_mpap_gene_sets)),
  linear_mPAP_up = gene %in% cm_mpap_gene_sets[["Linear mPAP up"]],
  threshold_mPAP_ge35_up = gene %in% cm_mpap_gene_sets[["mPAP >=35 up"]],
  linear_mPAP_down = gene %in% cm_mpap_gene_sets[["Linear mPAP down"]],
  threshold_mPAP_ge35_down = gene %in% cm_mpap_gene_sets[["mPAP >=35 down"]]
)

write.csv(
  cm_mpap_overlap_table,
  file = here::here(output_dir_data, "Cardiomyocytes_mPAP_linear_vs_threshold_overlap_genes.csv"),
  quote = FALSE,
  row.names = FALSE
)

cm_mpap_overlap_summary <- tibble(
  comparison = c(
    "Linear up and threshold up",
    "Linear down and threshold down",
    "Linear up and threshold down",
    "Linear down and threshold up"
  ),
  n_genes = c(
    sum(cm_mpap_overlap_table$linear_mPAP_up & cm_mpap_overlap_table$threshold_mPAP_ge35_up),
    sum(cm_mpap_overlap_table$linear_mPAP_down & cm_mpap_overlap_table$threshold_mPAP_ge35_down),
    sum(cm_mpap_overlap_table$linear_mPAP_up & cm_mpap_overlap_table$threshold_mPAP_ge35_down),
    sum(cm_mpap_overlap_table$linear_mPAP_down & cm_mpap_overlap_table$threshold_mPAP_ge35_up)
  )
)

write.csv(
  cm_mpap_overlap_summary,
  file = here::here(output_dir_data, "Cardiomyocytes_mPAP_linear_vs_threshold_overlap_summary.csv"),
  quote = FALSE,
  row.names = FALSE
)

cm_mpap_compare <- cm_mpap_linear %>%
  dplyr::select(gene, logFC_linear = logFC, FDR_linear = FDR, PValue_linear = PValue) %>%
  dplyr::inner_join(
    cm_mpap_threshold %>%
      dplyr::select(gene, logFC_threshold = logFC, FDR_threshold = FDR, PValue_threshold = PValue),
    by = "gene"
  ) %>%
  dplyr::mutate(
    linear_significant = FDR_linear < 0.05 & abs(logFC_linear) > 0.5,
    threshold_significant = FDR_threshold < 0.05 & abs(logFC_threshold) > 0.5,
    significance_group = dplyr::case_when(
      linear_significant & threshold_significant ~ "Both",
      linear_significant & !threshold_significant ~ "Linear only",
      !linear_significant & threshold_significant ~ "mPAP >=35 only",
      TRUE ~ "Not significant"
    ),
    neglog10FDR_linear = -log10(pmax(FDR_linear, .Machine$double.xmin)),
    neglog10FDR_threshold = -log10(pmax(FDR_threshold, .Machine$double.xmin)),
    neglog10FDR_linear = pmin(neglog10FDR_linear, 50),
    neglog10FDR_threshold = pmin(neglog10FDR_threshold, 50)
  )

write.csv(
  cm_mpap_compare,
  file = here::here(output_dir_data, "Cardiomyocytes_mPAP_linear_vs_threshold_all_genes.csv"),
  quote = FALSE,
  row.names = FALSE
)

cm_mpap_logFC_cor <- cor(
  cm_mpap_compare$logFC_linear,
  cm_mpap_compare$logFC_threshold,
  use = "complete.obs",
  method = "pearson"
)

cm_mpap_FDR_cor <- cor(
  cm_mpap_compare$neglog10FDR_linear,
  cm_mpap_compare$neglog10FDR_threshold,
  use = "complete.obs",
  method = "pearson"
)
```

```{r Cardiomyocytes_mPAP_linear_threshold_upset, fig.width=4.35, fig.height=4.5, dpi=300}
cm_mpap_comb_mat <- ComplexHeatmap::make_comb_mat(cm_mpap_gene_sets, mode = "intersect")

# Keep only real overlaps, so the top bars do not repeat the single-set sizes shown on the right.
cm_mpap_comb_mat_overlap <- cm_mpap_comb_mat[ComplexHeatmap::comb_degree(cm_mpap_comb_mat) >= 2]

ComplexHeatmap::UpSet(
  cm_mpap_comb_mat_overlap,
  column_title = "Cardiomyocytes: mPAP linear vs mPAP >=35",
  comb_order = order(ComplexHeatmap::comb_size(cm_mpap_comb_mat_overlap), decreasing = TRUE),
  top_annotation = ComplexHeatmap::upset_top_annotation(
    cm_mpap_comb_mat_overlap,
    add_numbers = TRUE
  ),
  right_annotation = ComplexHeatmap::upset_right_annotation(
    cm_mpap_comb_mat_overlap,
    add_numbers = TRUE
  )
)
```

```{r Cardiomyocytes_mPAP_logFC_correlation, fig.width=4, fig.height=4, dpi=300}
cm_mpap_logFC_plot <- ggplot(
  cm_mpap_compare,
  aes(x = logFC_linear, y = logFC_threshold)
) +
  coord_equal(xlim = c(-6, 6), ylim = c(-6, 6)) +
  scale_x_continuous(breaks = c(-6, -3, 0, 3, 6)) +
  scale_y_continuous(breaks = c(-6, -3, 0, 3, 6)) +
  geom_point(aes(color = significance_group), size = 0.7, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.5) +
  scale_color_manual(
    values = c(
      "Both" = "black",
      "Linear only" = "#004C8C",
      "mPAP >=35 only" = "#A63A2A",
      "Not significant" = "grey75"
    )
  ) +
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    hjust = -0.05,
    vjust = 1.2,
    size = 3.5,
    label = paste0("Pearson r = ", round(cm_mpap_logFC_cor, 2))
  ) +
  theme_classic() +
  theme(
    text = element_text(size = 12),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Cardiomyocytes",
    x = "mPAP linear log2FC",
    y = "mPAP >=35 log2FC"
  )

cm_mpap_logFC_plot

ggsave(
  filename = here::here(output_dir_figs, "Cardiomyocytes_mPAP_linear_vs_threshold_logFC_correlation.png"),
  plot = cm_mpap_logFC_plot,
  width = 4,
  height = 4,
  dpi = 600
)
```

```{r Cardiomyocytes_mPAP_FDR_correlation, fig.width=4, fig.height=4, dpi=300}
cm_mpap_FDR_plot <- ggplot(
  cm_mpap_compare,
  aes(x = neglog10FDR_linear, y = neglog10FDR_threshold)
) +
  geom_point(aes(color = significance_group), size = 0.7, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 0.4) +
  scale_color_manual(
    values = c(
      "Both" = "black",
      "Linear only" = "#004C8C",
      "mPAP >=35 only" = "#A63A2A",
      "Not significant" = "grey75"
    )
  ) +
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    hjust = -0.05,
    vjust = 1.2,
    size = 3.5,
    label = paste0("Pearson r = ", round(cm_mpap_FDR_cor, 2))
  ) +
  theme_classic() +
  theme(
    text = element_text(size = 12),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Cardiomyocytes",
    x = expression("mPAP linear " * -log[10] * " FDR"),
    y = expression("mPAP >=35 " * -log[10] * " FDR")
  )

cm_mpap_FDR_plot

ggsave(
  filename = here::here(output_dir_figs, "Cardiomyocytes_mPAP_linear_vs_threshold_FDR_correlation.png"),
  plot = cm_mpap_FDR_plot,
  width = 4,
  height = 4,
  dpi = 600
)
```

## RV compensation and decompensation

Patients with mPAP >20 mmHg were classified as compensated or
decompensated using a TAPSE/sPAP threshold of 0.36.

```{r RV_state_definition_prepare}
required_rv_state_columns <- c(
  "Sample",
  "Age_years",
  "mPAP_mmHg",
  "sPAP_mmHg",
  "TAPSE_mm"
)

missing_rv_state_columns <- setdiff(
  required_rv_state_columns,
  colnames(metadata_DCM)
)

if (length(missing_rv_state_columns) > 0L) {
  stop(
    "Missing metadata columns: ",
    paste(
      missing_rv_state_columns,
      collapse = ", "
    )
  )
}

rv_state_metadata <- metadata_DCM %>%
  dplyr::select(
    dplyr::all_of(
      required_rv_state_columns
    )
  ) %>%
  dplyr::distinct() %>%
  dplyr::mutate(
    TAPSE_sPAP =
      TAPSE_mm /
      sPAP_mmHg
  ) %>%
  dplyr::filter(
    is.finite(mPAP_mmHg),
    is.finite(TAPSE_sPAP)
  ) %>%
  dplyr::mutate(
    RV_state = dplyr::case_when(
      mPAP_mmHg <= 20 ~
        "Not overloaded",
      TAPSE_sPAP > 0.36 ~
        "Compensated",
      TAPSE_sPAP <= 0.36 ~
        "Decompensated",
      TRUE ~ NA_character_
    ),
    RV_state = factor(
      RV_state,
      levels = c(
        "Not overloaded",
        "Compensated",
        "Decompensated"
      )
    )
  ) %>%
  dplyr::arrange(Sample)

stopifnot(
  !anyDuplicated(
    rv_state_metadata$Sample
  )
)

rv_state_counts <- rv_state_metadata %>%
  dplyr::count(
    RV_state,
    .drop = FALSE
  )

rv_state_labels <- stats::setNames(
  paste0(
    as.character(
      rv_state_counts$RV_state
    ),
    " (n=",
    rv_state_counts$n,
    ")"
  ),
  as.character(
    rv_state_counts$RV_state
  )
)

rv_state_cols <- c(
  "Not overloaded" = "#BDBDBD",
  "Compensated" = "#E38C82",
  "Decompensated" = "#9E3434"
)

write.csv(
  rv_state_metadata,
  file = here::here(
    output_dir_data,
    paste0(
      "RV_state_assignments_",
      "mPAP20_TAPSEsPAP036.csv"
    )
  ),
  quote = FALSE,
  row.names = FALSE
)

reactable(
  rv_state_metadata,
  filterable = TRUE,
  searchable = TRUE,
  showPageSizeOptions = TRUE
)
```

```{r RV_state_definition_plot, fig.width=4, fig.height=5.3, dpi=300}
rv_state_plot <- ggplot(
  rv_state_metadata,
  aes(
    x = mPAP_mmHg,
    y = TAPSE_sPAP,
    fill = RV_state
  )
) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = 20,
    ymin = -Inf,
    ymax = Inf,
    fill = rv_state_cols[
      "Not overloaded"
    ],
    alpha = 0.12
  ) +
  annotate(
    "rect",
    xmin = 20,
    xmax = Inf,
    ymin = 0.36,
    ymax = Inf,
    fill = rv_state_cols[
      "Compensated"
    ],
    alpha = 0.10
  ) +
  annotate(
    "rect",
    xmin = 20,
    xmax = Inf,
    ymin = -Inf,
    ymax = 0.36,
    fill = rv_state_cols[
      "Decompensated"
    ],
    alpha = 0.08
  ) +
  geom_vline(
    xintercept = 20,
    linetype = "dashed",
    linewidth = 0.5
  ) +
  geom_hline(
    yintercept = 0.36,
    linetype = "dashed",
    linewidth = 0.5
  ) +
  geom_point(
    shape = 21,
    size = 3,
    stroke = 0.4,
    color = "black"
  ) +
  scale_fill_manual(
    values = rv_state_cols,
    breaks = levels(
      rv_state_metadata$RV_state
    ),
    labels = rv_state_labels[
      levels(
        rv_state_metadata$RV_state
      )
    ],
    drop = FALSE
  ) +
  labs(
    title = "RV state definition",
    x = "mPAP (mmHg)",
    y = "TAPSE/sPAP (mm/mmHg)",
    fill = NULL
  ) +
theme_classic(
  base_size = 12
) +
theme(
  plot.title = element_text(
    hjust = 0.5,
    face = "bold",
    size = 13
  ),
  legend.position = "bottom",
  legend.direction = "vertical",
  legend.justification = "center",
  legend.box = "vertical",
  legend.text = element_text(
    size = 9
  ),
  legend.spacing.y = unit(
    0.05,
    "cm"
  ),
  legend.margin = margin(
    t = 4,
    r = 0,
    b = 0,
    l = 0
  )
) +
guides(
  fill = guide_legend(
    ncol = 1,
    byrow = TRUE,
    title = NULL,
    override.aes = list(
      shape = 21,
      size = 3,
      color = "black"
    )
  )
)

rv_state_plot

ggsave(
  filename = here::here(
    output_dir_figs,
    paste0(
      "RV_state_definition_",
      "mPAP20_TAPSEsPAP036.png"
    )
  ),
  plot = rv_state_plot,
  width = 3.6,
  height = 4.5,
  dpi = 600
)
```

## Cardiomyocyte decompensation differential expression

Positive log2 fold changes indicate higher expression in decompensated
than compensated RVs.

```{r Cardiomyocytes_decompensation_DE}
cm_decomp_metadata <- rv_state_metadata %>%
  dplyr::filter(
    RV_state %in% c(
      "Compensated",
      "Decompensated"
    ),
    !is.na(Age_years)
  ) %>%
  dplyr::mutate(
    RV_state = factor(
      as.character(RV_state),
      levels = c(
        "Compensated",
        "Decompensated"
      )
    )
  )

cm_decomp_pseudocounts <- readRDS(
  here::here(
    output_dir_data,
    "Cardiomyocytes_pseudocounts.rds"
  )
)

cm_decomp_percent_stats <- read.csv(
  here::here(
    output_dir_data,
    "Cardiomyocytes_percent_stats.csv"
  )
)

cm_decomp_samples <- intersect(
  cm_decomp_metadata$Sample,
  colnames(
    cm_decomp_pseudocounts$counts
  )
)

cm_decomp_metadata <- cm_decomp_metadata %>%
  dplyr::filter(
    Sample %in% cm_decomp_samples
  ) %>%
  dplyr::arrange(
    match(
      Sample,
      cm_decomp_samples
    )
  )

cm_decomp_pseudocounts <-
  cm_decomp_pseudocounts[
    ,
    cm_decomp_metadata$Sample
  ]

stopifnot(
  identical(
    cm_decomp_metadata$Sample,
    colnames(
      cm_decomp_pseudocounts$counts
    )
  )
)

cm_decomp_group_counts <- table(
  cm_decomp_metadata$RV_state
)

if (
  any(
    cm_decomp_group_counts < 2L
  )
) {
  stop(
    "Fewer than two samples are available ",
    "in one of the RV-state groups."
  )
}

cm_decomp_pseudocounts <- normLibSizes(
  cm_decomp_pseudocounts
)

cm_decomp_design <- model.matrix(
  ~ RV_state + Age_years,
  data = cm_decomp_metadata
)

if (
  qr(cm_decomp_design)$rank <
    ncol(cm_decomp_design)
) {
  stop(
    "The compensated/decompensated ",
    "design matrix is not full rank."
  )
}

cm_decomp_pseudocounts <- estimateDisp(
  cm_decomp_pseudocounts,
  cm_decomp_design
)

cm_decomp_fit <- glmQLFit(
  cm_decomp_pseudocounts,
  cm_decomp_design,
  robust = TRUE
)

cm_decomp_test <- glmQLFTest(
  cm_decomp_fit,
  coef = "RV_stateDecompensated"
)

cm_decomp_results <- topTags(
  cm_decomp_test,
  n = Inf
)$table

if (
  !"gene" %in%
    colnames(cm_decomp_results)
) {
  cm_decomp_results <-
    cm_decomp_results %>%
    tibble::rownames_to_column(
      "gene"
    )
}

cm_decomp_results <- cm_decomp_results %>%
  dplyr::left_join(
    cm_decomp_percent_stats %>%
      dplyr::select(
        gene,
        percentCells
      ),
    by = "gene"
  ) %>%
  dplyr::arrange(FDR)

cm_decomp_significant <-
  cm_decomp_results %>%
  dplyr::filter(
    FDR < 0.05,
    abs(logFC) > 0.5
  )

n_compensated <- unname(
  cm_decomp_group_counts[
    "Compensated"
  ]
)

n_decompensated <- unname(
  cm_decomp_group_counts[
    "Decompensated"
  ]
)

cm_decomp_summary <- data.frame(
  cell_type = "Cardiomyocytes",
  n_compensated = n_compensated,
  n_decompensated = n_decompensated,
  n_upregulated = sum(
    cm_decomp_significant$logFC > 0.5
  ),
  n_downregulated = sum(
    cm_decomp_significant$logFC < -0.5
  ),
  n_total = nrow(
    cm_decomp_significant
  ),
  minimum_FDR = min(
    cm_decomp_results$FDR,
    na.rm = TRUE
  )
)

write.csv(
  cm_decomp_results,
  file = here::here(
    output_dir_data,
    paste0(
      "Cardiomyocytes_DE_Results_",
      "Decompensated_vs_Compensated.csv"
    )
  ),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  cm_decomp_significant,
  file = here::here(
    output_dir_data,
    paste0(
      "Cardiomyocytes_DE_Significant_",
      "Decompensated_vs_Compensated.csv"
    )
  ),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  cm_decomp_summary,
  file = here::here(
    output_dir_data,
    paste0(
      "Cardiomyocytes_DE_Summary_",
      "Decompensated_vs_Compensated.csv"
    )
  ),
  quote = FALSE,
  row.names = FALSE
)

reactable(
  cm_decomp_summary
)
```

## Decompensation differential expression in all cell types

Repeat the compensated versus decompensated comparison in all cell types included in the continuous differential expression analysis. The cardiomyocyte analysis above is kept unchanged and is used for the figure.

```{r Decompensation_DE_all_cell_types, eval=TRUE}
decomp_cell_types <- c(
  "Fibroblasts",
  "Endothelial_cells",
  "Pericytes",
  "Macrophages",
  "T_lymphocytes",
  "Smooth_muscle_cells",
  "Neuronal_cells",
  "Endocardial_cells"
)

decomp_all_summary <- cm_decomp_summary

for (celltype in decomp_cell_types) {
  decomp_metadata <- rv_state_metadata %>%
    dplyr::filter(
      RV_state %in% c(
        "Compensated",
        "Decompensated"
      ),
      !is.na(Age_years)
    ) %>%
    dplyr::mutate(
      RV_state = factor(
        as.character(RV_state),
        levels = c(
          "Compensated",
          "Decompensated"
        )
      )
    )

  decomp_pseudocounts <- readRDS(
    here::here(output_dir_data, paste0(celltype, "_pseudocounts.rds"))
  )

  decomp_percent_stats <- read.csv(
    here::here(output_dir_data, paste0(celltype, "_percent_stats.csv"))
  )

  decomp_samples <- intersect(
    decomp_metadata$Sample,
    colnames(decomp_pseudocounts$counts)
  )

  decomp_metadata <- decomp_metadata %>%
    dplyr::filter(Sample %in% decomp_samples) %>%
    dplyr::arrange(match(Sample, decomp_samples))

  decomp_pseudocounts <- decomp_pseudocounts[, decomp_metadata$Sample]

  stopifnot(
    identical(
      decomp_metadata$Sample,
      colnames(decomp_pseudocounts$counts)
    )
  )

  decomp_group_counts <- table(decomp_metadata$RV_state)

  if (any(decomp_group_counts < 2L)) {
    stop(
      "Fewer than two samples are available in one of the RV-state groups for ",
      celltype,
      "."
    )
  }

  decomp_pseudocounts <- normLibSizes(decomp_pseudocounts)
  decomp_design <- model.matrix(~ RV_state + Age_years, data = decomp_metadata)

  if (qr(decomp_design)$rank < ncol(decomp_design)) {
    stop(
      "The compensated/decompensated design matrix is not full rank for ",
      celltype,
      "."
    )
  }

  decomp_pseudocounts <- estimateDisp(decomp_pseudocounts, decomp_design)
  decomp_fit <- glmQLFit(decomp_pseudocounts, decomp_design, robust = TRUE)
  decomp_test <- glmQLFTest(decomp_fit, coef = "RV_stateDecompensated")

  decomp_results <- topTags(decomp_test, n = Inf)$table

  if (!"gene" %in% colnames(decomp_results)) {
    decomp_results <- decomp_results %>%
      tibble::rownames_to_column("gene")
  }

  decomp_results <- decomp_results %>%
    dplyr::left_join(
      decomp_percent_stats %>%
        dplyr::select(gene, percentCells),
      by = "gene"
    ) %>%
    dplyr::arrange(FDR)

  decomp_significant <- decomp_results %>%
    dplyr::filter(
      FDR < 0.05,
      abs(logFC) > 0.5
    )

  n_compensated_celltype <- unname(
    decomp_group_counts["Compensated"]
  )
  n_decompensated_celltype <- unname(
    decomp_group_counts["Decompensated"]
  )

  decomp_summary <- data.frame(
    cell_type = celltype,
    n_compensated = n_compensated_celltype,
    n_decompensated = n_decompensated_celltype,
    n_upregulated = sum(decomp_significant$logFC > 0.5),
    n_downregulated = sum(decomp_significant$logFC < -0.5),
    n_total = nrow(decomp_significant),
    minimum_FDR = min(decomp_results$FDR, na.rm = TRUE)
  )

  write.csv(
    decomp_results,
    file = here::here(
      output_dir_data,
      paste0(
        celltype,
        "_DE_Results_Decompensated_vs_Compensated.csv"
      )
    ),
    quote = FALSE,
    row.names = FALSE
  )

  write.csv(
    decomp_significant,
    file = here::here(
      output_dir_data,
      paste0(
        celltype,
        "_DE_Significant_Decompensated_vs_Compensated.csv"
      )
    ),
    quote = FALSE,
    row.names = FALSE
  )

  write.csv(
    decomp_summary,
    file = here::here(
      output_dir_data,
      paste0(
        celltype,
        "_DE_Summary_Decompensated_vs_Compensated.csv"
      )
    ),
    quote = FALSE,
    row.names = FALSE
  )

  decomp_all_summary <- rbind(
    decomp_all_summary,
    decomp_summary
  )
}

write.csv(
  decomp_all_summary,
  file = here::here(output_dir_data, "Decompensated_vs_Compensated_all_cell_types_summary.csv"),
  quote = FALSE,
  row.names = FALSE
)

reactable(
  decomp_all_summary,
  filterable = TRUE,
  searchable = TRUE,
  showPageSizeOptions = TRUE
)

rm(
  celltype, decomp_metadata, decomp_pseudocounts,
  decomp_percent_stats, decomp_samples, decomp_group_counts,
  decomp_design, decomp_fit, decomp_test, decomp_results,
  decomp_significant, n_compensated_celltype,
  n_decompensated_celltype, decomp_summary
)
```

```{r Cardiomyocytes_decompensation_volcano_prepare}
cm_decomp_keyvals <- rep(
  "grey",
  nrow(cm_decomp_results)
)

names(cm_decomp_keyvals) <- rep(
  "Not significant",
  nrow(cm_decomp_results)
)

cm_decomp_keyvals[
  cm_decomp_results$logFC > 0.5 &
    cm_decomp_results$FDR < 0.05
] <- "#A63A2A"

names(cm_decomp_keyvals)[
  cm_decomp_keyvals == "#A63A2A"
] <- "Increased"

cm_decomp_keyvals[
  cm_decomp_results$logFC < -0.5 &
    cm_decomp_results$FDR < 0.05
] <- "#004C8C"

names(cm_decomp_keyvals)[
  cm_decomp_keyvals == "#004C8C"
] <- "Decreased"

cm_decomp_comparison_label <- paste0(
  "Decompensated (n=",
  n_decompensated,
  ") vs compensated (n=",
  n_compensated,
  ")"
)

cm_decomp_caption <- paste0(
  nrow(cm_decomp_significant),
  " genes at FDR <0.05 and ",
  "|log2FC| >0.5"
)

cm_decomp_volcano <- EnhancedVolcano(
  cm_decomp_results,
  lab = cm_decomp_results$gene,
  x = "logFC",
  y = "FDR",
  labSize = 0,
  titleLabSize = 13,
  subtitleLabSize = 8,
  captionLabSize = 8,
  axisLabSize = 12,
  pointSize = 0.5,
  FCcutoff = 0.5,
  pCutoff = 0.05,
  xlim = c(-5, 5),
  ylim = c(0, 3.5),
  colCustom = cm_decomp_keyvals,
  colAlpha = 1,
  drawConnectors = FALSE,
  title = "Cardiomyocytes",
  subtitle = cm_decomp_comparison_label,
  caption = cm_decomp_caption
) +
  theme(
    legend.position = "none",
    plot.title = element_text(
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 8
    ),
    plot.caption = element_text(
      hjust = 0.5,
      size = 8
    )
  )

ggsave(
  filename = here::here(
    output_dir_figs,
    paste0(
      "Cardiomyocytes_volcano_",
      "Decompensated_vs_Compensated.png"
    )
  ),
  plot = cm_decomp_volcano,
  width = 3,
  height = 4.5,
  dpi = 600
)
```

```{r Cardiomyocytes_decompensation_volcano, fig.width=3, fig.height=4.5, dpi=300}
cm_decomp_volcano
```

## Supplementary data export

```{r Supplementary_data_export, eval=TRUE}
supp_cell_types <- c(
  "Cardiomyocytes",
  "Fibroblasts",
  "Endothelial_cells",
  "Pericytes",
  "Macrophages",
  "T_lymphocytes",
  "Smooth_muscle_cells",
  "Neuronal_cells",
  "Endocardial_cells"
)

supp_parameters <- colnames(metadata_DCM)[7:ncol(metadata_DCM)]

supp_data <- list()

supp_data[["Differential_abundance"]] <- differential_abundance

supp_data[["Cell_type_counts"]] <- as.data.frame(props$Counts) %>%
  tibble::rownames_to_column("cell_type")

supp_data[["Cell_type_proportions"]] <- as.data.frame(props$Proportions) %>%
  tibble::rownames_to_column("cell_type")

for (celltype in supp_cell_types) {
  continuous_results <- list()

  for (parameter in supp_parameters) {
    result_file <- here::here(
      output_dir_data,
      paste0(celltype, "_DE_Results_", parameter, ".csv")
    )

    if (!file.exists(result_file)) {
      stop("Missing continuous DE result file: ", result_file)
    }

    result_temp <- read.csv(result_file) %>%
      dplyr::mutate(parameter = .env$parameter, .before = 1)

    continuous_results[[parameter]] <- result_temp
  }

  supp_data[[paste0("DE_", celltype)]] <- dplyr::bind_rows(continuous_results)
}

mPAP_threshold_results <- list()
decompensation_results <- list()

for (celltype in supp_cell_types) {
  mPAP_threshold_file <- here::here(
    output_dir_data,
    paste0(celltype, "_DE_Results_threshold_mPAP_ge35.csv")
  )

  if (!file.exists(mPAP_threshold_file)) {
    stop("Missing mPAP threshold DE result file: ", mPAP_threshold_file)
  }

  mPAP_threshold_results[[celltype]] <- read.csv(mPAP_threshold_file) %>%
    dplyr::mutate(cell_type = .env$celltype, .before = 1)

  decompensation_file <- here::here(
    output_dir_data,
    paste0(celltype, "_DE_Results_Decompensated_vs_Compensated.csv")
  )

  if (!file.exists(decompensation_file)) {
    stop("Missing decompensation DE result file: ", decompensation_file)
  }

  decompensation_results[[celltype]] <- read.csv(decompensation_file) %>%
    dplyr::mutate(cell_type = .env$celltype, .before = 1)
}

supp_data[["mPAP35_DE_all"]] <- dplyr::bind_rows(mPAP_threshold_results)
supp_data[["RV_state_DE_all"]] <- dplyr::bind_rows(decompensation_results)
supp_data[["RV_state_assignments"]] <- rv_state_metadata %>%
  dplyr::mutate(RV_state = as.character(RV_state))
supp_data[["mPAP_linear_threshold"]] <- cm_mpap_compare
supp_data[["mPAP_overlap_genes"]] <- cm_mpap_overlap_table
supp_data[["mPAP_overlap_summary"]] <- cm_mpap_overlap_summary

DEG_counts <- lapply(
  supp_cell_types,
  function(celltype) {
    read.csv(
      here::here(output_dir_data, paste0(celltype, "_n_de_genes.csv"))
    ) %>%
      dplyr::mutate(cell_type = .env$celltype, .before = 1)
  }
) %>%
  dplyr::bind_rows()

supp_data[["DEG_counts"]] <- DEG_counts
supp_data[["mPAP35_DEG_counts"]] <- mPAP_threshold_all_summary
supp_data[["RV_state_DEG_counts"]] <- decomp_all_summary

writexl::write_xlsx(
  supp_data,
  here::here(output_dir_data, "DA_DE_all_cell_types_SuppData.xlsx")
)
```

