---
title: "Differential expression analysis in all cell types using fibrosis"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "DE_Fibrosis_all_cell_types"

# Load libraries
library(here)
library(readr)
library(readxl)
library(xlsx)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(ArchR)
library(clustree)
library(harmony)
library(gprofiler2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationHub)
library(ReactomePA)
library(statmod)
library(edgeR)
library(speckle)
library(EnhancedVolcano)
library(decoupleR)
library(OmnipathR)
library(dorothea)
library(enrichplot)
library(png)
library(reactable)
library(UpSetR)
library(ComplexHeatmap)
library(rtracklayer)
library(cowplot)

#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```


## DE Fibrosis

Load seurat object

```{r read DCM_RV_integrated}
DCM_RV_integrated <- readRDS(here::here("output", "QC_integration_annotation_26", "DCM_RV_integrated.rds"))
```

```{r add fibrosis}
Fibrosis_data <- read_excel(here::here("data", "Massons_quantification.xlsx"))
colnames(Fibrosis_data)[3] <- "Fibrosis_percent"
Fibrosis_data <- Fibrosis_data %>%
  dplyr::filter(Sample %in% unique(DCM_RV_integrated$Sample))
Fibrosis_data_DCM <- dplyr::filter(Fibrosis_data, Group=="DCM")
add_metadata <- left_join(DCM_RV_integrated[["Sample"]], dplyr::select(Fibrosis_data, c("Sample", "Fibrosis_percent")))
row.names(add_metadata) <- row.names(DCM_RV_integrated[[]])
DCM_RV_integrated <- AddMetaData(DCM_RV_integrated, metadata = add_metadata)
rm(add_metadata)
table(DCM_RV_integrated$Fibrosis_percent)
```

```{r Speckle_all}
# Initialize an empty data frame for storing differential abundance results
differential_abundance <- data.frame()
props <- getTransformedProps(DCM_RV_integrated$cell_type, DCM_RV_integrated$Sample, transform="logit")

props$Counts <- props$Counts[, Fibrosis_data_DCM$Sample, drop = FALSE]
props$TransformedProps <- props$TransformedProps[, Fibrosis_data_DCM$Sample, drop = FALSE]
props$Proportions <- props$Proportions[, Fibrosis_data_DCM$Sample, drop = FALSE]
Fibrosis_data_DCM <- Fibrosis_data_DCM %>% 
    arrange(Sample, colnames(props$Proportions))
design <- model.matrix(~ Fibrosis_data_DCM[["Fibrosis_percent"]])
fit <- lmFit(props$TransformedProps, design)
fit <- eBayes(fit, robust=TRUE)
differential_abundance <- topTable(fit, n = Inf, coef = 2)
differential_abundance$cell_state <- rownames(differential_abundance)

rm(props, fit)

write.csv(differential_abundance, file = here::here(output_dir_data, "Differential_Abundance_Fibrosis.csv"), quote=F, row.names = F)

reactable(differential_abundance, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```


```{r cluster_names}
cluster_names <- c("Cardiomyocytes", "Fibroblasts", "Endothelial cells", "Pericytes", "Macrophages", "Lymphocytes", "Smooth muscle cells", "Neuronal cells", "Endocardial cells")
```


Remove genes expressed in <1% of cells

```{r DE_fibrosis, eval=TRUE}
results_DE_fibrosis <- list()
signif_DE_fibrosis <- list()

for (cluster in cluster_names) {
data <- subset(DCM_RV_integrated, cell_type == cluster & Sample %in% Fibrosis_data_DCM$Sample)  
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- data[which(rownames(data) %in% keep_genes),]
rm(keep_genes, non_zero_cells, expr_matrix, total_cells, percent_cells)
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(Fibrosis_data_DCM$Sample, pseudocounts$samples$sample)]
all(Fibrosis_data_DCM$Sample == pseudocounts$samples$sample)
keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
pseudocounts <- normLibSizes(pseudocounts)
Fibrosis_data_subset <- Fibrosis_data_DCM %>%
  dplyr::filter(Sample %in% pseudocounts$samples$sample)
design <- model.matrix(~ Fibrosis_data_subset[["Fibrosis_percent"]])
pseudocounts <- estimateDisp(pseudocounts, design)
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
results_DE_fibrosis[[cluster]] <- topTags(fit, n = Inf)$table
results_DE_fibrosis[[cluster]] <- merge(results_DE_fibrosis[[cluster]], percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_DE_fibrosis[[cluster]]$up_down <- ifelse(
  results_DE_fibrosis[[cluster]]$logFC > 0.5 & results_DE_fibrosis[[cluster]]$FDR < 0.05, "Increased",
  ifelse(results_DE_fibrosis[[cluster]]$logFC < -0.5 & results_DE_fibrosis[[cluster]]$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )
signif_DE_fibrosis[[cluster]] <- results_DE_fibrosis[[cluster]] %>%
  dplyr::filter(FDR < 0.05) %>%
  dplyr::arrange(FDR)
}
```


Volcano plots

```{r get_legend}
get_legend <- function(my_plot) {
  tmp <- ggplotGrob(my_plot)
  legend <- tmp$grobs[[which(sapply(tmp$grobs, function(x) x$name) == "guide-box")]]
  return(legend)
}
```

```{r Volcano_up_down, fig.height=9, fig.width=16}
volcano <- list()
for (i in seq_along(results_DE_fibrosis)) {
  keyvals <- rep("grey", nrow(results_DE_fibrosis[[i]]))
  names(keyvals) <- rep("Not significant", nrow(results_DE_fibrosis[[i]]))
  
  keyvals[results_DE_fibrosis[[i]]$up_down == "Increased"] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Increased"
  keyvals[results_DE_fibrosis[[i]]$up_down == "Decreased"] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Decreased"

  volcano[[i]] <- EnhancedVolcano(results_DE_fibrosis[[i]],
                  lab = results_DE_fibrosis[[i]]$gene,
                  x = "logFC",
                  y = "FDR",
                  labSize = 0,
                  legendLabSize = 15,
                  titleLabSize = 16,
                  subtitleLabSize = 14,
                  axisLabSize = 12,
                  captionLabSize = 9,
                  pointSize = 1,
                  FCcutoff = 0.5,
                  pCutoff  = 0.05,
                  ylim = c(0, 5),
                  colAlpha = 1,
                  drawConnectors = FALSE,
                  subtitle = NULL,
                  title = names(results_DE_fibrosis)[i],
                  colCustom = keyvals
  ) + theme(legend.position = "none")
  names(volcano)[i] <- names(results_DE_fibrosis)[i]
}

# Extract the legend from the first plot
legend <- cowplot::get_legend(volcano[[1]] + theme(legend.position = "right"))

# Combine the individual volcano plots
combined_plot <- patchwork::wrap_plots(volcano, ncol = 5)

# Use ggdraw and draw_plot to overlay the legend in the bottom-right corner
final_plot <- cowplot::ggdraw() +
  cowplot::draw_plot(combined_plot, 0, 0, 1, 1) +    # Place the combined plot to fill the whole canvas
  cowplot::draw_plot(legend, 0.77, 0.15, 0.25, 0.25)    # Position the legend in the bottom-right

# Print the final plot
print(final_plot)
```




## GSEA

```{r reorder_GO_by_pvalue}
reorder_GO_by_pvalue <- function(enrichGO_result) {
  go_results <- enrichGO_result@result
  go_results_sorted <- go_results[order(go_results$p.adjust), ]
  enrichGO_result@result <- go_results_sorted
  return(enrichGO_result)
}
```

```{r classify_terms}
classify_terms <- function(df) {
  df$sign <- ifelse(df$NES > 0, "activated", "suppressed")
  return(df)
}
```

```{r cluster_names_2}
cluster_names_2 <- c("Cardiomyocytes", "Fibroblasts", "Endothelial cells", "Pericytes", "Macrophages", "Lymphocytes", "Smooth muscle cells", "Neuronal cells", "Endocardial cells")
```


```{r GSEA_all, eval=FALSE}
ranked_genes <- list()
GSEA_all <- list()
for (i in seq_along(cluster_names)) {
  gene_counts <- rowSums(subset(DCM_RV_integrated, cell_type == cluster_names_2[i])@assays$RNA@counts > 0)
  universe_genes <- names(gene_counts[gene_counts >= 3])
  universe_entrez <- bitr(universe_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)$ENTREZID
  rm(gene_counts)
  
  up_genes <- filter(signif_DE_fibrosis[[i]], logFC > 0.5)$gene
  up_genes <- bitr(up_genes, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")$ENTREZID
  down_genes <- filter(signif_DE_fibrosis[[i]], logFC < -0.5)$gene
  down_genes <- bitr(down_genes, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")$ENTREZID
  
  ranked_genes[[i]] <- results_DE_fibrosis[[i]]
  names(ranked_genes)[i] <- cluster_names[i]
  ranked_genes[[i]]$metric <- ranked_genes[[i]]$logFC*-log10(ranked_genes[[i]]$PValue)
  entrezid <- bitr(ranked_genes[[i]]$gene, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db", drop = TRUE)
  ranked_genes[[i]] <- merge(ranked_genes[[i]], entrezid, by.x = "gene", by.y = "SYMBOL")
  ranked_genes[[i]] <- ranked_genes[[i]][!duplicated(ranked_genes[[i]]$metric), ]
  write.csv(ranked_genes[[i]], file = here::here(output_dir_data, paste0(cluster_names[i], "_ranked_genes.csv")), quote=F, row.names = F)
  
  genelist <- ranked_genes[[i]]$metric
  names(genelist) <- ranked_genes[[i]]$ENTREZID
  genelist <- genelist[order(genelist, decreasing = T)]
  
  GSEA_all[i] <- gseGO(geneList = genelist,
                       OrgDb = org.Hs.eg.db,
                       ont = "ALL",
                       keyType = "ENTREZID",
                       nPermSimple = 1000000,
                       minGSSize = 10,
                       eps = 0)
  GSEA_all[[i]] <- reorder_GO_by_pvalue(GSEA_all[[i]])
  names(GSEA_all)[i] <- cluster_names[i]
  saveRDS(GSEA_all[i], here::here(output_dir_data, paste0(cluster_names[i], "_GSEA_GO_Fibrosis.rds")))
}

rm(universe_genes, universe_entrez, up_genes, down_genes, genelist)

saveRDS(ranked_genes, here::here(output_dir_data, "ranked_genes_list_Fibrosis.rds"))
saveRDS(GSEA_all, here::here(output_dir_data, "GSEA_all_list_Fibrosis.rds"))
```

```{r load GSEA_all}
GSEA_all <- readRDS(here::here(output_dir_data, "GSEA_all_list_Fibrosis.rds"))
```

```{r GSEA_Cardiomyocytes, fig.height=8, fig.width=8}
GSEA_simplified <- simplify(
  GSEA_all[["Cardiomyocytes"]],
  cutoff = 0.75,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)

GSEA_simplified <- reorder_GO_by_pvalue(GSEA_simplified)

dotplot(GSEA_simplified, showCategory = 8, split=".sign", title = paste0("Cardiomyocytes", "\nGO Enrichment Analysis", "\nmPAP-associated genes"), 
        label_format = 30, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
```

```{r GSEA_Cardiomyocytes_emapplot, fig.height=10, fig.width=20}
GSEA_simplified@result <- classify_terms(GSEA_simplified@result)

activated_terms <- subset(GSEA_simplified@result, sign == "activated")
suppressed_terms <- subset(GSEA_simplified@result, sign == "suppressed")

GSEA_activated <- GSEA_simplified
GSEA_activated@result <- activated_terms
GSEA_activated <- pairwise_termsim(GSEA_activated)

GSEA_suppressed <- GSEA_simplified
GSEA_suppressed@result <- suppressed_terms
GSEA_suppressed <- pairwise_termsim(GSEA_suppressed)

p1 <- emapplot(GSEA_activated, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

p2 <- emapplot(GSEA_suppressed, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Suppressed Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

combined_plot <- p1 | p2
print(combined_plot)
rm(GSEA_simplified, activated_terms, suppressed_terms, GSEA_activated, GSEA_suppressed, p1, p2)
```

```{r GSEA_Fibroblasts, fig.height=8, fig.width=8}
GSEA_simplified <- simplify(
  GSEA_all[["Fibroblasts"]],
  cutoff = 0.75,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)

GSEA_simplified <- reorder_GO_by_pvalue(GSEA_simplified)

dotplot(GSEA_simplified, showCategory = 8, split=".sign", title = paste0("Fibroblasts", "\nGO Enrichment Analysis", "\nmPAP-associated genes"), 
        label_format = 30, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
```

```{r GSEA_Fibroblasts_emapplot, fig.height=10, fig.width=20}
GSEA_simplified@result <- classify_terms(GSEA_simplified@result)

activated_terms <- subset(GSEA_simplified@result, sign == "activated")
suppressed_terms <- subset(GSEA_simplified@result, sign == "suppressed")

GSEA_activated <- GSEA_simplified
GSEA_activated@result <- activated_terms
GSEA_activated <- pairwise_termsim(GSEA_activated)

GSEA_suppressed <- GSEA_simplified
GSEA_suppressed@result <- suppressed_terms
GSEA_suppressed <- pairwise_termsim(GSEA_suppressed)

p1 <- emapplot(GSEA_activated, showCategory = 70, cex_label_category = 0.7, min_edge = 0.20) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

p2 <- emapplot(GSEA_suppressed, showCategory = 70, cex_label_category = 0.7, min_edge = 0.20) +
  ggtitle("Suppressed Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

combined_plot <- p1 | p2
print(combined_plot)
rm(GSEA_simplified, activated_terms, suppressed_terms, GSEA_activated, GSEA_suppressed, p1, p2)
```

```{r GSEA_Endothelial_cells, fig.height=8, fig.width=8}
GSEA_simplified <- simplify(
  GSEA_all[["Endothelial cells"]],
  cutoff = 0.75,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)

GSEA_simplified <- reorder_GO_by_pvalue(GSEA_simplified)

dotplot(GSEA_simplified, showCategory = 8, split=".sign", title = paste0("Endothelial cells", "\nGO Enrichment Analysis", "\nmPAP-associated genes"), 
        label_format = 40, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
```

```{r GSEA_Endothelial_cells_emapplot, fig.height=10, fig.width=20}
GSEA_simplified@result <- classify_terms(GSEA_simplified@result)

activated_terms <- subset(GSEA_simplified@result, sign == "activated")
suppressed_terms <- subset(GSEA_simplified@result, sign == "suppressed")

GSEA_activated <- GSEA_simplified
GSEA_activated@result <- activated_terms
GSEA_activated <- pairwise_termsim(GSEA_activated)

GSEA_suppressed <- GSEA_simplified
GSEA_suppressed@result <- suppressed_terms
#GSEA_suppressed <- pairwise_termsim(GSEA_suppressed)

p1 <- emapplot(GSEA_activated, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

#p2 <- emapplot(GSEA_suppressed, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Suppressed Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

combined_plot <- p1 
print(combined_plot)
rm(GSEA_simplified, activated_terms, suppressed_terms, GSEA_activated, GSEA_suppressed, p1, p2)
```

```{r GSEA_Pericytes, fig.height=8, fig.width=8}
GSEA_simplified <- simplify(
  GSEA_all[["Pericytes"]],
  cutoff = 0.75,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)

GSEA_simplified <- reorder_GO_by_pvalue(GSEA_simplified)

dotplot(GSEA_simplified, showCategory = 8, split=".sign", title = paste0("Pericytes", "\nGO Enrichment Analysis", "\nmPAP-associated genes"), 
        label_format = 30, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
```

```{r GSEA_Pericytes_emapplot, fig.height=10, fig.width=20}
GSEA_simplified@result <- classify_terms(GSEA_simplified@result)

activated_terms <- subset(GSEA_simplified@result, sign == "activated")
suppressed_terms <- subset(GSEA_simplified@result, sign == "suppressed")

GSEA_activated <- GSEA_simplified
GSEA_activated@result <- activated_terms
GSEA_activated <- pairwise_termsim(GSEA_activated)

GSEA_suppressed <- GSEA_simplified
GSEA_suppressed@result <- suppressed_terms
GSEA_suppressed <- pairwise_termsim(GSEA_suppressed)

p1 <- emapplot(GSEA_activated, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

p2 <- emapplot(GSEA_suppressed, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Suppressed Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

combined_plot <- p1 | p2
print(combined_plot)
rm(GSEA_simplified, activated_terms, suppressed_terms, GSEA_activated, GSEA_suppressed, p1, p2)
```

```{r GSEA_Macrophages, fig.height=8, fig.width=8}
GSEA_simplified <- simplify(
  GSEA_all[["Macrophages"]],
  cutoff = 0.75,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)

GSEA_simplified <- reorder_GO_by_pvalue(GSEA_simplified)

dotplot(GSEA_simplified, showCategory = 8, split=".sign", title = paste0("Macrophages", "\nGO Enrichment Analysis", "\nmPAP-associated genes"), 
        label_format = 30, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
```

```{r GSEA_Macrophages_emapplot, fig.height=10, fig.width=20}
GSEA_simplified@result <- classify_terms(GSEA_simplified@result)

activated_terms <- subset(GSEA_simplified@result, sign == "activated")
suppressed_terms <- subset(GSEA_simplified@result, sign == "suppressed")

GSEA_activated <- GSEA_simplified
GSEA_activated@result <- activated_terms
GSEA_activated <- pairwise_termsim(GSEA_activated)

GSEA_suppressed <- GSEA_simplified
GSEA_suppressed@result <- suppressed_terms
GSEA_suppressed <- pairwise_termsim(GSEA_suppressed)

p1 <- emapplot(GSEA_activated, showCategory = 70, cex_label_category = 0.7, min_edge = 0.20) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

p2 <- emapplot(GSEA_suppressed, showCategory = 70, cex_label_category = 0.7, min_edge = 0.20) +
  ggtitle("Suppressed Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

combined_plot <- p1 | p2
print(combined_plot)
rm(GSEA_simplified, activated_terms, suppressed_terms, GSEA_activated, GSEA_suppressed, p1, p2)
```

```{r GSEA_Lymphocytes, fig.height=8, fig.width=8}
GSEA_simplified <- simplify(
  GSEA_all[["Lymphocytes"]],
  cutoff = 0.75,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)

GSEA_simplified <- reorder_GO_by_pvalue(GSEA_simplified)

dotplot(GSEA_simplified, showCategory = 8, split=".sign", title = paste0("Lymphocytes", "\nGO Enrichment Analysis", "\nmPAP-associated genes"), 
        label_format = 30, font.size = 15) +
  theme(plot.title = element_text( face = "bold", size = 18, hjust = 0.5),
        axis.text.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 18)) +
  facet_grid(.~.sign)
```

```{r GSEA_Lymphocytes_emapplot, fig.height=10, fig.width=10}
GSEA_simplified@result <- classify_terms(GSEA_simplified@result)

activated_terms <- subset(GSEA_simplified@result, sign == "activated")

GSEA_activated <- GSEA_simplified
GSEA_activated@result <- activated_terms
GSEA_activated <- pairwise_termsim(GSEA_activated)

p1 <- emapplot(GSEA_activated, showCategory = 100, cex_label_category = 0.7, min_edge = 0.15) +
  ggtitle("Activated Terms") +
  theme(plot.title = element_text(face = "bold", size = 18, hjust = 0.5))

print(p1)
rm(GSEA_simplified, activated_terms, suppressed_terms, GSEA_activated, GSEA_suppressed, p1, p2)
```

```{r FAP, fig.width=3, fig.height=3}
data <- subset(DCM_RV_integrated, cell_type == "Fibroblasts" & Sample %in% Fibrosis_data_DCM$Sample)  
expr_matrix <- GetAssayData(object = data, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)
keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- data[which(rownames(data) %in% keep_genes),]
rm(keep_genes, non_zero_cells, expr_matrix, total_cells, percent_cells)
pseudocounts <- Seurat2PB(data, sample="Sample", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(Fibrosis_data_DCM$Sample, pseudocounts$samples$sample)]
all(Fibrosis_data_DCM$Sample == pseudocounts$samples$sample)
keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
pseudocounts <- normLibSizes(pseudocounts)

counts <- cpm(pseudocounts, log = TRUE)

metadata_matched <- Fibrosis_data_DCM[match(colnames(counts), Fibrosis_data_DCM$Sample), ]

stopifnot(all(colnames(counts) == metadata_matched$Sample))

order_idx <- order(metadata_matched$Group)

counts <- counts[, order_idx]
metadata_matched <- metadata_matched[order_idx, ]

data <- data.frame(FAP = t(counts)[,"FAP"],
                   Fibrosis = metadata_matched$Fibrosis_percent)

ggplot(data, aes_string(x = "Fibrosis", y = "FAP")) +
    geom_point() +
    theme_minimal() +
    labs(
      x = "Fibrosis (%)", 
      y = "FAP (cpm)"
    )
```


```{r FAP_2, fig.width=8, fig.height=4}
vlnplot <- VlnPlot(DCM_RV_integrated, group.by = "cell_type", features = "FAP", add.noise = F)
vlnplot[[1]]$layers[[2]]$aes_params$alpha <- 0.2
vlnplot
```
