---
title: "Endothelial cells Koenig Reichart manuscript"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
current_file <- "Endothelial_cells_Koenig_Reichart_manuscript"

library(here)
library(readxl)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(tibble)
library(patchwork)
library(edgeR)
library(limma)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(ggpubr)
library(reactable)
library(scales)
library(cowplot)
library(speckle)
library(broom)
library(harmony)
library(clustree)
library(qs2)
library(decoupleR)
library(OmnipathR)
library(dorothea)

output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"), recursive = TRUE)
output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs, recursive = TRUE)

base_data_dir <- here::here("data", "Heart_datasets")

cell_type_cols <- c(
  "#d05050", "#33A02C", "#1F78B4", "#CAB2D6", "#FF7F00", "#B2DF8A",
  "#B15928", "#FFD000", "#FA95C0", "#FDBF6F", "#6A3D9A", "#7aCDF9", "#E6E600"
)

group_cols <- c("H" = "grey85", "HF" = "grey45")
sex_cols <- c("F" = "#6C8EBF", "M" = "#8AB17D")
dataset_cols <- c("Koenig" = "#CAB2D6", "Reichart" = "#B2DF8A")
heat_cols <- circlize::colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

wrap_2lines <- function(x, width = 35) {
  vapply(x, function(s) {
    s <- as.character(s)
    if (nchar(s) <= width) return(s)
    cut <- stringr::str_locate_all(substr(s, 1, width), " ")[[1]]
    cut <- if (nrow(cut) == 0) width else max(cut[, 1])
    line1 <- stringr::str_trim(substr(s, 1, cut))
    rest  <- stringr::str_trim(substr(s, cut + 1, nchar(s)))
    if (nchar(rest) > width) rest <- paste0(substr(rest, 1, width - 1), "…")
    paste(line1, rest, sep = "\n")
  }, FUN.VALUE = character(1))
}


format_p_label <- function(p) {
  if (length(p) == 0 || is.na(p)) return("NA")
  dplyr::case_when(
    p < 0.0001 ~ "****",
    p < 0.001  ~ "***",
    p < 0.01   ~ "**",
    p < 0.05   ~ "*",
    TRUE ~ sprintf("%.4f", p)
  )
}

format_p_fontface <- function(p) {
  if (length(p) == 0 || is.na(p)) return("plain")
  dplyr::case_when(
    p < 0.05 ~ "bold",
    TRUE ~ "plain"
  )
}


volcano_cols <- c(
  "Decreased" = "#004C8C",
  "Not significant" = "grey75",
  "Increased" = "#A63A2A"
)
```

## Load datasets

```{r load datasets}
load(file.path(base_data_dir, "Koenig_2022", "GSE183852_DCM_Nuclei.Robj"))
Koenig <- nuclei
rm(nuclei)

Reichart <- readRDS(file.path(base_data_dir, "Reichart_2022", "DCM_ACM_heart_cell_atlas.rds"))

# Reichart uses Ensembl IDs in rownames, so replace them with gene symbols
Reichart_gene_names <- data.frame(
  ensembl_id = rownames(Reichart),
  gene_symbol = Reichart@assays[["RNA"]]@meta.features$feature_name
)

rownames(Reichart@assays[["RNA"]]@counts) <- Reichart@assays[["RNA"]]@meta.features$feature_name
rownames(Reichart@assays[["RNA"]]@data) <- Reichart@assays[["RNA"]]@meta.features$feature_name
```

## Add metadata

```{r add metadata}
Clinical_data_Koenig <- readxl::read_excel(
  file.path(base_data_dir, "Koenig_2022", "Koenig_et_al_2022_Selected_clinical_info.xlsx")
)
add_metadata_Koenig <- dplyr::left_join(Koenig[["orig.ident"]], Clinical_data_Koenig)
row.names(add_metadata_Koenig) <- row.names(Koenig[[]])
Koenig <- AddMetaData(Koenig, metadata = add_metadata_Koenig)
rm(add_metadata_Koenig, Clinical_data_Koenig)

Clinical_data_Reichart <- readxl::read_excel(
  file.path(base_data_dir, "Reichart_2022", "Reichart_et_al_Science_2022_Selected_clinical_info.xlsx")
)
add_metadata_Reichart <- dplyr::left_join(Reichart[["donor_id"]], Clinical_data_Reichart)
row.names(add_metadata_Reichart) <- row.names(Reichart[[]])
Reichart <- AddMetaData(Reichart, metadata = add_metadata_Reichart)
rm(add_metadata_Reichart, Clinical_data_Reichart)

Reichart$Sample <- Reichart$donor_id
Reichart$Group <- dplyr::case_when(
  Reichart$disease == "normal" ~ "HC",
  Reichart$disease == "dilated cardiomyopathy" ~ "DCM",
  TRUE ~ as.character(Reichart$disease)
)

Koenig$Sample <- Koenig$orig.ident
Koenig$Group <- dplyr::case_when(
  Koenig$Condition == "Donor" ~ "HC",
  Koenig$Condition == "DCM" ~ "DCM",
  TRUE ~ as.character(Koenig$Condition)
)

Koenig$cell_type_plot <- Koenig$Names

Reichart$cell_type_plot <- dplyr::case_when(
  Reichart$cell_type == "endothelial cell" & Reichart$cell_states == "EC7.0" ~ "endocardial cell",
  Reichart$cell_type == "endothelial cell" & Reichart$cell_states == "EC8.0" ~ "lymphatic cell",
  Reichart$cell_type == "endothelial cell" & Reichart$cell_states == "Meso" ~ "mesothelial cell",
  Reichart$cell_type == "endothelial cell" & Reichart$cell_states %in% c("EC1.0", "EC2.0", "EC5.0", "EC6.0") ~ "endothelial cell",
  TRUE ~ as.character(Reichart$cell_type)
)

Reichart <- subset(
  Reichart,
  subset = cell_type_plot != "unknown"
)
```

## Whole-dataset UMAPs

```{r whole dataset umaps, fig.width=12, fig.height=6}
koenig_levels <- sort(unique(as.character(Koenig$cell_type_plot)))
koenig_cols <- setNames(rep(cell_type_cols, length.out = length(koenig_levels)), koenig_levels)

reichart_levels <- sort(unique(as.character(Reichart$cell_type_plot)))
reichart_cols <- setNames(rep(cell_type_cols, length.out = length(reichart_levels)), reichart_levels)

p_koenig <- DimPlot(
  Koenig,
  group.by = "cell_type_plot",
  reduction = "umap",
  cols = koenig_cols,
  label = FALSE,
  shuffle = TRUE,
  pt.size = 1.2
) +
  NoLegend() +
  ggtitle(expression(italic("Koenig et al.") ~ " Nature CVR 2022")) +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5)
  )

p_koenig <- LabelClusters(
  p_koenig,
  id = "cell_type_plot",
  fontface = "bold",
  size = 5.2,
  repel = TRUE
)

p_reichart <- DimPlot(
  Reichart,
  group.by = "cell_type_plot",
  reduction = "umap",
  cols = reichart_cols,
  label = FALSE,
  shuffle = TRUE,
  pt.size = 1.2
) +
  NoLegend() +
  ggtitle(expression(italic("Reichart et al.") ~ " Science 2022")) +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5)
  )

p_reichart <- LabelClusters(
  p_reichart,
  id = "cell_type_plot",
  fontface = "bold",
  size = 5.2,
  repel = TRUE
)

p_koenig + p_reichart
```

```{r endothelial highlight umaps, fig.width=12, fig.height=6}
koenig_ec_cells <- rownames(Koenig@meta.data)[Koenig$Names %in% c("Endothelium")]
reichart_ec_cells <- rownames(Reichart@meta.data)[Reichart$cell_type_plot %in% c("endothelial cell")]

p_koenig_ec <- DimPlot(
  Koenig,
  reduction = "umap",
  cells.highlight = koenig_ec_cells,
  cols.highlight = "#1F78B4",
  cols = "grey85",
  sizes.highlight = 0.01,
  shuffle = TRUE,
  pt.size = 0.01,
  raster = FALSE
) +
  NoLegend() +
  ggtitle(expression(italic("Koenig et al.") ~ " Nature CVR 2022")) +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.position = "none"
  )

p_reichart_ec <- DimPlot(
  Reichart,
  reduction = "umap",
  cells.highlight = reichart_ec_cells,
  cols.highlight = "#1F78B4",
  cols = "grey85",
  sizes.highlight = 0.01,
  shuffle = TRUE,
  pt.size = 0.01,
  raster = FALSE
) +
  NoLegend() +
  ggtitle(expression(italic("Reichart et al.") ~ " Science 2022")) +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.position = "none"
  )

p_koenig_ec + p_reichart_ec
```

## Endothelial-cell subsets

```{r endothelial subsets}
Koenig_EC <- subset(Koenig, Names == "Endothelium")
Reichart_EC <- subset(
  Reichart,
  subset = cell_type_plot == "endothelial cell" & disease %in% c("normal", "dilated cardiomyopathy") & tissue == "heart left ventricle"
)

Koenig_EC$Dataset <- "Koenig"
Reichart_EC$Dataset <- "Reichart"

Koenig_EC$Group <- ifelse(Koenig_EC$Condition == "Donor", "H", "HF")
Reichart_EC$Group <- ifelse(Reichart_EC$disease == "normal", "H", "HF")

Koenig_EC$Sex <- as.character(Koenig_EC$Sex)
Reichart_EC$Sex <- as.character(Reichart_EC$sex)

Koenig_EC$Sex[Koenig_EC$Sex %in% c("Male", "male", "m")] <- "M"
Koenig_EC$Sex[Koenig_EC$Sex %in% c("Female", "female", "f")] <- "F"
Reichart_EC$Sex[Reichart_EC$Sex %in% c("Male", "male", "m")] <- "M"
Reichart_EC$Sex[Reichart_EC$Sex %in% c("Female", "female", "f")] <- "F"

age_midpoints <- c(
  "0-9" = 5,
  "10-19" = 15,
  "20-29" = 25,
  "30-39" = 35,
  "40-49" = 45,
  "50-59" = 55,
  "60-69" = 65,
  "70-79" = 75
)

Koenig_EC$Age <- as.numeric(as.character(Koenig_EC$Age))
Reichart_EC$Age <- unname(age_midpoints[as.character(Reichart_EC$Age_range)])
Reichart_EC$Age <- as.numeric(Reichart_EC$Age)
```

## Sample-level metadata

```{r sample metadata}
metadata_Koenig <- Koenig_EC@meta.data %>%
  dplyr::select(Sample = orig.ident, Age, Sex, Group, Dataset) %>%
  dplyr::distinct()

metadata_Reichart <- Reichart_EC@meta.data %>%
  dplyr::select(Sample = donor_id, Age, Sex, Group, Dataset) %>%
  dplyr::distinct()

metadata_Koenig$Group <- factor(metadata_Koenig$Group, levels = c("H", "HF"))
metadata_Reichart$Group <- factor(metadata_Reichart$Group, levels = c("H", "HF"))
metadata_Koenig$Sex <- factor(metadata_Koenig$Sex, levels = c("F", "M"))
metadata_Reichart$Sex <- factor(metadata_Reichart$Sex, levels = c("F", "M"))
metadata_Koenig$Dataset <- factor(metadata_Koenig$Dataset, levels = c("Koenig", "Reichart"))
metadata_Reichart$Dataset <- factor(metadata_Reichart$Dataset, levels = c("Koenig", "Reichart"))

reactable(metadata_Koenig)
reactable(metadata_Reichart)
```

## Pseudobulk per dataset

```{r pseudobulk koenig}
expr_matrix_Koenig <- GetAssayData(Koenig_EC, assay = "RNA", slot = "counts")
non_zero_cells_Koenig <- rowSums(expr_matrix_Koenig > 0)
percent_cells_Koenig <- non_zero_cells_Koenig / ncol(expr_matrix_Koenig) * 100
percent_stats_Koenig <- data.frame(gene = rownames(expr_matrix_Koenig), percentCells = percent_cells_Koenig)
keep_genes_Koenig <- percent_stats_Koenig$gene[percent_stats_Koenig$percentCells > 1]
Koenig_EC_pb_input <- Koenig_EC[keep_genes_Koenig, ]

pseudobulk_Koenig <- Seurat2PB(Koenig_EC_pb_input, sample = "orig.ident", cluster = "Names")
colnames(pseudobulk_Koenig) <- pseudobulk_Koenig$samples$sample
pseudobulk_Koenig <- pseudobulk_Koenig[, match(metadata_Koenig$Sample, pseudobulk_Koenig$samples$sample)]

keep_samples_Koenig <- pseudobulk_Koenig$samples$lib.size > 5e4
pseudobulk_Koenig <- pseudobulk_Koenig[, keep_samples_Koenig]
metadata_Koenig <- metadata_Koenig %>% dplyr::filter(Sample %in% colnames(pseudobulk_Koenig))
metadata_Koenig <- metadata_Koenig %>% dplyr::slice(match(colnames(pseudobulk_Koenig), Sample))
stopifnot(all(metadata_Koenig$Sample == colnames(pseudobulk_Koenig)))
```

```{r pseudobulk reichart}
expr_matrix_Reichart <- GetAssayData(Reichart_EC, assay = "RNA", slot = "counts")
non_zero_cells_Reichart <- rowSums(expr_matrix_Reichart > 0)
percent_cells_Reichart <- non_zero_cells_Reichart / ncol(expr_matrix_Reichart) * 100
percent_stats_Reichart <- data.frame(gene = rownames(expr_matrix_Reichart), percentCells = percent_cells_Reichart)
keep_genes_Reichart <- percent_stats_Reichart$gene[percent_stats_Reichart$percentCells > 1]
Reichart_EC_pb_input <- Reichart_EC[keep_genes_Reichart, ]

pseudobulk_Reichart <- Seurat2PB(Reichart_EC_pb_input, sample = "donor_id", cluster = "cell_type_plot")
colnames(pseudobulk_Reichart) <- pseudobulk_Reichart$samples$sample
pseudobulk_Reichart <- pseudobulk_Reichart[, match(metadata_Reichart$Sample, pseudobulk_Reichart$samples$sample)]

keep_samples_Reichart <- pseudobulk_Reichart$samples$lib.size > 5e4
pseudobulk_Reichart <- pseudobulk_Reichart[, keep_samples_Reichart]
metadata_Reichart <- metadata_Reichart %>% dplyr::filter(Sample %in% colnames(pseudobulk_Reichart))
metadata_Reichart <- metadata_Reichart %>% dplyr::slice(match(colnames(pseudobulk_Reichart), Sample))
stopifnot(all(metadata_Reichart$Sample == colnames(pseudobulk_Reichart)))
```

## Merge pseudobulks and run differential expression

```{r merge pseudobulk}
common_genes_pb <- intersect(rownames(pseudobulk_Koenig), rownames(pseudobulk_Reichart))
length(common_genes_pb)

combined_counts <- cbind(
  pseudobulk_Koenig[common_genes_pb, ]$counts,
  pseudobulk_Reichart[common_genes_pb, ]$counts
)

metadata_combined <- dplyr::bind_rows(metadata_Koenig, metadata_Reichart)
metadata_combined <- metadata_combined %>%
  dplyr::filter(Sample %in% colnames(combined_counts)) %>%
  dplyr::slice(match(colnames(combined_counts), Sample))

metadata_combined <- metadata_combined %>%
  dplyr::filter(!is.na(Age), !is.na(Sex), !is.na(Group), !is.na(Dataset))
combined_counts <- combined_counts[, metadata_combined$Sample]

metadata_combined$Group <- factor(metadata_combined$Group, levels = c("H", "HF"))
metadata_combined$Sex <- factor(metadata_combined$Sex, levels = c("F", "M"))
metadata_combined$Dataset <- factor(metadata_combined$Dataset, levels = c("Koenig", "Reichart"))

pseudobulk_combined <- edgeR::DGEList(counts = combined_counts)
pseudobulk_combined$samples <- cbind(pseudobulk_combined$samples, metadata_combined)

# Model: HF vs H, corrected for age, sex, and dataset
# Merging pseudobulks is fine here because the actual test is still sample-level,
# and dataset is included explicitly in the design.
design_combined <- model.matrix(~ Group + Age + Sex + Dataset, data = metadata_combined)
colnames(design_combined)

keep_genes_combined <- filterByExpr(pseudobulk_combined, design = design_combined)
pseudobulk_combined <- pseudobulk_combined[keep_genes_combined, , keep.lib.sizes = FALSE]
pseudobulk_combined <- normLibSizes(pseudobulk_combined)
pseudobulk_combined <- estimateDisp(pseudobulk_combined, design_combined)

logcpm_combined <- edgeR::cpm(pseudobulk_combined, log = TRUE, prior.count = 1)
```

```{r differential expression}
fit_combined <- glmQLFit(pseudobulk_combined, design_combined, robust = TRUE)
qlf_combined <- glmQLFTest(fit_combined, coef = "GroupHF")

results_combined <- topTags(qlf_combined, n = Inf)$table
results_combined <- results_combined %>%
  tibble::rownames_to_column("gene") %>%
  dplyr::mutate(metric = logFC * -log10(PValue))

results_combined$significance <- "Not significant"
results_combined$significance[results_combined$FDR < 0.05 & results_combined$logFC > 0.5] <- "Increased"
results_combined$significance[results_combined$FDR < 0.05 & results_combined$logFC < -0.5] <- "Decreased"
results_combined$significance <- factor(results_combined$significance,
                                        levels = c("Decreased", "Not significant", "Increased"))

signif_combined <- results_combined %>%
  dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>%
  dplyr::arrange(desc(metric))

write.csv(results_combined,
          file = file.path(output_dir_data, "EC_DE_results_Koenig_Reichart_combined.csv"),
          quote = FALSE, row.names = FALSE)
write.csv(signif_combined,
          file = file.path(output_dir_data, "EC_DE_signif_Koenig_Reichart_combined.csv"),
          quote = FALSE, row.names = FALSE)
```

## Volcano plot

```{r volcano, fig.width=4, fig.height=5}
n_h  <- sum(metadata_combined$Group == "H")
n_hf <- sum(metadata_combined$Group == "HF")
EnhancedVolcano(
  results_combined,
  lab = results_combined$gene,
  x = "logFC",
  y = "FDR",
  labSize = 0,
  legendLabSize = 12,
  titleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 0.8,
  FCcutoff = 0.5,
  pCutoff = 0.05,
  ylim = c(0, 20),
  xlim = c(-3.75, 3.75),
  colAlpha = 1,
  drawConnectors = FALSE,
  ylab = bquote(-Log[10]~FDR),
  subtitle = NULL,
  title = paste0("Endothelial cells\nHF (n=", n_hf, ") vs H (n=", n_h, ")"),
  colCustom = setNames(
    c("#004C8C", "grey75", "#A63A2A")[match(
      results_combined$significance,
      c("Decreased", "Not significant", "Increased")
    )],
    as.character(results_combined$significance)
  )
) +
  theme(legend.position = "none")
```

## Heatmaps

```{r order heatmap columns}
column_order <- metadata_combined %>%
  dplyr::arrange(Group, Dataset, Age, Sex) %>%
  dplyr::pull(Sample)

metadata_heatmap <- metadata_combined %>%
  dplyr::filter(Sample %in% column_order) %>%
  dplyr::slice(match(column_order, Sample))

ha <- HeatmapAnnotation(
  Group = metadata_heatmap$Group,
  Age = metadata_heatmap$Age,
  Sex = metadata_heatmap$Sex,
  Dataset = metadata_heatmap$Dataset,
  col = list(
    Group = group_cols,
    Age = circlize::colorRamp2(
      c(min(metadata_heatmap$Age, na.rm = TRUE), max(metadata_heatmap$Age, na.rm = TRUE)),
      c("grey90", "black")
    ),
    Sex = sex_cols,
    Dataset = dataset_cols
  ),
  annotation_name_gp = grid::gpar(fontsize = 10, fontface = "bold")
)
```

```{r heatmap all degs, fig.width=9, fig.height=12}
mat_all_deg <- logcpm_combined[rownames(logcpm_combined) %in% signif_combined$gene, column_order, drop = FALSE]
mat_all_deg <- mat_all_deg[match(signif_combined$gene, rownames(mat_all_deg)), , drop = FALSE]
mat_all_deg_z <- t(scale(t(mat_all_deg)))
mat_all_deg_z[is.na(mat_all_deg_z)] <- 0

Heatmap(
  mat_all_deg_z,
  name = "Z-score",
  top_annotation = ha,
  col = heat_cols,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  show_column_names = TRUE,
  column_names_gp = grid::gpar(fontsize = 5),
  row_names_gp = grid::gpar(fontsize = 6),
  column_title = "All significant DEGs",
  column_title_gp = grid::gpar(fontsize = 12, fontface = "bold")
)
```

```{r heatmap top 25 up down, fig.width=9, fig.height=8}
top_up <- results_combined %>%
  dplyr::filter(FDR < 0.05, logFC > 0.5) %>%
  dplyr::arrange(desc(metric)) %>%
  dplyr::slice_head(n = 25)

top_down <- results_combined %>%
  dplyr::filter(FDR < 0.05, logFC < -0.5) %>%
  dplyr::arrange(metric) %>%
  dplyr::slice_head(n = 25)

top_genes <- dplyr::bind_rows(top_up, top_down)

mat_top_deg <- logcpm_combined[rownames(logcpm_combined) %in% top_genes$gene, column_order, drop = FALSE]
mat_top_deg <- mat_top_deg[match(top_genes$gene, rownames(mat_top_deg)), , drop = FALSE]
mat_top_deg_z <- t(scale(t(mat_top_deg)))
mat_top_deg_z[is.na(mat_top_deg_z)] <- 0

Heatmap(
  mat_top_deg_z,
  name = "Z-score",
  top_annotation = ha,
  col = heat_cols,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  column_names_gp = grid::gpar(fontsize = 5),
  row_names_gp = grid::gpar(fontsize = 10),
  column_title = "Top 25 up and top 25 down genes",
  column_title_gp = grid::gpar(fontsize = 16, fontface = "bold")
)
```

## GSEA

## GSEA - run once and save objects

```{r gsea run and save, eval=FALSE}
ranked_genes <- results_combined %>%
  dplyr::select(gene, metric) %>%
  dplyr::filter(!is.na(metric))

entrezid <- bitr(
  ranked_genes$gene,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db,
  drop = TRUE
)

ranked_genes_entrez <- dplyr::left_join(
  ranked_genes,
  entrezid,
  by = c("gene" = "SYMBOL")
) %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::group_by(ENTREZID) %>%
  dplyr::slice_max(order_by = abs(metric), n = 1, with_ties = FALSE) %>%
  dplyr::ungroup()

genelist <- ranked_genes_entrez$metric
names(genelist) <- ranked_genes_entrez$ENTREZID
genelist <- sort(genelist, decreasing = TRUE)

GSEA_GO_ALL <- gseGO(
  geneList = genelist,
  OrgDb = org.Hs.eg.db,
  ont = "ALL",
  keyType = "ENTREZID",
  minGSSize = 10,
  maxGSSize = 500,
  eps = 0,
  nPermSimple = 10000
)

GSEA_GO_ALL <- clusterProfiler::setReadable(
  GSEA_GO_ALL,
  OrgDb = org.Hs.eg.db
)

GSEA_GO_ALL_simplified <- clusterProfiler::simplify(
  GSEA_GO_ALL,
  cutoff = 0.6,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)

GSEA_KEGG <- gseKEGG(
  geneList = genelist,
  organism = "hsa",
  keyType = "ncbi-geneid",
  minGSSize = 10,
  maxGSSize = 500,
  eps = 0,
  nPermSimple = 10000
)

GSEA_KEGG <- clusterProfiler::setReadable(
  GSEA_KEGG,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
)

GSEA_REACTOME <- ReactomePA::gsePathway(
  geneList = genelist,
  organism = "human",
  minGSSize = 10,
  maxGSSize = 500,
  eps = 0,
  pvalueCutoff = 1,
  pAdjustMethod = "BH",
  verbose = FALSE
)

GSEA_REACTOME <- clusterProfiler::setReadable(
  GSEA_REACTOME,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
)

wp_df <- msigdbr::msigdbr(
  species = "Homo sapiens",
  collection = "C2",
  subcollection = "CP:WIKIPATHWAYS"
)

term2gene_wp <- wp_df %>%
  dplyr::select(gs_name, ncbi_gene) %>%
  dplyr::filter(!is.na(ncbi_gene)) %>%
  dplyr::distinct()

term2name_wp <- wp_df %>%
  dplyr::select(gs_name, gs_description) %>%
  dplyr::distinct()

GSEA_WIKIPATHWAYS <- clusterProfiler::GSEA(
  geneList = genelist,
  TERM2GENE = term2gene_wp,
  TERM2NAME = term2name_wp,
  minGSSize = 10,
  maxGSSize = 500,
  eps = 0,
  pvalueCutoff = 1,
  pAdjustMethod = "BH",
  verbose = FALSE
)

qs2::qs_save(
  ranked_genes,
  file.path(output_dir_data, "ranked_genes.qs")
)

qs2::qs_save(
  ranked_genes_entrez,
  file.path(output_dir_data, "ranked_genes_entrez.qs")
)

qs2::qs_save(
  genelist,
  file.path(output_dir_data, "genelist.qs")
)

qs2::qs_save(
  GSEA_GO_ALL,
  file.path(output_dir_data, "GSEA_GO_ALL.qs")
)

qs2::qs_save(
  GSEA_GO_ALL_simplified,
  file.path(output_dir_data, "GSEA_GO_ALL_simplified.qs")
)

qs2::qs_save(
  GSEA_KEGG,
  file.path(output_dir_data, "GSEA_KEGG.qs")
)

qs2::qs_save(
  GSEA_REACTOME,
  file.path(output_dir_data, "GSEA_REACTOME.qs")
)

qs2::qs_save(
  GSEA_WIKIPATHWAYS,
  file.path(output_dir_data, "GSEA_WIKIPATHWAYS.qs")
)

write.csv(
  GSEA_GO_ALL@result,
  file = file.path(output_dir_data, "GSEA_GO_ALL_full_results.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  GSEA_GO_ALL_simplified@result,
  file = file.path(output_dir_data, "GSEA_GO_ALL_simplified_results.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  GSEA_KEGG@result,
  file = file.path(output_dir_data, "GSEA_KEGG_results.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  GSEA_REACTOME@result,
  file = file.path(output_dir_data, "GSEA_REACTOME_results.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  GSEA_WIKIPATHWAYS@result,
  file = file.path(output_dir_data, "GSEA_WIKIPATHWAYS_results.csv"),
  quote = FALSE,
  row.names = FALSE
)
```

```{r gsea load saved}
ranked_genes <- qs2::qs_read(
  file.path(output_dir_data, "ranked_genes.qs")
)

ranked_genes_entrez <- qs2::qs_read(
  file.path(output_dir_data, "ranked_genes_entrez.qs")
)

genelist <- qs2::qs_read(
  file.path(output_dir_data, "genelist.qs")
)

GSEA_GO_ALL <- qs2::qs_read(
  file.path(output_dir_data, "GSEA_GO_ALL.qs")
)

GSEA_GO_ALL_simplified <- qs2::qs_read(
  file.path(output_dir_data, "GSEA_GO_ALL_simplified.qs")
)

GSEA_KEGG <- qs2::qs_read(
  file.path(output_dir_data, "GSEA_KEGG.qs")
)

GSEA_REACTOME <- qs2::qs_read(
  file.path(output_dir_data, "GSEA_REACTOME.qs")
)

GSEA_WIKIPATHWAYS <- qs2::qs_read(
  file.path(output_dir_data, "GSEA_WIKIPATHWAYS.qs")
)
```

## Helper function for dotplot / emapplot / barplot

```{r gsea helper functions}
make_gsea_dotplot <- function(gsea_obj, title_text, show_n = 20, label_format = 45, font_size = 12) {
  if (is.null(gsea_obj) || nrow(gsea_obj@result) == 0) return(NULL)

  dotplot(
    gsea_obj,
    showCategory = show_n,
    split = ".sign",
    label_format = label_format,
    font.size = font_size,
    title = title_text
  ) +
    facet_grid(. ~ .sign) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
      strip.text = element_text(face = "bold", size = 12)
    )
}

make_gsea_emapplot <- function(gsea_obj, show_n = 30) {
  if (is.null(gsea_obj) || nrow(gsea_obj@result) == 0) return(NULL)

  gsea_res <- gsea_obj@result

  gsea_up <- gsea_obj
  gsea_up@result <- gsea_res %>% dplyr::filter(NES > 0, p.adjust < 0.05)

  gsea_down <- gsea_obj
  gsea_down@result <- gsea_res %>% dplyr::filter(NES < 0, p.adjust < 0.05)

  plot_list_emap <- list()

  if (nrow(gsea_up@result) > 1) {
    plot_list_emap[["up"]] <- emapplot(
      pairwise_termsim(gsea_up),
      showCategory = min(show_n, nrow(gsea_up@result)),
      size_category = 0.7,
      min_edge = 0.25
    ) +
      ggtitle("Activated terms") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))
  }

  if (nrow(gsea_down@result) > 1) {
    plot_list_emap[["down"]] <- emapplot(
      pairwise_termsim(gsea_down),
      showCategory = min(show_n, nrow(gsea_down@result)),
      size_category = 0.7,
      min_edge = 0.25
    ) +
      ggtitle("Suppressed terms") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))
  }

  if (length(plot_list_emap) == 2) {
    return(plot_list_emap[["up"]] + plot_list_emap[["down"]])
  } else if (length(plot_list_emap) == 1) {
    return(plot_list_emap[[1]])
  } else {
    return(NULL)
  }
}

make_gsea_barplot <- function(gsea_obj, label_width = 40, top_n = 10, base_size = 11) {
  if (is.null(gsea_obj) || nrow(gsea_obj@result) == 0) return(NULL)

  gsea_bar_df <- gsea_obj@result %>%
    as_tibble() %>%
    dplyr::filter(!is.na(NES), !is.na(p.adjust), p.adjust < 0.05)

  if (nrow(gsea_bar_df) == 0) return(NULL)

  gsea_bar_df <- gsea_bar_df %>%
    dplyr::mutate(score = -log10(p.adjust) * sign(NES))

  gsea_bar_df_pos <- gsea_bar_df %>%
    dplyr::filter(NES > 0) %>%
    dplyr::arrange(desc(NES)) %>%
    dplyr::slice_head(n = top_n)

  gsea_bar_df_neg <- gsea_bar_df %>%
    dplyr::filter(NES < 0) %>%
    dplyr::arrange(NES) %>%
    dplyr::slice_head(n = top_n)

  gsea_bar_df_plot <- dplyr::bind_rows(gsea_bar_df_pos, gsea_bar_df_neg)

  if (nrow(gsea_bar_df_plot) == 0) return(NULL)

  gsea_bar_df_plot$Description <- factor(
    gsea_bar_df_plot$Description,
    levels = gsea_bar_df_plot$Description[order(gsea_bar_df_plot$NES)]
  )

  L <- unname(quantile(abs(gsea_bar_df_plot$score), probs = 0.80, na.rm = TRUE))
  if (!is.finite(L) || L == 0) L <- max(abs(gsea_bar_df_plot$score), na.rm = TRUE)

  ggplot(gsea_bar_df_plot, aes(x = Description, y = NES, fill = score)) +
    geom_col(width = 0.9) +
    scale_fill_gradient2(
      low = "#004C8C",
      mid = "white",
      high = "#A63A2A",
      midpoint = 0,
      limits = c(-L, L),
      oob = scales::squish,
      name = expression(-log[10](adj.~p) %*% sign(NES))
    ) +
    scale_x_discrete(labels = function(x) wrap_2lines(x, width = label_width)) +
    coord_flip() +
    labs(x = NULL, y = "Normalized enrichment score (NES)") +
    theme_bw(base_size = base_size) +
    theme(
      text = element_text(color = "black"),
      axis.text = element_text(color = "black"),
      axis.title = element_text(color = "black", size = 13),
      plot.title = element_blank(),
      axis.text.y = element_text(size = 11),
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 9)
    ) +
    guides(fill = guide_colorbar(
      barwidth = unit(6, "cm"),
      barheight = unit(0.3, "cm"),
      title.position = "top",
      title.hjust = 0.5
    ))
}
```

## GO dotplot

```{r gsea go dotplot, fig.width=10, fig.height=8}
p_go_dot <- make_gsea_dotplot(
  GSEA_GO_ALL_simplified,
  title_text = "Endothelial cells\nGene ontology enrichment analysis\nHF vs H",
  show_n = 20
)

if (!is.null(p_go_dot)) {
  p_go_dot
}
```

## GO emapplot

```{r gsea go emapplot, fig.width=16, fig.height=8}
p_go_emap <- make_gsea_emapplot(GSEA_GO_ALL_simplified, show_n = 30)

if (!is.null(p_go_emap)) {
  p_go_emap
}
```

## GO barplot

```{r gsea go barplot style, fig.width=6, fig.height=5}
p_go_bar <- make_gsea_barplot(GSEA_GO_ALL_simplified, label_width = 43)

if (!is.null(p_go_bar)) {
  p_go_bar
}
```

## KEGG dotplot

```{r gsea kegg dotplot, fig.width=10, fig.height=8}
p_kegg_dot <- make_gsea_dotplot(
  GSEA_KEGG,
  title_text = "Endothelial cells\nKEGG pathway enrichment analysis\nHF vs H",
  show_n = 20,
  label_format = 60
)

if (!is.null(p_kegg_dot)) {
  p_kegg_dot
}
```

## KEGG emapplot

```{r gsea kegg emapplot, fig.width=16, fig.height=8}
p_kegg_emap <- make_gsea_emapplot(GSEA_KEGG, show_n = 30)

if (!is.null(p_kegg_emap)) {
  p_kegg_emap
}
```

## KEGG barplot

```{r gsea kegg barplot style, fig.width=6, fig.height=5}
p_kegg_bar <- make_gsea_barplot(GSEA_KEGG, label_width = 43)

if (!is.null(p_kegg_bar)) {
  p_kegg_bar
}
```

## Reactome dotplot

```{r gsea reactome dotplot, fig.width=10, fig.height=8}
p_reactome_dot <- make_gsea_dotplot(
  GSEA_REACTOME,
  title_text = "Endothelial cells\nReactome pathway enrichment analysis\nHF vs H",
  show_n = 15,
  label_format = 60
)

if (!is.null(p_reactome_dot)) {
  p_reactome_dot
}
```

## Reactome emapplot

```{r gsea reactome emapplot, fig.width=16, fig.height=8}
p_reactome_emap <- make_gsea_emapplot(GSEA_REACTOME, show_n = 30)

if (!is.null(p_reactome_emap)) {
  p_reactome_emap
}
```

## Reactome barplot

```{r gsea reactome barplot style, fig.width=7.3, fig.height=5}
p_reactome_bar <- make_gsea_barplot(GSEA_REACTOME, label_width = 59)

if (!is.null(p_reactome_bar)) {
  p_reactome_bar
}
```

## WikiPathways dotplot

```{r gsea wikipathways dotplot, fig.width=10, fig.height=8}
p_wiki_dot <- make_gsea_dotplot(
  GSEA_WIKIPATHWAYS,
  title_text = "Endothelial cells\nWikiPathways enrichment analysis\nHF vs H",
  show_n = 20,
  label_format = 65
)

if (!is.null(p_wiki_dot)) {
  p_wiki_dot
}
```

## WikiPathways emapplot

```{r gsea wikipathways emapplot, fig.width=16, fig.height=8}
p_wiki_emap <- make_gsea_emapplot(GSEA_WIKIPATHWAYS, show_n = 30)

if (!is.null(p_wiki_emap)) {
  p_wiki_emap
}
```

## WikiPathways barplot

```{r gsea wikipathways barplot style, fig.width=6, fig.height=5}
p_wiki_bar <- make_gsea_barplot(GSEA_WIKIPATHWAYS, label_width = 43)

if (!is.null(p_wiki_bar)) {
  p_wiki_bar
}
```

## Cell cycle scoring in each dataset separately

```{r cell cycle scoring koenig}
s.genes <- Seurat::cc.genes.updated.2019$s.genes
g2m.genes <- Seurat::cc.genes.updated.2019$g2m.genes

Koenig_EC <- CellCycleScoring(
  Koenig_EC,
  s.features = s.genes,
  g2m.features = g2m.genes,
  set.ident = FALSE
)

Reichart_EC <- CellCycleScoring(
  Reichart_EC,
  s.features = s.genes,
  g2m.features = g2m.genes,
  set.ident = FALSE
)
```

## UMAPs colored by cell-cycle phase

```{r cell cycle umaps, fig.width=12, fig.height=6}
p_koenig_cc <- DimPlot(
  Koenig_EC,
  reduction = "umap",
  group.by = "Phase",
  shuffle = TRUE,
  pt.size = 0.7,
  raster = FALSE
) +
  ggtitle(expression(italic("Koenig et al.") ~ " Nature CVR 2022")) +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 10, face = "bold")
  )

p_reichart_cc <- DimPlot(
  Reichart_EC,
  reduction = "umap",
  group.by = "Phase",
  shuffle = TRUE,
  pt.size = 0.7,
  raster = FALSE
) +
  ggtitle(expression(italic("Reichart et al.") ~ " Science 2022")) +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 10, face = "bold")
  )

(p_koenig_cc + p_reichart_cc) + patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "right")
```

## Phase proportions per sample

```{r cell cycle percentages per sample}
Koenig_EC$Sample <- Koenig_EC$orig.ident
Reichart_EC$Sample <- Reichart_EC$donor_id

df_phase_koenig <- Koenig_EC@meta.data %>%
  dplyr::group_by(Sample, Dataset, Group, Sex, Age, Phase) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(Sample, Dataset, Group, Sex, Age) %>%
  dplyr::mutate(
    total_n = sum(n),
    pct = n / total_n * 100
  ) %>%
  dplyr::ungroup()

df_phase_reichart <- Reichart_EC@meta.data %>%
  dplyr::group_by(Sample, Dataset, Group, Sex, Age, Phase) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(Sample, Dataset, Group, Sex, Age) %>%
  dplyr::mutate(
    total_n = sum(n),
    pct = n / total_n * 100
  ) %>%
  dplyr::ungroup()

df_phase <- dplyr::bind_rows(df_phase_koenig, df_phase_reichart)

df_phase$Group <- factor(df_phase$Group, levels = c("H", "HF"))
df_phase$Phase <- factor(df_phase$Phase, levels = c("G1", "S", "G2M"))
df_phase$Dataset <- factor(df_phase$Dataset, levels = c("Koenig", "Reichart"))
df_phase$Sex <- factor(df_phase$Sex)
df_phase$Age <- as.numeric(df_phase$Age)
```

## Cell-cycle statistics

```{r cell cycle stats}
phase_stats <- list()

for (ph in c("G1", "S", "G2M")) {
  df_sub <- df_phase %>% dplyr::filter(Phase == ph)

  fit_phase <- glm(
    cbind(n, total_n - n) ~ Group + Age + Sex + Dataset,
    family = quasibinomial,
    data = df_sub
  )

  phase_stats[[ph]] <- broom::tidy(fit_phase)
}

phase_stats_df <- dplyr::bind_rows(phase_stats, .id = "Phase")

write.csv(
  phase_stats_df,
  file = file.path(output_dir_data, "Cell_cycle_phase_stats.csv"),
  quote = FALSE,
  row.names = FALSE
)

phase_stats_df
```

## Cell-cycle barplots

```{r cell cycle barplots, fig.width=8, fig.height=3.5}
group_cols_cycle <- c("H" = "#D9D9D9", "HF" = "#7F7F7F")

plots_phase <- list()

for (ph in c("G1", "S", "G2M")) {
  df_sub <- df_phase %>% dplyr::filter(Phase == ph)

  fit_phase <- glm(
    cbind(n, total_n - n) ~ Group + Age + Sex + Dataset,
    family = quasibinomial,
    data = df_sub
  )

  fit_tidy <- broom::tidy(fit_phase)

  p_group <- fit_tidy %>%
    dplyr::filter(term == "GroupHF") %>%
    dplyr::pull(p.value)

  p_label <- format_p_label(p_group)

  p <- ggplot(df_sub, aes(x = Group, y = pct, fill = Group)) +
    stat_summary(fun = mean, geom = "bar", width = 0.65, color = "black") +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.15, linewidth = 0.5) +
    geom_jitter(
      width = 0.10,
      size = 1.1,
      shape = 21,
      stroke = 0.35,
      color = "black"
    ) +
    geom_segment(aes(x = 1, xend = 2, y = 46.4, yend = 46.4), inherit.aes = FALSE) +
    annotate("text", x = 1.5, y = 47.3, label = p_label, size = 5, fontface = format_p_fontface(p_group)) +
    scale_fill_manual(values = group_cols_cycle) +
    labs(title = ph, x = NULL, y = "Percentage of cells") +
    theme_classic(base_size = 12) +
    scale_y_continuous(limits = c(0, 50), expand = c(0, 0)) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.title.y = element_text(face = "bold"),
      legend.position = "none"
    )

  plots_phase[[ph]] <- p
}

plots_phase[["G1"]] + plots_phase[["S"]] + plots_phase[["G2M"]]
```

## Genes and scores of interest

```{r gene sets}
aerobic_respiration_row <- GSEA_GO_ALL@result %>%
  as_tibble() %>%
  dplyr::filter(Description == "aerobic respiration") %>%
  dplyr::slice(1)

if (nrow(aerobic_respiration_row) == 0) {
  stop("GO term 'aerobic respiration' was not found in GSEA_GO_ALL@result.")
}

aerobic_respiration_go_id <- aerobic_respiration_row$ID[1]

aerobic_respiration_entrez <- GSEA_GO_ALL@geneSets[[aerobic_respiration_go_id]]

aerobic_respiration_genes_df <- bitr(
  aerobic_respiration_entrez,
  fromType = "ENTREZID",
  toType = "SYMBOL",
  OrgDb = org.Hs.eg.db,
  drop = TRUE
)

aerobic_respiration_genes <- unique(aerobic_respiration_genes_df$SYMBOL)

write.csv(
  data.frame(gene = aerobic_respiration_genes),
  file = file.path(output_dir_data, "GO_aerobic_respiration_genes.csv"),
  quote = FALSE,
  row.names = FALSE
)

angiogenesis_genes <- c("ESM1", "ANGPT2", "KDR", "APLN", "CXCR4", "PGF", "DLL4", "KIT")

redox_no_genes <- c("NOX4", "NOS3", "SOD1")

mito_homeostasis_genes <- c("TFAM", "PPARGC1A")

apoptotic_priming_genes <- c(
  "BCL2L11", "BBC3", "PMAIP1", "BAX",
  "BID", "BAD", "BMF", "BIK",
  "BNIP3", "BNIP3L"
)

all_genes_of_interest <- unique(c(
  angiogenesis_genes,
  redox_no_genes,
  mito_homeostasis_genes,
  apoptotic_priming_genes,
  aerobic_respiration_genes
))
```

```{r pseudobulk scores}
Koenig_EC$Sample <- Koenig_EC$orig.ident
Reichart_EC$Sample <- Reichart_EC$donor_id

Koenig_EC$cell_type <- "Endothelium"
Reichart_EC$cell_type <- "Endothelium"

pb_koenig_scores <- Seurat2PB(
  Koenig_EC,
  sample = "Sample",
  cluster = "cell_type"
)

pb_reichart_scores <- Seurat2PB(
  Reichart_EC,
  sample = "Sample",
  cluster = "cell_type"
)

colnames(pb_koenig_scores) <- pb_koenig_scores$samples$sample
colnames(pb_reichart_scores) <- pb_reichart_scores$samples$sample

common_genes_scores <- intersect(rownames(pb_koenig_scores), rownames(pb_reichart_scores))

counts_combined_scores <- cbind(
  pb_koenig_scores[common_genes_scores, ]$counts,
  pb_reichart_scores[common_genes_scores, ]$counts
)

metadata_scores <- dplyr::bind_rows(
  Koenig_EC@meta.data %>%
    dplyr::select(Sample, Dataset, Group, Sex, Age) %>%
    dplyr::distinct(),
  Reichart_EC@meta.data %>%
    dplyr::select(Sample, Dataset, Group, Sex, Age) %>%
    dplyr::distinct()
)

metadata_scores <- metadata_scores %>%
  dplyr::distinct(Sample, .keep_all = TRUE) %>%
  dplyr::filter(Sample %in% colnames(counts_combined_scores))

metadata_scores <- metadata_scores[match(colnames(counts_combined_scores), metadata_scores$Sample), ]

stopifnot(all(metadata_scores$Sample == colnames(counts_combined_scores)))

metadata_scores$Group <- factor(metadata_scores$Group, levels = c("H", "HF"))
metadata_scores$Sex <- factor(metadata_scores$Sex)
metadata_scores$Dataset <- factor(metadata_scores$Dataset, levels = c("Koenig", "Reichart"))
metadata_scores$Age <- as.numeric(metadata_scores$Age)

dge_scores <- edgeR::DGEList(counts = counts_combined_scores)
dge_scores <- edgeR::calcNormFactors(dge_scores)

logcpm_scores <- edgeR::cpm(dge_scores, log = TRUE, prior.count = 1)
```

## Keep only genes of interest for scores and barplots

```{r score matrices}
all_genes_of_interest_present <- all_genes_of_interest[all_genes_of_interest %in% rownames(logcpm_scores)]
all_genes_of_interest_missing <- setdiff(all_genes_of_interest, rownames(logcpm_scores))

logcpm_subset_scores <- logcpm_scores[
  rownames(logcpm_scores) %in% all_genes_of_interest_present,
  ,
  drop = FALSE
]

logcpm_subset_z <- t(scale(t(logcpm_subset_scores)))
logcpm_subset_z[is.na(logcpm_subset_z)] <- 0

angiogenesis_score <- colMeans(
  logcpm_subset_z[rownames(logcpm_subset_z) %in% angiogenesis_genes, , drop = FALSE]
)

apoptotic_priming_score <- colMeans(
  logcpm_subset_z[rownames(logcpm_subset_z) %in% apoptotic_priming_genes, , drop = FALSE]
)

aerobic_respiration_score <- colMeans(
  logcpm_subset_z[rownames(logcpm_subset_z) %in% aerobic_respiration_genes, , drop = FALSE]
)

score_df <- metadata_scores %>%
  dplyr::mutate(
    angiogenesis_score = angiogenesis_score[Sample],
    apoptotic_priming_score = apoptotic_priming_score[Sample],
    aerobic_respiration_score = aerobic_respiration_score[Sample]
  )

write.csv(
  score_df,
  file = file.path(output_dir_data, "EC_gene_set_scores_unfilteredPB.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  data.frame(gene_missing = all_genes_of_interest_missing),
  file = file.path(output_dir_data, "EC_genes_of_interest_missing_unfilteredPB.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  data.frame(gene = setdiff(aerobic_respiration_genes, rownames(logcpm_scores))),
  file = file.path(output_dir_data, "GO_aerobic_respiration_genes_missing_in_scores_matrix.csv"),
  quote = FALSE,
  row.names = FALSE
)
```

## Pseudobulk score boxplots

```{r aerobic respiration score, fig.width=3, fig.height=3.7}
fit_aerobic_score <- lm(aerobic_respiration_score ~ Group + Age + Sex + Dataset, data = score_df)
fit_aerobic_tidy <- broom::tidy(fit_aerobic_score)
print(fit_aerobic_tidy)

p_aerobic <- fit_aerobic_tidy %>%
  dplyr::filter(term == "GroupHF") %>%
  dplyr::pull(p.value)

p_aerobic_label <- format_p_label(p_aerobic)

y_max_aerobic <- max(score_df$aerobic_respiration_score, na.rm = TRUE)
y_min_aerobic <- min(score_df$aerobic_respiration_score, na.rm = TRUE)
y_range_aerobic <- y_max_aerobic - y_min_aerobic
if (!is.finite(y_range_aerobic) || y_range_aerobic == 0) y_range_aerobic <- 1

ggplot(score_df, aes(x = Group, y = aerobic_respiration_score, fill = Group)) +
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    color = "black",
    linewidth = 0.5
  ) +
  geom_jitter(
    width = 0.10,
    size = 1.5,
    shape = 21,
    stroke = 0.4,
    color = "black"
  ) +
  geom_segment(
    aes(
      x = 1,
      xend = 2,
      y = y_max_aerobic + 0.10 * y_range_aerobic,
      yend = y_max_aerobic + 0.10 * y_range_aerobic
    ),
    inherit.aes = FALSE
  ) +
  annotate(
    "text",
    x = 1.5,
    y = y_max_aerobic + 0.16 * y_range_aerobic,
    label = p_aerobic_label,
    size = 5,
    fontface = format_p_fontface(p_aerobic)
  ) +
  scale_fill_manual(values = group_cols) +
  coord_cartesian(
    ylim = c(
      y_min_aerobic - 0.05 * y_range_aerobic,
      y_max_aerobic + 0.22 * y_range_aerobic
    )
  ) +
  labs(title = "Aerobic respiration score", x = NULL, y = "Score") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "none"
  )
```

```{r angiogenesis score, fig.width=3, fig.height=3.7}
fit_angio_score <- lm(angiogenesis_score ~ Group + Age + Sex + Dataset, data = score_df)
fit_angio_tidy <- broom::tidy(fit_angio_score)
print(fit_angio_tidy)

p_angio <- fit_angio_tidy %>%
  dplyr::filter(term == "GroupHF") %>%
  dplyr::pull(p.value)

p_angio_label <- format_p_label(p_angio)

y_max_angio <- max(score_df$angiogenesis_score, na.rm = TRUE)
y_min_angio <- min(score_df$angiogenesis_score, na.rm = TRUE)
y_range_angio <- y_max_angio - y_min_angio
if (!is.finite(y_range_angio) || y_range_angio == 0) y_range_angio <- 1

ggplot(score_df, aes(x = Group, y = angiogenesis_score, fill = Group)) +
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    color = "black",
    linewidth = 0.5
  ) +
  geom_jitter(
    width = 0.10,
    size = 1.5,
    shape = 21,
    stroke = 0.4,
    color = "black"
  ) +
  geom_segment(
    aes(
      x = 1,
      xend = 2,
      y = y_max_angio + 0.10 * y_range_angio,
      yend = y_max_angio + 0.10 * y_range_angio
    ),
    inherit.aes = FALSE
  ) +
  annotate(
    "text",
    x = 1.5,
    y = y_max_angio + 0.16 * y_range_angio,
    label = p_angio_label,
    size = 5,
    fontface = format_p_fontface(p_angio)
  ) +
  scale_fill_manual(values = group_cols) +
  coord_cartesian(
    ylim = c(
      y_min_angio - 0.05 * y_range_angio,
      y_max_angio + 0.22 * y_range_angio
    )
  ) +
  labs(title = "Angiogenesis score", x = NULL, y = "Score") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "none"
  )
```

```{r apoptotic priming score, fig.width=3, fig.height=3.7}
fit_apop_score <- lm(apoptotic_priming_score ~ Group + Age + Sex + Dataset, data = score_df)
fit_apop_tidy <- broom::tidy(fit_apop_score)
print(fit_apop_tidy)

p_apop <- fit_apop_tidy %>%
  dplyr::filter(term == "GroupHF") %>%
  dplyr::pull(p.value)

p_apop_label <- format_p_label(p_apop)

y_max_apop <- max(score_df$apoptotic_priming_score, na.rm = TRUE)
y_min_apop <- min(score_df$apoptotic_priming_score, na.rm = TRUE)
y_range_apop <- y_max_apop - y_min_apop
if (!is.finite(y_range_apop) || y_range_apop == 0) y_range_apop <- 1

ggplot(score_df, aes(x = Group, y = apoptotic_priming_score, fill = Group)) +
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    color = "black",
    linewidth = 0.5
  ) +
  geom_jitter(
    width = 0.10,
    size = 1.5,
    shape = 21,
    stroke = 0.4,
    color = "black"
  ) +
  geom_segment(
    aes(
      x = 1,
      xend = 2,
      y = y_max_apop + 0.10 * y_range_apop,
      yend = y_max_apop + 0.10 * y_range_apop
    ),
    inherit.aes = FALSE
  ) +
  annotate(
    "text",
    x = 1.5,
    y = y_max_apop + 0.16 * y_range_apop,
    label = p_apop_label,
    size = 5,
    fontface = format_p_fontface(p_apop)
  ) +
  scale_fill_manual(values = group_cols) +
  coord_cartesian(
    ylim = c(
      y_min_apop - 0.05 * y_range_apop,
      y_max_apop + 0.22 * y_range_apop
    )
  ) +
  labs(title = "Apoptotic priming score", x = NULL, y = "Score") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "none"
  )
```


## Feature plots for angiogenesis genes in each dataset separately

```{r feature plots angiogenesis, fig.width=11, fig.height=4.8}
for (gene in angiogenesis_genes) {

  p_list <- list()

  if (gene %in% rownames(Koenig_EC)) {
    p_koenig <- FeaturePlot(
      Koenig_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(
        bquote(atop(bold(.(gene)), italic("Koenig et al.") ~ " Nature CVR 2022"))
      ) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Koenig"]] <- p_koenig
  }

  if (gene %in% rownames(Reichart_EC)) {
    p_reichart <- FeaturePlot(
      Reichart_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(
        bquote(atop(bold(.(gene)), italic("Reichart et al.") ~ " Science 2022"))
      ) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Reichart"]] <- p_reichart
  }

  if (length(p_list) == 2) {
    print(p_list[["Koenig"]] | p_list[["Reichart"]])
  } else if (length(p_list) == 1) {
    print(p_list[[1]])
  } else {
    message("Gene not found in either dataset: ", gene)
  }
}
```

## Feature plots for redox / NO genes in each dataset separately

```{r feature plots redox no, fig.width=11, fig.height=4.8}
for (gene in redox_no_genes) {

  p_list <- list()

  if (gene %in% rownames(Koenig_EC)) {
    p_koenig <- FeaturePlot(
      Koenig_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(
        bquote(atop(bold(.(gene)), italic("Koenig et al.") ~ " Nature CVR 2022"))
      ) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Koenig"]] <- p_koenig
  }

  if (gene %in% rownames(Reichart_EC)) {
    p_reichart <- FeaturePlot(
      Reichart_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(
        bquote(atop(bold(.(gene)), italic("Reichart et al.") ~ " Science 2022"))
      ) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Reichart"]] <- p_reichart
  }

  if (length(p_list) == 2) {
    print(p_list[["Koenig"]] | p_list[["Reichart"]])
  } else if (length(p_list) == 1) {
    print(p_list[[1]])
  } else {
    message("Gene not found in either dataset: ", gene)
  }
}
```

## Feature plots for mitochondrial homeostasis genes in each dataset separately

```{r feature plots mito homeostasis, fig.width=11, fig.height=4.8}
for (gene in mito_homeostasis_genes) {

  p_list <- list()

  if (gene %in% rownames(Koenig_EC)) {
    p_koenig <- FeaturePlot(
      Koenig_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(
        bquote(atop(bold(.(gene)), italic("Koenig et al.") ~ " Nature CVR 2022"))
      ) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Koenig"]] <- p_koenig
  }

  if (gene %in% rownames(Reichart_EC)) {
    p_reichart <- FeaturePlot(
      Reichart_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(
        bquote(atop(bold(.(gene)), italic("Reichart et al.") ~ " Science 2022"))
      ) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Reichart"]] <- p_reichart
  }

  if (length(p_list) == 2) {
    print(p_list[["Koenig"]] | p_list[["Reichart"]])
  } else if (length(p_list) == 1) {
    print(p_list[[1]])
  } else {
    message("Gene not found in either dataset: ", gene)
  }
}
```

## Feature plots for apoptotic priming genes in each dataset separately

```{r feature plots apoptosis, fig.width=11, fig.height=4.8}
for (gene in apoptotic_priming_genes) {

  p_list <- list()

  if (gene %in% rownames(Koenig_EC)) {
    p_koenig <- FeaturePlot(
      Koenig_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(
        bquote(atop(bold(.(gene)), italic("Koenig et al.") ~ " Nature CVR 2022"))
      ) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Koenig"]] <- p_koenig
  }

  if (gene %in% rownames(Reichart_EC)) {
    p_reichart <- FeaturePlot(
      Reichart_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(
        bquote(atop(bold(.(gene)), italic("Reichart et al.") ~ " Science 2022"))
      ) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Reichart"]] <- p_reichart
  }

  if (length(p_list) == 2) {
    print(p_list[["Koenig"]] | p_list[["Reichart"]])
  } else if (length(p_list) == 1) {
    print(p_list[[1]])
  } else {
    message("Gene not found in either dataset: ", gene)
  }
}
```


## Gene-wise pseudobulk logCPM plots with model-based p values

```{r gene wise dataframe}
gene_barplot_df <- logcpm_scores %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene") %>%
  tidyr::pivot_longer(-gene, names_to = "Sample", values_to = "logCPM") %>%
  dplyr::left_join(metadata_scores, by = "Sample")
```


```{r barplots angiogenesis genes, fig.width=2.66, fig.height=3.7}
for (gene in angiogenesis_genes) {
  if (gene %in% unique(gene_barplot_df$gene)) {
    df_sub <- gene_barplot_df[gene_barplot_df$gene == gene, ]
    fit_gene <- lm(logCPM ~ Group + Age + Sex + Dataset, data = df_sub)
    fit_gene_tidy <- broom::tidy(fit_gene)
    print(fit_gene_tidy)

    p_gene <- fit_gene_tidy %>%
      dplyr::filter(term == "GroupHF") %>%
      dplyr::pull(p.value)

    p_gene_label <- format_p_label(p_gene)

    y_max_gene <- max(df_sub$logCPM, na.rm = TRUE)
    y_min_gene <- min(df_sub$logCPM, na.rm = TRUE)
    y_range_gene <- y_max_gene - y_min_gene
    if (!is.finite(y_range_gene) || y_range_gene == 0) y_range_gene <- 1

    p <- ggplot(df_sub, aes(x = Group, y = logCPM, fill = Group)) +
      geom_boxplot(
        width = 0.55,
        outlier.shape = NA,
        color = "black",
        linewidth = 0.5
      ) +
      geom_jitter(
        width = 0.10,
        size = 1.5,
        shape = 21,
        stroke = 0.4,
        color = "black"
      ) +
      geom_segment(
        aes(
          x = 1,
          xend = 2,
          y = y_max_gene + 0.10 * y_range_gene,
          yend = y_max_gene + 0.10 * y_range_gene
        ),
        inherit.aes = FALSE
      ) +
      annotate(
        "text",
        x = 1.5,
        y = y_max_gene + 0.16 * y_range_gene,
        label = p_gene_label,
        size = 5,
        fontface = "bold"
      ) +
      scale_fill_manual(values = group_cols) +
      coord_cartesian(
        ylim = c(
          y_min_gene - 0.05 * y_range_gene,
          y_max_gene + 0.22 * y_range_gene
        )
      ) +
      labs(title = gene, x = NULL, y = "logCPM") +
      theme_classic(base_size = 12) +
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.position = "none"
      )

    print(p)
  }
}
```


```{r barplots redox no genes, fig.width=2.66, fig.height=3.7}
for (gene in redox_no_genes) {
  if (gene %in% unique(gene_barplot_df$gene)) {
    df_sub <- gene_barplot_df[gene_barplot_df$gene == gene, ]
    fit_gene <- lm(logCPM ~ Group + Age + Sex + Dataset, data = df_sub)
    fit_gene_tidy <- broom::tidy(fit_gene)
    print(fit_gene_tidy)

    p_gene <- fit_gene_tidy %>%
      dplyr::filter(term == "GroupHF") %>%
      dplyr::pull(p.value)

    p_gene_label <- format_p_label(p_gene)

    y_max_gene <- max(df_sub$logCPM, na.rm = TRUE)
    y_min_gene <- min(df_sub$logCPM, na.rm = TRUE)
    y_range_gene <- y_max_gene - y_min_gene
    if (!is.finite(y_range_gene) || y_range_gene == 0) y_range_gene <- 1

    p <- ggplot(df_sub, aes(x = Group, y = logCPM, fill = Group)) +
      geom_boxplot(
        width = 0.55,
        outlier.shape = NA,
        color = "black",
        linewidth = 0.5
      ) +
      geom_jitter(
        width = 0.10,
        size = 1.5,
        shape = 21,
        stroke = 0.4,
        color = "black"
      ) +
      geom_segment(
        aes(
          x = 1,
          xend = 2,
          y = y_max_gene + 0.10 * y_range_gene,
          yend = y_max_gene + 0.10 * y_range_gene
        ),
        inherit.aes = FALSE
      ) +
      annotate(
        "text",
        x = 1.5,
        y = y_max_gene + 0.16 * y_range_gene,
        label = p_gene_label,
        size = 5,
        fontface = "bold"
      ) +
      scale_fill_manual(values = group_cols) +
      coord_cartesian(
        ylim = c(
          y_min_gene - 0.05 * y_range_gene,
          y_max_gene + 0.22 * y_range_gene
        )
      ) +
      labs(title = gene, x = NULL, y = "logCPM") +
      theme_classic(base_size = 12) +
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.position = "none"
      )

    print(p)
  }
}
```


```{r barplots mito homeostasis genes, fig.width=2.66, fig.height=3.7}
for (gene in mito_homeostasis_genes) {
  if (gene %in% unique(gene_barplot_df$gene)) {
    df_sub <- gene_barplot_df[gene_barplot_df$gene == gene, ]
    fit_gene <- lm(logCPM ~ Group + Age + Sex + Dataset, data = df_sub)
    fit_gene_tidy <- broom::tidy(fit_gene)
    print(fit_gene_tidy)

    p_gene <- fit_gene_tidy %>%
      dplyr::filter(term == "GroupHF") %>%
      dplyr::pull(p.value)

    p_gene_label <- format_p_label(p_gene)

    y_max_gene <- max(df_sub$logCPM, na.rm = TRUE)
    y_min_gene <- min(df_sub$logCPM, na.rm = TRUE)
    y_range_gene <- y_max_gene - y_min_gene
    if (!is.finite(y_range_gene) || y_range_gene == 0) y_range_gene <- 1

    p <- ggplot(df_sub, aes(x = Group, y = logCPM, fill = Group)) +
      geom_boxplot(
        width = 0.55,
        outlier.shape = NA,
        color = "black",
        linewidth = 0.5
      ) +
      geom_jitter(
        width = 0.10,
        size = 1.5,
        shape = 21,
        stroke = 0.4,
        color = "black"
      ) +
      geom_segment(
        aes(
          x = 1,
          xend = 2,
          y = y_max_gene + 0.10 * y_range_gene,
          yend = y_max_gene + 0.10 * y_range_gene
        ),
        inherit.aes = FALSE
      ) +
      annotate(
        "text",
        x = 1.5,
        y = y_max_gene + 0.16 * y_range_gene,
        label = p_gene_label,
        size = 5,
        fontface = "bold"
      ) +
      scale_fill_manual(values = group_cols) +
      coord_cartesian(
        ylim = c(
          y_min_gene - 0.05 * y_range_gene,
          y_max_gene + 0.22 * y_range_gene
        )
      ) +
      labs(title = gene, x = NULL, y = "logCPM") +
      theme_classic(base_size = 12) +
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.position = "none"
      )

    print(p)
  }
}
```


```{r barplots apoptotic priming genes, fig.width=2.66, fig.height=3.7}
for (gene in apoptotic_priming_genes) {
  if (gene %in% unique(gene_barplot_df$gene)) {
    df_sub <- gene_barplot_df[gene_barplot_df$gene == gene, ]
    fit_gene <- lm(logCPM ~ Group + Age + Sex + Dataset, data = df_sub)
    fit_gene_tidy <- broom::tidy(fit_gene)
    print(fit_gene_tidy)

    p_gene <- fit_gene_tidy %>%
      dplyr::filter(term == "GroupHF") %>%
      dplyr::pull(p.value)

    p_gene_label <- format_p_label(p_gene)

    y_max_gene <- max(df_sub$logCPM, na.rm = TRUE)
    y_min_gene <- min(df_sub$logCPM, na.rm = TRUE)
    y_range_gene <- y_max_gene - y_min_gene
    if (!is.finite(y_range_gene) || y_range_gene == 0) y_range_gene <- 1

    p <- ggplot(df_sub, aes(x = Group, y = logCPM, fill = Group)) +
      geom_boxplot(
        width = 0.55,
        outlier.shape = NA,
        color = "black",
        linewidth = 0.5
      ) +
      geom_jitter(
        width = 0.10,
        size = 1.5,
        shape = 21,
        stroke = 0.4,
        color = "black"
      ) +
      geom_segment(
        aes(
          x = 1,
          xend = 2,
          y = y_max_gene + 0.10 * y_range_gene,
          yend = y_max_gene + 0.10 * y_range_gene
        ),
        inherit.aes = FALSE
      ) +
      annotate(
        "text",
        x = 1.5,
        y = y_max_gene + 0.16 * y_range_gene,
        label = p_gene_label,
        size = 5,
        fontface = "bold"
      ) +
      scale_fill_manual(values = group_cols) +
      coord_cartesian(
        ylim = c(
          y_min_gene - 0.05 * y_range_gene,
          y_max_gene + 0.22 * y_range_gene
        )
      ) +
      labs(title = gene, x = NULL, y = "logCPM") +
      theme_classic(base_size = 12) +
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.position = "none"
      )

    print(p)
  }
}
```

## Cell-cycle program scores 

```{r cell cycle gene sets}
s_phase_genes <- Seurat::cc.genes.updated.2019$s.genes
g2m_phase_genes <- Seurat::cc.genes.updated.2019$g2m.genes

s_phase_genes_present <- intersect(s_phase_genes, rownames(logcpm_scores))
g2m_phase_genes_present <- intersect(g2m_phase_genes, rownames(logcpm_scores))

s_phase_genes_missing <- setdiff(s_phase_genes, rownames(logcpm_scores))
g2m_phase_genes_missing <- setdiff(g2m_phase_genes, rownames(logcpm_scores))

length(s_phase_genes_present)
length(g2m_phase_genes_present)
```

## Build score matrix for cell-cycle programs

```{r cell cycle score matrix}
cell_cycle_genes_all <- unique(c(
  s_phase_genes_present,
  g2m_phase_genes_present
  
))

logcpm_subset_cc <- logcpm_scores[
  rownames(logcpm_scores) %in% cell_cycle_genes_all,
  ,
  drop = FALSE
]

logcpm_subset_cc_z <- t(scale(t(logcpm_subset_cc)))
logcpm_subset_cc_z[is.na(logcpm_subset_cc_z)] <- 0

s_phase_score <- colMeans(
  logcpm_subset_cc_z[rownames(logcpm_subset_cc_z) %in% s_phase_genes_present, , drop = FALSE]
)

g2m_phase_score <- colMeans(
  logcpm_subset_cc_z[rownames(logcpm_subset_cc_z) %in% g2m_phase_genes_present, , drop = FALSE]
)

score_df_cc <- metadata_scores %>%
  dplyr::mutate(
    S_phase_score = s_phase_score[Sample],
    G2M_phase_score = g2m_phase_score[Sample]
  )

write.csv(
  score_df_cc,
  file = file.path(output_dir_data, "EC_cell_cycle_program_scores.csv"),
  quote = FALSE,
  row.names = FALSE
)
```

## Cell-cycle program score boxplots

```{r s phase score, fig.width=2.8, fig.height=3.7}
fit_s_score <- lm(S_phase_score ~ Group + Age + Sex + Dataset, data = score_df_cc)
fit_s_tidy <- broom::tidy(fit_s_score)
print(fit_s_tidy)

p_s <- fit_s_tidy %>%
  dplyr::filter(term == "GroupHF") %>%
  dplyr::pull(p.value)

p_s_label <- format_p_label(p_s)

y_max_s <- max(score_df_cc$S_phase_score, na.rm = TRUE)
y_min_s <- min(score_df_cc$S_phase_score, na.rm = TRUE)
y_range_s <- y_max_s - y_min_s
if (!is.finite(y_range_s) || y_range_s == 0) y_range_s <- 1

ggplot(score_df_cc, aes(x = Group, y = S_phase_score, fill = Group)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, color = "black", linewidth = 0.5) +
  geom_jitter(width = 0.10, size = 1.5, shape = 21, stroke = 0.4, color = "black") +
  geom_segment(
    aes(x = 1, xend = 2, y = y_max_s + 0.10 * y_range_s, yend = y_max_s + 0.10 * y_range_s),
    inherit.aes = FALSE
  ) +
  annotate("text", x = 1.5, y = y_max_s + 0.16 * y_range_s, label = p_s_label, size = 5, fontface = "bold") +
  scale_fill_manual(values = group_cols) +
  coord_cartesian(ylim = c(y_min_s - 0.05 * y_range_s, y_max_s + 0.22 * y_range_s)) +
  labs(title = "S-phase genes", x = NULL, y = "Score") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "none"
  )
```

```{r g2m phase score, fig.width=2.8, fig.height=3.7}
fit_g2m_score <- lm(G2M_phase_score ~ Group + Age + Sex + Dataset, data = score_df_cc)
fit_g2m_tidy <- broom::tidy(fit_g2m_score)
print(fit_g2m_tidy)

p_g2m <- fit_g2m_tidy %>%
  dplyr::filter(term == "GroupHF") %>%
  dplyr::pull(p.value)

p_g2m_label <- format_p_label(p_g2m)

y_max_g2m <- max(score_df_cc$G2M_phase_score, na.rm = TRUE)
y_min_g2m <- min(score_df_cc$G2M_phase_score, na.rm = TRUE)
y_range_g2m <- y_max_g2m - y_min_g2m
if (!is.finite(y_range_g2m) || y_range_g2m == 0) y_range_g2m <- 1

ggplot(score_df_cc, aes(x = Group, y = G2M_phase_score, fill = Group)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, color = "black", linewidth = 0.5) +
  geom_jitter(width = 0.10, size = 1.5, shape = 21, stroke = 0.4, color = "black") +
  geom_segment(
    aes(x = 1, xend = 2, y = y_max_g2m + 0.10 * y_range_g2m, yend = y_max_g2m + 0.10 * y_range_g2m),
    inherit.aes = FALSE
  ) +
  annotate("text", x = 1.5, y = y_max_g2m + 0.16 * y_range_g2m, label = p_g2m_label, size = 5, fontface = "bold") +
  scale_fill_manual(values = group_cols) +
  coord_cartesian(ylim = c(y_min_g2m - 0.05 * y_range_g2m, y_max_g2m + 0.22 * y_range_g2m)) +
  labs(title = "G2/M-phase genes", x = NULL, y = "Score") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "none"
  )
```

## Representative genes for each program

```{r representative genes for cell cycle programs}
s_phase_genes_plot <- c("MCM2", "MCM5", "PCNA", "TYMS", "RRM2", "TK1")
g2m_phase_genes_plot <- c("TOP2A", "UBE2C", "CENPF", "BIRC5", "CDC20", "AURKB")
```

```{r barplots s phase genes, fig.width=2.66, fig.height=3.7}
for (gene in s_phase_genes_plot[s_phase_genes_plot %in% unique(gene_barplot_df$gene)]) {

  df_sub <- gene_barplot_df[gene_barplot_df$gene == gene, ]

  fit_gene <- lm(logCPM ~ Group + Age + Sex + Dataset, data = df_sub)
  fit_gene_tidy <- broom::tidy(fit_gene)
  print(fit_gene_tidy)

  p_gene <- fit_gene_tidy %>%
    dplyr::filter(term == "GroupHF") %>%
    dplyr::pull(p.value)

  p_gene_label <- format_p_label(p_gene)

  y_max_gene <- max(df_sub$logCPM, na.rm = TRUE)
  y_min_gene <- min(df_sub$logCPM, na.rm = TRUE)
  y_range_gene <- y_max_gene - y_min_gene
  if (!is.finite(y_range_gene) || y_range_gene == 0) y_range_gene <- 1

  p <- ggplot(df_sub, aes(x = Group, y = logCPM, fill = Group)) +
    geom_boxplot(width = 0.55, outlier.shape = NA, color = "black", linewidth = 0.5) +
    geom_jitter(width = 0.10, size = 1.5, shape = 21, stroke = 0.4, color = "black") +
    geom_segment(aes(x = 1, xend = 2, y = y_max_gene + 0.10 * y_range_gene, yend = y_max_gene + 0.10 * y_range_gene), inherit.aes = FALSE) +
    annotate("text", x = 1.5, y = y_max_gene + 0.16 * y_range_gene, label = p_gene_label, size = 5, fontface = "bold") +
    scale_fill_manual(values = group_cols) +
    coord_cartesian(ylim = c(y_min_gene - 0.05 * y_range_gene, y_max_gene + 0.22 * y_range_gene)) +
    labs(title = gene, x = NULL, y = "logCPM") +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.title.y = element_text(face = "bold"),
      legend.position = "none"
    )

  print(p)
}
```

```{r barplots g2m phase genes, fig.width=2.66, fig.height=3.7}
for (gene in g2m_phase_genes_plot[g2m_phase_genes_plot %in% unique(gene_barplot_df$gene)]) {

  df_sub <- gene_barplot_df[gene_barplot_df$gene == gene, ]

  fit_gene <- lm(logCPM ~ Group + Age + Sex + Dataset, data = df_sub)
  fit_gene_tidy <- broom::tidy(fit_gene)
  print(fit_gene_tidy)

  p_gene <- fit_gene_tidy %>%
    dplyr::filter(term == "GroupHF") %>%
    dplyr::pull(p.value)

  p_gene_label <- format_p_label(p_gene)

  y_max_gene <- max(df_sub$logCPM, na.rm = TRUE)
  y_min_gene <- min(df_sub$logCPM, na.rm = TRUE)
  y_range_gene <- y_max_gene - y_min_gene
  if (!is.finite(y_range_gene) || y_range_gene == 0) y_range_gene <- 1

  p <- ggplot(df_sub, aes(x = Group, y = logCPM, fill = Group)) +
    geom_boxplot(width = 0.55, outlier.shape = NA, color = "black", linewidth = 0.5) +
    geom_jitter(width = 0.10, size = 1.5, shape = 21, stroke = 0.4, color = "black") +
    geom_segment(aes(x = 1, xend = 2, y = y_max_gene + 0.10 * y_range_gene, yend = y_max_gene + 0.10 * y_range_gene), inherit.aes = FALSE) +
    annotate("text", x = 1.5, y = y_max_gene + 0.16 * y_range_gene, label = p_gene_label, size = 5, fontface = "bold") +
    scale_fill_manual(values = group_cols) +
    coord_cartesian(ylim = c(y_min_gene - 0.05 * y_range_gene, y_max_gene + 0.22 * y_range_gene)) +
    labs(title = gene, x = NULL, y = "logCPM") +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.title.y = element_text(face = "bold"),
      legend.position = "none"
    )

  print(p)
}
```

## Endothelial-cell subclustering across datasets

```{r merge endothelial objects}
Koenig_EC_sub <- Koenig_EC
Reichart_EC_sub <- Reichart_EC

Koenig_EC_sub$Dataset <- "Koenig"
Reichart_EC_sub$Dataset <- "Reichart"

Koenig_EC_sub$Sample <- Koenig_EC_sub$orig.ident
Reichart_EC_sub$Sample <- Reichart_EC_sub$donor_id

Koenig_EC_sub$percent.mt <- Koenig_EC_sub$percent.mito
Reichart_EC_sub$percent.mt <- Reichart_EC_sub$percent_mito

Koenig_EC_sub@assays$SCT <- NULL
DefaultAssay(Koenig_EC_sub) <- "RNA"
Koenig_EC_sub$Group <- factor(Koenig_EC_sub$Group, levels = c("H", "HF"))
Reichart_EC_sub$Group <- factor(Reichart_EC_sub$Group, levels = c("H", "HF"))
Koenig_EC_sub$Dataset <- factor(Koenig_EC_sub$Dataset, levels = c("Koenig", "Reichart"))
Reichart_EC_sub$Dataset <- factor(Reichart_EC_sub$Dataset, levels = c("Koenig", "Reichart"))

Koenig_EC_sub <- RenameCells(Koenig_EC_sub, add.cell.id = "Koenig")
Reichart_EC_sub <- RenameCells(Reichart_EC_sub, add.cell.id = "Reichart")

common_genes_ec <- intersect(rownames(Koenig_EC_sub), rownames(Reichart_EC_sub))
length(common_genes_ec)

EC_combined <- merge(
  x = Koenig_EC_sub[common_genes_ec, ],
  y = Reichart_EC_sub[common_genes_ec, ],
  merge.data = FALSE
)

rm(Koenig_EC_sub, Reichart_EC_sub)
```

```{r qc metadata combined endothelial}
EC_combined$nCount_RNA <- Matrix::colSums(EC_combined@assays$RNA@counts)
EC_combined$nFeature_RNA <- Matrix::colSums(EC_combined@assays$RNA@counts > 0)

EC_combined@meta.data %>%
  dplyr::count(Dataset, Group)

EC_combined@meta.data %>%
  dplyr::count(Dataset, Sample, sort = TRUE) %>%
  reactable::reactable()
```

```{r sctransform endothelial combined}
DefaultAssay(EC_combined) <- "RNA"
EC_list <- SplitObject(EC_combined, split.by = "Sample")

for (i in seq_along(EC_list)) {
  EC_list[[i]] <- SCTransform(
    EC_list[[i]],
    vst.flavor = "v2",
    vars.to.regress = "percent.mt",
    return.only.var.genes = TRUE,
    variable.features.n = 3000,
    verbose = FALSE
  )
}

features_ec <- SelectIntegrationFeatures(object.list = EC_list, nfeatures = 3000)

EC_combined <- merge(
  x = EC_list[[1]],
  y = EC_list[2:length(EC_list)],
  merge.data = TRUE
)
VariableFeatures(EC_combined) <- features_ec

rm(EC_list)
```

```{r pca endothelial combined, fig.width=6, fig.height=4}
EC_combined <- RunPCA(EC_combined, npcs = 50, verbose = FALSE)
ElbowPlot(EC_combined, ndims = 50)
```

```{r harmony endothelial combined}
EC_combined <- harmony::RunHarmony(
  object = EC_combined,
  assay.use = "SCT",
  group.by.vars = c("Sample", "Dataset"),
  dims.use = 1:50,
  plot_convergence = TRUE,
  .options = harmony::harmony_options(max.iter.cluster = 100)
)
```

```{r}
resolutions_to_keep <- paste0("SCT_snn.res.", c("0.1", "0.2", "0.3", "0.4", "0.5", "0.6", "0.7", "0.8"))

sct_res_cols <- grep("^SCT_snn[\\._]res\\.", colnames(EC_combined@meta.data), value = TRUE)

cols_to_remove <- setdiff(sct_res_cols, resolutions_to_keep)

if (length(cols_to_remove) > 0) {
  EC_combined@meta.data <- EC_combined@meta.data[, !colnames(EC_combined@meta.data) %in% cols_to_remove, drop = FALSE]
}

grep("^SCT_snn[\\._]res\\.", colnames(EC_combined@meta.data), value = TRUE)
```

```{r clustering endothelial combined}
DefaultAssay(EC_combined) <- "SCT"

EC_combined <- RunUMAP(
  EC_combined,
  reduction = "harmony",
  dims = 1:20
)

EC_combined <- FindNeighbors(
  EC_combined,
  reduction = "harmony",
  dims = 1:20
)

EC_combined <- FindClusters(
  EC_combined,
  resolution = seq(0.1, 0.8, by = 0.1)
)

Idents(EC_combined) <- "SCT_snn_res.0.3"
```

```{r clustree, fig.width=8, fig.height=10}
clustree::clustree(EC_combined@meta.data[,grep("SCT_snn_res", colnames(EC_combined@meta.data))], prefix = "SCT_snn_res.")
```

```{r umaps endothelial combined, fig.width=12, fig.height=10}
p1 <- DimPlot(
  EC_combined,
  reduction = "umap",
  group.by = "SCT_snn_res.0.3",
  label = TRUE,
  repel = TRUE,
  shuffle = TRUE,
  pt.size = 0.1
) +
  NoLegend() +
  ggtitle("Combined endothelial subclusters") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

p2 <- DimPlot(
  EC_combined,
  reduction = "umap",
  group.by = "Dataset",
  cols = dataset_cols,
  shuffle = TRUE,
  pt.size = 0.1
) +
  ggtitle("Dataset") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

p3 <- DimPlot(
  EC_combined,
  reduction = "umap",
  group.by = "Group",
  cols = group_cols,
  shuffle = TRUE,
  pt.size = 0.1
) +
  ggtitle("Group") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

p4 <- DimPlot(
  EC_combined,
  reduction = "umap",
  group.by = "Sample",
  shuffle = TRUE,
  pt.size = 0.1
) +
  NoLegend() +
  ggtitle("Sample") +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

(p1 | p2) / (p3 | p4)
```

```{r cell_type_cols}
cell_type_cols <- c("#d05050", "#33A02C", "#1F78B4", "#CAB2D6", "#FF7F00", "#B2DF8A", "#B15928", "#FFD000", "#FA95C0", "#FDBF6F", "#6A3D9A", "#7aCDF9", "#E6E600")
```

```{r EC_combined_qc, fig.height=5, fig.width=15}
VlnPlot_scCustom(
  EC_combined,
  features = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
  num_columns = 3,
  pt.size = 0,
  colors_use = cell_type_cols
) + theme(axis.title.x = element_blank())
```

```{r endothelial canonical marker featureplots, fig.height=4, fig.width=12}
markers_capillary <- c("RGCC", "CA4", "KDR")
markers_venous <- c("ACKR1", "SELP", "VWF")
markers_arterial <- c("EFNB2", "GJA5", "CXCL12")
markers_angiogenic <- c("KDR", "ANGPT2", "APLN")
markers_inflammatory <- c("IFITM3", "IFI6", "CXCL10")
markers_proliferating <- c("MKI67", "TOP2A", "CENPF")

marker_lists <- list(
  Capillary = markers_capillary,
  Venous = markers_venous,
  Arterial = markers_arterial,
  Angiogenic_activated = markers_angiogenic,
  Interferon_inflammatory = markers_inflammatory,
  Proliferating = markers_proliferating
)

marker_lists_present <- lapply(marker_lists, function(x) x[x %in% rownames(EC_combined)])
marker_lists_missing <- lapply(marker_lists, function(x) setdiff(x, rownames(EC_combined)))
marker_lists_missing

for (nm in names(marker_lists_present)) {
  genes_use <- marker_lists_present[[nm]]

  if (length(genes_use) == 0) {
    message("Skipping ", nm, ": no markers found in object")
    next
  }

  p <- FeaturePlot(
    EC_combined,
    features = genes_use,
    reduction = "umap",
    order = TRUE,
    ncol = length(genes_use)
  ) &
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.text = element_blank(),
      axis.ticks = element_blank()
    )

  print(p)
}
```

```{r cluster proportions by dataset and group, fig.width=12, fig.height=5}
df_prop_ec <- EC_combined@meta.data %>%
  dplyr::count(Dataset, Group, Sample, cluster = SCT_snn_res.0.3) %>%
  dplyr::group_by(Dataset, Group, Sample) %>%
  dplyr::mutate(freq = n / sum(n)) %>%
  dplyr::ungroup()

p_prop_dataset <- ggplot(df_prop_ec, aes(x = cluster, y = freq, fill = Dataset)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.2, alpha = 0.8) +
  scale_fill_manual(values = dataset_cols) +
  labs(x = "Cluster", y = "Fraction of endothelial cells") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.title = element_blank()
  )

p_prop_group <- ggplot(df_prop_ec, aes(x = cluster, y = freq, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.2, alpha = 0.8) +
  scale_fill_manual(values = group_cols) +
  labs(x = "Cluster", y = "Fraction of endothelial cells") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.title = element_blank()
  )

p_prop_dataset | p_prop_group
```

```{r markers endothelial combined_2}
DefaultAssay(EC_combined) <- "RNA"
EC_combined <- NormalizeData(EC_combined)
Idents(EC_combined) <- "SCT_snn_res.0.3"

EC_markers_combined <- FindAllMarkers(
  EC_combined,
  assay = "RNA",
  only.pos = TRUE,
  min.pct = 0.1,
  logfc.threshold = 0.25,
  return.thresh = 0.05
)

EC_markers_combined <- EC_markers_combined %>%
  dplyr::arrange(cluster, desc(avg_log2FC))

EC_markers_top10_combined <- EC_markers_combined %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 10, with_ties = FALSE) %>%
  dplyr::ungroup()

EC_markers_top3_combined <- EC_markers_combined %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 3, with_ties = FALSE) %>%
  dplyr::ungroup()

EC_markers_combined %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_head(n = 20) %>%
  reactable::reactable()
```

```{r write cluster markers_2, eval=FALSE}
write.csv(
  EC_markers_combined,
  here::here(output_dir_data, "EC_markers_combined_all.csv"),
  row.names = FALSE
)
write.csv(
  EC_markers_top10_combined,
  here::here(output_dir_data, "EC_markers_combined_top10.csv"),
  row.names = FALSE
)
write.csv(
  EC_markers_top3_combined,
  here::here(output_dir_data, "EC_markers_combined_top3.csv"),
  row.names = FALSE
)
```

```{r annotate endothelial cell states}
EC_combined@meta.data[EC_combined@meta.data$SCT_snn_res.0.3 %in% c("0"), "cell_state"] = "EC_capillary"
EC_combined@meta.data[EC_combined@meta.data$SCT_snn_res.0.3 %in% c("1"), "cell_state"] = "EC_arterial_PCSK5"
EC_combined@meta.data[EC_combined@meta.data$SCT_snn_res.0.3 %in% c("2"), "cell_state"] = "EC_venous"
EC_combined@meta.data[EC_combined@meta.data$SCT_snn_res.0.3 %in% c("3"), "cell_state"] = "EC_interferon"
EC_combined@meta.data[EC_combined@meta.data$SCT_snn_res.0.3 %in% c("4"), "cell_state"] = "EC_activated"
EC_combined@meta.data[EC_combined@meta.data$SCT_snn_res.0.3 %in% c("5"), "cell_state"] = "EC_inflammatory"
EC_combined@meta.data[EC_combined@meta.data$SCT_snn_res.0.3 %in% c("6"), "cell_state"] = "EC_arterial_NKAIN2"
EC_combined@meta.data[EC_combined@meta.data$SCT_snn_res.0.3 %in% c("7"), "cell_state"] = "EC_CM_contamination"

EC_combined$cell_state <- factor(EC_combined$cell_state, levels = c("EC_capillary", "EC_arterial_PCSK5", "EC_venous", "EC_interferon", "EC_activated", "EC_inflammatory", "EC_arterial_NKAIN2", "EC_CM_contamination"))
Idents(EC_combined) <- EC_combined$cell_state
```

```{r}
EC_combined <- subset(EC_combined, cell_state %in% c("EC_capillary", "EC_arterial_PCSK5", "EC_venous", "EC_interferon", "EC_activated", "EC_inflammatory", "EC_arterial_NKAIN2"))
```

```{r save EC_combined_subclustering qs2, eval=FALSE}
DefaultAssay(EC_combined) <- "RNA"
qs_save(
  EC_combined,
  file = here::here(output_dir_data, "EC_combined.qs2"),
  nthreads = 32
)
```

```{r read EC_combined_subclustering qs2}
EC_combined <- qs_read(
  file = here::here(output_dir_data, "EC_combined.qs2"),
  nthreads = 32
)
```

```{r UMAP cell states, fig.width=7, fig.height=6}
plot_group <- if ("cell_state" %in% colnames(EC_combined@meta.data)) "cell_state" else "SCT_snn_res.0.3"

p <- DimPlot(
  EC_combined,
  group.by = plot_group,
  reduction = "umap",
  cols = cell_type_cols,
  label = FALSE,
  shuffle = TRUE,
  pt.size = 0.1
) +
  NoLegend() +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_blank()
  )

LabelClusters(
  p,
  id = plot_group,
  fontface = "bold",
  size = 5,
  repel = TRUE
)
```

```{r markers endothelial combined}
EC_markers_combined <- FindAllMarkers(
  EC_combined,
  assay = "RNA",
  only.pos = TRUE,
  min.pct = 0.1,
  logfc.threshold = 0.25,
  return.thresh = 0.05
)

EC_markers_combined <- EC_markers_combined %>%
  dplyr::arrange(cluster, desc(avg_log2FC))

EC_markers_top10_combined <- EC_markers_combined %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 10, with_ties = FALSE) %>%
  dplyr::ungroup()

EC_markers_top3_combined <- EC_markers_combined %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 3, with_ties = FALSE) %>%
  dplyr::ungroup()

EC_markers_combined %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_head(n = 20) %>%
  reactable::reactable()
```

```{r write cluster markers, eval=FALSE}
write.csv(
  EC_markers_combined,
  here::here(output_dir_data, "EC_markers_combined_all_cell_state.csv"),
  row.names = FALSE
)
write.csv(
  EC_markers_top10_combined,
  here::here(output_dir_data, "EC_markers_combined_top10_cell_state.csv"),
  row.names = FALSE
)
write.csv(
  EC_markers_top3_combined,
  here::here(output_dir_data, "EC_markers_combined_top3_cell_state.csv"),
  row.names = FALSE
)
```

```{r dotplot top markers, fig.width=10, fig.height=5}
id_backup <- Idents(EC_combined)
Idents(EC_combined) <- factor(Idents(EC_combined), levels = rev(levels(Idents(EC_combined))))

p_dot <- DotPlot_scCustom(
  EC_combined,
  features = unique(EC_markers_top3_combined$gene),
  x_lab_rotate = TRUE
)

p_dot + theme(
  axis.text.x = element_text(size = 11, angle = 60, vjust = 1, hjust = 1),
  axis.text.y = element_text(size = 12)
)

Idents(EC_combined) <- id_backup
```

```{r heatmap top markers, fig.width=11, fig.height=11}
heatmap_cols <- colorRampPalette(RColorBrewer::brewer.pal(9, "RdBu"))(256)
heatmap_cols <- rev(heatmap_cols[1:256])

set.seed(123)
cells_use_heatmap <- colnames(EC_combined)
if (length(cells_use_heatmap) > 20000) {
  cells_use_heatmap <- sample(cells_use_heatmap, 20000)
}

DoHeatmap(
  subset(EC_combined, cells = cells_use_heatmap),
  label = TRUE,
  draw.lines = FALSE,
  features = unique(EC_markers_top10_combined$gene),
  assay = "SCT",
  size = 3.5
) +
  scale_fill_gradientn(colours = heatmap_cols) +
  theme(
    text = element_text(size = 12),
    axis.text.y = element_text(size = 8),
    plot.margin = unit(c(0, 0, 0, 0), "inches")
  ) +
  guides(fill = guide_colorbar(barwidth = 1, barheight = 4)) +
  guides(color = "none")
```

```{r violin selected markers, fig.width=15, fig.height=24}
selected_marker_features <- unique(c(
  markers_capillary,
  markers_venous,
  markers_arterial,
  markers_angiogenic,
  markers_inflammatory,
  markers_proliferating
))
selected_marker_features <- selected_marker_features[selected_marker_features %in% rownames(EC_combined)]

VlnPlot_scCustom(
  EC_combined,
  features = selected_marker_features,
  num_columns = 3,
  colors_use = cell_type_cols,
  pt.size = 0.001
)
```

```{r featureplot selected markers, fig.width=12, fig.height=20}
FeaturePlot_scCustom(
  EC_combined,
  features = unique(EC_markers_top3_combined$gene),
  num_columns = 3,
  order = TRUE
)
```

```{r cluster proportions metadata}
cluster_var <- "cell_state"

df_prop_ec <- EC_combined@meta.data %>%
  dplyr::mutate(cluster = .data[[cluster_var]]) %>%
  dplyr::count(Dataset, Group, Sample, Sex, Age, cluster, name = "n") %>%
  dplyr::group_by(Dataset, Group, Sample) %>%
  dplyr::mutate(total_n = sum(n),
                freq = n / total_n) %>%
  dplyr::ungroup()

df_prop_ec$cluster <- factor(
  df_prop_ec$cluster,
  levels = sort(unique(as.character(df_prop_ec$cluster)))
)

df_prop_ec$Group <- factor(df_prop_ec$Group, levels = c("H", "HF"))
df_prop_ec$Dataset <- factor(df_prop_ec$Dataset, levels = c("Koenig", "Reichart"))
df_prop_ec$Sex <- factor(df_prop_ec$Sex, levels = c("F", "M"))
df_prop_ec$Age <- as.numeric(df_prop_ec$Age)
```

```{r cluster proportions speckle h vs hf}
meta_abund <- EC_combined@meta.data %>%
  dplyr::mutate(cluster = .data[[cluster_var]]) %>%
  dplyr::group_by(Sample) %>%
  dplyr::summarise(
    Dataset = dplyr::first(Dataset),
    Group   = dplyr::first(Group),
    Sex     = dplyr::first(Sex),
    Age     = dplyr::first(Age),
    .groups = "drop"
  ) %>%
  dplyr::filter(!is.na(Group), !is.na(Age), !is.na(Sex), !is.na(Dataset))

meta_abund$Group <- factor(meta_abund$Group, levels = c("H", "HF"))
meta_abund$Sex <- factor(meta_abund$Sex, levels = c("F", "M"))
meta_abund$Dataset <- factor(meta_abund$Dataset, levels = c("Koenig", "Reichart"))
meta_abund$Age <- as.numeric(meta_abund$Age)

props_ec <- speckle::getTransformedProps(
  clusters = EC_combined@meta.data[[cluster_var]],
  sample = EC_combined$Sample,
  transform = "logit"
)

sample_order <- colnames(props_ec$TransformedProps)
meta_abund <- meta_abund %>%
  dplyr::filter(Sample %in% sample_order) %>%
  dplyr::arrange(match(Sample, sample_order))

props_ec$Counts <- props_ec$Counts[, meta_abund$Sample, drop = FALSE]
props_ec$Proportions <- props_ec$Proportions[, meta_abund$Sample, drop = FALSE]
props_ec$TransformedProps <- props_ec$TransformedProps[, meta_abund$Sample, drop = FALSE]

design_abund <- model.matrix(~ Group + Age + Sex + Dataset, data = meta_abund)

fit_abund <- limma::lmFit(props_ec$TransformedProps, design_abund)
fit_abund <- limma::eBayes(fit_abund, robust = TRUE)

abund_h_vs_hf <- limma::topTable(
  fit_abund,
  coef = "GroupHF",
  number = Inf,
  sort.by = "none"
) %>%
  tibble::rownames_to_column("cluster") %>%
  dplyr::mutate(
    cluster = factor(cluster, levels = levels(df_prop_ec$cluster)),
    fdr_label = vapply(adj.P.Val, format_p_label, character(1)),
    fdr_fontface = vapply(adj.P.Val, format_p_fontface, character(1))
  )

write.csv(
  abund_h_vs_hf,
  file = file.path(output_dir_data, "EC_subclusters_differential_abundance_H_vs_HF.csv"),
  row.names = FALSE,
  quote = FALSE
)

abund_h_vs_hf

```

```{r cluster proportions by dataset and group_2, fig.width=9, fig.height=4.5}
pd <- position_dodge(width = 0.75)
pjd <- position_jitterdodge(jitter.width = 0.12, dodge.width = 0.75, seed = 123)

p_prop_dataset <- ggplot(
  df_prop_ec,
  aes(x = cluster, y = freq, fill = Dataset)
) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.8,
    width = 0.7,
    position = pd
  ) +
  geom_point(
    position = pjd,
    size = 0.8,
    alpha = 0.65,
    shape = 16
  ) +
  scale_fill_manual(values = dataset_cols) +
  labs(x = "Cluster", y = "Fraction of endothelial cells") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.title = element_blank()
  )

y_annot <- df_prop_ec %>%
  dplyr::group_by(cluster) %>%
  dplyr::summarise(y = max(freq, na.rm = TRUE) + 0.03, .groups = "drop") %>%
  dplyr::left_join(
    abund_h_vs_hf %>% dplyr::select(cluster, adj.P.Val, fdr_label, fdr_fontface),
    by = "cluster"
  )

p_prop_group <- ggplot(
  df_prop_ec,
  aes(x = cluster, y = freq, fill = Group)
) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.8,
    width = 0.7,
    position = pd
  ) +
  geom_point(
    position = pjd,
    size = 0.8,
    alpha = 0.65,
    shape = 16
  ) +
  geom_segment(
    data = y_annot,
    aes(
      x = as.numeric(cluster) - 0.18,
      xend = as.numeric(cluster) + 0.18,
      y = y,
      yend = y
    ),
    inherit.aes = FALSE,
    linewidth = 0.4
  ) +
  geom_text(
    data = y_annot,
    aes(
      x = cluster,
      y = y + 0.025,
      label = fdr_label,
      fontface = fdr_fontface
    ),
    inherit.aes = FALSE,
    size = 3.2
  ) +
  scale_fill_manual(values = group_cols) +
  labs(x = "Cluster", y = "Fraction of endothelial cells") +
  coord_cartesian(
    ylim = c(0, max(y_annot$y + 0.04, na.rm = TRUE))
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.title = element_blank()
  )

p_prop_dataset | p_prop_group
```

### Over-representation analysis from cluster markers

```{r prepare marker gene lists}
df_markers_for_enrich <- EC_markers_combined[, c("cluster", "gene")]
gene_lists_by_cluster <- split(df_markers_for_enrich$gene, df_markers_for_enrich$cluster)

# Background: genes detected in at least 3 cells
DefaultAssay(EC_combined) <- "RNA"
gene_counts <- rowSums(EC_combined@assays$RNA@counts > 0)
universe_genes <- names(gene_counts[gene_counts >= 3])

universe_entrez <- clusterProfiler::bitr(
  universe_genes,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
) %>%
  dplyr::pull(ENTREZID) %>%
  unique()

gene_lists_entrez <- lapply(gene_lists_by_cluster, function(x) {
  clusterProfiler::bitr(
    x,
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db
  ) %>%
    dplyr::pull(ENTREZID) %>%
    unique()
})
```

```{r GO pathways dotplot, fig.width=9, fig.height=8}
GO_clusters <- compareCluster(
  geneCluster = gene_lists_entrez,
  fun = "enrichGO",
  ont = "BP",
  OrgDb = "org.Hs.eg.db",
  universe = universe_entrez
)
GO_clusters <- gsfilter(GO_clusters, by = "Count", min = 3)

dotplot(
  GO_clusters,
  label_format = 50,
  title = "GO biological process over-representation analysis"
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.text.y = element_text(lineheight = 0.5)
  )
```

```{r KEGG pathways dotplot, fig.width=9, fig.height=8}
KEGG_clusters <- compareCluster(
  geneCluster = gene_lists_entrez,
  fun = "enrichKEGG",
  organism = "hsa",
  universe = universe_entrez
)
KEGG_clusters <- gsfilter(KEGG_clusters, by = "Count", min = 3)

dotplot(
  KEGG_clusters,
  label_format = 45,
  title = "KEGG over-representation analysis"
) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

### Transcription factor activity inference

```{r tf network}
net_tf <- decoupleR::get_collectri(organism = "human", split_complexes = FALSE)
```

```{r tf network filter by expressed genes}
expr_matrix <- GetAssayData(object = EC_combined, assay = "RNA", slot = "counts")
non_zero_cells <- rowSums(expr_matrix > 0)
total_cells <- ncol(expr_matrix)
percent_cells <- (non_zero_cells / total_cells) * 100
percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

net_tf <- net_tf %>%
  dplyr::filter(source %in% dplyr::filter(percent_stats, percentCells > 1)$gene)
```

```{r tf colors}
TF_cols <- rev(colorRampPalette(RColorBrewer::brewer.pal(5, "BrBG"))(101))
```

```{r TF_tfsulm, eval=FALSE}
mat <- as.matrix(EC_combined@assays$RNA@data)

acts_tf <- decoupleR::run_ulm(
  mat = mat,
  net = net_tf,
  .source = "source",
  .target = "target",
  .mor = "mor",
  minsize = 10
)

EC_combined[["tfsulm"]] <- acts_tf %>%
  tidyr::pivot_wider(
    id_cols = "source",
    names_from = "condition",
    values_from = "score"
  ) %>%
  tibble::column_to_rownames("source") %>%
  Seurat::CreateAssayObject(.)

DefaultAssay(EC_combined) <- "tfsulm"
EC_combined <- ScaleData(EC_combined)
EC_combined@assays$tfsulm@data <- EC_combined@assays$tfsulm@scale.data
```

```{r tf heatmap top activities, fig.width=7, fig.height=4}
DefaultAssay(EC_combined) <- "tfsulm"

n_tfs <- 20

df_tf <- t(as.matrix(EC_combined@assays$tfsulm@data)) %>%
  as.data.frame() %>%
  dplyr::mutate(cluster = Seurat::Idents(EC_combined)) %>%
  tidyr::pivot_longer(cols = -cluster, names_to = "source", values_to = "score") %>%
  dplyr::group_by(cluster, source) %>%
  dplyr::summarise(mean = mean(score), .groups = "drop")

tfs_use <- df_tf %>%
  dplyr::group_by(source) %>%
  dplyr::summarise(sd_mean = stats::sd(mean), .groups = "drop") %>%
  dplyr::arrange(desc(abs(sd_mean))) %>%
  dplyr::slice_head(n = n_tfs) %>%
  dplyr::pull(source)

top_tf_mat <- df_tf %>%
  dplyr::filter(source %in% tfs_use) %>%
  tidyr::pivot_wider(id_cols = cluster, names_from = source, values_from = mean) %>%
  tibble::column_to_rownames("cluster") %>%
  as.matrix()

my_breaks_tf <- c(
  seq(-2, 0, length.out = ceiling(101 / 2) + 1),
  seq(0.05, 2, length.out = floor(101 / 2))
)

pheatmap::pheatmap(
  mat = top_tf_mat,
  color = TF_cols,
  border_color = "white",
  breaks = my_breaks_tf,
  cellwidth = 15,
  cellheight = 15,
  treeheight_row = 20,
  treeheight_col = 20
)
```

```{r tf featureplots representative, fig.width=12, fig.height=4}
rep_tfs <- intersect(c("RORA", "CIITA", "JUN", "SP1", "STAT3"), rownames(EC_combined@assays$tfsulm))
DefaultAssay(EC_combined) <- "RNA"
rep_tf_expr <- rep_tfs[rep_tfs %in% rownames(EC_combined)]

for (tf in rep_tfs) {
  DefaultAssay(EC_combined) <- "tfsulm"
  p1 <- DimPlot(
    EC_combined,
    reduction = "umap",
    label = TRUE,
    pt.size = 0.1
  ) + NoLegend() + ggtitle("Cell states")

  p2 <- FeaturePlot(
    EC_combined,
    features = tf,
    pt.size = 0.1,
    order = TRUE
  ) +
    scale_colour_gradient2(low = TF_cols[1], high = TF_cols[101], mid = "white") +
    ggtitle(paste(tf, "activity"))

  if (tf %in% rep_tf_expr) {
    DefaultAssay(EC_combined) <- "RNA"
    p3 <- FeaturePlot_scCustom(
      EC_combined,
      features = tf,
      pt.size = 0.1
    ) + ggtitle(paste(tf, "expression"))

    print(p1 | p2 | p3)
  } else {
    print(p1 | p2)
  }
}
```

### Pathway activity inference

```{r pathway network}
net_path <- decoupleR::get_progeny(organism = "human", top = 100000)
```

```{r pathway colors}
Pathway_cols <- colorRampPalette(c("#28A2EB", "white", "#FEC233"))(101)
```

```{r Pathway_pathwaysmlm, eval=FALSE}
DefaultAssay(EC_combined) <- "RNA"
mat <- as.matrix(EC_combined@assays$RNA@data)

acts_pathway <- decoupleR::run_mlm(
  mat = mat,
  net = net_path,
  .source = "source",
  .target = "target",
  .mor = "weight",
  minsize = 10
)

EC_combined[["pathwaysmlm"]] <- acts_pathway %>%
  tidyr::pivot_wider(
    id_cols = "source",
    names_from = "condition",
    values_from = "score"
  ) %>%
  tibble::column_to_rownames("source") %>%
  Seurat::CreateAssayObject(.)

DefaultAssay(EC_combined) <- "pathwaysmlm"
EC_combined <- ScaleData(EC_combined)
EC_combined@assays$pathwaysmlm@data <- EC_combined@assays$pathwaysmlm@scale.data
```

```{r pathway heatmap, fig.width=7, fig.height=4}
DefaultAssay(EC_combined) <- "pathwaysmlm"

df_pathway <- t(as.matrix(EC_combined@assays$pathwaysmlm@data)) %>%
  as.data.frame() %>%
  dplyr::mutate(cluster = Seurat::Idents(EC_combined)) %>%
  tidyr::pivot_longer(cols = -cluster, names_to = "source", values_to = "score") %>%
  dplyr::group_by(cluster, source) %>%
  dplyr::summarise(mean = mean(score), .groups = "drop")

pathway_mat <- df_pathway %>%
  tidyr::pivot_wider(id_cols = cluster, names_from = source, values_from = mean) %>%
  tibble::column_to_rownames("cluster") %>%
  as.matrix()

my_breaks_path <- c(
  seq(-1.25, 0, length.out = ceiling(101 / 2) + 1),
  seq(0.05, 1.25, length.out = floor(101 / 2))
)

pheatmap::pheatmap(
  mat = pathway_mat,
  color = Pathway_cols,
  border_color = "white",
  breaks = my_breaks_path,
  cellwidth = 20,
  cellheight = 20,
  treeheight_row = 20,
  treeheight_col = 20
)
```

```{r pathway featureplots representative, fig.width=12, fig.height=4}
DefaultAssay(EC_combined) <- "pathwaysmlm"
rep_pathways <- intersect(c("TGFb", "TNFa", "NFkB", "VEGF", "MAPK"), rownames(EC_combined@assays$pathwaysmlm))

for (pw in rep_pathways) {
  p1 <- DimPlot(
    EC_combined,
    reduction = "umap",
    label = TRUE,
    pt.size = 0.1
  ) + NoLegend() + ggtitle("Cell states")

  p2 <- FeaturePlot(
    EC_combined,
    features = pw,
    pt.size = 0.1,
    order = TRUE
  ) +
    scale_colour_gradientn(colors = Pathway_cols) +
    ggtitle(paste(pw, "activity"))

  print(p1 | p2)
}
```

```{r save EC_combined final, eval=FALSE}
DefaultAssay(EC_combined) <- "RNA"
qs_save(
  EC_combined,
  file = here::here(output_dir_data, "EC_combined.qs2"),
  nthreads = 32
)
```

## Endothelial-cell subset programs

```{r subset gene sets}
angiogenesis_genes_subset_score <- c("ESM1", "ANGPT2", "KDR", "APLN", "CXCR4", "PGF", "DLL4", "KIT")
angiogenesis_genes_subset_plot  <- angiogenesis_genes_subset_score

apoptotic_priming_genes_subset <- c(
  "BAX", "BAK1", "BBC3", "BCL2L11", "PMAIP1", "BID",
  "CASP3", "CASP9", "APAF1", "DIABLO", "CYCS"
)

redox_no_genes_subset <- redox_no_genes
mito_homeostasis_genes_subset <- mito_homeostasis_genes

aerobic_respiration_genes_subset <- aerobic_respiration_genes

phase_gene_sets_subset <- list(
  S_phase_score = s_phase_genes_present,
  G2M_phase_score = g2m_phase_genes_present
)

subset_score_gene_sets <- list(
  Angiogenesis_score = angiogenesis_genes_subset_score,
  Apoptotic_priming_score = apoptotic_priming_genes_subset,
  Aerobic_respiration_score = aerobic_respiration_genes_subset
)

subset_gene_plot_list <- list(
  angiogenesis = angiogenesis_genes_subset_plot,
  apoptosis = apoptotic_priming_genes_subset,
  redox_no = redox_no_genes_subset,
  mito_homeostasis = mito_homeostasis_genes_subset
)
```

```{r subset metadata and helper functions}
DefaultAssay(EC_combined) <- "RNA"
Idents(EC_combined) <- EC_combined$cell_state

EC_combined$cell_state <- factor(
  EC_combined$cell_state,
  levels = levels(EC_combined$cell_state)
)
EC_combined$Group <- factor(EC_combined$Group, levels = c("H", "HF"))
EC_combined$Dataset <- factor(EC_combined$Dataset, levels = c("Koenig", "Reichart"))
EC_combined$Sex <- factor(EC_combined$Sex, levels = c("F", "M"))
EC_combined$Age <- as.numeric(EC_combined$Age)

subset_levels <- levels(droplevels(EC_combined$cell_state))
subset_cols <- setNames(cell_type_cols[seq_along(subset_levels)], subset_levels)

sample_subset_meta <- EC_combined@meta.data %>%
  dplyr::select(Sample, cell_state, Group, Sex, Age, Dataset) %>%
  dplyr::distinct() %>%
  dplyr::filter(!is.na(Sample), !is.na(cell_state), !is.na(Group), !is.na(Sex), !is.na(Age), !is.na(Dataset))

add_module_score_clean <- function(object, features, score_name, assay = NULL, nbin = 24, ctrl = 100, seed = 1) {
  if (is.null(assay)) {
    assay <- Seurat::DefaultAssay(object)
  }

  features_present <- intersect(features, rownames(object[[assay]]))
  if (length(features_present) == 0) {
    warning(paste0("No genes present for ", score_name, ". Filling with NA."))
    object[[score_name]] <- NA_real_
    return(object)
  }

  object <- Seurat::AddModuleScore(
    object = object,
    features = list(features_present),
    assay = assay,
    name = paste0(score_name, "_tmp"),
    nbin = nbin,
    ctrl = ctrl,
    seed = seed
  )

  tmp_col <- paste0(score_name, "_tmp1")
  object[[score_name]] <- object[[tmp_col]]
  object[[tmp_col]] <- NULL

  object
}

set_plot_layer_alpha <- function(p, alpha_value = 0.7) {
  for (i in seq_along(p$layers)) {
    geom_name <- class(p$layers[[i]]$geom)[1]
    if (geom_name %in% c("GeomPoint", "GeomJitter", "GeomScattermore")) {
      p$layers[[i]]$aes_params$alpha <- alpha_value
      if (!is.null(p$layers[[i]]$geom_params)) {
        p$layers[[i]]$geom_params$alpha <- alpha_value
      }
    }
  }
  p
}

plot_subset_overall_sc <- function(
  seurat_object,
  feature,
  title_text,
  ylab_text,
  point_size = 0.15,
  point_alpha = 0.7
) {
  p <- scCustomize::VlnPlot_scCustom(
    seurat_object = seurat_object,
    features = feature,
    group.by = "cell_state",
    colors_use = subset_cols,
    pt.size = point_size,
    raster = FALSE,
    add.noise = TRUE
  )

  p <- set_plot_layer_alpha(p, alpha_value = point_alpha)

  p +
    labs(title = title_text, x = NULL, y = ylab_text) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.title.y = element_text(face = "bold"),
      legend.position = "none"
    )
}


plot_hvhf_by_subset <- function(df, value_col, title_text, ylab_text, facet_nrow = 2) {
  df <- df %>%
    dplyr::filter(!is.na(.data[[value_col]])) %>%
    dplyr::mutate(
      cell_state = factor(cell_state, levels = subset_levels),
      Group = factor(Group, levels = c("H", "HF"))
    )

  p_df <- lapply(subset_levels, function(cs) {
    df_sub <- df %>% dplyr::filter(cell_state == cs)
    if (nrow(df_sub) == 0 || dplyr::n_distinct(df_sub$Group) < 2) {
      return(data.frame(cell_state = cs, p_value = NA_real_))
    }

    fit <- tryCatch(
      lm(stats::as.formula(paste0(value_col, " ~ Group + Age + Sex + Dataset")), data = df_sub),
      error = function(e) NULL
    )

    p_val <- if (!is.null(fit)) {
      broom::tidy(fit) %>%
        dplyr::filter(term == "GroupHF") %>%
        dplyr::pull(p.value)
    } else {
      NA_real_
    }

    data.frame(cell_state = cs, p_value = ifelse(length(p_val) == 0, NA_real_, p_val))
  }) %>% dplyr::bind_rows()

  anno_df <- df %>%
    dplyr::group_by(cell_state) %>%
    dplyr::summarise(
      y_max = max(.data[[value_col]], na.rm = TRUE),
      y_min = min(.data[[value_col]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      y_rng = y_max - y_min,
      y_rng = ifelse(!is.finite(y_rng) | y_rng == 0, 1, y_rng),
      y_line = y_max + 0.10 * y_rng,
      y_text = y_max + 0.16 * y_rng
    ) %>%
    dplyr::left_join(p_df, by = "cell_state") %>%
    dplyr::mutate(
      p_label = vapply(p_value, format_p_label, character(1)),
      p_fontface = vapply(p_value, format_p_fontface, character(1))
    )

  ggplot(df, aes(x = Group, y = .data[[value_col]], fill = Group)) +
    geom_boxplot(width = 0.55, outlier.shape = NA, color = "black", linewidth = 0.45) +
    geom_jitter(width = 0.10, size = 1.3, shape = 21, stroke = 0.35, color = "black") +
    geom_segment(
      data = anno_df,
      aes(x = 1, xend = 2, y = y_line, yend = y_line),
      inherit.aes = FALSE
    ) +
    geom_text(
      data = anno_df,
      aes(x = 1.5, y = y_text, label = p_label, fontface = p_fontface),
      inherit.aes = FALSE,
      size = 4.2
    ) +
    scale_fill_manual(values = group_cols) +
    facet_wrap(~cell_state, scales = "free_y", nrow = facet_nrow) +
    labs(title = title_text, x = NULL, y = ylab_text) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.title.y = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      legend.position = "none"
    )
}
```

## Single-cell scores and expression per subset

```{r subset single-cell scores and expression prepare}
EC_combined_scores <- EC_combined

for (score_name in names(phase_gene_sets_subset)) {
  EC_combined_scores <- add_module_score_clean(
    object = EC_combined_scores,
    features = phase_gene_sets_subset[[score_name]],
    score_name = score_name,
    assay = "RNA"
  )
}

for (score_name in names(subset_score_gene_sets)) {
  EC_combined_scores <- add_module_score_clean(
    object = EC_combined_scores,
    features = subset_score_gene_sets[[score_name]],
    score_name = score_name,
    assay = "RNA"
  )
}

df_score_sc <- EC_combined_scores@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  dplyr::transmute(
    cell_id,
    Sample,
    Dataset,
    Group,
    Sex,
    Age,
    cell_state,
    S_phase_score,
    G2M_phase_score,
    Angiogenesis_score,
    Apoptotic_priming_score,
    Aerobic_respiration_score
  ) %>%
  dplyr::mutate(
    cell_state = factor(cell_state, levels = subset_levels),
    Group = factor(Group, levels = c("H", "HF")),
    Dataset = factor(Dataset, levels = c("Koenig", "Reichart")),
    Sex = factor(Sex, levels = c("F", "M")),
    Age = as.numeric(Age)
  )

genes_for_single_cell_plots <- unique(c(
  angiogenesis_genes_subset_plot,
  apoptotic_priming_genes_subset,
  redox_no_genes_subset,
  mito_homeostasis_genes_subset
))
genes_for_single_cell_plots <- genes_for_single_cell_plots[genes_for_single_cell_plots %in% rownames(EC_combined_scores)]

expr_sc <- Seurat::FetchData(
  object = EC_combined_scores,
  vars = c("cell_state", "Group", "Dataset", "Sex", "Age", "Sample", genes_for_single_cell_plots),
  slot = "data"
) %>%
  tibble::rownames_to_column("cell_id")

subset_gene_df_sc <- expr_sc %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(genes_for_single_cell_plots),
    names_to = "gene",
    values_to = "expr"
  ) %>%
  dplyr::mutate(
    program = dplyr::case_when(
      gene %in% angiogenesis_genes_subset_plot ~ "angiogenesis",
      gene %in% apoptotic_priming_genes_subset ~ "apoptosis",
      TRUE ~ NA_character_
    ),
    cell_state = factor(cell_state, levels = subset_levels),
    Group = factor(Group, levels = c("H", "HF")),
    Dataset = factor(Dataset, levels = c("Koenig", "Reichart")),
    Sex = factor(Sex, levels = c("F", "M")),
    Age = as.numeric(Age)
  )

write.csv(
  df_score_sc,
  file = file.path(output_dir_data, "EC_subsets_single_cell_scores.csv"),
  row.names = FALSE,
  quote = FALSE
)

write.csv(
  subset_gene_df_sc,
  file = file.path(output_dir_data, "EC_subsets_single_cell_gene_expression.csv"),
  row.names = FALSE,
  quote = FALSE
)
```

## Subset pseudobulk for H vs HF comparisons

```{r subset pseudobulk prepare}
# Lower library-size threshold than the whole-endothelium analysis, because each sample is now split by subset.
min_subset_libsize <- 10000

pb_subsets <- Seurat2PB(
  EC_combined,
  sample = "Sample",
  cluster = "cell_state"
)

pb_subset_counts <- tryCatch(pb_subsets$counts, error = function(e) NULL)
if (is.null(pb_subset_counts)) {
  pb_subset_counts <- tryCatch(pb_subsets[, ]$counts, error = function(e) NULL)
}
if (is.null(pb_subset_counts)) {
  stop("Could not extract pseudobulk counts from pb_subsets.")
}

pb_subset_meta <- tryCatch(as.data.frame(pb_subsets$samples), error = function(e) NULL)
if (is.null(pb_subset_meta)) {
  pb_subset_meta <- tryCatch(as.data.frame(SummarizedExperiment::colData(pb_subsets)), error = function(e) NULL)
}
if (is.null(pb_subset_meta)) {
  stop("Could not extract pseudobulk sample metadata from pb_subsets.")
}

pb_subset_meta <- pb_subset_meta %>%
  dplyr::transmute(
    Sample = sample,
    cell_state = cluster
  ) %>%
  dplyr::left_join(sample_subset_meta, by = c("Sample", "cell_state"))

colnames(pb_subset_counts) <- paste(pb_subset_meta$Sample, pb_subset_meta$cell_state, sep = "__")
pb_subset_meta$pb_id <- colnames(pb_subset_counts)
rownames(pb_subset_meta) <- pb_subset_meta$pb_id

subset_libsizes <- colSums(pb_subset_counts)
pb_subset_meta$libsize <- subset_libsizes[pb_subset_meta$pb_id]

pb_subset_meta_keep <- pb_subset_meta %>%
  dplyr::filter(!is.na(Group), !is.na(Sex), !is.na(Age), !is.na(Dataset), libsize >= min_subset_libsize)

pb_subset_counts_keep <- pb_subset_counts[, pb_subset_meta_keep$pb_id, drop = FALSE]

stopifnot(all(colnames(pb_subset_counts_keep) == pb_subset_meta_keep$pb_id))

pb_subset_meta_keep$cell_state <- factor(pb_subset_meta_keep$cell_state, levels = subset_levels)
pb_subset_meta_keep$Group <- factor(pb_subset_meta_keep$Group, levels = c("H", "HF"))
pb_subset_meta_keep$Dataset <- factor(pb_subset_meta_keep$Dataset, levels = c("Koenig", "Reichart"))
pb_subset_meta_keep$Sex <- factor(pb_subset_meta_keep$Sex, levels = c("F", "M"))
pb_subset_meta_keep$Age <- as.numeric(pb_subset_meta_keep$Age)

write.csv(
  pb_subset_meta,
  file = file.path(output_dir_data, "EC_subsets_pseudobulk_metadata_all.csv"),
  row.names = FALSE,
  quote = FALSE
)

write.csv(
  pb_subset_meta_keep,
  file = file.path(output_dir_data, "EC_subsets_pseudobulk_metadata_kept.csv"),
  row.names = FALSE,
  quote = FALSE
)

write.csv(
  pb_subset_meta %>% dplyr::filter(!(pb_id %in% pb_subset_meta_keep$pb_id)),
  file = file.path(output_dir_data, "EC_subsets_pseudobulk_metadata_filtered_out.csv"),
  row.names = FALSE,
  quote = FALSE
)

dge_subset <- edgeR::DGEList(counts = pb_subset_counts_keep)
dge_subset <- edgeR::calcNormFactors(dge_subset)
logcpm_subset <- edgeR::cpm(dge_subset, log = TRUE, prior.count = 1)
```

```{r subset scores and gene matrices}
subset_score_genes_all <- unique(unlist(subset_score_gene_sets, use.names = FALSE))
score_genes_present <- subset_score_genes_all[subset_score_genes_all %in% rownames(logcpm_subset)]
score_genes_missing <- setdiff(subset_score_genes_all, rownames(logcpm_subset))

logcpm_subset_scores <- logcpm_subset[
  rownames(logcpm_subset) %in% score_genes_present,
  ,
  drop = FALSE
]

logcpm_subset_scores_z <- t(scale(t(logcpm_subset_scores)))
logcpm_subset_scores_z[is.na(logcpm_subset_scores_z)] <- 0

safe_colmean_score <- function(gene_vec, score_mat, ids) {
  genes_here <- intersect(gene_vec, rownames(score_mat))
  if (length(genes_here) == 0) {
    return(rep(NA_real_, length(ids)))
  }
  out <- colMeans(score_mat[genes_here, , drop = FALSE])
  as.numeric(out[ids])
}

subset_score_df <- pb_subset_meta_keep %>%
  dplyr::mutate(
    Angiogenesis_score = safe_colmean_score(angiogenesis_genes_subset_score, logcpm_subset_scores_z, pb_id),
    Apoptotic_priming_score = safe_colmean_score(apoptotic_priming_genes_subset, logcpm_subset_scores_z, pb_id),
    Aerobic_respiration_score = safe_colmean_score(aerobic_respiration_genes_subset, logcpm_subset_scores_z, pb_id)
  )

phase_score_df <- df_score_sc %>%
  dplyr::group_by(Sample, Dataset, Group, Sex, Age, cell_state) %>%
  dplyr::summarise(
    S_phase_score = mean(S_phase_score, na.rm = TRUE),
    G2M_phase_score = mean(G2M_phase_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    cell_state = factor(cell_state, levels = subset_levels),
    Group = factor(Group, levels = c("H", "HF")),
    Dataset = factor(Dataset, levels = c("Koenig", "Reichart")),
    Sex = factor(Sex, levels = c("F", "M")),
    Age = as.numeric(Age)
  )

subset_gene_df <- purrr::map_dfr(names(subset_gene_plot_list), function(program_name) {
  genes_here <- subset_gene_plot_list[[program_name]]
  genes_here <- genes_here[genes_here %in% rownames(logcpm_subset)]
  if (length(genes_here) == 0) return(NULL)

  purrr::map_dfr(genes_here, function(g) {
    pb_subset_meta_keep %>%
      dplyr::transmute(
        pb_id,
        Sample,
        Dataset,
        Group,
        Sex,
        Age,
        cell_state,
        program = program_name,
        gene = g,
        logCPM = as.numeric(logcpm_subset[g, pb_id])
      )
  })
})

write.csv(
  phase_score_df,
  file = file.path(output_dir_data, "EC_subsets_phase_scores_per_sample.csv"),
  row.names = FALSE,
  quote = FALSE
)

write.csv(
  subset_score_df,
  file = file.path(output_dir_data, "EC_subsets_gene_set_scores.csv"),
  row.names = FALSE,
  quote = FALSE
)

write.csv(
  subset_gene_df,
  file = file.path(output_dir_data, "EC_subsets_gene_logCPM.csv"),
  row.names = FALSE,
  quote = FALSE
)

write.csv(
  data.frame(gene_missing = score_genes_missing),
  file = file.path(output_dir_data, "EC_subsets_score_genes_missing.csv"),
  row.names = FALSE,
  quote = FALSE
)
```

## Phase scores per subset

```{r s score subset differences, fig.width=8.5, fig.height=3.8}
print(
  plot_subset_overall_sc(
    seurat_object = EC_combined_scores,
    feature = "S_phase_score",
    title_text = "S-phase score across subsets",
    ylab_text = "Score",
    point_size = 0.15,
    point_alpha = 0.0
  )
)
```

```{r s score h vs hf within subset, fig.width=14, fig.height=6.5}
print(
  plot_hvhf_by_subset(
    df = phase_score_df,
    value_col = "S_phase_score",
    title_text = "S-phase score by group within each subset",
    ylab_text = "Score",
    facet_nrow = 2
  )
)
```

```{r g2m score subset differences, fig.width=8.5, fig.height=3.8}
print(
  plot_subset_overall_sc(
    seurat_object = EC_combined_scores,
    feature = "G2M_phase_score",
    title_text = "G2/M-phase score across subsets",
    ylab_text = "Score",
    point_size = 0.15,
    point_alpha = 0
  )
)
```

```{r g2m score h vs hf within subset, fig.width=14, fig.height=6.5}
print(
  plot_hvhf_by_subset(
    df = phase_score_df,
    value_col = "G2M_phase_score",
    title_text = "G2/M-phase score by group within each subset",
    ylab_text = "Score",
    facet_nrow = 2
  )
)
```

## Gene-set scores per subset

```{r aerobic respiration score subset differences, fig.width=8.5, fig.height=3.8}
print(
  plot_subset_overall_sc(
    seurat_object = EC_combined_scores,
    feature = "Aerobic_respiration_score",
    title_text = "Aerobic respiration score across subsets",
    ylab_text = "Score",
    point_size = 0.15,
    point_alpha = 0
  )
)
```

```{r aerobic respiration score h vs hf within subset, fig.width=14, fig.height=6.5}
print(
  plot_hvhf_by_subset(
    df = subset_score_df,
    value_col = "Aerobic_respiration_score",
    title_text = "Aerobic respiration score by group within each subset"
    ,
    ylab_text = "Score",
    facet_nrow = 2
  )
)
```

```{r angiogenesis score subset differences, fig.width=8.5, fig.height=3.8}
print(
  plot_subset_overall_sc(
    seurat_object = EC_combined_scores,
    feature = "Angiogenesis_score",
    title_text = "Angiogenesis score across subsets",
    ylab_text = "Score",
    point_size = 0.15,
    point_alpha = 0
  )
)
```

```{r angiogenesis score h vs hf within subset, fig.width=14, fig.height=6.5}
print(
  plot_hvhf_by_subset(
    df = subset_score_df,
    value_col = "Angiogenesis_score",
    title_text = "Angiogenesis score by group within each subset",
    ylab_text = "Score",
    facet_nrow = 2
  )
)
```

```{r apoptotic priming score subset differences, fig.width=8.5, fig.height=3.8}
print(
  plot_subset_overall_sc(
    seurat_object = EC_combined_scores,
    feature = "Apoptotic_priming_score",
    title_text = "Apoptotic priming score across subsets",
    ylab_text = "Score",
    point_size = 0.15,
    point_alpha = 0
  )
)
```

```{r apoptotic priming score h vs hf within subset, fig.width=14, fig.height=6.5}
print(
  plot_hvhf_by_subset(
    df = subset_score_df,
    value_col = "Apoptotic_priming_score",
    title_text = "Apoptotic priming score by group within each subset",
    ylab_text = "Score",
    facet_nrow = 2
  )
)
```

## Individual angiogenesis and apoptosis genes per subset

```{r angiogenesis genes subset differences, fig.width=8.5, fig.height=3.8}
for (gene_here in angiogenesis_genes_subset_plot[angiogenesis_genes_subset_plot %in% rownames(EC_combined_scores)]) {
  p <- plot_subset_overall_sc(
    seurat_object = EC_combined_scores,
    feature = gene_here,
    title_text = paste0(gene_here, " across subsets"),
    ylab_text = "Log-normalized expression",
    point_size = 0.15,
    point_alpha = 0.1
  )
  print(p)
}
```

```{r angiogenesis genes h vs hf within subset, fig.width=14, fig.height=6.5}
for (gene_here in angiogenesis_genes_subset_plot[angiogenesis_genes_subset_plot %in% unique(subset_gene_df$gene)]) {
  df_sub <- subset_gene_df %>% dplyr::filter(program == "angiogenesis", gene == gene_here)

  p <- plot_hvhf_by_subset(
    df = df_sub,
    value_col = "logCPM",
    title_text = paste0(gene_here, " by group within each subset"),
    ylab_text = "logCPM",
    facet_nrow = 2
  )
  print(p)
}
```

```{r apoptosis genes subset differences, fig.width=8.5, fig.height=3.8}
for (gene_here in apoptotic_priming_genes_subset[apoptotic_priming_genes_subset %in% rownames(EC_combined_scores)]) {
  p <- plot_subset_overall_sc(
    seurat_object = EC_combined_scores,
    feature = gene_here,
    title_text = paste0(gene_here, " across subsets"),
    ylab_text = "Log-normalized expression",
    point_size = 0.15,
    point_alpha = 0.2
  )
  print(p)
}
```

```{r apoptosis genes h vs hf within subset, fig.width=14, fig.height=6.5}
for (gene_here in apoptotic_priming_genes_subset[apoptotic_priming_genes_subset %in% unique(subset_gene_df$gene)]) {
  df_sub <- subset_gene_df %>% dplyr::filter(program == "apoptosis", gene == gene_here)

  p <- plot_hvhf_by_subset(
    df = df_sub,
    value_col = "logCPM",
    title_text = paste0(gene_here, " by group within each subset"),
    ylab_text = "logCPM",
    facet_nrow = 2
  )
  print(p)
}
```

## EC_activated subset: differential expression and GO GSEA

```{r ec_activated subset pseudobulk prepare}
DefaultAssay(EC_combined) <- "RNA"

EC_activated <- subset(
  EC_combined,
  subset = cell_state == "EC_activated"
)

EC_activated_meta <- EC_activated@meta.data %>%
  dplyr::select(Sample, Group, Sex, Age, Dataset) %>%
  dplyr::distinct() %>%
  dplyr::filter(!is.na(Sample), !is.na(Group), !is.na(Sex), !is.na(Age), !is.na(Dataset))

EC_activated_pb <- Seurat2PB(
  EC_activated,
  sample = "Sample",
  cluster = "cell_state"
)

colnames(EC_activated_pb) <- EC_activated_pb$samples$sample
EC_activated_pb <- EC_activated_pb[, match(EC_activated_meta$Sample, EC_activated_pb$samples$sample)]

keep_samples_EC_activated <- EC_activated_pb$samples$lib.size > 2e4
EC_activated_pb <- EC_activated_pb[, keep_samples_EC_activated]

EC_activated_meta <- EC_activated_meta %>%
  dplyr::filter(Sample %in% colnames(EC_activated_pb)) %>%
  dplyr::slice(match(colnames(EC_activated_pb), Sample))

stopifnot(all(EC_activated_meta$Sample == colnames(EC_activated_pb)))

EC_activated_meta$Group <- factor(EC_activated_meta$Group, levels = c("H", "HF"))
EC_activated_meta$Sex <- factor(EC_activated_meta$Sex, levels = c("F", "M"))
EC_activated_meta$Dataset <- factor(EC_activated_meta$Dataset, levels = c("Koenig", "Reichart"))
EC_activated_meta$Age <- as.numeric(EC_activated_meta$Age)

pseudobulk_EC_activated <- edgeR::DGEList(counts = EC_activated_pb$counts)
pseudobulk_EC_activated$samples <- cbind(pseudobulk_EC_activated$samples, EC_activated_meta)

design_EC_activated <- model.matrix(~ Group + Age + Sex + Dataset, data = EC_activated_meta)
colnames(design_EC_activated)

keep_genes_EC_activated <- edgeR::filterByExpr(
  pseudobulk_EC_activated,
  design = design_EC_activated
)

pseudobulk_EC_activated <- pseudobulk_EC_activated[
  keep_genes_EC_activated,
  ,
  keep.lib.sizes = FALSE
]
pseudobulk_EC_activated <- edgeR::normLibSizes(pseudobulk_EC_activated)
pseudobulk_EC_activated <- edgeR::estimateDisp(
  pseudobulk_EC_activated,
  design_EC_activated
)

logcpm_EC_activated <- edgeR::cpm(
  pseudobulk_EC_activated,
  log = TRUE,
  prior.count = 1
)
```

```{r ec_activated differential expression}
fit_EC_activated <- edgeR::glmQLFit(
  pseudobulk_EC_activated,
  design_EC_activated,
  robust = TRUE
)

qlf_EC_activated <- edgeR::glmQLFTest(
  fit_EC_activated,
  coef = "GroupHF"
)

results_EC_activated <- edgeR::topTags(qlf_EC_activated, n = Inf)$table %>%
  tibble::rownames_to_column("gene") %>%
  dplyr::mutate(metric = logFC * -log10(PValue))

results_EC_activated$significance <- "Not significant"
results_EC_activated$significance[
  results_EC_activated$FDR < 0.05 & results_EC_activated$logFC > 0.5
] <- "Increased"
results_EC_activated$significance[
  results_EC_activated$FDR < 0.05 & results_EC_activated$logFC < -0.5
] <- "Decreased"
results_EC_activated$significance <- factor(
  results_EC_activated$significance,
  levels = c("Decreased", "Not significant", "Increased")
)

signif_EC_activated <- results_EC_activated %>%
  dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>%
  dplyr::arrange(desc(metric))

write.csv(
  results_EC_activated,
  file = file.path(output_dir_data, "EC_activated_DE_results_Koenig_Reichart_combined.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  signif_EC_activated,
  file = file.path(output_dir_data, "EC_activated_DE_signif_Koenig_Reichart_combined.csv"),
  quote = FALSE,
  row.names = FALSE
)
```

### EC_activated volcano plot

```{r ec_activated volcano, fig.width=4, fig.height=5}
n_h_EC_activated <- sum(EC_activated_meta$Group == "H")
n_hf_EC_activated <- sum(EC_activated_meta$Group == "HF")

EnhancedVolcano(
  results_EC_activated,
  lab = results_EC_activated$gene,
  x = "logFC",
  y = "FDR",
  labSize = 0,
  legendLabSize = 12,
  titleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 0.8,
  FCcutoff = 0.5,
  pCutoff = 0.05,
  ylim = c(0, 20),
  xlim = c(-3.75, 3.75),
  colAlpha = 1,
  drawConnectors = FALSE,
  ylab = bquote(-Log[10]~FDR),
  subtitle = NULL,
  title = paste0("EC_activated\nHF (n=", n_hf_EC_activated, ") vs H (n=", n_h_EC_activated, ")"),
  colCustom = setNames(
    c("#004C8C", "grey75", "#A63A2A")[match(
      results_EC_activated$significance,
      c("Decreased", "Not significant", "Increased")
    )],
    as.character(results_EC_activated$significance)
  )
) +
  theme(legend.position = "none")
```

### EC_activated heatmaps

```{r ec_activated order heatmap columns}
column_order_EC_activated <- EC_activated_meta %>%
  dplyr::arrange(Group, Dataset, Age, Sex) %>%
  dplyr::pull(Sample)

metadata_heatmap_EC_activated <- EC_activated_meta %>%
  dplyr::filter(Sample %in% column_order_EC_activated) %>%
  dplyr::slice(match(column_order_EC_activated, Sample))

ha_EC_activated <- ComplexHeatmap::HeatmapAnnotation(
  Group = metadata_heatmap_EC_activated$Group,
  Age = metadata_heatmap_EC_activated$Age,
  Sex = metadata_heatmap_EC_activated$Sex,
  Dataset = metadata_heatmap_EC_activated$Dataset,
  col = list(
    Group = group_cols,
    Age = circlize::colorRamp2(
      c(
        min(metadata_heatmap_EC_activated$Age, na.rm = TRUE),
        max(metadata_heatmap_EC_activated$Age, na.rm = TRUE)
      ),
      c("grey90", "black")
    ),
    Sex = sex_cols,
    Dataset = dataset_cols
  ),
  annotation_name_gp = grid::gpar(fontsize = 10, fontface = "bold")
)
```

```{r ec_activated heatmap all degs, fig.width=9, fig.height=12}
mat_all_deg_EC_activated <- logcpm_EC_activated[
  rownames(logcpm_EC_activated) %in% signif_EC_activated$gene,
  column_order_EC_activated,
  drop = FALSE
]

mat_all_deg_EC_activated <- mat_all_deg_EC_activated[
  match(signif_EC_activated$gene, rownames(mat_all_deg_EC_activated)),
  ,
  drop = FALSE
]

mat_all_deg_EC_activated_z <- t(scale(t(mat_all_deg_EC_activated)))
mat_all_deg_EC_activated_z[is.na(mat_all_deg_EC_activated_z)] <- 0

ComplexHeatmap::Heatmap(
  mat_all_deg_EC_activated_z,
  name = "Z-score",
  top_annotation = ha_EC_activated,
  col = heat_cols,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  show_column_names = TRUE,
  column_names_gp = grid::gpar(fontsize = 5),
  row_names_gp = grid::gpar(fontsize = 6),
  column_title = "EC_activated: all significant DEGs",
  column_title_gp = grid::gpar(fontsize = 12, fontface = "bold")
)
```

```{r ec_activated heatmap top 25 up down, fig.width=9, fig.height=8}
top_up_EC_activated <- results_EC_activated %>%
  dplyr::filter(FDR < 0.05, logFC > 0.5) %>%
  dplyr::arrange(desc(metric)) %>%
  dplyr::slice_head(n = 25)

top_down_EC_activated <- results_EC_activated %>%
  dplyr::filter(FDR < 0.05, logFC < -0.5) %>%
  dplyr::arrange(metric) %>%
  dplyr::slice_head(n = 25)

top_genes_EC_activated <- dplyr::bind_rows(top_up_EC_activated, top_down_EC_activated)

mat_top_deg_EC_activated <- logcpm_EC_activated[
  rownames(logcpm_EC_activated) %in% top_genes_EC_activated$gene,
  column_order_EC_activated,
  drop = FALSE
]

mat_top_deg_EC_activated <- mat_top_deg_EC_activated[
  match(top_genes_EC_activated$gene, rownames(mat_top_deg_EC_activated)),
  ,
  drop = FALSE
]

mat_top_deg_EC_activated_z <- t(scale(t(mat_top_deg_EC_activated)))
mat_top_deg_EC_activated_z[is.na(mat_top_deg_EC_activated_z)] <- 0

ComplexHeatmap::Heatmap(
  mat_top_deg_EC_activated_z,
  name = "Z-score",
  top_annotation = ha_EC_activated,
  col = heat_cols,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  column_names_gp = grid::gpar(fontsize = 5),
  row_names_gp = grid::gpar(fontsize = 10),
  column_title = "EC_activated: top 25 up and top 25 down genes",
  column_title_gp = grid::gpar(fontsize = 16, fontface = "bold")
)
```

### EC_activated GO GSEA

```{r ec_activated gsea run and save, eval=TRUE}
ranked_genes_EC_activated <- results_EC_activated %>%
  dplyr::select(gene, metric) %>%
  dplyr::filter(!is.na(metric))

entrezid_EC_activated <- clusterProfiler::bitr(
  ranked_genes_EC_activated$gene,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db,
  drop = TRUE
)

ranked_genes_entrez_EC_activated <- dplyr::left_join(
  ranked_genes_EC_activated,
  entrezid_EC_activated,
  by = c("gene" = "SYMBOL")
) %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::group_by(ENTREZID) %>%
  dplyr::slice_max(order_by = abs(metric), n = 1, with_ties = FALSE) %>%
  dplyr::ungroup()

genelist_EC_activated <- ranked_genes_entrez_EC_activated$metric
names(genelist_EC_activated) <- ranked_genes_entrez_EC_activated$ENTREZID
genelist_EC_activated <- sort(genelist_EC_activated, decreasing = TRUE)

GSEA_GO_EC_activated <- clusterProfiler::gseGO(
  geneList = genelist_EC_activated,
  OrgDb = org.Hs.eg.db,
  ont = "ALL",
  keyType = "ENTREZID",
  minGSSize = 10,
  maxGSSize = 500,
  eps = 0,
  nPermSimple = 10000
)

GSEA_GO_EC_activated <- clusterProfiler::setReadable(
  GSEA_GO_EC_activated,
  OrgDb = org.Hs.eg.db
)

GSEA_GO_EC_activated_simplified <- clusterProfiler::simplify(
  GSEA_GO_EC_activated,
  cutoff = 0.6,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)

qs2::qs_save(
  ranked_genes_EC_activated,
  file.path(output_dir_data, "ranked_genes_EC_activated.qs")
)

qs2::qs_save(
  ranked_genes_entrez_EC_activated,
  file.path(output_dir_data, "ranked_genes_entrez_EC_activated.qs")
)

qs2::qs_save(
  genelist_EC_activated,
  file.path(output_dir_data, "genelist_EC_activated.qs")
)

qs2::qs_save(
  GSEA_GO_EC_activated,
  file.path(output_dir_data, "GSEA_GO_EC_activated.qs")
)

qs2::qs_save(
  GSEA_GO_EC_activated_simplified,
  file.path(output_dir_data, "GSEA_GO_EC_activated_simplified.qs")
)

write.csv(
  GSEA_GO_EC_activated@result,
  file = file.path(output_dir_data, "GSEA_GO_EC_activated_full_results.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  GSEA_GO_EC_activated_simplified@result,
  file = file.path(output_dir_data, "GSEA_GO_EC_activated_simplified_results.csv"),
  quote = FALSE,
  row.names = FALSE
)
```

```{r ec_activated gsea load saved}
ranked_genes_EC_activated <- qs2::qs_read(
  file.path(output_dir_data, "ranked_genes_EC_activated.qs")
)

ranked_genes_entrez_EC_activated <- qs2::qs_read(
  file.path(output_dir_data, "ranked_genes_entrez_EC_activated.qs")
)

genelist_EC_activated <- qs2::qs_read(
  file.path(output_dir_data, "genelist_EC_activated.qs")
)

GSEA_GO_EC_activated <- qs2::qs_read(
  file.path(output_dir_data, "GSEA_GO_EC_activated.qs")
)

GSEA_GO_EC_activated_simplified <- qs2::qs_read(
  file.path(output_dir_data, "GSEA_GO_EC_activated_simplified.qs")
)
```

```{r ec_activated go dotplot, fig.width=10, fig.height=8}
p_go_dot_EC_activated <- make_gsea_dotplot(
  GSEA_GO_EC_activated_simplified,
  title_text = "EC_activated\nGene ontology enrichment analysis\nHF vs H",
  show_n = 20
)

if (!is.null(p_go_dot_EC_activated)) {
  p_go_dot_EC_activated
}
```
