---
title: "Endothelial cells Koenig Reichart manuscript"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
current_file <- "Endothelial_cells_Koenig_Reichart_manuscript"

library(here)
library(readxl)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(tibble)
library(patchwork)
library(edgeR)
library(limma)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(ggpubr)
library(reactable)
library(scales)
library(cowplot)
library(speckle)
library(broom)

output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"), recursive = TRUE)
output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs, recursive = TRUE)

base_data_dir <- here::here("data", "Heart_datasets")

cell_type_cols <- c(
  "#d05050", "#33A02C", "#1F78B4", "#CAB2D6", "#FF7F00", "#B2DF8A",
  "#B15928", "#FFD000", "#FA95C0", "#FDBF6F", "#6A3D9A", "#7aCDF9", "#E6E600"
)

group_cols <- c("H" = "grey85", "HF" = "grey45")
sex_cols <- c("F" = "#6C8EBF", "M" = "#8AB17D")
dataset_cols <- c("Koenig" = "#CAB2D6", "Reichart" = "#B2DF8A")
heat_cols <- circlize::colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

wrap_2lines <- function(x, width = 35) {
  vapply(x, function(s) {
    s <- as.character(s)
    if (nchar(s) <= width) return(s)
    cut <- stringr::str_locate_all(substr(s, 1, width), " ")[[1]]
    cut <- if (nrow(cut) == 0) width else max(cut[, 1])
    line1 <- stringr::str_trim(substr(s, 1, cut))
    rest  <- stringr::str_trim(substr(s, cut + 1, nchar(s)))
    if (nchar(rest) > width) rest <- paste0(substr(rest, 1, width - 1), "…")
    paste(line1, rest, sep = "\n")
  }, FUN.VALUE = character(1))
}


format_p_label <- function(p) {
  if (length(p) == 0 || is.na(p)) return("NA")
  dplyr::case_when(
    p < 0.0001 ~ "****",
    p < 0.001  ~ "***",
    p < 0.01   ~ "**",
    p < 0.05   ~ "*",
    TRUE ~ sprintf("%.4f", p)
  )
}

format_p_fontface <- function(p) {
  if (length(p) == 0 || is.na(p)) return("plain")
  dplyr::case_when(
    p < 0.05 ~ "bold",
    TRUE ~ "plain"
  )
}


volcano_cols <- c(
  "Decreased" = "#004C8C",
  "Not significant" = "grey75",
  "Increased" = "#A63A2A"
)
```

## Load datasets

```{r load datasets}
load(file.path(base_data_dir, "Koenig_2022", "GSE183852_DCM_Nuclei.Robj"))
Koenig <- nuclei
rm(nuclei)

Reichart <- readRDS(file.path(base_data_dir, "Reichart_2022", "DCM_ACM_heart_cell_atlas.rds"))

# Reichart uses Ensembl IDs in rownames, so replace them with gene symbols
Reichart_gene_names <- data.frame(
  ensembl_id = rownames(Reichart),
  gene_symbol = Reichart@assays[["RNA"]]@meta.features$feature_name
)

rownames(Reichart@assays[["RNA"]]@counts) <- Reichart@assays[["RNA"]]@meta.features$feature_name
rownames(Reichart@assays[["RNA"]]@data) <- Reichart@assays[["RNA"]]@meta.features$feature_name
```

## Add metadata

```{r add metadata}
Clinical_data_Koenig <- readxl::read_excel(
  file.path(base_data_dir, "Koenig_2022", "Koenig_et_al_2022_Selected_clinical_info.xlsx")
)
add_metadata_Koenig <- dplyr::left_join(Koenig[["orig.ident"]], Clinical_data_Koenig)
row.names(add_metadata_Koenig) <- row.names(Koenig[[]])
Koenig <- AddMetaData(Koenig, metadata = add_metadata_Koenig)
rm(add_metadata_Koenig, Clinical_data_Koenig)

Clinical_data_Reichart <- readxl::read_excel(
  file.path(base_data_dir, "Reichart_2022", "Reichart_et_al_Science_2022_Selected_clinical_info.xlsx")
)
add_metadata_Reichart <- dplyr::left_join(Reichart[["donor_id"]], Clinical_data_Reichart)
row.names(add_metadata_Reichart) <- row.names(Reichart[[]])
Reichart <- AddMetaData(Reichart, metadata = add_metadata_Reichart)
rm(add_metadata_Reichart, Clinical_data_Reichart)

Reichart$Sample <- Reichart$donor_id
Reichart$Group <- dplyr::case_when(
  Reichart$disease == "normal" ~ "HC",
  Reichart$disease == "dilated cardiomyopathy" ~ "DCM",
  TRUE ~ as.character(Reichart$disease)
)

Koenig$Sample <- Koenig$orig.ident
Koenig$Group <- dplyr::case_when(
  Koenig$Condition == "Donor" ~ "HC",
  Koenig$Condition == "DCM" ~ "DCM",
  TRUE ~ as.character(Koenig$Condition)
)

Koenig$cell_type_plot <- Koenig$Names

Reichart$cell_type_plot <- dplyr::case_when(
  Reichart$cell_type == "endothelial cell" & Reichart$cell_states == "EC7.0" ~ "endocardial cell",
  Reichart$cell_type == "endothelial cell" & Reichart$cell_states == "EC8.0" ~ "lymphatic cell",
  Reichart$cell_type == "endothelial cell" & Reichart$cell_states == "Meso" ~ "mesothelial cell",
  Reichart$cell_type == "endothelial cell" & Reichart$cell_states %in% c("EC1.0", "EC2.0", "EC5.0", "EC6.0") ~ "endothelial cell",
  TRUE ~ as.character(Reichart$cell_type)
)

Reichart <- subset(
  Reichart,
  subset = cell_type_plot != "unknown"
)
```

## Whole-dataset UMAPs

```{r whole dataset umaps, fig.width=12, fig.height=6}
koenig_levels <- sort(unique(as.character(Koenig$cell_type_plot)))
koenig_cols <- setNames(rep(cell_type_cols, length.out = length(koenig_levels)), koenig_levels)

reichart_levels <- sort(unique(as.character(Reichart$cell_type_plot)))
reichart_cols <- setNames(rep(cell_type_cols, length.out = length(reichart_levels)), reichart_levels)

p_koenig <- DimPlot(
  Koenig,
  group.by = "cell_type_plot",
  reduction = "umap",
  cols = koenig_cols,
  label = FALSE,
  shuffle = TRUE,
  pt.size = 1.2
) +
  NoLegend() +
  ggtitle(expression(italic("Koenig et al.") ~ " Nature CVR 2022")) +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5)
  )

p_koenig <- LabelClusters(
  p_koenig,
  id = "cell_type_plot",
  fontface = "bold",
  size = 5.2,
  repel = TRUE
)

p_reichart <- DimPlot(
  Reichart,
  group.by = "cell_type_plot",
  reduction = "umap",
  cols = reichart_cols,
  label = FALSE,
  shuffle = TRUE,
  pt.size = 1.2
) +
  NoLegend() +
  ggtitle(expression(italic("Reichart et al.") ~ " Science 2022")) +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5)
  )

p_reichart <- LabelClusters(
  p_reichart,
  id = "cell_type_plot",
  fontface = "bold",
  size = 5.2,
  repel = TRUE
)

p_koenig + p_reichart
```

```{r endothelial highlight umaps, fig.width=12, fig.height=6}
koenig_ec_cells <- rownames(Koenig@meta.data)[Koenig$Names %in% c("Endothelium")]
reichart_ec_cells <- rownames(Reichart@meta.data)[Reichart$cell_type_plot %in% c("endothelial cell")]

p_koenig_ec <- DimPlot(
  Koenig,
  reduction = "umap",
  cells.highlight = koenig_ec_cells,
  cols.highlight = "#1F78B4",
  cols = "grey85",
  sizes.highlight = 0.01,
  shuffle = TRUE,
  pt.size = 0.01,
  raster = FALSE
) +
  NoLegend() +
  ggtitle(expression(italic("Koenig et al.") ~ " Nature CVR 2022")) +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.position = "none"
  )

p_reichart_ec <- DimPlot(
  Reichart,
  reduction = "umap",
  cells.highlight = reichart_ec_cells,
  cols.highlight = "#1F78B4",
  cols = "grey85",
  sizes.highlight = 0.01,
  shuffle = TRUE,
  pt.size = 0.01,
  raster = FALSE
) +
  NoLegend() +
  ggtitle(expression(italic("Reichart et al.") ~ " Science 2022")) +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.position = "none"
  )

p_koenig_ec + p_reichart_ec
```

## Endothelial-cell subsets

```{r endothelial subsets}
Koenig_EC <- subset(Koenig, Names == "Endothelium")
Reichart_EC <- subset(
  Reichart,
  subset = cell_type_plot == "endothelial cell" & disease %in% c("normal", "dilated cardiomyopathy") & tissue == "heart left ventricle"
)

Koenig_EC$Dataset <- "Koenig"
Reichart_EC$Dataset <- "Reichart"

Koenig_EC$Group <- ifelse(Koenig_EC$Condition == "Donor", "H", "HF")
Reichart_EC$Group <- ifelse(Reichart_EC$disease == "normal", "H", "HF")

Koenig_EC$Sex <- as.character(Koenig_EC$Sex)
Reichart_EC$Sex <- as.character(Reichart_EC$sex)

Koenig_EC$Sex[Koenig_EC$Sex %in% c("Male", "male", "m")] <- "M"
Koenig_EC$Sex[Koenig_EC$Sex %in% c("Female", "female", "f")] <- "F"
Reichart_EC$Sex[Reichart_EC$Sex %in% c("Male", "male", "m")] <- "M"
Reichart_EC$Sex[Reichart_EC$Sex %in% c("Female", "female", "f")] <- "F"

age_midpoints <- c(
  "0-9" = 5,
  "10-19" = 15,
  "20-29" = 25,
  "30-39" = 35,
  "40-49" = 45,
  "50-59" = 55,
  "60-69" = 65,
  "70-79" = 75
)

Koenig_EC$Age <- as.numeric(as.character(Koenig_EC$Age))
Reichart_EC$Age <- unname(age_midpoints[as.character(Reichart_EC$Age_range)])
Reichart_EC$Age <- as.numeric(Reichart_EC$Age)
```

## Sample-level metadata

```{r sample metadata}
metadata_Koenig <- Koenig_EC@meta.data %>%
  dplyr::select(Sample = orig.ident, Age, Sex, Group, Dataset) %>%
  dplyr::distinct()

metadata_Reichart <- Reichart_EC@meta.data %>%
  dplyr::select(Sample = donor_id, Age, Sex, Group, Dataset) %>%
  dplyr::distinct()

metadata_Koenig$Group <- factor(metadata_Koenig$Group, levels = c("H", "HF"))
metadata_Reichart$Group <- factor(metadata_Reichart$Group, levels = c("H", "HF"))
metadata_Koenig$Sex <- factor(metadata_Koenig$Sex, levels = c("F", "M"))
metadata_Reichart$Sex <- factor(metadata_Reichart$Sex, levels = c("F", "M"))
metadata_Koenig$Dataset <- factor(metadata_Koenig$Dataset, levels = c("Koenig", "Reichart"))
metadata_Reichart$Dataset <- factor(metadata_Reichart$Dataset, levels = c("Koenig", "Reichart"))

reactable(metadata_Koenig)
reactable(metadata_Reichart)
```

## Pseudobulk per dataset

```{r pseudobulk koenig}
expr_matrix_Koenig <- GetAssayData(Koenig_EC, assay = "RNA", slot = "counts")
non_zero_cells_Koenig <- rowSums(expr_matrix_Koenig > 0)
percent_cells_Koenig <- non_zero_cells_Koenig / ncol(expr_matrix_Koenig) * 100
percent_stats_Koenig <- data.frame(gene = rownames(expr_matrix_Koenig), percentCells = percent_cells_Koenig)
keep_genes_Koenig <- percent_stats_Koenig$gene[percent_stats_Koenig$percentCells > 1]
Koenig_EC_pb_input <- Koenig_EC[keep_genes_Koenig, ]

pseudobulk_Koenig <- Seurat2PB(Koenig_EC_pb_input, sample = "orig.ident", cluster = "Names")
colnames(pseudobulk_Koenig) <- pseudobulk_Koenig$samples$sample
pseudobulk_Koenig <- pseudobulk_Koenig[, match(metadata_Koenig$Sample, pseudobulk_Koenig$samples$sample)]

keep_samples_Koenig <- pseudobulk_Koenig$samples$lib.size > 5e4
pseudobulk_Koenig <- pseudobulk_Koenig[, keep_samples_Koenig]
metadata_Koenig <- metadata_Koenig %>% dplyr::filter(Sample %in% colnames(pseudobulk_Koenig))
metadata_Koenig <- metadata_Koenig %>% dplyr::slice(match(colnames(pseudobulk_Koenig), Sample))
stopifnot(all(metadata_Koenig$Sample == colnames(pseudobulk_Koenig)))
```

```{r pseudobulk reichart}
expr_matrix_Reichart <- GetAssayData(Reichart_EC, assay = "RNA", slot = "counts")
non_zero_cells_Reichart <- rowSums(expr_matrix_Reichart > 0)
percent_cells_Reichart <- non_zero_cells_Reichart / ncol(expr_matrix_Reichart) * 100
percent_stats_Reichart <- data.frame(gene = rownames(expr_matrix_Reichart), percentCells = percent_cells_Reichart)
keep_genes_Reichart <- percent_stats_Reichart$gene[percent_stats_Reichart$percentCells > 1]
Reichart_EC_pb_input <- Reichart_EC[keep_genes_Reichart, ]

pseudobulk_Reichart <- Seurat2PB(Reichart_EC_pb_input, sample = "donor_id", cluster = "cell_type_plot")
colnames(pseudobulk_Reichart) <- pseudobulk_Reichart$samples$sample
pseudobulk_Reichart <- pseudobulk_Reichart[, match(metadata_Reichart$Sample, pseudobulk_Reichart$samples$sample)]

keep_samples_Reichart <- pseudobulk_Reichart$samples$lib.size > 5e4
pseudobulk_Reichart <- pseudobulk_Reichart[, keep_samples_Reichart]
metadata_Reichart <- metadata_Reichart %>% dplyr::filter(Sample %in% colnames(pseudobulk_Reichart))
metadata_Reichart <- metadata_Reichart %>% dplyr::slice(match(colnames(pseudobulk_Reichart), Sample))
stopifnot(all(metadata_Reichart$Sample == colnames(pseudobulk_Reichart)))
```

## Merge pseudobulks and run differential expression

```{r merge pseudobulk}
common_genes_pb <- intersect(rownames(pseudobulk_Koenig), rownames(pseudobulk_Reichart))
length(common_genes_pb)

combined_counts <- cbind(
  pseudobulk_Koenig[common_genes_pb, ]$counts,
  pseudobulk_Reichart[common_genes_pb, ]$counts
)

metadata_combined <- dplyr::bind_rows(metadata_Koenig, metadata_Reichart)
metadata_combined <- metadata_combined %>%
  dplyr::filter(Sample %in% colnames(combined_counts)) %>%
  dplyr::slice(match(colnames(combined_counts), Sample))

metadata_combined <- metadata_combined %>%
  dplyr::filter(!is.na(Age), !is.na(Sex), !is.na(Group), !is.na(Dataset))
combined_counts <- combined_counts[, metadata_combined$Sample]

metadata_combined$Group <- factor(metadata_combined$Group, levels = c("H", "HF"))
metadata_combined$Sex <- factor(metadata_combined$Sex, levels = c("F", "M"))
metadata_combined$Dataset <- factor(metadata_combined$Dataset, levels = c("Koenig", "Reichart"))

pseudobulk_combined <- edgeR::DGEList(counts = combined_counts)
pseudobulk_combined$samples <- cbind(pseudobulk_combined$samples, metadata_combined)

# Model: HF vs H, corrected for age, sex, and dataset
# Merging pseudobulks is fine here because the actual test is still sample-level,
# and dataset is included explicitly in the design.
design_combined <- model.matrix(~ Group + Age + Sex + Dataset, data = metadata_combined)
colnames(design_combined)

keep_genes_combined <- filterByExpr(pseudobulk_combined, design = design_combined)
pseudobulk_combined <- pseudobulk_combined[keep_genes_combined, , keep.lib.sizes = FALSE]
pseudobulk_combined <- normLibSizes(pseudobulk_combined)
pseudobulk_combined <- estimateDisp(pseudobulk_combined, design_combined)

logcpm_combined <- edgeR::cpm(pseudobulk_combined, log = TRUE, prior.count = 1)
```

```{r differential expression}
fit_combined <- glmQLFit(pseudobulk_combined, design_combined, robust = TRUE)
qlf_combined <- glmQLFTest(fit_combined, coef = "GroupHF")

results_combined <- topTags(qlf_combined, n = Inf)$table
results_combined <- results_combined %>%
  tibble::rownames_to_column("gene") %>%
  dplyr::mutate(metric = logFC * -log10(PValue))

results_combined$significance <- "Not significant"
results_combined$significance[results_combined$FDR < 0.05 & results_combined$logFC > 0.5] <- "Increased"
results_combined$significance[results_combined$FDR < 0.05 & results_combined$logFC < -0.5] <- "Decreased"
results_combined$significance <- factor(results_combined$significance,
                                        levels = c("Decreased", "Not significant", "Increased"))

signif_combined <- results_combined %>%
  dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>%
  dplyr::arrange(desc(metric))

write.csv(results_combined,
          file = file.path(output_dir_data, "EC_DE_results_Koenig_Reichart_combined.csv"),
          quote = FALSE, row.names = FALSE)
write.csv(signif_combined,
          file = file.path(output_dir_data, "EC_DE_signif_Koenig_Reichart_combined.csv"),
          quote = FALSE, row.names = FALSE)
```

## Volcano plot

```{r volcano, fig.width=4, fig.height=5}
n_h  <- sum(metadata_combined$Group == "H")
n_hf <- sum(metadata_combined$Group == "HF")
EnhancedVolcano(
  results_combined,
  lab = results_combined$gene,
  x = "logFC",
  y = "FDR",
  labSize = 0,
  legendLabSize = 12,
  titleLabSize = 14,
  axisLabSize = 12,
  captionLabSize = 0,
  pointSize = 0.8,
  FCcutoff = 0.5,
  pCutoff = 0.05,
  ylim = c(0, 20),
  xlim = c(-3.75, 3.75),
  colAlpha = 1,
  drawConnectors = FALSE,
  ylab = bquote(-Log[10]~FDR),
  subtitle = NULL,
  title = paste0("Endothelial cells\nHF (n=", n_hf, ") vs H (n=", n_h, ")"),
  colCustom = setNames(
    c("#004C8C", "grey75", "#A63A2A")[match(
      results_combined$significance,
      c("Decreased", "Not significant", "Increased")
    )],
    as.character(results_combined$significance)
  )
) +
  theme(legend.position = "none")
```

## Heatmaps

```{r order heatmap columns}
column_order <- metadata_combined %>%
  dplyr::arrange(Group, Dataset, Age, Sex) %>%
  dplyr::pull(Sample)

metadata_heatmap <- metadata_combined %>%
  dplyr::filter(Sample %in% column_order) %>%
  dplyr::slice(match(column_order, Sample))

ha <- HeatmapAnnotation(
  Group = metadata_heatmap$Group,
  Age = metadata_heatmap$Age,
  Sex = metadata_heatmap$Sex,
  Dataset = metadata_heatmap$Dataset,
  col = list(
    Group = group_cols,
    Age = circlize::colorRamp2(
      c(min(metadata_heatmap$Age, na.rm = TRUE), max(metadata_heatmap$Age, na.rm = TRUE)),
      c("grey90", "black")
    ),
    Sex = sex_cols,
    Dataset = dataset_cols
  ),
  annotation_name_gp = grid::gpar(fontsize = 10, fontface = "bold")
)
```

```{r heatmap all degs, fig.width=9, fig.height=12}
mat_all_deg <- logcpm_combined[rownames(logcpm_combined) %in% signif_combined$gene, column_order, drop = FALSE]
mat_all_deg <- mat_all_deg[match(signif_combined$gene, rownames(mat_all_deg)), , drop = FALSE]
mat_all_deg_z <- t(scale(t(mat_all_deg)))
mat_all_deg_z[is.na(mat_all_deg_z)] <- 0

Heatmap(
  mat_all_deg_z,
  name = "Z-score",
  top_annotation = ha,
  col = heat_cols,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  show_column_names = TRUE,
  column_names_gp = grid::gpar(fontsize = 5),
  row_names_gp = grid::gpar(fontsize = 6),
  column_title = "All significant DEGs",
  column_title_gp = grid::gpar(fontsize = 12, fontface = "bold")
)
```

```{r heatmap top 25 up down, fig.width=9, fig.height=8}
top_up <- results_combined %>%
  dplyr::filter(FDR < 0.05, logFC > 0.5) %>%
  dplyr::arrange(desc(metric)) %>%
  dplyr::slice_head(n = 25)

top_down <- results_combined %>%
  dplyr::filter(FDR < 0.05, logFC < -0.5) %>%
  dplyr::arrange(metric) %>%
  dplyr::slice_head(n = 25)

top_genes <- dplyr::bind_rows(top_up, top_down)

mat_top_deg <- logcpm_combined[rownames(logcpm_combined) %in% top_genes$gene, column_order, drop = FALSE]
mat_top_deg <- mat_top_deg[match(top_genes$gene, rownames(mat_top_deg)), , drop = FALSE]
mat_top_deg_z <- t(scale(t(mat_top_deg)))
mat_top_deg_z[is.na(mat_top_deg_z)] <- 0

Heatmap(
  mat_top_deg_z,
  name = "Z-score",
  top_annotation = ha,
  col = heat_cols,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  column_names_gp = grid::gpar(fontsize = 5),
  row_names_gp = grid::gpar(fontsize = 10),
  column_title = "Top 25 up and top 25 down genes",
  column_title_gp = grid::gpar(fontsize = 16, fontface = "bold")
)
```

## GSEA

## GSEA - run once and save objects

```{r gsea run and save, eval=FALSE}
ranked_genes <- results_combined %>%
  dplyr::select(gene, metric) %>%
  dplyr::filter(!is.na(metric))

entrezid <- bitr(
  ranked_genes$gene,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db,
  drop = TRUE
)

ranked_genes_entrez <- dplyr::left_join(
  ranked_genes,
  entrezid,
  by = c("gene" = "SYMBOL")
) %>%
  dplyr::filter(!is.na(ENTREZID)) %>%
  dplyr::group_by(ENTREZID) %>%
  dplyr::slice_max(order_by = abs(metric), n = 1, with_ties = FALSE) %>%
  dplyr::ungroup()

genelist <- ranked_genes_entrez$metric
names(genelist) <- ranked_genes_entrez$ENTREZID
genelist <- sort(genelist, decreasing = TRUE)

GSEA_GO_ALL <- gseGO(
  geneList = genelist,
  OrgDb = org.Hs.eg.db,
  ont = "ALL",
  keyType = "ENTREZID",
  minGSSize = 10,
  maxGSSize = 500,
  eps = 0,
  nPermSimple = 10000
)

GSEA_GO_ALL <- clusterProfiler::setReadable(
  GSEA_GO_ALL,
  OrgDb = org.Hs.eg.db
)

GSEA_GO_ALL_simplified <- clusterProfiler::simplify(
  GSEA_GO_ALL,
  cutoff = 0.6,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)

GSEA_KEGG <- gseKEGG(
  geneList = genelist,
  organism = "hsa",
  keyType = "ncbi-geneid",
  minGSSize = 10,
  maxGSSize = 500,
  eps = 0,
  nPermSimple = 10000
)

GSEA_KEGG <- clusterProfiler::setReadable(
  GSEA_KEGG,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
)

GSEA_REACTOME <- ReactomePA::gsePathway(
  geneList = genelist,
  organism = "human",
  minGSSize = 10,
  maxGSSize = 500,
  eps = 0,
  pvalueCutoff = 1,
  pAdjustMethod = "BH",
  verbose = FALSE
)

GSEA_REACTOME <- clusterProfiler::setReadable(
  GSEA_REACTOME,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
)

wp_df <- msigdbr::msigdbr(
  species = "Homo sapiens",
  collection = "C2",
  subcollection = "CP:WIKIPATHWAYS"
)

term2gene_wp <- wp_df %>%
  dplyr::select(gs_name, ncbi_gene) %>%
  dplyr::filter(!is.na(ncbi_gene)) %>%
  dplyr::distinct()

term2name_wp <- wp_df %>%
  dplyr::select(gs_name, gs_description) %>%
  dplyr::distinct()

GSEA_WIKIPATHWAYS <- clusterProfiler::GSEA(
  geneList = genelist,
  TERM2GENE = term2gene_wp,
  TERM2NAME = term2name_wp,
  minGSSize = 10,
  maxGSSize = 500,
  eps = 0,
  pvalueCutoff = 1,
  pAdjustMethod = "BH",
  verbose = FALSE
)

qs2::qs_save(
  ranked_genes,
  file.path(output_dir_data, "ranked_genes.qs")
)

qs2::qs_save(
  ranked_genes_entrez,
  file.path(output_dir_data, "ranked_genes_entrez.qs")
)

qs2::qs_save(
  genelist,
  file.path(output_dir_data, "genelist.qs")
)

qs2::qs_save(
  GSEA_GO_ALL,
  file.path(output_dir_data, "GSEA_GO_ALL.qs")
)

qs2::qs_save(
  GSEA_GO_ALL_simplified,
  file.path(output_dir_data, "GSEA_GO_ALL_simplified.qs")
)

qs2::qs_save(
  GSEA_KEGG,
  file.path(output_dir_data, "GSEA_KEGG.qs")
)

qs2::qs_save(
  GSEA_REACTOME,
  file.path(output_dir_data, "GSEA_REACTOME.qs")
)

qs2::qs_save(
  GSEA_WIKIPATHWAYS,
  file.path(output_dir_data, "GSEA_WIKIPATHWAYS.qs")
)

write.csv(
  GSEA_GO_ALL@result,
  file = file.path(output_dir_data, "GSEA_GO_ALL_full_results.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  GSEA_GO_ALL_simplified@result,
  file = file.path(output_dir_data, "GSEA_GO_ALL_simplified_results.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  GSEA_KEGG@result,
  file = file.path(output_dir_data, "GSEA_KEGG_results.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  GSEA_REACTOME@result,
  file = file.path(output_dir_data, "GSEA_REACTOME_results.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  GSEA_WIKIPATHWAYS@result,
  file = file.path(output_dir_data, "GSEA_WIKIPATHWAYS_results.csv"),
  quote = FALSE,
  row.names = FALSE
)
```

```{r gsea load saved}
ranked_genes <- qs2::qs_read(
  file.path(output_dir_data, "ranked_genes.qs")
)

ranked_genes_entrez <- qs2::qs_read(
  file.path(output_dir_data, "ranked_genes_entrez.qs")
)

genelist <- qs2::qs_read(
  file.path(output_dir_data, "genelist.qs")
)

GSEA_GO_ALL <- qs2::qs_read(
  file.path(output_dir_data, "GSEA_GO_ALL.qs")
)

GSEA_GO_ALL_simplified <- qs2::qs_read(
  file.path(output_dir_data, "GSEA_GO_ALL_simplified.qs")
)

GSEA_KEGG <- qs2::qs_read(
  file.path(output_dir_data, "GSEA_KEGG.qs")
)

GSEA_REACTOME <- qs2::qs_read(
  file.path(output_dir_data, "GSEA_REACTOME.qs")
)

GSEA_WIKIPATHWAYS <- qs2::qs_read(
  file.path(output_dir_data, "GSEA_WIKIPATHWAYS.qs")
)
```

## Helper function for dotplot / emapplot / barplot

```{r gsea helper functions}
make_gsea_dotplot <- function(gsea_obj, title_text, show_n = 20, label_format = 45, font_size = 12) {
  if (is.null(gsea_obj) || nrow(gsea_obj@result) == 0) return(NULL)

  dotplot(
    gsea_obj,
    showCategory = show_n,
    split = ".sign",
    label_format = label_format,
    font.size = font_size,
    title = title_text
  ) +
    facet_grid(. ~ .sign) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
      strip.text = element_text(face = "bold", size = 12)
    )
}

make_gsea_emapplot <- function(gsea_obj, show_n = 30) {
  if (is.null(gsea_obj) || nrow(gsea_obj@result) == 0) return(NULL)

  gsea_res <- gsea_obj@result

  gsea_up <- gsea_obj
  gsea_up@result <- gsea_res %>% dplyr::filter(NES > 0, p.adjust < 0.05)

  gsea_down <- gsea_obj
  gsea_down@result <- gsea_res %>% dplyr::filter(NES < 0, p.adjust < 0.05)

  plot_list_emap <- list()

  if (nrow(gsea_up@result) > 1) {
    plot_list_emap[["up"]] <- emapplot(
      pairwise_termsim(gsea_up),
      showCategory = min(show_n, nrow(gsea_up@result)),
      size_category = 0.7,
      min_edge = 0.25
    ) +
      ggtitle("Activated terms") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))
  }

  if (nrow(gsea_down@result) > 1) {
    plot_list_emap[["down"]] <- emapplot(
      pairwise_termsim(gsea_down),
      showCategory = min(show_n, nrow(gsea_down@result)),
      size_category = 0.7,
      min_edge = 0.25
    ) +
      ggtitle("Suppressed terms") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))
  }

  if (length(plot_list_emap) == 2) {
    return(plot_list_emap[["up"]] + plot_list_emap[["down"]])
  } else if (length(plot_list_emap) == 1) {
    return(plot_list_emap[[1]])
  } else {
    return(NULL)
  }
}

make_gsea_barplot <- function(gsea_obj, label_width = 40, top_n = 10, base_size = 11) {
  if (is.null(gsea_obj) || nrow(gsea_obj@result) == 0) return(NULL)

  gsea_bar_df <- gsea_obj@result %>%
    as_tibble() %>%
    dplyr::filter(!is.na(NES), !is.na(p.adjust), p.adjust < 0.05)

  if (nrow(gsea_bar_df) == 0) return(NULL)

  gsea_bar_df <- gsea_bar_df %>%
    dplyr::mutate(score = -log10(p.adjust) * sign(NES))

  gsea_bar_df_pos <- gsea_bar_df %>%
    dplyr::filter(NES > 0) %>%
    dplyr::arrange(desc(NES)) %>%
    dplyr::slice_head(n = top_n)

  gsea_bar_df_neg <- gsea_bar_df %>%
    dplyr::filter(NES < 0) %>%
    dplyr::arrange(NES) %>%
    dplyr::slice_head(n = top_n)

  gsea_bar_df_plot <- dplyr::bind_rows(gsea_bar_df_pos, gsea_bar_df_neg)

  if (nrow(gsea_bar_df_plot) == 0) return(NULL)

  gsea_bar_df_plot$Description <- factor(
    gsea_bar_df_plot$Description,
    levels = gsea_bar_df_plot$Description[order(gsea_bar_df_plot$NES)]
  )

  L <- unname(quantile(abs(gsea_bar_df_plot$score), probs = 0.80, na.rm = TRUE))
  if (!is.finite(L) || L == 0) L <- max(abs(gsea_bar_df_plot$score), na.rm = TRUE)

  ggplot(gsea_bar_df_plot, aes(x = Description, y = NES, fill = score)) +
    geom_col(width = 0.9) +
    scale_fill_gradient2(
      low = "#004C8C",
      mid = "white",
      high = "#A63A2A",
      midpoint = 0,
      limits = c(-L, L),
      oob = scales::squish,
      name = expression(-log[10](adj.~p) %*% sign(NES))
    ) +
    scale_x_discrete(labels = function(x) wrap_2lines(x, width = label_width)) +
    coord_flip() +
    labs(x = NULL, y = "Normalized enrichment score (NES)") +
    theme_bw(base_size = base_size) +
    theme(
      text = element_text(color = "black"),
      axis.text = element_text(color = "black"),
      axis.title = element_text(color = "black", size = 13),
      plot.title = element_blank(),
      axis.text.y = element_text(size = 11),
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 9)
    ) +
    guides(fill = guide_colorbar(
      barwidth = unit(6, "cm"),
      barheight = unit(0.3, "cm"),
      title.position = "top",
      title.hjust = 0.5
    ))
}
```

## GO dotplot

```{r gsea go dotplot, fig.width=10, fig.height=8}
p_go_dot <- make_gsea_dotplot(
  GSEA_GO_ALL_simplified,
  title_text = "Endothelial cells\nGene ontology enrichment analysis\nHF vs H",
  show_n = 20
)

if (!is.null(p_go_dot)) {
  p_go_dot
}
```

## GO emapplot

```{r gsea go emapplot, fig.width=16, fig.height=8}
p_go_emap <- make_gsea_emapplot(GSEA_GO_ALL_simplified, show_n = 30)

if (!is.null(p_go_emap)) {
  p_go_emap
}
```

## GO barplot

```{r gsea go barplot style, fig.width=6, fig.height=5}
p_go_bar <- make_gsea_barplot(GSEA_GO_ALL_simplified, label_width = 43)

if (!is.null(p_go_bar)) {
  p_go_bar
}
```

## KEGG dotplot

```{r gsea kegg dotplot, fig.width=10, fig.height=8}
p_kegg_dot <- make_gsea_dotplot(
  GSEA_KEGG,
  title_text = "Endothelial cells\nKEGG pathway enrichment analysis\nHF vs H",
  show_n = 20,
  label_format = 60
)

if (!is.null(p_kegg_dot)) {
  p_kegg_dot
}
```

## KEGG emapplot

```{r gsea kegg emapplot, fig.width=16, fig.height=8}
p_kegg_emap <- make_gsea_emapplot(GSEA_KEGG, show_n = 30)

if (!is.null(p_kegg_emap)) {
  p_kegg_emap
}
```

## KEGG barplot

```{r gsea kegg barplot style, fig.width=6, fig.height=5}
p_kegg_bar <- make_gsea_barplot(GSEA_KEGG, label_width = 43)

if (!is.null(p_kegg_bar)) {
  p_kegg_bar
}
```

## Reactome dotplot

```{r gsea reactome dotplot, fig.width=10, fig.height=8}
p_reactome_dot <- make_gsea_dotplot(
  GSEA_REACTOME,
  title_text = "Endothelial cells\nReactome pathway enrichment analysis\nHF vs H",
  show_n = 15,
  label_format = 60
)

if (!is.null(p_reactome_dot)) {
  p_reactome_dot
}
```

## Reactome emapplot

```{r gsea reactome emapplot, fig.width=16, fig.height=8}
p_reactome_emap <- make_gsea_emapplot(GSEA_REACTOME, show_n = 30)

if (!is.null(p_reactome_emap)) {
  p_reactome_emap
}
```

## Reactome barplot

```{r gsea reactome barplot style, fig.width=7.3, fig.height=5}
p_reactome_bar <- make_gsea_barplot(GSEA_REACTOME, label_width = 59)

if (!is.null(p_reactome_bar)) {
  p_reactome_bar
}
```

## WikiPathways dotplot

```{r gsea wikipathways dotplot, fig.width=10, fig.height=8}
p_wiki_dot <- make_gsea_dotplot(
  GSEA_WIKIPATHWAYS,
  title_text = "Endothelial cells\nWikiPathways enrichment analysis\nHF vs H",
  show_n = 20,
  label_format = 65
)

if (!is.null(p_wiki_dot)) {
  p_wiki_dot
}
```

## WikiPathways emapplot

```{r gsea wikipathways emapplot, fig.width=16, fig.height=8}
p_wiki_emap <- make_gsea_emapplot(GSEA_WIKIPATHWAYS, show_n = 30)

if (!is.null(p_wiki_emap)) {
  p_wiki_emap
}
```

## WikiPathways barplot

```{r gsea wikipathways barplot style, fig.width=6, fig.height=5}
p_wiki_bar <- make_gsea_barplot(GSEA_WIKIPATHWAYS, label_width = 43)

if (!is.null(p_wiki_bar)) {
  p_wiki_bar
}
```

## Cell cycle scoring in each dataset separately

```{r cell cycle scoring koenig}
s.genes <- Seurat::cc.genes.updated.2019$s.genes
g2m.genes <- Seurat::cc.genes.updated.2019$g2m.genes

Koenig_EC <- CellCycleScoring(
  Koenig_EC,
  s.features = s.genes,
  g2m.features = g2m.genes,
  set.ident = FALSE
)

Reichart_EC <- CellCycleScoring(
  Reichart_EC,
  s.features = s.genes,
  g2m.features = g2m.genes,
  set.ident = FALSE
)
```

## UMAPs colored by cell-cycle phase

```{r cell cycle umaps, fig.width=12, fig.height=6}
p_koenig_cc <- DimPlot(
  Koenig_EC,
  reduction = "umap",
  group.by = "Phase",
  shuffle = TRUE,
  pt.size = 0.7,
  raster = FALSE
) +
  ggtitle(expression(italic("Koenig et al.") ~ " Nature CVR 2022")) +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 10, face = "bold")
  )

p_reichart_cc <- DimPlot(
  Reichart_EC,
  reduction = "umap",
  group.by = "Phase",
  shuffle = TRUE,
  pt.size = 0.7,
  raster = FALSE
) +
  ggtitle(expression(italic("Reichart et al.") ~ " Science 2022")) +
  theme(
    axis.text = element_text(size = 8, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 10, face = "bold")
  )

(p_koenig_cc + p_reichart_cc) + patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "right")
```

## Phase proportions per sample

```{r cell cycle percentages per sample}
Koenig_EC$Sample <- Koenig_EC$orig.ident
Reichart_EC$Sample <- Reichart_EC$donor_id

df_phase_koenig <- Koenig_EC@meta.data %>%
  dplyr::group_by(Sample, Dataset, Group, Sex, Age, Phase) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(Sample, Dataset, Group, Sex, Age) %>%
  dplyr::mutate(
    total_n = sum(n),
    pct = n / total_n * 100
  ) %>%
  dplyr::ungroup()

df_phase_reichart <- Reichart_EC@meta.data %>%
  dplyr::group_by(Sample, Dataset, Group, Sex, Age, Phase) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(Sample, Dataset, Group, Sex, Age) %>%
  dplyr::mutate(
    total_n = sum(n),
    pct = n / total_n * 100
  ) %>%
  dplyr::ungroup()

df_phase <- dplyr::bind_rows(df_phase_koenig, df_phase_reichart)

df_phase$Group <- factor(df_phase$Group, levels = c("H", "HF"))
df_phase$Phase <- factor(df_phase$Phase, levels = c("G1", "S", "G2M"))
df_phase$Dataset <- factor(df_phase$Dataset, levels = c("Koenig", "Reichart"))
df_phase$Sex <- factor(df_phase$Sex)
df_phase$Age <- as.numeric(df_phase$Age)
```

## Cell-cycle statistics

```{r cell cycle stats}
phase_stats <- list()

for (ph in c("G1", "S", "G2M")) {
  df_sub <- df_phase %>% dplyr::filter(Phase == ph)

  fit_phase <- glm(
    cbind(n, total_n - n) ~ Group + Age + Sex + Dataset,
    family = quasibinomial,
    data = df_sub
  )

  phase_stats[[ph]] <- broom::tidy(fit_phase)
}

phase_stats_df <- dplyr::bind_rows(phase_stats, .id = "Phase")

write.csv(
  phase_stats_df,
  file = file.path(output_dir_data, "Cell_cycle_phase_stats.csv"),
  quote = FALSE,
  row.names = FALSE
)

phase_stats_df
```

## Cell-cycle barplots

```{r cell cycle barplots, fig.width=8, fig.height=3.5}
group_cols_cycle <- c("H" = "#D9D9D9", "HF" = "#7F7F7F")

plots_phase <- list()

for (ph in c("G1", "S", "G2M")) {
  df_sub <- df_phase %>% dplyr::filter(Phase == ph)

  fit_phase <- glm(
    cbind(n, total_n - n) ~ Group + Age + Sex + Dataset,
    family = quasibinomial,
    data = df_sub
  )

  fit_tidy <- broom::tidy(fit_phase)

  p_group <- fit_tidy %>%
    dplyr::filter(term == "GroupHF") %>%
    dplyr::pull(p.value)

  p_label <- format_p_label(p_group)

  p <- ggplot(df_sub, aes(x = Group, y = pct, fill = Group)) +
    stat_summary(fun = mean, geom = "bar", width = 0.65, color = "black") +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.15, linewidth = 0.5) +
    geom_jitter(
      width = 0.10,
      size = 1.1,
      shape = 21,
      stroke = 0.35,
      color = "black"
    ) +
    geom_segment(aes(x = 1, xend = 2, y = 46.4, yend = 46.4), inherit.aes = FALSE) +
    annotate("text", x = 1.5, y = 47.3, label = p_label, size = 5, fontface = format_p_fontface(p_group)) +
    scale_fill_manual(values = group_cols_cycle) +
    labs(title = ph, x = NULL, y = "Percentage of cells") +
    theme_classic(base_size = 12) +
    scale_y_continuous(limits = c(0, 50), expand = c(0, 0)) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.title.y = element_text(face = "bold"),
      legend.position = "none"
    )

  plots_phase[[ph]] <- p
}

plots_phase[["G1"]] + plots_phase[["S"]] + plots_phase[["G2M"]]
```

## Genes and scores of interest

```{r gene sets}
aerobic_respiration_row <- GSEA_GO_ALL@result %>%
  as_tibble() %>%
  dplyr::filter(Description == "aerobic respiration") %>%
  dplyr::slice(1)

if (nrow(aerobic_respiration_row) == 0) {
  stop("GO term 'aerobic respiration' was not found in GSEA_GO_ALL@result.")
}

aerobic_respiration_go_id <- aerobic_respiration_row$ID[1]

aerobic_respiration_entrez <- GSEA_GO_ALL@geneSets[[aerobic_respiration_go_id]]

aerobic_respiration_genes_df <- bitr(
  aerobic_respiration_entrez,
  fromType = "ENTREZID",
  toType = "SYMBOL",
  OrgDb = org.Hs.eg.db,
  drop = TRUE
)

aerobic_respiration_genes <- unique(aerobic_respiration_genes_df$SYMBOL)

write.csv(
  data.frame(gene = aerobic_respiration_genes),
  file = file.path(output_dir_data, "GO_aerobic_respiration_genes.csv"),
  quote = FALSE,
  row.names = FALSE
)

angiogenesis_genes <- c("ESM1", "ANGPT2", "FLT1", "KDR", "APLN", "EDN1")

redox_no_genes <- c("NOX4", "NOS3", "SOD1")

mito_homeostasis_genes <- c("TFAM", "PPARGC1A")

apoptotic_priming_genes <- c(
  "BCL2L11", "BBC3", "PMAIP1", "BAX",
  "BID", "BAD", "BMF", "BIK",
  "BNIP3", "BNIP3L"
)

all_genes_of_interest <- unique(c(
  angiogenesis_genes,
  redox_no_genes,
  mito_homeostasis_genes,
  apoptotic_priming_genes,
  aerobic_respiration_genes
))
```

```{r pseudobulk scores}
Koenig_EC$Sample <- Koenig_EC$orig.ident
Reichart_EC$Sample <- Reichart_EC$donor_id

Koenig_EC$cell_type <- "Endothelium"
Reichart_EC$cell_type <- "Endothelium"

pb_koenig_scores <- Seurat2PB(
  Koenig_EC,
  sample = "Sample",
  cluster = "cell_type"
)

pb_reichart_scores <- Seurat2PB(
  Reichart_EC,
  sample = "Sample",
  cluster = "cell_type"
)

colnames(pb_koenig_scores) <- pb_koenig_scores$samples$sample
colnames(pb_reichart_scores) <- pb_reichart_scores$samples$sample

common_genes_scores <- intersect(rownames(pb_koenig_scores), rownames(pb_reichart_scores))

counts_combined_scores <- cbind(
  pb_koenig_scores[common_genes_scores, ]$counts,
  pb_reichart_scores[common_genes_scores, ]$counts
)

metadata_scores <- dplyr::bind_rows(
  Koenig_EC@meta.data %>%
    dplyr::select(Sample, Dataset, Group, Sex, Age) %>%
    dplyr::distinct(),
  Reichart_EC@meta.data %>%
    dplyr::select(Sample, Dataset, Group, Sex, Age) %>%
    dplyr::distinct()
)

metadata_scores <- metadata_scores %>%
  dplyr::distinct(Sample, .keep_all = TRUE) %>%
  dplyr::filter(Sample %in% colnames(counts_combined_scores))

metadata_scores <- metadata_scores[match(colnames(counts_combined_scores), metadata_scores$Sample), ]

stopifnot(all(metadata_scores$Sample == colnames(counts_combined_scores)))

metadata_scores$Group <- factor(metadata_scores$Group, levels = c("H", "HF"))
metadata_scores$Sex <- factor(metadata_scores$Sex)
metadata_scores$Dataset <- factor(metadata_scores$Dataset, levels = c("Koenig", "Reichart"))
metadata_scores$Age <- as.numeric(metadata_scores$Age)

dge_scores <- edgeR::DGEList(counts = counts_combined_scores)
dge_scores <- edgeR::calcNormFactors(dge_scores)

logcpm_scores <- edgeR::cpm(dge_scores, log = TRUE, prior.count = 1)
```

## Keep only genes of interest for scores and barplots

```{r score matrices}
all_genes_of_interest_present <- all_genes_of_interest[all_genes_of_interest %in% rownames(logcpm_scores)]
all_genes_of_interest_missing <- setdiff(all_genes_of_interest, rownames(logcpm_scores))

logcpm_subset_scores <- logcpm_scores[
  rownames(logcpm_scores) %in% all_genes_of_interest_present,
  ,
  drop = FALSE
]

logcpm_subset_z <- t(scale(t(logcpm_subset_scores)))
logcpm_subset_z[is.na(logcpm_subset_z)] <- 0

angiogenesis_score <- colMeans(
  logcpm_subset_z[rownames(logcpm_subset_z) %in% angiogenesis_genes, , drop = FALSE]
)

apoptotic_priming_score <- colMeans(
  logcpm_subset_z[rownames(logcpm_subset_z) %in% apoptotic_priming_genes, , drop = FALSE]
)

aerobic_respiration_score <- colMeans(
  logcpm_subset_z[rownames(logcpm_subset_z) %in% aerobic_respiration_genes, , drop = FALSE]
)

score_df <- metadata_scores %>%
  dplyr::mutate(
    angiogenesis_score = angiogenesis_score[Sample],
    apoptotic_priming_score = apoptotic_priming_score[Sample],
    aerobic_respiration_score = aerobic_respiration_score[Sample]
  )

write.csv(
  score_df,
  file = file.path(output_dir_data, "EC_gene_set_scores_unfilteredPB.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  data.frame(gene_missing = all_genes_of_interest_missing),
  file = file.path(output_dir_data, "EC_genes_of_interest_missing_unfilteredPB.csv"),
  quote = FALSE,
  row.names = FALSE
)

write.csv(
  data.frame(gene = setdiff(aerobic_respiration_genes, rownames(logcpm_scores))),
  file = file.path(output_dir_data, "GO_aerobic_respiration_genes_missing_in_scores_matrix.csv"),
  quote = FALSE,
  row.names = FALSE
)
```

## Pseudobulk score boxplots

```{r aerobic respiration score, fig.width=3, fig.height=3.7}
fit_aerobic_score <- lm(aerobic_respiration_score ~ Group + Age + Sex + Dataset, data = score_df)
fit_aerobic_tidy <- broom::tidy(fit_aerobic_score)
print(fit_aerobic_tidy)

p_aerobic <- fit_aerobic_tidy %>%
  dplyr::filter(term == "GroupHF") %>%
  dplyr::pull(p.value)

p_aerobic_label <- format_p_label(p_aerobic)

y_max_aerobic <- max(score_df$aerobic_respiration_score, na.rm = TRUE)
y_min_aerobic <- min(score_df$aerobic_respiration_score, na.rm = TRUE)
y_range_aerobic <- y_max_aerobic - y_min_aerobic
if (!is.finite(y_range_aerobic) || y_range_aerobic == 0) y_range_aerobic <- 1

ggplot(score_df, aes(x = Group, y = aerobic_respiration_score, fill = Group)) +
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    color = "black",
    linewidth = 0.5
  ) +
  geom_jitter(
    width = 0.10,
    size = 1.5,
    shape = 21,
    stroke = 0.4,
    color = "black"
  ) +
  geom_segment(
    aes(
      x = 1,
      xend = 2,
      y = y_max_aerobic + 0.10 * y_range_aerobic,
      yend = y_max_aerobic + 0.10 * y_range_aerobic
    ),
    inherit.aes = FALSE
  ) +
  annotate(
    "text",
    x = 1.5,
    y = y_max_aerobic + 0.16 * y_range_aerobic,
    label = p_aerobic_label,
    size = 5,
    fontface = format_p_fontface(p_aerobic)
  ) +
  scale_fill_manual(values = group_cols) +
  coord_cartesian(
    ylim = c(
      y_min_aerobic - 0.05 * y_range_aerobic,
      y_max_aerobic + 0.22 * y_range_aerobic
    )
  ) +
  labs(title = "Aerobic respiration score", x = NULL, y = "Score") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "none"
  )
```

```{r angiogenesis score, fig.width=3, fig.height=3.7}
fit_angio_score <- lm(angiogenesis_score ~ Group + Age + Sex + Dataset, data = score_df)
fit_angio_tidy <- broom::tidy(fit_angio_score)
print(fit_angio_tidy)

p_angio <- fit_angio_tidy %>%
  dplyr::filter(term == "GroupHF") %>%
  dplyr::pull(p.value)

p_angio_label <- format_p_label(p_angio)

y_max_angio <- max(score_df$angiogenesis_score, na.rm = TRUE)
y_min_angio <- min(score_df$angiogenesis_score, na.rm = TRUE)
y_range_angio <- y_max_angio - y_min_angio
if (!is.finite(y_range_angio) || y_range_angio == 0) y_range_angio <- 1

ggplot(score_df, aes(x = Group, y = angiogenesis_score, fill = Group)) +
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    color = "black",
    linewidth = 0.5
  ) +
  geom_jitter(
    width = 0.10,
    size = 1.5,
    shape = 21,
    stroke = 0.4,
    color = "black"
  ) +
  geom_segment(
    aes(
      x = 1,
      xend = 2,
      y = y_max_angio + 0.10 * y_range_angio,
      yend = y_max_angio + 0.10 * y_range_angio
    ),
    inherit.aes = FALSE
  ) +
  annotate(
    "text",
    x = 1.5,
    y = y_max_angio + 0.16 * y_range_angio,
    label = p_angio_label,
    size = 5,
    fontface = format_p_fontface(p_angio)
  ) +
  scale_fill_manual(values = group_cols) +
  coord_cartesian(
    ylim = c(
      y_min_angio - 0.05 * y_range_angio,
      y_max_angio + 0.22 * y_range_angio
    )
  ) +
  labs(title = "Angiogenesis score", x = NULL, y = "Score") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "none"
  )
```

```{r apoptotic priming score, fig.width=3, fig.height=3.7}
fit_apop_score <- lm(apoptotic_priming_score ~ Group + Age + Sex + Dataset, data = score_df)
fit_apop_tidy <- broom::tidy(fit_apop_score)
print(fit_apop_tidy)

p_apop <- fit_apop_tidy %>%
  dplyr::filter(term == "GroupHF") %>%
  dplyr::pull(p.value)

p_apop_label <- format_p_label(p_apop)

y_max_apop <- max(score_df$apoptotic_priming_score, na.rm = TRUE)
y_min_apop <- min(score_df$apoptotic_priming_score, na.rm = TRUE)
y_range_apop <- y_max_apop - y_min_apop
if (!is.finite(y_range_apop) || y_range_apop == 0) y_range_apop <- 1

ggplot(score_df, aes(x = Group, y = apoptotic_priming_score, fill = Group)) +
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    color = "black",
    linewidth = 0.5
  ) +
  geom_jitter(
    width = 0.10,
    size = 1.5,
    shape = 21,
    stroke = 0.4,
    color = "black"
  ) +
  geom_segment(
    aes(
      x = 1,
      xend = 2,
      y = y_max_apop + 0.10 * y_range_apop,
      yend = y_max_apop + 0.10 * y_range_apop
    ),
    inherit.aes = FALSE
  ) +
  annotate(
    "text",
    x = 1.5,
    y = y_max_apop + 0.16 * y_range_apop,
    label = p_apop_label,
    size = 5,
    fontface = format_p_fontface(p_apop)
  ) +
  scale_fill_manual(values = group_cols) +
  coord_cartesian(
    ylim = c(
      y_min_apop - 0.05 * y_range_apop,
      y_max_apop + 0.22 * y_range_apop
    )
  ) +
  labs(title = "Apoptotic priming score", x = NULL, y = "Score") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "none"
  )
```


## Feature plots for genes of interest in each dataset separately

```{r feature plots angiogenesis, fig.width=11, fig.height=4.8}
for (gene in angiogenesis_genes) {

  p_list <- list()

  if (gene %in% rownames(Koenig_EC)) {
    p_koenig <- FeaturePlot(
      Koenig_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(expression(italic("Koenig et al.") ~ " Nature CVR 2022")) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Koenig"]] <- p_koenig
  }

  if (gene %in% rownames(Reichart_EC)) {
    p_reichart <- FeaturePlot(
      Reichart_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(expression(italic("Reichart et al.") ~ " Science 2022")) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Reichart"]] <- p_reichart
  }

  if (length(p_list) == 2) {
    print(p_list[["Koenig"]] | p_list[["Reichart"]])
  } else if (length(p_list) == 1) {
    print(p_list[[1]])
  } else {
    message("Gene not found in either dataset: ", gene)
  }
}
```


## Feature plots for genes of interest in each dataset separately

```{r feature plots redox no, fig.width=11, fig.height=4.8}
for (gene in redox_no_genes) {

  p_list <- list()

  if (gene %in% rownames(Koenig_EC)) {
    p_koenig <- FeaturePlot(
      Koenig_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(expression(italic("Koenig et al.") ~ " Nature CVR 2022")) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Koenig"]] <- p_koenig
  }

  if (gene %in% rownames(Reichart_EC)) {
    p_reichart <- FeaturePlot(
      Reichart_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(expression(italic("Reichart et al.") ~ " Science 2022")) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Reichart"]] <- p_reichart
  }

  if (length(p_list) == 2) {
    print(p_list[["Koenig"]] | p_list[["Reichart"]])
  } else if (length(p_list) == 1) {
    print(p_list[[1]])
  } else {
    message("Gene not found in either dataset: ", gene)
  }
}
```


## Feature plots for genes of interest in each dataset separately

```{r feature plots mito homeostasis, fig.width=11, fig.height=4.8}
for (gene in mito_homeostasis_genes) {

  p_list <- list()

  if (gene %in% rownames(Koenig_EC)) {
    p_koenig <- FeaturePlot(
      Koenig_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(expression(italic("Koenig et al.") ~ " Nature CVR 2022")) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Koenig"]] <- p_koenig
  }

  if (gene %in% rownames(Reichart_EC)) {
    p_reichart <- FeaturePlot(
      Reichart_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(expression(italic("Reichart et al.") ~ " Science 2022")) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Reichart"]] <- p_reichart
  }

  if (length(p_list) == 2) {
    print(p_list[["Koenig"]] | p_list[["Reichart"]])
  } else if (length(p_list) == 1) {
    print(p_list[[1]])
  } else {
    message("Gene not found in either dataset: ", gene)
  }
}
```


## Feature plots for genes of interest in each dataset separately

```{r feature plots apoptosis, fig.width=11, fig.height=4.8}
for (gene in apoptotic_priming_genes) {

  p_list <- list()

  if (gene %in% rownames(Koenig_EC)) {
    p_koenig <- FeaturePlot(
      Koenig_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(expression(italic("Koenig et al.") ~ " Nature CVR 2022")) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Koenig"]] <- p_koenig
  }

  if (gene %in% rownames(Reichart_EC)) {
    p_reichart <- FeaturePlot(
      Reichart_EC,
      features = gene,
      order = TRUE,
      cols = c("grey90", "#A63A2A"),
      pt.size = 0.15,
      combine = TRUE,
      raster = FALSE
    ) +
      ggtitle(expression(italic("Reichart et al.") ~ " Science 2022")) +
      theme(
        axis.text = element_text(size = 8, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 18, hjust = 0.5)
      )

    p_list[["Reichart"]] <- p_reichart
  }

  if (length(p_list) == 2) {
    print(p_list[["Koenig"]] | p_list[["Reichart"]])
  } else if (length(p_list) == 1) {
    print(p_list[[1]])
  } else {
    message("Gene not found in either dataset: ", gene)
  }
}
```


## Gene-wise pseudobulk logCPM plots with model-based p values

```{r gene wise dataframe}
gene_barplot_df <- logcpm_scores %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene") %>%
  tidyr::pivot_longer(-gene, names_to = "Sample", values_to = "logCPM") %>%
  dplyr::left_join(metadata_scores, by = "Sample")
```


```{r barplots angiogenesis genes, fig.width=2.66, fig.height=3.7}
for (gene in angiogenesis_genes) {
  if (gene %in% unique(gene_barplot_df$gene)) {
    df_sub <- gene_barplot_df[gene_barplot_df$gene == gene, ]
    fit_gene <- lm(logCPM ~ Group + Age + Sex + Dataset, data = df_sub)
    fit_gene_tidy <- broom::tidy(fit_gene)
    print(fit_gene_tidy)

    p_gene <- fit_gene_tidy %>%
      dplyr::filter(term == "GroupHF") %>%
      dplyr::pull(p.value)

    p_gene_label <- format_p_label(p_gene)

    y_max_gene <- max(df_sub$logCPM, na.rm = TRUE)
    y_min_gene <- min(df_sub$logCPM, na.rm = TRUE)
    y_range_gene <- y_max_gene - y_min_gene
    if (!is.finite(y_range_gene) || y_range_gene == 0) y_range_gene <- 1

    p <- ggplot(df_sub, aes(x = Group, y = logCPM, fill = Group)) +
      geom_boxplot(
        width = 0.55,
        outlier.shape = NA,
        color = "black",
        linewidth = 0.5
      ) +
      geom_jitter(
        width = 0.10,
        size = 1.5,
        shape = 21,
        stroke = 0.4,
        color = "black"
      ) +
      geom_segment(
        aes(
          x = 1,
          xend = 2,
          y = y_max_gene + 0.10 * y_range_gene,
          yend = y_max_gene + 0.10 * y_range_gene
        ),
        inherit.aes = FALSE
      ) +
      annotate(
        "text",
        x = 1.5,
        y = y_max_gene + 0.16 * y_range_gene,
        label = p_gene_label,
        size = 5,
        fontface = "bold"
      ) +
      scale_fill_manual(values = group_cols) +
      coord_cartesian(
        ylim = c(
          y_min_gene - 0.05 * y_range_gene,
          y_max_gene + 0.22 * y_range_gene
        )
      ) +
      labs(title = gene, x = NULL, y = "logCPM") +
      theme_classic(base_size = 12) +
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.position = "none"
      )

    print(p)
  }
}
```


```{r barplots redox no genes, fig.width=2.66, fig.height=3.7}
for (gene in redox_no_genes) {
  if (gene %in% unique(gene_barplot_df$gene)) {
    df_sub <- gene_barplot_df[gene_barplot_df$gene == gene, ]
    fit_gene <- lm(logCPM ~ Group + Age + Sex + Dataset, data = df_sub)
    fit_gene_tidy <- broom::tidy(fit_gene)
    print(fit_gene_tidy)

    p_gene <- fit_gene_tidy %>%
      dplyr::filter(term == "GroupHF") %>%
      dplyr::pull(p.value)

    p_gene_label <- format_p_label(p_gene)

    y_max_gene <- max(df_sub$logCPM, na.rm = TRUE)
    y_min_gene <- min(df_sub$logCPM, na.rm = TRUE)
    y_range_gene <- y_max_gene - y_min_gene
    if (!is.finite(y_range_gene) || y_range_gene == 0) y_range_gene <- 1

    p <- ggplot(df_sub, aes(x = Group, y = logCPM, fill = Group)) +
      geom_boxplot(
        width = 0.55,
        outlier.shape = NA,
        color = "black",
        linewidth = 0.5
      ) +
      geom_jitter(
        width = 0.10,
        size = 1.5,
        shape = 21,
        stroke = 0.4,
        color = "black"
      ) +
      geom_segment(
        aes(
          x = 1,
          xend = 2,
          y = y_max_gene + 0.10 * y_range_gene,
          yend = y_max_gene + 0.10 * y_range_gene
        ),
        inherit.aes = FALSE
      ) +
      annotate(
        "text",
        x = 1.5,
        y = y_max_gene + 0.16 * y_range_gene,
        label = p_gene_label,
        size = 5,
        fontface = "bold"
      ) +
      scale_fill_manual(values = group_cols) +
      coord_cartesian(
        ylim = c(
          y_min_gene - 0.05 * y_range_gene,
          y_max_gene + 0.22 * y_range_gene
        )
      ) +
      labs(title = gene, x = NULL, y = "logCPM") +
      theme_classic(base_size = 12) +
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.position = "none"
      )

    print(p)
  }
}
```


```{r barplots mito homeostasis genes, fig.width=2.66, fig.height=3.7}
for (gene in mito_homeostasis_genes) {
  if (gene %in% unique(gene_barplot_df$gene)) {
    df_sub <- gene_barplot_df[gene_barplot_df$gene == gene, ]
    fit_gene <- lm(logCPM ~ Group + Age + Sex + Dataset, data = df_sub)
    fit_gene_tidy <- broom::tidy(fit_gene)
    print(fit_gene_tidy)

    p_gene <- fit_gene_tidy %>%
      dplyr::filter(term == "GroupHF") %>%
      dplyr::pull(p.value)

    p_gene_label <- format_p_label(p_gene)

    y_max_gene <- max(df_sub$logCPM, na.rm = TRUE)
    y_min_gene <- min(df_sub$logCPM, na.rm = TRUE)
    y_range_gene <- y_max_gene - y_min_gene
    if (!is.finite(y_range_gene) || y_range_gene == 0) y_range_gene <- 1

    p <- ggplot(df_sub, aes(x = Group, y = logCPM, fill = Group)) +
      geom_boxplot(
        width = 0.55,
        outlier.shape = NA,
        color = "black",
        linewidth = 0.5
      ) +
      geom_jitter(
        width = 0.10,
        size = 1.5,
        shape = 21,
        stroke = 0.4,
        color = "black"
      ) +
      geom_segment(
        aes(
          x = 1,
          xend = 2,
          y = y_max_gene + 0.10 * y_range_gene,
          yend = y_max_gene + 0.10 * y_range_gene
        ),
        inherit.aes = FALSE
      ) +
      annotate(
        "text",
        x = 1.5,
        y = y_max_gene + 0.16 * y_range_gene,
        label = p_gene_label,
        size = 5,
        fontface = "bold"
      ) +
      scale_fill_manual(values = group_cols) +
      coord_cartesian(
        ylim = c(
          y_min_gene - 0.05 * y_range_gene,
          y_max_gene + 0.22 * y_range_gene
        )
      ) +
      labs(title = gene, x = NULL, y = "logCPM") +
      theme_classic(base_size = 12) +
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.position = "none"
      )

    print(p)
  }
}
```


```{r barplots apoptotic priming genes, fig.width=2.66, fig.height=3.7}
for (gene in apoptotic_priming_genes) {
  if (gene %in% unique(gene_barplot_df$gene)) {
    df_sub <- gene_barplot_df[gene_barplot_df$gene == gene, ]
    fit_gene <- lm(logCPM ~ Group + Age + Sex + Dataset, data = df_sub)
    fit_gene_tidy <- broom::tidy(fit_gene)
    print(fit_gene_tidy)

    p_gene <- fit_gene_tidy %>%
      dplyr::filter(term == "GroupHF") %>%
      dplyr::pull(p.value)

    p_gene_label <- format_p_label(p_gene)

    y_max_gene <- max(df_sub$logCPM, na.rm = TRUE)
    y_min_gene <- min(df_sub$logCPM, na.rm = TRUE)
    y_range_gene <- y_max_gene - y_min_gene
    if (!is.finite(y_range_gene) || y_range_gene == 0) y_range_gene <- 1

    p <- ggplot(df_sub, aes(x = Group, y = logCPM, fill = Group)) +
      geom_boxplot(
        width = 0.55,
        outlier.shape = NA,
        color = "black",
        linewidth = 0.5
      ) +
      geom_jitter(
        width = 0.10,
        size = 1.5,
        shape = 21,
        stroke = 0.4,
        color = "black"
      ) +
      geom_segment(
        aes(
          x = 1,
          xend = 2,
          y = y_max_gene + 0.10 * y_range_gene,
          yend = y_max_gene + 0.10 * y_range_gene
        ),
        inherit.aes = FALSE
      ) +
      annotate(
        "text",
        x = 1.5,
        y = y_max_gene + 0.16 * y_range_gene,
        label = p_gene_label,
        size = 5,
        fontface = "bold"
      ) +
      scale_fill_manual(values = group_cols) +
      coord_cartesian(
        ylim = c(
          y_min_gene - 0.05 * y_range_gene,
          y_max_gene + 0.22 * y_range_gene
        )
      ) +
      labs(title = gene, x = NULL, y = "logCPM") +
      theme_classic(base_size = 12) +
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.position = "none"
      )

    print(p)
  }
}
```

## Cell-cycle program scores 

```{r cell cycle gene sets}
s_phase_genes <- Seurat::cc.genes.updated.2019$s.genes
g2m_phase_genes <- Seurat::cc.genes.updated.2019$g2m.genes

s_phase_genes_present <- intersect(s_phase_genes, rownames(logcpm_scores))
g2m_phase_genes_present <- intersect(g2m_phase_genes, rownames(logcpm_scores))

s_phase_genes_missing <- setdiff(s_phase_genes, rownames(logcpm_scores))
g2m_phase_genes_missing <- setdiff(g2m_phase_genes, rownames(logcpm_scores))

length(s_phase_genes_present)
length(g2m_phase_genes_present)
```

## Build score matrix for cell-cycle programs

```{r cell cycle score matrix}
cell_cycle_genes_all <- unique(c(
  s_phase_genes_present,
  g2m_phase_genes_present
  
))

logcpm_subset_cc <- logcpm_scores[
  rownames(logcpm_scores) %in% cell_cycle_genes_all,
  ,
  drop = FALSE
]

logcpm_subset_cc_z <- t(scale(t(logcpm_subset_cc)))
logcpm_subset_cc_z[is.na(logcpm_subset_cc_z)] <- 0

s_phase_score <- colMeans(
  logcpm_subset_cc_z[rownames(logcpm_subset_cc_z) %in% s_phase_genes_present, , drop = FALSE]
)

g2m_phase_score <- colMeans(
  logcpm_subset_cc_z[rownames(logcpm_subset_cc_z) %in% g2m_phase_genes_present, , drop = FALSE]
)

score_df_cc <- metadata_scores %>%
  dplyr::mutate(
    S_phase_score = s_phase_score[Sample],
    G2M_phase_score = g2m_phase_score[Sample]
  )

write.csv(
  score_df_cc,
  file = file.path(output_dir_data, "EC_cell_cycle_program_scores.csv"),
  quote = FALSE,
  row.names = FALSE
)
```

## Cell-cycle program score boxplots

```{r s phase score, fig.width=2.8, fig.height=3.7}
fit_s_score <- lm(S_phase_score ~ Group + Age + Sex + Dataset, data = score_df_cc)
fit_s_tidy <- broom::tidy(fit_s_score)
print(fit_s_tidy)

p_s <- fit_s_tidy %>%
  dplyr::filter(term == "GroupHF") %>%
  dplyr::pull(p.value)

p_s_label <- format_p_label(p_s)

y_max_s <- max(score_df_cc$S_phase_score, na.rm = TRUE)
y_min_s <- min(score_df_cc$S_phase_score, na.rm = TRUE)
y_range_s <- y_max_s - y_min_s
if (!is.finite(y_range_s) || y_range_s == 0) y_range_s <- 1

ggplot(score_df_cc, aes(x = Group, y = S_phase_score, fill = Group)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, color = "black", linewidth = 0.5) +
  geom_jitter(width = 0.10, size = 1.5, shape = 21, stroke = 0.4, color = "black") +
  geom_segment(
    aes(x = 1, xend = 2, y = y_max_s + 0.10 * y_range_s, yend = y_max_s + 0.10 * y_range_s),
    inherit.aes = FALSE
  ) +
  annotate("text", x = 1.5, y = y_max_s + 0.16 * y_range_s, label = p_s_label, size = 5, fontface = "bold") +
  scale_fill_manual(values = group_cols) +
  coord_cartesian(ylim = c(y_min_s - 0.05 * y_range_s, y_max_s + 0.22 * y_range_s)) +
  labs(title = "S-phase genes", x = NULL, y = "Score") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "none"
  )
```

```{r g2m phase score, fig.width=2.8, fig.height=3.7}
fit_g2m_score <- lm(G2M_phase_score ~ Group + Age + Sex + Dataset, data = score_df_cc)
fit_g2m_tidy <- broom::tidy(fit_g2m_score)
print(fit_g2m_tidy)

p_g2m <- fit_g2m_tidy %>%
  dplyr::filter(term == "GroupHF") %>%
  dplyr::pull(p.value)

p_g2m_label <- format_p_label(p_g2m)

y_max_g2m <- max(score_df_cc$G2M_phase_score, na.rm = TRUE)
y_min_g2m <- min(score_df_cc$G2M_phase_score, na.rm = TRUE)
y_range_g2m <- y_max_g2m - y_min_g2m
if (!is.finite(y_range_g2m) || y_range_g2m == 0) y_range_g2m <- 1

ggplot(score_df_cc, aes(x = Group, y = G2M_phase_score, fill = Group)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, color = "black", linewidth = 0.5) +
  geom_jitter(width = 0.10, size = 1.5, shape = 21, stroke = 0.4, color = "black") +
  geom_segment(
    aes(x = 1, xend = 2, y = y_max_g2m + 0.10 * y_range_g2m, yend = y_max_g2m + 0.10 * y_range_g2m),
    inherit.aes = FALSE
  ) +
  annotate("text", x = 1.5, y = y_max_g2m + 0.16 * y_range_g2m, label = p_g2m_label, size = 5, fontface = "bold") +
  scale_fill_manual(values = group_cols) +
  coord_cartesian(ylim = c(y_min_g2m - 0.05 * y_range_g2m, y_max_g2m + 0.22 * y_range_g2m)) +
  labs(title = "G2/M-phase genes", x = NULL, y = "Score") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "none"
  )
```

## Representative genes for each program

```{r representative genes for cell cycle programs}
s_phase_genes_plot <- c("MCM2", "MCM5", "PCNA", "TYMS", "RRM2", "TK1")
g2m_phase_genes_plot <- c("TOP2A", "UBE2C", "CENPF", "BIRC5", "CDC20", "AURKB")
```

```{r barplots s phase genes, fig.width=2.66, fig.height=3.7}
for (gene in s_phase_genes_plot[s_phase_genes_plot %in% unique(gene_barplot_df$gene)]) {

  df_sub <- gene_barplot_df[gene_barplot_df$gene == gene, ]

  fit_gene <- lm(logCPM ~ Group + Age + Sex + Dataset, data = df_sub)
  fit_gene_tidy <- broom::tidy(fit_gene)
  print(fit_gene_tidy)

  p_gene <- fit_gene_tidy %>%
    dplyr::filter(term == "GroupHF") %>%
    dplyr::pull(p.value)

  p_gene_label <- format_p_label(p_gene)

  y_max_gene <- max(df_sub$logCPM, na.rm = TRUE)
  y_min_gene <- min(df_sub$logCPM, na.rm = TRUE)
  y_range_gene <- y_max_gene - y_min_gene
  if (!is.finite(y_range_gene) || y_range_gene == 0) y_range_gene <- 1

  p <- ggplot(df_sub, aes(x = Group, y = logCPM, fill = Group)) +
    geom_boxplot(width = 0.55, outlier.shape = NA, color = "black", linewidth = 0.5) +
    geom_jitter(width = 0.10, size = 1.5, shape = 21, stroke = 0.4, color = "black") +
    geom_segment(aes(x = 1, xend = 2, y = y_max_gene + 0.10 * y_range_gene, yend = y_max_gene + 0.10 * y_range_gene), inherit.aes = FALSE) +
    annotate("text", x = 1.5, y = y_max_gene + 0.16 * y_range_gene, label = p_gene_label, size = 5, fontface = "bold") +
    scale_fill_manual(values = group_cols) +
    coord_cartesian(ylim = c(y_min_gene - 0.05 * y_range_gene, y_max_gene + 0.22 * y_range_gene)) +
    labs(title = gene, x = NULL, y = "logCPM") +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.title.y = element_text(face = "bold"),
      legend.position = "none"
    )

  print(p)
}
```

```{r barplots g2m phase genes, fig.width=2.66, fig.height=3.7}
for (gene in g2m_phase_genes_plot[g2m_phase_genes_plot %in% unique(gene_barplot_df$gene)]) {

  df_sub <- gene_barplot_df[gene_barplot_df$gene == gene, ]

  fit_gene <- lm(logCPM ~ Group + Age + Sex + Dataset, data = df_sub)
  fit_gene_tidy <- broom::tidy(fit_gene)
  print(fit_gene_tidy)

  p_gene <- fit_gene_tidy %>%
    dplyr::filter(term == "GroupHF") %>%
    dplyr::pull(p.value)

  p_gene_label <- format_p_label(p_gene)

  y_max_gene <- max(df_sub$logCPM, na.rm = TRUE)
  y_min_gene <- min(df_sub$logCPM, na.rm = TRUE)
  y_range_gene <- y_max_gene - y_min_gene
  if (!is.finite(y_range_gene) || y_range_gene == 0) y_range_gene <- 1

  p <- ggplot(df_sub, aes(x = Group, y = logCPM, fill = Group)) +
    geom_boxplot(width = 0.55, outlier.shape = NA, color = "black", linewidth = 0.5) +
    geom_jitter(width = 0.10, size = 1.5, shape = 21, stroke = 0.4, color = "black") +
    geom_segment(aes(x = 1, xend = 2, y = y_max_gene + 0.10 * y_range_gene, yend = y_max_gene + 0.10 * y_range_gene), inherit.aes = FALSE) +
    annotate("text", x = 1.5, y = y_max_gene + 0.16 * y_range_gene, label = p_gene_label, size = 5, fontface = "bold") +
    scale_fill_manual(values = group_cols) +
    coord_cartesian(ylim = c(y_min_gene - 0.05 * y_range_gene, y_max_gene + 0.22 * y_range_gene)) +
    labs(title = gene, x = NULL, y = "logCPM") +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.title.y = element_text(face = "bold"),
      legend.position = "none"
    )

  print(p)
}
```
