---
title: "LV_Klaudia"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "LV_Klaudia"

# Load libraries
library(here)
library(readr)
library(readxl)
library(xlsx)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(ArchR)
library(clustree)
library(harmony)
library(gprofiler2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationHub)
library(ReactomePA)
library(statmod)
library(edgeR)
library(speckle)
library(EnhancedVolcano)
library(decoupleR)
library(SeuratDisk)
library(dorothea)
library(enrichplot)
library(png)
library(reactable)
library(UpSetR)
library(ComplexHeatmap)
library(rtracklayer)
library(cowplot)
library(nichenetr)
library(multinichenetr)
library(muscat)


#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```

## Koenig

```{r Import Koenig}
load("/scratch/gbonaz/Heart_datasets/Koenig_2022/GSE183852_DCM_Nuclei.Robj")
Koenig <- nuclei
rm(nuclei)
DefaultAssay(Koenig) <- "RNA"
Koenig <- NormalizeData(Koenig)
```

```{r add_metadata Koenig}
Clinical_data <- read_excel("/scratch/gbonaz/Heart_datasets/Koenig_2022/Koenig_et_al_2022_Selected_clinical_info.xlsx")
add_metadata <- left_join(Koenig[["orig.ident"]], Clinical_data)
row.names(add_metadata) <- row.names(Koenig[[]])
Koenig <- AddMetaData(Koenig, metadata = add_metadata)
rm(add_metadata, Clinical_data)

Koenig@meta.data <- Koenig@meta.data %>%
  mutate(Group = case_when(
    Condition == "Donor" ~ "HC",
    Condition == "DCM" ~ "DCM"
  ))

Koenig$Sample <- make.names(Koenig$orig.ident)
```

## Chaffin

```{r Import Chaffin}
#library(zellkonverter)

#sce <- readH5AD("/scratch/gbonaz/Heart_datasets/Chaffin_2022/human_dcm_hcm_scportal_03.17.2022.h5ad")

#Chaffin <- as.Seurat(sce, counts = "X", data = NULL)
#DefaultAssay(Chaffin) <- "originalexp"

#saveRDS(Chaffin, "/scratch/gbonaz/Heart_datasets/Chaffin_2022/Chaffin.rds")
Chaffin <- readRDS("/scratch/gbonaz/Heart_datasets/Chaffin_2022/Chaffin.rds")
```

```{r add_metadata Chaffin}
Clinical_data <- read_excel("/scratch/gbonaz/Heart_datasets/Chaffin_2022/Chaffin_et al_2022_Selected_clinical_info.xlsx")
add_metadata <- left_join(Chaffin[["donor_id"]], Clinical_data)
row.names(add_metadata) <- row.names(Chaffin[[]])
Chaffin <- AddMetaData(Chaffin, metadata = add_metadata)
rm(add_metadata, Clinical_data)



Chaffin$donor_id <- make.names(Chaffin$donor_id)
```

```{r subset Chaffin}
Chaffin <- subset(
  Chaffin, 
  disease %in% c("DCM", "NF")
)
```

```{r Chaffin Group}
Chaffin@meta.data <- Chaffin@meta.data %>%
  mutate(Group = case_when(
    disease == "NF" ~ "HC",
    disease == "DCM" ~ "DCM"
    ))
```

## Reichart

```{r Import Reichart}
Reichart <- readRDS("/scratch/gbonaz/Heart_datasets/Reichart_2022/DCM_ACM_heart_cell_atlas.rds")
Reichart_gene_names <- data.frame(
  ensembl_id = rownames(Reichart),
  gene_symbol = Reichart@assays[["RNA"]]@meta.features$feature_name
)
rownames(Reichart@assays[["RNA"]]@data) <- Reichart@assays[["RNA"]]@meta.features$feature_name
rownames(Reichart@assays[["RNA"]]@counts) <- Reichart@assays[["RNA"]]@meta.features$feature_name

Idents(Reichart) <- Reichart$cell_type
```

```{r add metadata Reichart}
Clinical_data <- read_excel("/scratch/gbonaz/Heart_datasets/Reichart_2022/Reichart_et_al_Science_2022_Selected_clinical_info.xlsx")
add_metadata <- left_join(Reichart[["donor_id"]], Clinical_data)
row.names(add_metadata) <- row.names(Reichart[[]])
Reichart <- AddMetaData(Reichart, metadata = add_metadata)
rm(add_metadata, Clinical_data)
```

```{r subset Reichart}
Reichart <- subset(
  Reichart, 
  disease %in% c("dilated cardiomyopathy", "normal") &
    tissue == "heart left ventricle" &
    cell_type != "unknown"
)

table(Reichart$cell_type, Reichart$disease)
```

```{r Reichart Group}
Reichart@meta.data <- Reichart@meta.data %>%
  mutate(Group = case_when(
    disease == "normal" ~ "HC",
    disease == "dilated cardiomyopathy" ~ "DCM"
  ))
```


## Dotplots

```{r cell_type}
Reichart$cell_type <- factor(Reichart$cell_type, levels = c("cardiac muscle cell",
                                                                    "fibroblast of cardiac tissue",
                                                                    "endothelial cell",
                                                                    "mural cell",
                                                                    "myeloid cell",
                                                                    "lymphocyte",
                                                                    "cardiac neuron",
                                                                    "mast cell",
                                                                    "adipocyte"
                                                                    ))

Koenig$Names <- factor(Koenig$Names, levels = c("Cardiomyocytes",
                                                "Fibroblasts",
                                                "Endothelium",
                                                "Pericytes",
                                                "Smooth_Muscle",
                                                "Macrophages",
                                                "Monocytes",
                                                "Endocardium",
                                                "T/NK_Cells",
                                                "B_Cells",
                                                "Neurons",
                                                "Lymphatic",
                                                "Epicardium",
                                                "Mast_Cells",
                                                "Adipocytes"
                                                ))

Chaffin$cell_type_leiden0.6 <- factor(Chaffin$cell_type_leiden0.6, levels = c("Cardiomyocyte_I",
                                                "Cardiomyocyte_II",
                                                "Cardiomyocyte_III",
                                                "Fibroblast_I",
                                                "Fibroblast_II",
                                                "Activated_fibroblast",
                                                "Endothelial_I",
                                                "Endothelial_II",
                                                "Endothelial_III",
                                                "Pericyte_I",
                                                "Pericyte_II",
                                                "VSMC",
                                                "Macrophage",
                                                "Proliferating_macrophage",
                                                "Lymphocyte",
                                                "Endocardial",
                                                "Neuronal",
                                                "Lymphatic_endothelial",
                                                "Epicardial",
                                                "Mast_cell",
                                                "Adipocyte"
                                                ))
```

```{r gene lists}
P1ar <- c("ADORA1", "ADORA2A", "ADORA2B", "ADORA3")
P2Xr <- c("P2RX1", "P2RX2", "P2RX3", "P2RX4", "P2RX5", "P2RX6", "P2RX7")
P2Yr <- c("P2RY1","P2RY2","P2RY4","P2RY6","P2RY11","P2RY12","P2RY13","P2RY14")
Ectoenz <- c("ENTPD1","NT5E","ENPP1","ENPP2","ENPP3","CECR1","ADA1","DPP4","ALPP","PNP")
NucleoTrans <- c("SLC29A1","SLC29A2","SLC29A3","SLC29A4",
  "SLC28A1","SLC28A2","SLC28A3","ABCC4","ABCC5")
SecMess <- c("GNAI1","GNAQ","PLCB","CAMK2","PRKC","CREB1","MAPK1","PIK3R1")
IntrEnz <- c("AMPD1","XDH","PRKAA1","PRKAA2")
Activation <- c("ACTA2","TAGLN","POSTN","MYH11","CNN1","COL1A1",
  "COL3A1","SPARC","THY1","FAP","PDPN","LOX")
RegGenes <- c("TGFB1","TGFB2","TGFB3","TGFBR1","TGFBR2",
  "SMAD2","SMAD3","SMAD4","EDN1","CTGF",
  "PDGFB","PDGFRA","PDGFRB","IL11","YAP1",
  "TEAD1","TEAD4","STAT3","STAT1","SNAI1",
  "TWIST1","ZEB1","ZEB2","FOSL1","JUNB",
  "SOX9","TGFB1I1","SERPINE1")
EndoEMT <- c("PECAM1","CDH5","VWF","CLDN5","TJP1","ACTA2",
  "TAGLN","FN1","VIM","COL1A1","CDH2",
  "SNAI1","TWIST1","ZEB1","ZEB2")
SignPathways <- c("TGFB1","TGFB2","TGFB3","TGFBR1","TGFBR2",
  "SMAD2","SMAD3","SMAD4",
  # BMP family (main ligands)
  "BMP2","BMP4","BMP5","BMP6","BMP7","BMP8A","BMP8B","BMP9","BMP10",
  # Notch receptors
  "NOTCH1","NOTCH2","NOTCH3","NOTCH4",
  # WNT ligands (canonical set)
  "WNT1","WNT2","WNT2B","WNT3","WNT3A","WNT4","WNT5A","WNT5B","WNT6",
  "WNT7A","WNT7B","WNT8A","WNT8B","WNT9A","WNT9B","WNT10A","WNT10B","WNT11","WNT16",
  "VEGFA","HIF1A","IL1B","TNF","JAK1",
  "STAT1","STAT3","RELA","PIK3CA","AKT1",
  "MAPK1","YAP1","TEAD1","TEAD4")
Senescence <- c("CDKN2A","CDKN1A","TP53","MDM2","GLB1",
  "SERPINE1","IL6","CXCL8","MMP1","MMP3")

```

```{r title_grob}
title_grob <- function(txt){
  ggplot() + labs(title = txt) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 18))
}
```

```{r Dotplot_P1ar, fig.height=10, fig.width=8}
p1 <- DotPlot_scCustom(
  seurat_object = Reichart,
  features = rev(P1ar),
  group.by = "cell_type",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p2 <- DotPlot_scCustom(
  seurat_object = Koenig,
  features = rev(P1ar),
  group.by = "Names",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p3 <- DotPlot_scCustom(
  seurat_object = Chaffin,
  features = rev(P1ar),
  group.by = "cell_type_leiden0.6",
  assay = "originalexp",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

final_plot <- title_grob("Reichart et al. 2022") / p1 /
              title_grob("Koenig et al. 2022") / p2 /
              title_grob("Chaffin et al. 2022") / p3

final_plot + 
  plot_layout(heights = c(0.07, 1, 0.07, 1, 0.07, 1)) &
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8)
  )
```

```{r Dotplot_P2Xr, fig.height=10, fig.width=8}
p1 <- DotPlot_scCustom(
  seurat_object = Reichart,
  features = rev(P2Xr),
  group.by = "cell_type",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p2 <- DotPlot_scCustom(
  seurat_object = Koenig,
  features = rev(P2Xr),
  group.by = "Names",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p3 <- DotPlot_scCustom(
  seurat_object = Chaffin,
  features = rev(P2Xr),
  group.by = "cell_type_leiden0.6",
  assay = "originalexp",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

final_plot <- title_grob("Reichart et al. 2022") / p1 /
              title_grob("Koenig et al. 2022") / p2 /
              title_grob("Chaffin et al. 2022") / p3

final_plot + 
  plot_layout(heights = c(0.07, 1, 0.07, 1, 0.07, 1)) &
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8)
  )
```

```{r Dotplot_P2Yr, fig.height=10, fig.width=8}
p1 <- DotPlot_scCustom(
  seurat_object = Reichart,
  features = rev(P2Yr),
  group.by = "cell_type",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p2 <- DotPlot_scCustom(
  seurat_object = Koenig,
  features = rev(P2Yr),
  group.by = "Names",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p3 <- DotPlot_scCustom(
  seurat_object = Chaffin,
  features = rev(P2Yr),
  group.by = "cell_type_leiden0.6",
  assay = "originalexp",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

final_plot <- title_grob("Reichart et al. 2022") / p1 /
              title_grob("Koenig et al. 2022") / p2 /
              title_grob("Chaffin et al. 2022") / p3

final_plot + 
  plot_layout(heights = c(0.07, 1, 0.07, 1, 0.07, 1)) &
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8)
  )
```

```{r Dotplot_Ectoenz, fig.height=10, fig.width=8}
p1 <- DotPlot_scCustom(
  seurat_object = Reichart,
  features = rev(Ectoenz),
  group.by = "cell_type",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p2 <- DotPlot_scCustom(
  seurat_object = Koenig,
  features = rev(Ectoenz),
  group.by = "Names",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p3 <- DotPlot_scCustom(
  seurat_object = Chaffin,
  features = rev(Ectoenz),
  group.by = "cell_type_leiden0.6",
  assay = "originalexp",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

final_plot <- title_grob("Reichart et al. 2022") / p1 /
              title_grob("Koenig et al. 2022") / p2 /
              title_grob("Chaffin et al. 2022") / p3

final_plot + 
  plot_layout(heights = c(0.07, 1, 0.07, 1, 0.07, 1)) &
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8)
  )
```

```{r Dotplot_NucleoTrans, fig.height=10, fig.width=8}
p1 <- DotPlot_scCustom(
  seurat_object = Reichart,
  features = rev(NucleoTrans),
  group.by = "cell_type",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p2 <- DotPlot_scCustom(
  seurat_object = Koenig,
  features = rev(NucleoTrans),
  group.by = "Names",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p3 <- DotPlot_scCustom(
  seurat_object = Chaffin,
  features = rev(NucleoTrans),
  group.by = "cell_type_leiden0.6",
  assay = "originalexp",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

final_plot <- title_grob("Reichart et al. 2022") / p1 /
              title_grob("Koenig et al. 2022") / p2 /
              title_grob("Chaffin et al. 2022") / p3

final_plot + 
  plot_layout(heights = c(0.07, 1, 0.07, 1, 0.07, 1)) &
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8)
  )
```

```{r Dotplot_SecMess, fig.height=10, fig.width=8}
p1 <- DotPlot_scCustom(
  seurat_object = Reichart,
  features = rev(SecMess),
  group.by = "cell_type",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p2 <- DotPlot_scCustom(
  seurat_object = Koenig,
  features = rev(SecMess),
  group.by = "Names",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p3 <- DotPlot_scCustom(
  seurat_object = Chaffin,
  features = rev(SecMess),
  group.by = "cell_type_leiden0.6",
  assay = "originalexp",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

final_plot <- title_grob("Reichart et al. 2022") / p1 /
              title_grob("Koenig et al. 2022") / p2 /
              title_grob("Chaffin et al. 2022") / p3

final_plot + 
  plot_layout(heights = c(0.07, 1, 0.07, 1, 0.07, 1)) &
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8)
  )
```

```{r Dotplot_IntrEnz, fig.height=10, fig.width=8}
p1 <- DotPlot_scCustom(
  seurat_object = Reichart,
  features = rev(IntrEnz),
  group.by = "cell_type",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p2 <- DotPlot_scCustom(
  seurat_object = Koenig,
  features = rev(IntrEnz),
  group.by = "Names",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p3 <- DotPlot_scCustom(
  seurat_object = Chaffin,
  features = rev(IntrEnz),
  group.by = "cell_type_leiden0.6",
  assay = "originalexp",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

final_plot <- title_grob("Reichart et al. 2022") / p1 /
              title_grob("Koenig et al. 2022") / p2 /
              title_grob("Chaffin et al. 2022") / p3

final_plot + 
  plot_layout(heights = c(0.07, 1, 0.07, 1, 0.07, 1)) &
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8)
  )
```

```{r Dotplot_Activation, fig.height=12, fig.width=8}
p1 <- DotPlot_scCustom(
  seurat_object = Reichart,
  features = rev(Activation),
  group.by = "cell_type",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p2 <- DotPlot_scCustom(
  seurat_object = Koenig,
  features = rev(Activation),
  group.by = "Names",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p3 <- DotPlot_scCustom(
  seurat_object = Chaffin,
  features = rev(Activation),
  group.by = "cell_type_leiden0.6",
  assay = "originalexp",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

final_plot <- title_grob("Reichart et al. 2022") / p1 /
              title_grob("Koenig et al. 2022") / p2 /
              title_grob("Chaffin et al. 2022") / p3

final_plot + 
  plot_layout(heights = c(0.07, 1, 0.07, 1, 0.07, 1)) &
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8)
  )
```

```{r Dotplot_RegGenes, fig.height=18, fig.width=8}
p1 <- DotPlot_scCustom(
  seurat_object = Reichart,
  features = rev(RegGenes),
  group.by = "cell_type",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p2 <- DotPlot_scCustom(
  seurat_object = Koenig,
  features = rev(RegGenes),
  group.by = "Names",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p3 <- DotPlot_scCustom(
  seurat_object = Chaffin,
  features = rev(RegGenes),
  group.by = "cell_type_leiden0.6",
  assay = "originalexp",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

final_plot <- title_grob("Reichart et al. 2022") / p1 /
              title_grob("Koenig et al. 2022") / p2 /
              title_grob("Chaffin et al. 2022") / p3

final_plot + 
  plot_layout(heights = c(0.07, 1, 0.07, 1, 0.07, 1)) &
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8)
  )
```

```{r Dotplot_EndoEMT, fig.height=13, fig.width=8}
p1 <- DotPlot_scCustom(
  seurat_object = Reichart,
  features = rev(EndoEMT),
  group.by = "cell_type",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p2 <- DotPlot_scCustom(
  seurat_object = Koenig,
  features = rev(EndoEMT),
  group.by = "Names",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p3 <- DotPlot_scCustom(
  seurat_object = Chaffin,
  features = rev(EndoEMT),
  group.by = "cell_type_leiden0.6",
  assay = "originalexp",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

final_plot <- title_grob("Reichart et al. 2022") / p1 /
              title_grob("Koenig et al. 2022") / p2 /
              title_grob("Chaffin et al. 2022") / p3

final_plot + 
  plot_layout(heights = c(0.07, 1, 0.07, 1, 0.07, 1)) &
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8)
  )
```

```{r Dotplot_SignPathways, fig.height=32, fig.width=8}
p1 <- DotPlot_scCustom(
  seurat_object = Reichart,
  features = rev(SignPathways),
  group.by = "cell_type",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p2 <- DotPlot_scCustom(
  seurat_object = Koenig,
  features = rev(SignPathways),
  group.by = "Names",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p3 <- DotPlot_scCustom(
  seurat_object = Chaffin,
  features = rev(SignPathways),
  group.by = "cell_type_leiden0.6",
  assay = "originalexp",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

final_plot <- title_grob("Reichart et al. 2022") / p1 /
              title_grob("Koenig et al. 2022") / p2 /
              title_grob("Chaffin et al. 2022") / p3

final_plot + 
  plot_layout(heights = c(0.07, 1, 0.07, 1, 0.07, 1)) &
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8)
  )
```

```{r Dotplot_Senescence, fig.height=12, fig.width=8}
p1 <- DotPlot_scCustom(
  seurat_object = Reichart,
  features = rev(Senescence),
  group.by = "cell_type",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p2 <- DotPlot_scCustom(
  seurat_object = Koenig,
  features = rev(Senescence),
  group.by = "Names",
  assay = "RNA",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

p3 <- DotPlot_scCustom(
  seurat_object = Chaffin,
  features = rev(Senescence),
  group.by = "cell_type_leiden0.6",
  assay = "originalexp",
  x_lab_rotate = TRUE,
  flip_axes = TRUE
)

final_plot <- title_grob("Reichart et al. 2022") / p1 /
              title_grob("Koenig et al. 2022") / p2 /
              title_grob("Chaffin et al. 2022") / p3

final_plot + 
  plot_layout(heights = c(0.07, 1, 0.07, 1, 0.07, 1)) &
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 8)
  )
```

```{r all_lists}
all_lists <- list(
  P1ar = P1ar,
  P2Xr = P2Xr,
  P2Yr = P2Yr,
  Ectoenz = Ectoenz,
  NucleoTrans = NucleoTrans,
  SecMess = SecMess,
  IntrEnz = IntrEnz,
  Activation = Activation,
  RegGenes = RegGenes,
  EndoEMT = EndoEMT,
  SignPathways = SignPathways,
  Senescence = Senescence
)

genes_df <- bind_rows(lapply(names(all_lists), function(lst) {
  data.frame(
    gene = all_lists[[lst]],
    list = lst,
    stringsAsFactors = FALSE
  )
})) %>%
  distinct(gene, .keep_all = TRUE)

head(genes_df, 20)
```

```{r heatmap_cols}
heatmap_cols <- colorRampPalette(c("#004C8C", "white", "#A63A2A"))(101)

ann_colors <- list(
  Group = c("HC" = "#999999", "DCM" = "#B44D4D"))
```

## Koenig Cardiomyocytes DE DCM vs HC

```{r}
cm_K <- subset(
  Koenig, 
  Names == "Cardiomyocytes"
)

table(cm_K$Names, cm_K$Condition)
```

```{r, fig.width=5, fig.height=4}
DimPlot(cm_K, group.by = c("Group"), reduction = "umap", label = F, shuffle = TRUE, pt.size = 0.1) 
```

```{r}
cm_K_metadata <- cm_K@meta.data %>%
  dplyr::select(orig.ident, Group, Age, Sex, BMI, Pre_Op_EF, Pre_Op_CO_L_min, HTN) %>%
  unique()

rownames(cm_K_metadata) <- cm_K_metadata$orig.ident

reactable(cm_K_metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r}
expr_matrix <- GetAssayData(object = cm_K, assay = "RNA", slot = "counts")

non_zero_cells <- rowSums(expr_matrix > 0)

total_cells <- ncol(expr_matrix)

percent_cells <- (non_zero_cells / total_cells) * 100

percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- cm_K[which(rownames(cm_K) %in% keep_genes),]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```


```{r}
DefaultAssay(data) <- "RNA"

pseudocounts <- Seurat2PB(data, sample="orig.ident", cluster = "Names")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(cm_K_metadata$orig.ident, pseudocounts$samples$sample)]
all(cm_K_metadata$orig.ident == pseudocounts$samples$sample)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```


```{r}
cm_K_metadata$Group <- factor(cm_K_metadata$Group, levels = c("HC", "DCM"))
design<- model.matrix(~ Group, data = cm_K_metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_cm_K <- topTags(fit, n = Inf)$table
results_cm_K <- merge(results_cm_K, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_cm_K$up_down <- ifelse(
  results_cm_K$logFC > 0.5 & results_cm_K$FDR < 0.05, "Increased",
  ifelse(results_cm_K$logFC < -0.5 & results_cm_K$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )

signif_cm_K <- results_cm_K %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)


results_cm_K_goi <- results_cm_K %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene") 

signif_cm_K_goi <- signif_cm_K %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene")
  
write.csv(results_cm_K, file = here::here(output_dir_data, "cm_Koenig_DE_Results.csv"), quote=F, row.names = F)
write.csv(results_cm_K_goi, file = here::here(output_dir_data, "cm_Koenig_DE_Results_goi.csv"), quote=F, row.names = F)
write.csv(signif_cm_K, file = here::here(output_dir_data, "cm_Koenig_DE_Significant.csv"), quote=F, row.names = F)
write.csv(signif_cm_K_goi, file = here::here(output_dir_data, "cm_Koenig_DE_Significant_goi.csv"), quote=F, row.names = F)
```

```{r Volcano_cm_Koenig, fig.height=6, fig.width=4}
keyvals <- rep("grey", nrow(results_cm_K))
names(keyvals) <- rep("Not significant", nrow(results_cm_K))

keyvals[results_cm_K$up_down == "Increased"] <- "#d62728"
names(keyvals)[keyvals == "#d62728"] <- "Increased"
keyvals[results_cm_K$up_down == "Decreased"] <- "#1f77b4"
names(keyvals)[keyvals == "#1f77b4"] <- "Decreased"

volcano <- EnhancedVolcano(results_cm_K,
                lab = results_cm_K$gene,
                x = "logFC",
                y = "FDR",
                labSize = 0,
                legendLabSize = 8.5,
                titleLabSize = 15,
                subtitleLabSize = 14,
                axisLabSize = 12,
                captionLabSize = 9,
                pointSize = 0.5,
                FCcutoff = 0.5,
                pCutoff  = 0.05,
                ylim = c(0, 9),
                colAlpha = 1,
                drawConnectors = FALSE,
                subtitle = NULL,
                title = "LV Cardiomyocytes\nDCM (n=13) vs HC (n=25)",
                colCustom = keyvals
) + theme(legend.position = "bottom")

volcano
```

```{r Heatmap_cm_Koenig, fig.height=12, fig.width=10, warning=FALSE}
metadata_heatmap <- cm_K_metadata %>% 
  dplyr::select(orig.ident, Group) %>%
  arrange(Group)

counts_cm_K <- cpm(pseudocounts, log = TRUE)
write.csv(counts_cm_K, "cm_Koenig_logCPM.csv")

idx <- match(genes_df$gene, rownames(counts_cm_K))   
keep <- !is.na(idx)
missing_genes <- genes_df$gene[!keep]
if (length(missing_genes)) {
  message("Missing genes (not in counts_cm_K): ",
          paste(missing_genes, collapse = ", "))
}

counts_K_heatmap <- counts_cm_K[idx[keep], , drop = FALSE]
counts_K_heatmap <- counts_K_heatmap[, metadata_heatmap$orig.ident, drop = FALSE]

row_annot <- genes_df[keep, c("gene", "list")]
rownames(row_annot) <- row_annot$gene
row_annot$gene <- NULL

ann_col <- metadata_heatmap %>% 
  dplyr::select(Group)

runs <- rle(as.character(row_annot$list))
gaps_row <- cumsum(runs$lengths)
gaps_row <- gaps_row[-length(gaps_row)]

# genes in the same order as the heatmap rows
lab_row <- rownames(counts_K_heatmap)

# stars by FDR (from your signif table)
sig_map <- signif_cm_K_goi %>%
  dplyr::select(gene, FDR) %>%
  dplyr::mutate(stars = dplyr::case_when(
    FDR < 0.001 ~ "***",
    FDR < 0.01  ~ "**",
    FDR < 0.05  ~ "*",
    TRUE ~ ""
  )) %>%
  dplyr::distinct(gene, .keep_all = TRUE)

stars_vec <- sig_map$stars[match(lab_row, sig_map$gene)]
stars_vec[is.na(stars_vec)] <- ""

# arrows by logFC sign (use results for all genes)
lfc_map <- results_cm_K %>% dplyr::select(gene, logFC)
lfc_vec <- lfc_map$logFC[match(lab_row, lfc_map$gene)]
arrow_vec <- ifelse(is.na(lfc_vec), "",
                    ifelse(lfc_vec > 0, "↑",
                           ifelse(lfc_vec < 0, "↓", "")))

#(optional) only show arrows for significant genes:
arrow_vec[stars_vec == ""] <- ""

# combine: GENENAME [space] ARROW + STARS
labels_row <- paste0(lab_row,
                     ifelse(arrow_vec != "", paste0(" ", arrow_vec), ""),
                     stars_vec)

pheatmap::pheatmap(
  counts_K_heatmap,
  color = heatmap_cols,
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,
  annotation_legend = TRUE,
  border_color = NA,
  fontsize = 16,
  fontsize_row = 12,
  fontsize_col = 8,
  cellheight = 11,
  cellwidth = 12,
  gaps_row = gaps_row,
  gaps_col = c(0,0,0,0,0, sum(metadata_heatmap$Group == "HC")),
  main = "Koenig et al. 2022 - Cardiomyocytes",
  angle_col = 45,
  labels_row = labels_row   # ← arrows + stars
)
```


## Koenig Fibroblasts DE DCM vs HC

```{r}
fb_K <- subset(
  Koenig, 
  Names == "Fibroblasts"
)

table(fb_K$Names, fb_K$Condition)
```

```{r, fig.width=5, fig.height=4}
DimPlot(fb_K, group.by = c("Group"), reduction = "umap", label = F, shuffle = TRUE, pt.size = 0.1) 
```

```{r}
fb_K_metadata <- fb_K@meta.data %>%
  dplyr::select(orig.ident, Group, Age, Sex, BMI, Pre_Op_EF, Pre_Op_CO_L_min, HTN) %>%
  unique()

rownames(fb_K_metadata) <- fb_K_metadata$orig.ident

reactable(fb_K_metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r}
expr_matrix <- GetAssayData(object = fb_K, assay = "RNA", slot = "counts")

non_zero_cells <- rowSums(expr_matrix > 0)

total_cells <- ncol(expr_matrix)

percent_cells <- (non_zero_cells / total_cells) * 100

percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- fb_K[which(rownames(fb_K) %in% keep_genes),]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```


```{r}
DefaultAssay(data) <- "RNA"

pseudocounts <- Seurat2PB(data, sample="orig.ident", cluster = "Names")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(fb_K_metadata$orig.ident, pseudocounts$samples$sample)]
all(fb_K_metadata$orig.ident == pseudocounts$samples$sample)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```


```{r}
fb_K_metadata$Group <- factor(fb_K_metadata$Group, levels = c("HC", "DCM"))
design<- model.matrix(~ Group, data = fb_K_metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_fb_K <- topTags(fit, n = Inf)$table
results_fb_K <- merge(results_fb_K, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_fb_K$up_down <- ifelse(
  results_fb_K$logFC > 0.5 & results_fb_K$FDR < 0.05, "Increased",
  ifelse(results_fb_K$logFC < -0.5 & results_fb_K$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )

signif_fb_K <- results_fb_K %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)


results_fb_K_goi <- results_fb_K %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene") 

signif_fb_K_goi <- signif_fb_K %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene")
  
write.csv(results_fb_K, file = here::here(output_dir_data, "fb_Koenig_DE_Results.csv"), quote=F, row.names = F)
write.csv(results_fb_K_goi, file = here::here(output_dir_data, "fb_Koenig_DE_Results_goi.csv"), quote=F, row.names = F)
write.csv(signif_fb_K, file = here::here(output_dir_data, "fb_Koenig_DE_Significant.csv"), quote=F, row.names = F)
write.csv(signif_fb_K_goi, file = here::here(output_dir_data, "fb_Koenig_DE_Significant_goi.csv"), quote=F, row.names = F)
```

```{r Volcano_fb_Koenig, fig.height=6, fig.width=4}
keyvals <- rep("grey", nrow(results_fb_K))
names(keyvals) <- rep("Not significant", nrow(results_fb_K))

keyvals[results_fb_K$up_down == "Increased"] <- "#d62728"
names(keyvals)[keyvals == "#d62728"] <- "Increased"
keyvals[results_fb_K$up_down == "Decreased"] <- "#1f77b4"
names(keyvals)[keyvals == "#1f77b4"] <- "Decreased"

volcano <- EnhancedVolcano(results_fb_K,
                lab = results_fb_K$gene,
                x = "logFC",
                y = "FDR",
                labSize = 0,
                legendLabSize = 8.5,
                titleLabSize = 15,
                subtitleLabSize = 14,
                axisLabSize = 12,
                captionLabSize = 9,
                pointSize = 0.5,
                FCcutoff = 0.5,
                pCutoff  = 0.05,
                ylim = c(0, 9),
                colAlpha = 1,
                drawConnectors = FALSE,
                subtitle = NULL,
                title = "LV Fibroblasts\nDCM (n=13) vs HC (n=25)",
                colCustom = keyvals
) + theme(legend.position = "bottom")

volcano
```

```{r Heatmap_fb_Koenig, fig.height=14, fig.width=10, warning=FALSE}
metadata_heatmap <- fb_K_metadata %>% 
  dplyr::select(orig.ident, Group) %>%
  arrange(Group)

counts_fb_K <- cpm(pseudocounts, log = TRUE)
write.csv(counts_fb_K, "fb_Koenig_logCPM.csv")

idx <- match(genes_df$gene, rownames(counts_fb_K))   
keep <- !is.na(idx)
missing_genes <- genes_df$gene[!keep]
if (length(missing_genes)) {
  message("Missing genes (not in counts_fb_K): ",
          paste(missing_genes, collapse = ", "))
}

counts_K_heatmap <- counts_fb_K[idx[keep], , drop = FALSE]
counts_K_heatmap <- counts_K_heatmap[, metadata_heatmap$orig.ident, drop = FALSE]

row_annot <- genes_df[keep, c("gene", "list")]
rownames(row_annot) <- row_annot$gene
row_annot$gene <- NULL

ann_col <- metadata_heatmap %>% 
  dplyr::select(Group)

runs <- rle(as.character(row_annot$list))
gaps_row <- cumsum(runs$lengths)
gaps_row <- gaps_row[-length(gaps_row)]

# genes in the same order as the heatmap rows
lab_row <- rownames(counts_K_heatmap)

# stars by FDR (from your signif table)
sig_map <- signif_fb_K_goi %>%
  dplyr::select(gene, FDR) %>%
  dplyr::mutate(stars = dplyr::case_when(
    FDR < 0.001 ~ "***",
    FDR < 0.01  ~ "**",
    FDR < 0.05  ~ "*",
    TRUE ~ ""
  )) %>%
  dplyr::distinct(gene, .keep_all = TRUE)

stars_vec <- sig_map$stars[match(lab_row, sig_map$gene)]
stars_vec[is.na(stars_vec)] <- ""

# arrows by logFC sign (use results for all genes)
lfc_map <- results_fb_K %>% dplyr::select(gene, logFC)
lfc_vec <- lfc_map$logFC[match(lab_row, lfc_map$gene)]
arrow_vec <- ifelse(is.na(lfc_vec), "",
                    ifelse(lfc_vec > 0, "↑",
                           ifelse(lfc_vec < 0, "↓", "")))

#(optional) only show arrows for significant genes:
arrow_vec[stars_vec == ""] <- ""

# combine: GENENAME [space] ARROW + STARS
labels_row <- paste0(lab_row,
                     ifelse(arrow_vec != "", paste0(" ", arrow_vec), ""),
                     stars_vec)

pheatmap::pheatmap(
  counts_K_heatmap,
  color = heatmap_cols,
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,
  annotation_legend = TRUE,
  border_color = NA,
  fontsize = 16,
  fontsize_row = 12,
  fontsize_col = 8,
  cellheight = 11,
  cellwidth = 12,
  gaps_row = gaps_row,
  gaps_col = c(0,0,0,0,0, sum(metadata_heatmap$Group == "HC")),
  main = "Koenig et al. 2022 - Fibroblasts",
  angle_col = 45,
  labels_row = labels_row   # ← arrows + stars
)
```


## Koenig Endothelial cells DE DCM vs HC

```{r}
ec_K <- subset(
  Koenig, 
  Names == "Endothelium"
)

table(ec_K$Names, ec_K$Condition)
```

```{r, fig.width=5, fig.height=4}
DimPlot(ec_K, group.by = c("Group"), reduction = "umap", label = F, shuffle = TRUE, pt.size = 0.1) 
```

```{r}
ec_K_metadata <- ec_K@meta.data %>%
  dplyr::select(orig.ident, Group, Age, Sex, BMI, Pre_Op_EF, Pre_Op_CO_L_min, HTN) %>%
  unique()

rownames(ec_K_metadata) <- ec_K_metadata$orig.ident

reactable(ec_K_metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r}
expr_matrix <- GetAssayData(object = ec_K, assay = "RNA", slot = "counts")

non_zero_cells <- rowSums(expr_matrix > 0)

total_cells <- ncol(expr_matrix)

percent_cells <- (non_zero_cells / total_cells) * 100

percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- ec_K[which(rownames(ec_K) %in% keep_genes),]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```


```{r}
DefaultAssay(data) <- "RNA"

pseudocounts <- Seurat2PB(data, sample="orig.ident", cluster = "Names")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(ec_K_metadata$orig.ident, pseudocounts$samples$sample)]
all(ec_K_metadata$orig.ident == pseudocounts$samples$sample)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```


```{r}
ec_K_metadata$Group <- factor(ec_K_metadata$Group, levels = c("HC", "DCM"))
design<- model.matrix(~ Group, data = ec_K_metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_ec_K <- topTags(fit, n = Inf)$table
results_ec_K <- merge(results_ec_K, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_ec_K$up_down <- ifelse(
  results_ec_K$logFC > 0.5 & results_ec_K$FDR < 0.05, "Increased",
  ifelse(results_ec_K$logFC < -0.5 & results_ec_K$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )

signif_ec_K <- results_ec_K %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)


results_ec_K_goi <- results_ec_K %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene") 

signif_ec_K_goi <- signif_ec_K %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene")
  
write.csv(results_ec_K, file = here::here(output_dir_data, "ec_Koenig_DE_Results.csv"), quote=F, row.names = F)
write.csv(results_ec_K_goi, file = here::here(output_dir_data, "ec_Koenig_DE_Results_goi.csv"), quote=F, row.names = F)
write.csv(signif_ec_K, file = here::here(output_dir_data, "ec_Koenig_DE_Significant.csv"), quote=F, row.names = F)
write.csv(signif_ec_K_goi, file = here::here(output_dir_data, "ec_Koenig_DE_Significant_goi.csv"), quote=F, row.names = F)
```

```{r Volcano_ec_Koenig, fig.height=6, fig.width=4}
keyvals <- rep("grey", nrow(results_ec_K))
names(keyvals) <- rep("Not significant", nrow(results_ec_K))

keyvals[results_ec_K$up_down == "Increased"] <- "#d62728"
names(keyvals)[keyvals == "#d62728"] <- "Increased"
keyvals[results_ec_K$up_down == "Decreased"] <- "#1f77b4"
names(keyvals)[keyvals == "#1f77b4"] <- "Decreased"

volcano <- EnhancedVolcano(results_ec_K,
                lab = results_ec_K$gene,
                x = "logFC",
                y = "FDR",
                labSize = 0,
                legendLabSize = 8.5,
                titleLabSize = 15,
                subtitleLabSize = 14,
                axisLabSize = 12,
                captionLabSize = 9,
                pointSize = 0.5,
                FCcutoff = 0.5,
                pCutoff  = 0.05,
                ylim = c(0, 9),
                colAlpha = 1,
                drawConnectors = FALSE,
                subtitle = NULL,
                title = "LV Endothelial cells\nDCM (n=13) vs HC (n=25)",
                colCustom = keyvals
) + theme(legend.position = "bottom")

volcano
```

```{r Heatmap_ec_Koenig, fig.height=13, fig.width=10, warning=FALSE}
metadata_heatmap <- ec_K_metadata %>% 
  dplyr::select(orig.ident, Group) %>%
  arrange(Group)

counts_ec_K <- cpm(pseudocounts, log = TRUE)
write.csv(counts_ec_K, "ec_Koenig_logCPM.csv")

idx <- match(genes_df$gene, rownames(counts_ec_K))   
keep <- !is.na(idx)
missing_genes <- genes_df$gene[!keep]
if (length(missing_genes)) {
  message("Missing genes (not in counts_ec_K): ",
          paste(missing_genes, collapse = ", "))
}

counts_K_heatmap <- counts_ec_K[idx[keep], , drop = FALSE]
counts_K_heatmap <- counts_K_heatmap[, metadata_heatmap$orig.ident, drop = FALSE]

row_annot <- genes_df[keep, c("gene", "list")]
rownames(row_annot) <- row_annot$gene
row_annot$gene <- NULL

ann_col <- metadata_heatmap %>% 
  dplyr::select(Group)

runs <- rle(as.character(row_annot$list))
gaps_row <- cumsum(runs$lengths)
gaps_row <- gaps_row[-length(gaps_row)]

# genes in the same order as the heatmap rows
lab_row <- rownames(counts_K_heatmap)

# stars by FDR (from your signif table)
sig_map <- signif_ec_K_goi %>%
  dplyr::select(gene, FDR) %>%
  dplyr::mutate(stars = dplyr::case_when(
    FDR < 0.001 ~ "***",
    FDR < 0.01  ~ "**",
    FDR < 0.05  ~ "*",
    TRUE ~ ""
  )) %>%
  dplyr::distinct(gene, .keep_all = TRUE)

stars_vec <- sig_map$stars[match(lab_row, sig_map$gene)]
stars_vec[is.na(stars_vec)] <- ""

# arrows by logFC sign (use results for all genes)
lfc_map <- results_ec_K %>% dplyr::select(gene, logFC)
lfc_vec <- lfc_map$logFC[match(lab_row, lfc_map$gene)]
arrow_vec <- ifelse(is.na(lfc_vec), "",
                    ifelse(lfc_vec > 0, "↑",
                           ifelse(lfc_vec < 0, "↓", "")))

#(optional) only show arrows for significant genes:
arrow_vec[stars_vec == ""] <- ""

# combine: GENENAME [space] ARROW + STARS
labels_row <- paste0(lab_row,
                     ifelse(arrow_vec != "", paste0(" ", arrow_vec), ""),
                     stars_vec)

pheatmap::pheatmap(
  counts_K_heatmap,
  color = heatmap_cols,
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,
  annotation_legend = TRUE,
  border_color = NA,
  fontsize = 16,
  fontsize_row = 12,
  fontsize_col = 8,
  cellheight = 11,
  cellwidth = 12,
  gaps_row = gaps_row,
  gaps_col = c(0,0,0,0,0, sum(metadata_heatmap$Group == "HC")),
  main = "Koenig et al. 2022 - Endothelial cells",
  angle_col = 45,
  labels_row = labels_row   # ← arrows + stars
)
```



## Chaffin Cardiomyocytes DE DCM vs HC

```{r}
cm_C <- subset(
  Chaffin, 
  cell_type_leiden0.6 %in% c("Cardiomyocyte_I",
                             "Cardiomyocyte_II",
                             "Cardiomyocyte_III")
)
cm_C$cell_type <- "Cardiomyocytes"

table(cm_C$cell_type, cm_C$Group)
```

```{r, fig.width=5, fig.height=4}
DimPlot(cm_C, group.by = c("Group"), reduction = "X_umap", label = F, shuffle = TRUE, pt.size = 0.1) 
```

```{r}
cm_C_metadata <- cm_C@meta.data %>%
  dplyr::select(donor_id, Group, sex, age, lvef, Body_Mass_Index_kg_m2, Heart_Weight_g, Left_Ventricular_Mass_g, Left_Ventricular_Mass_Index_g_m2, Left_Ventricular_End_Systolic_Dimension_cm, Left_Ventricular_End_Diastolic_Dimension_cm, Posterior_Wall_Thickness_cm) %>%
  unique()

rownames(cm_C_metadata) <- cm_C_metadata$donor_id

reactable(cm_C_metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r}
cm_C <- RenameAssays(cm_C, originalexp = "RNA")

expr_matrix <- GetAssayData(object = cm_C, assay = "RNA", slot = "counts")

non_zero_cells <- rowSums(expr_matrix > 0)

total_cells <- ncol(expr_matrix)

percent_cells <- (non_zero_cells / total_cells) * 100

percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- cm_C[which(rownames(cm_C) %in% keep_genes),]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```


```{r}
DefaultAssay(data) <- "RNA"

pseudocounts <- Seurat2PB(data, sample="donor_id", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(cm_C_metadata$donor_id, pseudocounts$samples$sample)]
all(cm_C_metadata$donor_id == pseudocounts$samples$sample)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```


```{r}
cm_C_metadata$Group <- factor(cm_C_metadata$Group, levels = c("HC", "DCM"))
design<- model.matrix(~ Group, data = cm_C_metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_cm_C <- topTags(fit, n = Inf)$table
results_cm_C <- merge(results_cm_C, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_cm_C$up_down <- ifelse(
  results_cm_C$logFC > 0.5 & results_cm_C$FDR < 0.05, "Increased",
  ifelse(results_cm_C$logFC < -0.5 & results_cm_C$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )

signif_cm_C <- results_cm_C %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)


results_cm_C_goi <- results_cm_C %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene") 

signif_cm_C_goi <- signif_cm_C %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene")
  
write.csv(results_cm_C, file = here::here(output_dir_data, "cm_Chaffin_DE_Results.csv"), quote=F, row.names = F)
write.csv(results_cm_C_goi, file = here::here(output_dir_data, "cm_Chaffin_DE_Results_goi.csv"), quote=F, row.names = F)
write.csv(signif_cm_C, file = here::here(output_dir_data, "cm_Chaffin_DE_Significant.csv"), quote=F, row.names = F)
write.csv(signif_cm_C_goi, file = here::here(output_dir_data, "cm_Chaffin_DE_Significant_goi.csv"), quote=F, row.names = F)
```

```{r Volcano_cm_Chaffin, fig.height=6, fig.width=4}
keyvals <- rep("grey", nrow(results_cm_C))
names(keyvals) <- rep("Not significant", nrow(results_cm_C))

keyvals[results_cm_C$up_down == "Increased"] <- "#d62728"
names(keyvals)[keyvals == "#d62728"] <- "Increased"
keyvals[results_cm_C$up_down == "Decreased"] <- "#1f77b4"
names(keyvals)[keyvals == "#1f77b4"] <- "Decreased"

volcano <- EnhancedVolcano(results_cm_C,
                lab = results_cm_C$gene,
                x = "logFC",
                y = "FDR",
                labSize = 0,
                legendLabSize = 8.5,
                titleLabSize = 15,
                subtitleLabSize = 14,
                axisLabSize = 12,
                captionLabSize = 9,
                pointSize = 0.5,
                FCcutoff = 0.5,
                pCutoff  = 0.05,
                ylim = c(0, 9),
                colAlpha = 1,
                drawConnectors = FALSE,
                subtitle = NULL,
                title = "LV Cardiomyocytes\nDCM (n=11) vs HC (n=16)",
                colCustom = keyvals
) + theme(legend.position = "bottom")

volcano
```

```{r Heatmap_cm_Chaffin, fig.height=12, fig.width=8, warning=FALSE}
metadata_heatmap <- cm_C_metadata %>% 
  dplyr::select(donor_id, Group) %>%
  arrange(Group)

counts_cm_C <- cpm(pseudocounts, log = TRUE)
write.csv(counts_cm_C, "cm_Chaffin_logCPM.csv")

idx <- match(genes_df$gene, rownames(counts_cm_C))   
keep <- !is.na(idx)
missing_genes <- genes_df$gene[!keep]
if (length(missing_genes)) {
  message("Missing genes (not in counts_cm_C): ",
          paste(missing_genes, collapse = ", "))
}

counts_C_heatmap <- counts_cm_C[idx[keep], , drop = FALSE]
counts_C_heatmap <- counts_C_heatmap[, metadata_heatmap$donor_id, drop = FALSE]

row_annot <- genes_df[keep, c("gene", "list")]
rownames(row_annot) <- row_annot$gene
row_annot$gene <- NULL

ann_col <- metadata_heatmap %>% 
  dplyr::select(Group)

runs <- rle(as.character(row_annot$list))
gaps_row <- cumsum(runs$lengths)
gaps_row <- gaps_row[-length(gaps_row)]

# genes in the same order as the heatmap rows
lab_row <- rownames(counts_C_heatmap)

# stars by FDR (from your signif table)
sig_map <- signif_cm_C_goi %>%
  dplyr::select(gene, FDR) %>%
  dplyr::mutate(stars = dplyr::case_when(
    FDR < 0.001 ~ "***",
    FDR < 0.01  ~ "**",
    FDR < 0.05  ~ "*",
    TRUE ~ ""
  )) %>%
  dplyr::distinct(gene, .keep_all = TRUE)

stars_vec <- sig_map$stars[match(lab_row, sig_map$gene)]
stars_vec[is.na(stars_vec)] <- ""

# arrows by logFC sign (use results for all genes)
lfc_map <- results_cm_C %>% dplyr::select(gene, logFC)
lfc_vec <- lfc_map$logFC[match(lab_row, lfc_map$gene)]
arrow_vec <- ifelse(is.na(lfc_vec), "",
                    ifelse(lfc_vec > 0, "↑",
                           ifelse(lfc_vec < 0, "↓", "")))

#(optional) only show arrows for significant genes:
arrow_vec[stars_vec == ""] <- ""

# combine: GENENAME [space] ARROW + STARS
labels_row <- paste0(lab_row,
                     ifelse(arrow_vec != "", paste0(" ", arrow_vec), ""),
                     stars_vec)

pheatmap::pheatmap(
  counts_C_heatmap,
  color = heatmap_cols,
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,
  annotation_legend = TRUE,
  border_color = NA,
  fontsize = 16,
  fontsize_row = 12,
  fontsize_col = 8,
  cellheight = 11,
  cellwidth = 12,
  gaps_row = gaps_row,
  gaps_col = c(sum(metadata_heatmap$Group == "HC")),
  main = "Chaffin et al. 2022 - Cardiomyocytes",
  angle_col = 45,
  labels_row = labels_row   # ← arrows + stars
)
```




## Chaffin Fibroblasts DE DCM vs HC

```{r}
fb_C <- subset(
  Chaffin, 
  cell_type_leiden0.6 %in% c("Fibroblast_I",
                             "Fibroblast_II",
                             "Activated_fibroblast")
)
fb_C$cell_type <- "Fibroblasts"

table(fb_C$cell_type, fb_C$Group)
```

```{r, fig.width=5, fig.height=4}
DimPlot(fb_C, group.by = c("Group"), reduction = "X_umap", label = F, shuffle = TRUE, pt.size = 0.1) 
```

```{r}
fb_C_metadata <- fb_C@meta.data %>%
  dplyr::select(donor_id, Group, sex, age, lvef, Body_Mass_Index_kg_m2, Heart_Weight_g, Left_Ventricular_Mass_g, Left_Ventricular_Mass_Index_g_m2, Left_Ventricular_End_Systolic_Dimension_cm, Left_Ventricular_End_Diastolic_Dimension_cm, Posterior_Wall_Thickness_cm) %>%
  unique()

rownames(fb_C_metadata) <- fb_C_metadata$donor_id

reactable(fb_C_metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r}
fb_C <- RenameAssays(fb_C, originalexp = "RNA")

expr_matrix <- GetAssayData(object = fb_C, assay = "RNA", slot = "counts")

non_zero_cells <- rowSums(expr_matrix > 0)

total_cells <- ncol(expr_matrix)

percent_cells <- (non_zero_cells / total_cells) * 100

percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- fb_C[which(rownames(fb_C) %in% keep_genes),]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```


```{r}
DefaultAssay(data) <- "RNA"

pseudocounts <- Seurat2PB(data, sample="donor_id", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(fb_C_metadata$donor_id, pseudocounts$samples$sample)]
all(fb_C_metadata$donor_id == pseudocounts$samples$sample)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```


```{r}
fb_C_metadata$Group <- factor(fb_C_metadata$Group, levels = c("HC", "DCM"))
design<- model.matrix(~ Group, data = fb_C_metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_fb_C <- topTags(fit, n = Inf)$table
results_fb_C <- merge(results_fb_C, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_fb_C$up_down <- ifelse(
  results_fb_C$logFC > 0.5 & results_fb_C$FDR < 0.05, "Increased",
  ifelse(results_fb_C$logFC < -0.5 & results_fb_C$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )

signif_fb_C <- results_fb_C %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)


results_fb_C_goi <- results_fb_C %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene") 

signif_fb_C_goi <- signif_fb_C %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene")
  
write.csv(results_fb_C, file = here::here(output_dir_data, "fb_Chaffin_DE_Results.csv"), quote=F, row.names = F)
write.csv(results_fb_C_goi, file = here::here(output_dir_data, "fb_Chaffin_DE_Results_goi.csv"), quote=F, row.names = F)
write.csv(signif_fb_C, file = here::here(output_dir_data, "fb_Chaffin_DE_Significant.csv"), quote=F, row.names = F)
write.csv(signif_fb_C_goi, file = here::here(output_dir_data, "fb_Chaffin_DE_Significant_goi.csv"), quote=F, row.names = F)
```

```{r Volcano_fb_Chaffin, fig.height=6, fig.width=4}
keyvals <- rep("grey", nrow(results_fb_C))
names(keyvals) <- rep("Not significant", nrow(results_fb_C))

keyvals[results_fb_C$up_down == "Increased"] <- "#d62728"
names(keyvals)[keyvals == "#d62728"] <- "Increased"
keyvals[results_fb_C$up_down == "Decreased"] <- "#1f77b4"
names(keyvals)[keyvals == "#1f77b4"] <- "Decreased"

volcano <- EnhancedVolcano(results_fb_C,
                lab = results_fb_C$gene,
                x = "logFC",
                y = "FDR",
                labSize = 0,
                legendLabSize = 8.5,
                titleLabSize = 15,
                subtitleLabSize = 14,
                axisLabSize = 12,
                captionLabSize = 9,
                pointSize = 0.5,
                FCcutoff = 0.5,
                pCutoff  = 0.05,
                ylim = c(0, 9),
                colAlpha = 1,
                drawConnectors = FALSE,
                subtitle = NULL,
                title = "LV Fibroblasts\nDCM (n=11) vs HC (n=16)",
                colCustom = keyvals
) + theme(legend.position = "bottom")

volcano
```

```{r Heatmap_fb_Chaffin, fig.height=10, fig.width=8, warning=FALSE}
metadata_heatmap <- fb_C_metadata %>% 
  dplyr::select(donor_id, Group) %>%
  arrange(Group)

counts_fb_C <- cpm(pseudocounts, log = TRUE)
write.csv(counts_fb_C, "fb_Chaffin_logCPM.csv")

idx <- match(genes_df$gene, rownames(counts_fb_C))   
keep <- !is.na(idx)
missing_genes <- genes_df$gene[!keep]
if (length(missing_genes)) {
  message("Missing genes (not in counts_fb_C): ",
          paste(missing_genes, collapse = ", "))
}

counts_C_heatmap <- counts_fb_C[idx[keep], , drop = FALSE]
counts_C_heatmap <- counts_C_heatmap[, metadata_heatmap$donor_id, drop = FALSE]

row_annot <- genes_df[keep, c("gene", "list")]
rownames(row_annot) <- row_annot$gene
row_annot$gene <- NULL

ann_col <- metadata_heatmap %>% 
  dplyr::select(Group)

runs <- rle(as.character(row_annot$list))
gaps_row <- cumsum(runs$lengths)
gaps_row <- gaps_row[-length(gaps_row)]

# genes in the same order as the heatmap rows
lab_row <- rownames(counts_C_heatmap)

# stars by FDR (from your signif table)
sig_map <- signif_fb_C_goi %>%
  dplyr::select(gene, FDR) %>%
  dplyr::mutate(stars = dplyr::case_when(
    FDR < 0.001 ~ "***",
    FDR < 0.01  ~ "**",
    FDR < 0.05  ~ "*",
    TRUE ~ ""
  )) %>%
  dplyr::distinct(gene, .keep_all = TRUE)

stars_vec <- sig_map$stars[match(lab_row, sig_map$gene)]
stars_vec[is.na(stars_vec)] <- ""

# arrows by logFC sign (use results for all genes)
lfc_map <- results_fb_C %>% dplyr::select(gene, logFC)
lfc_vec <- lfc_map$logFC[match(lab_row, lfc_map$gene)]
arrow_vec <- ifelse(is.na(lfc_vec), "",
                    ifelse(lfc_vec > 0, "↑",
                           ifelse(lfc_vec < 0, "↓", "")))

#(optional) only show arrows for significant genes:
arrow_vec[stars_vec == ""] <- ""

# combine: GENENAME [space] ARROW + STARS
labels_row <- paste0(lab_row,
                     ifelse(arrow_vec != "", paste0(" ", arrow_vec), ""),
                     stars_vec)

pheatmap::pheatmap(
  counts_C_heatmap,
  color = heatmap_cols,
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,
  annotation_legend = TRUE,
  border_color = NA,
  fontsize = 16,
  fontsize_row = 12,
  fontsize_col = 8,
  cellheight = 11,
  cellwidth = 12,
  gaps_row = gaps_row,
  gaps_col = c(sum(metadata_heatmap$Group == "HC")),
  main = "Chaffin et al. 2022 - Fibroblasts",
  angle_col = 45,
  labels_row = labels_row   # ← arrows + stars
)
```


## Chaffin Endothelial cells DE DCM vs HC

```{r}
ec_C <- subset(
  Chaffin, 
  cell_type_leiden0.6 %in% c("Endothelial_I",
                             "Endothelial_II",
                             "Endothelial_III")
)
ec_C$cell_type <- "Endothelial cells"

table(ec_C$cell_type, ec_C$Group)
```

```{r, fig.width=5, fig.height=4}
DimPlot(ec_C, group.by = c("Group"), reduction = "X_umap", label = F, shuffle = TRUE, pt.size = 0.1) 
```

```{r}
ec_C_metadata <- ec_C@meta.data %>%
  dplyr::select(donor_id, Group, sex, age, lvef, Body_Mass_Index_kg_m2, Heart_Weight_g, Left_Ventricular_Mass_g, Left_Ventricular_Mass_Index_g_m2, Left_Ventricular_End_Systolic_Dimension_cm, Left_Ventricular_End_Diastolic_Dimension_cm, Posterior_Wall_Thickness_cm) %>%
  unique()

rownames(ec_C_metadata) <- ec_C_metadata$donor_id

reactable(ec_C_metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r}
ec_C <- RenameAssays(ec_C, originalexp = "RNA")

expr_matrix <- GetAssayData(object = ec_C, assay = "RNA", slot = "counts")

non_zero_cells <- rowSums(expr_matrix > 0)

total_cells <- ncol(expr_matrix)

percent_cells <- (non_zero_cells / total_cells) * 100

percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- ec_C[which(rownames(ec_C) %in% keep_genes),]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```


```{r}
DefaultAssay(data) <- "RNA"

pseudocounts <- Seurat2PB(data, sample="donor_id", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(ec_C_metadata$donor_id, pseudocounts$samples$sample)]
all(ec_C_metadata$donor_id == pseudocounts$samples$sample)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```


```{r}
ec_C_metadata$Group <- factor(ec_C_metadata$Group, levels = c("HC", "DCM"))
design<- model.matrix(~ Group, data = ec_C_metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_ec_C <- topTags(fit, n = Inf)$table
results_ec_C <- merge(results_ec_C, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_ec_C$up_down <- ifelse(
  results_ec_C$logFC > 0.5 & results_ec_C$FDR < 0.05, "Increased",
  ifelse(results_ec_C$logFC < -0.5 & results_ec_C$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )

signif_ec_C <- results_ec_C %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)


results_ec_C_goi <- results_ec_C %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene") 

signif_ec_C_goi <- signif_ec_C %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene")
  
write.csv(results_ec_C, file = here::here(output_dir_data, "ec_Chaffin_DE_Results.csv"), quote=F, row.names = F)
write.csv(results_ec_C_goi, file = here::here(output_dir_data, "ec_Chaffin_DE_Results_goi.csv"), quote=F, row.names = F)
write.csv(signif_ec_C, file = here::here(output_dir_data, "ec_Chaffin_DE_Significant.csv"), quote=F, row.names = F)
write.csv(signif_ec_C_goi, file = here::here(output_dir_data, "ec_Chaffin_DE_Significant_goi.csv"), quote=F, row.names = F)
```

```{r Volcano_ec_Chaffin, fig.height=6, fig.width=4}
keyvals <- rep("grey", nrow(results_ec_C))
names(keyvals) <- rep("Not significant", nrow(results_ec_C))

keyvals[results_ec_C$up_down == "Increased"] <- "#d62728"
names(keyvals)[keyvals == "#d62728"] <- "Increased"
keyvals[results_ec_C$up_down == "Decreased"] <- "#1f77b4"
names(keyvals)[keyvals == "#1f77b4"] <- "Decreased"

volcano <- EnhancedVolcano(results_ec_C,
                lab = results_ec_C$gene,
                x = "logFC",
                y = "FDR",
                labSize = 0,
                legendLabSize = 8.5,
                titleLabSize = 15,
                subtitleLabSize = 14,
                axisLabSize = 12,
                captionLabSize = 9,
                pointSize = 0.5,
                FCcutoff = 0.5,
                pCutoff  = 0.05,
                ylim = c(0, 9),
                colAlpha = 1,
                drawConnectors = FALSE,
                subtitle = NULL,
                title = "LV Endothelial cells\nDCM (n=11) vs HC (n=16)",
                colCustom = keyvals
) + theme(legend.position = "bottom")

volcano
```

```{r Heatmap_ec_Chaffin, fig.height=10, fig.width=8, warning=FALSE}
metadata_heatmap <- ec_C_metadata %>% 
  dplyr::select(donor_id, Group) %>%
  arrange(Group)

counts_ec_C <- cpm(pseudocounts, log = TRUE)
write.csv(counts_ec_C, "ec_Chaffin_logCPM.csv")

idx <- match(genes_df$gene, rownames(counts_ec_C))   
keep <- !is.na(idx)
missing_genes <- genes_df$gene[!keep]
if (length(missing_genes)) {
  message("Missing genes (not in counts_ec_C): ",
          paste(missing_genes, collapse = ", "))
}

counts_C_heatmap <- counts_ec_C[idx[keep], , drop = FALSE]
counts_C_heatmap <- counts_C_heatmap[, metadata_heatmap$donor_id, drop = FALSE]

row_annot <- genes_df[keep, c("gene", "list")]
rownames(row_annot) <- row_annot$gene
row_annot$gene <- NULL

ann_col <- metadata_heatmap %>% 
  dplyr::select(Group)

runs <- rle(as.character(row_annot$list))
gaps_row <- cumsum(runs$lengths)
gaps_row <- gaps_row[-length(gaps_row)]

# genes in the same order as the heatmap rows
lab_row <- rownames(counts_C_heatmap)

# stars by FDR (from your signif table)
sig_map <- signif_ec_C_goi %>%
  dplyr::select(gene, FDR) %>%
  dplyr::mutate(stars = dplyr::case_when(
    FDR < 0.001 ~ "***",
    FDR < 0.01  ~ "**",
    FDR < 0.05  ~ "*",
    TRUE ~ ""
  )) %>%
  dplyr::distinct(gene, .keep_all = TRUE)

stars_vec <- sig_map$stars[match(lab_row, sig_map$gene)]
stars_vec[is.na(stars_vec)] <- ""

# arrows by logFC sign (use results for all genes)
lfc_map <- results_ec_C %>% dplyr::select(gene, logFC)
lfc_vec <- lfc_map$logFC[match(lab_row, lfc_map$gene)]
arrow_vec <- ifelse(is.na(lfc_vec), "",
                    ifelse(lfc_vec > 0, "↑",
                           ifelse(lfc_vec < 0, "↓", "")))

#(optional) only show arrows for significant genes:
arrow_vec[stars_vec == ""] <- ""

# combine: GENENAME [space] ARROW + STARS
labels_row <- paste0(lab_row,
                     ifelse(arrow_vec != "", paste0(" ", arrow_vec), ""),
                     stars_vec)

pheatmap::pheatmap(
  counts_C_heatmap,
  color = heatmap_cols,
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,
  annotation_legend = TRUE,
  border_color = NA,
  fontsize = 16,
  fontsize_row = 12,
  fontsize_col = 8,
  cellheight = 11,
  cellwidth = 12,
  gaps_row = gaps_row,
  gaps_col = c(sum(metadata_heatmap$Group == "HC")),
  main = "Chaffin et al. 2022 - Endothelial cells",
  angle_col = 45,
  labels_row = labels_row   # ← arrows + stars
)
```




## Reichart Cardiomyocytes DE DCM vs HC

```{r}
cm_R <- subset(
  Reichart, 
  cell_type == "cardiac muscle cell"
)

table(cm_R$cell_type, cm_R$Group)
```

```{r, fig.width=5, fig.height=4}
DimPlot(cm_R, group.by = c("Group"), reduction = "umap", label = F, shuffle = TRUE, pt.size = 0.1, raster=F) 
```

```{r}
cm_R_metadata <- cm_R@meta.data %>%
  dplyr::select(donor_id, Group, Age_range, BMI, LVEF_percent, LVIDd_mm, LVIDs_mm, RVEDD_mm, TAPSE_mm, PAH, Hypertension, COPD) %>%
  unique()

rownames(cm_R_metadata) <- cm_R_metadata$donor_id

reactable(cm_R_metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r}
expr_matrix <- GetAssayData(object = cm_R, assay = "RNA", slot = "counts")

non_zero_cells <- rowSums(expr_matrix > 0)

total_cells <- ncol(expr_matrix)

percent_cells <- (non_zero_cells / total_cells) * 100

percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- cm_R[which(rownames(cm_R) %in% keep_genes),]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```


```{r}
DefaultAssay(data) <- "RNA"

pseudocounts <- Seurat2PB(data, sample="donor_id", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(cm_R_metadata$donor_id, pseudocounts$samples$sample)]
all(cm_R_metadata$donor_id == pseudocounts$samples$sample)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```


```{r}
cm_R_metadata$Group <- factor(cm_R_metadata$Group, levels = c("HC", "DCM"))
design<- model.matrix(~ Group, data = cm_R_metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_cm_R <- topTags(fit, n = Inf)$table
results_cm_R <- merge(results_cm_R, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_cm_R$up_down <- ifelse(
  results_cm_R$logFC > 0.5 & results_cm_R$FDR < 0.05, "Increased",
  ifelse(results_cm_R$logFC < -0.5 & results_cm_R$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )

signif_cm_R <- results_cm_R %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)


results_cm_R_goi <- results_cm_R %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene") 

signif_cm_R_goi <- signif_cm_R %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene")
  
write.csv(results_cm_R, file = here::here(output_dir_data, "cm_Reichart_DE_Results.csv"), quote=F, row.names = F)
write.csv(results_cm_R_goi, file = here::here(output_dir_data, "cm_Reichart_DE_Results_goi.csv"), quote=F, row.names = F)
write.csv(signif_cm_R, file = here::here(output_dir_data, "cm_Reichart_DE_Significant.csv"), quote=F, row.names = F)
write.csv(signif_cm_R_goi, file = here::here(output_dir_data, "cm_Reichart_DE_Significant_goi.csv"), quote=F, row.names = F)
```

```{r Volcano_cm_Reichart, fig.height=6, fig.width=4}
keyvals <- rep("grey", nrow(results_cm_R))
names(keyvals) <- rep("Not significant", nrow(results_cm_R))

keyvals[results_cm_R$up_down == "Increased"] <- "#d62728"
names(keyvals)[keyvals == "#d62728"] <- "Increased"
keyvals[results_cm_R$up_down == "Decreased"] <- "#1f77b4"
names(keyvals)[keyvals == "#1f77b4"] <- "Decreased"

volcano <- EnhancedVolcano(results_cm_R,
                lab = results_cm_R$gene,
                x = "logFC",
                y = "FDR",
                labSize = 0,
                legendLabSize = 8.5,
                titleLabSize = 15,
                subtitleLabSize = 14,
                axisLabSize = 12,
                captionLabSize = 9,
                pointSize = 0.5,
                FCcutoff = 0.5,
                pCutoff  = 0.05,
                ylim = c(0, 9),
                colAlpha = 1,
                drawConnectors = FALSE,
                subtitle = NULL,
                title = "LV Cardiomyocytes\nDCM (n=50) vs HC (n=18)",
                colCustom = keyvals
) + theme(legend.position = "bottom")

volcano
```

```{r Heatmap_cm_Reichart, fig.height=14, fig.width=15, warning=FALSE}
metadata_heatmap <- cm_R_metadata %>% 
  dplyr::select(donor_id, Group) %>%
  arrange(Group)

counts_cm_R <- cpm(pseudocounts, log = TRUE)
write.csv(counts_cm_R, "cm_Reichart_logCPM.csv")

idx <- match(genes_df$gene, rownames(counts_cm_R))   
keep <- !is.na(idx)
missing_genes <- genes_df$gene[!keep]
if (length(missing_genes)) {
  message("Missing genes (not in counts_cm_R): ",
          paste(missing_genes, collapse = ", "))
}

counts_K_heatmap <- counts_cm_R[idx[keep], , drop = FALSE]
counts_K_heatmap <- counts_K_heatmap[, metadata_heatmap$donor_id, drop = FALSE]

row_annot <- genes_df[keep, c("gene", "list")]
rownames(row_annot) <- row_annot$gene
row_annot$gene <- NULL

ann_col <- metadata_heatmap %>% 
  dplyr::select(Group)

runs <- rle(as.character(row_annot$list))
gaps_row <- cumsum(runs$lengths)
gaps_row <- gaps_row[-length(gaps_row)]

# genes in the same order as the heatmap rows
lab_row <- rownames(counts_K_heatmap)

# stars by FDR (from your signif table)
sig_map <- signif_cm_R_goi %>%
  dplyr::select(gene, FDR) %>%
  dplyr::mutate(stars = dplyr::case_when(
    FDR < 0.001 ~ "***",
    FDR < 0.01  ~ "**",
    FDR < 0.05  ~ "*",
    TRUE ~ ""
  )) %>%
  dplyr::distinct(gene, .keep_all = TRUE)

stars_vec <- sig_map$stars[match(lab_row, sig_map$gene)]
stars_vec[is.na(stars_vec)] <- ""

# arrows by logFC sign (use results for all genes)
lfc_map <- results_cm_R %>% dplyr::select(gene, logFC)
lfc_vec <- lfc_map$logFC[match(lab_row, lfc_map$gene)]
arrow_vec <- ifelse(is.na(lfc_vec), "",
                    ifelse(lfc_vec > 0, "↑",
                           ifelse(lfc_vec < 0, "↓", "")))

#(optional) only show arrows for significant genes:
arrow_vec[stars_vec == ""] <- ""

# combine: GENENAME [space] ARROW + STARS
labels_row <- paste0(lab_row,
                     ifelse(arrow_vec != "", paste0(" ", arrow_vec), ""),
                     stars_vec)

pheatmap::pheatmap(
  counts_K_heatmap,
  color = heatmap_cols,
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,
  annotation_legend = TRUE,
  border_color = NA,
  fontsize = 16,
  fontsize_row = 12,
  fontsize_col = 8,
  cellheight = 11,
  cellwidth = 12,
  gaps_row = gaps_row,
  gaps_col = c(0,0,0,0,0, sum(metadata_heatmap$Group == "HC")),
  main = "Reichart et al. 2022 - Cardiomyocytes",
  angle_col = 45,
  labels_row = labels_row   # ← arrows + stars
)
```




## Reichart Fibroblasts DE DCM vs HC

```{r}
fb_R <- subset(
  Reichart, 
  cell_type == "fibroblast of cardiac tissue"
)

table(fb_R$cell_type, fb_R$Group)
```

```{r, fig.width=5, fig.height=4}
DimPlot(fb_R, group.by = c("Group"), reduction = "umap", label = F, shuffle = TRUE, pt.size = 0.1, raster=F) 
```

```{r}
fb_R_metadata <- fb_R@meta.data %>%
  dplyr::select(donor_id, Group, Age_range, BMI, LVEF_percent, LVIDd_mm, LVIDs_mm, RVEDD_mm, TAPSE_mm, PAH, Hypertension, COPD) %>%
  unique()

rownames(fb_R_metadata) <- fb_R_metadata$donor_id

reactable(fb_R_metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r}
expr_matrix <- GetAssayData(object = fb_R, assay = "RNA", slot = "counts")

non_zero_cells <- rowSums(expr_matrix > 0)

total_cells <- ncol(expr_matrix)

percent_cells <- (non_zero_cells / total_cells) * 100

percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- fb_R[which(rownames(fb_R) %in% keep_genes),]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```


```{r}
DefaultAssay(data) <- "RNA"

pseudocounts <- Seurat2PB(data, sample="donor_id", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(fb_R_metadata$donor_id, pseudocounts$samples$sample)]
all(fb_R_metadata$donor_id == pseudocounts$samples$sample)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```


```{r}
fb_R_metadata$Group <- factor(fb_R_metadata$Group, levels = c("HC", "DCM"))
design<- model.matrix(~ Group, data = fb_R_metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_fb_R <- topTags(fit, n = Inf)$table
results_fb_R <- merge(results_fb_R, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_fb_R$up_down <- ifelse(
  results_fb_R$logFC > 0.5 & results_fb_R$FDR < 0.05, "Increased",
  ifelse(results_fb_R$logFC < -0.5 & results_fb_R$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )

signif_fb_R <- results_fb_R %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)


results_fb_R_goi <- results_fb_R %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene") 

signif_fb_R_goi <- signif_fb_R %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene")
  
write.csv(results_fb_R, file = here::here(output_dir_data, "fb_Reichart_DE_Results.csv"), quote=F, row.names = F)
write.csv(results_fb_R_goi, file = here::here(output_dir_data, "fb_Reichart_DE_Results_goi.csv"), quote=F, row.names = F)
write.csv(signif_fb_R, file = here::here(output_dir_data, "fb_Reichart_DE_Significant.csv"), quote=F, row.names = F)
write.csv(signif_fb_R_goi, file = here::here(output_dir_data, "fb_Reichart_DE_Significant_goi.csv"), quote=F, row.names = F)
```

```{r Volcano_fb_Reichart, fig.height=6, fig.width=4}
keyvals <- rep("grey", nrow(results_fb_R))
names(keyvals) <- rep("Not significant", nrow(results_fb_R))

keyvals[results_fb_R$up_down == "Increased"] <- "#d62728"
names(keyvals)[keyvals == "#d62728"] <- "Increased"
keyvals[results_fb_R$up_down == "Decreased"] <- "#1f77b4"
names(keyvals)[keyvals == "#1f77b4"] <- "Decreased"

volcano <- EnhancedVolcano(results_fb_R,
                lab = results_fb_R$gene,
                x = "logFC",
                y = "FDR",
                labSize = 0,
                legendLabSize = 8.5,
                titleLabSize = 15,
                subtitleLabSize = 14,
                axisLabSize = 12,
                captionLabSize = 9,
                pointSize = 0.5,
                FCcutoff = 0.5,
                pCutoff  = 0.05,
                ylim = c(0, 9),
                colAlpha = 1,
                drawConnectors = FALSE,
                subtitle = NULL,
                title = "LV Fibroblasts\nDCM (n=50) vs HC (n=18)",
                colCustom = keyvals
) + theme(legend.position = "bottom")

volcano
```

```{r Heatmap_fb_Reichart, fig.height=13, fig.width=15, warning=FALSE}
metadata_heatmap <- fb_R_metadata %>% 
  dplyr::select(donor_id, Group) %>%
  arrange(Group)

counts_fb_R <- cpm(pseudocounts, log = TRUE)
write.csv(counts_fb_R, "fb_Reichart_logCPM.csv")

idx <- match(genes_df$gene, rownames(counts_fb_R))   
keep <- !is.na(idx)
missing_genes <- genes_df$gene[!keep]
if (length(missing_genes)) {
  message("Missing genes (not in counts_fb_R): ",
          paste(missing_genes, collapse = ", "))
}

counts_K_heatmap <- counts_fb_R[idx[keep], , drop = FALSE]
counts_K_heatmap <- counts_K_heatmap[, metadata_heatmap$donor_id, drop = FALSE]

row_annot <- genes_df[keep, c("gene", "list")]
rownames(row_annot) <- row_annot$gene
row_annot$gene <- NULL

ann_col <- metadata_heatmap %>% 
  dplyr::select(Group)

runs <- rle(as.character(row_annot$list))
gaps_row <- cumsum(runs$lengths)
gaps_row <- gaps_row[-length(gaps_row)]

# genes in the same order as the heatmap rows
lab_row <- rownames(counts_K_heatmap)

# stars by FDR (from your signif table)
sig_map <- signif_fb_R_goi %>%
  dplyr::select(gene, FDR) %>%
  dplyr::mutate(stars = dplyr::case_when(
    FDR < 0.001 ~ "***",
    FDR < 0.01  ~ "**",
    FDR < 0.05  ~ "*",
    TRUE ~ ""
  )) %>%
  dplyr::distinct(gene, .keep_all = TRUE)

stars_vec <- sig_map$stars[match(lab_row, sig_map$gene)]
stars_vec[is.na(stars_vec)] <- ""

# arrows by logFC sign (use results for all genes)
lfc_map <- results_fb_R %>% dplyr::select(gene, logFC)
lfc_vec <- lfc_map$logFC[match(lab_row, lfc_map$gene)]
arrow_vec <- ifelse(is.na(lfc_vec), "",
                    ifelse(lfc_vec > 0, "↑",
                           ifelse(lfc_vec < 0, "↓", "")))

#(optional) only show arrows for significant genes:
arrow_vec[stars_vec == ""] <- ""

# combine: GENENAME [space] ARROW + STARS
labels_row <- paste0(lab_row,
                     ifelse(arrow_vec != "", paste0(" ", arrow_vec), ""),
                     stars_vec)

pheatmap::pheatmap(
  counts_K_heatmap,
  color = heatmap_cols,
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,
  annotation_legend = TRUE,
  border_color = NA,
  fontsize = 16,
  fontsize_row = 12,
  fontsize_col = 8,
  cellheight = 11,
  cellwidth = 12,
  gaps_row = gaps_row,
  gaps_col = c(0,0,0,0,0, sum(metadata_heatmap$Group == "HC")),
  main = "Reichart et al. 2022 - Fibroblasts",
  angle_col = 45,
  labels_row = labels_row   # ← arrows + stars
)
```



## Reichart Endothelial cells DE DCM vs HC

```{r}
ec_R <- subset(
  Reichart, 
  cell_type == "endothelial cell"
)

table(ec_R$cell_type, ec_R$Group)
```

```{r, fig.width=5, fig.height=4}
DimPlot(ec_R, group.by = c("Group"), reduction = "umap", label = F, shuffle = TRUE, pt.size = 0.1, raster=F) 
```

```{r}
ec_R_metadata <- ec_R@meta.data %>%
  dplyr::select(donor_id, Group, Age_range, BMI, LVEF_percent, LVIDd_mm, LVIDs_mm, RVEDD_mm, TAPSE_mm, PAH, Hypertension, COPD) %>%
  unique()

rownames(ec_R_metadata) <- ec_R_metadata$donor_id

reactable(ec_R_metadata, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```

```{r}
expr_matrix <- GetAssayData(object = ec_R, assay = "RNA", slot = "counts")

non_zero_cells <- rowSums(expr_matrix > 0)

total_cells <- ncol(expr_matrix)

percent_cells <- (non_zero_cells / total_cells) * 100

percent_stats <- data.frame(gene = rownames(expr_matrix), percentCells = percent_cells)

rownames(percent_stats) <- percent_stats$gene

keep_genes <- rownames(dplyr::filter(percent_stats, percentCells > 1))
data <- ec_R[which(rownames(ec_R) %in% keep_genes),]
rm(keep_genes)

reactable(percent_stats, 
          filterable = TRUE,
          searchable = TRUE,
          showPageSizeOptions = TRUE)
```


```{r}
DefaultAssay(data) <- "RNA"

pseudocounts <- Seurat2PB(data, sample="donor_id", cluster = "cell_type")
colnames(pseudocounts) <- pseudocounts$samples$sample
pseudocounts <- pseudocounts[, match(ec_R_metadata$donor_id, pseudocounts$samples$sample)]
all(ec_R_metadata$donor_id == pseudocounts$samples$sample)

keep_samples <- pseudocounts$samples$lib.size > 5e4
pseudocounts <- pseudocounts[, keep_samples]
keep_genes <- filterByExpr(pseudocounts)
pseudocounts <- pseudocounts[keep_genes, ]
rm(keep_samples, keep_genes, data)
```


```{r}
ec_R_metadata$Group <- factor(ec_R_metadata$Group, levels = c("HC", "DCM"))
design<- model.matrix(~ Group, data = ec_R_metadata)
pseudocounts <- normLibSizes(pseudocounts)
pseudocounts <- estimateDisp(pseudocounts, design)
  
fit <- glmQLFit(pseudocounts, design, robust=TRUE)
fit <- glmQLFTest(fit, coef = 2)
  
results_ec_R <- topTags(fit, n = Inf)$table
results_ec_R <- merge(results_ec_R, percent_stats[, c("gene", "percentCells")], by = "gene", all.x = FALSE)
results_ec_R$up_down <- ifelse(
  results_ec_R$logFC > 0.5 & results_ec_R$FDR < 0.05, "Increased",
  ifelse(results_ec_R$logFC < -0.5 & results_ec_R$FDR < 0.05, "Decreased",
         "Not significant"
         )
  )

signif_ec_R <- results_ec_R %>% dplyr::filter(FDR < 0.05, abs(logFC) > 0.5) %>% 
    dplyr::arrange(FDR)


results_ec_R_goi <- results_ec_R %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene") 

signif_ec_R_goi <- signif_ec_R %>%
  filter(gene %in% genes_df$gene) %>%
  left_join(genes_df, by = "gene")
  
write.csv(results_ec_R, file = here::here(output_dir_data, "ec_Reichart_DE_Results.csv"), quote=F, row.names = F)
write.csv(results_ec_R_goi, file = here::here(output_dir_data, "ec_Reichart_DE_Results_goi.csv"), quote=F, row.names = F)
write.csv(signif_ec_R, file = here::here(output_dir_data, "ec_Reichart_DE_Significant.csv"), quote=F, row.names = F)
write.csv(signif_ec_R_goi, file = here::here(output_dir_data, "ec_Reichart_DE_Significant_goi.csv"), quote=F, row.names = F)
```

```{r Volcano_ec_Reichart, fig.height=6, fig.width=4}
keyvals <- rep("grey", nrow(results_ec_R))
names(keyvals) <- rep("Not significant", nrow(results_ec_R))

keyvals[results_ec_R$up_down == "Increased"] <- "#d62728"
names(keyvals)[keyvals == "#d62728"] <- "Increased"
keyvals[results_ec_R$up_down == "Decreased"] <- "#1f77b4"
names(keyvals)[keyvals == "#1f77b4"] <- "Decreased"

volcano <- EnhancedVolcano(results_ec_R,
                lab = results_ec_R$gene,
                x = "logFC",
                y = "FDR",
                labSize = 0,
                legendLabSize = 8.5,
                titleLabSize = 15,
                subtitleLabSize = 14,
                axisLabSize = 12,
                captionLabSize = 9,
                pointSize = 0.5,
                FCcutoff = 0.5,
                pCutoff  = 0.05,
                ylim = c(0, 9),
                colAlpha = 1,
                drawConnectors = FALSE,
                subtitle = NULL,
                title = "LV Endothelial cells\nDCM (n=50) vs HC (n=18)",
                colCustom = keyvals
) + theme(legend.position = "bottom")

volcano
```

```{r Heatmap_ec_Reichart, fig.height=11, fig.width=15, warning=FALSE}
metadata_heatmap <- ec_R_metadata %>% 
  dplyr::select(donor_id, Group) %>%
  arrange(Group)

counts_ec_R <- cpm(pseudocounts, log = TRUE)
write.csv(counts_ec_R, "ec_Reichart_logCPM.csv")

idx <- match(genes_df$gene, rownames(counts_ec_R))   
keep <- !is.na(idx)
missing_genes <- genes_df$gene[!keep]
if (length(missing_genes)) {
  message("Missing genes (not in counts_ec_R): ",
          paste(missing_genes, collapse = ", "))
}

counts_K_heatmap <- counts_ec_R[idx[keep], , drop = FALSE]
counts_K_heatmap <- counts_K_heatmap[, metadata_heatmap$donor_id, drop = FALSE]

row_annot <- genes_df[keep, c("gene", "list")]
rownames(row_annot) <- row_annot$gene
row_annot$gene <- NULL

ann_col <- metadata_heatmap %>% 
  dplyr::select(Group)

runs <- rle(as.character(row_annot$list))
gaps_row <- cumsum(runs$lengths)
gaps_row <- gaps_row[-length(gaps_row)]

# genes in the same order as the heatmap rows
lab_row <- rownames(counts_K_heatmap)

# stars by FDR (from your signif table)
sig_map <- signif_ec_R_goi %>%
  dplyr::select(gene, FDR) %>%
  dplyr::mutate(stars = dplyr::case_when(
    FDR < 0.001 ~ "***",
    FDR < 0.01  ~ "**",
    FDR < 0.05  ~ "*",
    TRUE ~ ""
  )) %>%
  dplyr::distinct(gene, .keep_all = TRUE)

stars_vec <- sig_map$stars[match(lab_row, sig_map$gene)]
stars_vec[is.na(stars_vec)] <- ""

# arrows by logFC sign (use results for all genes)
lfc_map <- results_ec_R %>% dplyr::select(gene, logFC)
lfc_vec <- lfc_map$logFC[match(lab_row, lfc_map$gene)]
arrow_vec <- ifelse(is.na(lfc_vec), "",
                    ifelse(lfc_vec > 0, "↑",
                           ifelse(lfc_vec < 0, "↓", "")))

#(optional) only show arrows for significant genes:
arrow_vec[stars_vec == ""] <- ""

# combine: GENENAME [space] ARROW + STARS
labels_row <- paste0(lab_row,
                     ifelse(arrow_vec != "", paste0(" ", arrow_vec), ""),
                     stars_vec)

pheatmap::pheatmap(
  counts_K_heatmap,
  color = heatmap_cols,
  scale = "row",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_colnames = TRUE,
  annotation_legend = TRUE,
  border_color = NA,
  fontsize = 16,
  fontsize_row = 12,
  fontsize_col = 8,
  cellheight = 11,
  cellwidth = 12,
  gaps_row = gaps_row,
  gaps_col = c(0,0,0,0,0, sum(metadata_heatmap$Group == "HC")),
  main = "Reichart et al. 2022 - Endothelial cells",
  angle_col = 45,
  labels_row = labels_row   # ← arrows + stars
)
```


