---
title: "LVAD_OLINK"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
current_file <- "LVAD_OLINK"

library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(readxl)
library(qs2)
library(glmnet)
library(ggpubr)
library(patchwork)
library(pROC)
library(scales)
library(tibble)
library(broom)
library(SummarizedExperiment)

output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"), recursive = TRUE)

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs, recursive = TRUE)

cols_groups <- c("Healthy" = "#004C8C", "LVAD" = "#A63A2A")
cols_tapse <- c("≥0.32" = "#9CA3AF", "<0.32" = "#A63A2A")

# absolute paths
path_ssc <- "/home/gbonaz/data/SSc-PAH"
path_dcm <- "/home/gbonaz/data/DCM_snRNAseq"
path_lvad <- "/home/gbonaz/data/DCM_snRNAseq/data/OLINK_LVAD"

# candidate proteins
panel_tapse_lasso <- c("CLSTN2", "SIGLEC8", "LRRN1", "CXCL17", "IL15")
panel_mpap <- c("C1QTNF1", "LDB3")
panel_tapse_extra <- c("C1QTNF1", "LDB3")
proteins_to_keep <- unique(c(panel_tapse_lasso, panel_mpap, panel_tapse_extra))
```

## Import LVAD Olink NPX data

```{r read_lvad_npx}
npx_raw <- read.delim(
  file.path(path_lvad, "pre_quant.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

LVAD_clinical <- read.csv(
  file.path(path_lvad, "LVAD organ database full.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
```

## Clean LVAD clinical data

```{r clean_lvad_clinical}
LVAD_clinical_clean <- LVAD_clinical %>%
  mutate(across(
    .cols = -c(`LVAD no.`, `Data implantacji`, Etiologia),
    .fns = ~ case_when(
      . == ">60" ~ 60,
      . == "" ~ NA_real_,
      TRUE ~ suppressWarnings(as.numeric(.))
    )
  )) %>%
  dplyr::rename(
    LVAD.no. = `LVAD no.`,
    Age = Age,
    TAPSE_sPAP = `TAPSE/sPAP`,
    mPAP = mPAP
  ) %>%
  dplyr::filter(LVAD.no. != "LV35") %>%
  mutate(
    TAPSE_group = case_when(
      is.na(TAPSE_sPAP) ~ NA_character_,
      TAPSE_sPAP < 0.32 ~ "<0.32",
      TAPSE_sPAP >= 0.32 ~ "≥0.32"
    ),
    TAPSE_group = factor(TAPSE_group, levels = c("≥0.32", "<0.32")),
    Disease_group = "LVAD"
  )

glimpse(LVAD_clinical_clean)
```

## Clean LVAD Olink NPX data and build wide matrix

```{r clean_lvad_npx}
npx_lvad_all <- npx_raw %>%
  filter(SampleType == "SAMPLE") %>%
  filter(AssayType == "assay") %>%
  filter(AssayQC == "PASS", SampleQC == "PASS") %>%
  dplyr::select(SampleID, Group_Name1, Assay, NPX)

npx_lvad_all <- npx_lvad_all %>%
  filter(Assay %in% proteins_to_keep)

npx_matrix_all <- npx_lvad_all %>%
  tidyr::pivot_wider(
    names_from = Assay,
    values_from = NPX
  )

glimpse(npx_matrix_all)
```

## Build LVAD-only validation table

```{r build_lvad_validation_table}
npx_matrix_lvad <- npx_matrix_all %>%
  filter(Group_Name1 == "LVAD") %>%
  dplyr::rename(LVAD.no. = SampleID)

lvad_data_full <- LVAD_clinical_clean %>%
  inner_join(npx_matrix_lvad, by = "LVAD.no.")

dim(lvad_data_full)

lvad_data_full %>%
  dplyr::select(LVAD.no., Age, TAPSE_sPAP, mPAP, all_of(proteins_to_keep)) %>%
  head()
```

## Build all-sample table for LVAD versus healthy plots

```{r build_allsample_table}
all_samples_wide <- npx_matrix_all %>%
  mutate(
    Disease_group = case_when(
      Group_Name1 == "LVAD" ~ "LVAD",
      Group_Name1 == "Healthy" ~ "Healthy",
      TRUE ~ Group_Name1
    )
  )

all_samples_wide$Disease_group <- factor(all_samples_wide$Disease_group, levels = c("Healthy", "LVAD"))

table(all_samples_wide$Disease_group)
```

## Import SSc-PAH Olink cohorts

```{r import_ssc_olink}
data_z <- qs_read(file.path(path_ssc, "data", "o37198_2025-02-27_olink_SummarizedExperiment.qs2"))
clin_z <- read_excel(file.path(path_ssc, "data", "Metadata_Olink_analysis.xlsx"))

meta_z <- colData(data_z) %>% data.frame()
isOutlierColumn <- meta_z$Plate == "P0194" & meta_z$Col %in% c("3", "4")
use <- meta_z$SampleType %in% c("Serum") & meta_z$Group %in% c("SSc-PAH", "SSc-noPAH") & !isOutlierColumn

data_z <- data_z[, use]
meta_z <- meta_z[use, ]
rownames(meta_z) <- gsub("_P0194", "", rownames(meta_z))
meta_z$Barcode <- rownames(meta_z)
rownames(data_z) <- data_z@elementMetadata@listData[["Gene"]]
rownames(data_z@assays@data@listData[["npx"]]) <- data_z@elementMetadata@listData[["Gene"]]
data_z <- assay(data_z, "npx")
data_z <- as.data.frame(data_z[-c(1:3), ])
colnames(data_z) <- rownames(meta_z)

clin_z <- clin_z %>%
  dplyr::filter(Barcode %in% colnames(data_z)) %>%
  mutate(
    Age = as.numeric(Age),
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    Sex = factor(Sex)
  )


data_p <- read.delim(
  file.path(path_ssc, "data", "OLINK_PL", "quant.xls"),
  sep = "\t",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

clin_p <- read_excel(file.path(path_ssc, "data", "OLINK_PL", "Metadata_Olink_PL.xlsx"))

rownames(data_p) <- sub("_.*", "", data_p$SampleID)
data_p <- data_p[, -1]
keep <- intersect(colnames(data_p), clin_p$Barcode)
data_p <- data_p[, keep, drop = FALSE]

clin_p <- clin_p %>%
  dplyr::filter(Barcode %in% colnames(data_p)) %>%
  mutate(
    Age = as.numeric(Age),
    TAPSE_sPAP = as.numeric(TAPSE_sPAP),
    Sex = factor(Sex)
  )
```

## Refit the TAPSE/sPAP LASSO model in the SSc-PAH cohort

```{r fit_tapse_lasso_ssc}
cutoff <- 0.32

train_p <- clin_p %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_p))

test_z <- clin_z %>%
  dplyr::filter(Group == "SSc-PAH") %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(Age), !is.na(Sex)) %>%
  dplyr::filter(Barcode %in% colnames(data_z))

y_tr <- ifelse(as.numeric(train_p$TAPSE_sPAP) < cutoff, 1, 0)
y_te <- ifelse(as.numeric(test_z$TAPSE_sPAP) < cutoff, 1, 0)

Xtr_prot <- t(as.matrix(data_p[panel_tapse_lasso, train_p$Barcode, drop = FALSE]))
Xte_prot <- t(as.matrix(data_z[panel_tapse_lasso, test_z$Barcode, drop = FALSE]))

cov_tr <- data.frame(Age = as.numeric(train_p$Age), Sex = factor(train_p$Sex))
cov_te <- data.frame(Age = as.numeric(test_z$Age), Sex = factor(test_z$Sex, levels = levels(cov_tr$Sex)))

Xtr_cov <- model.matrix(~ Age + Sex, data = cov_tr)[, -1, drop = FALSE]
Xte_cov <- model.matrix(~ Age + Sex, data = cov_te)[, -1, drop = FALSE]

Xtr <- cbind(Xtr_cov, Xtr_prot)
Xte <- cbind(Xte_cov, Xte_prot)

pf <- c(rep(0, ncol(Xtr_cov)), rep(1, ncol(Xtr_prot)))

set.seed(1)
cvfit_tapse <- cv.glmnet(
  x = Xtr,
  y = y_tr,
  family = "binomial",
  alpha = 1,
  penalty.factor = pf,
  standardize = TRUE,
  nfolds = 5,
  type.measure = "auc"
)

lambdas <- cvfit_tapse$glmnet.fit$lambda
beta_mat <- as.matrix(coef(cvfit_tapse$glmnet.fit, s = lambdas))
feat_names <- rownames(beta_mat)
prot_idx <- which(feat_names %in% panel_tapse_lasso)
nprot <- apply(beta_mat[prot_idx, , drop = FALSE] != 0, 2, sum)

ok <- which(nprot >= 3 & nprot <= 10)
if (length(ok) > 0) {
  j <- ok[which.min(cvfit_tapse$cvm[match(lambdas[ok], cvfit_tapse$lambda)])]
  lambda_use <- lambdas[j]
} else {
  j <- which.min(abs(nprot - 5))
  lambda_use <- lambdas[j]
}

coef_tapse_lasso <- as.matrix(coef(cvfit_tapse, s = lambda_use))
coef_tapse_lasso <- data.frame(
  feature = rownames(coef_tapse_lasso),
  coefficient = coef_tapse_lasso[, 1],
  row.names = NULL
)
coef_tapse_lasso

write.csv(
  coef_tapse_lasso,
  file = here::here(output_dir_data, "coefficients_TAPSE_lasso_Poland_train.csv"),
  row.names = FALSE,
  quote = FALSE
)
```

## Check training and test performance in the original SSc-PAH cohorts

```{r check_tapse_lasso_original_cohorts, fig.width=5.2, fig.height=3.8}
train_prob <- as.numeric(predict(cvfit_tapse, newx = Xtr, s = lambda_use, type = "response"))
test_prob <- as.numeric(predict(cvfit_tapse, newx = Xte, s = lambda_use, type = "response"))

roc_tr <- pROC::roc(y_tr, train_prob, quiet = TRUE)
roc_te <- pROC::roc(y_te, test_prob, quiet = TRUE)

auc_tr <- as.numeric(pROC::auc(roc_tr))
auc_te <- as.numeric(pROC::auc(roc_te))

ggplot() +
  geom_line(
    data = data.frame(tpr = rev(roc_tr$sensitivities), fpr = rev(1 - roc_tr$specificities)),
    aes(fpr, tpr),
    color = "red"
  ) +
  geom_line(
    data = data.frame(tpr = rev(roc_te$sensitivities), fpr = rev(1 - roc_te$specificities)),
    aes(fpr, tpr),
    color = "black"
  ) +
  geom_abline(linetype = 2) +
  labs(
    x = "False positive rate",
    y = "True positive rate",
    title = paste0("TAPSE/sPAP lasso\nPoland train AUC = ", round(auc_tr, 2),
                   " | Zurich test AUC = ", round(auc_te, 2))
  ) +
  theme_classic(base_size = 12)
```

## Score the LVAD cohort with the TAPSE/sPAP LASSO model

```{r score_lvad_with_tapse_lasso}
# full score: only available in LVAD samples with Age
cov_lvad <- data.frame(
  Age = as.numeric(lvad_data_full$Age),
  Sex = factor(rep(levels(cov_tr$Sex)[1], nrow(lvad_data_full)), levels = levels(cov_tr$Sex))
)

Xlvad_cov <- model.matrix(~ Age + Sex, data = cov_lvad)[, -1, drop = FALSE]
Xlvad_prot <- as.matrix(lvad_data_full[, panel_tapse_lasso, drop = FALSE])
Xlvad <- cbind(Xlvad_cov, Xlvad_prot)

lvad_data_full$lasso_prob_full <- as.numeric(
  predict(cvfit_tapse, newx = Xlvad, s = lambda_use, type = "response")
)

lvad_data_full$lasso_eta_full <- as.numeric(
  predict(cvfit_tapse, newx = Xlvad, s = lambda_use, type = "link")
)

# protein-only score: can also be used for HC because clinical data are not available there
coef_tbl <- coef_tapse_lasso %>%
  dplyr::filter(feature != "(Intercept)")

intercept_lasso <- coef_tapse_lasso %>%
  dplyr::filter(feature == "(Intercept)") %>%
  dplyr::pull(coefficient)

protein_coef_tbl <- coef_tbl %>%
  dplyr::filter(feature %in% panel_tapse_lasso)

all_samples_score <- all_samples_wide
all_samples_score$lasso_eta_protein_only <- intercept_lasso

for (i in seq_len(nrow(protein_coef_tbl))) {
  this_prot <- protein_coef_tbl$feature[i]
  this_beta <- protein_coef_tbl$coefficient[i]
  all_samples_score$lasso_eta_protein_only <- all_samples_score$lasso_eta_protein_only + all_samples_score[[this_prot]] * this_beta
}

all_samples_score$lasso_prob_protein_only <- plogis(all_samples_score$lasso_eta_protein_only)

lvad_data_full <- lvad_data_full %>%
  left_join(
    all_samples_score %>%
      dplyr::select(SampleID, lasso_eta_protein_only, lasso_prob_protein_only),
    by = c("LVAD.no." = "SampleID")
  )
```

## Correlation of individual TAPSE/sPAP candidate proteins with TAPSE/sPAP in LVAD

```{r corr_tapse_candidates_lvad}
cor_results_tapse <- data.frame()

for (prot in panel_tapse_lasso) {
  df <- lvad_data_full %>%
    dplyr::select(LVAD.no., Age, TAPSE_sPAP, all_of(prot)) %>%
    dplyr::rename(NPX = all_of(prot)) %>%
    dplyr::filter(!is.na(TAPSE_sPAP), !is.na(NPX))

  ct <- cor.test(df$NPX, df$TAPSE_sPAP, method = "spearman", exact = FALSE)
  fit <- lm(NPX ~ TAPSE_sPAP + Age, data = df)
  sm <- summary(fit)

  cor_results_tapse <- rbind(
    cor_results_tapse,
    data.frame(
      protein = prot,
      n = nrow(df),
      spearman_rho = unname(ct$estimate),
      spearman_p = ct$p.value,
      beta_TAPSE_sPAP_adjAge = sm$coefficients["TAPSE_sPAP", "Estimate"],
      p_TAPSE_sPAP_adjAge = sm$coefficients["TAPSE_sPAP", "Pr(>|t|)"],
      beta_Age = sm$coefficients["Age", "Estimate"],
      p_Age = sm$coefficients["Age", "Pr(>|t|)"],
      r_squared = sm$r.squared
    )
  )
}

cor_results_tapse$FDR_spearman <- p.adjust(cor_results_tapse$spearman_p, method = "BH")
cor_results_tapse$FDR_adjAge <- p.adjust(cor_results_tapse$p_TAPSE_sPAP_adjAge, method = "BH")
cor_results_tapse <- cor_results_tapse %>% arrange(FDR_spearman)
cor_results_tapse

write.csv(
  cor_results_tapse,
  file = here::here(output_dir_data, "correlations_TAPSE_candidates_LVAD.csv"),
  row.names = FALSE,
  quote = FALSE
)
```

## Correlation plots for TAPSE/sPAP candidate proteins

```{r corr_tapse_candidates_plots, fig.width=12, fig.height=7}
plot_list_tapse <- list()

for (prot in panel_tapse_lasso) {
  ann <- cor_results_tapse %>% dplyr::filter(protein == prot)

  df <- lvad_data_full %>%
    dplyr::select(LVAD.no., TAPSE_sPAP, all_of(prot)) %>%
    dplyr::rename(NPX = all_of(prot)) %>%
    dplyr::filter(!is.na(TAPSE_sPAP), !is.na(NPX))

  plot_list_tapse[[prot]] <- ggplot(df, aes(x = TAPSE_sPAP, y = NPX)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.5) +
    theme_classic(base_size = 12) +
    labs(
      title = paste0(
        prot,
        "\nρ = ", round(ann$spearman_rho, 2),
        "  p = ", signif(ann$spearman_p, 2)
      ),
      x = "TAPSE/sPAP",
      y = "NPX"
    )
}

wrap_plots(plot_list_tapse, ncol = 3)
```

## TAPSE/sPAP LASSO score versus TAPSE/sPAP in LVAD

```{r corr_tapse_lasso_score_lvad, fig.width=9, fig.height=3.8}
score_results_tapse <- lvad_data_full %>%
  dplyr::select(LVAD.no., Age, TAPSE_sPAP, lasso_eta_full, lasso_prob_full, lasso_eta_protein_only, lasso_prob_protein_only) %>%
  dplyr::filter(!is.na(TAPSE_sPAP), !is.na(lasso_eta_full))

ct_eta <- cor.test(score_results_tapse$lasso_eta_full, score_results_tapse$TAPSE_sPAP, method = "spearman", exact = FALSE)
fit_eta <- lm(lasso_eta_full ~ TAPSE_sPAP + Age, data = score_results_tapse)
summary(fit_eta)

p1 <- ggplot(score_results_tapse, aes(x = TAPSE_sPAP, y = lasso_eta_full)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5) +
  theme_classic(base_size = 12) +
  labs(
    title = paste0("Lasso score (eta)\nρ = ", round(unname(ct_eta$estimate), 2),
                   "  p = ", signif(ct_eta$p.value, 2)),
    x = "TAPSE/sPAP",
    y = "Score"
  )

ct_prob <- cor.test(score_results_tapse$lasso_prob_full, score_results_tapse$TAPSE_sPAP, method = "spearman", exact = FALSE)

p2 <- ggplot(score_results_tapse, aes(x = TAPSE_sPAP, y = lasso_prob_full)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5) +
  theme_classic(base_size = 12) +
  labs(
    title = paste0("Lasso score (probability)\nρ = ", round(unname(ct_prob$estimate), 2),
                   "  p = ", signif(ct_prob$p.value, 2)),
    x = "TAPSE/sPAP",
    y = "Predicted probability"
  )

p1 + p2
```

## Boxplots by TAPSE/sPAP group in LVAD

```{r boxplots_tapse_lvad, fig.width=13, fig.height=8}
box_list_tapse <- list()

for (prot in panel_tapse_lasso) {
  df <- lvad_data_full %>%
    dplyr::select(LVAD.no., TAPSE_group, all_of(prot)) %>%
    dplyr::rename(NPX = all_of(prot)) %>%
    dplyr::filter(!is.na(TAPSE_group), !is.na(NPX))

  box_list_tapse[[prot]] <- ggplot(df, aes(x = TAPSE_group, y = NPX, fill = TAPSE_group)) +
    geom_boxplot(outlier.shape = NA, width = 0.65) +
    geom_jitter(width = 0.12, size = 1.8) +
    stat_compare_means(comparisons = list(c("≥0.32", "<0.32")), method = "wilcox.test", label = "p.format") +
    scale_fill_manual(values = cols_tapse) +
    theme_classic(base_size = 12) +
    theme(legend.position = "none") +
    labs(title = prot, x = "TAPSE/sPAP group", y = "NPX")
}

box_list_tapse[["Lasso_score"]] <- ggplot(
  lvad_data_full %>% dplyr::filter(!is.na(TAPSE_group), !is.na(lasso_eta_full)),
  aes(x = TAPSE_group, y = lasso_eta_full, fill = TAPSE_group)
) +
  geom_boxplot(outlier.shape = NA, width = 0.65) +
  geom_jitter(width = 0.12, size = 1.8) +
  stat_compare_means(comparisons = list(c("≥0.32", "<0.32")), method = "wilcox.test", label = "p.format") +
  scale_fill_manual(values = cols_tapse) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none") +
  labs(title = "Lasso score", x = "TAPSE/sPAP group", y = "Score")

wrap_plots(box_list_tapse, ncol = 3)
```

## HC versus LVAD for TAPSE/sPAP candidate proteins and protein-only score

```{r boxplots_hc_vs_lvad_tapse, fig.width=13, fig.height=8}
box_list_hc_lvad_tapse <- list()

for (prot in panel_tapse_lasso) {
  df <- all_samples_wide %>%
    dplyr::select(SampleID, Disease_group, all_of(prot)) %>%
    dplyr::rename(NPX = all_of(prot)) %>%
    dplyr::filter(Disease_group %in% c("Healthy", "LVAD")) %>%
    dplyr::filter(!is.na(Disease_group), !is.na(NPX))

  box_list_hc_lvad_tapse[[prot]] <- ggplot(df, aes(x = Disease_group, y = NPX, fill = Disease_group)) +
    geom_boxplot(outlier.shape = NA, width = 0.65) +
    geom_jitter(width = 0.12, size = 1.8) +
    stat_compare_means(comparisons = list(c("Healthy", "LVAD")), method = "wilcox.test", label = "p.format") +
    scale_fill_manual(values = cols_groups) +
    theme_classic(base_size = 12) +
    theme(legend.position = "none") +
    labs(title = prot, x = NULL, y = "NPX")
}

box_list_hc_lvad_tapse[["Protein_only_score"]] <- ggplot(
  all_samples_score %>%
    dplyr::filter(Disease_group %in% c("Healthy", "LVAD"), !is.na(lasso_eta_protein_only)),
  aes(x = Disease_group, y = lasso_eta_protein_only, fill = Disease_group)
) +
  geom_boxplot(outlier.shape = NA, width = 0.65) +
  geom_jitter(width = 0.12, size = 1.8) +
  stat_compare_means(comparisons = list(c("Healthy", "LVAD")), method = "wilcox.test", label = "p.format") +
  scale_fill_manual(values = cols_groups) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none") +
  labs(title = "Protein-only lasso score", x = NULL, y = "Score")

wrap_plots(box_list_hc_lvad_tapse, ncol = 3)
```

## Correlation of mPAP biomarkers with mPAP in LVAD

```{r corr_mpap_candidates_lvad}
cor_results_mpap <- data.frame()

for (prot in panel_mpap) {
  df <- lvad_data_full %>%
    dplyr::select(LVAD.no., Age, mPAP, all_of(prot)) %>%
    dplyr::rename(NPX = all_of(prot)) %>%
    dplyr::filter(!is.na(mPAP), !is.na(NPX))

  ct <- cor.test(df$NPX, df$mPAP, method = "spearman", exact = FALSE)
  fit <- lm(NPX ~ mPAP + Age, data = df)
  sm <- summary(fit)

  cor_results_mpap <- rbind(
    cor_results_mpap,
    data.frame(
      protein = prot,
      n = nrow(df),
      spearman_rho = unname(ct$estimate),
      spearman_p = ct$p.value,
      beta_mPAP_adjAge = sm$coefficients["mPAP", "Estimate"],
      p_mPAP_adjAge = sm$coefficients["mPAP", "Pr(>|t|)"],
      beta_Age = sm$coefficients["Age", "Estimate"],
      p_Age = sm$coefficients["Age", "Pr(>|t|)"],
      r_squared = sm$r.squared
    )
  )
}

cor_results_mpap$FDR_spearman <- p.adjust(cor_results_mpap$spearman_p, method = "BH")
cor_results_mpap$FDR_adjAge <- p.adjust(cor_results_mpap$p_mPAP_adjAge, method = "BH")
cor_results_mpap

write.csv(
  cor_results_mpap,
  file = here::here(output_dir_data, "correlations_mPAP_candidates_LVAD.csv"),
  row.names = FALSE,
  quote = FALSE
)
```

## Correlation plots for mPAP biomarkers

```{r corr_mpap_candidates_plots, fig.width=8, fig.height=3.8}
plot_list_mpap <- list()

for (prot in panel_mpap) {
  ann <- cor_results_mpap %>% dplyr::filter(protein == prot)

  df <- lvad_data_full %>%
    dplyr::select(LVAD.no., mPAP, all_of(prot)) %>%
    dplyr::rename(NPX = all_of(prot)) %>%
    dplyr::filter(!is.na(mPAP), !is.na(NPX))

  plot_list_mpap[[prot]] <- ggplot(df, aes(x = mPAP, y = NPX)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.5) +
    theme_classic(base_size = 12) +
    labs(
      title = paste0(
        prot,
        "\nρ = ", round(ann$spearman_rho, 2),
        "  p = ", signif(ann$spearman_p, 2)
      ),
      x = "mPAP",
      y = "NPX"
    )
}

plot_list_mpap[[1]] + plot_list_mpap[[2]]
```

## HC versus LVAD for mPAP biomarkers

```{r boxplots_hc_vs_lvad_mpap, fig.width=7.5, fig.height=3.8}
plot_list_mpap_hc <- list()

for (prot in panel_mpap) {
  df <- all_samples_wide %>%
    dplyr::select(SampleID, Disease_group, all_of(prot)) %>%
    dplyr::rename(NPX = all_of(prot)) %>%
    dplyr::filter(Disease_group %in% c("Healthy", "LVAD")) %>%
    dplyr::filter(!is.na(Disease_group), !is.na(NPX))

  plot_list_mpap_hc[[prot]] <- ggplot(df, aes(x = Disease_group, y = NPX, fill = Disease_group)) +
    geom_boxplot(outlier.shape = NA, width = 0.65) +
    geom_jitter(width = 0.12, size = 1.8) +
    stat_compare_means(comparisons = list(c("Healthy", "LVAD")), method = "wilcox.test", label = "p.format") +
    scale_fill_manual(values = cols_groups) +
    theme_classic(base_size = 12) +
    theme(legend.position = "none") +
    labs(title = prot, x = NULL, y = "NPX")
}

plot_list_mpap_hc[[1]] + plot_list_mpap_hc[[2]]
```

## Save final validation tables

```{r save_final_tables}
write.csv(
  lvad_data_full,
  file = here::here(output_dir_data, "LVAD_validation_table_with_scores.csv"),
  row.names = FALSE,
  quote = FALSE
)

write.csv(
  all_samples_score,
  file = here::here(output_dir_data, "LVAD_and_Healthy_NPX_with_protein_only_score.csv"),
  row.names = FALSE,
  quote = FALSE
)
```

## Correlation of C1QTNF1 and LDB3 with TAPSE/sPAP in LVAD

```{r corr_mpap_markers_with_tapse_lvad}
cor_results_tapse_extra <- data.frame()

for (prot in panel_tapse_extra) {
  df <- lvad_data_full %>%
    dplyr::select(LVAD.no., Age, TAPSE_sPAP, all_of(prot)) %>%
    dplyr::rename(NPX = all_of(prot)) %>%
    dplyr::filter(!is.na(TAPSE_sPAP), !is.na(NPX))

  ct <- cor.test(df$NPX, df$TAPSE_sPAP, method = "spearman", exact = FALSE)
  fit <- lm(NPX ~ TAPSE_sPAP + Age, data = df)
  sm <- summary(fit)

  cor_results_tapse_extra <- rbind(
    cor_results_tapse_extra,
    data.frame(
      protein = prot,
      n = nrow(df),
      spearman_rho = unname(ct$estimate),
      spearman_p = ct$p.value,
      beta_TAPSE_sPAP_adjAge = sm$coefficients["TAPSE_sPAP", "Estimate"],
      p_TAPSE_sPAP_adjAge = sm$coefficients["TAPSE_sPAP", "Pr(>|t|)"],
      beta_Age = sm$coefficients["Age", "Estimate"],
      p_Age = sm$coefficients["Age", "Pr(>|t|)"],
      r_squared = sm$r.squared
    )
  )
}

cor_results_tapse_extra$FDR_spearman <- p.adjust(cor_results_tapse_extra$spearman_p, method = "BH")
cor_results_tapse_extra$FDR_adjAge <- p.adjust(cor_results_tapse_extra$p_TAPSE_sPAP_adjAge, method = "BH")
cor_results_tapse_extra
```

```{r corr_mpap_markers_with_tapse_plots, fig.width=8, fig.height=3.8}
plot_list_tapse_extra <- list()

for (prot in panel_tapse_extra) {
  ann <- cor_results_tapse_extra %>% dplyr::filter(protein == prot)

  df <- lvad_data_full %>%
    dplyr::select(LVAD.no., TAPSE_sPAP, all_of(prot)) %>%
    dplyr::rename(NPX = all_of(prot)) %>%
    dplyr::filter(!is.na(TAPSE_sPAP), !is.na(NPX))

  plot_list_tapse_extra[[prot]] <- ggplot(df, aes(x = TAPSE_sPAP, y = NPX)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.5) +
    theme_classic(base_size = 12) +
    labs(
      title = paste0(
        prot,
        "\nρ = ", round(ann$spearman_rho, 2),
        "  p = ", signif(ann$spearman_p, 2)
      ),
      x = "TAPSE/sPAP",
      y = "NPX"
    )
}

plot_list_tapse_extra[[1]] + plot_list_tapse_extra[[2]]
```


## Boxplots of C1QTNF1 and LDB3 by TAPSE/sPAP group in LVAD

```{r boxplots_mpap_markers_by_tapse_group, fig.width=7.5, fig.height=4}
plot_list_tapse_group_extra <- list()

for (prot in panel_tapse_extra) {
  df <- lvad_data_full %>%
    dplyr::select(LVAD.no., TAPSE_group, all_of(prot)) %>%
    dplyr::rename(NPX = all_of(prot)) %>%
    dplyr::filter(!is.na(TAPSE_group), !is.na(NPX))

  plot_list_tapse_group_extra[[prot]] <- ggplot(df, aes(x = TAPSE_group, y = NPX, fill = TAPSE_group)) +
    geom_boxplot(outlier.shape = NA, width = 0.65) +
    geom_jitter(width = 0.12, size = 1.8) +
    stat_compare_means(
      comparisons = list(c("≥0.32", "<0.32")),
      method = "wilcox.test",
      label = "p.format"
    ) +
    scale_fill_manual(values = cols_tapse) +
    theme_classic(base_size = 12) +
    theme(legend.position = "none") +
    labs(title = prot, x = "TAPSE/sPAP group", y = "NPX")
}

plot_list_tapse_group_extra[[1]] + plot_list_tapse_group_extra[[2]]
```

