---
title: "Proteomics integration and matched multi-omics analysis with MOFA v3"
author: "Gino Bonazza"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  dpi = 300,
  fig.align = "center"
)
```

# Overview

This workflow fits two independent MOFA models.

1. **Integrated proteomics model:** Olink Reveal and LC-MS/MS are integrated in patients measured by both platforms. Factors are tested against continuous clinical parameters and between compensated and decompensated RV states. Factor 1 is interpreted by Gene Ontology gene set enrichment analysis of the complete ranked weights.

2. **Matched three-view model:** pseudobulk snRNA-seq, Olink and LC-MS/MS are integrated in patients measured in all three modalities. The analysis distinguishes modality-specific from shared molecular variation. Factors are tested against RV state, continuous clinical parameters and histological fibrosis. Factor 2 is interpreted separately in the transcriptomic and proteomic layers by Gene Ontology gene set enrichment analysis.

Proteomic matrices and retained male DCM samples are imported from the preceding proteomics quality-control workflow. BCA concentration is removed from the Olink view before model fitting. MOFA training is unsupervised; RV state and clinical variables are used only in downstream association models. Age and sample storage time are included as covariates.

RV states are defined as:

- **Not overloaded:** mPAP <= 20 mmHg;
- **Compensated:** mPAP > 20 mmHg and TAPSE/sPAP > 0.32;
- **Decompensated:** mPAP > 20 mmHg and TAPSE/sPAP <= 0.32.

Patients with missing mPAP receive `NA` for `RV_group`. Patients with mPAP > 20 mmHg and missing TAPSE/sPAP also receive `NA`.

# Setup

```{r libraries}
library(here)
library(qs2)
library(Seurat)
library(SeuratObject)
library(Matrix)
library(edgeR)
library(DESeq2)
library(SummarizedExperiment)
library(readxl)
library(MOFA2)
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(stringr)
library(matrixStats)
library(limma)
library(ggplot2)
library(patchwork)
library(emmeans)
library(broom)
library(reactable)
library(clusterProfiler)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(GO.db)

```

```{r analysis choices}
current_file <- "MOFA_integrative_analysis_v3"

base_output_dir <- here::here("output", current_file)
preprocessing_output_dir <- file.path(base_output_dir, "preprocessed_views")
proteomics_output_dir <- file.path(base_output_dir, "proteomics_model")
matched_output_dir <- file.path(base_output_dir, "matched_three_view_model")

base_figure_dir <- here::here(
  "docs",
  "figure",
  paste0(current_file, ".Rmd")
)
proteomics_figure_dir <- file.path(base_figure_dir, "proteomics_model")
matched_figure_dir <- file.path(base_figure_dir, "matched_three_view_model")

dir.create(preprocessing_output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(proteomics_output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(matched_output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(proteomics_figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(matched_figure_dir, recursive = TRUE, showWarnings = FALSE)

clinical_file <- here::here("data", "DCM_Clinical_data_83.xlsx")
seurat_file <- here::here(
  "output",
  "Add_cell_states_remove_contamination",
  "DCM_RV_cell_states_RNA.qs2"
)
proteomics_qc_file <- here::here(
  "output",
  "Proteomics_quality_control_v3",
  "Proteomics_QC_complete_objects.rds"
)

mPAP_overload_cutoff <- 20
coupling_cutoff <- 0.32

group_colors <- c(
  "Not overloaded" = "#BDBDBD",
  "Compensated" = "#E38C82",
  "Decompensated" = "#9E3434"
)

rna_min_nuclei_fraction <- 0.01
min_rna_library_size <- 5e4
n_rna_features <- 2000
exclude_mitochondrial_and_ribosomal_genes <- TRUE

min_ms_observed_fraction <- 0.50
n_qs_threads <- 16

candidate_factors <- 5:15
tuning_seeds <- c(20260715, 20260716, 20260717)
last_factor_r2_cutoff <- 2
n_consecutive_low_models <- 3

tuning_maxiter <- 5000
final_maxiter <- 10000
mofa_convergence_mode <- "medium"
mofa_freq_elbo <- 5

factor_fdr_cutoff <- 0.05
exploratory_factor_fdr <- 0.10
n_top_weights <- 30
n_factors_for_plots <- 3

clinical_parameters <- c(
  "mPAP_mmHg",
  "sPAP_mmHg",
  "PCW_mmHg",
  "PVR_WU",
  "TAPSE_mm",
  "RVIT_mm",
  "LVEF_percent",
  "LVEDD_mm",
  "TAPSE_sPAP"
)

pretty_parameter_names <- c(
  mPAP_mmHg = "mPAP (mmHg)",
  sPAP_mmHg = "sPAP (mmHg)",
  PCW_mmHg = "PCWP (mmHg)",
  PVR_WU = "PVR (WU)",
  TAPSE_mm = "TAPSE (mm)",
  RVIT_mm = "RVIT (mm)",
  LVEF_percent = "LVEF (%)",
  LVEDD_mm = "LVEDD (mm)",
  TAPSE_sPAP = "TAPSE/sPAP"
)

proteomics_gsea_factor_table <- tibble::tribble(
  ~Factor,   ~orientation, ~interpretation,
  "Factor1", 1,            "Factor1"
)

rv_group_comparison_levels <- c(
  "Compensated",
  "Decompensated"
)

format_fdr_label <- function(x) {
  dplyr::case_when(
    is.na(x) ~ "FDR = NA",
    x < 0.001 ~ "FDR < 0.001",
    TRUE ~ sprintf("FDR = %.3f", x)
  )
}

matched_shared_r2_cutoff <- 1

processed_proteomics_file <- file.path(
  preprocessing_output_dir,
  "processed_Olink_and_LCMS_views.rds"
)
processed_rna_file <- file.path(
  preprocessing_output_dir,
  "processed_RNA_pseudobulk_view.rds"
)

proteomics_input_file <- file.path(
  proteomics_output_dir,
  "MOFA_proteomics_input.rds"
)
proteomics_tuning_summary_file <- file.path(
  proteomics_output_dir,
  "MOFA_proteomics_factor_tuning_all_runs.csv"
)
proteomics_best_tuning_summary_file <- file.path(
  proteomics_output_dir,
  "MOFA_proteomics_factor_tuning_best_seed_per_K.csv"
)
proteomics_selected_settings_file <- file.path(
  proteomics_output_dir,
  "MOFA_proteomics_selected_model_settings.csv"
)
proteomics_final_model_file <- file.path(
  proteomics_output_dir,
  "RV_DCM_proteomics_MOFA_model.hdf5"
)
proteomics_final_model_rds <- file.path(
  proteomics_output_dir,
  "RV_DCM_proteomics_MOFA_model_with_metadata.rds"
)

matched_input_file <- file.path(
  matched_output_dir,
  "MOFA_matched_three_view_input.rds"
)
matched_tuning_summary_file <- file.path(
  matched_output_dir,
  "MOFA_matched_factor_tuning_all_runs.csv"
)
matched_best_tuning_summary_file <- file.path(
  matched_output_dir,
  "MOFA_matched_factor_tuning_best_seed_per_K.csv"
)
matched_selected_settings_file <- file.path(
  matched_output_dir,
  "MOFA_matched_selected_model_settings.csv"
)
matched_final_model_file <- file.path(
  matched_output_dir,
  "RV_DCM_matched_three_view_MOFA_model.hdf5"
)
matched_final_model_rds <- file.path(
  matched_output_dir,
  "RV_DCM_matched_three_view_MOFA_model_with_metadata.rds"
)
```

```{r additional factor and fibrosis analysis choices}
fibrosis_file <- here::here(
  "data",
  "Massons_quantification.xlsx"
)

n_top_factor_loadings_each_direction <- 20

enrichment_simplify_cutoff <- 0.70
enrichment_simplify_padj_cutoff <- 0.10
enrichment_simplify_max_terms <- 300
gsea_min_ranked_genes <- 20

matched_gsea_factor_table <-
  tibble::tribble(
    ~Factor,   ~orientation, ~interpretation,
    "Factor2", 1,            "Factor2"
  )
```

```{r mofa factor helper functions}
mofa_view_colors <- c(
  "RNA" = "#33A02C",
  "OLINK" = "#E38C82",
  "MS" = "#2166AC"
)


standardise_mofa_view_name <- function(view) {
  dplyr::case_when(
    view == "RNA_pseudobulk" ~ "RNA",
    view == "Olink" ~ "OLINK",
    view == "LCMS" ~ "MS",
    TRUE ~ as.character(view)
  )
}


clean_mofa_feature_name <- function(
    feature
) {
  stringr::str_remove(
    as.character(feature),
    "_(RNA_pseudobulk|RNA|Olink|OLINK|LCMS|MS)$"
  )
}


format_p_value <- function(
    p_value,
    prefix = "P"
) {
  dplyr::case_when(
    is.na(p_value) ~
      paste0(prefix, " = NA"),
    p_value < 0.001 ~
      paste0(prefix, " < 0.001"),
    TRUE ~
      sprintf(
        "%s = %.3f",
        prefix,
        p_value
      )
  )
}


prepare_top_factor_loadings <- function(
    weight_table,
    factor_name = "Factor1",
    n_each_direction = 20
) {
  factor_weights <- weight_table %>%
    dplyr::filter(
      Factor == factor_name,
      is.finite(Weight)
    ) %>%
    dplyr::mutate(
      Modality =
        standardise_mofa_view_name(
          View
        ),
      Feature_clean =
        clean_mofa_feature_name(
          Feature
        ),
      Feature_label = paste(
        Modality,
        Feature_clean,
        sep = "_"
      )
    )

  positive_weights <- factor_weights %>%
    dplyr::filter(
      Weight > 0
    ) %>%
    dplyr::slice_max(
      order_by = Weight,
      n = n_each_direction,
      with_ties = FALSE
    )

  negative_weights <- factor_weights %>%
    dplyr::filter(
      Weight < 0
    ) %>%
    dplyr::slice_min(
      order_by = Weight,
      n = n_each_direction,
      with_ties = FALSE
    )

  dplyr::bind_rows(
    negative_weights,
    positive_weights
  ) %>%
    dplyr::arrange(
      Weight
    ) %>%
    dplyr::mutate(
      Feature_label = factor(
        Feature_label,
        levels = unique(
          Feature_label
        )
      )
    )
}


plot_top_factor_loadings <- function(
    loading_data,
    title_text
) {
  ggplot(
    loading_data,
    aes(
      x = Weight,
      y = Feature_label
    )
  ) +
    geom_vline(
      xintercept = 0,
      linewidth = 0.35,
      color = "grey40"
    ) +
    geom_segment(
      aes(
        x = 0,
        xend = Weight,
        yend = Feature_label
      ),
      linewidth = 0.65,
      color = "grey65"
    ) +
    geom_point(
      aes(
        fill = Modality
      ),
      shape = 21,
      size = 3,
      color = "black",
      stroke = 0.3
    ) +
    scale_fill_manual(
      values = mofa_view_colors,
      drop = FALSE
    ) +
    labs(
      title = title_text,
      x = "MOFA feature weight",
      y = NULL,
      fill = "Modality"
    ) +
    theme_classic() +
    theme(
      plot.title = element_text(
        hjust = 0.5,
        face = "bold"
      ),
      legend.position = "bottom"
    )
}


scale_mofa_weights_within_view <- function(
    weight_table,
    factor_table
) {
  weight_table %>%
    dplyr::filter(
      Factor %in%
        factor_table$Factor,
      is.finite(Weight)
    ) %>%
    dplyr::mutate(
      Feature_original =
        as.character(Feature),
      Feature =
        clean_mofa_feature_name(
          Feature
        )
    ) %>%
    dplyr::inner_join(
      factor_table,
      by = "Factor"
    ) %>%
    dplyr::group_by(
      Factor,
      View
    ) %>%
    dplyr::mutate(
      Oriented_weight =
        Weight * orientation,
      maximum_absolute_weight =
        max(
          abs(Oriented_weight),
          na.rm = TRUE
        ),
      Scaled_weight =
        dplyr::if_else(
          maximum_absolute_weight > 0,
          Oriented_weight /
            maximum_absolute_weight,
          0
        ),
      Absolute_scaled_weight =
        abs(Scaled_weight)
    ) %>%
    dplyr::ungroup()
}


prepare_mofa_ranked_gene_list <- function(
    weight_table,
    factor_name,
    orientation
) {
  ranked_symbols <-
    scale_mofa_weights_within_view(
      weight_table =
        weight_table,
      factor_table =
        tibble::tibble(
          Factor = factor_name,
          orientation = orientation,
          interpretation =
            factor_name
        )
    ) %>%
    dplyr::group_by(
      Factor,
      Feature
    ) %>%
    dplyr::summarise(
      Ranked_weight =
        mean(
          Scaled_weight,
          na.rm = TRUE
        ),
      .groups = "drop"
    ) %>%
    dplyr::filter(
      is.finite(
        Ranked_weight
      ),
      Ranked_weight != 0
    )

  symbol_to_entrez <- bitr(
    ranked_symbols$Feature,
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db
  ) %>%
    dplyr::distinct(
      SYMBOL,
      ENTREZID
    ) %>%
    dplyr::mutate(
      ENTREZID =
        as.character(
          ENTREZID
        )
    )

  ranked_entrez <-
    ranked_symbols %>%
    dplyr::inner_join(
      symbol_to_entrez,
      by = c(
        "Feature" =
          "SYMBOL"
      )
    ) %>%
    dplyr::group_by(
      ENTREZID
    ) %>%
    dplyr::slice_max(
      order_by =
        abs(Ranked_weight),
      n = 1,
      with_ties = FALSE
    ) %>%
    dplyr::ungroup() %>%
    dplyr::arrange(
      dplyr::desc(
        Ranked_weight
      )
    )

  gene_list <-
    ranked_entrez$Ranked_weight

  names(gene_list) <-
    ranked_entrez$ENTREZID

  sort(
    gene_list,
    decreasing = TRUE
  )
}


prefix_go_descriptions <- function(
    enrichment_result
) {
  if (
    is.null(enrichment_result) ||
    nrow(
      enrichment_result@result
    ) == 0L ||
    !"ONTOLOGY" %in%
      colnames(
        enrichment_result@result
      )
  ) {
    return(enrichment_result)
  }

  already_prefixed <-
    stringr::str_detect(
      enrichment_result@result$
        Description,
      "^\\[(BP|CC|MF)\\] "
    )

  enrichment_result@result$
    Description[
      !already_prefixed
    ] <- paste0(
      "[",
      enrichment_result@result$
        ONTOLOGY[
          !already_prefixed
        ],
      "] ",
      enrichment_result@result$
        Description[
          !already_prefixed
        ]
    )

  enrichment_result
}


run_mofa_gsea <- function(
    weight_table,
    factor_table
) {
  gsea_objects <- list()
  ranked_lists <- list()

  for (
    factor_index in
    seq_len(
      nrow(factor_table)
    )
  ) {
    factor_name <-
      factor_table$Factor[
        factor_index
      ]

    orientation <-
      factor_table$orientation[
        factor_index
      ]

    gene_list <-
      prepare_mofa_ranked_gene_list(
        weight_table =
          weight_table,
        factor_name =
          factor_name,
        orientation =
          orientation
      )

    ranked_lists[[
      factor_name
    ]] <- gene_list

    if (
      length(gene_list) <
        gsea_min_ranked_genes
    ) {
      next
    }

    go_result <- tryCatch(
      gseGO(
        geneList = gene_list,
        ont = "ALL",
        OrgDb = org.Hs.eg.db,
        keyType = "ENTREZID",
        exponent = 1,
        minGSSize = 5,
        maxGSSize = 500,
        eps = 1e-10,
        pvalueCutoff = 1,
        pAdjustMethod = "BH",
        verbose = FALSE,
        seed = TRUE,
        by = "fgsea"
      ),
      error = function(e) {
        warning(
          "GO GSEA failed for ",
          factor_name,
          ": ",
          conditionMessage(e)
        )
        NULL
      }
    )

    go_result <-
      prefix_go_descriptions(
        go_result
      )

    gsea_objects[[
      paste(
        factor_name,
        "GO_ALL",
        sep = "__"
      )
    ]] <- go_result
  }

  list(
    results =
      extract_gsea_results(
        gsea_objects
      ),
    objects =
      gsea_objects,
    ranked_lists =
      ranked_lists
  )
}


extract_gsea_results <- function(
    object_list
) {
  output <- tibble::tibble()

  for (
    object_name in
    names(object_list)
  ) {
    current_object <-
      object_list[[
        object_name
      ]]

    if (
      is.null(current_object) ||
      nrow(
        as.data.frame(
          current_object
        )
      ) == 0L
    ) {
      next
    }

    name_parts <-
      strsplit(
        object_name,
        "__",
        fixed = TRUE
      )[[1]]

    current_table <-
      as.data.frame(
        current_object
      ) %>%
      tibble::as_tibble() %>%
      dplyr::mutate(
        Factor =
          name_parts[1],
        Direction =
          dplyr::if_else(
            NES >= 0,
            "Positive",
            "Negative"
          ),
        database =
          name_parts[2]
      )

    output <-
      dplyr::bind_rows(
        output,
        current_table
      )
  }

  output
}


subset_enrichment_object <- function(
    enrichment_result,
    selected_ids
) {
  reduced_result <-
    enrichment_result

  reduced_result@result <-
    enrichment_result@result[
      enrichment_result@result$ID %in%
        selected_ids,
      ,
      drop = FALSE
    ]

  if (
    "geneSets" %in%
      methods::slotNames(
        reduced_result
      )
  ) {
    reduced_result@geneSets <-
      enrichment_result@geneSets[
        intersect(
          selected_ids,
          names(
            enrichment_result@geneSets
          )
        )
      ]
  }

  reduced_result
}


simplify_go_all_result <- function(
    enrichment_result,
    cutoff =
      enrichment_simplify_cutoff
) {
  if (
    is.null(enrichment_result) ||
    nrow(
      enrichment_result@result
    ) < 2L
  ) {
    return(enrichment_result)
  }

  result_table <-
    as.data.frame(
      enrichment_result
    ) %>%
    dplyr::filter(
      is.finite(p.adjust)
    ) %>%
    dplyr::arrange(
      p.adjust
    )

  selected_table <-
    result_table %>%
    dplyr::filter(
      p.adjust <
        enrichment_simplify_padj_cutoff
    )

  if (nrow(selected_table) < 2L) {
    selected_table <-
      result_table %>%
      dplyr::slice_head(
        n = min(
          enrichment_simplify_max_terms,
          nrow(result_table)
        )
      )
  }

  if (
    nrow(selected_table) >
      enrichment_simplify_max_terms
  ) {
    selected_table <-
      selected_table %>%
      dplyr::slice_head(
        n =
          enrichment_simplify_max_terms
      )
  }

  selected_object <-
    subset_enrichment_object(
      enrichment_result =
        enrichment_result,
      selected_ids =
        selected_table$ID
    )

  selected_result_table <-
    selected_object@result

  if (
    !"ONTOLOGY" %in%
      colnames(
        selected_result_table
      )
  ) {
    return(
      tryCatch(
        clusterProfiler::simplify(
          selected_object,
          cutoff = cutoff,
          by = "p.adjust",
          select_fun = min,
          measure = "Wang"
        ),
        error = function(e) {
          warning(
            "GO simplification failed: ",
            conditionMessage(e)
          )
          selected_object
        }
      )
    )
  }

  ontology_parts <- list()

  for (
    current_ontology in
    c(
      "BP",
      "CC",
      "MF"
    )
  ) {
    current_ids <-
      selected_result_table$ID[
        selected_result_table$
          ONTOLOGY ==
          current_ontology
      ]

    if (length(current_ids) == 0L) {
      next
    }

    current_object <-
      subset_enrichment_object(
        enrichment_result =
          selected_object,
        selected_ids =
          current_ids
      )

    if (
      "ontology" %in%
        methods::slotNames(
          current_object
        )
    ) {
      current_object@ontology <-
        current_ontology
    }

    if (
      "setType" %in%
        methods::slotNames(
          current_object
        )
    ) {
      current_object@setType <-
        current_ontology
    }

    ontology_parts[[
      current_ontology
    ]] <- tryCatch(
      clusterProfiler::simplify(
        current_object,
        cutoff = cutoff,
        by = "p.adjust",
        select_fun = min,
        measure = "Wang"
      ),
      error = function(e) {
        warning(
          "GO simplification failed for ",
          current_ontology,
          ": ",
          conditionMessage(e)
        )
        current_object
      }
    )
  }

  if (length(ontology_parts) == 0L) {
    return(selected_object)
  }

  combined_object <-
    selected_object

  combined_object@result <-
    do.call(
      rbind,
      lapply(
        ontology_parts,
        function(x) {
          x@result
        }
      )
    )

  if (
    "geneSets" %in%
      methods::slotNames(
        combined_object
      )
  ) {
    combined_object@geneSets <-
      unlist(
        lapply(
          ontology_parts,
          function(x) {
            x@geneSets
          }
        ),
        recursive = FALSE
      )
  }

  if (
    "ontology" %in%
      methods::slotNames(
        combined_object
      )
  ) {
    combined_object@ontology <-
      "ALL"
  }

  if (
    "setType" %in%
      methods::slotNames(
        combined_object
      )
  ) {
    combined_object@setType <-
      "ALL"
  }

  combined_object
}


simplify_enrichment_objects <- function(
    object_list
) {
  lapply(
    names(object_list),
    function(object_name) {
      current_object <-
        object_list[[
          object_name
        ]]

      if (
        stringr::str_ends(
          object_name,
          "GO_ALL"
        )
      ) {
        simplify_go_all_result(
          current_object,
          cutoff =
            enrichment_simplify_cutoff
        )
      } else {
        current_object
      }
    }
  ) %>%
    stats::setNames(
      names(object_list)
    )
}


fit_factor_fibrosis_models <- function(
    factor_scores,
    factor_names
) {
  required_columns <- c(
    "Factor",
    "Factor_value",
    "Fibrosis_percent",
    "Age_years",
    "Sample_age_days"
  )

  missing_columns <- setdiff(
    required_columns,
    colnames(factor_scores)
  )

  if (length(missing_columns) > 0L) {
    stop(
      "Missing fibrosis-analysis columns: ",
      paste(
        missing_columns,
        collapse = ", "
      )
    )
  }

  results <- tibble::tibble()

  for (
    factor_name in
    factor_names
  ) {
    factor_data <-
      factor_scores %>%
      dplyr::filter(
        Factor == factor_name
      ) %>%
      dplyr::select(
        Factor_value,
        Fibrosis_percent,
        Age_years,
        Sample_age_days
      ) %>%
      tidyr::drop_na()

    if (
      nrow(factor_data) < 6L ||
      stats::sd(
        factor_data$Fibrosis_percent
      ) == 0
    ) {
      next
    }

    factor_data <- factor_data %>%
      dplyr::mutate(
        fibrosis_scaled =
          as.numeric(
            scale(
              Fibrosis_percent
            )
          ),
        age_scaled =
          as.numeric(
            scale(Age_years)
          ),
        storage_scaled =
          as.numeric(
            scale(
              Sample_age_days
            )
          )
      )

    fibrosis_fit <- lm(
      Factor_value ~
        fibrosis_scaled +
        age_scaled +
        storage_scaled,
      data = factor_data
    )

    fibrosis_coefficient <-
      broom::tidy(
        fibrosis_fit,
        conf.int = TRUE
      ) %>%
      dplyr::filter(
        term ==
          "fibrosis_scaled"
      )

    results <- dplyr::bind_rows(
      results,
      tibble::tibble(
        Factor = factor_name,
        n = nrow(factor_data),
        beta =
          fibrosis_coefficient$
            estimate,
        standard_error =
          fibrosis_coefficient$
            std.error,
        confidence_interval_low =
          fibrosis_coefficient$
            conf.low,
        confidence_interval_high =
          fibrosis_coefficient$
            conf.high,
        t_statistic =
          fibrosis_coefficient$
            statistic,
        p_value =
          fibrosis_coefficient$
            p.value,
        model_R2 =
          summary(
            fibrosis_fit
          )$r.squared,
        adjusted_model_R2 =
          summary(
            fibrosis_fit
          )$adj.r.squared
      )
    )
  }

  results %>%
    dplyr::mutate(
      FDR = p.adjust(
        p_value,
        method = "BH"
      )
    ) %>%
    dplyr::arrange(FDR)
}


plot_factor_fibrosis_associations <- function(
    fibrosis_results,
    factor_names,
    title_text
) {
  plot_data <- fibrosis_results %>%
    dplyr::mutate(
      Factor = factor(
        Factor,
        levels = rev(
          factor_names
        )
      ),
      label =
        format_fdr_label(FDR)
    )

  ggplot(
    plot_data,
    aes(
      x = beta,
      y = Factor
    )
  ) +
    geom_vline(
      xintercept = 0,
      linetype = 2,
      color = "grey50"
    ) +
    geom_segment(
      aes(
        x =
          confidence_interval_low,
        xend =
          confidence_interval_high,
        yend = Factor
      ),
      linewidth = 0.7,
      color = "grey35"
    ) +
    geom_point(
      aes(
        fill = beta
      ),
      shape = 21,
      size = 4,
      color = "black",
      stroke = 0.35
    ) +
    geom_text(
      aes(
        x = Inf,
        label = label
      ),
      hjust = 1.05,
      size = 3.1
    ) +
    scale_fill_gradient2(
      low = "#2166AC",
      mid = "white",
      high = "#B2182B",
      midpoint = 0
    ) +
    scale_x_continuous(
      expand = expansion(
        mult = c(
          0.08,
          0.35
        )
      )
    ) +
    labs(
      title = title_text,
      x =
        "Adjusted coefficient per 1 SD increase\nin histological fibrosis",
      y = NULL,
      fill = "Coefficient"
    ) +
    theme_classic() +
    theme(
      plot.title = element_text(
        hjust = 0.5,
        face = "bold"
      )
    )
}


```

```{r load filtered proteomics QC object}
if (!file.exists(proteomics_qc_file)) {
  stop(
    "The filtered proteomics QC object was not found: ",
    proteomics_qc_file
  )
}

proteomics_qc <- readRDS(
  proteomics_qc_file
)

filtered_proteomics <-
  proteomics_qc$after_exclusion

retained_dcm_samples <-
  filtered_proteomics$clinical %>%
  dplyr::filter(
    Group == "DCM"
  ) %>%
  dplyr::pull(Sample) %>%
  unique()
```

```{r histological fibrosis data}
fibrosis_data <- readxl::read_excel(
  fibrosis_file
) %>%
  dplyr::select(
    -dplyr::matches(
      "^Unnamed"
    )
  )

if (
  !"Fibrosis_percent" %in%
    colnames(fibrosis_data)
) {
  colnames(fibrosis_data)[3L] <-
    "Fibrosis_percent"
}

fibrosis_data <- fibrosis_data %>%
  dplyr::mutate(
    Sample =
      as.character(Sample),
    Fibrosis_percent =
      as.numeric(
        Fibrosis_percent
      )
  ) %>%
  dplyr::filter(
    Group == "DCM"
  ) %>%
  dplyr::distinct(
    Sample,
    .keep_all = TRUE
  )

write.csv(
  fibrosis_data,
  file.path(
    base_output_dir,
    "histological_fibrosis_data.csv"
  ),
  row.names = FALSE
)
```


# Clinical metadata and RV states

```{r clinical metadata}
clin_all <- readxl::read_excel(clinical_file) %>%
  dplyr::select(-dplyr::matches("^Unnamed")) %>%
  dplyr::mutate(
    Sample = as.character(Sample),
    Collection_date = as.Date(Collection_date),
    Used_date = as.Date(Used_date),
    Sample_age_days = as.numeric(Sample_age_days),
    Age_years = as.numeric(Age_years),
    LVEF_percent = as.numeric(LVEF_percent),
    LVID_mm = as.numeric(LVID_mm),
    LVEDD_mm = as.numeric(LVEDD_mm),
    RVOT_mm = as.numeric(RVOT_mm),
    TAPSE_mm = as.numeric(TAPSE_mm),
    RVIT_mm = as.numeric(RVIT_mm),
    RAA_cm2 = as.numeric(RAA_cm2),
    RVS_cm.s = as.numeric(RVS_cm.s),
    RVSP_mmHg = as.numeric(RVSP_mmHg),
    sPAP_mmHg = as.numeric(sPAP_mmHg),
    mPAP_mmHg = as.numeric(mPAP_mmHg),
    PCW_mmHg = as.numeric(PCW_mmHg),
    PVR_WU = as.numeric(PVR_WU),
    TPG_mmHg = as.numeric(TPG_mmHg),
    CO_l.min = as.numeric(CO_l.min),
    TAPSE_sPAP = TAPSE_mm / sPAP_mmHg
  ) %>%
  dplyr::filter(
    Group == "DCM",
    Sample %in% retained_dcm_samples,
    toupper(
      trimws(
        as.character(Sex)
      )
    ) != "F"
  ) %>%
  dplyr::distinct(Sample, .keep_all = TRUE) %>%
  dplyr::mutate(
    RV_group = dplyr::case_when(
      is.na(mPAP_mmHg) ~ NA_character_,
      mPAP_mmHg <= mPAP_overload_cutoff ~
        "Not overloaded",
      mPAP_mmHg > mPAP_overload_cutoff &
        is.na(TAPSE_sPAP) ~
        NA_character_,
      mPAP_mmHg > mPAP_overload_cutoff &
        TAPSE_sPAP > coupling_cutoff ~
        "Compensated",
      mPAP_mmHg > mPAP_overload_cutoff &
        TAPSE_sPAP <= coupling_cutoff ~
        "Decompensated",
      TRUE ~ NA_character_
    ),
    RV_group = factor(
      RV_group,
      levels = c(
        "Not overloaded",
        "Compensated",
        "Decompensated"
      )
    ),
    Sex = factor(Sex),
    Etiology = factor(Etiology)
  )

clin_all %>%
  dplyr::count(RV_group, .drop = FALSE)
```

```{r clinical group overview, fig.height=3, fig.width=5}
clinical_group_plot_data <- clin_all %>%
  dplyr::filter(
    RV_group %in%
      rv_group_comparison_levels
  ) %>%
  dplyr::mutate(
    RV_group = factor(
      as.character(RV_group),
      levels = rv_group_comparison_levels
    )
  )

p_group_mpap <- clinical_group_plot_data %>%
  dplyr::filter(
    !is.na(mPAP_mmHg)
  ) %>%
  ggplot(
    aes(
      x = RV_group,
      y = mPAP_mmHg,
      fill = RV_group
    )
  ) +
  geom_boxplot(
    outlier.shape = NA,
    width = 0.55,
    color = "grey20"
  ) +
  geom_jitter(
    width = 0.10,
    height = 0,
    size = 2.2,
    shape = 21,
    color = "black",
    stroke = 0.25
  ) +
  scale_fill_manual(
    values = group_colors[
      rv_group_comparison_levels
    ],
    drop = FALSE
  ) +
  labs(
    x = NULL,
    y = "mPAP (mmHg)"
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 20,
      hjust = 1
    )
  )

p_group_coupling <- clinical_group_plot_data %>%
  dplyr::filter(
    !is.na(TAPSE_sPAP)
  ) %>%
  ggplot(
    aes(
      x = RV_group,
      y = TAPSE_sPAP,
      fill = RV_group
    )
  ) +
  geom_hline(
    yintercept = coupling_cutoff,
    linetype = 2
  ) +
  geom_boxplot(
    outlier.shape = NA,
    width = 0.55,
    color = "grey20"
  ) +
  geom_jitter(
    width = 0.10,
    height = 0,
    size = 2.2,
    shape = 21,
    color = "black",
    stroke = 0.25
  ) +
  scale_fill_manual(
    values = group_colors[
      rv_group_comparison_levels
    ],
    drop = FALSE
  ) +
  labs(
    x = NULL,
    y = "TAPSE/sPAP"
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 20,
      hjust = 1
    )
  )

p_group_mpap | p_group_coupling
```

# Preprocess Olink proteomics

The sample-filtered Olink matrix is imported from the proteomics quality-control workflow. All usable Olink proteins are retained. BCA concentration is regressed from the Olink abundance matrix before either MOFA model is fitted.

```{r read olink data}
olink_expr <-
  filtered_proteomics$olink

olink_sample_meta <-
  filtered_proteomics$olink_metadata %>%
  dplyr::mutate(
    Sample = as.character(Sample),
    BCA_mg_ml = as.numeric(BCA_mg_ml),
    Batch = factor(Batch)
  ) %>%
  dplyr::distinct(
    Sample,
    .keep_all = TRUE
  )

olink_samples <- clin_all$Sample[
  clin_all$Sample %in%
    colnames(olink_expr)
]

olink_expr <- olink_expr[
  ,
  olink_samples,
  drop = FALSE
]

olink_meta <- clin_all %>%
  dplyr::inner_join(
    olink_sample_meta %>%
      dplyr::select(
        Sample,
        Batch,
        BCA_mg_ml
      ),
    by = "Sample"
  ) %>%
  dplyr::filter(
    Sample %in% colnames(olink_expr),
    !is.na(BCA_mg_ml)
  ) %>%
  dplyr::slice(
    match(
      colnames(olink_expr),
      Sample
    )
  ) %>%
  dplyr::distinct(
    Sample,
    .keep_all = TRUE
  )

olink_expr <- olink_expr[
  ,
  olink_meta$Sample,
  drop = FALSE
]

olink_observed <- matrixStats::rowSums2(
  is.finite(olink_expr)
)

olink_variance_original <- matrixStats::rowVars(
  olink_expr,
  na.rm = TRUE
)

keep_olink_features <-
  olink_observed >= 2 &
  is.finite(olink_variance_original) &
  olink_variance_original > 0

olink_expr <- olink_expr[
  keep_olink_features,
  ,
  drop = FALSE
]

dim(olink_expr)
```

```{r regress bca from olink}
bca_covariate <- matrix(
  as.numeric(
    scale(olink_meta$BCA_mg_ml)
  ),
  ncol = 1,
  dimnames = list(
    olink_meta$Sample,
    "BCA_scaled"
  )
)

olink_mofa <- limma::removeBatchEffect(
  olink_expr,
  covariates = bca_covariate
)

olink_variance_corrected <- matrixStats::rowVars(
  olink_mofa,
  na.rm = TRUE
)

keep_corrected_olink_features <-
  is.finite(olink_variance_corrected) &
  olink_variance_corrected > 0

olink_mofa <- olink_mofa[
  keep_corrected_olink_features,
  ,
  drop = FALSE
]

olink_feature_table <- tibble::tibble(
  feature = rownames(olink_expr),
  variance_before_BCA_correction =
    matrixStats::rowVars(
      olink_expr,
      na.rm = TRUE
    ),
  variance_after_BCA_correction =
    olink_variance_corrected,
  selected_for_MOFA =
    rownames(olink_expr) %in%
      rownames(olink_mofa)
)

bca_correlations_before <- apply(
  olink_expr,
  1,
  stats::cor,
  y = olink_meta$BCA_mg_ml,
  use = "pairwise.complete.obs"
)

bca_correlations_after <- apply(
  olink_mofa,
  1,
  stats::cor,
  y = olink_meta$BCA_mg_ml,
  use = "pairwise.complete.obs"
)

olink_bca_correction_diagnostic <-
  tibble::tibble(
    feature = names(
      bca_correlations_before
    ),
    correlation_before =
      as.numeric(
        bca_correlations_before
      ),
    correlation_after =
      as.numeric(
        bca_correlations_after[
          names(
            bca_correlations_before
          )
        ]
      )
  )

write.csv(
  olink_feature_table,
  file.path(
    preprocessing_output_dir,
    "Olink_feature_selection_all_proteins.csv"
  ),
  row.names = FALSE
)

write.csv(
  olink_bca_correction_diagnostic,
  file.path(
    preprocessing_output_dir,
    "Olink_BCA_correction_diagnostic.csv"
  ),
  row.names = FALSE
)

dim(olink_mofa)
```

```{r Olink BCA correction diagnostic, fig.height=4, fig.width=4}
olink_bca_correction_plot_data <-
  olink_bca_correction_diagnostic %>%
  tidyr::pivot_longer(
    cols = c(
      correlation_before,
      correlation_after
    ),
    names_to = "stage",
    values_to = "correlation"
  ) %>%
  dplyr::filter(
    is.finite(correlation)
  ) %>%
  dplyr::mutate(
    stage = factor(
      stage,
      levels = c(
        "correlation_before",
        "correlation_after"
      ),
      labels = c(
        "Before correction",
        "After correction"
      )
    ),
    absolute_correlation =
      abs(correlation)
  )

ggplot(
  olink_bca_correction_plot_data,
  aes(
    x = stage,
    y = absolute_correlation,
    fill = stage
  )
) +
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA
  ) +
  geom_jitter(
    width = 0.12,
    height = 0,
    size = 0.6,
    alpha = 0.18
  ) +
  scale_fill_manual(
    values = c(
      "Before correction" = "#F2A7A3",
      "After correction" = "#63C5C9"
    )
  ) +
  labs(
    x = NULL,
    y = "Absolute protein-wise correlation\nwith BCA concentration"
  ) +
  theme_classic() +
  theme(
    legend.position = "none"
  )
```


# Preprocess LC-MS/MS proteomics

The sample-filtered and log2-transformed LC-MS/MS matrix is imported from the proteomics quality-control workflow. Proteins observed in at least half of the retained samples are used for MOFA.

```{r read mass spectrometry}
ms_expr <-
  filtered_proteomics$ms

ms_samples <- clin_all$Sample[
  clin_all$Sample %in%
    colnames(ms_expr)
]

ms_expr <- ms_expr[
  ,
  ms_samples,
  drop = FALSE
]

ms_observed_fraction <- matrixStats::rowMeans2(
  is.finite(ms_expr)
)

ms_variance <- matrixStats::rowVars(
  ms_expr,
  na.rm = TRUE
)

keep_ms_features <-
  ms_observed_fraction >=
    min_ms_observed_fraction &
  is.finite(ms_variance) &
  ms_variance > 0

ms_mofa <- ms_expr[
  keep_ms_features,
  ,
  drop = FALSE
]

ms_feature_table <- tibble::tibble(
  feature = rownames(ms_expr),
  observed_fraction = ms_observed_fraction,
  variance = ms_variance,
  selected_for_MOFA = keep_ms_features
) %>%
  dplyr::arrange(
    dplyr::desc(variance)
  )

write.csv(
  ms_feature_table,
  file.path(
    preprocessing_output_dir,
    "MS_feature_selection.csv"
  ),
  row.names = FALSE
)

dim(ms_mofa)
```

```{r save processed proteomics views}
saveRDS(
  list(
    Olink = olink_mofa,
    LCMS = ms_mofa,
    Olink_original = olink_expr,
    Olink_metadata = olink_meta,
    Olink_BCA_diagnostic =
      olink_bca_correction_diagnostic,
    QC_source_file =
      proteomics_qc_file
  ),
  processed_proteomics_file
)
```

# Model 1: integrated Olink and LC-MS/MS proteomics

Only patients measured by both proteomic platforms are used. This avoids using platform availability to define factors and makes every proteomics factor directly estimable from both assays.

## Assemble the proteomics-only MOFA input

```{r assemble proteomics input}
proteomics_samples <- clin_all$Sample[
  clin_all$Sample %in%
    intersect(
      colnames(olink_mofa),
      colnames(ms_mofa)
    )
]

proteomics_data <- list(
  Olink = olink_mofa[
    ,
    proteomics_samples,
    drop = FALSE
  ],
  LCMS = ms_mofa[
    ,
    proteomics_samples,
    drop = FALSE
  ]
)

lapply(
  proteomics_data,
  dim
)
```

```{r proteomics sample overview}
proteomics_sample_overview <- tibble::tibble(
  Sample = proteomics_samples,
  Olink = TRUE,
  LCMS = TRUE
) %>%
  dplyr::left_join(
    clin_all %>%
      dplyr::select(
        Sample,
        RV_group,
        mPAP_mmHg,
        TAPSE_mm,
        RVIT_mm,
        TAPSE_sPAP
      ),
    by = "Sample"
  )

write.csv(
  proteomics_sample_overview,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_samples.csv"
  ),
  row.names = FALSE
)

proteomics_sample_overview %>%
  dplyr::count(
    RV_group,
    .drop = FALSE
  )
```

```{r save proteomics input checkpoint}
saveRDS(
  proteomics_data,
  proteomics_input_file
)
```

## Checkpoint: load the proteomics input

The saved proteomics input provides a restart point after preprocessing.

```{r load proteomics input checkpoint}
proteomics_data <- readRDS(
  proteomics_input_file
)

lapply(
  proteomics_data,
  dim
)
```

## Create the untrained proteomics model

```{r create proteomics mofa object}
proteomics_untrained <- MOFA2::create_mofa(
  proteomics_data
)

proteomics_untrained
```

```{r proteomics input overview, fig.height=4.5, fig.width=8}
MOFA2::plot_data_overview(
  proteomics_untrained
)
```

## Tune the number of proteomics factors

Three seeds are fitted for every candidate number of factors. Each complete execution refits all tuning models and overwrites the corresponding tuning files.

```{r tune proteomics factor number, message=FALSE, warning=FALSE, results='hide'}
proteomics_tuning_results <-
  tibble::tibble()

for (n_factors in candidate_factors) {
  for (tuning_seed in tuning_seeds) {
    tuning_model_file <- file.path(
      proteomics_output_dir,
      paste0(
        "MOFA_proteomics_tuning_K",
        sprintf("%02d", n_factors),
        "_seed",
        tuning_seed,
        ".hdf5"
      )
    )

    if (file.exists(tuning_model_file)) {
      unlink(tuning_model_file)
    }

    message(
      "Starting proteomics model: K = ",
      n_factors,
      ", seed = ",
      tuning_seed
    )

    tuning_object <- MOFA2::create_mofa(
      proteomics_data
    )

    tuning_data_options <-
      MOFA2::get_default_data_options(
        tuning_object
      )

    tuning_data_options$scale_views <- TRUE

    tuning_model_options <-
      MOFA2::get_default_model_options(
        tuning_object
      )

    tuning_model_options$num_factors <-
      n_factors

    tuning_model_options$likelihoods[] <-
      "gaussian"

    tuning_training_options <-
      MOFA2::get_default_training_options(
        tuning_object
      )

    tuning_training_options$maxiter <-
      tuning_maxiter

    tuning_training_options$convergence_mode <-
      mofa_convergence_mode

    tuning_training_options$freqELBO <-
      mofa_freq_elbo

    tuning_training_options$drop_factor_threshold <-
      -1

    tuning_training_options$seed <-
      tuning_seed

    tuning_training_options$verbose <-
      FALSE

    tuning_object <- MOFA2::prepare_mofa(
      object = tuning_object,
      data_options = tuning_data_options,
      model_options = tuning_model_options,
      training_options =
        tuning_training_options
    )

    MOFA2::run_mofa(
      tuning_object,
      outfile = tuning_model_file,
      use_basilisk = TRUE
    )

    rm(tuning_object)
    gc()

    tuning_model <- MOFA2::load_model(
      tuning_model_file,
      sort_factors = TRUE,
      remove_inactive_factors = FALSE
    )

    tuning_r2 <-
      MOFA2::get_variance_explained(
        tuning_model
      )$r2_per_factor[[1]]

    weakest_factor <-
      tail(
        rownames(tuning_r2),
        1
      )

    weakest_factor_r2 <-
      tuning_r2[
        weakest_factor,
        ,
        drop = TRUE
      ]

    new_tuning_result <- tibble::tibble(
      n_factors = n_factors,
      seed = tuning_seed,
      final_ELBO =
        MOFA2::get_elbo(tuning_model),
      weakest_factor = weakest_factor,
      weakest_factor_total_R2 =
        sum(
          weakest_factor_r2,
          na.rm = TRUE
        ),
      weakest_factor_Olink_R2 =
        unname(
          weakest_factor_r2["Olink"]
        ),
      weakest_factor_LCMS_R2 =
        unname(
          weakest_factor_r2["LCMS"]
        ),
      n_factors_total_R2_at_least_cutoff =
        sum(
          rowSums(
            tuning_r2,
            na.rm = TRUE
          ) >= last_factor_r2_cutoff
        ),
      model_file = tuning_model_file
    )

    proteomics_tuning_results <-
      dplyr::bind_rows(
        proteomics_tuning_results,
        new_tuning_result
      ) %>%
      dplyr::arrange(
        n_factors,
        seed
      )

    write.csv(
      proteomics_tuning_results,
      proteomics_tuning_summary_file,
      row.names = FALSE
    )

    message(
      "Finished proteomics model: K = ",
      n_factors,
      ", seed = ",
      tuning_seed
    )

    rm(tuning_model)
    gc()
  }
}

proteomics_tuning_results
```

## Select the proteomics factor number

```{r select proteomics best tuning run per K}
proteomics_best_tuning_runs <-
  proteomics_tuning_results %>%
  dplyr::group_by(n_factors) %>%
  dplyr::slice_max(
    order_by = final_ELBO,
    n = 1,
    with_ties = FALSE
  ) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(n_factors)

write.csv(
  proteomics_best_tuning_runs,
  proteomics_best_tuning_summary_file,
  row.names = FALSE
)

proteomics_low_weakest_factor <-
  proteomics_best_tuning_runs$
    weakest_factor_total_R2 <
  last_factor_r2_cutoff

proteomics_low_runs <- vapply(
  seq_along(
    proteomics_low_weakest_factor
  ),
  function(i) {
    end_i <-
      i +
      n_consecutive_low_models -
      1

    end_i <=
      length(
        proteomics_low_weakest_factor
      ) &&
      all(
        proteomics_low_weakest_factor[
          i:end_i
        ]
      )
  },
  logical(1)
)

proteomics_first_low_run <-
  which(proteomics_low_runs)[1]

if (
  length(proteomics_first_low_run) == 1 &&
  !is.na(proteomics_first_low_run)
) {
  proteomics_selected_n_factors <- max(
    proteomics_best_tuning_runs$
      n_factors[
        proteomics_first_low_run
      ] - 1,
    min(candidate_factors)
  )

  proteomics_factor_plateau_reached <-
    TRUE
} else {
  proteomics_selected_n_factors <-
    max(candidate_factors)

  proteomics_factor_plateau_reached <-
    FALSE

  warning(
    paste0(
      "No sequence of ",
      n_consecutive_low_models,
      " consecutive proteomics models with weakest-factor variance below ",
      last_factor_r2_cutoff,
      "% was observed. The maximum tested factor number was selected."
    )
  )
}

proteomics_selected_seed <-
  proteomics_best_tuning_runs %>%
  dplyr::filter(
    n_factors ==
      proteomics_selected_n_factors
  ) %>%
  dplyr::pull(seed)

proteomics_selected_model_settings <-
  tibble::tibble(
    selected_n_factors =
      proteomics_selected_n_factors,
    selected_seed =
      proteomics_selected_seed,
    factor_plateau_reached =
      proteomics_factor_plateau_reached,
    weakest_factor_R2_cutoff =
      last_factor_r2_cutoff,
    consecutive_low_models_required =
      n_consecutive_low_models,
    tuning_maxiter = tuning_maxiter,
    final_maxiter = final_maxiter,
    convergence_mode =
      mofa_convergence_mode
  )

write.csv(
  proteomics_selected_model_settings,
  proteomics_selected_settings_file,
  row.names = FALSE
)

proteomics_selected_model_settings
```

```{r proteomics factor tuning plot, fig.height=3.3, fig.width=4.4}
ggplot() +
  geom_line(
    data = proteomics_tuning_results,
    aes(
      x = n_factors,
      y = weakest_factor_total_R2,
      group = factor(seed)
    ),
    color = "grey70"
  ) +
  geom_point(
    data = proteomics_tuning_results,
    aes(
      x = n_factors,
      y = weakest_factor_total_R2
    ),
    color = "grey60",
    size = 1.8
  ) +
  geom_line(
    data = proteomics_best_tuning_runs,
    aes(
      x = n_factors,
      y = weakest_factor_total_R2
    ),
    linewidth = 0.8
  ) +
  geom_point(
    data = proteomics_best_tuning_runs,
    aes(
      x = n_factors,
      y = weakest_factor_total_R2
    ),
    size = 2.7
  ) +
  geom_hline(
    yintercept =
      last_factor_r2_cutoff,
    linetype = 2
  ) +
  geom_vline(
    xintercept =
      proteomics_selected_n_factors,
    linetype = 3
  ) +
  scale_x_continuous(
    breaks = candidate_factors
  ) +
  labs(
    x = "Number of factors in fitted model",
    y = "Variance explained by weakest factor (%)"
  ) +
  theme_classic()
```

## Train the final proteomics model

```{r train final proteomics mofa model, message=FALSE, warning=FALSE, results='hide'}
proteomics_data <- readRDS(
  proteomics_input_file
)

if (file.exists(proteomics_final_model_file)) {
  unlink(proteomics_final_model_file)
}

proteomics_object <-
  MOFA2::create_mofa(
    proteomics_data
  )

proteomics_data_options <-
  MOFA2::get_default_data_options(
    proteomics_object
  )

proteomics_data_options$scale_views <-
  TRUE

proteomics_model_options <-
  MOFA2::get_default_model_options(
    proteomics_object
  )

proteomics_model_options$num_factors <-
  proteomics_selected_n_factors

proteomics_model_options$likelihoods[] <-
  "gaussian"

proteomics_training_options <-
  MOFA2::get_default_training_options(
    proteomics_object
  )

proteomics_training_options$maxiter <-
  final_maxiter

proteomics_training_options$convergence_mode <-
  mofa_convergence_mode

proteomics_training_options$freqELBO <-
  mofa_freq_elbo

proteomics_training_options$drop_factor_threshold <-
  -1

proteomics_training_options$seed <-
  proteomics_selected_seed

proteomics_training_options$verbose <-
  FALSE

proteomics_object <- MOFA2::prepare_mofa(
  object = proteomics_object,
  data_options =
    proteomics_data_options,
  model_options =
    proteomics_model_options,
  training_options =
    proteomics_training_options
)

MOFA2::run_mofa(
  proteomics_object,
  outfile =
    proteomics_final_model_file,
  use_basilisk = TRUE
)

rm(proteomics_object)
gc()
```

## Load the proteomics model and attach metadata

```{r load proteomics model and attach metadata}
proteomics_model <- MOFA2::load_model(
  proteomics_final_model_file,
  sort_factors = TRUE,
  remove_inactive_factors = FALSE
)

proteomics_model_samples <-
  MOFA2::samples_names(
    proteomics_model
  )[[1]]

proteomics_model_metadata <-
  clin_all %>%
  dplyr::left_join(
    olink_sample_meta %>%
      dplyr::select(
        Sample,
        Batch,
        BCA_mg_ml
      ),
    by = "Sample"
  ) %>%
  dplyr::filter(
    Sample %in%
      proteomics_model_samples
  ) %>%
  dplyr::slice(
    match(
      proteomics_model_samples,
      Sample
    )
  ) %>%
  dplyr::rename(
    sample = Sample
  )

samples_metadata(proteomics_model) <-
  as.data.frame(
    proteomics_model_metadata
  )

saveRDS(
  proteomics_model,
  proteomics_final_model_rds
)

write.csv(
  proteomics_model_metadata,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_sample_metadata.csv"
  ),
  row.names = FALSE
)
```

## Checkpoint: reload the final proteomics model

The saved final proteomics model provides a restart point for downstream analyses.

```{r reload final proteomics model checkpoint}
proteomics_model <- readRDS(
  proteomics_final_model_rds
)

proteomics_model_metadata <-
  samples_metadata(
    proteomics_model
  )

proteomics_selected_model_settings <-
  read.csv(
    proteomics_selected_settings_file,
    stringsAsFactors = FALSE
  )

proteomics_tuning_results <-
  read.csv(
    proteomics_tuning_summary_file,
    stringsAsFactors = FALSE
  )

proteomics_best_tuning_runs <-
  read.csv(
    proteomics_best_tuning_summary_file,
    stringsAsFactors = FALSE
  )
```

# Proteomics model overview

```{r proteomics trained model overview, fig.height=4.5, fig.width=8}
MOFA2::plot_data_overview(
  proteomics_model
)
```

```{r proteomics variance explained, fig.height=3.2, fig.width=3.6}
proteomics_r2_heatmap_matrix <-
  MOFA2::get_variance_explained(
    proteomics_model
  )$r2_per_factor[[1]]

proteomics_r2_heatmap_data <-
  as.data.frame(
    proteomics_r2_heatmap_matrix
  ) %>%
  tibble::rownames_to_column(
    "Factor"
  ) %>%
  tidyr::pivot_longer(
    cols = -Factor,
    names_to = "View",
    values_to = "R2"
  ) %>%
  dplyr::mutate(
    Factor = factor(
      Factor,
      levels = rownames(
        proteomics_r2_heatmap_matrix
      )
    ),
    View = factor(
      View,
      levels = colnames(
        proteomics_r2_heatmap_matrix
      )
    ),
    text_color = dplyr::if_else(
      R2 >=
        0.55 *
        max(
          R2,
          na.rm = TRUE
        ),
      "white",
      "black"
    )
  )

ggplot(
  proteomics_r2_heatmap_data,
  aes(
    x = View,
    y = Factor,
    fill = R2
  )
) +
  geom_tile(
    color = "grey40",
    linewidth = 0.3
  ) +
  geom_text(
    aes(
      label = sprintf(
        "%.1f%%",
        R2
      ),
      color = text_color
    ),
    size = 3.2
  ) +
  scale_color_identity() +
  scale_fill_gradient(
    low = "white",
    high = "navy"
  ) +
  labs(
    title =
      "",
    x = NULL,
    y = NULL,
    fill = "Variance\nexplained (%)"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    ),
  axis.text.x = element_text(
    size = 11
  ),
  axis.text.y = element_text(
    size = 11
  )
  )
```

```{r proteomics total variance explained, fig.height=4.5, fig.width=7}
MOFA2::plot_variance_explained(
  proteomics_model,
  x = "view",
  y = "factor",
  plot_total = TRUE
)[[2]]
```

```{r proteomics factor correlations, fig.height=4, fig.width=5}
MOFA2::plot_factor_cor(
  proteomics_model
) +
  labs(
    title = "Proteomics factor correlations"
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    )
  )
```

```{r extract proteomics variance explained}
proteomics_variance_explained <-
  MOFA2::get_variance_explained(
    proteomics_model
  )

proteomics_r2_total <-
  proteomics_variance_explained$
    r2_total[[1]]

proteomics_r2_per_factor <-
  proteomics_variance_explained$
    r2_per_factor[[1]]

proteomics_r2_total_table <-
  tibble::enframe(
    proteomics_r2_total,
    name = "View",
    value = "Total_R2"
  )

proteomics_r2_per_factor_table <-
  as.data.frame(
    proteomics_r2_per_factor
  ) %>%
  tibble::rownames_to_column(
    "Factor"
  ) %>%
  tidyr::pivot_longer(
    cols = -Factor,
    names_to = "View",
    values_to = "R2"
  )

write.csv(
  proteomics_r2_total_table,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_total_variance_explained.csv"
  ),
  row.names = FALSE
)

write.csv(
  proteomics_r2_per_factor_table,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_variance_explained_per_factor.csv"
  ),
  row.names = FALSE
)
```

# Extract proteomics factor scores and weights

```{r extract proteomics factors and weights}
proteomics_factor_scores <-
  MOFA2::get_factors(
    proteomics_model,
    as.data.frame = TRUE
  ) %>%
  dplyr::rename(
    Sample = sample,
    Factor = factor,
    Factor_value = value
  ) %>%
  dplyr::left_join(
    as.data.frame(
      proteomics_model_metadata
    ) %>%
      dplyr::rename(
        Sample = sample
      ),
    by = "Sample"
  )

proteomics_factor_scores_wide <-
  proteomics_factor_scores %>%
  dplyr::select(
    Sample,
    Factor,
    Factor_value
  ) %>%
  tidyr::pivot_wider(
    names_from = Factor,
    values_from = Factor_value
  )

proteomics_factor_weights <-
  MOFA2::get_weights(
    proteomics_model,
    as.data.frame = TRUE
  ) %>%
  dplyr::rename(
    View = view,
    Feature = feature,
    Factor = factor,
    Weight = value
  )

write.csv(
  proteomics_factor_scores,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_factor_scores_long.csv"
  ),
  row.names = FALSE
)

write.csv(
  proteomics_factor_scores_wide,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_factor_scores_wide.csv"
  ),
  row.names = FALSE
)

write.csv(
  proteomics_factor_weights,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_feature_weights_all.csv"
  ),
  row.names = FALSE
)
```

```{r proteomics factor1 top loadings, fig.height=7, fig.width=5}
proteomics_factor1_top_loadings <-
  prepare_top_factor_loadings(
    weight_table =
      proteomics_factor_weights,
    factor_name = "Factor1",
    n_each_direction =
      n_top_factor_loadings_each_direction
  )

write.csv(
  proteomics_factor1_top_loadings,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_Factor1_top_loadings.csv"
  ),
  row.names = FALSE
)

proteomics_factor1_loading_plot <-
  plot_top_factor_loadings(
    loading_data =
      proteomics_factor1_top_loadings,
    title_text =
      "Factor 1 top loadings"
  )

proteomics_factor1_loading_plot

```

# Proteomics factors in compensated and decompensated RV states

```{r proteomics group association models}
proteomics_factor_names <-
  MOFA2::factors_names(
    proteomics_model
  )

proteomics_group_global_results <-
  tibble::tibble()

proteomics_group_pairwise_results <-
  tibble::tibble()

for (
  factor_name in
  proteomics_factor_names
) {
  factor_data <-
    proteomics_factor_scores %>%
    dplyr::filter(
      Factor == factor_name,
      RV_group %in%
        rv_group_comparison_levels
    ) %>%
    dplyr::select(
      Sample,
      Factor_value,
      RV_group,
      Age_years,
      Sample_age_days
    ) %>%
    tidyr::drop_na() %>%
    dplyr::mutate(
      RV_group = factor(
        as.character(RV_group),
        levels =
          rv_group_comparison_levels
      )
    ) %>%
    droplevels()

  if (
    nlevels(factor_data$RV_group) < 2L
  ) {
    next
  }

  group_counts <-
    table(
      factor_data$RV_group
    )

  group_fit <- lm(
    Factor_value ~
      RV_group +
      scale(Age_years) +
      scale(Sample_age_days),
    data = factor_data
  )

  group_test <- drop1(
    group_fit,
    test = "F"
  )

  proteomics_group_global_results <-
    dplyr::bind_rows(
      proteomics_group_global_results,
      tibble::tibble(
        Factor = factor_name,
        comparison =
          "Compensated versus Decompensated",
        n = nrow(factor_data),
        n_compensated =
          unname(
            group_counts["Compensated"]
          ),
        n_decompensated =
          unname(
            group_counts["Decompensated"]
          ),
        F_statistic =
          group_test[
            "RV_group",
            "F value"
          ],
        p_value =
          group_test[
            "RV_group",
            "Pr(>F)"
          ],
        model_R2 =
          summary(group_fit)$r.squared,
        adjusted_model_R2 =
          summary(group_fit)$adj.r.squared
      )
    )

  factor_emmeans <- emmeans::emmeans(
    group_fit,
    ~ RV_group
  )

  factor_contrasts <-
    emmeans::contrast(
      factor_emmeans,
      method = list(
        "Decompensated - Compensated" =
          c(-1, 1)
      ),
      adjust = "none"
    ) %>%
    summary(
      infer = c(TRUE, TRUE)
    ) %>%
    as.data.frame() %>%
    dplyr::mutate(
      Factor = factor_name,
      n = nrow(factor_data),
      n_compensated =
        unname(
          group_counts["Compensated"]
        ),
      n_decompensated =
        unname(
          group_counts["Decompensated"]
        )
    ) %>%
    dplyr::select(
      Factor,
      contrast,
      estimate,
      SE,
      df,
      lower.CL,
      upper.CL,
      t.ratio,
      p.value,
      n,
      n_compensated,
      n_decompensated
    )

  proteomics_group_pairwise_results <-
    dplyr::bind_rows(
      proteomics_group_pairwise_results,
      factor_contrasts
    )
}

proteomics_group_global_results <-
  proteomics_group_global_results %>%
  dplyr::mutate(
    FDR = p.adjust(
      p_value,
      method = "BH"
    )
  ) %>%
  dplyr::arrange(FDR)

proteomics_group_pairwise_results <-
  proteomics_group_pairwise_results %>%
  dplyr::mutate(
    FDR = p.adjust(
      p.value,
      method = "BH"
    )
  ) %>%
  dplyr::arrange(FDR)

write.csv(
  proteomics_group_global_results,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_factors_compensated_vs_decompensated_global.csv"
  ),
  row.names = FALSE
)

write.csv(
  proteomics_group_pairwise_results,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_factors_compensated_vs_decompensated.csv"
  ),
  row.names = FALSE
)

reactable::reactable(
  proteomics_group_pairwise_results,
  searchable = TRUE,
  filterable = TRUE,
  defaultPageSize = 15
)
```

```{r proteomics group comparison results table}
proteomics_group_pairwise_results %>%
  dplyr::select(
    Factor,
    contrast,
    estimate,
    SE,
    lower.CL,
    upper.CL,
    t.ratio,
    p.value,
    FDR,
    n,
    n_compensated,
    n_decompensated
  ) %>%
  dplyr::arrange(FDR) %>%
  reactable::reactable(
    searchable = TRUE,
    filterable = TRUE,
    defaultPageSize = 15
  )
```

```{r proteomics factors by rv group, fig.height=5, fig.width=7}
proteomics_group_plot_data <-
  proteomics_factor_scores %>%
  dplyr::filter(
    RV_group %in%
      rv_group_comparison_levels
  ) %>%
  dplyr::mutate(
    RV_group = factor(
      as.character(RV_group),
      levels =
        rv_group_comparison_levels
    ),
    Factor = factor(
      Factor,
      levels =
        proteomics_factor_names
    )
  )

proteomics_group_fdr_labels <-
  proteomics_group_pairwise_results %>%
  dplyr::mutate(
    Factor = factor(
      Factor,
      levels =
        proteomics_factor_names
    ),
    label =
      format_fdr_label(FDR)
  )

ggplot(
  proteomics_group_plot_data,
  aes(
    x = RV_group,
    y = Factor_value,
    fill = RV_group
  )
) +
  geom_violin(
    trim = FALSE,
    alpha = 0.30,
    color = NA
  ) +
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    color = "grey20"
  ) +
  geom_jitter(
    width = 0.10,
    height = 0,
    size = 1.8,
    shape = 21,
    color = "black",
    stroke = 0.25
  ) +
  geom_text(
    data =
      proteomics_group_fdr_labels,
    aes(
      x = 1.5,
      y = Inf,
      label = label
    ),
    inherit.aes = FALSE,
    vjust = 1.25,
    size = 3.2
  ) +
  scale_fill_manual(
    values = group_colors[
      rv_group_comparison_levels
    ],
    drop = FALSE
  ) +
  scale_y_continuous(
    expand = expansion(
      mult = c(0.05, 0.16)
    )
  ) +
  facet_wrap(
    ~ Factor,
    scales = "free_y",
    ncol = 4
  ) +
  labs(
    x = NULL,
    y = "Factor score"
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    strip.text =
      element_text(face = "bold")
  )
```

```{r proteomics Factor1 compensated versus decompensated, fig.height=2.5, fig.width=2.5}
proteomics_factor1_group_plot_data <-
  proteomics_factor_scores %>%
  dplyr::filter(
    Factor == "Factor1",
    RV_group %in%
      rv_group_comparison_levels
  ) %>%
  dplyr::mutate(
    RV_group = factor(
      as.character(RV_group),
      levels =
        rv_group_comparison_levels
    )
  )

proteomics_factor1_group_p_value <-
  proteomics_group_pairwise_results %>%
  dplyr::filter(
    Factor == "Factor1"
  ) %>%
  dplyr::pull(
    p.value
  )

proteomics_factor1_group_label <-
  format_p_value(
    dplyr::first(
      proteomics_factor1_group_p_value
    )
  )

proteomics_factor1_group_plot <-
  ggplot(
    proteomics_factor1_group_plot_data,
    aes(
      x = RV_group,
      y = Factor_value,
      fill = RV_group
    )
  ) +
  geom_violin(
    trim = FALSE,
    alpha = 0.30,
    color = NA
  ) +
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    color = "grey20"
  ) +
  geom_jitter(
    width = 0.10,
    height = 0,
    size = 2,
    shape = 21,
    color = "black",
    stroke = 0.25
  ) +
  annotate(
    "text",
    x = 1.5,
    y = Inf,
    label =
      proteomics_factor1_group_label,
    vjust = 1.2,
    size = 3.5
  ) +
  scale_fill_manual(
    values = group_colors[
      rv_group_comparison_levels
    ],
    drop = FALSE
  ) +
  scale_y_continuous(
    expand = expansion(
      mult = c(
        0.05,
        0.16
      )
    )
  ) +
  labs(
    x = NULL,
    y = "Factor 1 score"
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    )
  )

proteomics_factor1_group_plot

```


# Proteomics factor associations with continuous clinical parameters

Each parameter is tested in a separate model:

`factor score ~ scaled clinical parameter + scaled age + scaled sample storage time`

```{r proteomics continuous clinical association models}
proteomics_continuous_results <-
  tibble::tibble()

for (
  factor_name in
  proteomics_factor_names
) {
  for (
    clinical_parameter in
    clinical_parameters
  ) {
    factor_data <-
      proteomics_factor_scores %>%
      dplyr::filter(
        Factor == factor_name
      ) %>%
      dplyr::select(
        Sample,
        Factor_value,
        Age_years,
        Sample_age_days,
        dplyr::all_of(
          clinical_parameter
        )
      ) %>%
      tidyr::drop_na()

    factor_data <- factor_data %>%
      dplyr::mutate(
        clinical_scaled =
          as.numeric(
            scale(
              .data[[clinical_parameter]]
            )
          ),
        age_scaled =
          as.numeric(
            scale(Age_years)
          ),
        storage_scaled =
          as.numeric(
            scale(
              Sample_age_days
            )
          )
      )

    continuous_fit <- lm(
      Factor_value ~
        clinical_scaled +
        age_scaled +
        storage_scaled,
      data = factor_data
    )

    clinical_coefficient <-
      broom::tidy(
        continuous_fit,
        conf.int = TRUE
      ) %>%
      dplyr::filter(
        term ==
          "clinical_scaled"
      )

    proteomics_continuous_results <-
      dplyr::bind_rows(
        proteomics_continuous_results,
        tibble::tibble(
          Factor = factor_name,
          parameter =
            clinical_parameter,
          n = nrow(factor_data),
          beta =
            clinical_coefficient$
              estimate,
          standard_error =
            clinical_coefficient$
              std.error,
          confidence_interval_low =
            clinical_coefficient$
              conf.low,
          confidence_interval_high =
            clinical_coefficient$
              conf.high,
          t_statistic =
            clinical_coefficient$
              statistic,
          p_value =
            clinical_coefficient$
              p.value,
          model_R2 =
            summary(
              continuous_fit
            )$r.squared,
          adjusted_model_R2 =
            summary(
              continuous_fit
            )$adj.r.squared
        )
      )
  }
}

proteomics_continuous_results <-
  proteomics_continuous_results %>%
  dplyr::group_by(parameter) %>%
  dplyr::mutate(
    FDR_within_parameter =
      p.adjust(
        p_value,
        method = "BH"
      )
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    FDR_global =
      p.adjust(
        p_value,
        method = "BH"
      ),
    parameter_label =
      unname(
        pretty_parameter_names[
          parameter
        ]
      )
  ) %>%
  dplyr::arrange(
    FDR_global
  )

write.csv(
  proteomics_continuous_results,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_factors_vs_continuous_clinical_parameters.csv"
  ),
  row.names = FALSE
)

reactable::reactable(
  proteomics_continuous_results,
  searchable = TRUE,
  filterable = TRUE
)
```

```{r proteomics continuous association dotplot, fig.height=4, fig.width=6.5}
proteomics_continuous_plot_data <-
  proteomics_continuous_results %>%
  dplyr::mutate(
    parameter_label = factor(
      parameter_label,
      levels = unname(
        pretty_parameter_names[
          clinical_parameters
        ]
      )
    ),
    Factor = factor(
      Factor,
      levels = rev(
        proteomics_factor_names
      )
    ),
    dot_size = -log10(
      pmax(
        FDR_global,
        .Machine$double.xmin
      )
    ),
    outlined =
      FDR_global <
      exploratory_factor_fdr
  )

ggplot(
  proteomics_continuous_plot_data,
  aes(
    x = parameter_label,
    y = Factor,
    size = dot_size,
    fill = beta
  )
) +
  geom_point(
    shape = 21,
    color = "grey60",
    stroke = 0.3
  ) +
  geom_point(
    data = dplyr::filter(
      proteomics_continuous_plot_data,
      outlined
    ),
    shape = 21,
    fill = NA,
    color = "black",
    stroke = 1,
    show.legend = FALSE
  ) +
  scale_size_continuous(
    range = c(2.5, 8)
  ) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0
  ) +
  labs(
    x = NULL,
    y = NULL,
    size =
      expression(
        -log[10](FDR)
      ),
    fill =
      "Coefficient\nper 1 SD"
  ) +
  theme_classic(
  base_size = 14
) +
theme(
  axis.text.x = element_text(
    angle = 45,
    hjust = 1,
    size = 12
  ),
  axis.text.y = element_text(
    size = 12
  ),
  legend.title = element_text(
    size = 12
  ),
  legend.text = element_text(
    size = 11
  )
)
  
```

```{r proteomics strongest clinical scatter plots, fig.height=3, fig.width=9}
proteomics_strongest_associations <-
  proteomics_continuous_results %>%
  dplyr::filter(
    FDR_global < 0.20
  ) %>%
  dplyr::arrange(
    FDR_global
  ) %>%
  dplyr::slice_head(
    n = 12
  )

proteomics_clinical_scatter_plots <-
  list()

for (
  i in seq_len(
    nrow(
      proteomics_strongest_associations
    )
  )
) {
  factor_name <-
    proteomics_strongest_associations$
      Factor[i]

  clinical_parameter <-
    proteomics_strongest_associations$
      parameter[i]

  adjusted_beta <-
    proteomics_strongest_associations$
      beta[i]

  global_fdr <-
    proteomics_strongest_associations$
      FDR_global[i]

  association_label <-
    paste0(
      "\u03B2adj = ",
      sprintf(
        "%.2f",
        adjusted_beta
      ),
      "\n",
      dplyr::case_when(
        global_fdr < 0.001 ~
          "FDR < 0.001",
        TRUE ~
          paste0(
            "FDR = ",
            sprintf(
              "%.3f",
              global_fdr
            )
          )
      )
    )

  plot_data <-
    proteomics_factor_scores %>%
    dplyr::filter(
      Factor == factor_name,
      !is.na(
        .data[[clinical_parameter]]
      )
    )

  proteomics_clinical_scatter_plots[[i]] <-
    ggplot(
      plot_data,
      aes(
        x =
          .data[[clinical_parameter]],
        y = Factor_value
      )
    ) +
    geom_point(
      size = 2
    ) +
    geom_smooth(
      method = "lm",
      se = TRUE
    ) +
    annotate(
      geom = "text",
      x = Inf,
      y = Inf,
      label =
        association_label,
      hjust = 1.08,
      vjust = 1.15,
      size = 3.5
    ) +
    labs(
      title = factor_name,
      x =
        pretty_parameter_names[[clinical_parameter]],
      y = "Factor score"
    ) +
    scale_x_continuous(
      expand = expansion(
        mult = c(
          0.05,
          0.08
        )
      )
    ) +
    scale_y_continuous(
      expand = expansion(
        mult = c(
          0.05,
          0.20
        )
      )
    ) +
    theme_classic(
      base_size = 12
    )
}

if (
  length(
    proteomics_clinical_scatter_plots
  ) > 0
) {
  patchwork::wrap_plots(
    proteomics_clinical_scatter_plots,
    ncol = 3
  )
}
```

# Proteomics technical diagnostics

BCA is tested as a post hoc diagnostic. Its linear effect was removed from the Olink input matrix, but not from LC-MS/MS; associations with LC-MS/MS-dominant factors can therefore remain.

```{r proteomics technical diagnostic models}
proteomics_technical_results <-
  tibble::tibble()

technical_variables <- c(
  "BCA_mg_ml",
  "Age_years",
  "Sample_age_days"
)

for (
  factor_name in
  proteomics_factor_names
) {
  for (
    technical_variable in
    technical_variables
  ) {
    if (
      technical_variable ==
      "BCA_mg_ml"
    ) {
      technical_data <-
        proteomics_factor_scores %>%
        dplyr::filter(
          Factor == factor_name
        ) %>%
        dplyr::select(
          Factor_value,
          BCA_mg_ml,
          Age_years,
          Sample_age_days
        ) %>%
        tidyr::drop_na() %>%
        dplyr::mutate(
          technical_scaled =
            as.numeric(
              scale(BCA_mg_ml)
            ),
          age_scaled =
            as.numeric(
              scale(Age_years)
            ),
          storage_scaled =
            as.numeric(
              scale(
                Sample_age_days
              )
            )
        )

      technical_fit <- lm(
        Factor_value ~
          technical_scaled +
          age_scaled +
          storage_scaled,
        data = technical_data
      )
    } else if (
      technical_variable ==
      "Age_years"
    ) {
      technical_data <-
        proteomics_factor_scores %>%
        dplyr::filter(
          Factor == factor_name
        ) %>%
        dplyr::select(
          Factor_value,
          Age_years,
          Sample_age_days
        ) %>%
        tidyr::drop_na() %>%
        dplyr::mutate(
          technical_scaled =
            as.numeric(
              scale(Age_years)
            ),
          storage_scaled =
            as.numeric(
              scale(
                Sample_age_days
              )
            )
        )

      technical_fit <- lm(
        Factor_value ~
          technical_scaled +
          storage_scaled,
        data = technical_data
      )
    } else {
      technical_data <-
        proteomics_factor_scores %>%
        dplyr::filter(
          Factor == factor_name
        ) %>%
        dplyr::select(
          Factor_value,
          Sample_age_days,
          Age_years
        ) %>%
        tidyr::drop_na() %>%
        dplyr::mutate(
          technical_scaled =
            as.numeric(
              scale(
                Sample_age_days
              )
            ),
          age_scaled =
            as.numeric(
              scale(Age_years)
            )
        )

      technical_fit <- lm(
        Factor_value ~
          technical_scaled +
          age_scaled,
        data = technical_data
      )
    }

    technical_coefficient <-
      broom::tidy(
        technical_fit,
        conf.int = TRUE
      ) %>%
      dplyr::filter(
        term ==
          "technical_scaled"
      )

    proteomics_technical_results <-
      dplyr::bind_rows(
        proteomics_technical_results,
        tibble::tibble(
          Factor = factor_name,
          technical_variable =
            technical_variable,
          n = nrow(technical_data),
          beta =
            technical_coefficient$
              estimate,
          standard_error =
            technical_coefficient$
              std.error,
          confidence_interval_low =
            technical_coefficient$
              conf.low,
          confidence_interval_high =
            technical_coefficient$
              conf.high,
          p_value =
            technical_coefficient$
              p.value
        )
      )
  }
}

proteomics_technical_results <-
  proteomics_technical_results %>%
  dplyr::group_by(
    technical_variable
  ) %>%
  dplyr::mutate(
    FDR = p.adjust(
      p_value,
      method = "BH"
    )
  ) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(FDR)

write.csv(
  proteomics_technical_results,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_factor_technical_diagnostics.csv"
  ),
  row.names = FALSE
)
```

```{r proteomics technical diagnostic dotplot, fig.height=5, fig.width=7}
proteomics_technical_plot_data <-
  proteomics_technical_results %>%
  dplyr::mutate(
    technical_variable = factor(
      technical_variable,
      levels = c(
        "Age_years",
        "BCA_mg_ml",
        "Sample_age_days"
      ),
      labels = c(
        "Age",
        "BCA concentration",
        "Sample storage time"
      )
    ),
    Factor = factor(
      Factor,
      levels = rev(
        proteomics_factor_names
      )
    ),
    dot_size = -log10(
      pmax(
        FDR,
        .Machine$double.xmin
      )
    ),
    outlined =
      FDR <
      exploratory_factor_fdr
  )

ggplot(
  proteomics_technical_plot_data,
  aes(
    x = technical_variable,
    y = Factor,
    size = dot_size,
    fill = beta
  )
) +
  geom_point(
    shape = 21,
    color = "grey60",
    stroke = 0.3
  ) +
  geom_point(
    data = dplyr::filter(
      proteomics_technical_plot_data,
      outlined
    ),
    shape = 21,
    fill = NA,
    color = "black",
    stroke = 1,
    show.legend = FALSE
  ) +
  scale_size_continuous(
    range = c(2.5, 8)
  ) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0
  ) +
  labs(
    x = NULL,
    y = NULL,
    size =
      expression(
        -log[10](FDR)
      ),
    fill = "Coefficient"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    )
  )
```

# Proteomics factor summary

```{r proteomics factor summary}
proteomics_factor_total_r2 <-
  proteomics_r2_per_factor_table %>%
  dplyr::group_by(Factor) %>%
  dplyr::summarise(
    total_R2_across_views =
      sum(R2, na.rm = TRUE),
    maximum_view_R2 =
      max(R2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::arrange(
    dplyr::desc(
      total_R2_across_views
    )
  )

proteomics_factor_summary <-
  proteomics_factor_total_r2 %>%
  dplyr::left_join(
    proteomics_group_global_results %>%
      dplyr::select(
        Factor,
        RV_group_p = p_value,
        RV_group_FDR = FDR
      ),
    by = "Factor"
  ) %>%
  dplyr::left_join(
    proteomics_continuous_results %>%
      dplyr::group_by(Factor) %>%
      dplyr::slice_min(
        FDR_global,
        n = 1,
        with_ties = FALSE
      ) %>%
      dplyr::ungroup() %>%
      dplyr::select(
        Factor,
        strongest_clinical_parameter =
          parameter,
        strongest_clinical_beta =
          beta,
        strongest_clinical_FDR =
          FDR_global
      ),
    by = "Factor"
  ) %>%
  dplyr::arrange(
    dplyr::desc(
      total_R2_across_views
    )
  )

write.csv(
  proteomics_factor_summary,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_main_factor_summary.csv"
  ),
  row.names = FALSE
)

reactable::reactable(
  proteomics_factor_summary,
  searchable = TRUE,
  filterable = TRUE
)
```

# Proteomics Factor 1 gene set enrichment analysis

Factor 1 was analysed by Gene Ontology GSEA using the complete ranked weight vector. Olink and LC-MS/MS weights were scaled within each platform before aggregation.

```{r proteomics Factor1 GSEA inputs}
proteomics_factor_enrichment_table <-
  proteomics_gsea_factor_table %>%
  dplyr::filter(
    Factor %in%
      proteomics_factor_names
  )

proteomics_enrichment_weights <-
  proteomics_factor_weights %>%
  dplyr::filter(
    View %in%
      c(
        "Olink",
        "LCMS"
      ),
    Factor %in%
      proteomics_factor_enrichment_table$
        Factor
  )
```

```{r proteomics Factor1 GO GSEA, message=TRUE, warning=TRUE}
proteomics_factor1_gsea <-
  run_mofa_gsea(
    weight_table =
      proteomics_enrichment_weights,
    factor_table =
      proteomics_factor_enrichment_table
  )

proteomics_factor1_gsea_simplified_objects <-
  simplify_enrichment_objects(
    proteomics_factor1_gsea$objects
  )

proteomics_factor1_gsea_results <-
  proteomics_factor1_gsea$results

proteomics_factor1_gsea_simplified_results <-
  extract_gsea_results(
    proteomics_factor1_gsea_simplified_objects
  )

write.csv(
  proteomics_factor1_gsea_results,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_Factor1_GO_GSEA.csv"
  ),
  row.names = FALSE
)

write.csv(
  proteomics_factor1_gsea_simplified_results,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_Factor1_GO_GSEA_simplified.csv"
  ),
  row.names = FALSE
)

saveRDS(
  list(
    GSEA =
      proteomics_factor1_gsea$objects,
    GSEA_simplified =
      proteomics_factor1_gsea_simplified_objects,
    ranked_lists =
      proteomics_factor1_gsea$ranked_lists
  ),
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_Factor1_GO_GSEA_objects.rds"
  )
)
```

```{r proteomics Factor1 simplified GO GSEA barplot, fig.height=4.8, fig.width=6.05, dpi=300}
top_n_positive <- 8
top_n_negative <- 8
gsea_fdr_cutoff <- 0.05


wrap_gsea_labels <- function(
    x,
    width = 41
) {
  vapply(
    x,
    function(label) {
      label <-
        stringr::str_squish(
          as.character(label)
        )

      if (
        nchar(label) <= width
      ) {
        return(label)
      }

      words <-
        stringr::str_split(
          label,
          "\\s+"
        )[[1]]

      lines <- c(
        "",
        "",
        ""
      )

      current_line <- 1L

      for (
        word in words
      ) {
        candidate <-
          if (
            lines[current_line] == ""
          ) {
            word
          } else {
            paste(
              lines[current_line],
              word
            )
          }

        if (
          nchar(candidate) <= width
        ) {
          lines[current_line] <-
            candidate
        } else if (
          current_line < 3L
        ) {
          current_line <-
            current_line + 1L

          lines[current_line] <-
            word
        } else {
          lines[current_line] <-
            paste0(
              substr(
                lines[current_line],
                1,
                width - 1
              ),
              "…"
            )

          break
        }
      }

      if (
        nchar(lines[3]) > width
      ) {
        lines[3] <-
          paste0(
            substr(
              lines[3],
              1,
              width - 1
            ),
            "…"
          )
      }

      paste(
        lines[
          lines != ""
        ],
        collapse = "\n"
      )
    },
    FUN.VALUE = character(1)
  )
}


proteomics_factor1_go_gsea_simplified <-
  proteomics_factor1_gsea_simplified_objects[[
    "Factor1__GO_ALL"
  ]]


if (
  is.null(
    proteomics_factor1_go_gsea_simplified
  ) ||
  nrow(
    as.data.frame(
      proteomics_factor1_go_gsea_simplified
    )
  ) == 0L
) {
  message(
    "The simplified Factor 1 GO GSEA result is empty."
  )
} else {
  proteomics_factor1_go_gsea_results <-
    as.data.frame(
      proteomics_factor1_go_gsea_simplified
    ) %>%
    tibble::as_tibble() %>%
    dplyr::filter(
      !is.na(NES),
      !is.na(p.adjust),
      p.adjust <
        gsea_fdr_cutoff
    )

  if (
    nrow(
      proteomics_factor1_go_gsea_results
    ) == 0L
  ) {
    message(
      "No significant terms with FDR < ",
      gsea_fdr_cutoff,
      "."
    )
  } else {
    positive_terms <-
      proteomics_factor1_go_gsea_results %>%
      dplyr::filter(
        NES > 0
      ) %>%
      dplyr::arrange(
        dplyr::desc(NES)
      ) %>%
      dplyr::slice_head(
        n = top_n_positive
      )

    negative_terms <-
      proteomics_factor1_go_gsea_results %>%
      dplyr::filter(
        NES < 0
      ) %>%
      dplyr::arrange(NES) %>%
      dplyr::slice_head(
        n = top_n_negative
      )

    proteomics_factor1_go_gsea_plot_data <-
      dplyr::bind_rows(
        positive_terms,
        negative_terms
      ) %>%
      dplyr::mutate(
        Description_clean =
          stringr::str_remove(
            Description,
            "^\\[(BP|CC|MF)\\]\\s*"
          ),
        p_for_colour =
          dplyr::if_else(
            "pvalue" %in%
              colnames(
                proteomics_factor1_go_gsea_results
              ) &
              is.finite(pvalue) &
              pvalue > 0,
            pvalue,
            p.adjust
          ),
        colour_score =
          -log10(
            pmax(
              p_for_colour,
              .Machine$double.xmin
            )
          ) *
          sign(NES),
        Description_wrapped =
          wrap_gsea_labels(
            Description_clean,
            width = 41
          )
      ) %>%
      dplyr::arrange(NES) %>%
      dplyr::mutate(
        Description_wrapped =
          factor(
            Description_wrapped,
            levels =
              unique(
                Description_wrapped
              )
          )
      )

    colour_limit <-
      unname(
        stats::quantile(
          abs(
            proteomics_factor1_go_gsea_plot_data$
              colour_score
          ),
          probs = 0.80,
          na.rm = TRUE
        )
      )

    if (
      !is.finite(colour_limit) ||
      colour_limit == 0
    ) {
      colour_limit <-
        max(
          abs(
            proteomics_factor1_go_gsea_plot_data$
              colour_score
          ),
          na.rm = TRUE
        )
    }

    proteomics_factor1_go_gsea_plot <-
      ggplot(
        proteomics_factor1_go_gsea_plot_data,
        aes(
          x =
            Description_wrapped,
          y = NES,
          fill =
            colour_score
        )
      ) +
      geom_col(
        width = 0.9
      ) +
      coord_flip() +
      scale_fill_gradient2(
        low = "#004C8C",
        mid = "white",
        high = "#A63A2A",
        midpoint = 0,
        limits = c(
          -colour_limit,
          colour_limit
        ),
        oob =
          scales::squish,
        name =
          expression(
            -log[10](p) %*% sign(NES)
          )
      ) +
      labs(
        title =
          "Proteomics | GSEA | Gene Ontology",
        x = NULL,
        y =
          "Normalized enrichment score (NES)"
      ) +
      theme_bw(
        base_size = 10
      ) +
      theme(
        text =
          element_text(
            color = "black"
          ),
        plot.title =
          element_text(
            hjust = 0.5,
            face = "bold",
            size = 11
          ),
        axis.text =
          element_text(
            color = "black"
          ),
        axis.title.x =
          element_text(
            color = "black",
            size = 12
          ),
        axis.text.y =
          element_text(
            size = 10.2,
            color = "black",
            lineheight = 0.6
          ),
        panel.grid.minor =
          element_blank(),
        legend.position =
          "bottom",
        legend.direction =
          "horizontal",
        legend.title =
          element_text(
            size = 9
          ),
        legend.text =
          element_text(
            size = 9
          )
      ) +
      guides(
        fill =
          guide_colorbar(
            barwidth =
              grid::unit(
                6,
                "cm"
              ),
            barheight =
              grid::unit(
                0.3,
                "cm"
              ),
            title.position =
              "top",
            title.hjust = 0.5
          )
      )

    proteomics_factor1_go_gsea_plot
  }
}
```

## Factor 1 core-enrichment features

```{r proteomics Factor1 selected GSEA core genes}
factor1_go_gsea <-
  proteomics_factor1_gsea$
    objects[[
      "Factor1__GO_ALL"
    ]]

if (is.null(factor1_go_gsea)) {
  stop(
    "The Factor1 GO GSEA object was not found."
  )
}

factor1_orientation <-
  proteomics_gsea_factor_table %>%
  dplyr::filter(
    Factor == "Factor1"
  ) %>%
  dplyr::pull(
    orientation
  )

if (
  length(factor1_orientation) != 1L
) {
  factor1_orientation <- 1
}

factor1_go_gsea_table <-
  as.data.frame(
    factor1_go_gsea
  ) %>%
  tibble::as_tibble() %>%
  dplyr::mutate(
    Description_clean =
      stringr::str_remove(
        Description,
        "^\\[(BP|CC|MF)\\]\\s*"
      ),
    Description_key =
      stringr::str_to_lower(
        Description_clean
      )
  )

target_gsea_terms <-
  factor1_go_gsea_table %>%
  dplyr::filter(
    Description_key %in%
      c(
        "immune response",
        "oxidative phosphorylation"
      )
  ) %>%
  dplyr::group_by(
    Description_key
  ) %>%
  dplyr::slice_min(
    order_by = p.adjust,
    n = 1,
    with_ties = FALSE
  ) %>%
  dplyr::ungroup()

if (
  nrow(target_gsea_terms) != 2L
) {
  stop(
    "One or both selected GSEA terms were not found."
  )
}

target_core_genes <-
  purrr::map_dfr(
    seq_len(
      nrow(
        target_gsea_terms
      )
    ),
    function(term_index) {
      core_entrez <-
        stringr::str_split(
          target_gsea_terms$
            core_enrichment[
              term_index
            ],
          pattern = "/"
        )[[1]]

      core_mapping <-
        bitr(
          core_entrez,
          fromType = "ENTREZID",
          toType = "SYMBOL",
          OrgDb = org.Hs.eg.db
        ) %>%
        dplyr::distinct(
          ENTREZID,
          SYMBOL
        )

      core_mapping %>%
        dplyr::transmute(
          GSEA_term =
            target_gsea_terms$
              Description_clean[
                term_index
              ],
          GO_ID =
            target_gsea_terms$
              ID[
                term_index
              ],
          NES =
            target_gsea_terms$
              NES[
                term_index
              ],
          GSEA_p_value =
            target_gsea_terms$
              pvalue[
                term_index
              ],
          GSEA_FDR =
            target_gsea_terms$
              p.adjust[
                term_index
              ],
          ENTREZID,
          Feature = SYMBOL
        )
    }
  )

factor1_proteomics_weights <-
  proteomics_factor_weights %>%
  dplyr::filter(
    Factor == "Factor1",
    View %in%
      c(
        "Olink",
        "LCMS"
      ),
    is.finite(Weight)
  ) %>%
  dplyr::mutate(
    Feature =
      clean_mofa_feature_name(
        Feature
      ),
    Modality =
      standardise_mofa_view_name(
        View
      ),
    Oriented_weight =
      Weight *
      factor1_orientation
  ) %>%
  dplyr::group_by(
    View
  ) %>%
  dplyr::mutate(
    Scaled_weight =
      Oriented_weight /
      max(
        abs(
          Oriented_weight
        ),
        na.rm = TRUE
      )
  ) %>%
  dplyr::ungroup()

factor1_core_feature_table <-
  target_core_genes %>%
  dplyr::inner_join(
    factor1_proteomics_weights,
    by = "Feature"
  ) %>%
  dplyr::mutate(
    Sign_consistent =
      sign(
        Scaled_weight
      ) ==
      sign(
        NES
      ),
    Directional_score =
      Scaled_weight *
      sign(
        NES
      )
  ) %>%
  dplyr::arrange(
    GSEA_term,
    dplyr::desc(
      Directional_score
    )
  )

factor1_core_top5_per_view <-
  factor1_core_feature_table %>%
  dplyr::filter(
    Sign_consistent
  ) %>%
  dplyr::group_by(
    GSEA_term,
    View
  ) %>%
  dplyr::slice_max(
    order_by =
      Directional_score,
    n = 5,
    with_ties = FALSE
  ) %>%
  dplyr::mutate(
    Rank_within_term_and_view =
      dplyr::row_number()
  ) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(
    GSEA_term,
    View,
    Rank_within_term_and_view
  )

write.csv(
  factor1_core_feature_table,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_Factor1_selected_GSEA_all_core_features.csv"
  ),
  row.names = FALSE
)

write.csv(
  factor1_core_top5_per_view,
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_Factor1_selected_GSEA_top5_per_view.csv"
  ),
  row.names = FALSE
)
```

```{r proteomics Factor1 selected GSEA heatmap, fig.height=6, fig.width=10}
selected_gsea_heatmap_features <-
  factor1_core_top5_per_view %>%
  dplyr::mutate(
    Pathway = dplyr::case_when(
      stringr::str_to_lower(GSEA_term) ==
        "immune response" ~
        "Immune response",

      stringr::str_to_lower(GSEA_term) ==
        "oxidative phosphorylation" ~
        "Oxidative phosphorylation",

      TRUE ~ GSEA_term
    ),

    Pathway = factor(
      Pathway,
      levels = c(
        "Immune response",
        "Oxidative phosphorylation"
      )
    ),

    View_order = factor(
      View,
      levels = c(
        "Olink",
        "LCMS"
      )
    )
  ) %>%
  dplyr::arrange(
    Pathway,
    View_order,
    Rank_within_term_and_view
  )



proteomics_gsea_heatmap_sample_metadata <-
  clin_all %>%
  dplyr::filter(
    Sample %in%
      intersect(
        colnames(
          proteomics_data$Olink
        ),
        colnames(
          proteomics_data$LCMS
        )
      )
  ) %>%
  dplyr::mutate(
    RV_group = factor(
      as.character(RV_group),
      levels = c(
        "Not overloaded",
        "Compensated",
        "Decompensated"
      )
    )
  ) %>%
  dplyr::arrange(
    is.na(TAPSE_mm),
    dplyr::desc(TAPSE_mm),
    Sample
  )


proteomics_gsea_heatmap_sample_order <-
  proteomics_gsea_heatmap_sample_metadata$
    Sample


extract_factor1_gsea_heatmap_matrix <- function(
    feature_table,
    data_views,
    sample_order
) {
  extracted_rows <- vector(
    mode = "list",
    length = nrow(feature_table)
  )

  row_found <- logical(
    nrow(feature_table)
  )

  resolved_features <- character(
    nrow(feature_table)
  )

  for (
    feature_index in
    seq_len(nrow(feature_table))
  ) {
    current_view <-
      as.character(
        feature_table$View[
          feature_index
        ]
      )

    current_feature <-
      as.character(
        feature_table$Feature[
          feature_index
        ]
      )

    if (
      !current_view %in%
        names(data_views)
    ) {
      next
    }

    feature_candidates <- unique(
      c(
        current_feature,

        paste0(
          current_feature,
          "_",
          current_view
        ),

        paste0(
          current_feature,
          "_",
          dplyr::case_when(
            current_view == "Olink" ~
              "OLINK",

            current_view == "LCMS" ~
              "MS",

            TRUE ~
              current_view
          )
        )
      )
    )

    resolved_feature <-
      feature_candidates[
        feature_candidates %in%
          rownames(
            data_views[[
              current_view
            ]]
          )
      ][1]

    if (
      length(resolved_feature) == 0L ||
      is.na(resolved_feature)
    ) {
      next
    }

    extracted_rows[[
      feature_index
    ]] <-
      as.numeric(
        data_views[[
          current_view
        ]][
          resolved_feature,
          sample_order,
          drop = TRUE
        ]
      )

    resolved_features[
      feature_index
    ] <- resolved_feature

    row_found[
      feature_index
    ] <- TRUE
  }

  if (!any(row_found)) {
    stop(
      "None of the selected features were found in the proteomics matrices."
    )
  }

  if (any(!row_found)) {
    warning(
      "The following selected features were not found and were omitted: ",
      paste(
        paste0(
          feature_table$Feature[
            !row_found
          ],
          " (",
          feature_table$View[
            !row_found
          ],
          ")"
        ),
        collapse = ", "
      )
    )
  }

  retained_feature_table <-
    feature_table[
      row_found,
      ,
      drop = FALSE
    ] %>%
    dplyr::mutate(
      Resolved_feature =
        resolved_features[
          row_found
        ],

      Display_feature =
        Feature,

      Modality =
        dplyr::case_when(
          View == "Olink" ~
            "OLINK",

          View == "LCMS" ~
            "MS",

          TRUE ~
            as.character(View)
        ),

      Internal_row_id =
        make.unique(
          paste(
            Pathway,
            View,
            Feature,
            sep = "__"
          )
        )
    )

  expression_matrix <-
    do.call(
      rbind,
      extracted_rows[
        row_found
      ]
    )

  colnames(expression_matrix) <-
    sample_order

  rownames(expression_matrix) <-
    retained_feature_table$
      Internal_row_id

  list(
    matrix =
      expression_matrix,

    feature_table =
      retained_feature_table
  )
}


prepare_factor1_gsea_heatmap_z_scores <- function(
    expression_matrix
) {
  imputed_matrix <-
    expression_matrix

  for (
    row_index in
    seq_len(nrow(imputed_matrix))
  ) {
    row_values <-
      imputed_matrix[
        row_index,
        ,
        drop = TRUE
      ]

    finite_values <-
      row_values[
        is.finite(row_values)
      ]

    if (
      length(finite_values) == 0L
    ) {
      next
    }

    row_values[
      !is.finite(row_values)
    ] <-
      stats::median(
        finite_values
      )

    imputed_matrix[
      row_index,
    ] <- row_values
  }

  row_means <-
    rowMeans(
      imputed_matrix,
      na.rm = TRUE
    )

  row_standard_deviations <-
    apply(
      imputed_matrix,
      1,
      stats::sd,
      na.rm = TRUE
    )

  row_standard_deviations[
    !is.finite(
      row_standard_deviations
    ) |
      row_standard_deviations == 0
  ] <- 1

  z_score_matrix <-
    sweep(
      imputed_matrix,
      1,
      row_means,
      "-"
    )

  z_score_matrix <-
    sweep(
      z_score_matrix,
      1,
      row_standard_deviations,
      "/"
    )

  pmax(
    pmin(
      z_score_matrix,
      2
    ),
    -2
  )
}


make_grey_annotation_scale <- function(
    values,
    dark_at = c(
      "low",
      "high"
    )
) {
  dark_at <-
    match.arg(
      dark_at
    )

  finite_values <-
    values[
      is.finite(values)
    ]

  value_range <-
    if (
      length(finite_values) == 0L
    ) {
      c(0, 1)
    } else {
      range(
        finite_values,
        na.rm = TRUE
      )
    }

  if (
    value_range[1] ==
      value_range[2]
  ) {
    value_range <-
      value_range +
      c(
        -0.5,
        0.5
      )
  }

  annotation_colours <-
    if (
      dark_at == "low"
    ) {
      c(
        "grey10",
        "grey95"
      )
    } else {
      c(
        "grey95",
        "grey10"
      )
    }

  circlize::colorRamp2(
    value_range,
    annotation_colours
  )
}


selected_gsea_heatmap_data <-
  extract_factor1_gsea_heatmap_matrix(
    feature_table =
      selected_gsea_heatmap_features,

    data_views =
      proteomics_data,

    sample_order =
      proteomics_gsea_heatmap_sample_order
  )


selected_gsea_heatmap_matrix <-
  prepare_factor1_gsea_heatmap_z_scores(
    selected_gsea_heatmap_data$
      matrix
  )


selected_gsea_feature_metadata <-
  selected_gsea_heatmap_data$
    feature_table


tapse_colour_function <-
  make_grey_annotation_scale(
    proteomics_gsea_heatmap_sample_metadata$
      TAPSE_mm,

    dark_at = "low"
  )


rvit_colour_function <-
  make_grey_annotation_scale(
    proteomics_gsea_heatmap_sample_metadata$
      RVIT_mm,

    dark_at = "high"
  )


tapse_spap_colour_function <-
  make_grey_annotation_scale(
    proteomics_gsea_heatmap_sample_metadata$
      TAPSE_sPAP,

    dark_at = "low"
  )


rv_group_colors <- c(
  "Not overloaded" =
    unname(
      group_colors[
        "Not overloaded"
      ]
    ),

  "Compensated" =
    unname(
      group_colors[
        "Compensated"
      ]
    ),

  "Decompensated" =
    unname(
      group_colors[
        "Decompensated"
      ]
    )
)



selected_gsea_top_annotation <-
  ComplexHeatmap::HeatmapAnnotation(
    `TAPSE (mm)` =
      ComplexHeatmap::anno_simple(
        proteomics_gsea_heatmap_sample_metadata$
          TAPSE_mm,

        col =
          tapse_colour_function,

        na_col =
          "white",

        height =
          grid::unit(
            4.5,
            "mm"
          )
      ),

    `RVIT (mm)` =
      ComplexHeatmap::anno_simple(
        proteomics_gsea_heatmap_sample_metadata$
          RVIT_mm,

        col =
          rvit_colour_function,

        na_col =
          "white",

        height =
          grid::unit(
            4.5,
            "mm"
          )
      ),

    `TAPSE/sPAP` =
      ComplexHeatmap::anno_simple(
        proteomics_gsea_heatmap_sample_metadata$
          TAPSE_sPAP,

        col =
          tapse_spap_colour_function,

        na_col =
          "white",

        height =
          grid::unit(
            4.5,
            "mm"
          )
      ),

    `RV state` =
      ComplexHeatmap::anno_simple(
        as.character(
          proteomics_gsea_heatmap_sample_metadata$
            RV_group
        ),

        col =
          rv_group_colors,

        na_col =
          "white",

        height =
          grid::unit(
            4.5,
            "mm"
          )
      ),

    annotation_name_gp =
      grid::gpar(
        fontsize = 11,
        fontface = "bold",
        col = "black"
      ),

    annotation_name_side =
      "right",

    gap =
      grid::unit(
        1,
        "mm"
      )
  )


pathway_colors <- c(
  "Immune response" =
    "#6A3D9A",

  "Oxidative phosphorylation" =
    "#238B45"
)


pathway_slice_order <-
  unique(
    as.character(
      selected_gsea_feature_metadata$
        Pathway
    )
  )


pathway_display_labels <- c(
  "Immune response" =
    "Immune\nresponse",

  "Oxidative phosphorylation" =
    "Oxidative\nphosphorylation"
)


selected_gsea_left_annotation <-
  ComplexHeatmap::rowAnnotation(
    Pathway =
      ComplexHeatmap::anno_block(
        gp =
          grid::gpar(
            fill =
              unname(
                pathway_colors[
                  pathway_slice_order
                ]
              ),

            col = NA
          ),

        labels =
          unname(
            pathway_display_labels[
              pathway_slice_order
            ]
          ),

        labels_gp =
          grid::gpar(
            fontsize = 9,
            fontface = "bold",
            col = "white",
            lineheight = 0.9
          ),

        labels_rot = 90,

        width =
          grid::unit(
            8,
            "mm"
          )
      ),

    show_annotation_name =
      FALSE
  )


modality_colors <- c(
  "RNA" =
    "#009E73",

  "OLINK" =
    "#E69F00",

  "MS" =
    "#7B61A8"
)


selected_gsea_feature_metadata <-
  selected_gsea_feature_metadata %>%
  dplyr::mutate(
    Modality = factor(
      Modality,
      levels = c(
        "RNA",
        "OLINK",
        "MS"
      )
    )
  )


selected_gsea_right_annotation <-
  ComplexHeatmap::rowAnnotation(
    Modality =
      selected_gsea_feature_metadata$
        Modality,

    col = list(
      Modality =
        modality_colors
    ),

    show_legend =
      FALSE,

    annotation_name_gp =
      grid::gpar(
        fontsize = 10,
        fontface = "bold",
        col = "black"
      ),

    simple_anno_size =
      grid::unit(
        4.5,
        "mm"
      )
  )


heatmap_colour_function <-
  circlize::colorRamp2(
    c(
      -2,
      0,
      2
    ),
    c(
      "#2166AC",
      "white",
      "#B2182B"
    )
  )


selected_gsea_heatmap <-
  ComplexHeatmap::Heatmap(
    selected_gsea_heatmap_matrix,

    name =
      "Row Z-score",

    col =
      heatmap_colour_function,

    na_col =
      "grey90",

    show_heatmap_legend =
      FALSE,

    cluster_rows =
      FALSE,

    cluster_columns =
      FALSE,

    row_split =
      selected_gsea_feature_metadata$
        Pathway,

    row_gap =
      grid::unit(
        2,
        "mm"
      ),

    cluster_row_slices =
      FALSE,

    top_annotation =
      selected_gsea_top_annotation,

    left_annotation =
      selected_gsea_left_annotation,

    right_annotation =
      selected_gsea_right_annotation,

    row_labels =
      selected_gsea_feature_metadata$
        Display_feature,

    row_names_side =
      "left",

    row_names_gp =
      grid::gpar(
        fontsize = 10,
        col = "black"
      ),

    column_names_rot =
      45,

    column_names_gp =
      grid::gpar(
        fontsize = 7,
        col = "black"
      ),

    show_row_dend =
      FALSE,

    show_column_dend =
      FALSE,

    row_title =
      NULL,

    column_title =
      NULL,

    border =
      TRUE,

    height =
      grid::unit(
        nrow(
          selected_gsea_heatmap_matrix
        ) * 4,
        "mm"
      )
  )



row_z_score_legend <-
  ComplexHeatmap::Legend(
    title =
      "Row Z-score",

    col_fun =
      heatmap_colour_function,

    at = c(
      -2,
      0,
      2
    )
  )


modality_legend <-
  ComplexHeatmap::Legend(
    title =
      "Modality",

    at = c(
      "RNA",
      "OLINK",
      "MS"
    ),

    legend_gp =
      grid::gpar(
        fill =
          unname(
            modality_colors[
              c(
                "RNA",
                "OLINK",
                "MS"
              )
            ]
          )
      )
  )


rv_state_legend <-
  ComplexHeatmap::Legend(
    title =
      "RV state",

    at =
      names(
        rv_group_colors
      ),

    legend_gp =
      grid::gpar(
        fill =
          unname(
            rv_group_colors
          )
      )
  )


legend_column_1 <-
  ComplexHeatmap::packLegend(
    row_z_score_legend,
    modality_legend,
    rv_state_legend,

    direction =
      "vertical",

    gap =
      grid::unit(
        4,
        "mm"
      )
  )



tapse_legend <-
  ComplexHeatmap::Legend(
    title =
      "TAPSE (mm)",

    col_fun =
      tapse_colour_function
  )


rvit_legend <-
  ComplexHeatmap::Legend(
    title =
      "RVIT (mm)",

    col_fun =
      rvit_colour_function
  )


tapse_spap_legend <-
  ComplexHeatmap::Legend(
    title =
      "TAPSE/sPAP",

    col_fun =
      tapse_spap_colour_function
  )


legend_column_2 <-
  ComplexHeatmap::packLegend(
    tapse_legend,
    rvit_legend,
    tapse_spap_legend,

    direction =
      "vertical",

    gap =
      grid::unit(
        4,
        "mm"
      )
  )


selected_gsea_combined_legends <-
  ComplexHeatmap::packLegend(
    legend_column_1,
    legend_column_2,

    direction =
      "horizontal",

    gap =
      grid::unit(
        8,
        "mm"
      )
  )


ComplexHeatmap::draw(
  selected_gsea_heatmap,

  heatmap_legend_side =
    "right",

  annotation_legend_side =
    "right",

  heatmap_legend_list =
    list(
      selected_gsea_combined_legends
    ),

  merge_legends =
    FALSE
)




write.csv(
  selected_gsea_feature_metadata,

  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_Factor1_selected_GSEA_top5_per_view_heatmap_features.csv"
  ),

  row.names = FALSE
)


write.csv(
  proteomics_gsea_heatmap_sample_metadata,

  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_Factor1_selected_GSEA_top5_per_view_heatmap_sample_order.csv"
  ),

  row.names = FALSE
)
```

# Save the complete proteomics analysis

```{r save complete proteomics analysis}
saveRDS(
  list(
    model =
      proteomics_model,
    model_file =
      proteomics_final_model_file,
    input_file =
      proteomics_input_file,
    metadata =
      proteomics_model_metadata,
    selected_model_settings =
      proteomics_selected_model_settings,
    factor_tuning_results =
      proteomics_tuning_results,
    best_tuning_runs =
      proteomics_best_tuning_runs,
    factor_scores =
      proteomics_factor_scores,
    factor_weights =
      proteomics_factor_weights,
    factor1_top_loadings =
      proteomics_factor1_top_loadings,
    variance_explained =
      proteomics_r2_per_factor_table,
    group_results =
      proteomics_group_global_results,
    group_pairwise_results =
      proteomics_group_pairwise_results,
    continuous_results =
      proteomics_continuous_results,
    technical_results =
      proteomics_technical_results,
    factor_summary =
      proteomics_factor_summary,
    factor1_gsea = list(
      results =
        proteomics_factor1_gsea_results,
      simplified_results =
        proteomics_factor1_gsea_simplified_results,
      objects =
        proteomics_factor1_gsea$objects,
      simplified_objects =
        proteomics_factor1_gsea_simplified_objects,
      ranked_lists =
        proteomics_factor1_gsea$ranked_lists,
      core_features =
        factor1_core_feature_table,
      selected_heatmap_features =
        factor1_core_top5_per_view
    )
  ),
  file.path(
    proteomics_output_dir,
    "MOFA_proteomics_complete_analysis_objects.rds"
  )
)
```

# Preprocess pseudobulk snRNA-seq

The RNA view is needed only for the matched three-view model. Genes are retained when detected in at least 1% of nuclei from the DCM samples, pseudobulk counts are aggregated per patient, DESeq2 VST is applied, and the 2,000 most variable transformed genes are selected.

## Read the Seurat object

```{r read snrna object}
DCM_RV <- qs2::qs_read(
  seurat_file,
  nthreads = n_qs_threads
)

DefaultAssay(DCM_RV) <- "RNA"
DCM_RV$pseudobulk <- "Pseudobulk"
```

## Filter genes using DCM nuclei

```{r rna nucleus level gene filtering}
dcm_cells <- colnames(DCM_RV)[
  DCM_RV$Sample %in%
    clin_all$Sample
]

rna_counts <- SeuratObject::LayerData(
  DCM_RV,
  assay = "RNA",
  layer = "counts"
)[
  ,
  dcm_cells,
  drop = FALSE
]

n_rna_nuclei <- ncol(rna_counts)

minimum_nuclei <- ceiling(
  rna_min_nuclei_fraction *
    n_rna_nuclei
)

rna_nuclei_detected <-
  Matrix::rowSums(
    rna_counts > 0
  )

rna_detection_table <-
  tibble::tibble(
    feature = rownames(rna_counts),
    n_nuclei_detected =
      as.numeric(
        rna_nuclei_detected
      ),
    fraction_nuclei_detected =
      as.numeric(
        rna_nuclei_detected
      ) /
      n_rna_nuclei,
    passes_1_percent_filter =
      n_nuclei_detected >=
      minimum_nuclei
  )

if (
  exclude_mitochondrial_and_ribosomal_genes
) {
  rna_detection_table <-
    rna_detection_table %>%
    dplyr::mutate(
      mitochondrial_or_ribosomal =
        grepl(
          "^MT-|^RP[SL]",
          feature
        ),
      retained_before_VST =
        passes_1_percent_filter &
        !mitochondrial_or_ribosomal
    )
} else {
  rna_detection_table <-
    rna_detection_table %>%
    dplyr::mutate(
      mitochondrial_or_ribosomal =
        FALSE,
      retained_before_VST =
        passes_1_percent_filter
    )
}

rna_genes_for_pseudobulk <-
  rna_detection_table %>%
  dplyr::filter(
    retained_before_VST
  ) %>%
  dplyr::pull(feature)

write.csv(
  rna_detection_table,
  file.path(
    preprocessing_output_dir,
    "RNA_nucleus_detection_filter.csv"
  ),
  row.names = FALSE
)

c(
  nuclei = n_rna_nuclei,
  minimum_nuclei = minimum_nuclei,
  retained_genes =
    length(
      rna_genes_for_pseudobulk
    )
)
```

## Generate sample-level pseudobulk counts

```{r create rna pseudobulk}
rna_pb <- edgeR::Seurat2PB(
  DCM_RV,
  sample = "Sample",
  cluster = "pseudobulk"
)

colnames(rna_pb) <-
  rna_pb$samples$sample

rna_samples <- intersect(
  clin_all$Sample,
  colnames(rna_pb)
)

rna_pb <- rna_pb[
  ,
  rna_samples,
  keep.lib.sizes = FALSE
]

keep_rna_samples <-
  rna_pb$samples$lib.size >
  min_rna_library_size

rna_pb <- rna_pb[
  ,
  keep_rna_samples,
  keep.lib.sizes = FALSE
]

rna_genes_for_pseudobulk <-
  intersect(
    rna_genes_for_pseudobulk,
    rownames(rna_pb)
  )

rna_pb <- rna_pb[
  rna_genes_for_pseudobulk,
  ,
  keep.lib.sizes = FALSE
]

rm(rna_counts)
gc()

dim(rna_pb)
```

## Apply VST and select 2,000 variable genes

```{r vst transform rna pseudobulk}
rna_count_matrix <-
  as.matrix(
    rna_pb$counts
  )

storage.mode(
  rna_count_matrix
) <- "integer"

rna_coldata <- data.frame(
  Sample =
    colnames(
      rna_count_matrix
    ),
  row.names =
    colnames(
      rna_count_matrix
    )
)

rna_dds <-
  DESeq2::DESeqDataSetFromMatrix(
    countData =
      rna_count_matrix,
    colData =
      rna_coldata,
    design = ~ 1
  )

rna_vst_object <-
  DESeq2::vst(
    rna_dds,
    blind = TRUE
  )

rna_vst_matrix <-
  SummarizedExperiment::assay(
    rna_vst_object
  )

rna_variance <-
  matrixStats::rowVars(
    rna_vst_matrix
  )

keep_finite_rna_variance <-
  is.finite(rna_variance) &
  rna_variance > 0

rna_vst_matrix <-
  rna_vst_matrix[
    keep_finite_rna_variance,
    ,
    drop = FALSE
  ]

rna_variance <-
  rna_variance[
    keep_finite_rna_variance
  ]

rna_feature_order <- order(
  rna_variance,
  decreasing = TRUE
)

rna_feature_order <- head(
  rna_feature_order,
  min(
    n_rna_features,
    length(
      rna_feature_order
    )
  )
)

rna_mofa <-
  rna_vst_matrix[
    rna_feature_order,
    ,
    drop = FALSE
  ]

rna_feature_table <-
  tibble::tibble(
    feature =
      rownames(
        rna_vst_matrix
      ),
    VST_variance =
      as.numeric(
        rna_variance
      ),
    selected_for_MOFA =
      rownames(
        rna_vst_matrix
      ) %in%
      rownames(rna_mofa)
  ) %>%
  dplyr::arrange(
    dplyr::desc(
      VST_variance
    )
  )

write.csv(
  rna_feature_table,
  file.path(
    preprocessing_output_dir,
    "RNA_VST_feature_selection.csv"
  ),
  row.names = FALSE
)

saveRDS(
  list(
    RNA_pseudobulk = rna_mofa,
    RNA_VST_all_filtered_genes =
      rna_vst_matrix,
    feature_table =
      rna_feature_table
  ),
  processed_rna_file
)

dim(rna_mofa)
```

# Model 2: matched RNA, Olink and LC-MS/MS

Only patients measured in all three modalities are included. This model is used to evaluate whether major factors are shared between transcriptomic and proteomic layers.

## Assemble the matched three-view input

```{r assemble matched three view input}
matched_samples <- Reduce(
  intersect,
  list(
    colnames(rna_mofa),
    colnames(olink_mofa),
    colnames(ms_mofa)
  )
)

matched_samples <- clin_all$Sample[
  clin_all$Sample %in%
    matched_samples
]

matched_data <- list(
  RNA_pseudobulk = rna_mofa[
    ,
    matched_samples,
    drop = FALSE
  ],
  Olink = olink_mofa[
    ,
    matched_samples,
    drop = FALSE
  ],
  LCMS = ms_mofa[
    ,
    matched_samples,
    drop = FALSE
  ]
)

length(matched_samples)

lapply(
  matched_data,
  dim
)
```

```{r matched sample overview}
matched_sample_overview <-
  tibble::tibble(
    Sample = matched_samples,
    RNA_pseudobulk = TRUE,
    Olink = TRUE,
    LCMS = TRUE
  ) %>%
  dplyr::left_join(
    clin_all %>%
      dplyr::select(
        Sample,
        RV_group,
        mPAP_mmHg,
        TAPSE_mm,
        RVIT_mm,
        TAPSE_sPAP
      ),
    by = "Sample"
  )

write.csv(
  matched_sample_overview,
  file.path(
    matched_output_dir,
    "MOFA_matched_three_view_samples.csv"
  ),
  row.names = FALSE
)

matched_sample_overview %>%
  dplyr::count(
    RV_group,
    .drop = FALSE
  )
```

```{r save matched input checkpoint}
saveRDS(
  matched_data,
  matched_input_file
)
```

## Checkpoint: load the matched input

```{r load matched input checkpoint}
matched_data <- readRDS(
  matched_input_file
)

lapply(
  matched_data,
  dim
)
```

## Create the untrained matched model

```{r create matched mofa object}
matched_untrained <-
  MOFA2::create_mofa(
    matched_data
  )

matched_untrained
```

```{r matched input overview, fig.height=4.5, fig.width=8}
MOFA2::plot_data_overview(
  matched_untrained
)
```

## Tune the number of matched-model factors

```{r tune matched factor number, message=FALSE, warning=FALSE, results='hide'}
matched_tuning_results <-
  tibble::tibble()

for (n_factors in candidate_factors) {
  for (tuning_seed in tuning_seeds) {
    tuning_model_file <- file.path(
      matched_output_dir,
      paste0(
        "MOFA_matched_tuning_K",
        sprintf("%02d", n_factors),
        "_seed",
        tuning_seed,
        ".hdf5"
      )
    )

    if (file.exists(tuning_model_file)) {
      unlink(tuning_model_file)
    }

    message(
      "Starting matched model: K = ",
      n_factors,
      ", seed = ",
      tuning_seed
    )

    tuning_object <- MOFA2::create_mofa(
      matched_data
    )

    tuning_data_options <-
      MOFA2::get_default_data_options(
        tuning_object
      )

    tuning_data_options$scale_views <- TRUE

    tuning_model_options <-
      MOFA2::get_default_model_options(
        tuning_object
      )

    tuning_model_options$num_factors <-
      n_factors

    tuning_model_options$likelihoods[] <-
      "gaussian"

    tuning_training_options <-
      MOFA2::get_default_training_options(
        tuning_object
      )

    tuning_training_options$maxiter <-
      tuning_maxiter

    tuning_training_options$convergence_mode <-
      mofa_convergence_mode

    tuning_training_options$freqELBO <-
      mofa_freq_elbo

    tuning_training_options$drop_factor_threshold <-
      -1

    tuning_training_options$seed <-
      tuning_seed

    tuning_training_options$verbose <-
      FALSE

    tuning_object <- MOFA2::prepare_mofa(
      object = tuning_object,
      data_options =
        tuning_data_options,
      model_options =
        tuning_model_options,
      training_options =
        tuning_training_options
    )

    MOFA2::run_mofa(
      tuning_object,
      outfile = tuning_model_file,
      use_basilisk = TRUE
    )

    rm(tuning_object)
    gc()

    tuning_model <- MOFA2::load_model(
      tuning_model_file,
      sort_factors = TRUE,
      remove_inactive_factors = FALSE
    )

    tuning_r2 <-
      MOFA2::get_variance_explained(
        tuning_model
      )$r2_per_factor[[1]]

    weakest_factor <-
      tail(
        rownames(tuning_r2),
        1
      )

    weakest_factor_r2 <-
      tuning_r2[
        weakest_factor,
        ,
        drop = TRUE
      ]

    new_tuning_result <- tibble::tibble(
      n_factors = n_factors,
      seed = tuning_seed,
      final_ELBO =
        MOFA2::get_elbo(tuning_model),
      weakest_factor = weakest_factor,
      weakest_factor_total_R2 =
        sum(
          weakest_factor_r2,
          na.rm = TRUE
        ),
      weakest_factor_RNA_R2 =
        unname(
          weakest_factor_r2[
            "RNA_pseudobulk"
          ]
        ),
      weakest_factor_Olink_R2 =
        unname(
          weakest_factor_r2[
            "Olink"
          ]
        ),
      weakest_factor_LCMS_R2 =
        unname(
          weakest_factor_r2[
            "LCMS"
          ]
        ),
      n_factors_total_R2_at_least_cutoff =
        sum(
          rowSums(
            tuning_r2,
            na.rm = TRUE
          ) >= last_factor_r2_cutoff
        ),
      model_file = tuning_model_file
    )

    matched_tuning_results <-
      dplyr::bind_rows(
        matched_tuning_results,
        new_tuning_result
      ) %>%
      dplyr::arrange(
        n_factors,
        seed
      )

    write.csv(
      matched_tuning_results,
      matched_tuning_summary_file,
      row.names = FALSE
    )

    message(
      "Finished matched model: K = ",
      n_factors,
      ", seed = ",
      tuning_seed
    )

    rm(tuning_model)
    gc()
  }
}

matched_tuning_results
```

## Select the matched-model factor number

```{r select matched best tuning run per K}
matched_best_tuning_runs <-
  matched_tuning_results %>%
  dplyr::group_by(n_factors) %>%
  dplyr::slice_max(
    order_by = final_ELBO,
    n = 1,
    with_ties = FALSE
  ) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(n_factors)

write.csv(
  matched_best_tuning_runs,
  matched_best_tuning_summary_file,
  row.names = FALSE
)

matched_low_weakest_factor <-
  matched_best_tuning_runs$
    weakest_factor_total_R2 <
  last_factor_r2_cutoff

matched_low_runs <- vapply(
  seq_along(
    matched_low_weakest_factor
  ),
  function(i) {
    end_i <-
      i +
      n_consecutive_low_models -
      1

    end_i <=
      length(
        matched_low_weakest_factor
      ) &&
      all(
        matched_low_weakest_factor[
          i:end_i
        ]
      )
  },
  logical(1)
)

matched_first_low_run <-
  which(matched_low_runs)[1]

if (
  length(matched_first_low_run) == 1 &&
  !is.na(matched_first_low_run)
) {
  matched_selected_n_factors <- max(
    matched_best_tuning_runs$
      n_factors[
        matched_first_low_run
      ] - 1,
    min(candidate_factors)
  )

  matched_factor_plateau_reached <-
    TRUE
} else {
  matched_selected_n_factors <-
    max(candidate_factors)

  matched_factor_plateau_reached <-
    FALSE

  warning(
    paste0(
      "No sequence of ",
      n_consecutive_low_models,
      " consecutive matched models with weakest-factor variance below ",
      last_factor_r2_cutoff,
      "% was observed. The maximum tested factor number was selected."
    )
  )
}

matched_selected_seed <-
  matched_best_tuning_runs %>%
  dplyr::filter(
    n_factors ==
      matched_selected_n_factors
  ) %>%
  dplyr::pull(seed)

matched_selected_model_settings <-
  tibble::tibble(
    selected_n_factors =
      matched_selected_n_factors,
    selected_seed =
      matched_selected_seed,
    factor_plateau_reached =
      matched_factor_plateau_reached,
    weakest_factor_R2_cutoff =
      last_factor_r2_cutoff,
    consecutive_low_models_required =
      n_consecutive_low_models,
    tuning_maxiter = tuning_maxiter,
    final_maxiter = final_maxiter,
    convergence_mode =
      mofa_convergence_mode
  )

write.csv(
  matched_selected_model_settings,
  matched_selected_settings_file,
  row.names = FALSE
)

matched_selected_model_settings
```

```{r matched factor tuning plot, fig.height=3.3, fig.width=4.4}
ggplot() +
  geom_line(
    data = matched_tuning_results,
    aes(
      x = n_factors,
      y = weakest_factor_total_R2,
      group = factor(seed)
    ),
    color = "grey70"
  ) +
  geom_point(
    data = matched_tuning_results,
    aes(
      x = n_factors,
      y = weakest_factor_total_R2
    ),
    color = "grey60",
    size = 1.8
  ) +
  geom_line(
    data = matched_best_tuning_runs,
    aes(
      x = n_factors,
      y = weakest_factor_total_R2
    ),
    linewidth = 0.8
  ) +
  geom_point(
    data = matched_best_tuning_runs,
    aes(
      x = n_factors,
      y = weakest_factor_total_R2
    ),
    size = 2.7
  ) +
  geom_hline(
    yintercept =
      last_factor_r2_cutoff,
    linetype = 2
  ) +
  geom_vline(
    xintercept =
      matched_selected_n_factors,
    linetype = 3
  ) +
  scale_x_continuous(
    breaks = candidate_factors
  ) +
  labs(
    x = "Number of factors in fitted model",
    y = "Variance explained by weakest factor (%)"
  ) +
  theme_classic()
```

## Train the final matched three-view model

```{r train final matched mofa model, message=FALSE, warning=FALSE, results='hide'}
matched_data <- readRDS(
  matched_input_file
)

if (file.exists(matched_final_model_file)) {
  unlink(matched_final_model_file)
}

matched_object <-
  MOFA2::create_mofa(
    matched_data
  )

matched_data_options <-
  MOFA2::get_default_data_options(
    matched_object
  )

matched_data_options$scale_views <-
  TRUE

matched_model_options <-
  MOFA2::get_default_model_options(
    matched_object
  )

matched_model_options$num_factors <-
  matched_selected_n_factors

matched_model_options$likelihoods[] <-
  "gaussian"

matched_training_options <-
  MOFA2::get_default_training_options(
    matched_object
  )

matched_training_options$maxiter <-
  final_maxiter

matched_training_options$convergence_mode <-
  mofa_convergence_mode

matched_training_options$freqELBO <-
  mofa_freq_elbo

matched_training_options$drop_factor_threshold <-
  -1

matched_training_options$seed <-
  matched_selected_seed

matched_training_options$verbose <-
  FALSE

matched_object <- MOFA2::prepare_mofa(
  object = matched_object,
  data_options =
    matched_data_options,
  model_options =
    matched_model_options,
  training_options =
    matched_training_options
)

MOFA2::run_mofa(
  matched_object,
  outfile =
    matched_final_model_file,
  use_basilisk = TRUE
)

rm(matched_object)
gc()
```

## Load the matched model and attach metadata

```{r load matched model and attach metadata}
matched_model <- MOFA2::load_model(
  matched_final_model_file,
  sort_factors = TRUE,
  remove_inactive_factors = FALSE
)

matched_model_samples <-
  MOFA2::samples_names(
    matched_model
  )[[1]]

matched_model_metadata <-
  clin_all %>%
  dplyr::left_join(
    olink_sample_meta %>%
      dplyr::select(
        Sample,
        Batch,
        BCA_mg_ml
      ),
    by = "Sample"
  ) %>%
  dplyr::filter(
    Sample %in%
      matched_model_samples
  ) %>%
  dplyr::slice(
    match(
      matched_model_samples,
      Sample
    )
  ) %>%
  dplyr::rename(
    sample = Sample
  )

samples_metadata(matched_model) <-
  as.data.frame(
    matched_model_metadata
  )

saveRDS(
  matched_model,
  matched_final_model_rds
)

write.csv(
  matched_model_metadata,
  file.path(
    matched_output_dir,
    "MOFA_matched_sample_metadata.csv"
  ),
  row.names = FALSE
)
```

## Checkpoint: reload the final matched model

```{r reload final matched model checkpoint}
matched_model <- readRDS(
  matched_final_model_rds
)

matched_model_metadata <-
  samples_metadata(
    matched_model
  )

matched_selected_model_settings <-
  read.csv(
    matched_selected_settings_file,
    stringsAsFactors = FALSE
  )

matched_tuning_results <-
  read.csv(
    matched_tuning_summary_file,
    stringsAsFactors = FALSE
  )

matched_best_tuning_runs <-
  read.csv(
    matched_best_tuning_summary_file,
    stringsAsFactors = FALSE
  )
```

# Matched-model variance decomposition

```{r matched trained model overview, fig.height=4.5, fig.width=8}
MOFA2::plot_data_overview(
  matched_model
)
```

```{r matched variance explained, fig.height=2.6, fig.width=5}
matched_r2_heatmap_matrix <-
  MOFA2::get_variance_explained(
    matched_model
  )$r2_per_factor[[1]]

matched_r2_heatmap_data <-
  as.data.frame(
    matched_r2_heatmap_matrix
  ) %>%
  tibble::rownames_to_column(
    "Factor"
  ) %>%
  tidyr::pivot_longer(
    cols = -Factor,
    names_to = "View",
    values_to = "R2"
  ) %>%
  dplyr::mutate(
    Factor = factor(
      Factor,
      levels = rownames(
        matched_r2_heatmap_matrix
      )
    ),
    View = factor(
      View,
      levels = colnames(
        matched_r2_heatmap_matrix
      )
    ),
    text_color = dplyr::if_else(
      R2 >=
        0.55 *
        max(
          R2,
          na.rm = TRUE
        ),
      "white",
      "black"
    )
  )

ggplot(
  matched_r2_heatmap_data,
  aes(
    x = View,
    y = Factor,
    fill = R2
  )
) +
  geom_tile(
    color = "grey40",
    linewidth = 0.3
  ) +
  geom_text(
    aes(
      label = sprintf(
        "%.1f%%",
        R2
      ),
      color = text_color
    ),
    size = 3.2
  ) +
  scale_color_identity() +
  scale_fill_gradient(
    low = "white",
    high = "navy"
  ) +
  labs(
    title =
      "",
    x = NULL,
    y = NULL,
    fill = "Variance\nexplained (%)"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    ),
  axis.text.x = element_text(
    size = 11
  ),
  axis.text.y = element_text(
    size = 11
  )
  )
```

```{r matched factor correlations, fig.height=3.5, fig.width=4}
MOFA2::plot_factor_cor(
  matched_model
) +
  labs(
    title =
      "Matched multi-omics factor correlations"
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold"
    )
  )
```

```{r extract matched variance explained}
matched_variance_explained <-
  MOFA2::get_variance_explained(
    matched_model
  )

matched_r2_total <-
  matched_variance_explained$
    r2_total[[1]]

matched_r2_per_factor <-
  matched_variance_explained$
    r2_per_factor[[1]]

matched_r2_total_table <-
  tibble::enframe(
    matched_r2_total,
    name = "View",
    value = "Total_R2"
  )

matched_r2_per_factor_table <-
  as.data.frame(
    matched_r2_per_factor
  ) %>%
  tibble::rownames_to_column(
    "Factor"
  ) %>%
  tidyr::pivot_longer(
    cols = -Factor,
    names_to = "View",
    values_to = "R2"
  )

write.csv(
  matched_r2_total_table,
  file.path(
    matched_output_dir,
    "MOFA_matched_total_variance_explained.csv"
  ),
  row.names = FALSE
)

write.csv(
  matched_r2_per_factor_table,
  file.path(
    matched_output_dir,
    "MOFA_matched_variance_explained_per_factor.csv"
  ),
  row.names = FALSE
)
```

## Quantify RNA-specific, proteomics-specific and shared factors

```{r matched factor layer structure}
matched_factor_layer_table <-
  matched_r2_per_factor_table %>%
  tidyr::pivot_wider(
    names_from = View,
    values_from = R2,
    values_fill = 0
  ) %>%
  dplyr::mutate(
    Proteomics_R2 =
      Olink + LCMS,
    Total_R2 =
      RNA_pseudobulk +
      Proteomics_R2,
    RNA_fraction =
      dplyr::if_else(
        Total_R2 > 0,
        RNA_pseudobulk /
          Total_R2,
        NA_real_
      ),
    Proteomics_fraction =
      dplyr::if_else(
        Total_R2 > 0,
        Proteomics_R2 /
          Total_R2,
        NA_real_
      ),
    Meaningful_RNA_and_proteomics =
      RNA_pseudobulk >=
        matched_shared_r2_cutoff &
      Proteomics_R2 >=
        matched_shared_r2_cutoff
  ) %>%
  dplyr::arrange(
    dplyr::desc(Total_R2)
  )

write.csv(
  matched_factor_layer_table,
  file.path(
    matched_output_dir,
    "MOFA_matched_factor_layer_structure.csv"
  ),
  row.names = FALSE
)

reactable::reactable(
  matched_factor_layer_table,
  searchable = TRUE,
  filterable = TRUE
)
```

```{r matched modality contribution heatmap, fig.height=7, fig.width=8}
matched_factor_order <-
  matched_factor_layer_table$
    Factor

matched_r2_plot_data <-
  matched_r2_per_factor_table %>%
  dplyr::mutate(
    Factor = factor(
      Factor,
      levels = rev(
        matched_factor_order
      )
    ),
    View = factor(
      View,
      levels = c(
        "RNA_pseudobulk",
        "Olink",
        "LCMS"
      )
    ),
    text_color = dplyr::if_else(
      R2 >=
        0.55 *
        max(
          R2,
          na.rm = TRUE
        ),
      "white",
      "black"
    )
  )

ggplot(
  matched_r2_plot_data,
  aes(
    x = View,
    y = Factor,
    fill = R2
  )
) +
  geom_tile(
    color = "grey40",
    linewidth = 0.3
  ) +
  geom_text(
    aes(
      label = sprintf(
        "%.1f%%",
        R2
      ),
      color = text_color
    ),
    size = 3
  ) +
  scale_color_identity() +
  scale_fill_gradient(
    low = "white",
    high = "navy"
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Variance\nexplained (%)"
  ) +
  theme_classic()
```

```{r matched proportional modality contribution, fig.height=6, fig.width=9}
matched_layer_fraction_data <-
  matched_factor_layer_table %>%
  dplyr::select(
    Factor,
    RNA_pseudobulk,
    Olink,
    LCMS
  ) %>%
  tidyr::pivot_longer(
    cols = -Factor,
    names_to = "View",
    values_to = "R2"
  ) %>%
  dplyr::mutate(
    Factor = factor(
      Factor,
      levels =
        matched_factor_order
    )
  )

ggplot(
  matched_layer_fraction_data,
  aes(
    x = Factor,
    y = R2,
    fill = View
  )
) +
  geom_col(
    position = "fill"
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = NULL,
    y = "Fraction of factor variance",
    fill = "View"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

```{r matched rna versus proteomics variance, fig.height=6, fig.width=7}
ggplot(
  matched_factor_layer_table,
  aes(
    x = RNA_pseudobulk,
    y = Proteomics_R2,
    label = Factor
  )
) +
  geom_vline(
    xintercept =
      matched_shared_r2_cutoff,
    linetype = 2
  ) +
  geom_hline(
    yintercept =
      matched_shared_r2_cutoff,
    linetype = 2
  ) +
  geom_point(
    aes(
      shape =
        Meaningful_RNA_and_proteomics
    ),
    size = 3
  ) +
  geom_text(
    nudge_y = 0.35,
    size = 3
  ) +
  labs(
    x = "RNA variance explained (%)",
    y = "Olink + LC-MS/MS variance explained (%)",
    shape = paste0(
      "At least ",
      matched_shared_r2_cutoff,
      "% in both layers"
    )
  ) +
  theme_classic()
```

# Extract matched-model factor scores and weights

```{r extract matched factors and weights}
matched_factor_scores <-
  MOFA2::get_factors(
    matched_model,
    as.data.frame = TRUE
  ) %>%
  dplyr::rename(
    Sample = sample,
    Factor = factor,
    Factor_value = value
  ) %>%
  dplyr::left_join(
    as.data.frame(
      matched_model_metadata
    ) %>%
      dplyr::rename(
        Sample = sample
      ),
    by = "Sample"
  )

matched_factor_scores_wide <-
  matched_factor_scores %>%
  dplyr::select(
    Sample,
    Factor,
    Factor_value
  ) %>%
  tidyr::pivot_wider(
    names_from = Factor,
    values_from = Factor_value
  )

matched_factor_weights <-
  MOFA2::get_weights(
    matched_model,
    as.data.frame = TRUE
  ) %>%
  dplyr::rename(
    View = view,
    Feature = feature,
    Factor = factor,
    Weight = value
  )

matched_factor_names <-
  MOFA2::factors_names(
    matched_model
  )

write.csv(
  matched_factor_scores,
  file.path(
    matched_output_dir,
    "MOFA_matched_factor_scores_long.csv"
  ),
  row.names = FALSE
)

write.csv(
  matched_factor_scores_wide,
  file.path(
    matched_output_dir,
    "MOFA_matched_factor_scores_wide.csv"
  ),
  row.names = FALSE
)

write.csv(
  matched_factor_weights,
  file.path(
    matched_output_dir,
    "MOFA_matched_feature_weights_all.csv"
  ),
  row.names = FALSE
)
```

```{r add fibrosis to matched factor scores}
matched_factor_scores <-
  matched_factor_scores %>%
  dplyr::left_join(
    fibrosis_data %>%
      dplyr::select(
        Sample,
        Fibrosis_percent
      ),
    by = "Sample"
  )
```

# Matched multi-omics Factor 1 and Factor 2 loadings

```{r matched Factor1 top loadings, fig.height=7, fig.width=5}
matched_factor1_top_loadings <-
  prepare_top_factor_loadings(
    weight_table =
      matched_factor_weights,
    factor_name = "Factor1",
    n_each_direction =
      n_top_factor_loadings_each_direction
  )

write.csv(
  matched_factor1_top_loadings,
  file.path(
    matched_output_dir,
    "MOFA_matched_Factor1_top_loadings.csv"
  ),
  row.names = FALSE
)

matched_factor1_loading_plot <-
  plot_top_factor_loadings(
    loading_data =
      matched_factor1_top_loadings,
    title_text =
      "Multi-omics Factor 1 top loadings"
  )

print(
  matched_factor1_loading_plot
)

```

```{r matched Factor2 top loadings, fig.height=7, fig.width=5}
matched_factor2_top_loadings <-
  prepare_top_factor_loadings(
    weight_table =
      matched_factor_weights,
    factor_name = "Factor2",
    n_each_direction =
      n_top_factor_loadings_each_direction
  )

write.csv(
  matched_factor2_top_loadings,
  file.path(
    matched_output_dir,
    "MOFA_matched_Factor2_top_loadings.csv"
  ),
  row.names = FALSE
)

matched_factor2_loading_plot <-
  plot_top_factor_loadings(
    loading_data =
      matched_factor2_top_loadings,
    title_text =
      "Multi-omics Factor 2 top loadings"
  )

print(
  matched_factor2_loading_plot
)

```

```{r matched group association models}
matched_group_global_results <-
  tibble::tibble()

matched_group_pairwise_results <-
  tibble::tibble()

for (
  factor_name in
  matched_factor_names
) {
  factor_data <-
    matched_factor_scores %>%
    dplyr::filter(
      Factor == factor_name,
      RV_group %in%
        rv_group_comparison_levels
    ) %>%
    dplyr::select(
      Sample,
      Factor_value,
      RV_group,
      Age_years,
      Sample_age_days
    ) %>%
    tidyr::drop_na() %>%
    dplyr::mutate(
      RV_group = factor(
        as.character(RV_group),
        levels =
          rv_group_comparison_levels
      )
    ) %>%
    droplevels()

  if (
    nlevels(factor_data$RV_group) < 2L
  ) {
    next
  }

  group_counts <-
    table(
      factor_data$RV_group
    )

  group_fit <- lm(
    Factor_value ~
      RV_group +
      scale(Age_years) +
      scale(Sample_age_days),
    data = factor_data
  )

  group_test <- drop1(
    group_fit,
    test = "F"
  )

  matched_group_global_results <-
    dplyr::bind_rows(
      matched_group_global_results,
      tibble::tibble(
        Factor = factor_name,
        comparison =
          "Compensated versus Decompensated",
        n = nrow(factor_data),
        n_compensated =
          unname(
            group_counts["Compensated"]
          ),
        n_decompensated =
          unname(
            group_counts["Decompensated"]
          ),
        F_statistic =
          group_test[
            "RV_group",
            "F value"
          ],
        p_value =
          group_test[
            "RV_group",
            "Pr(>F)"
          ],
        model_R2 =
          summary(group_fit)$r.squared,
        adjusted_model_R2 =
          summary(group_fit)$adj.r.squared
      )
    )

  factor_emmeans <- emmeans::emmeans(
    group_fit,
    ~ RV_group
  )

  factor_contrasts <-
    emmeans::contrast(
      factor_emmeans,
      method = list(
        "Decompensated - Compensated" =
          c(-1, 1)
      ),
      adjust = "none"
    ) %>%
    summary(
      infer = c(TRUE, TRUE)
    ) %>%
    as.data.frame() %>%
    dplyr::mutate(
      Factor = factor_name,
      n = nrow(factor_data),
      n_compensated =
        unname(
          group_counts["Compensated"]
        ),
      n_decompensated =
        unname(
          group_counts["Decompensated"]
        )
    ) %>%
    dplyr::select(
      Factor,
      contrast,
      estimate,
      SE,
      df,
      lower.CL,
      upper.CL,
      t.ratio,
      p.value,
      n,
      n_compensated,
      n_decompensated
    )

  matched_group_pairwise_results <-
    dplyr::bind_rows(
      matched_group_pairwise_results,
      factor_contrasts
    )
}

matched_group_global_results <-
  matched_group_global_results %>%
  dplyr::mutate(
    FDR = p.adjust(
      p_value,
      method = "BH"
    )
  ) %>%
  dplyr::arrange(FDR)

matched_group_pairwise_results <-
  matched_group_pairwise_results %>%
  dplyr::mutate(
    FDR = p.adjust(
      p.value,
      method = "BH"
    )
  ) %>%
  dplyr::arrange(FDR)

write.csv(
  matched_group_global_results,
  file.path(
    matched_output_dir,
    "MOFA_matched_factors_compensated_vs_decompensated_global.csv"
  ),
  row.names = FALSE
)

write.csv(
  matched_group_pairwise_results,
  file.path(
    matched_output_dir,
    "MOFA_matched_factors_compensated_vs_decompensated.csv"
  ),
  row.names = FALSE
)

reactable::reactable(
  matched_group_pairwise_results,
  searchable = TRUE,
  filterable = TRUE,
  defaultPageSize = 15
)
```

```{r matched factors by rv group, fig.height=3.5, fig.width=7}
matched_group_plot_data <-
  matched_factor_scores %>%
  dplyr::filter(
    RV_group %in%
      rv_group_comparison_levels
  ) %>%
  dplyr::mutate(
    RV_group = factor(
      as.character(RV_group),
      levels =
        rv_group_comparison_levels
    ),
    Factor = factor(
      Factor,
      levels =
        matched_factor_names
    )
  )

matched_group_fdr_labels <-
  matched_group_pairwise_results %>%
  dplyr::mutate(
    Factor = factor(
      Factor,
      levels =
        matched_factor_names
    ),
    label =
      format_fdr_label(FDR)
  )

ggplot(
  matched_group_plot_data,
  aes(
    x = RV_group,
    y = Factor_value,
    fill = RV_group
  )
) +
  geom_violin(
    trim = FALSE,
    alpha = 0.30,
    color = NA
  ) +
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    color = "grey20"
  ) +
  geom_jitter(
    width = 0.10,
    height = 0,
    size = 1.8,
    shape = 21,
    color = "black",
    stroke = 0.25
  ) +
  geom_text(
    data =
      matched_group_fdr_labels,
    aes(
      x = 1.5,
      y = Inf,
      label = label
    ),
    inherit.aes = FALSE,
    vjust = 1.25,
    size = 3.2
  ) +
  scale_fill_manual(
    values = group_colors[
      rv_group_comparison_levels
    ],
    drop = FALSE
  ) +
  scale_y_continuous(
    expand = expansion(
      mult = c(0.05, 0.16)
    )
  ) +
  facet_wrap(
    ~ Factor,
    scales = "free_y",
    ncol = 4
  ) +
  labs(
    x = NULL,
    y = "Factor score"
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    strip.text =
      element_text(face = "bold")
  )
```

# Matched-model associations with continuous clinical parameters

```{r matched continuous clinical association models}
matched_continuous_results <-
  tibble::tibble()

for (
  factor_name in
  matched_factor_names
) {
  for (
    clinical_parameter in
    clinical_parameters
  ) {
    factor_data <-
      matched_factor_scores %>%
      dplyr::filter(
        Factor == factor_name
      ) %>%
      dplyr::select(
        Sample,
        Factor_value,
        Age_years,
        Sample_age_days,
        dplyr::all_of(
          clinical_parameter
        )
      ) %>%
      tidyr::drop_na()

    factor_data <- factor_data %>%
      dplyr::mutate(
        clinical_scaled =
          as.numeric(
            scale(
              .data[[clinical_parameter]]
            )
          ),
        age_scaled =
          as.numeric(
            scale(Age_years)
          ),
        storage_scaled =
          as.numeric(
            scale(
              Sample_age_days
            )
          )
      )

    continuous_fit <- lm(
      Factor_value ~
        clinical_scaled +
        age_scaled +
        storage_scaled,
      data = factor_data
    )

    clinical_coefficient <-
      broom::tidy(
        continuous_fit,
        conf.int = TRUE
      ) %>%
      dplyr::filter(
        term ==
          "clinical_scaled"
      )

    matched_continuous_results <-
      dplyr::bind_rows(
        matched_continuous_results,
        tibble::tibble(
          Factor = factor_name,
          parameter =
            clinical_parameter,
          n = nrow(factor_data),
          beta =
            clinical_coefficient$
              estimate,
          standard_error =
            clinical_coefficient$
              std.error,
          confidence_interval_low =
            clinical_coefficient$
              conf.low,
          confidence_interval_high =
            clinical_coefficient$
              conf.high,
          t_statistic =
            clinical_coefficient$
              statistic,
          p_value =
            clinical_coefficient$
              p.value,
          model_R2 =
            summary(
              continuous_fit
            )$r.squared,
          adjusted_model_R2 =
            summary(
              continuous_fit
            )$adj.r.squared
        )
      )
  }
}

matched_continuous_results <-
  matched_continuous_results %>%
  dplyr::group_by(parameter) %>%
  dplyr::mutate(
    FDR_within_parameter =
      p.adjust(
        p_value,
        method = "BH"
      )
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    FDR_global =
      p.adjust(
        p_value,
        method = "BH"
      ),
    parameter_label =
      unname(
        pretty_parameter_names[
          parameter
        ]
      )
  ) %>%
  dplyr::arrange(
    FDR_global
  )

write.csv(
  matched_continuous_results,
  file.path(
    matched_output_dir,
    "MOFA_matched_factors_vs_continuous_clinical_parameters.csv"
  ),
  row.names = FALSE
)
```

```{r matched continuous association dotplot, fig.height=3.5, fig.width=6.5}
matched_continuous_plot_data <-
  matched_continuous_results %>%
  dplyr::mutate(
    parameter_label = factor(
      parameter_label,
      levels = unname(
        pretty_parameter_names[
          clinical_parameters
        ]
      )
    ),
    Factor = factor(
      Factor,
      levels = rev(
        matched_factor_names
      )
    ),
    dot_size = -log10(
      pmax(
        FDR_global,
        .Machine$double.xmin
      )
    ),
    outlined =
      FDR_global <
      exploratory_factor_fdr
  )

ggplot(
  matched_continuous_plot_data,
  aes(
    x = parameter_label,
    y = Factor,
    size = dot_size,
    fill = beta
  )
) +
  geom_point(
    shape = 21,
    color = "grey60",
    stroke = 0.3
  ) +
  geom_point(
    data = dplyr::filter(
      matched_continuous_plot_data,
      outlined
    ),
    shape = 21,
    fill = NA,
    color = "black",
    stroke = 1,
    show.legend = FALSE
  ) +
  scale_size_continuous(
    range = c(2.5, 8)
  ) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0
  ) +
  labs(
    x = NULL,
    y = NULL,
    size =
      expression(
        -log[10](FDR)
      ),
    fill =
      "Coefficient\nper 1 SD"
  ) +
  theme_classic(
  base_size = 14
) +
theme(
  axis.text.x = element_text(
    angle = 45,
    hjust = 1,
    size = 12
  ),
  axis.text.y = element_text(
    size = 12
  ),
  legend.title = element_text(
    size = 12
  ),
  legend.text = element_text(
    size = 11
  ),
    plot.margin = margin(
    t = 30,
    r = 8,
    b = 5,
    l = 5,
    unit = "pt"
  )
)
```

```{r matched fibrosis association models}
matched_fibrosis_results <-
  fit_factor_fibrosis_models(
    factor_scores =
      matched_factor_scores,
    factor_names =
      matched_factor_names
  )

write.csv(
  matched_fibrosis_results,
  file.path(
    matched_output_dir,
    "MOFA_matched_factors_vs_histological_fibrosis.csv"
  ),
  row.names = FALSE
)

reactable::reactable(
  matched_fibrosis_results,
  searchable = TRUE,
  filterable = TRUE
)
```

```{r matched fibrosis association plot, fig.height=3, fig.width=5}
matched_fibrosis_plot <-
  plot_factor_fibrosis_associations(
    fibrosis_results =
      matched_fibrosis_results,
    factor_names =
      matched_factor_names,
    title_text =
      "Multi-omics factors and histological fibrosis"
  )+
  theme(
    axis.text.y = element_text(
      size = 12
    )
  )

matched_fibrosis_plot

```

```{r matched Factor2 fibrosis correlation, fig.height=2.5, fig.width=2.7}
matched_factor2_fibrosis_data <-
  matched_factor_scores %>%
  dplyr::filter(
    Factor == "Factor2",
    is.finite(
      Fibrosis_percent
    ),
    is.finite(
      Factor_value
    )
  )

matched_factor2_fibrosis_correlation <-
  stats::cor.test(
    matched_factor2_fibrosis_data$
      Fibrosis_percent,
    matched_factor2_fibrosis_data$
      Factor_value,
    method = "spearman",
    exact = FALSE
  )

matched_factor2_fibrosis_label <-
  paste0(
    "Spearman \u03c1 = ",
    sprintf(
      "%.2f",
      unname(
        matched_factor2_fibrosis_correlation$
          estimate
      )
    ),
    "\n",
    format_p_value(
      matched_factor2_fibrosis_correlation$
        p.value
    )
  )

matched_factor2_fibrosis_plot <-
  ggplot(
    matched_factor2_fibrosis_data,
    aes(
      x = Fibrosis_percent,
      y = Factor_value
    )
  ) +
  geom_point(
    size = 2.4
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    linewidth = 0.8
  ) +
  annotate(
  "text",
  x = 0.12,
  y = 4.3,
  label = matched_factor2_fibrosis_label,
  hjust = 0,
  vjust = 1,
  size = 3.5
) +
  labs(
    x = "Fibrotic area (%)",
    y = "Factor 2 score"
  ) +
  theme_classic()

matched_factor2_fibrosis_plot

write.csv(
  matched_factor2_fibrosis_data,
  file.path(
    matched_output_dir,
    "MOFA_matched_Factor2_histological_fibrosis_scatter_data.csv"
  ),
  row.names = FALSE
)

```

# Matched Factor 2 fibrosis-associated features

```{r matched Factor2 fibrosis custom heatmap, fig.height=9, fig.width=10}
clean_matched_feature_name <- function(
    feature
) {
  stringr::str_remove(
    as.character(feature),
    "_(RNA_pseudobulk|Olink|OLINK|LCMS|RNA|MS)$"
  )
}


factor2_fibrosis_beta <-
  matched_fibrosis_results %>%
  dplyr::filter(
    Factor == "Factor2"
  ) %>%
  dplyr::pull(
    beta
  )


if (
  length(factor2_fibrosis_beta) != 1L ||
  !is.finite(factor2_fibrosis_beta)
) {
  stop(
    "A valid fibrosis coefficient for Factor2 was not found."
  )
}


factor2_fibrosis_orientation <-
  ifelse(
    factor2_fibrosis_beta >= 0,
    1,
    -1
  )


matched_factor2_custom_sets <-
  tibble::tribble(
    ~Feature,   ~View,            ~Annotation,

    "FOSB",     "RNA_pseudobulk", "AP-1 / immediate-early activation",
    "FOS",      "RNA_pseudobulk", "AP-1 / immediate-early activation",
    "JUNB",     "RNA_pseudobulk", "AP-1 / immediate-early activation",
    "FOSL1",    "RNA_pseudobulk", "AP-1 / immediate-early activation",
    "EGR3",     "RNA_pseudobulk", "AP-1 / immediate-early activation",
    "MYC",      "RNA_pseudobulk", "AP-1 / immediate-early activation",

    "SERPINE1", "RNA_pseudobulk", "Profibrotic matricellular / TGF-beta-associated",
    "TNC",      "RNA_pseudobulk", "Profibrotic matricellular / TGF-beta-associated",
    "CCN2",     "RNA_pseudobulk", "Profibrotic matricellular / TGF-beta-associated",
    "ADAMTS1",  "RNA_pseudobulk", "Profibrotic matricellular / TGF-beta-associated",
    "LTBP3",    "Olink",          "Profibrotic matricellular / TGF-beta-associated",
    "ITGA11",   "Olink",          "Profibrotic matricellular / TGF-beta-associated",

    "VCAN",     "RNA_pseudobulk", "ECM organization and adhesion",
    "MATN3",    "Olink",          "ECM organization and adhesion",
    "LUM",      "LCMS",           "ECM organization and adhesion",
    "PRELP",    "LCMS",           "ECM organization and adhesion",
    "FBLN1",    "LCMS",           "ECM organization and adhesion",
    "ITGB1",    "LCMS",           "ECM organization and adhesion",
    "CD151",    "LCMS",           "ECM organization and adhesion",

    "VIM",      "RNA_pseudobulk", "Myofibroblast / contractile cytoskeleton",
    "TAGLN",    "RNA_pseudobulk", "Myofibroblast / contractile cytoskeleton",
    "TPM4",     "RNA_pseudobulk", "Myofibroblast / contractile cytoskeleton",
    "FLNA",     "LCMS",           "Myofibroblast / contractile cytoskeleton",
    "TAGLN",    "LCMS",           "Myofibroblast / contractile cytoskeleton",
    "CNN1",     "LCMS",           "Myofibroblast / contractile cytoskeleton",
    "S100A4",   "LCMS",           "Myofibroblast / contractile cytoskeleton",
    "PLS3",     "LCMS",           "Myofibroblast / contractile cytoskeleton",
    "TAGLN2",   "LCMS",           "Myofibroblast / contractile cytoskeleton",
    "ACTN1",    "LCMS",           "Myofibroblast / contractile cytoskeleton",
    "MYL6",     "LCMS",           "Myofibroblast / contractile cytoskeleton",
    "VCL",      "LCMS",           "Myofibroblast / contractile cytoskeleton",
    "MYH9",     "LCMS",           "Myofibroblast / contractile cytoskeleton",
    "ACTR3",    "LCMS",           "Myofibroblast / contractile cytoskeleton",
    "PARVA",    "LCMS",           "Myofibroblast / contractile cytoskeleton"
  ) %>%
  dplyr::mutate(
    Annotation = factor(
      Annotation,
      levels = c(
        "AP-1 / immediate-early activation",
        "Profibrotic matricellular / TGF-beta-associated",
        "ECM organization and adhesion",
        "Myofibroblast / contractile cytoskeleton"
      )
    )
  )


matched_factor2_weights_clean <-
  matched_factor_weights %>%
  dplyr::filter(
    Factor == "Factor2",
    View %in% c(
      "RNA_pseudobulk",
      "Olink",
      "LCMS"
    ),
    is.finite(Weight)
  ) %>%
  dplyr::mutate(
    Feature_clean =
      clean_matched_feature_name(
        Feature
      ),

    Fibrosis_oriented_weight =
      Weight *
      factor2_fibrosis_orientation
  )


matched_factor2_heatmap_features <-
  matched_factor2_custom_sets %>%
  dplyr::mutate(
    Feature_clean =
      Feature
  ) %>%
  dplyr::left_join(
    matched_factor2_weights_clean %>%
      dplyr::select(
        Feature_actual =
          Feature,

        Feature_clean,
        View,
        Weight,
        Fibrosis_oriented_weight
      ),

    by = c(
      "Feature_clean",
      "View"
    )
  )


missing_heatmap_features <-
  matched_factor2_heatmap_features %>%
  dplyr::filter(
    is.na(Weight)
  )


if (
  nrow(
    missing_heatmap_features
  ) > 0L
) {
  warning(
    "The following selected features were not found in the current Factor2 weight table: ",
    paste(
      paste0(
        missing_heatmap_features$
          Feature,
        " (",
        missing_heatmap_features$
          View,
        ")"
      ),
      collapse = ", "
    )
  )
}


matched_factor2_heatmap_features <-
  matched_factor2_heatmap_features %>%
  dplyr::filter(
    !is.na(Weight)
  ) %>%
  dplyr::mutate(
    Modality =
      dplyr::case_when(
        View == "RNA_pseudobulk" ~
          "RNA",

        View == "Olink" ~
          "OLINK",

        View == "LCMS" ~
          "MS",

        TRUE ~
          View
      ),

    Modality = factor(
      Modality,
      levels = c(
        "RNA",
        "OLINK",
        "MS"
      )
    )
  ) %>%
  dplyr::arrange(
    Annotation,
    dplyr::desc(
      Fibrosis_oriented_weight
    )
  )


matched_sample_intersection <-
  Reduce(
    intersect,
    lapply(
      matched_data[
        c(
          "RNA_pseudobulk",
          "Olink",
          "LCMS"
        )
      ],
      colnames
    )
  )


matched_fibrosis_sample_metadata <-
  fibrosis_data %>%
  dplyr::transmute(
    Sample =
      as.character(
        Sample
      ),

    Fibrotic_area_percent =
      as.numeric(
        Fibrosis_percent
      )
  ) %>%
  dplyr::filter(
    Sample %in%
      matched_sample_intersection,

    is.finite(
      Fibrotic_area_percent
    )
  ) %>%
  dplyr::arrange(
    Fibrotic_area_percent,
    Sample
  )


matched_fibrosis_sample_order <-
  matched_fibrosis_sample_metadata$
    Sample


extract_matched_factor2_heatmap_matrix <- function(
    feature_table,
    data_views,
    sample_order
) {
  extracted_rows <- vector(
    mode = "list",
    length = nrow(
      feature_table
    )
  )

  row_found <- logical(
    nrow(
      feature_table
    )
  )

  resolved_features <- character(
    nrow(
      feature_table
    )
  )

  for (
    feature_index in
    seq_len(
      nrow(
        feature_table
      )
    )
  ) {
    current_view <-
      as.character(
        feature_table$View[
          feature_index
        ]
      )

    current_feature <-
      as.character(
        feature_table$Feature_actual[
          feature_index
        ]
      )

    if (
      !current_view %in%
        names(
          data_views
        )
    ) {
      next
    }

    feature_candidates <- unique(
      c(
        current_feature,

        as.character(
          feature_table$Feature[
            feature_index
          ]
        ),

        paste0(
          as.character(
            feature_table$Feature[
              feature_index
            ]
          ),
          "_",
          current_view
        ),

        paste0(
          as.character(
            feature_table$Feature[
              feature_index
            ]
          ),
          "_",
          dplyr::case_when(
            current_view == "Olink" ~
              "OLINK",

            current_view == "LCMS" ~
              "MS",

            current_view == "RNA_pseudobulk" ~
              "RNA_pseudobulk",

            TRUE ~
              current_view
          )
        )
      )
    )

    resolved_feature <-
      feature_candidates[
        feature_candidates %in%
          rownames(
            data_views[[
              current_view
            ]]
          )
      ][1]

    if (
      length(
        resolved_feature
      ) == 0L ||
      is.na(
        resolved_feature
      )
    ) {
      next
    }

    extracted_rows[[
      feature_index
    ]] <-
      as.numeric(
        data_views[[
          current_view
        ]][
          resolved_feature,
          sample_order,
          drop = TRUE
        ]
      )

    resolved_features[
      feature_index
    ] <-
      resolved_feature

    row_found[
      feature_index
    ] <-
      TRUE
  }


  if (
    !any(
      row_found
    )
  ) {
    stop(
      "None of the selected matched-model features were found in matched_data."
    )
  }


  if (
    any(
      !row_found
    )
  ) {
    warning(
      "The following selected features were not found and were omitted: ",
      paste(
        paste0(
          feature_table$Feature[
            !row_found
          ],
          " (",
          feature_table$View[
            !row_found
          ],
          ")"
        ),
        collapse = ", "
      )
    )
  }


  retained_feature_table <-
    feature_table[
      row_found,
      ,
      drop = FALSE
    ] %>%
    dplyr::mutate(
      Resolved_feature =
        resolved_features[
          row_found
        ],

      Display_feature =
        Feature,

      Internal_row_id =
        make.unique(
          paste(
            Annotation,
            View,
            Feature,
            sep = "__"
          )
        )
    )


  expression_matrix <-
    do.call(
      rbind,
      extracted_rows[
        row_found
      ]
    )


  colnames(
    expression_matrix
  ) <-
    sample_order


  rownames(
    expression_matrix
  ) <-
    retained_feature_table$
      Internal_row_id


  list(
    matrix =
      expression_matrix,

    feature_table =
      retained_feature_table
  )
}


prepare_matched_heatmap_z_scores <- function(
    expression_matrix
) {
  imputed_matrix <-
    expression_matrix


  for (
    row_index in
    seq_len(
      nrow(
        imputed_matrix
      )
    )
  ) {
    row_values <-
      imputed_matrix[
        row_index,
        ,
        drop = TRUE
      ]


    finite_values <-
      row_values[
        is.finite(
          row_values
        )
      ]


    if (
      length(
        finite_values
      ) == 0L
    ) {
      next
    }


    row_values[
      !is.finite(
        row_values
      )
    ] <-
      stats::median(
        finite_values
      )


    imputed_matrix[
      row_index,
    ] <-
      row_values
  }


  row_means <-
    rowMeans(
      imputed_matrix,
      na.rm = TRUE
    )


  row_standard_deviations <-
    apply(
      imputed_matrix,
      1,
      stats::sd,
      na.rm = TRUE
    )


  row_standard_deviations[
    !is.finite(
      row_standard_deviations
    ) |
      row_standard_deviations == 0
  ] <- 1


  z_score_matrix <-
    sweep(
      imputed_matrix,
      1,
      row_means,
      "-"
    )


  z_score_matrix <-
    sweep(
      z_score_matrix,
      1,
      row_standard_deviations,
      "/"
    )


  pmax(
    pmin(
      z_score_matrix,
      2
    ),
    -2
  )
}


make_fibrosis_grey_scale <- function(
    values
) {
  finite_values <-
    values[
      is.finite(
        values
      )
    ]


  value_range <-
    if (
      length(
        finite_values
      ) == 0L
    ) {
      c(
        0,
        1
      )
    } else {
      range(
        finite_values,
        na.rm = TRUE
      )
    }


  if (
    value_range[1] ==
      value_range[2]
  ) {
    value_range <-
      value_range +
      c(
        -0.5,
        0.5
      )
  }


  circlize::colorRamp2(
    value_range,
    c(
      "grey95",
      "grey10"
    )
  )
}


matched_factor2_heatmap_data <-
  extract_matched_factor2_heatmap_matrix(
    feature_table =
      matched_factor2_heatmap_features,

    data_views =
      matched_data,

    sample_order =
      matched_fibrosis_sample_order
  )


matched_factor2_heatmap_matrix <-
  prepare_matched_heatmap_z_scores(
    matched_factor2_heatmap_data$
      matrix
  )


matched_factor2_feature_metadata <-
  matched_factor2_heatmap_data$
    feature_table


fibrosis_colour_function <-
  make_fibrosis_grey_scale(
    matched_fibrosis_sample_metadata$
      Fibrotic_area_percent
  )


matched_factor2_top_annotation <-
  ComplexHeatmap::HeatmapAnnotation(
    `Fibrotic area (%)` =
      ComplexHeatmap::anno_simple(
        matched_fibrosis_sample_metadata$
          Fibrotic_area_percent,

        col =
          fibrosis_colour_function,

        na_col =
          "white",

        height =
          grid::unit(
            4.5,
            "mm"
          )
      ),

    annotation_name_gp =
      grid::gpar(
        fontsize = 11,
        fontface = "bold",
        col = "black"
      ),

    annotation_name_side =
      "right"
  )


set_colors <- c(
  "AP-1 / immediate-early activation" =
    "#6A3D9A",

  "Profibrotic matricellular / TGF-beta-associated" =
    "#B35806",

  "ECM organization and adhesion" =
    "#1B7837",

  "Myofibroblast / contractile cytoskeleton" =
    "#2166AC"
)


annotation_slice_order <-
  unique(
    as.character(
      matched_factor2_feature_metadata$
        Annotation
    )
  )


annotation_display_labels <- c(
  "AP-1 / immediate-early activation" =
    "AP-1 /\nimmediate-early\nactivation",

  "Profibrotic matricellular / TGF-beta-associated" =
    "Profibrotic /\nTGF-beta-associated",

  "ECM organization and adhesion" =
    "ECM organization\nand adhesion",

  "Myofibroblast / contractile cytoskeleton" =
    "Myofibroblast /\ncontractile cytoskeleton"
)


matched_factor2_left_annotation <-
  ComplexHeatmap::rowAnnotation(
    Annotation =
      ComplexHeatmap::anno_block(
        gp =
          grid::gpar(
            fill =
              unname(
                set_colors[
                  annotation_slice_order
                ]
              ),

            col = NA
          ),

        labels =
          unname(
            annotation_display_labels[
              annotation_slice_order
            ]
          ),

        labels_gp =
          grid::gpar(
            fontsize = 8.5,
            fontface = "bold",
            col = "white",
            lineheight = 0.85
          ),

        labels_rot =
          90,

        width =
          grid::unit(
            9,
            "mm"
          )
      ),

    show_annotation_name =
      FALSE
  )


modality_colors <- c(
  "RNA" =
    "#009E73",

  "OLINK" =
    "#E69F00",

  "MS" =
    "#7B61A8"
)


matched_factor2_right_annotation <-
  ComplexHeatmap::rowAnnotation(
    Modality =
      matched_factor2_feature_metadata$
        Modality,

    col = list(
      Modality =
        modality_colors
    ),

    show_legend =
      FALSE,

    annotation_name_gp =
      grid::gpar(
        fontsize = 10,
        fontface = "bold",
        col = "black"
      ),

    simple_anno_size =
      grid::unit(
        4.5,
        "mm"
      )
  )


heatmap_colour_function <-
  circlize::colorRamp2(
    c(
      -2,
      0,
      2
    ),
    c(
      "#2166AC",
      "white",
      "#B2182B"
    )
  )


matched_factor2_custom_heatmap <-
  ComplexHeatmap::Heatmap(
    matched_factor2_heatmap_matrix,

    name =
      "Row Z-score",

    col =
      heatmap_colour_function,

    na_col =
      "grey90",

    show_heatmap_legend =
      FALSE,

    cluster_rows =
      FALSE,

    cluster_columns =
      FALSE,

    row_split =
      matched_factor2_feature_metadata$
        Annotation,

    row_gap =
      grid::unit(
        2,
        "mm"
      ),

    cluster_row_slices =
      FALSE,

    top_annotation =
      matched_factor2_top_annotation,

    left_annotation =
      matched_factor2_left_annotation,

    right_annotation =
      matched_factor2_right_annotation,

    row_labels =
      matched_factor2_feature_metadata$
        Display_feature,

    row_names_side =
      "left",

    row_names_gp =
      grid::gpar(
        fontsize = 9,
        col = "black"
      ),

    column_names_rot =
      45,

    column_names_gp =
      grid::gpar(
        fontsize = 7,
        col = "black"
      ),

    show_row_dend =
      FALSE,

    show_column_dend =
      FALSE,

    row_title =
      NULL,

    column_title =
      NULL,

    border =
      TRUE,

height =
  grid::unit(
    nrow(
      matched_factor2_heatmap_matrix
    ) * 6,
    "mm"
  )
)


row_z_score_legend <-
  ComplexHeatmap::Legend(
    title =
      "Row Z-score",

    col_fun =
      heatmap_colour_function,

    at = c(
      -2,
      0,
      2
    )
  )


modality_legend <-
  ComplexHeatmap::Legend(
    title =
      "Modality",

    at = c(
      "RNA",
      "OLINK",
      "MS"
    ),

    legend_gp =
      grid::gpar(
        fill =
          unname(
            modality_colors[
              c(
                "RNA",
                "OLINK",
                "MS"
              )
            ]
          )
      )
  )


fibrosis_legend <-
  ComplexHeatmap::Legend(
    title =
      "Fibrotic area (%)",

    col_fun =
      fibrosis_colour_function
  )


matched_factor2_combined_legends <-
  ComplexHeatmap::packLegend(
    row_z_score_legend,
    modality_legend,
    fibrosis_legend,

    direction =
      "vertical",

    gap =
      grid::unit(
        5,
        "mm"
      )
  )


ComplexHeatmap::draw(
  matched_factor2_custom_heatmap,

  heatmap_legend_side =
    "right",

  annotation_legend_side =
    "right",

  heatmap_legend_list =
    list(
      matched_factor2_combined_legends
    ),

  merge_legends =
    FALSE
)




write.csv(
  matched_factor2_feature_metadata,

  file.path(
    matched_output_dir,
    "MOFA_matched_Factor2_custom_fibrosis_heatmap_features.csv"
  ),

  row.names = FALSE
)


write.csv(
  matched_fibrosis_sample_metadata,

  file.path(
    matched_output_dir,
    "MOFA_matched_Factor2_custom_fibrosis_heatmap_sample_order.csv"
  ),

  row.names = FALSE
)
```

# Matched-model factor summary

```{r matched factor summary}
matched_factor_summary <-
  matched_factor_layer_table %>%
  dplyr::left_join(
    matched_group_global_results %>%
      dplyr::select(
        Factor,
        RV_group_p = p_value,
        RV_group_FDR = FDR
      ),
    by = "Factor"
  ) %>%
  dplyr::left_join(
    matched_continuous_results %>%
      dplyr::group_by(Factor) %>%
      dplyr::slice_min(
        FDR_global,
        n = 1,
        with_ties = FALSE
      ) %>%
      dplyr::ungroup() %>%
      dplyr::select(
        Factor,
        strongest_clinical_parameter =
          parameter,
        strongest_clinical_beta =
          beta,
        strongest_clinical_FDR =
          FDR_global
      ),
    by = "Factor"
  ) %>%
  dplyr::arrange(
    dplyr::desc(Total_R2)
  )

write.csv(
  matched_factor_summary,
  file.path(
    matched_output_dir,
    "MOFA_matched_main_factor_summary.csv"
  ),
  row.names = FALSE
)
 
reactable::reactable(
  matched_factor_summary,
  searchable = TRUE,
  filterable = TRUE
)
```

# Matched Factor 2 gene set enrichment analysis

Factor 2 was analysed separately in the transcriptomic and proteomic layers. RNA pseudobulk weights were analysed directly, while Olink and LC-MS/MS weights were scaled within each platform before aggregation.

```{r matched Factor2 GSEA inputs}
matched_gsea_factor_table <-
  matched_gsea_factor_table %>%
  dplyr::filter(
    Factor %in%
      matched_factor_names
  )

matched_rna_weights_for_enrichment <-
  matched_factor_weights %>%
  dplyr::filter(
    View == "RNA_pseudobulk",
    Factor %in%
      matched_gsea_factor_table$
        Factor
  )

matched_proteomics_weights_for_enrichment <-
  matched_factor_weights %>%
  dplyr::filter(
    View %in%
      c(
        "Olink",
        "LCMS"
      ),
    Factor %in%
      matched_gsea_factor_table$
        Factor
  )
```

```{r matched Factor2 RNA and proteomics GO GSEA, message=TRUE, warning=TRUE}
matched_rna_gsea <-
  run_mofa_gsea(
    weight_table =
      matched_rna_weights_for_enrichment,
    factor_table =
      matched_gsea_factor_table
  )

matched_proteomics_gsea <-
  run_mofa_gsea(
    weight_table =
      matched_proteomics_weights_for_enrichment,
    factor_table =
      matched_gsea_factor_table
  )

matched_rna_gsea_simplified_objects <-
  simplify_enrichment_objects(
    matched_rna_gsea$objects
  )

matched_proteomics_gsea_simplified_objects <-
  simplify_enrichment_objects(
    matched_proteomics_gsea$objects
  )

matched_rna_gsea_results <-
  matched_rna_gsea$results

matched_proteomics_gsea_results <-
  matched_proteomics_gsea$results

matched_rna_gsea_simplified_results <-
  extract_gsea_results(
    matched_rna_gsea_simplified_objects
  )

matched_proteomics_gsea_simplified_results <-
  extract_gsea_results(
    matched_proteomics_gsea_simplified_objects
  )

write.csv(
  matched_rna_gsea_results,
  file.path(
    matched_output_dir,
    "MOFA_matched_RNA_Factor2_GO_GSEA.csv"
  ),
  row.names = FALSE
)

write.csv(
  matched_rna_gsea_simplified_results,
  file.path(
    matched_output_dir,
    "MOFA_matched_RNA_Factor2_GO_GSEA_simplified.csv"
  ),
  row.names = FALSE
)

write.csv(
  matched_proteomics_gsea_results,
  file.path(
    matched_output_dir,
    "MOFA_matched_proteomics_Factor2_GO_GSEA.csv"
  ),
  row.names = FALSE
)

write.csv(
  matched_proteomics_gsea_simplified_results,
  file.path(
    matched_output_dir,
    "MOFA_matched_proteomics_Factor2_GO_GSEA_simplified.csv"
  ),
  row.names = FALSE
)

saveRDS(
  list(
    Transcriptomics = list(
      GSEA =
        matched_rna_gsea$objects,
      GSEA_simplified =
        matched_rna_gsea_simplified_objects,
      ranked_lists =
        matched_rna_gsea$ranked_lists
    ),
    Proteomics = list(
      GSEA =
        matched_proteomics_gsea$objects,
      GSEA_simplified =
        matched_proteomics_gsea_simplified_objects,
      ranked_lists =
        matched_proteomics_gsea$ranked_lists
    )
  ),
  file.path(
    matched_output_dir,
    "MOFA_matched_Factor2_separate_RNA_proteomics_GO_GSEA_objects.rds"
  )
)
```

```{r matched Factor2 simplified GO GSEA plots, fig.height=4.8, fig.width=12.1, dpi=300}
factor2_gsea_top_positive <- 8
factor2_gsea_top_negative <- 8
factor2_gsea_fdr_cutoff <- 0.05


wrap_factor2_gsea_labels <- function(
    x,
    width = 41
) {
  vapply(
    x,
    function(label) {
      label <-
        stringr::str_squish(
          as.character(label)
        )

      if (
        nchar(label) <= width
      ) {
        return(label)
      }

      words <-
        stringr::str_split(
          label,
          "\\s+"
        )[[1]]

      lines <- c(
        "",
        "",
        ""
      )

      current_line <- 1L

      for (
        word in words
      ) {
        candidate <-
          if (
            lines[current_line] == ""
          ) {
            word
          } else {
            paste(
              lines[current_line],
              word
            )
          }

        if (
          nchar(candidate) <= width
        ) {
          lines[current_line] <-
            candidate
        } else if (
          current_line < 3L
        ) {
          current_line <-
            current_line + 1L

          lines[current_line] <-
            word
        } else {
          lines[current_line] <-
            paste0(
              substr(
                lines[current_line],
                1,
                width - 1
              ),
              "…"
            )

          break
        }
      }

      if (
        nchar(lines[3]) > width
      ) {
        lines[3] <-
          paste0(
            substr(
              lines[3],
              1,
              width - 1
            ),
            "…"
          )
      }

      paste(
        lines[
          lines != ""
        ],
        collapse = "\n"
      )
    },
    FUN.VALUE = character(1)
  )
}


make_factor2_simplified_gsea_plot <- function(
    gsea_object,
    layer_label
) {
  if (
    is.null(gsea_object) ||
    nrow(
      as.data.frame(
        gsea_object
      )
    ) == 0L
  ) {
    return(
      ggplot() +
        annotate(
          "text",
          x = 1,
          y = 1,
          label =
            "No GSEA results"
        ) +
        labs(
          title =
            paste0(
              layer_label,
              " | GSEA | Gene Ontology"
            )
        ) +
        theme_void()
    )
  }

  gsea_results <-
    as.data.frame(
      gsea_object
    ) %>%
    tibble::as_tibble() %>%
    dplyr::filter(
      is.finite(NES),
      is.finite(p.adjust),
      p.adjust <
        factor2_gsea_fdr_cutoff
    )

  if (
    nrow(gsea_results) == 0L
  ) {
    return(
      ggplot() +
        annotate(
          "text",
          x = 1,
          y = 1,
          label =
            paste0(
              "No terms with FDR < ",
              factor2_gsea_fdr_cutoff
            )
        ) +
        labs(
          title =
            paste0(
              layer_label,
              " | GSEA | Gene Ontology"
            )
        ) +
        theme_void()
    )
  }

  positive_terms <-
    gsea_results %>%
    dplyr::filter(
      NES > 0
    ) %>%
    dplyr::arrange(
      dplyr::desc(NES)
    ) %>%
    dplyr::slice_head(
      n =
        factor2_gsea_top_positive
    )

  negative_terms <-
    gsea_results %>%
    dplyr::filter(
      NES < 0
    ) %>%
    dplyr::arrange(NES) %>%
    dplyr::slice_head(
      n =
        factor2_gsea_top_negative
    )

  plot_data <-
    dplyr::bind_rows(
      positive_terms,
      negative_terms
    ) %>%
    dplyr::mutate(
      Description_clean =
        stringr::str_remove(
          Description,
          "^\\[(BP|CC|MF)\\]\\s*"
        )
    )

  if (
    "pvalue" %in%
      colnames(plot_data)
  ) {
    plot_data <-
      plot_data %>%
      dplyr::mutate(
        p_for_colour =
          dplyr::if_else(
            is.finite(pvalue) &
              pvalue > 0,
            pvalue,
            p.adjust
          )
      )
  } else {
    plot_data <-
      plot_data %>%
      dplyr::mutate(
        p_for_colour =
          p.adjust
      )
  }

  plot_data <-
    plot_data %>%
    dplyr::mutate(
      colour_score =
        -log10(
          pmax(
            p_for_colour,
            .Machine$double.xmin
          )
        ) *
        sign(NES),

      Description_wrapped =
        wrap_factor2_gsea_labels(
          Description_clean,
          width = 41
        )
    ) %>%
    dplyr::arrange(NES) %>%
    dplyr::mutate(
      Description_wrapped =
        factor(
          Description_wrapped,
          levels =
            unique(
              Description_wrapped
            )
        )
    )

  colour_limit <-
    unname(
      stats::quantile(
        abs(
          plot_data$
            colour_score
        ),
        probs = 0.80,
        na.rm = TRUE
      )
    )

  if (
    length(colour_limit) != 1L ||
    !is.finite(colour_limit) ||
    colour_limit == 0
  ) {
    colour_limit <-
      max(
        abs(
          plot_data$
            colour_score
        ),
        na.rm = TRUE
      )
  }

  ggplot(
    plot_data,
    aes(
      x =
        Description_wrapped,
      y = NES,
      fill =
        colour_score
    )
  ) +
    geom_col(
      width = 0.9
    ) +
    coord_flip() +
    scale_fill_gradient2(
      low = "#004C8C",
      mid = "white",
      high = "#A63A2A",
      midpoint = 0,
      limits = c(
        -colour_limit,
        colour_limit
      ),
      oob =
        scales::squish,
      name =
        expression(
          -log[10](p) %*% sign(NES)
        )
    ) +
    labs(
      title =
        paste0(
          layer_label,
          " | GSEA | Gene Ontology"
        ),
      x = NULL,
      y =
        "Normalized enrichment score (NES)"
    ) +
    theme_bw(
      base_size = 10
    ) +
    theme(
      text =
        element_text(
          color = "black"
        ),

      plot.title =
        element_text(
          hjust = 0.5,
          face = "bold",
          size = 11
        ),

      axis.text =
        element_text(
          color = "black"
        ),

      axis.title.x =
        element_text(
          color = "black",
          size = 12
        ),

      axis.text.y =
        element_text(
          size = 9,
          color = "black",
          lineheight = 0.7
        ),

      panel.grid.minor =
        element_blank(),

      legend.position =
        "bottom",

      legend.direction =
        "horizontal",

      legend.title =
        element_text(
          size = 9
        ),

      legend.text =
        element_text(
          size = 9
        )
    ) +
    guides(
      fill =
        guide_colorbar(
          barwidth =
            grid::unit(
              6,
              "cm"
            ),

          barheight =
            grid::unit(
              0.3,
              "cm"
            ),

          title.position =
            "top",

          title.hjust =
            0.5
        )
    )
}


matched_factor2_rna_go_gsea_simplified <-
  matched_rna_gsea_simplified_objects[[
    "Factor2__GO_ALL"
  ]]


matched_factor2_proteomics_go_gsea_simplified <-
  matched_proteomics_gsea_simplified_objects[[
    "Factor2__GO_ALL"
  ]]


matched_factor2_rna_gsea_plot <-
  make_factor2_simplified_gsea_plot(
    gsea_object =
      matched_factor2_rna_go_gsea_simplified,

    layer_label =
      "Transcriptomics"
  )


matched_factor2_proteomics_gsea_plot <-
  make_factor2_simplified_gsea_plot(
    gsea_object =
      matched_factor2_proteomics_go_gsea_simplified,

    layer_label =
      "Proteomics"
  )


matched_factor2_simplified_gsea_plot <-
  matched_factor2_rna_gsea_plot |
  matched_factor2_proteomics_gsea_plot


print(
  matched_factor2_simplified_gsea_plot
)


```

# Supplementary Data export

```{r Supplementary_data_export, eval=TRUE}
mofa_suppdata <- list(
  Histological_fibrosis =
    fibrosis_data,

  Proteomics_samples =
    proteomics_sample_overview,

  Prot_tuning_all =
    proteomics_tuning_results,

  Prot_selected_model =
    proteomics_selected_model_settings,

  Prot_variance_factor =
    proteomics_r2_per_factor_table,

  Prot_factor_scores =
    proteomics_factor_scores,

  Prot_feature_weights =
    proteomics_factor_weights,

  Prot_RV_group =
    proteomics_group_pairwise_results,

  Prot_clinical_assoc =
    proteomics_continuous_results,

  Prot_technical_assoc =
    proteomics_technical_results,

  Prot_F1_GSEA =
    proteomics_factor1_gsea_results,

  Matched_samples =
    matched_sample_overview,

  Matched_tuning_all =
    matched_tuning_results,

  Matched_selected_model =
    matched_selected_model_settings,

  Matched_variance_factor =
    matched_r2_per_factor_table,

  Matched_layer_structure =
    matched_factor_layer_table,

  Matched_factor_scores =
    matched_factor_scores,

  Matched_feature_weights =
    matched_factor_weights,

  Matched_RV_group =
    matched_group_pairwise_results,

  Matched_clinical_assoc =
    matched_continuous_results,

  Matched_fibrosis_assoc =
    matched_fibrosis_results,

  Matched_RNA_F2_GSEA =
    matched_rna_gsea_results,

  Matched_prot_F2_GSEA =
    matched_proteomics_gsea_results
)

writexl::write_xlsx(
  mofa_suppdata,
  file.path(
    base_output_dir,
    "MOFA_v3_SuppData.xlsx"
  )
)
```
