---
title: "Proteomics differential abundance analysis v2"
author: "Gino Bonazza"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  dpi = 300,
  fig.align = "center"
)
```

# Overview

This workflow tests associations between RV tissue protein abundance and
continuous clinical parameters in the Olink Reveal and LC-MS/MS datasets.

Only male DCM patients retained after the proteomics quality-control workflow
and measured by both proteomic platforms are included. Olink and LC-MS/MS are
analysed separately using limma. Models are adjusted for age, with BCA protein
concentration included as an additional covariate in the Olink models.

The analysis produces the protein-count summary, RVIT and TAPSE volcano plots,
and LC-MS/MS heatmaps for proteins significantly associated with TAPSE or RVIT.
No PCA or enrichment analysis is performed.

# Setup

```{r libraries}
library(here)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(patchwork)
library(limma)
library(scales)
library(EnhancedVolcano)
library(ComplexHeatmap)
library(circlize)
library(writexl)
```

```{r analysis choices}
current_file <- "Proteomics_differential_abundance_v2"

output_dir <- here::here(
  "output",
  current_file
)

figure_dir <- here::here(
  "docs",
  "figure",
  paste0(current_file, ".Rmd")
)

dir.create(
  output_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  figure_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

proteomics_qc_file <- here::here(
  "output",
  "Proteomics_quality_control_v2",
  "Proteomics_QC_complete_objects.rds"
)

parameters <- c(
  "mPAP_mmHg",
  "sPAP_mmHg",
  "PVR_WU",
  "PCW_mmHg",
  "RVIT_mm",
  "TAPSE_mm",
  "LVEF_percent",
  "LVEDD_mm"
)

ph_parameters <- c(
  "mPAP_mmHg",
  "sPAP_mmHg",
  "PVR_WU",
  "PCW_mmHg"
)

rv_parameters <- c(
  "RVIT_mm",
  "TAPSE_mm"
)

lv_parameters <- c(
  "LVEF_percent",
  "LVEDD_mm"
)

ordered_parameters <- c(
  ph_parameters,
  rv_parameters,
  lv_parameters
)

pretty_parameter_names <- c(
  mPAP_mmHg = "mPAP (mmHg)",
  sPAP_mmHg = "sPAP (mmHg)",
  PVR_WU = "PVR (WU)",
  PCW_mmHg = "PCWP (mmHg)",
  RVIT_mm = "RVIT (mm)",
  TAPSE_mm = "TAPSE (mm)",
  LVEF_percent = "LVEF (%)",
  LVEDD_mm = "LVEDD (mm)"
)

fdr_cutoff <- 0.20
ms_min_observed_fraction <- 0.70
```

# Helper functions

```{r helper functions}
pretty_var_names <- c(
  mPAP_mmHg = "mPAP (mmHg)",
  sPAP_mmHg = "sPAP (mmHg)",
  PVR_WU = "PVR (WU)",
  PCW_mmHg = "PCWP (mmHg)",
  RVIT_mm = "RVIT (mm)",
  TAPSE_mm = "TAPSE (mm)",
  LVEF_percent = "LVEF (%)",
  LVEDD_mm = "LVEDD (mm)"
)

ordered_params <- ordered_parameters

`%||%` <- function(x, y) if (is.null(x)) y else x


make_clinical_volcano <- function(res, param, pretty_var_names, p_cutoff = 0.2) {
  keyvals <- rep("grey70", nrow(res))
  names(keyvals) <- rep("Not significant", nrow(res))
  keyvals[res$logFC > 0 & res$adj.P.Val < p_cutoff] <- "#A63A2A"
  names(keyvals)[keyvals == "#A63A2A"] <- "Positive"
  keyvals[res$logFC < 0 & res$adj.P.Val < p_cutoff] <- "#004C8C"
  names(keyvals)[keyvals == "#004C8C"] <- "Negative"

  EnhancedVolcano(
    res,
    lab = res$feature,
    x = "logFC",
    y = "adj.P.Val",
    labSize = 0,
    titleLabSize = 15,
    subtitleLabSize = 12,
    axisLabSize = 11,
    captionLabSize = 9,
    pointSize = 0.8,
    FCcutoff = 0,
    pCutoff = p_cutoff,
    ylim = c(0, max(2.5, ceiling(max(-log10(res$adj.P.Val), na.rm = TRUE) + 0.5))),
    colCustom = keyvals,
    colAlpha = 1,
    drawConnectors = FALSE,
    subtitle = NULL,
    title = paste0(pretty_var_names[[param]] %||% param, "\n", "n = ", unique(res$n_complete_samples)),
    xlab = "Effect size (rescaled 0-1 predictor)",
    ylab = "FDR"
  ) + theme(legend.position = "none")
}


build_sig_heatmap <- function(res_tbl,
                              expr_mat,
                              meta_df,
                              param,
                              assay_label,
                              sig_cutoff = 0.2,
                              filter_detect_frac = NULL) {

  meta_sub <- meta_df %>%
    dplyr::select(
      Sample,
      all_of(param)
    ) %>%
    dplyr::distinct(
      Sample,
      .keep_all = TRUE
    ) %>%
    dplyr::filter(
      Sample %in% colnames(expr_mat)
    ) %>%
    dplyr::filter(
      !is.na(.data[[param]])
    ) %>%
    dplyr::arrange(
      .data[[param]]
    )

  expr_sub <- expr_mat[
    ,
    meta_sub$Sample,
    drop = FALSE
  ]

  if (!is.null(filter_detect_frac)) {
    keep_rows <- rowMeans(
      !is.na(expr_sub)
    ) >= filter_detect_frac

    expr_sub <- expr_sub[
      keep_rows,
      ,
      drop = FALSE
    ]
  }

  sig_tbl <- res_tbl %>%
    dplyr::filter(
      adj.P.Val < sig_cutoff
    ) %>%
    dplyr::filter(
      feature %in% rownames(expr_sub)
    ) %>%
    dplyr::arrange(
      dplyr::desc(logFC)
    )

  if (nrow(sig_tbl) == 0) {
    message(
      "No significant features for ",
      assay_label,
      " - ",
      param
    )
    return(NULL)
  }

  mat <- expr_sub[
    sig_tbl$feature,
    meta_sub$Sample,
    drop = FALSE
  ]

  mat_z <- t(
    scale(
      t(mat)
    )
  )

  mat_z[is.nan(mat_z)] <- 0
  mat_z[!is.finite(mat_z)] <- 0

  row_split <- ifelse(
    sig_tbl$logFC > 0,
    paste0(
      "Positive association with ",
      pretty_var_names[[param]] %||% param
    ),
    paste0(
      "Negative association with ",
      pretty_var_names[[param]] %||% param
    )
  )

  ann_df <- data.frame(
    row.names = meta_sub$Sample
  )

  ann_df[[
    pretty_var_names[[param]] %||% param
  ]] <- meta_sub[[param]]

  ann_col_list <- list()

  ann_col_list[[
    pretty_var_names[[param]] %||% param
  ]] <- circlize::colorRamp2(
    c(
      min(
        meta_sub[[param]],
        na.rm = TRUE
      ),
      median(
        meta_sub[[param]],
        na.rm = TRUE
      ),
      max(
        meta_sub[[param]],
        na.rm = TRUE
      )
    ),
    c(
      "#DCEAF7",
      "#6AAED6",
      "#08306B"
    )
  )

  ha <- ComplexHeatmap::HeatmapAnnotation(
    df = ann_df,
    col = ann_col_list,
    annotation_name_gp = grid::gpar(
      fontsize = 9
    )
  )

  row_height_mm <- 2.5

  heatmap_height <- grid::unit(
    nrow(mat_z) *
      row_height_mm,
    "mm"
  )

  ComplexHeatmap::Heatmap(
    mat_z,
    name = "Z-score",
    col = circlize::colorRamp2(
      c(-2, 0, 2),
      c("#2166AC", "white", "#B2182B")
    ),
    top_annotation = ha,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_split = row_split,
    show_row_dend = FALSE,
    show_column_dend = FALSE,
    column_title = paste0(
      assay_label,
      " - significant features for ",
      pretty_var_names[[param]] %||% param
    ),
    row_title = NULL,
    column_names_rot = 45,
    column_names_gp = grid::gpar(
      fontsize = 8
    ),
    row_names_gp = grid::gpar(
      fontsize = 7
    ),
    height = heatmap_height
  )
}
```

# Import filtered proteomics data

```{r import QC object}
if (
  !file.exists(
    proteomics_qc_file
  )
) {
  stop(
    "The filtered proteomics QC object was not found: ",
    proteomics_qc_file
  )
}

proteomics_qc <- readRDS(
  proteomics_qc_file
)

filtered_proteomics <-
  proteomics_qc$after_exclusion
```

```{r matched proteomics samples}
olink_expr_all <-
  filtered_proteomics$olink

ms_expr_all <-
  filtered_proteomics$ms

dcm_samples <-
  filtered_proteomics$clinical %>%
  dplyr::filter(
    Group == "DCM",
    toupper(
      trimws(
        as.character(Sex)
      )
    ) != "F"
  ) %>%
  dplyr::pull(Sample) %>%
  unique()

matched_samples <- dcm_samples[
  dcm_samples %in%
    intersect(
      colnames(
        olink_expr_all
      ),
      colnames(
        ms_expr_all
      )
    )
]

olink_expr <- olink_expr_all[
  ,
  matched_samples,
  drop = FALSE
]

ms_expr <- ms_expr_all[
  ,
  matched_samples,
  drop = FALSE
]

clinical_matched <-
  filtered_proteomics$clinical %>%
  dplyr::filter(
    Sample %in%
      matched_samples
  ) %>%
  dplyr::slice(
    match(
      matched_samples,
      Sample
    )
  ) %>%
  dplyr::distinct(
    Sample,
    .keep_all = TRUE
  )

olink_meta <-
  clinical_matched %>%
  dplyr::left_join(
    filtered_proteomics$olink_metadata %>%
      dplyr::select(
        Sample,
        Batch,
        BCA_mg_ml
      ) %>%
      dplyr::distinct(
        Sample,
        .keep_all = TRUE
      ),
    by = "Sample"
  )

ms_meta <-
  clinical_matched

stopifnot(
  identical(
    colnames(olink_expr),
    matched_samples
  ),
  identical(
    colnames(ms_expr),
    matched_samples
  ),
  identical(
    olink_meta$Sample,
    matched_samples
  ),
  identical(
    ms_meta$Sample,
    matched_samples
  )
)

length(
  matched_samples
)

matched_samples
```

# Differential abundance analysis

## Olink

```{r Olink differential abundance}
covariates_olink <- c(
  "Age_years",
  "BCA_mg_ml"
)

olink_da_results <- list()
olink_da_summary <- list()

for (
  param in parameters
) {
  vars_needed <- unique(
    c(
      param,
      covariates_olink
    )
  )

  meta_da <- olink_meta %>%
    dplyr::select(
      Sample,
      all_of(
        vars_needed
      )
    ) %>%
    dplyr::distinct(
      Sample,
      .keep_all = TRUE
    ) %>%
    dplyr::filter(
      Sample %in%
        colnames(
          olink_expr
        )
    ) %>%
    dplyr::filter(
      complete.cases(.)
    ) %>%
    dplyr::mutate(
      dplyr::across(
        all_of(
          vars_needed
        ),
        scales::rescale
      )
    )

  expr_da <- olink_expr[
    ,
    meta_da$Sample,
    drop = FALSE
  ]

  design_df <- meta_da %>%
    dplyr::select(
      all_of(
        c(
          param,
          covariates_olink
        )
      )
    )

  design <- model.matrix(
    as.formula(
      paste0(
        "~ ",
        paste(
          colnames(
            design_df
          ),
          collapse = " + "
        )
      )
    ),
    data = design_df
  )

  fit <- limma::lmFit(
    expr_da,
    design
  )

  fit <- limma::eBayes(
    fit,
    robust = TRUE
  )

  results <- limma::topTable(
    fit,
    coef = param,
    number = Inf,
    sort.by = "P"
  ) %>%
    tibble::rownames_to_column(
      "feature"
    ) %>%
    dplyr::mutate(
      parameter = param,
      assay = "Olink",
      n_complete_samples =
        nrow(meta_da),
      n_tested_proteins =
        nrow(expr_da),
      regulation =
        dplyr::case_when(
          adj.P.Val <
            fdr_cutoff &
            logFC > 0 ~
            "Up",
          adj.P.Val <
            fdr_cutoff &
            logFC < 0 ~
            "Down",
          TRUE ~ "NS"
        )
    )

  olink_da_results[[
    param
  ]] <- results

  olink_da_summary[[
    param
  ]] <- tibble::tibble(
    assay = "Olink",
    parameter = param,
    direction = c(
      "Up",
      "Down"
    ),
    n_sig = c(
      sum(
        results$adj.P.Val <
          fdr_cutoff &
          results$logFC > 0,
        na.rm = TRUE
      ),
      sum(
        results$adj.P.Val <
          fdr_cutoff &
          results$logFC < 0,
        na.rm = TRUE
      )
    ),
    n_complete_samples =
      nrow(meta_da),
    n_tested_proteins =
      nrow(expr_da)
  )
}

olink_da_all <- dplyr::bind_rows(
  olink_da_results
)

olink_da_summary_df <-
  dplyr::bind_rows(
    olink_da_summary
  )
```

## LC-MS/MS

```{r LCMS differential abundance}
covariates_ms <- c(
  "Age_years"
)

ms_da_results <- list()
ms_da_summary <- list()

for (
  param in parameters
) {
  vars_needed <- unique(
    c(
      param,
      covariates_ms
    )
  )

  meta_da <- ms_meta %>%
    dplyr::select(
      Sample,
      all_of(
        vars_needed
      )
    ) %>%
    dplyr::distinct(
      Sample,
      .keep_all = TRUE
    ) %>%
    dplyr::filter(
      Sample %in%
        colnames(
          ms_expr
        )
    ) %>%
    dplyr::filter(
      complete.cases(.)
    ) %>%
    dplyr::mutate(
      dplyr::across(
        all_of(
          vars_needed
        ),
        scales::rescale
      )
    )

  expr_da_initial <- ms_expr[
    ,
    meta_da$Sample,
    drop = FALSE
  ]

  keep_proteins <- rowMeans(
    is.finite(
      expr_da_initial
    )
  ) >=
    ms_min_observed_fraction

  expr_da <- expr_da_initial[
    keep_proteins,
    ,
    drop = FALSE
  ]

  design_df <- meta_da %>%
    dplyr::select(
      all_of(
        c(
          param,
          covariates_ms
        )
      )
    )

  design <- model.matrix(
    as.formula(
      paste0(
        "~ ",
        paste(
          colnames(
            design_df
          ),
          collapse = " + "
        )
      )
    ),
    data = design_df
  )

  fit <- limma::lmFit(
    expr_da,
    design,
    na.action =
      na.exclude
  )

  fit <- limma::eBayes(
    fit,
    trend = TRUE,
    robust = TRUE
  )

  results <- limma::topTable(
    fit,
    coef = param,
    number = Inf,
    sort.by = "P"
  ) %>%
    tibble::rownames_to_column(
      "feature"
    ) %>%
    dplyr::mutate(
      parameter = param,
      assay = "LC-MS/MS",
      n_complete_samples =
        nrow(meta_da),
      n_tested_proteins =
        nrow(expr_da),
      regulation =
        dplyr::case_when(
          adj.P.Val <
            fdr_cutoff &
            logFC > 0 ~
            "Up",
          adj.P.Val <
            fdr_cutoff &
            logFC < 0 ~
            "Down",
          TRUE ~ "NS"
        )
    )

  ms_da_results[[
    param
  ]] <- results

  ms_da_summary[[
    param
  ]] <- tibble::tibble(
    assay = "LC-MS/MS",
    parameter = param,
    direction = c(
      "Up",
      "Down"
    ),
    n_sig = c(
      sum(
        results$adj.P.Val <
          fdr_cutoff &
          results$logFC > 0,
        na.rm = TRUE
      ),
      sum(
        results$adj.P.Val <
          fdr_cutoff &
          results$logFC < 0,
        na.rm = TRUE
      )
    ),
    n_complete_samples =
      nrow(meta_da),
    n_tested_proteins =
      nrow(expr_da)
  )
}

ms_da_all <- dplyr::bind_rows(
  ms_da_results
)

ms_da_summary_df <-
  dplyr::bind_rows(
    ms_da_summary
  )
```

# Number of significant proteins

```{r differential abundance summary, fig.width=3, fig.height=5, dpi=300}
deg_summary <- bind_rows(olink_da_summary_df, ms_da_summary_df) %>%
  dplyr::mutate(
    assay = dplyr::recode(
      assay,
      "LC-MS/MS" = "MS",
      "Mass spec" = "MS"
    ),
    parameter = factor(parameter, levels = ordered_params),
    parameter_label = factor(
      pretty_var_names[as.character(parameter)],
      levels = pretty_var_names[ordered_params]
    ),
    assay = factor(assay, levels = c("Olink", "MS")),
    direction = factor(direction, levels = c("Up", "Down"))
  )

ggplot(deg_summary, aes(x = assay, y = parameter_label, fill = n_sig)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = n_sig), size = 4) +
  scale_fill_gradient(low = "grey95", high = "#A63A2A") +
  facet_wrap(~ direction, ncol = 2) +
  labs(
    x = NULL,
    y = NULL,
    fill = "n proteins"
  ) +
  theme_classic(base_size = 13) +
  theme(
    axis.line = element_blank(),
    panel.border = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  guides(
    fill = guide_colorbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth = grid::unit(4.5, "cm"),
      barheight = grid::unit(0.4, "cm")
    )
  )
```

# RVIT volcano plots

```{r RVIT volcano plots, fig.width=6, fig.height=4, dpi=300}
p_RVIT_olink <- make_clinical_volcano(
  res = olink_da_results[["RVIT_mm"]],
  param = "RVIT_mm",
  pretty_var_names = pretty_var_names,
  p_cutoff = 0.2
) + ggtitle("Olink - RVIT")

p_RVIT_ms <- make_clinical_volcano(
  res = ms_da_results[["RVIT_mm"]],
  param = "RVIT_mm",
  pretty_var_names = pretty_var_names,
  p_cutoff = 0.2
) + ggtitle("MS - RVIT")

p_RVIT_olink + p_RVIT_ms
```

# TAPSE volcano plots

```{r TAPSE volcano plots, fig.width=6, fig.height=4, dpi=300}
p_TAPSE_olink <- make_clinical_volcano(
  res = olink_da_results[["TAPSE_mm"]],
  param = "TAPSE_mm",
  pretty_var_names = pretty_var_names,
  p_cutoff = 0.2
) + ggtitle("Olink - TAPSE")

p_TAPSE_ms <- make_clinical_volcano(
  res = ms_da_results[["TAPSE_mm"]],
  param = "TAPSE_mm",
  pretty_var_names = pretty_var_names,
  p_cutoff = 0.2
) + ggtitle("MS - TAPSE")

p_TAPSE_olink + p_TAPSE_ms
```

# LC-MS/MS TAPSE heatmap

```{r TAPSE heatmap, fig.width=9, fig.height=6.5, dpi=300}
ms_TAPSE_heat <- build_sig_heatmap(
  res_tbl = ms_da_results[["TAPSE_mm"]],
  expr_mat = ms_expr,
  meta_df = ms_meta,
  param = "TAPSE_mm",
  assay_label = "MS",
  sig_cutoff = 0.2,
  filter_detect_frac = 0.70
)

if (!is.null(ms_TAPSE_heat)) {
  ComplexHeatmap::draw(
    ms_TAPSE_heat,
    heatmap_legend_side = "right",
    annotation_legend_side = "right",
    padding = grid::unit(c(2, 8, 2, 2), "mm")
  )
}
```

# LC-MS/MS RVIT heatmap

```{r RVIT heatmap, fig.width=9, fig.height=9, dpi=300}
ms_RVIT_heat <- build_sig_heatmap(
  res_tbl = ms_da_results[["RVIT_mm"]],
  expr_mat = ms_expr,
  meta_df = ms_meta,
  param = "RVIT_mm",
  assay_label = "MS",
  sig_cutoff = 0.2,
  filter_detect_frac = 0.70
)

if (!is.null(ms_RVIT_heat)) {
  ComplexHeatmap::draw(
    ms_RVIT_heat,
    heatmap_legend_side = "right",
    annotation_legend_side = "right",
    padding = grid::unit(c(2, 8, 2, 2), "mm")
  )
}
```

# Supplementary Data export

```{r Supplementary_data_export, eval=TRUE}
matched_sample_table <- clinical_matched %>%
  dplyr::select(
    Sample,
    Age_years,
    mPAP_mmHg,
    sPAP_mmHg,
    PVR_WU,
    PCW_mmHg,
    RVIT_mm,
    TAPSE_mm,
    LVEF_percent,
    LVEDD_mm
  ) %>%
  dplyr::left_join(
    olink_meta %>%
      dplyr::select(
        Sample,
        BCA_mg_ml
      ),
    by = "Sample"
  )

olink_TAPSE_sig <- olink_da_results[[
  "TAPSE_mm"
]] %>%
  dplyr::filter(
    adj.P.Val <
      fdr_cutoff
  )

olink_RVIT_sig <- olink_da_results[[
  "RVIT_mm"
]] %>%
  dplyr::filter(
    adj.P.Val <
      fdr_cutoff
  )

lcms_TAPSE_sig <- ms_da_results[[
  "TAPSE_mm"
]] %>%
  dplyr::filter(
    adj.P.Val <
      fdr_cutoff
  )

lcms_RVIT_sig <- ms_da_results[[
  "RVIT_mm"
]] %>%
  dplyr::filter(
    adj.P.Val <
      fdr_cutoff
  )

writexl::write_xlsx(
  list(
    Matched_samples =
      matched_sample_table,
    DA_summary =
      dplyr::bind_rows(
        olink_da_summary_df,
        ms_da_summary_df
      ),
    Olink_all_results =
      olink_da_all,
    LCMS_all_results =
      ms_da_all,
    Olink_TAPSE_sig =
      olink_TAPSE_sig,
    Olink_RVIT_sig =
      olink_RVIT_sig,
    LCMS_TAPSE_sig =
      lcms_TAPSE_sig,
    LCMS_RVIT_sig =
      lcms_RVIT_sig
  ),
  file.path(
    output_dir,
    "Proteomics_differential_abundance_v2_SuppData.xlsx"
  )
)
```
