---
title: "Proteomics PCA and cell-state composition"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source="fold-hide"}
current_file <- "Proteomics_PCA_cell_state_composition"

library(here)
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(magrittr)
library(patchwork)
library(pheatmap)
library(scales)
library(qs2)
library(Seurat)
library(matrixStats)
library(OlinkAnalyze)
library(reactable)
library(ggrepel)
library(purrr)
library(stringr)

`%||%` <- function(x, y) if (is.null(x)) y else x

output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))
output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs, recursive = TRUE)

set.seed(20260430)
```

## Analysis choices

```{r analysis choices}
excluded_samples <- c("PK211", "P11", "PK69", "PK61", "PK63")
allowed_etiology <- c("DCM", "DCM_LVNC")

pcs_to_test <- 1:5
pcs_to_plot <- 1:5
n_pcs_scree <- 10
top_n_loadings <- 20

# Minimum number of matched proteomics samples in which a state must be present.
# This keeps rare states such as B cells if present in at least 10 matched samples.
min_samples_present <- 10

pretty_var_names <- c(
  Age_years = "Age (years)",
  Sample_age_days = "Sample age (days)",
  LVEF_percent = "LVEF (%)",
  LVEDD_mm = "LVEDD (mm)",
  TAPSE_mm = "TAPSE (mm)",
  RVIT_mm = "RVIT (mm)",
  sPAP_mmHg = "sPAP (mmHg)",
  mPAP_mmHg = "mPAP (mmHg)",
  PCW_mmHg = "PCW (mmHg)",
  PVR_WU = "PVR (WU)",
  BCA_mg_ml = "BCA (mg/ml)",
  Batch = "Batch"
)

heat_cols <- colorRampPalette(c("#2166AC", "white", "#B2182B"))(100)
```

## Helper functions

```{r helper functions}
find_existing_file <- function(candidates) {
  hit <- candidates[file.exists(candidates)][1]
  if (length(hit) == 0 || is.na(hit)) return(NA_character_)
  hit
}

read_olink_delim <- function(path) {
  x <- suppressMessages(readr::read_delim(path, delim = "\t", col_types = readr::cols(), progress = FALSE))
  if (ncol(x) <= 1) x <- suppressMessages(readr::read_delim(path, delim = ";", col_types = readr::cols(), progress = FALSE))
  if (ncol(x) <= 1) x <- suppressMessages(readr::read_csv(path, col_types = readr::cols(), progress = FALSE))
  x
}

simple_impute_for_pca <- function(mat) {
  out <- as.matrix(mat)
  global_floor <- suppressWarnings(stats::quantile(out, probs = 0.01, na.rm = TRUE))
  if (!is.finite(global_floor)) global_floor <- 0

  for (i in seq_len(nrow(out))) {
    miss <- is.na(out[i, ]) | !is.finite(out[i, ])
    if (!any(miss)) next

    obs <- out[i, !miss]
    if (length(obs) >= 3) {
      mu <- mean(obs, na.rm = TRUE)
      sdv <- stats::sd(obs, na.rm = TRUE)
      if (!is.finite(sdv) || sdv == 0) sdv <- 0.3
      out[i, miss] <- stats::rnorm(sum(miss), mean = mu - 1.8 * sdv, sd = max(0.05, 0.3 * sdv))
    } else {
      out[i, miss] <- global_floor
    }
  }

  out[!is.finite(out)] <- global_floor
  out
}

logit_prop <- function(x, eps = 0.001) {
  x <- pmin(pmax(x, eps), 1 - eps)
  log(x / (1 - x))
}

detect_sample_col <- function(obj) {
  candidates <- c("Sample", "sample", "orig.ident")
  hit <- candidates[candidates %in% colnames(obj@meta.data)][1]
  if (length(hit) == 0 || is.na(hit)) {
    stop("Could not detect a sample column. Check object metadata and set it manually.")
  }
  hit
}

detect_state_col <- function(obj) {
  candidates <- c(
    "cell_state", "subcluster", "cell_subtype", "cell_type_sub", "celltype_sub",
    "annotation_sub", "state", "cell_type"
  )
  hit <- candidates[candidates %in% colnames(obj@meta.data)][1]
  if (length(hit) == 0 || is.na(hit)) {
    stop("Could not detect a state column. Check object metadata and set it manually.")
  }
  hit
}

run_pca_from_matrix <- function(expr_mat, meta_df, assay_label) {
  sample_keep <- intersect(meta_df$Sample, colnames(expr_mat))

  meta_df <- meta_df %>%
    dplyr::filter(Sample %in% sample_keep) %>%
    dplyr::slice(match(sample_keep, Sample))

  expr_mat <- expr_mat[, meta_df$Sample, drop = FALSE]
  expr_pca <- simple_impute_for_pca(expr_mat)

  feature_sds <- matrixStats::rowSds(expr_pca, na.rm = TRUE)
  keep_feature <- is.finite(feature_sds) & feature_sds > 0
  expr_pca <- expr_pca[keep_feature, , drop = FALSE]

  pca <- prcomp(t(expr_pca), center = TRUE, scale. = TRUE)
  pvar <- 100 * (pca$sdev^2 / sum(pca$sdev^2))

  scores <- as.data.frame(pca$x) %>%
    tibble::rownames_to_column("Sample") %>%
    dplyr::left_join(meta_df, by = "Sample") %>%
    dplyr::mutate(assay = assay_label)

  list(
    pca = pca,
    pvar = pvar,
    scores = scores,
    expr_pca = expr_pca,
    meta = meta_df
  )
}

make_scree_plot <- function(pca_list, assay_label, n_pcs = 10) {
  n_pcs <- min(n_pcs, length(pca_list$pvar))

  df <- tibble(
    PC = factor(paste0("PC", seq_len(n_pcs)), levels = paste0("PC", seq_len(n_pcs))),
    variance = pca_list$pvar[seq_len(n_pcs)],
    cumulative = cumsum(pca_list$pvar[seq_len(n_pcs)])
  )

  ggplot(df, aes(x = PC, y = variance, group = 1)) +
    geom_col() +
    geom_line(aes(y = cumulative / max(cumulative) * max(variance))) +
    geom_point(aes(y = cumulative / max(cumulative) * max(variance))) +
    labs(
      title = paste0(assay_label, " scree / elbow plot"),
      x = NULL,
      y = "% variance explained"
    ) +
    theme_classic(base_size = 12)
}

make_pca_plot <- function(pca_list, color_by, title_text) {
  df <- pca_list$scores
  pvar <- pca_list$pvar

  if (!color_by %in% colnames(df)) {
    return(
      ggplot() +
        annotate("text", x = 1, y = 1, label = paste0(color_by, " not available")) +
        theme_void() +
        ggtitle(title_text)
    )
  }

  ggplot(df, aes(x = PC1, y = PC2, color = .data[[color_by]], label = Sample)) +
    geom_point(size = 3) +
    ggrepel::geom_text_repel(size = 2.5, max.overlaps = 30, show.legend = FALSE) +
    labs(
      title = title_text,
      x = paste0("PC1 (", round(pvar[1], 1), "%)"),
      y = paste0("PC2 (", round(pvar[2], 1), "%)"),
      color = pretty_var_names[[color_by]] %||% color_by
    ) +
    theme_classic(base_size = 12)
}

get_top_pca_features <- function(pca_obj, pcs = 1:5, top_n = 20) {
  loadings <- pca_obj$rotation[, paste0("PC", pcs), drop = FALSE]

  purrr::map_dfr(seq_along(pcs), function(i) {
    pc_name <- colnames(loadings)[i]
    vals <- loadings[, i]

    tibble(
      feature = names(vals),
      loading = as.numeric(vals),
      abs_loading = abs(as.numeric(vals)),
      PC = pc_name
    ) %>%
      dplyr::arrange(dplyr::desc(abs_loading)) %>%
      dplyr::slice_head(n = top_n)
  })
}

make_loading_heatmap <- function(pca_list, top_tbl, pc, title_text, annotation_col = NULL) {
  feat <- top_tbl %>%
    dplyr::filter(PC == pc) %>%
    dplyr::arrange(dplyr::desc(abs_loading)) %>%
    dplyr::pull(feature)

  hm <- pca_list$expr_pca[feat, , drop = FALSE]
  hm_z <- t(scale(t(hm)))
  hm_z[is.nan(hm_z)] <- 0
  hm_z[!is.finite(hm_z)] <- 0

  row_lab <- top_tbl %>%
    dplyr::filter(PC == pc) %>%
    dplyr::mutate(label = paste0(feature, " (", sprintf("%.3f", loading), ")")) %>%
    dplyr::select(feature, label)

  rownames(hm_z) <- row_lab$label[match(rownames(hm_z), row_lab$feature)]

  if (!is.null(annotation_col)) {
    annotation_col <- annotation_col[colnames(hm_z), , drop = FALSE]
  }

  pheatmap(
    hm_z,
    annotation_col = annotation_col,
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    show_rownames = TRUE,
    show_colnames = TRUE,
    fontsize_row = 8,
    fontsize_col = 8,
    border_color = NA,
    color = heat_cols,
    main = title_text
  )
}

extract_state_map <- function(file, family, prefix = NULL, state_col = NULL) {
  if (is.na(file)) {
    message("Skipping ", family, ": object not found.")
    return(NULL)
  }

  obj <- qs2::qs_read(file, nthreads = 32)

  if (is.null(state_col)) {
    state_col <- detect_state_col(obj)
  }

  meta <- obj@meta.data %>%
    tibble::rownames_to_column("cell") %>%
    dplyr::mutate(
      final_state = as.character(.data[[state_col]]),
      state_family = family
    ) %>%
    dplyr::filter(!is.na(final_state), final_state != "")

  if (!is.null(prefix)) {
    numeric_state <- grepl("^[0-9]+$", meta$final_state)
    meta$final_state[numeric_state] <- paste0(prefix, meta$final_state[numeric_state])
  }

  meta %>%
    dplyr::select(cell, final_state, state_family) %>%
    dplyr::distinct(cell, .keep_all = TRUE)
}

make_state_order <- function(state_info) {
  family_levels <- c(
    "Cardiomyocytes",
    "Fibroblasts",
    "Endothelial cells",
    "Mural cells",
    "Macrophages",
    "Lymphocytes",
    "Other"
  )

  state_info %>%
    dplyr::mutate(
      state_family = factor(state_family, levels = family_levels),
      state_number = readr::parse_number(final_state),
      other_order = dplyr::case_when(
        final_state == "Endocardial cells" ~ 1,
        final_state == "Adipocytes" ~ 2,
        final_state == "Mast cells" ~ 3,
        final_state == "Lymphatic cells" ~ 4,
        final_state == "Neuronal cells" ~ 5,
        TRUE ~ 99
      )
    ) %>%
    dplyr::arrange(state_family, other_order, state_number, final_state) %>%
    dplyr::pull(final_state)
}

fit_pc_state_models <- function(scores_df, prop_df, state_info, assay = c("Olink", "MS"), pcs = 1:5,
                                min_samples_present = 10) {
  assay <- match.arg(assay)

  dat <- scores_df %>%
    dplyr::left_join(prop_df, by = "Sample")

  pc_names <- intersect(paste0("PC", pcs), colnames(dat))
  prop_cols_all <- grep("^prop_STATE_", colnames(dat), value = TRUE)

  prevalence <- purrr::map_dfr(prop_cols_all, function(prop_col) {
    final_state <- sub("^prop_STATE_", "", prop_col)

    tibble(
      final_state = final_state,
      prop_col = prop_col,
      n_present = sum(dat[[prop_col]] > 0, na.rm = TRUE),
      max_prop = max(dat[[prop_col]], na.rm = TRUE)
    )
  })

  prop_cols <- prevalence %>%
    dplyr::filter(n_present >= min_samples_present) %>%
    dplyr::pull(prop_col)

  logit_cols <- gsub("^prop_", "logit_prop_", prop_cols)

  covars <- c("Age_years")
  if (assay == "Olink") covars <- c(covars, "BCA_mg_ml")
  covars <- covars[covars %in% colnames(dat)]

  res <- purrr::map_dfr(pc_names, function(pc) {
    purrr::map2_dfr(prop_cols, logit_cols, function(prop_col, logit_col) {
      final_state <- sub("^prop_STATE_", "", prop_col)
      cols_needed <- c(pc, prop_col, logit_col, covars)

      df <- dat %>%
        dplyr::select(Sample, dplyr::all_of(cols_needed)) %>%
        tidyr::drop_na()

      if (nrow(df) < 5 || stats::sd(df[[logit_col]], na.rm = TRUE) == 0) {
        return(tibble(
          assay = assay,
          PC = pc,
          final_state = final_state,
          n = nrow(df),
          estimate = NA_real_,
          p_value = NA_real_,
          rho_prop = NA_real_,
          p_prop = NA_real_
        ))
      }

      form <- reformulate(termlabels = c(pc, covars), response = logit_col)
      fit <- lm(form, data = df)
      coef_tab <- summary(fit)$coefficients

      ct <- suppressWarnings(cor.test(df[[pc]], df[[prop_col]], method = "spearman"))

      tibble(
        assay = assay,
        PC = pc,
        final_state = final_state,
        n = nrow(df),
        estimate = coef_tab[pc, "Estimate"],
        p_value = coef_tab[pc, "Pr(>|t|)"],
        rho_prop = unname(ct$estimate),
        p_prop = ct$p.value
      )
    })
  })

  res %>%
    dplyr::left_join(state_info, by = "final_state") %>%
    dplyr::left_join(prevalence %>% dplyr::select(final_state, n_present, max_prop), by = "final_state") %>%
    dplyr::mutate(
      p_adj = p.adjust(p_value, method = "BH"),
      neg_log10_padj = -log10(p_adj),
      signed_neg_log10_padj = sign(estimate) * neg_log10_padj
    ) %>%
    dplyr::arrange(p_adj, p_value)
}

make_pc_state_dotplot <- function(res_tbl, state_order, title_text) {
  plot_df <- res_tbl %>%
    dplyr::filter(!is.na(estimate), final_state %in% state_order) %>%
    dplyr::mutate(
      PC = factor(PC, levels = paste0("PC", pcs_to_test)),
      final_state = factor(final_state, levels = state_order),
      fdr_size = -log10(p_adj),
      fdr_outline = p_adj < 0.2
    )

  size_cap <- max(2, stats::quantile(plot_df$fdr_size, probs = 0.95, na.rm = TRUE))
  plot_df <- plot_df %>%
    dplyr::mutate(fdr_size_capped = pmin(fdr_size, size_cap))

  ggplot(plot_df, aes(x = final_state, y = PC)) +
    geom_point(
      aes(size = fdr_size_capped, fill = estimate),
      shape = 21,
      color = "grey75",
      stroke = 0.3
    ) +
    geom_point(
      data = plot_df %>% dplyr::filter(fdr_outline),
      aes(size = fdr_size_capped, fill = estimate),
      shape = 21,
      color = "black",
      stroke = 0.7
    ) +
    scale_fill_gradient2(
      low = "#2166AC",
      mid = "white",
      high = "#B2182B",
      midpoint = 0,
      name = "Beta"
    ) +
    scale_size_continuous(
      range = c(1.2, 6),
      name = "-log10(FDR)"
    ) +
    labs(
      title = title_text,
      x = NULL,
      y = NULL
    ) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1, size = 8),
      axis.text.y = element_text(size = 10),
      legend.position = "right"
    )
}
```

## Read metadata

```{r read metadata}
meta_file <- here::here("data", "Baza_dla_Gino_AP_short.xlsx")
if (!file.exists(meta_file)) stop("Could not find metadata file: ", meta_file)

clin_all <- readxl::read_excel(meta_file) %>%
  dplyr::select(-dplyr::matches("^Unnamed")) %>%
  dplyr::mutate(
    Collection_date = as.Date(Collection_date),
    Used_date = as.Date(Used_date),
    Sample_age_days = as.numeric(Sample_age_days),
    Age_years = as.numeric(Age_years),
    LVEF_percent = as.numeric(LVEF_percent),
    LVEDD_mm = as.numeric(LVEDD_mm),
    TAPSE_mm = as.numeric(TAPSE_mm),
    RVIT_mm = as.numeric(RVIT_mm),
    sPAP_mmHg = as.numeric(sPAP_mmHg),
    mPAP_mmHg = as.numeric(mPAP_mmHg),
    PCW_mmHg = as.numeric(PCW_mmHg),
    PVR_WU = as.numeric(PVR_WU),
    Sex = factor(Sex),
    Group = factor(Group),
    Etiology = factor(Etiology)
  ) %>%
  dplyr::filter(
    !Sample %in% excluded_samples,
    Group == "DCM",
    Etiology %in% allowed_etiology
  ) %>%
  dplyr::distinct(Sample, .keep_all = TRUE)
```

## Read Olink data

```{r read olink}
olink_meta_file <- here::here("data", "OLINK", "Samples_RV_Olink.xlsx")
if (!file.exists(olink_meta_file)) stop("Could not find Olink sample file: ", olink_meta_file)

olink_sample_meta <- readxl::read_excel(olink_meta_file) %>%
  dplyr::select(-dplyr::matches("^Unnamed")) %>%
  dplyr::filter(Olink == "Yes") %>%
  dplyr::filter(!Sample %in% excluded_samples) %>%
  dplyr::mutate(BCA_mg_ml = as.numeric(BCA_mg_ml)) %>%
  dplyr::distinct(Sample, .keep_all = TRUE)

olink_meta <- clin_all %>%
  dplyr::inner_join(
    olink_sample_meta %>% dplyr::select(Sample, Batch, BCA_mg_ml),
    by = "Sample"
  ) %>%
  dplyr::distinct(Sample, .keep_all = TRUE)

olink_raw <- read.delim(here::here("data", "OLINK", "quant.xls"), sep = "\t", check.names = FALSE)
rownames(olink_raw) <- sub("_.*", "", olink_raw$SampleID)

olink_expr <- olink_raw[, -1, drop = FALSE]
olink_expr <- as.matrix(olink_expr)
storage.mode(olink_expr) <- "numeric"

olink_keep <- intersect(olink_meta$Sample, colnames(olink_expr))
olink_meta <- olink_meta %>%
  dplyr::filter(Sample %in% olink_keep) %>%
  dplyr::slice(match(olink_keep, Sample))
olink_expr <- olink_expr[, olink_keep, drop = FALSE]
```

## Olink LOD filtering

```{r olink lod qc}
min_frac_above_lod <- 0.40

olink_lod_data_candidates <- c(
  here::here("data", "OLINK", "pre_quant.csv"),
  here::here("data", "OLINK", "pre_quant.tsv"),
  here::here("data", "OLINK", "npx.csv"),
  here::here("data", "OLINK", "npx.tsv")
)
olink_lod_data_file <- find_existing_file(olink_lod_data_candidates)

fixedLOD_candidates <- c(
  here::here("data", "OLINK", "Reveal_fixedLOD.csv"),
  here::here("data", "OLINK", "fixedLOD.csv"),
  here::here("data", "OLINK", "FixedLOD.csv")
)
fixedLOD_path <- find_existing_file(fixedLOD_candidates)

if (is.na(olink_lod_data_file)) stop("Could not find Olink long-format NPX file for LOD QC.")
if (is.na(fixedLOD_path)) stop("Could not find Olink fixed LOD file.")

npx_raw <- read_olink_delim(olink_lod_data_file)

npx_lod <- OlinkAnalyze::olink_lod(
  data = npx_raw,
  lod_file_path = fixedLOD_path,
  lod_method = "FixedLOD"
)

lod_info <- read_olink_delim(fixedLOD_path)
if (all(c("OlinkID", "DataAnalysisRefID") %in% colnames(lod_info))) {
  lod_info <- lod_info %>%
    dplyr::select(dplyr::any_of(c("OlinkID", "DataAnalysisRefID", "LODNPX", "LODCount", "LODMethod")))
  npx_lod <- npx_lod %>%
    dplyr::left_join(lod_info, by = c("OlinkID", "DataAnalysisRefID"))
}

for (nm in c("LOD", "LODCount", "Count", "NPX")) {
  if (!nm %in% colnames(npx_lod)) npx_lod[[nm]] <- NA_real_
}

npx_lod <- npx_lod %>%
  dplyr::mutate(
    above_LOD = dplyr::case_when(
      !is.na(LOD) ~ NPX > LOD,
      !is.na(LODCount) ~ Count > LODCount,
      TRUE ~ NA
    )
  )

feature_candidates <- intersect(c("Assay", "UniProt", "OlinkID"), colnames(npx_lod))
feature_match_stats <- purrr::map_dfr(feature_candidates, function(feature_col) {
  tibble(
    feature_col = feature_col,
    n_matches = sum(unique(as.character(npx_lod[[feature_col]])) %in% rownames(olink_expr))
  )
})

feature_col_for_lod <- feature_match_stats %>%
  dplyr::arrange(dplyr::desc(n_matches)) %>%
  dplyr::slice(1) %>%
  dplyr::pull(feature_col)

olink_lod_summary <- npx_lod %>%
  dplyr::mutate(
    Sample = as.character(SampleID),
    feature = as.character(.data[[feature_col_for_lod]]),
    above_LOD = dplyr::if_else(is.na(above_LOD), FALSE, above_LOD)
  ) %>%
  dplyr::filter(Sample %in% olink_meta$Sample, feature %in% rownames(olink_expr)) %>%
  dplyr::distinct(Sample, feature, .keep_all = TRUE) %>%
  dplyr::group_by(feature) %>%
  dplyr::summarise(
    n_samples_lod = dplyr::n(),
    n_above_LOD = sum(above_LOD, na.rm = TRUE),
    frac_above_LOD = n_above_LOD / n_samples_lod,
    .groups = "drop"
  ) %>%
  dplyr::mutate(keep_lod = frac_above_LOD >= min_frac_above_lod)

olink_keep_features_lod <- olink_lod_summary %>%
  dplyr::filter(keep_lod) %>%
  dplyr::pull(feature)

olink_expr <- olink_expr[intersect(rownames(olink_expr), olink_keep_features_lod), , drop = FALSE]
```

## Read mass spectrometry data

```{r read mass spec}
ms_raw <- read.csv(here::here("data", "Mass_spec", "20260319_MS_DIA_report_2uniquepeptides.csv"), check.names = FALSE)

prot_annot <- ms_raw[, c(
  "Protein.Group", "Protein.Names", "Genes", "First.Protein.Description",
  "N.Sequences", "N.Proteotypic.Sequences"
)]

sample_cols <- setdiff(colnames(ms_raw), colnames(prot_annot))
qc_cols <- grep("^QC", sample_cols, value = TRUE)
biol_cols <- setdiff(sample_cols, qc_cols)

ms_expr0 <- as.data.frame(ms_raw[, biol_cols, drop = FALSE])
rownames(ms_expr0) <- ms_raw$Genes
keep_rows <- !is.na(rownames(ms_expr0)) & rownames(ms_expr0) != ""
ms_expr0 <- ms_expr0[keep_rows, , drop = FALSE]
rownames(ms_expr0) <- make.unique(rownames(ms_expr0))
ms_expr0 <- as.matrix(ms_expr0)
storage.mode(ms_expr0) <- "numeric"

ms_meta <- clin_all %>%
  dplyr::filter(Sample %in% colnames(ms_expr0)) %>%
  dplyr::distinct(Sample, .keep_all = TRUE)

ms_expr0 <- ms_expr0[, ms_meta$Sample, drop = FALSE]
ms_expr <- log2(ms_expr0 + 1)
```

## Import cell states and create final cell-state assignment

```{r import cell states}
dcm_rv_file <- find_existing_file(c(
  here::here("output", "QC_integration_annotation", "DCM_RV.qs2"),
  here::here("output", "QC_integration_annotation", "DCM_RV_clean.qs2"),
  here::here("output", "QC_integration_annotation", "DCM_RV_annotated.qs2"),
  here::here("output", "QC_integration_annotation", "RV.qs2"),
  here::here("output", "DCM_RV.qs2"),
  here::here("data", "DCM_RV.qs2")
))

if (is.na(dcm_rv_file)) stop("Could not find the full DCM_RV object.")

DCM_RV <- qs2::qs_read(dcm_rv_file, nthreads = 32)

full_cell_type_col <- dplyr::case_when(
  "cell_type" %in% colnames(DCM_RV@meta.data) ~ "cell_type",
  "celltype" %in% colnames(DCM_RV@meta.data) ~ "celltype",
  "annotation" %in% colnames(DCM_RV@meta.data) ~ "annotation",
  TRUE ~ NA_character_
)

if (is.na(full_cell_type_col)) stop("Could not detect a general cell-type column in DCM_RV@meta.data.")

full_sample_col <- detect_sample_col(DCM_RV)

subcluster_files <- list(
  CM = find_existing_file(c(
    here::here("output", "Cardiomyocytes_subclustering", "CM.qs2"),
    here::here("output", "Cardiomyocytes_subclustering", "Cardiomyocytes.qs2")
  )),
  FB = find_existing_file(c(
    here::here("output", "Fibroblasts_subclustering", "FB.qs2"),
    here::here("output", "Fibroblasts_subclustering", "Fibroblasts.qs2")
  )),
  EC = find_existing_file(c(
    here::here("output", "Endothelial_cells_subclustering", "EC.qs2"),
    here::here("output", "Endothelial_cells_subclustering", "Endothelial_cells.qs2")
  )),
  Mural = find_existing_file(c(
    here::here("output", "Mural_cells_subclustering", "MC.qs2"),
    here::here("output", "Mural_cells_subclustering", "Mural.qs2"),
    here::here("output", "Mural_cells_subclustering", "Mural_cells.qs2")
  )),
  Macrophages = find_existing_file(c(
    here::here("output", "Macrophages_subclustering", "MP.qs2"),
    here::here("output", "Macrophages_subclustering", "Macrophages.qs2"),
    here::here("output", "Macrophages_subclustering", "Macrophage.qs2"),
    here::here("output", "Macrophages_subclustering", "Mac.qs2")
  )),
  Lymphocytes = find_existing_file(c(
    here::here("output", "Lymphocytes_subclustering", "LC.qs2"),
    here::here("output", "Lymphocytes_subclustering", "Lymphocytes.qs2"),
    here::here("output", "Lymphocytes_subclustering", "Lymph.qs2")
  ))
)

subcluster_files

state_map_list <- list(
  CM = extract_state_map(subcluster_files$CM, family = "Cardiomyocytes", prefix = "CM"),
  FB = extract_state_map(subcluster_files$FB, family = "Fibroblasts", prefix = "FB"),
  EC = extract_state_map(subcluster_files$EC, family = "Endothelial cells", prefix = "EC"),
  Mural = extract_state_map(subcluster_files$Mural, family = "Mural cells", prefix = "MC"),
  Macrophages = extract_state_map(subcluster_files$Macrophages, family = "Macrophages", prefix = "MP"),
  Lymphocytes = extract_state_map(subcluster_files$Lymphocytes, family = "Lymphocytes", prefix = "LC")
)

subcluster_state_map <- dplyr::bind_rows(state_map_list)

subclustered_main_types <- c(
  "Cardiomyocytes",
  "Fibroblasts",
  "Endothelial cells",
  "Pericytes",
  "Smooth muscle cells",
  "Macrophages",
  "T lymphocytes",
  "B lymphocytes"
)

non_subclustered_main_types <- c(
  "Adipocytes",
  "Endocardial cells",
  "Mast cells",
  "Lymphatic cells",
  "Neuronal cells"
)

full_meta_state <- DCM_RV@meta.data %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::mutate(
    Sample = as.character(.data[[full_sample_col]]),
    main_cell_type = as.character(.data[[full_cell_type_col]])
  ) %>%
  dplyr::left_join(subcluster_state_map, by = "cell") %>%
  dplyr::mutate(
    final_state = dplyr::case_when(
      !is.na(final_state) ~ final_state,
      main_cell_type %in% non_subclustered_main_types ~ main_cell_type,
      TRUE ~ NA_character_
    ),
    state_family = dplyr::case_when(
      !is.na(state_family) ~ state_family,
      main_cell_type %in% non_subclustered_main_types ~ "Other",
      TRUE ~ NA_character_
    )
  )

state_assignment_summary <- full_meta_state %>%
  dplyr::mutate(
    assignment = dplyr::case_when(
      !is.na(final_state) ~ "kept",
      main_cell_type %in% subclustered_main_types ~ "removed_from_subclustered_compartment",
      TRUE ~ "removed_other_or_unexpected"
    )
  ) %>%
  dplyr::count(main_cell_type, assignment, name = "n_cells") %>%
  dplyr::arrange(main_cell_type, assignment)

write.csv(state_assignment_summary, file.path(output_dir_data, "cell_state_assignment_summary.csv"), row.names = FALSE)

state_assignment_summary %>%
  reactable::reactable(filterable = TRUE, searchable = TRUE)
```

## Per-sample cell-state proportions

```{r cell state proportions}
state_meta_kept <- full_meta_state %>%
  dplyr::filter(!is.na(final_state), !is.na(Sample)) %>%
  dplyr::mutate(
    final_state = as.character(final_state),
    state_family = as.character(state_family)
  )

state_counts_long <- state_meta_kept %>%
  dplyr::count(Sample, state_family, final_state, name = "n_cells") %>%
  dplyr::group_by(Sample) %>%
  dplyr::mutate(
    total_cells = sum(n_cells),
    prop = n_cells / total_cells,
    logit_prop = logit_prop(prop)
  ) %>%
  dplyr::ungroup()

all_states <- sort(unique(state_counts_long$final_state))

state_prop_wide <- state_counts_long %>%
  dplyr::select(Sample, final_state, prop) %>%
  tidyr::complete(Sample, final_state = all_states, fill = list(prop = 0)) %>%
  tidyr::pivot_wider(names_from = final_state, values_from = prop, names_prefix = "prop_STATE_")

state_logit_wide <- state_counts_long %>%
  dplyr::select(Sample, final_state, prop) %>%
  tidyr::complete(Sample, final_state = all_states, fill = list(prop = 0)) %>%
  dplyr::mutate(logit_prop = logit_prop(prop)) %>%
  dplyr::select(Sample, final_state, logit_prop) %>%
  tidyr::pivot_wider(names_from = final_state, values_from = logit_prop, names_prefix = "logit_prop_STATE_")

all_state_props <- state_prop_wide %>%
  dplyr::left_join(state_logit_wide, by = "Sample")

state_info <- state_counts_long %>%
  dplyr::distinct(final_state, state_family) %>%
  dplyr::arrange(state_family, final_state)

state_order <- make_state_order(state_info)

write.csv(state_counts_long, file.path(output_dir_data, "cell_state_proportions_long.csv"), row.names = FALSE)
write.csv(all_state_props, file.path(output_dir_data, "cell_state_proportions_wide.csv"), row.names = FALSE)
write.csv(state_info, file.path(output_dir_data, "cell_state_info.csv"), row.names = FALSE)

state_counts_long %>%
  dplyr::group_by(state_family, final_state) %>%
  dplyr::summarise(
    total_cells = sum(n_cells),
    median_prop = median(prop),
    max_prop = max(prop),
    n_samples_present = sum(n_cells > 0),
    .groups = "drop"
  ) %>%
  dplyr::arrange(
    factor(state_family, levels = c("Cardiomyocytes", "Fibroblasts", "Endothelial cells", "Mural cells", "Macrophages", "Lymphocytes", "Other")),
    dplyr::desc(total_cells)
  ) %>%
  reactable::reactable(filterable = TRUE, searchable = TRUE)
```

## Matched proteomics samples and PCA

```{r matched proteomics pca}
sc_samples <- unique(all_state_props$Sample)

matched_olink_samples <- intersect(olink_meta$Sample, sc_samples)
matched_ms_samples <- intersect(ms_meta$Sample, sc_samples)

olink_meta_matched <- olink_meta %>%
  dplyr::filter(Sample %in% matched_olink_samples) %>%
  dplyr::slice(match(matched_olink_samples, Sample))

olink_expr_matched <- olink_expr[, matched_olink_samples, drop = FALSE]

ms_meta_matched <- ms_meta %>%
  dplyr::filter(Sample %in% matched_ms_samples) %>%
  dplyr::slice(match(matched_ms_samples, Sample))

ms_expr_matched <- ms_expr[, matched_ms_samples, drop = FALSE]

olink_pca <- run_pca_from_matrix(olink_expr_matched, olink_meta_matched, assay_label = "Olink")
ms_pca <- run_pca_from_matrix(ms_expr_matched, ms_meta_matched, assay_label = "MS")

olink_top_loadings <- get_top_pca_features(olink_pca$pca, pcs = pcs_to_plot, top_n = top_n_loadings)
ms_top_loadings <- get_top_pca_features(ms_pca$pca, pcs = pcs_to_plot, top_n = top_n_loadings)

write.csv(olink_top_loadings, file.path(output_dir_data, "Olink_top_loadings_matched.csv"), row.names = FALSE)
write.csv(ms_top_loadings, file.path(output_dir_data, "MS_top_loadings_matched.csv"), row.names = FALSE)

tibble(
  assay = c("Olink", "Mass spec"),
  n_samples = c(nrow(olink_pca$scores), nrow(ms_pca$scores)),
  n_features = c(nrow(olink_pca$expr_pca), nrow(ms_pca$expr_pca))
)
```

## Scree / elbow plots

```{r scree plots, fig.width=10, fig.height=4}
make_scree_plot(olink_pca, "Olink", n_pcs = n_pcs_scree) +
  make_scree_plot(ms_pca, "Mass spec", n_pcs = n_pcs_scree)
```

## PCA plots

```{r pca plots olink, fig.width=11, fig.height=8}
(make_pca_plot(olink_pca, "mPAP_mmHg", "Olink matched samples") |
 make_pca_plot(olink_pca, "TAPSE_mm", "Olink matched samples")) /
(make_pca_plot(olink_pca, "RVIT_mm", "Olink matched samples") |
 make_pca_plot(olink_pca, "Age_years", "Olink matched samples")) /
(make_pca_plot(olink_pca, "BCA_mg_ml", "Olink matched samples") |
 make_pca_plot(olink_pca, "Sample_age_days", "Olink matched samples"))
```

```{r pca plots ms, fig.width=11, fig.height=8}
(make_pca_plot(ms_pca, "mPAP_mmHg", "Mass spec matched samples") |
 make_pca_plot(ms_pca, "TAPSE_mm", "Mass spec matched samples")) /
(make_pca_plot(ms_pca, "RVIT_mm", "Mass spec matched samples") |
 make_pca_plot(ms_pca, "Age_years", "Mass spec matched samples")) /
(make_pca_plot(ms_pca, "Sample_age_days", "Mass spec matched samples") |
 make_pca_plot(ms_pca, "LVEF_percent", "Mass spec matched samples"))
```

## Heatmaps of top PC-loading proteins

```{r olink loading heatmaps, fig.width=6.5, fig.height=7}
olink_ann <- olink_meta_matched %>%
  dplyr::select(Sample, BCA_mg_ml) %>%
  tibble::column_to_rownames("Sample")

for (pc in paste0("PC", pcs_to_plot)) {
  make_loading_heatmap(
    pca_list = olink_pca,
    top_tbl = olink_top_loadings,
    pc = pc,
    title_text = paste0("Olink - top ", top_n_loadings, " ", pc, " proteins"),
    annotation_col = olink_ann
  )
}
```

```{r ms loading heatmaps, fig.width=6.5, fig.height=7}
for (pc in paste0("PC", pcs_to_plot)) {
  make_loading_heatmap(
    pca_list = ms_pca,
    top_tbl = ms_top_loadings,
    pc = pc,
    title_text = paste0("Mass spec - top ", top_n_loadings, " ", pc, " proteins"),
    annotation_col = NULL
  )
}
```

## Cell-state associations with proteomic PCs

```{r run pc state models}
all_state_olink_res <- fit_pc_state_models(
  scores_df = olink_pca$scores,
  prop_df = all_state_props,
  state_info = state_info,
  assay = "Olink",
  pcs = pcs_to_test,
  min_samples_present = min_samples_present
)

all_state_ms_res <- fit_pc_state_models(
  scores_df = ms_pca$scores,
  prop_df = all_state_props,
  state_info = state_info,
  assay = "MS",
  pcs = pcs_to_test,
  min_samples_present = min_samples_present
)

write.csv(all_state_olink_res, file.path(output_dir_data, "Olink_PC_vs_cell_states_results.csv"), row.names = FALSE)
write.csv(all_state_ms_res, file.path(output_dir_data, "MS_PC_vs_cell_states_results.csv"), row.names = FALSE)

all_state_olink_res %>%
  reactable::reactable(filterable = TRUE, searchable = TRUE)

all_state_ms_res %>%
  reactable::reactable(filterable = TRUE, searchable = TRUE)
```

## Dot plots of PC-cell-state associations

```{r pc state dotplots, fig.width=16, fig.height=5}
make_pc_state_dotplot(
  all_state_olink_res,
  state_order = state_order,
  title_text = "Olink PCs vs cell-state proportions"
)

make_pc_state_dotplot(
  all_state_ms_res,
  state_order = state_order,
  title_text = "Mass spec PCs vs cell-state proportions"
)
```

## Summary tables

```{r summary tables}
state_family_summary <- dplyr::bind_rows(all_state_olink_res, all_state_ms_res) %>%
  dplyr::group_by(assay, state_family) %>%
  dplyr::summarise(
    min_p_adj = min(p_adj, na.rm = TRUE),
    max_neg_log10_padj = max(-log10(p_adj), na.rm = TRUE),
    n_nominal = sum(p_value < 0.05, na.rm = TRUE),
    n_fdr_0_2 = sum(p_adj < 0.2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    min_p_adj = ifelse(is.infinite(min_p_adj), NA_real_, min_p_adj),
    max_neg_log10_padj = ifelse(is.infinite(max_neg_log10_padj), NA_real_, max_neg_log10_padj)
  )

top_state_associations <- dplyr::bind_rows(all_state_olink_res, all_state_ms_res) %>%
  dplyr::arrange(p_adj, p_value) %>%
  dplyr::select(assay, PC, final_state, state_family, n, n_present, estimate, p_value, p_adj, rho_prop, p_prop)

write.csv(state_family_summary, file.path(output_dir_data, "cell_state_family_summary.csv"), row.names = FALSE)
write.csv(top_state_associations, file.path(output_dir_data, "top_cell_state_associations.csv"), row.names = FALSE)

state_family_summary %>%
  reactable::reactable(filterable = TRUE, searchable = TRUE)

top_state_associations %>%
  reactable::reactable(filterable = TRUE, searchable = TRUE)
```

## Notes

```{r notes, results='asis'}
cat(
  "Testing summary:\n",
  "- PCA is performed separately for Olink and mass spectrometry using matched proteomics/single-nucleus samples.\n",
  "- Olink PCA heatmaps show only BCA as annotation. Mass-spec heatmaps have no top annotation.\n",
  "- Heatmap rows and columns are not clustered.\n",
  "- Cell states are assigned from the subclustered objects using cell IDs.\n",
  "- Non-subclustered cell types are kept as their original broad cell type.\n",
  "- Cells from subclustered compartments that are absent from the subclustered object are removed.\n",
  "- States must be present in at least ", min_samples_present, " matched samples to be tested.\n",
  "- Models use logit-transformed cell-state proportions.\n",
  "- Olink models: logit proportion ~ PC + Age_years + BCA_mg_ml.\n",
  "- Mass-spec models: logit proportion ~ PC + Age_years.\n",
  "- Dot-plot color shows the model beta. Dot size shows -log10(FDR). Black outlines indicate FDR < 0.2.\n",
  "- BH correction is applied across all tested PC-state associations within each platform.\n",
  sep = ""
)
```

## Focused proteomic PC and cell-state plots

This section focuses on the associations selected from the global screen:

- Mass spec PC1 vs FB5
- Mass spec PC2 vs CM5
- Olink PC2 vs CM3
- Olink PC2 vs CM4

It makes:
1. scatterplots of PC score vs raw state proportion, with p values from the adjusted logit-proportion model;
2. heatmaps of the top PC-loading proteins with the selected cell-state proportion shown as a simple top annotation;
3. scatterplots of the selected state proportion against the strongest PC-loading proteins;
4. optional GO over-representation analysis for the top-loading proteins.

```{r focused pc state setup}
library(clusterProfiler)
library(org.Hs.eg.db)

focused_pairs <- tibble::tribble(
  ~assay,   ~PC,    ~state,
  "MS",     "PC1",  "FB5",
  "MS",     "PC2",  "CM5",
  "Olink",  "PC2",  "CM3",
  "Olink",  "PC2",  "CM4"
)

focused_top_n_loadings <- 20
focused_top_n_gene_scatter <- 8

get_pair_objects <- function(assay) {
  if (assay == "MS") {
    list(
      pca = ms_pca,
      res = all_state_ms_res,
      top_loadings = ms_top_loadings,
      expression = ms_pca$expr_pca,
      scores = ms_pca$scores
    )
  } else if (assay == "Olink") {
    list(
      pca = olink_pca,
      res = all_state_olink_res,
      top_loadings = olink_top_loadings,
      expression = olink_pca$expr_pca,
      scores = olink_pca$scores
    )
  } else {
    stop("Unknown assay: ", assay)
  }
}

get_state_prop_col <- function(state) {
  paste0("prop_STATE_", state)
}

get_state_logit_col <- function(state) {
  paste0("logit_prop_STATE_", state)
}
```

## Scatterplots of selected PC-cell-state associations

```{r focused pc state scatter helper}
plot_focused_pc_state <- function(assay, pc, state) {
  obj <- get_pair_objects(assay)

  prop_col <- get_state_prop_col(state)
  logit_col <- get_state_logit_col(state)

  covars <- c("Age_years")
  if (assay == "Olink") covars <- c(covars, "BCA_mg_ml")
  covars <- covars[covars %in% colnames(obj$scores)]

  dat <- obj$scores %>%
    dplyr::left_join(all_state_props, by = "Sample") %>%
    dplyr::select(Sample, dplyr::all_of(c(pc, prop_col, logit_col, covars))) %>%
    tidyr::drop_na()

  if (nrow(dat) < 5) {
    return(ggplot() + annotate("text", x = 1, y = 1, label = "Too few samples") + theme_void())
  }

  form <- reformulate(termlabels = c(pc, covars), response = logit_col)
  fit <- lm(form, data = dat)
  coef_tab <- summary(fit)$coefficients

  beta <- coef_tab[pc, "Estimate"]
  pval <- coef_tab[pc, "Pr(>|t|)"]

  fdr <- obj$res %>%
    dplyr::filter(PC == pc, final_state == state) %>%
    dplyr::pull(p_adj)

  if (length(fdr) == 0) fdr <- NA_real_

  ct <- suppressWarnings(cor.test(dat[[pc]], dat[[prop_col]], method = "spearman"))

  ggplot(dat, aes(x = .data[[prop_col]], y = .data[[pc]], label = Sample)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 0.5, color = "black") +
    ggrepel::geom_text_repel(size = 2.5, max.overlaps = 30) +
    annotate(
      "text",
      x = Inf, y = Inf,
      label = paste0(
        "beta(logit) = ", sprintf("%.3f", beta),
        "\np = ", signif(pval, 2),
        "\nFDR = ", signif(fdr, 2),
        "\nrho(prop) = ", round(unname(ct$estimate), 2),
        "\nn = ", nrow(dat)
      ),
      hjust = 1.1,
      vjust = 1.2,
      size = 4
    ) +
    labs(
      title = paste0(assay, " ", pc, " vs ", state, " proportion"),
      x = paste0(state, " proportion"),
      y = paste0(assay, " ", pc, " score")
    ) +
    theme_classic(base_size = 12)
}
```

```{r focused pc state scatterplots, fig.width=11, fig.height=8}
focused_scatter_plots <- purrr::pmap(
  focused_pairs,
  function(assay, PC, state) {
    plot_focused_pc_state(assay, PC, state)
  }
)

wrap_plots(focused_scatter_plots, ncol = 2)
```

## Heatmaps of top PC-loading proteins with cell-state proportion annotation

```{r focused loading heatmap helper}
make_focused_loading_heatmap <- function(assay, pc, state, top_n = 20) {
  obj <- get_pair_objects(assay)

  top_tbl <- obj$top_loadings %>%
    dplyr::filter(PC == pc) %>%
    dplyr::arrange(dplyr::desc(abs_loading)) %>%
    dplyr::slice_head(n = top_n)

  feat <- top_tbl$feature
  hm <- obj$expression[feat, , drop = FALSE]

  hm_z <- t(scale(t(hm)))
  hm_z[is.nan(hm_z)] <- 0
  hm_z[!is.finite(hm_z)] <- 0

  rownames(hm_z) <- top_tbl %>%
    dplyr::mutate(label = paste0(feature, " (", sprintf("%.3f", loading), ")")) %>%
    dplyr::pull(label)

  prop_col <- get_state_prop_col(state)

  ann <- all_state_props %>%
    dplyr::filter(Sample %in% colnames(hm_z)) %>%
    dplyr::slice(match(colnames(hm_z), Sample)) %>%
    dplyr::select(Sample, dplyr::all_of(prop_col)) %>%
    dplyr::rename(!!paste0("prop_", state) := dplyr::all_of(prop_col))

  # Order columns by selected cell-state proportion
  sample_order <- ann %>%
    dplyr::arrange(.data[[paste0("prop_", state)]]) %>%
    dplyr::pull(Sample)

  hm_z <- hm_z[, sample_order, drop = FALSE]

  ann <- ann %>%
    dplyr::slice(match(sample_order, Sample)) %>%
    tibble::column_to_rownames("Sample")

  # Cluster rows, but hide the row dendrogram
  row_hclust <- hclust(dist(hm_z), method = "complete")

  pheatmap(
    hm_z,
    annotation_col = ann,
    cluster_rows = row_hclust,
    treeheight_row = 0,
    cluster_cols = FALSE,
    show_rownames = TRUE,
    show_colnames = TRUE,
    fontsize_row = 8,
    fontsize_col = 8,
    border_color = NA,
    color = heat_cols,
    main = paste0(
      assay, " ", pc, " top ", top_n,
      " proteins - ", state, " proportion"
    )
  )
}
```

```{r focused loading heatmaps, fig.width=6, fig.height=4}
purrr::pwalk(
  focused_pairs,
  function(assay, PC, state) {
    make_focused_loading_heatmap(
      assay = assay,
      pc = PC,
      state = state,
      top_n = focused_top_n_loadings
    )
  }
)
```

## Protein-level scatterplots for the top loading proteins

These plots ask whether the individual proteins driving a PC also track the selected cell-state proportion. They are descriptive and should not be overinterpreted one-by-one.

```{r focused protein scatter helper}
plot_top_loading_protein_vs_state <- function(assay, pc, state, top_n = 8) {
  obj <- get_pair_objects(assay)

  top_tbl <- obj$top_loadings %>%
    dplyr::filter(PC == pc) %>%
    dplyr::arrange(dplyr::desc(abs_loading)) %>%
    dplyr::slice_head(n = top_n)

  prop_col <- get_state_prop_col(state)

  plot_df <- obj$expression[top_tbl$feature, , drop = FALSE] %>%
    as.data.frame() %>%
    tibble::rownames_to_column("feature") %>%
    tidyr::pivot_longer(-feature, names_to = "Sample", values_to = "abundance") %>%
    dplyr::left_join(
      all_state_props %>%
        dplyr::select(Sample, dplyr::all_of(prop_col)),
      by = "Sample"
    ) %>%
    dplyr::left_join(
      top_tbl %>% dplyr::select(feature, loading),
      by = "feature"
    ) %>%
    dplyr::mutate(
      feature_label = paste0(feature, " (loading ", sprintf("%.3f", loading), ")")
    ) %>%
    tidyr::drop_na()

  cor_tbl <- plot_df %>%
    dplyr::group_by(feature_label) %>%
    dplyr::summarise(
      rho = suppressWarnings(cor(.data[[prop_col]], abundance, method = "spearman")),
      p = suppressWarnings(cor.test(.data[[prop_col]], abundance, method = "spearman")$p.value),
      .groups = "drop"
    ) %>%
    dplyr::mutate(label = paste0("rho=", round(rho, 2), "; p=", signif(p, 2)))

  plot_df <- plot_df %>%
    dplyr::left_join(cor_tbl, by = "feature_label")

  ggplot(plot_df, aes(x = .data[[prop_col]], y = abundance)) +
    geom_point(size = 1.8) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 0.4, color = "black") +
    facet_wrap(~ feature_label, scales = "free_y", ncol = 4) +
    labs(
      title = paste0(assay, " ", pc, ": top loading proteins vs ", state, " proportion"),
      x = paste0(state, " proportion"),
      y = "Protein abundance"
    ) +
    theme_classic(base_size = 10) +
    theme(
      strip.text = element_text(size = 8)
    )
}
```

```{r focused protein scatterplots, fig.width=13, fig.height=8}
plot_top_loading_protein_vs_state("MS", "PC1", "FB5", top_n = focused_top_n_gene_scatter)
plot_top_loading_protein_vs_state("MS", "PC2", "CM5", top_n = focused_top_n_gene_scatter)
plot_top_loading_protein_vs_state("Olink", "PC2", "CM3", top_n = focused_top_n_gene_scatter)
plot_top_loading_protein_vs_state("Olink", "PC2", "CM4", top_n = focused_top_n_gene_scatter)
```

## Optional: GO over-representation analysis of top PC-loading proteins

For mass spec this can be informative. For Olink, interpret carefully because the panel is targeted and not a genome-wide proteome background. For Olink, the better background is all Olink proteins that passed filtering.

```{r focused ORA helper}
run_pc_loading_ora <- function(assay, pc, top_n = 50, ont = "BP") {
  obj <- get_pair_objects(assay)

  genes <- obj$top_loadings %>%
    dplyr::filter(PC == pc) %>%
    dplyr::arrange(dplyr::desc(abs_loading)) %>%
    dplyr::slice_head(n = top_n) %>%
    dplyr::pull(feature) %>%
    unique()

  universe <- rownames(obj$expression)

  genes_entrez <- bitr(
    genes,
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db
  ) %>%
    dplyr::pull(ENTREZID) %>%
    unique()

  universe_entrez <- bitr(
    universe,
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db
  ) %>%
    dplyr::pull(ENTREZID) %>%
    unique()

  ego <- enrichGO(
    gene = genes_entrez,
    universe = universe_entrez,
    OrgDb = org.Hs.eg.db,
    ont = ont,
    pAdjustMethod = "BH",
    readable = TRUE
  )

  ego
}
```

```{r focused ORA, fig.width=8, fig.height=5}
ora_ms_pc1 <- run_pc_loading_ora("MS", "PC1", top_n = 50, ont = "BP")
ora_ms_pc2 <- run_pc_loading_ora("MS", "PC2", top_n = 50, ont = "BP")
ora_olink_pc2 <- run_pc_loading_ora("Olink", "PC2", top_n = 50, ont = "BP")

write.csv(as.data.frame(ora_ms_pc1), file.path(output_dir_data, "ORA_MS_PC1_top50_loadings_GO_BP.csv"), row.names = FALSE)
write.csv(as.data.frame(ora_ms_pc2), file.path(output_dir_data, "ORA_MS_PC2_top50_loadings_GO_BP.csv"), row.names = FALSE)
write.csv(as.data.frame(ora_olink_pc2), file.path(output_dir_data, "ORA_Olink_PC2_top50_loadings_GO_BP.csv"), row.names = FALSE)

dotplot(ora_ms_pc1, showCategory = 10) + ggtitle("MS PC1 top loading proteins")
dotplot(ora_ms_pc2, showCategory = 10) + ggtitle("MS PC2 top loading proteins")
dotplot(ora_olink_pc2, showCategory = 10) + ggtitle("Olink PC2 top loading proteins")
```

## Overlap between selected PC-loading proteins and subcluster markers

```{r selected PC-marker overlap tables}
top_n_marker_overlap <- 20

selected_marker_overlap <- tibble::tribble(
  ~assay,   ~PC,    ~cell_state, ~marker_file,
  "MS",     "PC1",  "FB5",       here::here("output", "Fibroblasts_subclustering", "FB_Markers_all_cell_state.csv"),
  "MS",     "PC2",  "CM5",       here::here("output", "Cardiomyocytes_subclustering", "CM_Markers_all_cell_state.csv"),
  "Olink",  "PC2",  "CM3",       here::here("output", "Cardiomyocytes_subclustering", "CM_Markers_all_cell_state.csv"),
  "Olink",  "PC2",  "CM4",       here::here("output", "Cardiomyocytes_subclustering", "CM_Markers_all_cell_state.csv")
)

# Fallback CM marker file names, in case the file name is slightly different.
if (!file.exists(here::here("output", "Cardiomyocytes_subclustering", "CM_Markers_all_cell_state.csv"))) {
  cm_marker_candidates <- c(
    here::here("output", "Cardiomyocytes_subclustering", "CM_Markers_all_cell_state.csv"),
    here::here("output", "Cardiomyocytes_subclustering", "Cardiomyocytes_Markers_all_cell_state.csv"),
    here::here("output", "Cardiomyocytes_subclustering", "CM_markers_all_cell_state.csv"),
    here::here("output", "Cardiomyocytes_subclustering", "markers_all_cell_state.csv")
  )
  cm_marker_file <- cm_marker_candidates[file.exists(cm_marker_candidates)][1]

  selected_marker_overlap <- selected_marker_overlap %>%
    dplyr::mutate(
      marker_file = dplyr::if_else(
        cell_state %in% c("CM3", "CM4", "CM5"),
        cm_marker_file,
        marker_file
      )
    )
}

read_marker_table_simple <- function(path) {
  if (length(path) == 0 || is.na(path) || !file.exists(path)) {
    stop("Marker file not found: ", path)
  }

  x <- read.csv(path, check.names = FALSE)

  # Fix empty column names from exported rownames
  empty_cols <- which(is.na(colnames(x)) | colnames(x) == "")
  if (length(empty_cols) > 0) {
    colnames(x)[empty_cols] <- paste0("unnamed_", empty_cols)
  }

  # If the first unnamed column is just rownames/index, drop it
  first_col <- colnames(x)[1]
  if (grepl("^unnamed_", first_col) || first_col %in% c("X", "...1")) {
    x <- x[, -1, drop = FALSE]
  }

  gene_col <- c("gene", "Gene", "genes", "symbol", "SYMBOL")
  gene_col <- gene_col[gene_col %in% colnames(x)][1]

  state_col <- c("cluster", "cell_state", "CellState", "state", "ident", "group")
  state_col <- state_col[state_col %in% colnames(x)][1]

  if (length(gene_col) == 0 || is.na(gene_col)) {
    stop(
      "Could not detect gene column in marker file: ", path,
      "\nAvailable columns are: ", paste(colnames(x), collapse = ", ")
    )
  }

  if (length(state_col) == 0 || is.na(state_col)) {
    stop(
      "Could not detect state/cluster column in marker file: ", path,
      "\nAvailable columns are: ", paste(colnames(x), collapse = ", ")
    )
  }

  x %>%
    dplyr::rename(
      marker_gene = dplyr::all_of(gene_col),
      marker_state = dplyr::all_of(state_col)
    ) %>%
    dplyr::mutate(
      marker_gene = as.character(marker_gene),
      marker_state = as.character(marker_state)
    )
}

get_top_pc_loading_simple <- function(assay, pc, top_n = 20) {
  loading_tbl <- if (assay == "MS") {
    ms_top_loadings
  } else {
    olink_top_loadings
  }

  loading_tbl %>%
    dplyr::filter(PC == pc) %>%
    dplyr::arrange(dplyr::desc(abs_loading)) %>%
    dplyr::slice_head(n = top_n) %>%
    dplyr::mutate(
      loading_rank = dplyr::row_number(),
      loading_gene = as.character(feature)
    )
}

make_pc_marker_overlap_simple <- function(assay, pc, cell_state, marker_file, top_n = 20) {
  top_tbl <- get_top_pc_loading_simple(assay, pc, top_n = top_n)

  marker_tbl <- read_marker_table_simple(marker_file) %>%
    dplyr::filter(marker_state == cell_state) %>%
    dplyr::distinct(marker_gene, .keep_all = TRUE)

  marker_cols <- intersect(
    c("marker_gene", "marker_state", "p_val", "avg_log2FC", "avg_logFC", "pct.1", "pct.2", "p_val_adj"),
    colnames(marker_tbl)
  )

  full_tbl <- top_tbl %>%
    dplyr::left_join(
      marker_tbl %>% dplyr::select(dplyr::all_of(marker_cols)),
      by = c("loading_gene" = "marker_gene")
    ) %>%
    dplyr::mutate(
      assay = assay,
      PC = pc,
      tested_state = cell_state,
      is_marker_of_state = !is.na(marker_state)
    ) %>%
    dplyr::select(
      assay, PC, tested_state,
      loading_rank, loading_gene, loading, abs_loading,
      is_marker_of_state,
      dplyr::everything()
    )

  summary_tbl <- tibble::tibble(
    assay = assay,
    PC = pc,
    tested_state = cell_state,
    top_n = top_n,
    n_top_pc_proteins = nrow(top_tbl),
    n_state_markers = nrow(marker_tbl),
    n_overlap = sum(full_tbl$is_marker_of_state),
    overlap_genes = paste(full_tbl$loading_gene[full_tbl$is_marker_of_state], collapse = "; ")
  )

  list(summary = summary_tbl, full = full_tbl, overlap = full_tbl %>% dplyr::filter(is_marker_of_state))
}

pc_marker_overlap_list <- purrr::pmap(
  selected_marker_overlap,
  function(assay, PC, cell_state, marker_file) {
    make_pc_marker_overlap_simple(
      assay = assay,
      pc = PC,
      cell_state = cell_state,
      marker_file = marker_file,
      top_n = top_n_marker_overlap
    )
  }
)

pc_marker_overlap_summary <- purrr::map_dfr(pc_marker_overlap_list, "summary")
pc_marker_overlap_full <- purrr::map_dfr(pc_marker_overlap_list, "full")
pc_marker_overlap_only <- purrr::map_dfr(pc_marker_overlap_list, "overlap")

write.csv(
  pc_marker_overlap_summary,
  file.path(output_dir_data, paste0("top", top_n_marker_overlap, "_PC_marker_overlap_summary.csv")),
  row.names = FALSE
)

write.csv(
  pc_marker_overlap_full,
  file.path(output_dir_data, paste0("top", top_n_marker_overlap, "_PC_marker_overlap_full.csv")),
  row.names = FALSE
)

write.csv(
  pc_marker_overlap_only,
  file.path(output_dir_data, paste0("top", top_n_marker_overlap, "_PC_marker_overlap_only.csv")),
  row.names = FALSE
)

pc_marker_overlap_summary %>%
  reactable::reactable(filterable = TRUE, searchable = TRUE)

pc_marker_overlap_only %>%
  reactable::reactable(filterable = TRUE, searchable = TRUE)
```

```{r selected PC-marker full tables}
pc_marker_overlap_full %>%
  dplyr::arrange(assay, PC, tested_state, loading_rank) %>%
  reactable::reactable(filterable = TRUE, searchable = TRUE)
```

```{r}
FB_combined <- qs2::qs_read(here::here("output", "Fibroblasts_subclustering", "FB.qs2"), nthreads = 32)
CM_combined <- qs2::qs_read(here::here("output", "Cardiomyocytes_subclustering", "CM.qs2"), nthreads = 32)
fb5_ms_genes <- c("FLNA", "ANXA1", "ITGB1", "TAGLN")
cm5_ms_genes <- c("TNNT2", "TNNI3", "MYL2", "MYL3")

cell_type_cols <- c(
  "#d05050", "#33A02C", "#1F78B4", "#CAB2D6",
  "#FF7F00", "#B2DF8A", "#B15928", "#FFD000",
  "#FA95C0", "#FDBF6F", "#6A3D9A", "#7aCDF9"
)

Idents(FB_combined) <- "cell_state"
Idents(CM_combined) <- "cell_state"
```

```{r}
style_vln_list <- function(p_list) {
  lapply(p_list, function(p) {
    for (j in seq_along(p$layers)) {
      geom_name <- class(p$layers[[j]]$geom)[1]

      if (geom_name %in% c("GeomPoint", "GeomJitter")) {
        p$layers[[j]]$aes_params$alpha <- 0.1
        p$layers[[j]]$aes_params$colour <- scales::alpha("black", 0.01)
      }
    }

    p +
      theme_classic() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", size = 12),
        axis.text.y = element_text(face = "bold", size = 10),
        axis.title.y = element_text(face = "bold"),
        axis.title.x = element_blank(),
        strip.text = element_text(face = "bold"),
        legend.position = "none"
      )
  })
}
```

```{r, fig.width=12, fig.height=2.8}
DefaultAssay(FB_combined) <- "RNA"
Idents(FB_combined) <- "cell_state"

fb_vln_list <- VlnPlot(
  FB_combined,
  features = fb5_ms_genes,
  group.by = "cell_state",
  assay = "RNA",
  cols = cell_type_cols,
  pt.size = 0.001,
  add.noise = FALSE,
  ncol = length(fb5_ms_genes),
  combine = FALSE
)

fb_vln_list <- style_vln_list(fb_vln_list)

wrap_plots(fb_vln_list, ncol = length(fb_vln_list))
```

```{r, fig.width=12, fig.height=2.8}
DefaultAssay(CM_combined) <- "RNA"
Idents(CM_combined) <- "cell_state"

cm_vln_list <- VlnPlot(
  CM_combined,
  features = cm5_ms_genes,
  group.by = "cell_state",
  assay = "RNA",
  cols = cell_type_cols,
  pt.size = 0.001,
  add.noise = FALSE,
  ncol = length(cm5_ms_genes),
  combine = FALSE
)

cm_vln_list <- style_vln_list(cm_vln_list)

wrap_plots(cm_vln_list, ncol = length(cm_vln_list))
```

```{r, fig.width=7, fig.height=4}
DefaultAssay(FB_combined) <- "SCT"
Idents(FB_combined) <- "cell_state"

DotPlot(
  FB_combined,
  dot.scale = 13,
  features = fb5_ms_genes,
  assay = "SCT",
  cols = c("grey90", "#A63A2A")
) +
  coord_flip() +
  theme_classic() +
  labs(title = "FB5-associated MS genes") +
  theme(
    axis.text.x = element_text(face = "bold", size = 19),
    axis.text.y = element_text(face = "bold", size = 19),
    axis.title = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

```{r, fig.width=7, fig.height=4}
DefaultAssay(CM_combined) <- "SCT"
Idents(CM_combined) <- "cell_state"

DotPlot(
  CM_combined,
  dot.scale = 13,
  features = cm5_ms_genes,
  assay = "SCT",
  cols = c("grey90", "#A63A2A")
) +
  coord_flip() +
  theme_classic() +
  labs(title = "CM5-associated MS genes") +
  theme(
    axis.text.x = element_text(face = "bold", size = 19),
    axis.text.y = element_text(face = "bold", size = 19),
    axis.title = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

```{r, fig.width=16.8, fig.height=4}
DefaultAssay(FB_combined) <- "SCT"

FeaturePlot(
  FB_combined,
  features = fb5_ms_genes,
  reduction = "umap",
  order = TRUE,
  cols = c("grey90", "#A63A2A"),
  pt.size = 0.5,
  ncol = 4,
  raster = FALSE
) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

```{r, fig.width=16.8, fig.height=4}
DefaultAssay(FB_combined) <- "SCT"

FeaturePlot(
  FB_combined,
  features = fb5_ms_genes,
  reduction = "umap",
  order = TRUE,
  cols = c("grey90", "#A63A2A"),
  pt.size = 0.5,
  ncol = 4,
  raster = FALSE
) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```


