---
title: "Proteomics quality control: Olink and LC-MS/MS"
author: "Gino Bonazza"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  dpi = 300,
  fig.align = "center"
)
```

# Overview

This workflow evaluates the quality of the Olink and LC-MS/MS proteomic
datasets before and after sample exclusion. Healthy control samples are
excluded from the initial analysis, which includes all DCM and ICM samples.

P11, PK61 and PK63 are subsequently excluded because of poor proteomic data
quality identified in the diagnostic plots. PK69 is excluded because duplicate
sample identifiers prevent unambiguous patient assignment. ICM samples are
excluded because the downstream analysis focuses on DCM.

Heatmaps include all usable proteins after removal of rows with insufficient
observations or zero variance. Missing LC-MS/MS values are imputed only for
visualisation. PCA is performed on proteins observed in at least 50% of
samples and uses row-median imputation.

# Setup

```{r libraries}
library(here)
library(readxl)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(limma)
library(matrixStats)
library(ComplexHeatmap)
library(circlize)
library(reactable)
library(scales)
```

```{r analysis choices}
current_file <- "Proteomics_quality_control"

output_dir <- here::here(
  "output",
  current_file
)

figure_dir <- here::here(
  "docs",
  "figure",
  paste0(current_file, ".Rmd")
)

dir.create(
  output_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  figure_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

clinical_file <- here::here(
  "data",
  "DCM_Clinical_data_83.xlsx"
)

olink_sample_file <- here::here(
  "data",
  "OLINK",
  "Samples_RV_Olink.xlsx"
)

olink_quant_file <- here::here(
  "data",
  "OLINK",
  "quant.xls"
)

ms_file <- here::here(
  "data",
  "Mass_spec",
  "20260319_MS_DIA_report_2uniquepeptides.csv"
)

ms_zero_is_missing <- TRUE
pca_min_observed_fraction <- 0.50
label_pca_samples <- TRUE

heatmap_column_fontsize <- 7
distribution_sample_fontsize <- 7
```

# Helper functions

```{r helper functions}
normalise_sample_id <- function(x) {
  trimws(
    as.character(x)
  )
}


make_group_palette <- function(values) {
  group_colours <- c(
    DCM = "#A63A2A",
    ICM = "#004C8C"
  )

  levels_present <- sort(
    unique(
      as.character(
        values[
          !is.na(values)
        ]
      )
    )
  )

  unexpected_levels <- setdiff(
    levels_present,
    names(group_colours)
  )

  if (length(unexpected_levels) > 0L) {
    stop(
      "Unexpected Group label(s): ",
      paste(
        unexpected_levels,
        collapse = ", "
      )
    )
  }

  group_colours[
    levels_present
  ]
}


safe_continuous_colour <- function(
    values,
    colours = c(
      "grey90",
      "grey60",
      "grey20"
    )
) {
  finite_values <- values[
    is.finite(values)
  ]

  if (length(finite_values) == 0L) {
    return(NULL)
  }

  breaks <- c(
    min(finite_values),
    stats::median(finite_values),
    max(finite_values)
  )

  if (length(unique(breaks)) < 3L) {
    spread <- max(
      abs(breaks[1L]) * 0.01,
      0.01
    )

    breaks <- c(
      breaks[1L] - spread,
      breaks[1L],
      breaks[1L] + spread
    )
  }

  circlize::colorRamp2(
    breaks,
    colours
  )
}


impute_rows_median <- function(mat) {
  output <- as.matrix(mat)

  for (i in seq_len(nrow(output))) {
    missing <- !is.finite(
      output[i, ]
    )

    if (!any(missing)) {
      next
    }

    observed <- output[
      i,
      !missing
    ]

    replacement <- if (
      length(observed) > 0L
    ) {
      stats::median(
        observed,
        na.rm = TRUE
      )
    } else {
      0
    }

    output[
      i,
      missing
    ] <- replacement
  }

  output[
    !is.finite(output)
  ] <- 0

  output
}


impute_rows_low_value <- function(mat) {
  output <- as.matrix(mat)

  global_floor <- suppressWarnings(
    stats::quantile(
      output,
      probs = 0.01,
      na.rm = TRUE
    )
  )

  if (!is.finite(global_floor)) {
    global_floor <- 0
  }

  for (i in seq_len(nrow(output))) {
    missing <- !is.finite(
      output[i, ]
    )

    if (!any(missing)) {
      next
    }

    observed <- output[
      i,
      !missing
    ]

    replacement <- if (
      length(observed) > 0L
    ) {
      minimum <- min(
        observed,
        na.rm = TRUE
      )

      spread <- stats::sd(
        observed,
        na.rm = TRUE
      )

      if (
        !is.finite(spread) ||
        spread == 0
      ) {
        spread <- 0.5
      }

      minimum - 0.5 * spread
    } else {
      global_floor
    }

    output[
      i,
      missing
    ] <- replacement
  }

  output[
    !is.finite(output)
  ] <- global_floor

  output
}


filter_usable_features <- function(
    mat,
    minimum_observed = 2L
) {
  mat <- as.matrix(mat)

  observed <- matrixStats::rowSums2(
    is.finite(mat)
  )

  variance <- matrixStats::rowVars(
    mat,
    na.rm = TRUE
  )

  keep <- observed >= minimum_observed &
    is.finite(variance) &
    variance > 0

  mat[
    keep,
    ,
    drop = FALSE
  ]
}


prepare_pca <- function(
    mat,
    metadata,
    assay_name,
    minimum_observed_fraction =
      pca_min_observed_fraction
) {
  mat <- as.matrix(mat)

  observed_fraction <- matrixStats::rowMeans2(
    is.finite(mat)
  )

  variance <- matrixStats::rowVars(
    mat,
    na.rm = TRUE
  )

  keep <- observed_fraction >=
    minimum_observed_fraction &
    is.finite(variance) &
    variance > 0

  pca_matrix <- mat[
    keep,
    ,
    drop = FALSE
  ]

  if (nrow(pca_matrix) < 2L) {
    stop(
      assay_name,
      ": fewer than two proteins are available for PCA."
    )
  }

  pca_matrix <- impute_rows_median(
    pca_matrix
  )

  pca <- stats::prcomp(
    t(pca_matrix),
    center = TRUE,
    scale. = TRUE
  )

  variance_explained <-
    pca$sdev^2 /
    sum(pca$sdev^2) *
    100

  plot_data <- data.frame(
    Sample = rownames(pca$x),
    PC1 = pca$x[, 1L],
    PC2 = pca$x[, 2L],
    stringsAsFactors = FALSE
  ) %>%
    dplyr::left_join(
      metadata,
      by = "Sample"
    )

  list(
    pca = pca,
    variance_explained =
      variance_explained,
    plot_data = plot_data,
    number_features =
      nrow(pca_matrix),
    assay_name = assay_name
  )
}


plot_pca_group <- function(
    pca_object,
    title_suffix
) {
  plot_data <- pca_object$plot_data

  group_colours <- make_group_palette(
    plot_data$Group
  )

  output <- ggplot(
    plot_data,
    aes(
      x = PC1,
      y = PC2,
      colour = Group
    )
  ) +
    geom_hline(
      yintercept = 0,
      colour = "grey85",
      linewidth = 0.3
    ) +
    geom_vline(
      xintercept = 0,
      colour = "grey85",
      linewidth = 0.3
    ) +
    geom_point(
      size = 3
    ) +
    scale_colour_manual(
      values = group_colours,
      na.value = "grey60"
    ) +
    labs(
      title = paste(
        pca_object$assay_name,
        title_suffix,
        sep = " — "
      ),
      subtitle = paste(
        pca_object$number_features,
        "proteins used"
      ),
      x = paste0(
        "PC1 (",
        round(
          pca_object$
            variance_explained[1L],
          1
        ),
        "%)"
      ),
      y = paste0(
        "PC2 (",
        round(
          pca_object$
            variance_explained[2L],
          1
        ),
        "%)"
      ),
      colour = "Group"
    ) +
    theme_classic()

  if (isTRUE(label_pca_samples)) {
    output <- output +
      ggrepel::geom_text_repel(
        aes(label = Sample),
        size = 2.7,
        max.overlaps = Inf,
        box.padding = 0.25,
        point.padding = 0.15,
        segment.size = 0.25,
        show.legend = FALSE
      )
  }

  output
}


make_sample_annotation <- function(
    metadata,
    samples,
    include_olink_technical = FALSE,
    include_sample_age = FALSE
) {
  ordered_metadata <- metadata %>%
    dplyr::slice(
      match(
        samples,
        Sample
      )
    )

  annotation_data <- data.frame(
    Group = factor(
      ordered_metadata$Group
    ),
    row.names = samples,
    check.names = FALSE
  )

  annotation_colours <- list(
    Group = make_group_palette(
      annotation_data$Group
    )
  )

  if (
    isTRUE(include_olink_technical) &&
    "Batch" %in% colnames(ordered_metadata)
  ) {
    annotation_data$Batch <- factor(
      ordered_metadata$Batch
    )

    batch_levels <- levels(
      droplevels(
        annotation_data$Batch
      )
    )

    if (length(batch_levels) > 0L) {
      batch_colours <- setNames(
        scales::hue_pal(
          h = c(30, 300),
          c = 80,
          l = 55
        )(
          length(batch_levels)
        ),
        batch_levels
      )

      annotation_colours$Batch <-
        batch_colours
    }
  }

  if (
    isTRUE(include_olink_technical) &&
    "BCA_mg_ml" %in% colnames(
      ordered_metadata
    )
  ) {
    bca <- as.numeric(
      ordered_metadata$BCA_mg_ml
    )

    if (any(is.finite(bca))) {
      annotation_data$BCA_mg_ml <- bca

      annotation_colours$BCA_mg_ml <-
        safe_continuous_colour(bca)
    }
  }

  if (
    isTRUE(include_sample_age) &&
    "Sample_age_days" %in%
      colnames(ordered_metadata)
  ) {
    sample_age <- as.numeric(
      ordered_metadata$Sample_age_days
    )

    if (any(is.finite(sample_age))) {
      annotation_data$Sample_age_days <-
        sample_age

      annotation_colours$
        Sample_age_days <-
        safe_continuous_colour(
          sample_age
        )
    }
  }

  ComplexHeatmap::HeatmapAnnotation(
    df = annotation_data,
    col = annotation_colours,
    na_col = "#E5E7EB"
  )
}


draw_expression_heatmap <- function(
    mat,
    metadata,
    assay_name,
    title_suffix,
    imputation = c(
      "median",
      "low"
    ),
    include_olink_technical = FALSE,
    include_sample_age = FALSE
) {
  imputation <- match.arg(
    imputation
  )

  mat <- filter_usable_features(
    mat
  )

  if (imputation == "median") {
    visualisation_matrix <-
      impute_rows_median(mat)
  } else {
    visualisation_matrix <-
      impute_rows_low_value(mat)
  }

  z_matrix <- t(
    scale(
      t(
        visualisation_matrix
      )
    )
  )

  z_matrix[
    !is.finite(z_matrix)
  ] <- 0

  annotation <- make_sample_annotation(
    metadata = metadata,
    samples = colnames(z_matrix),
    include_olink_technical =
      include_olink_technical,
    include_sample_age =
      include_sample_age
  )

  heatmap <- ComplexHeatmap::Heatmap(
    z_matrix,
    name = "z-score",
    top_annotation = annotation,
    show_row_names = FALSE,
    show_column_names = TRUE,
    column_names_gp = grid::gpar(
      fontsize = heatmap_column_fontsize
    ),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    use_raster = nrow(z_matrix) > 2000,
    raster_quality = 2,
    col = circlize::colorRamp2(
      c(-2, 0, 2),
      c(
        "#2166AC",
        "white",
        "#B2182B"
      )
    ),
    column_title = paste(
      assay_name,
      title_suffix,
      paste0(
        "(",
        nrow(z_matrix),
        " proteins; ",
        ncol(z_matrix),
        " samples)"
      ),
      sep = " — "
    )
  )

  ComplexHeatmap::draw(
    heatmap,
    heatmap_legend_side = "right",
    annotation_legend_side = "right"
  )

  invisible(heatmap)
}


matrix_to_long <- function(
    mat,
    metadata,
    value_name
) {
  as.data.frame(mat) %>%
    tibble::rownames_to_column(
      "Protein"
    ) %>%
    tidyr::pivot_longer(
      cols = -Protein,
      names_to = "Sample",
      values_to = value_name
    ) %>%
    dplyr::left_join(
      metadata %>%
        dplyr::select(
          Sample,
          Group
        ),
      by = "Sample"
    )
}


plot_sample_boxplots <- function(
    mat,
    metadata,
    assay_name,
    title_suffix,
    y_label
) {
  long_data <- matrix_to_long(
    mat,
    metadata,
    value_name = "Abundance"
  ) %>%
    dplyr::filter(
      is.finite(Abundance)
    )

  sample_order <- long_data %>%
    dplyr::group_by(Sample) %>%
    dplyr::summarise(
      median_abundance =
        stats::median(
          Abundance,
          na.rm = TRUE
        ),
      .groups = "drop"
    ) %>%
    dplyr::arrange(
      median_abundance
    ) %>%
    dplyr::pull(Sample)

  long_data$Sample <- factor(
    long_data$Sample,
    levels = sample_order
  )

  group_colours <- make_group_palette(
    long_data$Group
  )

  ggplot(
    long_data,
    aes(
      x = Sample,
      y = Abundance,
      fill = Group
    )
  ) +
    geom_boxplot(
      outlier.shape = NA,
      width = 0.75,
      linewidth = 0.25
    ) +
    scale_fill_manual(
      values = group_colours,
      na.value = "grey70"
    ) +
    labs(
      title = paste(
        assay_name,
        title_suffix,
        sep = " — "
      ),
      x = NULL,
      y = y_label,
      fill = "Group"
    ) +
    theme_classic() +
    theme(
      axis.text.x = element_text(
        angle = 90,
        hjust = 1,
        vjust = 0.5,
        size =
          distribution_sample_fontsize
      )
    )
}


plot_sample_densities <- function(
    mat,
    metadata,
    assay_name,
    title_suffix,
    x_label
) {
  long_data <- matrix_to_long(
    mat,
    metadata,
    value_name = "Abundance"
  ) %>%
    dplyr::filter(
      is.finite(Abundance)
    )

  group_colours <- make_group_palette(
    long_data$Group
  )

  ggplot(
    long_data,
    aes(
      x = Abundance,
      group = Sample,
      colour = Group
    )
  ) +
    geom_density(
      linewidth = 0.45,
      alpha = 0.65
    ) +
    scale_colour_manual(
      values = group_colours,
      na.value = "grey65"
    ) +
    labs(
      title = paste(
        assay_name,
        title_suffix,
        sep = " — "
      ),
      x = x_label,
      y = "Density",
      colour = "Group"
    ) +
    theme_classic()
}


calculate_sample_metrics <- function(
    mat,
    metadata,
    assay_name
) {
  mat <- as.matrix(mat)

  metrics <- tibble::tibble(
    Sample = colnames(mat),
    number_detected =
      matrixStats::colSums2(
        is.finite(mat)
      ),
    missing_fraction =
      matrixStats::colMeans2(
        !is.finite(mat)
      ),
    median_abundance =
      matrixStats::colMedians(
        mat,
        na.rm = TRUE
      ),
    mean_abundance =
      matrixStats::colMeans2(
        mat,
        na.rm = TRUE
      )
  ) %>%
    dplyr::left_join(
      metadata %>%
        dplyr::select(
          Sample,
          Group
        ),
      by = "Sample"
    ) %>%
    dplyr::mutate(
      assay = assay_name
    )

  metrics
}


plot_detected_per_sample <- function(
    metrics,
    assay_name,
    title_suffix
) {
  plot_data <- metrics %>%
    dplyr::arrange(
      number_detected
    ) %>%
    dplyr::mutate(
      Sample = factor(
        Sample,
        levels = Sample
      )
    )

  group_colours <- make_group_palette(
    plot_data$Group
  )

  ggplot(
    plot_data,
    aes(
      x = Sample,
      y = number_detected,
      fill = Group
    )
  ) +
    geom_col() +
    geom_hline(
      yintercept = stats::median(
        plot_data$number_detected,
        na.rm = TRUE
      ),
      linetype = 2,
      linewidth = 0.4
    ) +
    scale_fill_manual(
      values = group_colours,
      na.value = "grey70"
    ) +
    labs(
      title = paste(
        assay_name,
        title_suffix,
        sep = " — "
      ),
      subtitle =
        "Dashed line: cohort median",
      x = NULL,
      y = "Number of detected proteins",
      fill = "Group"
    ) +
    theme_classic() +
    theme(
      axis.text.x = element_text(
        angle = 90,
        hjust = 1,
        vjust = 0.5,
        size =
          distribution_sample_fontsize
      )
    )
}


plot_missing_fraction <- function(
    metrics,
    assay_name,
    title_suffix
) {
  plot_data <- metrics %>%
    dplyr::arrange(
      missing_fraction
    ) %>%
    dplyr::mutate(
      Sample = factor(
        Sample,
        levels = Sample
      )
    )

  group_colours <- make_group_palette(
    plot_data$Group
  )

  ggplot(
    plot_data,
    aes(
      x = Sample,
      y = missing_fraction,
      fill = Group
    )
  ) +
    geom_col() +
    scale_y_continuous(
      labels = scales::percent
    ) +
    scale_fill_manual(
      values = group_colours,
      na.value = "grey70"
    ) +
    labs(
      title = paste(
        assay_name,
        title_suffix,
        sep = " — "
      ),
      x = NULL,
      y = "Missing fraction",
      fill = "Group"
    ) +
    theme_classic() +
    theme(
      axis.text.x = element_text(
        angle = 90,
        hjust = 1,
        vjust = 0.5,
        size =
          distribution_sample_fontsize
      )
    )
}


plot_median_abundance <- function(
    metrics,
    assay_name,
    title_suffix,
    y_label
) {
  plot_data <- metrics %>%
    dplyr::arrange(
      median_abundance
    ) %>%
    dplyr::mutate(
      Sample = factor(
        Sample,
        levels = Sample
      )
    )

  group_colours <- make_group_palette(
    plot_data$Group
  )

  ggplot(
    plot_data,
    aes(
      x = Sample,
      y = median_abundance,
      fill = Group
    )
  ) +
    geom_col() +
    scale_fill_manual(
      values = group_colours,
      na.value = "grey70"
    ) +
    labs(
      title = paste(
        assay_name,
        title_suffix,
        sep = " — "
      ),
      x = NULL,
      y = y_label,
      fill = "Group"
    ) +
    theme_classic() +
    theme(
      axis.text.x = element_text(
        angle = 90,
        hjust = 1,
        vjust = 0.5,
        size =
          distribution_sample_fontsize
      )
    )
}


plot_feature_missingness <- function(
    mat,
    assay_name,
    title_suffix
) {
  missing_fraction <- matrixStats::rowMeans2(
    !is.finite(mat)
  )

  plot_data <- tibble::tibble(
    missing_fraction =
      missing_fraction
  )

  ggplot(
    plot_data,
    aes(
      x = missing_fraction
    )
  ) +
    geom_histogram(
      bins = 40,
      colour = "white",
      linewidth = 0.2
    ) +
    scale_x_continuous(
      labels = scales::percent
    ) +
    labs(
      title = paste(
        assay_name,
        title_suffix,
        sep = " — "
      ),
      x = "Missing fraction per protein",
      y = "Number of proteins"
    ) +
    theme_classic()
}


draw_sample_correlation_heatmap <- function(
    mat,
    metadata,
    assay_name,
    title_suffix,
    include_olink_technical = FALSE,
    include_sample_age = FALSE
) {
  correlation <- stats::cor(
    mat,
    use = "pairwise.complete.obs",
    method = "pearson"
  )

  annotation <- make_sample_annotation(
    metadata = metadata,
    samples = colnames(correlation),
    include_olink_technical =
      include_olink_technical,
    include_sample_age =
      include_sample_age
  )

  heatmap <- ComplexHeatmap::Heatmap(
    correlation,
    name = "Pearson r",
    top_annotation = annotation,
    show_row_names = TRUE,
    show_column_names = TRUE,
    row_names_gp = grid::gpar(
      fontsize = heatmap_column_fontsize
    ),
    column_names_gp = grid::gpar(
      fontsize = heatmap_column_fontsize
    ),
    col = circlize::colorRamp2(
      c(
        0,
        0.9,
        1
      ),
      c(
        "#2166AC",
        "white",
        "#B2182B"
      )
    ),
    column_title = paste(
      assay_name,
      title_suffix,
      sep = " — "
    )
  )

  ComplexHeatmap::draw(
    heatmap,
    heatmap_legend_side = "right",
    annotation_legend_side = "right"
  )

  invisible(heatmap)
}


prepare_qc_set <- function(
    excluded_samples,
    label
) {
  clinical_use <- clinical_non_control %>%
    dplyr::filter(
      !Sample %in%
        excluded_samples
    )

  olink_samples <- clinical_use$Sample[
    clinical_use$Sample %in%
      colnames(olink_expression_all)
  ]

  ms_samples <- clinical_use$Sample[
    clinical_use$Sample %in%
      colnames(ms_expression_log2_all)
  ]

  olink_matrix <- olink_expression_all[
    ,
    olink_samples,
    drop = FALSE
  ]

  ms_matrix <- ms_expression_log2_all[
    ,
    ms_samples,
    drop = FALSE
  ]

  olink_metadata <- clinical_use %>%
    dplyr::left_join(
      olink_sample_metadata %>%
        dplyr::select(
          Sample,
          Batch,
          BCA_mg_ml
        ),
      by = "Sample"
    ) %>%
    dplyr::filter(
      Sample %in%
        olink_samples
    ) %>%
    dplyr::slice(
      match(
        olink_samples,
        Sample
      )
    )

  ms_metadata <- clinical_use %>%
    dplyr::filter(
      Sample %in%
        ms_samples
    ) %>%
    dplyr::slice(
      match(
        ms_samples,
        Sample
      )
    )

  availability <- clinical_use %>%
    dplyr::transmute(
      Sample,
      Group,
      has_Olink =
        Sample %in%
        colnames(
          olink_expression_all
        ),
      has_LCMS =
        Sample %in%
        colnames(
          ms_expression_log2_all
        )
    ) %>%
    dplyr::mutate(
      availability = dplyr::case_when(
        has_Olink &
          has_LCMS ~
          "Both Olink and LC-MS/MS",
        has_Olink &
          !has_LCMS ~
          "Olink only",
        !has_Olink &
          has_LCMS ~
          "LC-MS/MS only",
        TRUE ~ "Neither"
      )
    ) %>%
    dplyr::arrange(
      availability,
      Group,
      Sample
    )

  availability_summary <- availability %>%
    dplyr::count(
      availability,
      name = "number_samples"
    ) %>%
    tidyr::complete(
      availability = c(
        "Both Olink and LC-MS/MS",
        "Olink only",
        "LC-MS/MS only",
        "Neither"
      ),
      fill = list(
        number_samples = 0L
      )
    ) %>%
    dplyr::bind_rows(
      tibble::tibble(
        availability =
          "Total non-control samples",
        number_samples =
          nrow(availability)
      )
    )

  list(
    label = label,
    excluded_samples =
      excluded_samples,
    clinical = clinical_use,
    olink = olink_matrix,
    ms = ms_matrix,
    olink_metadata =
      olink_metadata,
    ms_metadata =
      ms_metadata,
    availability =
      availability,
    availability_summary =
      availability_summary
  )
}


write_qc_tables <- function(
    qc_object,
    suffix
) {
  write.csv(
    qc_object$availability,
    file.path(
      output_dir,
      paste0(
        "sample_availability_",
        suffix,
        ".csv"
      )
    ),
    row.names = FALSE
  )

  write.csv(
    qc_object$availability_summary,
    file.path(
      output_dir,
      paste0(
        "sample_availability_summary_",
        suffix,
        ".csv"
      )
    ),
    row.names = FALSE
  )

  olink_metrics <- calculate_sample_metrics(
    qc_object$olink,
    qc_object$olink_metadata,
    assay_name = "Olink"
  )

  ms_metrics <- calculate_sample_metrics(
    qc_object$ms,
    qc_object$ms_metadata,
    assay_name = "LC-MS/MS"
  )

  write.csv(
    olink_metrics,
    file.path(
      output_dir,
      paste0(
        "Olink_sample_metrics_",
        suffix,
        ".csv"
      )
    ),
    row.names = FALSE
  )

  write.csv(
    ms_metrics,
    file.path(
      output_dir,
      paste0(
        "LCMS_sample_metrics_",
        suffix,
        ".csv"
      )
    ),
    row.names = FALSE
  )

  invisible(
    list(
      olink_metrics =
        olink_metrics,
      ms_metrics =
        ms_metrics
    )
  )
}
```

# Import clinical and proteomics data

```{r import clinical metadata}
clinical_all <- readxl::read_excel(
  clinical_file
) %>%
  dplyr::select(
    -dplyr::matches(
      "^Unnamed"
    )
  ) %>%
  dplyr::mutate(
    Sample = normalise_sample_id(
      Sample
    ),
    Group = trimws(
      as.character(
        Group
      )
    ),
    Sample_age_days =
      as.numeric(
        Sample_age_days
      ),
    Age_years = as.numeric(
      Age_years
    )
  ) %>%
  dplyr::filter(
    !is.na(Sample),
    nzchar(Sample)
  ) %>%
  dplyr::distinct(
    Sample,
    .keep_all = TRUE
  )

clinical_non_control <- clinical_all %>%
  dplyr::filter(
    Group != "HC"
  ) %>%
  dplyr::mutate(
    Group = factor(
      Group,
      levels = c(
        "DCM",
        "ICM"
      )
    )
  )

clinical_all %>%
  dplyr::count(
    Group,
    name = "number_samples"
  )
```

```{r import Olink data}
olink_sample_metadata <- readxl::read_excel(
  olink_sample_file
) %>%
  dplyr::select(
    -dplyr::matches(
      "^Unnamed"
    )
  ) %>%
  dplyr::mutate(
    Sample = normalise_sample_id(
      Sample
    ),
    BCA_mg_ml =
      as.numeric(
        BCA_mg_ml
      ),
    Batch = factor(
      Batch
    )
  ) %>%
  dplyr::distinct(
    Sample,
    .keep_all = TRUE
  )

olink_raw <- read.delim(
  olink_quant_file,
  sep = "\t",
  check.names = FALSE
)

if (
  !"SampleID" %in%
    colnames(olink_raw)
) {
  stop(
    "The Olink file does not contain a SampleID column."
  )
}

olink_feature_ids <- sub(
  "_.*",
  "",
  as.character(
    olink_raw$SampleID
  )
)

olink_expression_initial <- as.matrix(
  olink_raw[
    ,
    setdiff(
      colnames(olink_raw),
      "SampleID"
    ),
    drop = FALSE
  ]
)

storage.mode(
  olink_expression_initial
) <- "numeric"

colnames(
  olink_expression_initial
) <- normalise_sample_id(
  colnames(
    olink_expression_initial
  )
)

olink_expression_all <- limma::avereps(
  olink_expression_initial,
  ID = olink_feature_ids
)

olink_expression_all <-
  filter_usable_features(
    olink_expression_all
  )

dim(
  olink_expression_all
)
```

```{r import LCMS data}
ms_raw <- read.csv(
  ms_file,
  check.names = FALSE
)

ms_annotation_columns <- c(
  "Protein.Group",
  "Protein.Names",
  "Genes",
  "First.Protein.Description",
  "N.Sequences",
  "N.Proteotypic.Sequences"
)

missing_ms_annotation_columns <- setdiff(
  ms_annotation_columns,
  colnames(ms_raw)
)

if (
  length(
    missing_ms_annotation_columns
  ) > 0L
) {
  stop(
    "Missing expected LC-MS/MS annotation columns: ",
    paste(
      missing_ms_annotation_columns,
      collapse = ", "
    )
  )
}

ms_sample_columns <- setdiff(
  colnames(ms_raw),
  ms_annotation_columns
)

ms_qc_columns <- grep(
  "^QC",
  ms_sample_columns,
  value = TRUE
)

ms_biological_columns <- setdiff(
  ms_sample_columns,
  ms_qc_columns
)

ms_gene <- sub(
  ";.*$",
  "",
  as.character(
    ms_raw$Genes
  )
)

ms_gene <- trimws(
  ms_gene
)

keep_ms_rows <- !is.na(ms_gene) &
  nzchar(ms_gene)

ms_expression_initial <- as.matrix(
  ms_raw[
    keep_ms_rows,
    ms_biological_columns,
    drop = FALSE
  ]
)

storage.mode(
  ms_expression_initial
) <- "numeric"

colnames(
  ms_expression_initial
) <- normalise_sample_id(
  colnames(
    ms_expression_initial
  )
)

ms_gene <- ms_gene[
  keep_ms_rows
]

ms_expression_collapsed <- limma::avereps(
  ms_expression_initial,
  ID = ms_gene
)

ms_expression_log2_all <- log2(
  ms_expression_collapsed + 1
)

if (isTRUE(ms_zero_is_missing)) {
  ms_expression_log2_all[
    !is.finite(
      ms_expression_collapsed
    ) |
      ms_expression_collapsed <= 0
  ] <- NA_real_
} else {
  ms_expression_log2_all[
    !is.finite(
      ms_expression_log2_all
    )
  ] <- NA_real_
}

ms_expression_log2_all <-
  filter_usable_features(
    ms_expression_log2_all
  )

dim(
  ms_expression_log2_all
)
```

```{r identify unmatched raw samples}
raw_sample_inventory <- dplyr::bind_rows(
  tibble::tibble(
    Sample =
      colnames(
        olink_expression_all
      ),
    assay = "Olink"
  ),
  tibble::tibble(
    Sample =
      colnames(
        ms_expression_log2_all
      ),
    assay = "LC-MS/MS"
  )
) %>%
  dplyr::distinct() %>%
  dplyr::left_join(
    clinical_all %>%
      dplyr::select(
        Sample,
        Group
      ),
    by = "Sample"
  ) %>%
  dplyr::mutate(
    present_in_clinical_metadata =
      !is.na(Group),
    is_control =
      Group == "HC"
  ) %>%
  dplyr::arrange(
    assay,
    Sample
  )

write.csv(
  raw_sample_inventory,
  file.path(
    output_dir,
    "raw_proteomics_sample_inventory.csv"
  ),
  row.names = FALSE
)

raw_sample_inventory %>%
  dplyr::filter(
    !present_in_clinical_metadata |
      is_control
  ) %>%
  reactable::reactable(
    searchable = TRUE,
    filterable = TRUE,
    defaultPageSize = 15
  )
```

# Part 1: QC before sample exclusion

```{r prepare QC before exclusion}
qc_before <- prepare_qc_set(
  excluded_samples =
    character(0),
  label =
    "Before exclusion"
)

metrics_before <- write_qc_tables(
  qc_before,
  suffix = "before_exclusion"
)
```

## Sample availability before exclusion

```{r availability summary before}
qc_before$availability_summary %>%
  reactable::reactable(
    searchable = FALSE,
    pagination = FALSE
  )
```

```{r availability table before}
qc_before$availability %>%
  reactable::reactable(
    searchable = TRUE,
    filterable = TRUE,
    defaultPageSize = 20
  )
```

## Olink QC before exclusion

### NPX distribution by sample

```{r Olink boxplots before, fig.height=6.5, fig.width=15}
plot_sample_boxplots(
  mat = qc_before$olink,
  metadata =
    qc_before$olink_metadata,
  assay_name = "Olink",
  title_suffix =
    "NPX distributions",
  y_label = "NPX"
)
```

```{r Olink densities before, fig.height=5.5, fig.width=9}
plot_sample_densities(
  mat = qc_before$olink,
  metadata =
    qc_before$olink_metadata,
  assay_name = "Olink",
  title_suffix =
    "Sample density distributions",
  x_label = "NPX"
)
```

### Olink PCA

```{r Olink PCA object before}
olink_pca_before <- prepare_pca(
  mat = qc_before$olink,
  metadata =
    qc_before$olink_metadata,
  assay_name = "Olink"
)
```

```{r Olink PCA before, fig.height=7, fig.width=10}
plot_pca_group(
  olink_pca_before,
  title_suffix =
    "PCA"
)
```

### Olink sample correlation

```{r Olink correlation before, fig.height=11, fig.width=12}
draw_sample_correlation_heatmap(
  mat = qc_before$olink,
  metadata =
    qc_before$olink_metadata,
  assay_name = "Olink",
  title_suffix =
    "Sample correlations",
  include_olink_technical = TRUE
)
```

### Olink heatmap: all usable proteins and samples

```{r Olink heatmap before, fig.height=12, fig.width=14}
draw_expression_heatmap(
  mat = qc_before$olink,
  metadata =
    qc_before$olink_metadata,
  assay_name = "Olink",
  title_suffix =
    "All usable proteins",
  imputation = "median",
  include_olink_technical = TRUE
)
```

## LC-MS/MS QC before exclusion

### Number of proteins detected per sample

```{r LCMS detected proteins before, fig.height=6.5, fig.width=15}
plot_detected_per_sample(
  metrics =
    metrics_before$ms_metrics,
  assay_name = "LC-MS/MS",
  title_suffix =
    "Detected proteins"
)
```

### LC-MS/MS abundance distributions

```{r LCMS boxplots before, fig.height=6.5, fig.width=15}
plot_sample_boxplots(
  mat = qc_before$ms,
  metadata =
    qc_before$ms_metadata,
  assay_name = "LC-MS/MS",
  title_suffix =
    "Log2 abundance distributions",
  y_label = "Log2 abundance"
)
```

```{r LCMS densities before, fig.height=5.5, fig.width=9}
plot_sample_densities(
  mat = qc_before$ms,
  metadata =
    qc_before$ms_metadata,
  assay_name = "LC-MS/MS",
  title_suffix =
    "Sample density distributions",
  x_label = "Log2 abundance"
)
```

### Additional LC-MS/MS diagnostic plots

```{r LCMS diagnostics before, fig.height=10, fig.width=15}
(
  plot_missing_fraction(
    metrics =
      metrics_before$ms_metrics,
    assay_name = "LC-MS/MS",
    title_suffix =
      "Missingness"
  ) |
    plot_median_abundance(
      metrics =
        metrics_before$ms_metrics,
      assay_name = "LC-MS/MS",
      title_suffix =
        "Median abundance",
      y_label =
        "Median log2 abundance"
    )
) /
  plot_feature_missingness(
    mat = qc_before$ms,
    assay_name = "LC-MS/MS",
    title_suffix =
      "Protein missingness"
  )
```

### LC-MS/MS PCA

```{r LCMS PCA object before}
ms_pca_before <- prepare_pca(
  mat = qc_before$ms,
  metadata =
    qc_before$ms_metadata,
  assay_name = "LC-MS/MS"
)
```

```{r LCMS PCA before, fig.height=7, fig.width=10}
plot_pca_group(
  ms_pca_before,
  title_suffix =
    "PCA"
)
```

### LC-MS/MS sample correlation

```{r LCMS correlation before, fig.height=11, fig.width=12}
draw_sample_correlation_heatmap(
  mat = qc_before$ms,
  metadata =
    qc_before$ms_metadata,
  assay_name = "LC-MS/MS",
  title_suffix =
    "Sample correlations",
  include_sample_age = TRUE
)
```

### LC-MS/MS heatmap: all usable proteins and samples

```{r LCMS heatmap before, fig.height=12, fig.width=14}
draw_expression_heatmap(
  mat = qc_before$ms,
  metadata =
    qc_before$ms_metadata,
  assay_name = "LC-MS/MS",
  title_suffix =
    "All usable proteins",
  imputation = "low",
  include_sample_age = TRUE
)
```

# Sample exclusions

```{r define sample exclusions}
poor_quality_samples <- c(
  "P11",
  "PK61",
  "PK63"
)

ambiguous_identity_samples <- "PK69"

icm_samples <- clinical_non_control %>%
  dplyr::filter(
    Group == "ICM"
  ) %>%
  dplyr::pull(Sample)

sample_exclusions <- dplyr::bind_rows(
  tibble::tibble(
    Sample = poor_quality_samples,
    Exclusion_reason =
      "Poor proteomic data quality"
  ),
  tibble::tibble(
    Sample =
      ambiguous_identity_samples,
    Exclusion_reason =
      "Ambiguous patient assignment due to duplicate sample identifier"
  ),
  tibble::tibble(
    Sample = icm_samples,
    Exclusion_reason =
      "ICM excluded from the DCM-focused analysis"
  )
) %>%
  dplyr::group_by(Sample) %>%
  dplyr::summarise(
    Exclusion_reason = paste(
      unique(Exclusion_reason),
      collapse = "; "
    ),
    .groups = "drop"
  ) %>%
  dplyr::left_join(
    qc_before$availability,
    by = "Sample"
  ) %>%
  dplyr::arrange(
    Exclusion_reason,
    Sample
  )

samples_to_exclude <- unique(
  sample_exclusions$Sample
)

write.csv(
  sample_exclusions,
  file.path(
    output_dir,
    "sample_exclusions.csv"
  ),
  row.names = FALSE
)

reactable::reactable(
  sample_exclusions,
  searchable = TRUE,
  filterable = TRUE,
  pagination = FALSE
)
```

# Part 2: QC after sample exclusion

```{r prepare QC after exclusion}
qc_after <- prepare_qc_set(
  excluded_samples =
    samples_to_exclude,
  label =
    "After manual exclusion"
)

metrics_after <- write_qc_tables(
  qc_after,
  suffix = "after_exclusion"
)
```

## Sample availability after exclusion

```{r availability summary after}
qc_after$availability_summary %>%
  reactable::reactable(
    searchable = FALSE,
    pagination = FALSE
  )
```

```{r availability table after}
qc_after$availability %>%
  reactable::reactable(
    searchable = TRUE,
    filterable = TRUE,
    defaultPageSize = 20
  )
```

## Olink QC after exclusion

### NPX distribution by sample

```{r Olink boxplots after, fig.height=6.5, fig.width=15}
plot_sample_boxplots(
  mat = qc_after$olink,
  metadata =
    qc_after$olink_metadata,
  assay_name = "Olink",
  title_suffix =
    "NPX distributions",
  y_label = "NPX"
)
```

```{r Olink densities after, fig.height=5.5, fig.width=9}
plot_sample_densities(
  mat = qc_after$olink,
  metadata =
    qc_after$olink_metadata,
  assay_name = "Olink",
  title_suffix =
    "Sample density distributions",
  x_label = "NPX"
)
```

### Olink PCA

```{r Olink PCA object after}
olink_pca_after <- prepare_pca(
  mat = qc_after$olink,
  metadata =
    qc_after$olink_metadata,
  assay_name = "Olink"
)
```

```{r Olink PCA after, fig.height=7, fig.width=10}
plot_pca_group(
  olink_pca_after,
  title_suffix =
    "PCA"
)
```

### Olink sample correlation

```{r Olink correlation after, fig.height=11, fig.width=12}
draw_sample_correlation_heatmap(
  mat = qc_after$olink,
  metadata =
    qc_after$olink_metadata,
  assay_name = "Olink",
  title_suffix =
    "Sample correlations",
  include_olink_technical = TRUE
)
```

### Olink heatmap: all usable proteins and samples

```{r Olink heatmap after, fig.height=12, fig.width=14}
draw_expression_heatmap(
  mat = qc_after$olink,
  metadata =
    qc_after$olink_metadata,
  assay_name = "Olink",
  title_suffix =
    "All usable proteins",
  imputation = "median",
  include_olink_technical = TRUE
)
```

## LC-MS/MS QC after exclusion

### Number of proteins detected per sample

```{r LCMS detected proteins after, fig.height=3.5, fig.width=7}
plot_detected_per_sample(
  metrics =
    metrics_after$ms_metrics,
  assay_name = "LC-MS/MS",
  title_suffix =
    "Detected proteins"
)
```

### LC-MS/MS abundance distributions

```{r LCMS boxplots after, fig.height=6.5, fig.width=15}
plot_sample_boxplots(
  mat = qc_after$ms,
  metadata =
    qc_after$ms_metadata,
  assay_name = "LC-MS/MS",
  title_suffix =
    "Log2 abundance distributions",
  y_label = "Log2 abundance"
)
```

```{r LCMS densities after, fig.height=5.5, fig.width=9}
plot_sample_densities(
  mat = qc_after$ms,
  metadata =
    qc_after$ms_metadata,
  assay_name = "LC-MS/MS",
  title_suffix =
    "Sample density distributions",
  x_label = "Log2 abundance"
)
```

### Additional LC-MS/MS diagnostic plots

```{r LCMS diagnostics after, fig.height=10, fig.width=15}
(
  plot_missing_fraction(
    metrics =
      metrics_after$ms_metrics,
    assay_name = "LC-MS/MS",
    title_suffix =
      "Missingness"
  ) |
    plot_median_abundance(
      metrics =
        metrics_after$ms_metrics,
      assay_name = "LC-MS/MS",
      title_suffix =
        "Median abundance",
      y_label =
        "Median log2 abundance"
    )
) /
  plot_feature_missingness(
    mat = qc_after$ms,
    assay_name = "LC-MS/MS",
    title_suffix =
      "Protein missingness"
  )
```

### LC-MS/MS PCA

```{r LCMS PCA object after}
ms_pca_after <- prepare_pca(
  mat = qc_after$ms,
  metadata =
    qc_after$ms_metadata,
  assay_name = "LC-MS/MS"
)
```

```{r LCMS PCA after, fig.height=7, fig.width=10}
plot_pca_group(
  ms_pca_after,
  title_suffix =
    "PCA"
)
```

### LC-MS/MS sample correlation

```{r LCMS correlation after, fig.height=11, fig.width=12}
draw_sample_correlation_heatmap(
  mat = qc_after$ms,
  metadata =
    qc_after$ms_metadata,
  assay_name = "LC-MS/MS",
  title_suffix =
    "Sample correlations",
  include_sample_age = TRUE
)
```

### LC-MS/MS heatmap: all usable proteins and samples

```{r LCMS heatmap after, fig.height=12, fig.width=14}
draw_expression_heatmap(
  mat = qc_after$ms,
  metadata =
    qc_after$ms_metadata,
  assay_name = "LC-MS/MS",
  title_suffix =
    "All usable proteins",
  imputation = "low",
  include_sample_age = TRUE
)
```

# Before-versus-after sample counts

```{r compare before and after sample counts}
availability_comparison <- dplyr::bind_rows(
  qc_before$availability_summary %>%
    dplyr::mutate(
      QC_stage = "Before exclusion"
    ),
  qc_after$availability_summary %>%
    dplyr::mutate(
      QC_stage = "After exclusion"
    )
) %>%
  dplyr::select(
    QC_stage,
    availability,
    number_samples
  )

write.csv(
  availability_comparison,
  file.path(
    output_dir,
    "sample_availability_before_after_comparison.csv"
  ),
  row.names = FALSE
)

reactable::reactable(
  availability_comparison,
  searchable = FALSE,
  pagination = FALSE
)
```

# Save QC objects

```{r save QC objects}
saveRDS(
  list(
    settings = list(
      samples_to_exclude =
        samples_to_exclude,
      sample_exclusions =
        sample_exclusions,
      ms_zero_is_missing =
        ms_zero_is_missing,
      pca_min_observed_fraction =
        pca_min_observed_fraction
    ),
    raw_sample_inventory =
      raw_sample_inventory,
    before_exclusion =
      qc_before,
    after_exclusion =
      qc_after,
    metrics_before =
      metrics_before,
    metrics_after =
      metrics_after
  ),
  file.path(
    output_dir,
    "Proteomics_QC_complete_objects.rds"
  )
)
```

# Clinical correlation structure

```{r additional quality control libraries}
library(Hmisc)
library(ggtext)
library(variancePartition)
```

```{r additional quality control settings}
additional_qc_figure_dir <- file.path(
  figure_dir,
  "additional_quality_control"
)

dir.create(
  additional_qc_figure_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

clinical_variable_labels <- c(
  Age_years = "Age",
  Sample_age_days = "Sample age",
  mPAP_mmHg = "mPAP",
  sPAP_mmHg = "sPAP",
  PCW_mmHg = "PCWP",
  PVR_WU = "PVR",
  TAPSE_mm = "TAPSE",
  RVIT_mm = "RVIT",
  LVEF_percent = "LVEF",
  LVEDD_mm = "LVEDD",
  BCA_mg_ml = "BCA",
  Residuals = "Unexplained"
)

protein_heatmap_colors <-
  circlize::colorRamp2(
    c(
      -2,
      0,
      2
    ),
    c(
      "#2166AC",
      "white",
      "#B2182B"
    )
  )
```

```{r clinical correlation matrix, fig.height=7.8, fig.width=8}
clinical_correlation_variables <- c(
  "mPAP_mmHg",
  "sPAP_mmHg",
  "PCW_mmHg",
  "PVR_WU",
  "TAPSE_mm",
  "RVIT_mm",
  "LVEF_percent",
  "LVEDD_mm",
  "Age_years",
  "Sample_age_days"
)

final_proteomics_samples <-
  union(
    colnames(
      qc_after$olink
    ),
    colnames(
      qc_after$ms
    )
  )

clinical_correlation_variables <-
  intersect(
    clinical_correlation_variables,
    colnames(
      qc_after$clinical
    )
  )

clinical_correlation_data <-
  qc_after$clinical %>%
  dplyr::filter(
    Sample %in%
      final_proteomics_samples
  ) %>%
  dplyr::mutate(
    dplyr::across(
      dplyr::all_of(
        clinical_correlation_variables
      ),
      as.numeric
    )
  ) %>%
  dplyr::select(
    dplyr::all_of(
      clinical_correlation_variables
    )
  )

clinical_correlation_variables <-
  clinical_correlation_variables[
    vapply(
      clinical_correlation_data[
        clinical_correlation_variables
      ],
      function(values) {
        finite_values <-
          values[
            is.finite(values)
          ]

        length(finite_values) >= 3L &&
          stats::sd(
            finite_values
          ) > 0
      },
      logical(1)
    )
  ]

clinical_correlation_data <-
  clinical_correlation_data %>%
  dplyr::select(
    dplyr::all_of(
      clinical_correlation_variables
    )
  )

clinical_correlation_result <-
  Hmisc::rcorr(
    as.matrix(
      clinical_correlation_data
    ),
    type = "pearson"
  )

clinical_correlation_matrix <-
  clinical_correlation_result$r[
    clinical_correlation_variables,
    clinical_correlation_variables,
    drop = FALSE
  ]

clinical_correlation_p_matrix <-
  clinical_correlation_result$P[
    clinical_correlation_variables,
    clinical_correlation_variables,
    drop = FALSE
  ]

clinical_correlation_long <-
  as.data.frame(
    clinical_correlation_matrix
  ) %>%
  tibble::rownames_to_column(
    "variable_y"
  ) %>%
  tidyr::pivot_longer(
    cols = -variable_y,
    names_to = "variable_x",
    values_to = "correlation"
  ) %>%
  dplyr::left_join(
    as.data.frame(
      clinical_correlation_p_matrix
    ) %>%
      tibble::rownames_to_column(
        "variable_y"
      ) %>%
      tidyr::pivot_longer(
        cols = -variable_y,
        names_to = "variable_x",
        values_to = "p_value"
      ),
    by = c(
      "variable_y",
      "variable_x"
    )
  ) %>%
  dplyr::mutate(
    variable_x = factor(
      variable_x,
      levels =
        clinical_correlation_variables,
      labels =
        unname(
          clinical_variable_labels[
            clinical_correlation_variables
          ]
        )
    ),
    variable_y = factor(
      variable_y,
      levels =
        rev(
          clinical_correlation_variables
        ),
      labels =
        unname(
          clinical_variable_labels[
            rev(
              clinical_correlation_variables
            )
          ]
        )
    ),
    correlation_label =
      dplyr::if_else(
        is.finite(correlation),
        sprintf(
          "%.2f",
          correlation
        ),
        ""
      ),
    p_label =
      dplyr::case_when(
        is.na(p_value) ~ "",
        p_value < 0.001 ~
          "P < 0.001",
        TRUE ~
          sprintf(
            "P = %.3f",
            p_value
          )
      ),
    highlighted =
      is.finite(correlation) &
      is.finite(p_value) &
      abs(correlation) > 0.40 &
      p_value < 0.05,
    combined_label =
      dplyr::if_else(
        highlighted,
        paste0(
          "<b>",
          correlation_label,
          "</b><br><b>",
          p_label,
          "</b>"
        ),
        paste0(
          correlation_label,
          "<br>",
          p_label
        )
      )
  )

clinical_correlation_plot <-
  ggplot(
    clinical_correlation_long,
    aes(
      x = variable_x,
      y = variable_y,
      fill = correlation
    )
  ) +
  geom_tile(
    colour = "grey85",
    linewidth = 0.4
  ) +
  geom_tile(
    data =
      dplyr::filter(
        clinical_correlation_long,
        highlighted
      ),
    fill = NA,
    colour = "black",
    linewidth = 0.8
  ) +
  ggtext::geom_richtext(
    aes(
      label =
        combined_label
    ),
    fill = NA,
    label.colour = NA,
    size = 2.8,
    lineheight = 0.9
  ) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(
      -1,
      1
    ),
    oob =
      scales::squish,
    name =
      "Pearson r"
  ) +
  coord_fixed() +
  labs(
    x = NULL,
    y = NULL
  ) +
  theme_classic(
    base_size = 15
  ) +
  theme(
    axis.text.x =
      element_text(
        angle = 45,
        hjust = 1,
        face = "bold",
        colour = "black"
      ),
    axis.text.y =
      element_text(
        face = "bold",
        colour = "black"
      ),
    axis.line =
      element_blank()
  )

clinical_correlation_plot

write.csv(
  clinical_correlation_long,
  file.path(
    output_dir,
    "clinical_correlations.csv"
  ),
  row.names = FALSE
)
```

# Variance partitioning

The contribution of clinical and technical variables to protein abundance
variation was estimated separately for the Olink and LC-MS/MS datasets.

```{r variance partition helper functions}
run_proteomics_variance_partition <- function(
    expression_matrix,
    metadata,
    candidate_variables
) {
  variables <-
    intersect(
      candidate_variables,
      colnames(metadata)
    )

  analysis_metadata <-
    metadata %>%
    dplyr::select(
      Sample,
      dplyr::all_of(
        variables
      )
    ) %>%
    dplyr::distinct(
      Sample,
      .keep_all = TRUE
    ) %>%
    dplyr::filter(
      Sample %in%
        colnames(
          expression_matrix
        )
    ) %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(
          variables
        ),
        as.numeric
      )
    ) %>%
    tidyr::drop_na()

  variables <-
    variables[
      vapply(
        analysis_metadata[
          variables
        ],
        function(values) {
          is.finite(
            stats::sd(values)
          ) &&
            stats::sd(values) > 0
        },
        logical(1)
      )
    ]

  analysis_metadata <-
    analysis_metadata %>%
    dplyr::select(
      Sample,
      dplyr::all_of(
        variables
      )
    )

  if (
    length(variables) == 0L
  ) {
    stop(
      "No usable variables were available for variance partitioning."
    )
  }

  if (
    nrow(analysis_metadata) <=
      length(variables) + 2L
  ) {
    stop(
      "Too few complete samples for the selected variance-partition model."
    )
  }

  analysis_metadata <-
    analysis_metadata %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(
          variables
        ),
        ~ as.numeric(
          scale(.x)
        )
      )
    )

  analysis_matrix <-
    expression_matrix[
      ,
      analysis_metadata$Sample,
      drop = FALSE
    ]

  observed_values <-
    matrixStats::rowSums2(
      is.finite(
        analysis_matrix
      )
    )

  feature_variance <-
    matrixStats::rowVars(
      analysis_matrix,
      na.rm = TRUE
    )

  usable_features <-
    observed_values >= 3L &
    is.finite(feature_variance) &
    feature_variance > 0

  analysis_matrix <-
    analysis_matrix[
      usable_features,
      ,
      drop = FALSE
    ]

  variance_formula <-
    stats::reformulate(
      variables
    )

  variance_partition <-
    variancePartition::fitExtractVarPartModel(
      analysis_matrix,
      variance_formula,
      analysis_metadata
    )

  colnames(
    variance_partition
  ) <-
    gsub(
      "`",
      "",
      colnames(
        variance_partition
      ),
      fixed = TRUE
    )

  mean_variance <-
    sort(
      colMeans(
        variance_partition,
        na.rm = TRUE
      ),
      decreasing = TRUE
    )

  plot_data <-
    tibble::tibble(
      variable =
        names(
          mean_variance
        ),
      mean_fraction =
        as.numeric(
          mean_variance
        )
    ) %>%
    dplyr::mutate(
      variable_label =
        dplyr::recode(
          variable,
          !!!clinical_variable_labels,
          .default =
            variable
        ),
      variable_label =
        factor(
          variable_label,
          levels =
            rev(
              variable_label
            )
        )
    )

  list(
    variance_partition =
      variance_partition,
    plot_data =
      plot_data,
    metadata =
      analysis_metadata,
    number_features =
      nrow(
        analysis_matrix
      )
  )
}


plot_proteomics_variance_partition <- function(
    variance_result,
    title_text
) {
  ggplot(
    variance_result$plot_data,
    aes(
      x = variable_label,
      y = mean_fraction
    )
  ) +
    geom_col(
      fill = "grey35",
      width = 0.75
    ) +
    coord_flip() +
    scale_y_continuous(
      labels =
        scales::percent_format(
          accuracy = 1
        ),
      expand =
        expansion(
          mult = c(
            0,
            0.05
          )
        )
    ) +
    labs(
      title =
        title_text,
      x = NULL,
      y =
        "Mean fraction of variance explained"
    ) +
    theme_classic(
      base_size = 12
    ) +
    theme(
      plot.title =
        element_text(
          face = "bold",
          hjust = 0.5
        ),
      axis.text =
        element_text(
          colour = "black"
        )
    )
}
```

```{r Olink variance partition, fig.height=3.5, fig.width=5.5}
olink_variance_partition <-
  run_proteomics_variance_partition(
    expression_matrix =
      qc_after$olink,
    metadata =
      qc_after$olink_metadata,
    candidate_variables = c(
      "Age_years",
      "Sample_age_days",
      "BCA_mg_ml",
      "mPAP_mmHg",
      "RVIT_mm",
      "TAPSE_mm",
      "LVEF_percent",
      "LVEDD_mm"
    )
  )

olink_variance_partition_plot <-
  plot_proteomics_variance_partition(
    variance_result =
      olink_variance_partition,
    title_text =
      "Olink"
  )

olink_variance_partition_plot

write.csv(
  olink_variance_partition$
    plot_data,
  file.path(
    output_dir,
    "Olink_variance_partition.csv"
  ),
  row.names = FALSE
)
```

```{r LCMS variance partition, fig.height=3.5, fig.width=5.5}
ms_variance_partition <-
  run_proteomics_variance_partition(
    expression_matrix =
      qc_after$ms,
    metadata =
      qc_after$ms_metadata,
    candidate_variables = c(
      "Age_years",
      "Sample_age_days",
      "mPAP_mmHg",
      "RVIT_mm",
      "TAPSE_mm",
      "LVEF_percent",
      "LVEDD_mm"
    )
  )

ms_variance_partition_plot <-
  plot_proteomics_variance_partition(
    variance_result =
      ms_variance_partition,
    title_text =
      "LC-MS/MS"
  )

ms_variance_partition_plot

write.csv(
  ms_variance_partition$
    plot_data,
  file.path(
    output_dir,
    "LCMS_variance_partition.csv"
  ),
  row.names = FALSE
)
```

# LC-MS/MS missingness and abundance

```{r LCMS missingness abundance, fig.height=3.5, fig.width=5.2}
ms_missingness_abundance <-
  tibble::tibble(
    Protein =
      rownames(
        qc_after$ms
      ),
    Missing_fraction =
      matrixStats::rowMeans2(
        !is.finite(
          qc_after$ms
        )
      ),
    Median_log2_abundance =
      matrixStats::rowMedians(
        qc_after$ms,
        na.rm = TRUE
      )
  ) %>%
  dplyr::filter(
    is.finite(
      Missing_fraction
    ),
    is.finite(
      Median_log2_abundance
    )
  )

ms_missingness_abundance_plot <-
  ggplot(
    ms_missingness_abundance,
    aes(
      x =
        Median_log2_abundance,
      y =
        Missing_fraction
    )
  ) +
  geom_point(
    size = 1.3,
    alpha = 0.65
  ) +
  scale_y_continuous(
    labels =
      scales::percent_format(
        accuracy = 1
      ),
    limits = c(
      0,
      1
    )
  ) +
  labs(
    x =
      "Median log2 abundance",
    y =
      "Fraction of samples missing"
  ) +
  theme_classic(
    base_size = 12
  ) +
  theme(
    axis.text =
      element_text(
        colour = "black"
      )
  )

ms_missingness_abundance_plot

write.csv(
  ms_missingness_abundance,
  file.path(
    output_dir,
    "LCMS_missingness_abundance.csv"
  ),
  row.names = FALSE
)
```

# Global abundance structure

```{r LCMS all protein heatmap, fig.height=6, fig.width=8}
ms_heatmap_matrix <-
  filter_usable_features(
    qc_after$ms
  )

ms_heatmap_matrix <-
  impute_rows_low_value(
    ms_heatmap_matrix
  )

ms_heatmap_z <-
  t(
    scale(
      t(
        ms_heatmap_matrix
      )
    )
  )

ms_heatmap_z[
  !is.finite(
    ms_heatmap_z
  )
] <- 0


ms_heatmap_metadata <-
  qc_after$clinical %>%
  dplyr::select(
    Sample,
    Age_years,
    Sex
  ) %>%
  dplyr::filter(
    Sample %in%
      colnames(
        ms_heatmap_z
      )
  ) %>%
  dplyr::slice(
    match(
      colnames(
        ms_heatmap_z
      ),
      Sample
    )
  ) %>%
  dplyr::mutate(
    Age_years =
      as.numeric(
        Age_years
      ),
    Sex =
      factor(
        as.character(
          Sex
        )
      )
  )


stopifnot(
  all(
    ms_heatmap_metadata$
      Sample ==
      colnames(
        ms_heatmap_z
      )
  )
)


ms_age_color <-
  safe_continuous_colour(
    ms_heatmap_metadata$
      Age_years,
    colours = c(
      "grey95",
      "grey60",
      "grey10"
    )
  )


ms_sex_levels <-
  levels(
    droplevels(
      ms_heatmap_metadata$
        Sex
    )
  )


ms_sex_colors <-
  stats::setNames(
    c(
      "#009E73",
      "#E69F00",
      "#7B61A8",
      "#56B4E9"
    )[
      seq_along(
        ms_sex_levels
      )
    ],
    ms_sex_levels
  )


ms_heatmap_annotation <-
  ComplexHeatmap::HeatmapAnnotation(
    Age =
      ms_heatmap_metadata$
        Age_years,
    Sex =
      ms_heatmap_metadata$
        Sex,
    col = list(
      Age =
        ms_age_color,
      Sex =
        ms_sex_colors
    ),
    na_col = "white",
    annotation_name_gp =
      grid::gpar(
        fontsize = 10,
        fontface = "bold",
        col = "black"
      )
  )


ms_all_protein_heatmap <-
  ComplexHeatmap::Heatmap(
    ms_heatmap_z,
    name = "Z-score",
    col =
      protein_heatmap_colors,
    top_annotation =
      ms_heatmap_annotation,
    show_row_names = FALSE,
    show_column_names = TRUE,
    column_names_gp =
      grid::gpar(
        fontsize = 7,
        col = "black"
      ),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    use_raster =
      nrow(
        ms_heatmap_z
      ) > 2000,
    raster_quality = 2,
    column_title =
      "LC-MS/MS",
    column_title_gp =
      grid::gpar(
        fontsize = 12,
        fontface = "bold"
      ),
    heatmap_legend_param = list(
      title = "Z-score",
      at = c(
        -2,
        0,
        2
      )
    )
  )


ComplexHeatmap::draw(
  ms_all_protein_heatmap,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)



```

```{r Olink all protein heatmap, fig.height=6, fig.width=8.5}
olink_heatmap_matrix <-
  filter_usable_features(
    qc_after$olink
  )

olink_heatmap_matrix <-
  impute_rows_median(
    olink_heatmap_matrix
  )

olink_heatmap_z <-
  t(
    scale(
      t(
        olink_heatmap_matrix
      )
    )
  )

olink_heatmap_z[
  !is.finite(
    olink_heatmap_z
  )
] <- 0


olink_heatmap_metadata <-
  qc_after$olink_metadata %>%
  dplyr::select(
    Sample,
    Batch,
    BCA_mg_ml
  ) %>%
  dplyr::left_join(
    qc_after$clinical %>%
      dplyr::select(
        Sample,
        Age_years,
        Sex
      ),
    by = "Sample"
  ) %>%
  dplyr::filter(
    Sample %in%
      colnames(
        olink_heatmap_z
      )
  ) %>%
  dplyr::slice(
    match(
      colnames(
        olink_heatmap_z
      ),
      Sample
    )
  ) %>%
  dplyr::mutate(
    Age_years =
      as.numeric(
        Age_years
      ),
    Sex =
      factor(
        as.character(
          Sex
        )
      ),
    Batch =
      factor(
        as.character(
          Batch
        )
      ),
    BCA_mg_ml =
      as.numeric(
        BCA_mg_ml
      )
  )


stopifnot(
  all(
    olink_heatmap_metadata$
      Sample ==
      colnames(
        olink_heatmap_z
      )
  )
)


olink_age_color <-
  safe_continuous_colour(
    olink_heatmap_metadata$
      Age_years,
    colours = c(
      "grey95",
      "grey60",
      "grey10"
    )
  )


olink_bca_color <-
  safe_continuous_colour(
    olink_heatmap_metadata$
      BCA_mg_ml,
    colours = c(
      "grey95",
      "grey60",
      "grey10"
    )
  )


olink_sex_levels <-
  levels(
    droplevels(
      olink_heatmap_metadata$
        Sex
    )
  )


olink_sex_colors <-
  stats::setNames(
    c(
      "#009E73",
      "#E69F00",
      "#7B61A8",
      "#56B4E9"
    )[
      seq_along(
        olink_sex_levels
      )
    ],
    olink_sex_levels
  )


olink_batch_levels <-
  levels(
    droplevels(
      olink_heatmap_metadata$
        Batch
    )
  )


olink_batch_colors <-
  stats::setNames(
    c(
      "#56B4E9",
      "#CC79A7",
      "#009E73",
      "#E69F00"
    )[
      seq_along(
        olink_batch_levels
      )
    ],
    olink_batch_levels
  )


olink_heatmap_annotation <-
  ComplexHeatmap::HeatmapAnnotation(
    Age =
      olink_heatmap_metadata$
        Age_years,
    Sex =
      olink_heatmap_metadata$
        Sex,
    Batch =
      olink_heatmap_metadata$
        Batch,
    BCA =
      olink_heatmap_metadata$
        BCA_mg_ml,
    col = list(
      Age =
        olink_age_color,
      Sex =
        olink_sex_colors,
      Batch =
        olink_batch_colors,
      BCA =
        olink_bca_color
    ),
    na_col = "white",
    annotation_name_gp =
      grid::gpar(
        fontsize = 10,
        fontface = "bold",
        col = "black"
      )
  )


olink_all_protein_heatmap <-
  ComplexHeatmap::Heatmap(
    olink_heatmap_z,
    name = "Z-score",
    col =
      protein_heatmap_colors,
    top_annotation =
      olink_heatmap_annotation,
    show_row_names = FALSE,
    show_column_names = TRUE,
    column_names_gp =
      grid::gpar(
        fontsize = 7,
        col = "black"
      ),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    use_raster =
      nrow(
        olink_heatmap_z
      ) > 2000,
    raster_quality = 2,
    column_title =
      "Olink",
    column_title_gp =
      grid::gpar(
        fontsize = 12,
        fontface = "bold"
      ),
    heatmap_legend_param = list(
      title = "Z-score",
      at = c(
        -2,
        0,
        2
      )
    )
  )


ComplexHeatmap::draw(
  olink_all_protein_heatmap,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)


```

# Supplementary Data export

```{r Supplementary_data_export, eval=TRUE}
qc_suppdata <- list(
  Sample_exclusions =
    sample_exclusions,

  Availability_comparison =
    availability_comparison,

  Availability_after =
    qc_after$availability,

  Olink_sample_metrics =
    metrics_after$olink_metrics,

  LCMS_sample_metrics =
    metrics_after$ms_metrics,

  Clinical_correlations =
    clinical_correlation_long,

  Olink_variance =
    olink_variance_partition$plot_data,

  LCMS_variance =
    ms_variance_partition$plot_data,

  LCMS_missingness =
    ms_missingness_abundance
)

writexl::write_xlsx(
  qc_suppdata,
  file.path(
    output_dir,
    "Proteomics_quality_control_SuppData.xlsx"
  )
)
```
