---
title: "QC, integration, annotation all DCM samples"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "QC_integration_annotation"

# Load libraries
library(here)
library(readr)
library(readxl)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(clustree)
library(harmony)
library(speckle)
library(qs2)

#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```


## Quality Control

Load and merge cellbender and cellranger output

```{r load data, echo=T, results= "hide"}
cell_bender_merged <- Read_CellBender_h5_Multi_File(merge = TRUE,
                                                    data_dir = here::here("data", "Cellbender_output"),
                                                    custom_name = "_filtered.h5",
                                                    sample_list = c("cellbender_output_319801_1-337PK", 
                                                                    "cellbender_output_319801_2-357PK", 
                                                                    "cellbender_output_319801_3-345PK", 
                                                                    "cellbender_output_323041_1-369PK", 
                                                                    "cellbender_output_323041_2-375PK", 
                                                                    "cellbender_output_323041_3-385PK",
                                                                    "cellbender_output_327451_1-401PK", 
                                                                    "cellbender_output_327451_2-403PK", 
                                                                    "cellbender_output_327451_3-415PK", 
                                                                    "cellbender_output_330241_1-437PK", 
                                                                    "cellbender_output_330241_2-439PK", 
                                                                    "cellbender_output_330241_3-461PK", 
                                                                    "cellbender_output_330241_4-441PK_reprocessed", 
                                                                    "cellbender_output_330241_5-463PK", 
                                                                    "cellbender_output_351071_1-213PK", 
                                                                    "cellbender_output_351071_2-229PK", 
                                                                    "cellbender_output_351071_3-P3", 
                                                                    "cellbender_output_351071_4-P8", 
                                                                    "cellbender_output_352751_1-153PK", 
                                                                    "cellbender_output_352751_2-211PK", 
                                                                    "cellbender_output_352751_3-323PK", 
                                                                    "cellbender_output_352751_4-P5"
                                                                    ),
                                                    sample_names = c("PK337", 
                                                                     "PK357", 
                                                                     "PK345", 
                                                                     "PK369", 
                                                                     "PK375", 
                                                                     "PK385", 
                                                                     "PK401", 
                                                                     "PK403", 
                                                                     "PK415", 
                                                                     "PK437", 
                                                                     "PK439", 
                                                                     "PK461", 
                                                                     "PK441", 
                                                                     "PK463", 
                                                                     "PK213", 
                                                                     "PK229", 
                                                                     "P3", 
                                                                     "P8", 
                                                                     "PK153", 
                                                                     "PK211", 
                                                                     "PK323", 
                                                                     "P5")
                                                    )

cell_ranger_merged <- Read10X_h5_GEO(merge = TRUE,
                                     data_dir = here::here("data", "Cellranger_output"), 
                                     shared_suffix = "_filtered_feature_bc_matrix.h5",
                                     sample_list = c("319801_1-337PK", 
                                                     "319801_2-357PK", 
                                                     "319801_3-345PK", 
                                                     "323041_1-369PK", 
                                                     "323041_2-375PK", 
                                                     "323041_3-385PK", 
                                                     "327451_1-401PK", 
                                                     "327451_2-403PK", 
                                                     "327451_3-415PK", 
                                                     "330241_1-437PK", 
                                                     "330241_2-439PK", 
                                                     "330241_3-461PK", 
                                                     "330241_4-441PK_reprocessed", 
                                                     "330241_5-463PK", 
                                                     "351071_1-213PK", 
                                                     "351071_2-229PK", 
                                                     "351071_3-P3", 
                                                     "351071_4-P8", 
                                                     "352751_1-153PK", 
                                                     "352751_2-211PK", 
                                                     "352751_3-323PK", 
                                                     "352751_4-P5"),
                                     sample_names = c("PK337", 
                                                      "PK357", 
                                                      "PK345", 
                                                      "PK369", 
                                                      "PK375", 
                                                      "PK385", 
                                                      "PK401", 
                                                      "PK403", 
                                                      "PK415", 
                                                      "PK437", 
                                                      "PK439", 
                                                      "PK461", 
                                                      "PK441", 
                                                      "PK463", 
                                                      "PK213", 
                                                      "PK229", 
                                                      "P3", 
                                                      "P8", 
                                                      "PK153", 
                                                      "PK211", 
                                                      "PK323", 
                                                      "P5")
                                     )

DCM_RV_preQC <- Create_CellBender_Merged_Seurat(raw_cell_bender_matrix = cell_bender_merged, 
                                             raw_counts_matrix = cell_ranger_merged, 
                                             raw_assay_name = "RAW", 
                                             cell_bender_assay_name = "RNA")
rm(cell_bender_merged, cell_ranger_merged)
```

Add metadata: percentage of mitochondrial and ribosomal genes, clinical data

```{r add metadata}
DCM_RV_preQC[["percent.mt"]] <- PercentageFeatureSet(DCM_RV_preQC, pattern = "^MT-")
DCM_RV_preQC[["percent.rp"]] <- PercentageFeatureSet(DCM_RV_preQC, pattern = "^RP[SL]")
DCM_RV_preQC$Sample <- DCM_RV_preQC$orig.ident
Clinical_data <- read_excel(here::here("data", "DCM_Clinical_data_22.xlsx"))
add_metadata <- left_join(DCM_RV_preQC[["Sample"]], Clinical_data)
row.names(add_metadata) <- row.names(DCM_RV_preQC[[]])
DCM_RV_preQC <- AddMetaData(DCM_RV_preQC, metadata = add_metadata)
rm(add_metadata, Clinical_data)
metadata <- unique(subset(DCM_RV_preQC@meta.data, select = c(orig.ident, Sample:length(DCM_RV_preQC@meta.data))))
rownames(metadata) <- metadata$orig.ident
metadata
```

Check how much Cellbender has affected the dataset

```{r Cellbender, fig.height=4, fig.width=7}
table(DCM_RV_preQC$Sample)

DCM_RV_preQC <- Add_CellBender_Diff(seurat_object = DCM_RV_preQC, 
                                 raw_assay_name = "RAW", 
                                 cell_bender_assay_name = "RNA")

median_stats <- Median_Stats(seurat_object = DCM_RV_preQC, 
                             group_by_var = "orig.ident", 
                             median_var = c("nCount_Diff", "nFeature_Diff"))
median_stats

feature_diff <- CellBender_Feature_Diff(seurat_object = DCM_RV_preQC, 
                                        raw_assay = "RAW", 
                                        cell_bender_assay = "RNA")
head(feature_diff)

CellBender_Diff_Plot(feature_diff_df = feature_diff)
```

Check quality control parameters

```{r QC_pre_VlnPlots, fig.height=6, fig.width=15}
p1 <- VlnPlot(DCM_RV_preQC, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(DCM_RV_preQC, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(DCM_RV_preQC, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(DCM_RV_preQC, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_VlnPlots
```

```{r QC_pre_VlnPlots_zoom, fig.height=10, fig.width=15}
p1 <- VlnPlot(DCM_RV_preQC, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 20000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(DCM_RV_preQC, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(DCM_RV_preQC, features = "percent.mt", group.by = "Sample", pt.size = 0, y.max = 5) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(DCM_RV_preQC, features = "percent.rp", group.by = "Sample", pt.size = 0, y.max = 5) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_VlnPlots_zoom <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_VlnPlots_zoom
```

Doublets detection

```{r Doublets detection}
DCM_RV_preQC_sce <- as.SingleCellExperiment(DCM_RV_preQC)
DCM_RV_preQC_sce <- scDblFinder(DCM_RV_preQC_sce, samples="Sample", clusters = TRUE)
table(DCM_RV_preQC_sce@colData$scDblFinder.class)
DCM_RV_preQC <- as.Seurat(DCM_RV_preQC_sce, counts = "counts", data = "logcounts")
rm(DCM_RV_preQC_sce)
DCM_RV_preQC@meta.data[DCM_RV_preQC@meta.data$scDblFinder.class %in% "singlet", "scDblFinder.n"] = paste0("Singlets (n=", table(DCM_RV_preQC$scDblFinder.class)[1], ")")
DCM_RV_preQC@meta.data[DCM_RV_preQC@meta.data$scDblFinder.class %in% "doublet", "scDblFinder.n"] = paste0("Doublets (n=", table(DCM_RV_preQC$scDblFinder.class)[2], ")")
DCM_RV_preQC$scDblFinder.n <- factor(x = DCM_RV_preQC$scDblFinder.n, levels = c(rownames(table(DCM_RV_preQC$scDblFinder.n))[2], rownames(table(DCM_RV_preQC$scDblFinder.n))[1])) 
```

```{r QC_pre_Doublets, fig.height=3, fig.width=15}
p1 <- VlnPlot(DCM_RV_preQC, features = c("nCount_RNA"), split.by = "scDblFinder.n", group.by = "Sample", pt.size = 0, y.max = 50000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(DCM_RV_preQC, features = c("nFeature_RNA"), split.by = "scDblFinder.n", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) 

QC_pre_Doublets <- p1 + p2 + plot_layout(ncol = 2)

QC_pre_Doublets
```

```{r save DCM_RV_preQC, eval=FALSE}
qs_save(DCM_RV_preQC
        file = here::here(output_dir_data, "DCM_RV_preQC.qs2"),
        nthreads = 32
        )
```

```{r read DCM_RV_preQC, eval=TRUE}
DCM_RV_preQC <- qs_read(file = here::here(output_dir_data, "DCM_RV_preQC.qs2"),
        nthreads = 32
        )
```

Remove doublets

```{r Remove doublets}
DCM_RV <- subset(x = DCM_RV_preQC, subset = scDblFinder.class == "singlet")
table(DCM_RV@meta.data$scDblFinder.class)
rm(DCM_RV_preQC)
DCM_RV@assays[["RAW"]] <- NULL
DefaultAssay(DCM_RV) <- "RNA"
```

Check quality control parameters after removing doublets

```{r QC_pre_no_doublets_VlnPlots, fig.height=6, fig.width=15}
p1 <- VlnPlot(DCM_RV, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(DCM_RV, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(DCM_RV, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(DCM_RV, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_no_doublets_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_no_doublets_VlnPlots
```

```{r QC_pre_no_doublets_VlnPlots_zoom, fig.height=10, fig.width=15}
p1 <- VlnPlot(DCM_RV, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 20000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(DCM_RV, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(DCM_RV, features = "percent.mt", group.by = "Sample", pt.size = 0, y.max = 1.5) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(DCM_RV, features = "percent.rp", group.by = "Sample", pt.size = 0, y.max = 1.5) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_no_doublets_VlnPlots_zoom <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_no_doublets_VlnPlots_zoom
```

Filter based on number of counts, features and percentage of mitochondrial genes

```{r filter}
DCM_RV <- subset(DCM_RV, subset = 
                   nFeature_RNA > 500 &
                   percent.mt < 1.5 &
                   nCount_RNA > 1000 &
                   nCount_RNA < 100000)
table(DCM_RV$Sample)
```

Sample PK211 has a really low number of counts and features, too different compared to all other samples. Therefore it will be excluded from the analysis

```{r remove PK211}
DCM_RV <- subset(DCM_RV, subset = 
                   Sample != "PK211")

table(DCM_RV$Sample)
```

```{r cols}
cell_type_cols <- c("#d05050", "#33A02C", "#1F78B4", "#CAB2D6", "#FF7F00", "#B2DF8A", "#B15928", "#FFD000", "#FA95C0", "#FDBF6F", "#6A3D9A", "#7aCDF9", "#E6E600")  
sample_cols <- DiscretePalette_scCustomize(num_colors = 29, palette = "polychrome", shuffle_pal = TRUE, seed = 220510)
batch_cols <- DiscretePalette_scCustomize(num_colors = 9, palette = "polychrome", shuffle_pal = TRUE, seed = 220510)
cluster_cols <- DiscretePalette_scCustomize(num_colors = 15, palette = "polychrome", shuffle_pal = TRUE, seed = 220510)
```

```{r plot_cols, fig.height=5, fig.width=10}
PalettePlot(cell_type_cols) + 
  PalettePlot(sample_cols) + 
  PalettePlot(batch_cols) + 
  PalettePlot(cluster_cols)
```

Check quality control parameters

```{r QC_post_VlnPlots, fig.height=6, fig.width=15}
p1 <- VlnPlot(DCM_RV, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(DCM_RV, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(DCM_RV, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(DCM_RV, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_post_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_post_VlnPlots
```

```{r QC_post_VlnPlots_zoom, fig.height=10, fig.width=15}
p1 <- VlnPlot(DCM_RV, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 20000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(DCM_RV, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(DCM_RV, features = "percent.mt", group.by = "Sample", pt.size = 0, y.max = 1.5) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(DCM_RV, features = "percent.rp", group.by = "Sample", pt.size = 0, y.max = 1.5) + theme(axis.title.x = element_blank()) + NoLegend()

QC_post_VlnPlots_zoom <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_post_VlnPlots_zoom
```


## Integration and clustering

Normalization and scaling using SCTransform

```{r SCTransform, message=FALSE, warning=FALSE, echo = TRUE, results= "hide"}
DefaultAssay(DCM_RV) <- "RNA"
DCM_RV <- SplitObject(DCM_RV, split.by = "Sample")
for (i in 1:length(DCM_RV)) {
  DCM_RV[[i]] <- SCTransform(DCM_RV[[i]], 
                             vst.flavor = "v2", 
                             vars.to.regress = c("percent.mt"), 
                             return.only.var.genes = TRUE,
                             method = "glmGamPoi")
}
```

PCA

```{r PCA}
features <- SelectIntegrationFeatures(object.list = DCM_RV, nfeatures = 3000)
DCM_RV <- merge(x = DCM_RV[[1]], y = DCM_RV[2:length(DCM_RV)], merge.data=TRUE)
VariableFeatures(DCM_RV) <- features
DCM_RV <- RunPCA(DCM_RV)
```

```{r Elbow_not_integrated, fig.width=4, fig.height=4}
ElbowPlot(DCM_RV, ndims = 50)
```

Clustering without integration

```{r clustering not integrated}
DCM_RV <- RunUMAP(DCM_RV, dims = 1:50)
DCM_RV <- FindNeighbors(DCM_RV, dims = 1:50)
DCM_RV <- FindClusters(DCM_RV, resolution = seq(0.1, 0.8, by=0.1))
```

```{r Clustree_not_integrated, fig.width=6, fig.height=9}
clustree::clustree(DCM_RV@meta.data[,grep("SCT_snn_res", colnames(DCM_RV@meta.data))],
                   prefix = "SCT_snn_res.")
```

```{r UMAPs_not_integrated, fig.width=12, fig.height=10}
DimPlot(DCM_RV, reduction = "umap", shuffle = T,
        group.by = c("SCT_snn_res.0.1", "Sample", "Batch_processing", "Batch_sequencing"), ncol = 2)
```

Integrate the samples

```{r RunHarmony}
DCM_RV <- RunHarmony(DCM_RV, 
                     assay.use="SCT", 
                     group.by.vars = "Sample", 
                     dims.use = 1:50, 
                     plot_convergence=TRUE, 
                     .options = harmony_options(
                       max.iter.cluster = 100,
                       )
                     )
```

Clustering after integration

```{r clustering integrated}
DefaultAssay(DCM_RV) <- "SCT"
DCM_RV <- RunUMAP(DCM_RV, dims = 1:50, reduction = "harmony")
DCM_RV <- FindNeighbors(DCM_RV, dims = 1:50, reduction = "harmony")
DCM_RV <- FindClusters(DCM_RV, resolution = seq(0.1, 0.8, by=0.1))
```

```{r Clustree_integrated_low_quality, fig.width=6, fig.height=9}
clustree::clustree(DCM_RV@meta.data[,grep("SCT_snn_res", colnames(DCM_RV@meta.data))],
                   prefix = "SCT_snn_res.")
```

```{r UMAPs_integrated_low_quality, fig.width=12, fig.height=10}
DCM_RV$SCT_snn_res.0.2 <- factor(DCM_RV$SCT_snn_res.0.2, levels = c("0", "1", "2", "3", "4", "5","6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16"))

DimPlot(DCM_RV, reduction = "umap", shuffle = T,
        group.by = c("SCT_snn_res.0.2", "Sample", "Batch_processing", "Batch_sequencing"), ncol = 2, label=T)
```

Find the markers that characterize each cell population

```{r FindAllMarkers}
DefaultAssay(DCM_RV) <- "RNA"
DCM_RV <- NormalizeData(DCM_RV)
Idents(DCM_RV) <- "SCT_snn_res.0.2"
Markers <- FindAllMarkers(DCM_RV, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 1)
write.csv(Markers, here::here(output_dir_data, "DCM_RV_Markers_all_low_quality.csv"))
Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
write.csv(Markers_top10, here::here(output_dir_data, "DCM_RV_Markers_top10_low_quality.csv"))
Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
write.csv(Markers_top3, here::here(output_dir_data, "DCM_RV_Markers_top3_low_quality.csv"))
```

```{r Heatmap_low_quality, fig.width=15, fig.height=10}
heatmap_cols <- colorRampPalette(RColorBrewer::brewer.pal(9,"RdBu"))(256)
heatmap_cols <- rev(heatmap_cols[1:256])
Heatmap <- DoHeatmap(subset(DCM_RV, cells = sample(colnames(DCM_RV), 20000)), label = FALSE, draw.line = F, features = Markers_top10$gene, assay = "SCT") +
  scale_fill_gradientn(colours = heatmap_cols) +
  theme(axis.text.y = element_text(size = 5)) +
  theme(plot.margin = unit(c(0.1, 0, 0, 0), 
                           "inches"))
Heatmap
```

Check QC parameters in each cluster

```{r QC_VlnPlots_low_quality, fig.width=12, fig.height=6}
VlnPlot(DCM_RV, features = c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.rp", "scDblFinder.score"), ncol = 3, pt.size = 0) + theme(axis.title.x = element_blank())
```

```{r QC_VlnPlots_low_quality_2, fig.width=20, fig.height=6}
VlnPlot_scCustom(DCM_RV, features = dplyr::filter(Markers_top10, cluster == "10")$gene, num_columns = 5, pt.size = 0) + theme(axis.title.x = element_blank())
```

```{r QC_VlnPlots_low_quality_3, fig.width=20, fig.height=6}
VlnPlot_scCustom(DCM_RV, features = dplyr::filter(Markers_top10, cluster == "14")$gene, num_columns = 5, pt.size = 0) + theme(axis.title.x = element_blank())
```

```{r QC_VlnPlots_low_quality_4, fig.width=20, fig.height=6}
VlnPlot_scCustom(DCM_RV, features = dplyr::filter(Markers_top10, cluster == "16")$gene, num_columns = 5, pt.size = 0) + theme(axis.title.x = element_blank())
```

Remove low quality cells (clusters with low number of count and features and higher percentage of mt or rp genes)

```{r remove low quality, eval=FALSE}
DCM_RV <- subset(DCM_RV, subset = SCT_snn_res.0.2 %in% c("0", "1", "2", "3", "4", "5","6", "7", "8", "9", "11", "12", "13", "15"))
```

```{r save DCM_RV, eval=FALSE}
qs_save(DCM_RV,
        file = here::here(output_dir_data, "DCM_RV.qs2"),
        nthreads = 32
        )
```

```{r read DCM_RV}
DCM_RV <- qs_read(file = here::here(output_dir_data, "DCM_RV.qs2"),
        nthreads = 32
        )
```

Reintegrate and recluster

```{r Recluster, eval=FALSE, message=FALSE}
DefaultAssay(DCM_RV) <- "SCT"
DCM_RV <- SplitObject(DCM_RV, split.by = "Sample")
features <- SelectIntegrationFeatures(object.list = DCM_RV, nfeatures = 3000)
DCM_RV <- merge(x = DCM_RV[[1]], y = DCM_RV[2:length(DCM_RV)], merge.data=TRUE)
VariableFeatures(DCM_RV) <- features
DCM_RV <- RunPCA(DCM_RV, npcs = 50)
ElbowPlot(DCM_RV, ndims = 50)

DCM_RV <- RunHarmony(DCM_RV, 
                 assay.use="SCT", 
                 group.by.vars = "Sample", 
                 dims.use = 1:50, 
                 plot_convergence=TRUE, 
                 .options = harmony_options(
                   max.iter.cluster = 100,
                   )
                 )

DefaultAssay(DCM_RV) <- "SCT"
DCM_RV <- RunUMAP(DCM_RV, dims = 1:50, reduction = "harmony")
DCM_RV <- FindNeighbors(DCM_RV, dims = 1:50, reduction = "harmony")
DCM_RV <- FindClusters(DCM_RV, resolution = seq(0.1, 0.8, by=0.1))
```

```{r Clustree_integrated, fig.width=6, fig.height=9}
clustree::clustree(DCM_RV@meta.data[,grep("SCT_snn_res", colnames(DCM_RV@meta.data))],
                   prefix = "SCT_snn_res.")
```

```{r Clustree_Manuscript, fig.width=7, fig.height=8}
res_names <- sub("SCT_snn_res\\.", "", grep("SCT_snn_res", names(DCM_RV@meta.data), value = TRUE))

clustree_cols <- c("#8F7C00FF", "#5EF1F2FF", "#FFA8BBFF", "#F0A0FFFF", "#FFCC99FF", "#2BCE48FF", "#FFE100FF", "#9DCC00FF")

names(clustree_cols) <- res_names[seq_along(clustree_cols)]      

clustree(
  DCM_RV@meta.data[, grep("SCT_snn_res", names(DCM_RV@meta.data))],
  prefix = "SCT_snn_res."
) +
  scale_colour_manual(values = clustree_cols, name = "resolution") +  
  scale_fill_manual(values = clustree_cols,  name = "resolution")    
```

```{r UMAPs_integrated, fig.width=12, fig.height=10}
DCM_RV$SCT_snn_res.0.2 <- factor(DCM_RV$SCT_snn_res.0.2, levels = c("0", "1", "2", "3", "4", "5","6", "7", "8", "9", "10", "11", "12", "13"))
DimPlot(DCM_RV, reduction = "umap", shuffle = T,
        group.by = c("SCT_snn_res.0.2", "Sample", "Batch_processing", "Batch_sequencing"), ncol = 2)
```

Find the markers that characterize each cell population

```{r FindAllMarkers 2}
DefaultAssay(DCM_RV) <- "RNA"
Idents(DCM_RV) <- "SCT_snn_res.0.2"
Markers <- FindAllMarkers(DCM_RV, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 1)
write.csv(Markers, here::here(output_dir_data, "DCM_RV_Markers_all.csv"))
Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
write.csv(Markers_top10, here::here(output_dir_data, "DCM_RV_Markers_top10.csv"))
Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
write.csv(Markers_top3, here::here(output_dir_data, "DCM_RV_Markers_top3.csv"))
```

Check cell cycle genes

```{r Cell_cycle_UMAP, fig.width=5, fig.height=4}
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
DCM_RV <- CellCycleScoring(DCM_RV, s.features = s.genes, g2m.features = g2m.genes)
DimPlot(DCM_RV, reduction = "umap", shuffle = T,
        group.by = "Phase")
```

## Cell type annotation

Annotate the clusters based on the characteristic markers

```{r annotation}
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("0"), "cell_type"] = "Cardiomyocytes"
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("1"), "cell_type"] = "Fibroblasts"
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("2", "5"), "cell_type"] = "Endothelial cells"
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("3"), "cell_type"] = "Pericytes"
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("4"), "cell_type"] = "Macrophages"
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("6"), "cell_type"] = "T lymphocytes"
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("7"), "cell_type"] = "Smooth muscle cells"
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("8"), "cell_type"] = "Neuronal cells"
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("9"), "cell_type"] = "Endocardial cells"
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("10"), "cell_type"] = "Adipocytes"
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("11"), "cell_type"] = "Mast cells"
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("12"), "cell_type"] = "Lymphatic cells"
DCM_RV@meta.data[DCM_RV@meta.data$SCT_snn_res.0.2 %in% c("13"), "cell_type"] = "B lymphocytes"

DCM_RV$cell_type <- factor(DCM_RV$cell_type, levels = c("Cardiomyocytes", 
                                                        "Fibroblasts",
                                                        "Endothelial cells",
                                                        "Pericytes",
                                                        "Macrophages",
                                                        "T lymphocytes",
                                                        "Smooth muscle cells",
                                                        "Neuronal cells",
                                                        "Endocardial cells",
                                                        "Adipocytes",
                                                        "Mast cells",
                                                        "Lymphatic cells",
                                                        "B lymphocytes"))
Idents(DCM_RV) <- DCM_RV$cell_type
```

```{r QC_sample_Manuscript, fig.width=14, fig.height=6}
VlnPlot_scCustom(DCM_RV, features=c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.rp"), group.by="Sample", num_columns=2, pt.size = 0, colors_use = sample_cols) & theme(axis.title.x = element_blank())
```

```{r QC_batch_Manuscript, fig.width=14, fig.height=6}
VlnPlot_scCustom(DCM_RV, features=c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.rp"), group.by="Batch_processing", num_columns=2, pt.size = 0, colors_use = batch_cols) & theme(axis.title.x = element_blank())
```

```{r QC_cell_type_Manuscript, fig.width=14, fig.height=7}
VlnPlot_scCustom(DCM_RV, features=c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.rp"), group.by="cell_type", num_columns=2, pt.size = 0, colors_use = cell_type_cols) & theme(axis.title.x = element_blank())
```

```{r UMAP_cluster_Manuscript, fig.height=5, fig.width=5}
DimPlot_scCustom(DCM_RV, group.by = "SCT_snn_res.0.2", colors_use = cluster_cols)
```

```{r UMAP_sample_Manuscript, fig.height=5, fig.width=6.2}
DimPlot_scCustom(DCM_RV, group.by = "Sample", colors_use = sample_cols)
```

```{r UMAP_batch_Manuscript, fig.height=5, fig.width=5.3}
DimPlot_scCustom(DCM_RV, group.by = "Batch_processing", colors_use = batch_cols)
```

```{r FindAllMarkers cell_type}
DefaultAssay(DCM_RV) <- "RNA"
Idents(DCM_RV) <- "cell_type"
Markers <- FindAllMarkers(DCM_RV, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 1)
write.csv(Markers, here::here(output_dir_data, "DCM_RV_Markers_all_cell_type.csv"))
Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
write.csv(Markers_top10, here::here(output_dir_data, "DCM_RV_Markers_top10_cell_type.csv"))
Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
write.csv(Markers_top3, here::here(output_dir_data, "DCM_RV_Markers_top3_cell_type.csv"))
```

```{r Heatmap_Manuscript, fig.width=12, fig.height=10}
genes.use <- intersect(
  Markers_top10$gene,
  rownames(DCM_RV[["SCT"]]@scale.data)
)
DefaultAssay(DCM_RV) <- "SCT"
Heatmap <- DoHeatmap(subset(DCM_RV, cells = sample(colnames(DCM_RV), 10000)),
                     features = genes.use,
                     label = FALSE,
                     draw.line = TRUE,
                     group.by = "cell_type",
                     group.colors = cell_type_cols,
                     lines.width = 1) +
  scale_fill_gradientn(colours = heatmap_cols) + 
  theme(text = element_text(size = 15), 
        axis.text.y = element_text(size = 6))
Heatmap
```

```{r UMAP_cell_type, fig.width=6.5, fig.height=6}
p <- DimPlot(DCM_RV, group.by = "cell_type", reduction = "umap", label = F) + 
  NoLegend() + 
  theme(axis.text=element_text(size=10, face = "bold"), axis.title = element_text(size = 14, face = "bold")) + 
  theme(plot.title = element_blank())
LabelClusters(p, id = "cell_type", fontface = "bold", size = 5, repel = T)
```

```{r UMAP_cell_type_Manuscript, fig.width=6, fig.height=6}
p <- DimPlot(
        DCM_RV,
        group.by  = "cell_type",
        reduction = "umap",
        cols      = cell_type_cols,       
        label     = FALSE,
        shuffle=T,
        pt.size = 1.2
        ) +
     NoLegend() +
     theme(axis.text  = element_text(size = 8, face = "bold"),
           axis.title = element_text(size = 12, face = "bold"),
           plot.title = element_blank())

LabelClusters(p,
              id       = "cell_type",
              fontface = "bold",
              size     = 5.8,
              repel    = TRUE)
```

```{r DotPlot_markers_Manuscript, fig.width=10.5, fig.height=6}
Markers_top3 <- Markers_top3[-c(34),]
Idents(DCM_RV) <- factor(
  Idents(DCM_RV),
  levels = rev(levels(Idents(DCM_RV)))
)
p <- DotPlot_scCustom(
        DCM_RV,
        features       = Markers_top3$gene,
        x_lab_rotate   = TRUE
)

p + theme(
      axis.text.x  = element_text(size = 12,        
                                  angle = 60,          
                                  vjust = 1,
                                  hjust = 1),
      axis.text.y  = element_text(size = 15))         
```

```{r FeatPlot_markers, fig.width=18, fig.height=10}
FeaturePlot_scCustom(DCM_RV, features = c("RYR2", "COL6A3", "VWF", "RGS5", "MRC1", "PTPRC", "RGS6", "NRXN1", "PKHD1L1", "ADIPOQ", "IL18R1", "MMRN1", "POU2AF1"), num_columns = 5, order = F)
```

```{r Barplot_Sample_Manuscript, fig.height=6, fig.width=15, message=FALSE}
props <- getTransformedProps(DCM_RV$cell_type, DCM_RV$Sample, transform="logit")

par(mar = c(7, 5, 1, 20), xpd = TRUE)
barplot(props$Proportions, legend = FALSE, ylab = "Proportions", col = cell_type_cols, 
        cex.names = 1.8, las = 2, font.lab = 2, font.axis = 2, cex.axis = 1.2, cex.lab = 2)
legend("topright", inset = c(-0.375, -0.065), legend = rownames(props$Proportions), fill = cell_type_cols, bty = "n", cex = 2)
```

```{r Barplot_cell_type, fig.height=6, fig.width=12, message=FALSE}
props <- getTransformedProps(DCM_RV$Sample, DCM_RV$cell_type, transform="logit")

par(mar = c(13, 5, 1, 14), xpd = TRUE)
barplot(props$Proportions, legend = FALSE, ylab = "Proportions", col = sample_cols, 
        cex.names = 1.4, las = 2, font.lab = 2, font.axis = 2, cex.axis = 1, cex.lab = 1.4)
legend("topright", inset = c(-0.31, -0.065), legend = rownames(props$Proportions), fill = sample_cols, bty = "n", cex = 1.4, ncol = 2)
```

```{r}
dim(DCM_RV)
table(DCM_RV$cell_type)
```

```{r save DCM_RV final, eval = FALSE}
qs_save(DCM_RV,
        file = here::here(output_dir_data, "DCM_RV.qs2"),
        nthreads = 32
        )
```

```{r save DCM_RV_RNA, eval = FALSE}
DCM_RV_RNA <- DCM_RV
DCM_RV_RNA@assays[["SCT"]] <- NULL
DefaultAssay(DCM_RV_RNA) <- "RNA"
qs_save(DCM_RV_RNA,
        file = here::here(output_dir_data, "DCM_RV_RNA.qs2"),
        nthreads = 32
        )
rm(DCM_RV_RNA)
```

```{r save DCM_RV rds, eval = FALSE}
saveRDS(DCM_RV, 
        here::here(output_dir_data, "DCM_RV.rds"))
```
